### 2018-01-09 codes/R/cohorts_nc.Rnw
text file (11 items)

\documentclass[pdftex,10pt,twoside,a4paper]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amstext}% for resizing the symbol in math
\usepackage{amsthm}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[babel]{csquotes}
\usepackage{booktabs}
\usepackage{tocloft}
\usepackage{dsfont}
\usepackage{bbm}
\usepackage{xcolor}
\usepackage{accents}
\usepackage{setspace} % for spacing
\onehalfspacing
\usepackage{rotating}
\usepackage{fullpage}
\usepackage{pdflscape}  % for landscape page
\usepackage{lmodern}  %font
\usepackage{natbib}
\usepackage{array}
\usepackage{graphicx}
\usepackage{adjustbox}
\usepackage{tikz}
\usepackage{bm}
\usepackage{tabularx}
\usepackage{marvosym}
\usepackage{calc}
\usepackage{array}
\usepackage{verbatim}
\usepackage{multirow}
\usepackage[multiple]{footmisc}   % for multiple footnotes
\usepackage{textgreek}            % for greek letters in text

\setlength{\jot}{11pt} % increase vertical spacing in equations
\usepackage{txfonts}
\usepackage[T1]{fontenc}

% define new columns
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\renewcommand{\arraystretch}{1.2}%

% appendices and numeration
\usepackage[nottoc,numbib]{tocbibind} 		% bibliography in TOC
\usepackage[toc,page]{appendix} %appendix
\newcommand{\tocless}[2]{\bgroup\let\addcontentsline=\nocontentsline#1{#2}\egroup} % for empty section
\numberwithin{equation}{section} %for section eq numeration

\usepackage{hyperref}                            % references
\AtBeginDocument{%
\renewcommand\sectionautorefname{Section}        % custom autoref names
\renewcommand\subsectionautorefname{Section}
}
\newcommand{\aref}[1]{\hyperref[#1]{Appendix~\ref{#1}}} % for appendix sections
\usepackage{cleveref}

%tikz
\usepackage{tikz}
\usetikzlibrary{shapes,intersections}
\usetikzlibrary{positioning,arrows, arrows.meta}

%lists
\usepackage[inline]{enumitem}
\setlist{noitemsep}
\setlist[enumerate]{labelsep=*, leftmargin=*}
\setlist[itemize]{labelsep=*, leftmargin=*}
\usepackage[textfont={small},labelfont={bf,small},justification=centering]{caption}     % to allow multiple lines in captions, and customize the look

% page setup (from Racine)
\setlength{\topmargin}{0cm}
\setlength{\textheight}{24cm}
\setlength{\oddsidemargin}{.0cm}
\setlength{\evensidemargin}{.0cm}
\setlength{\textwidth}{16.5cm}
\setlength{\footskip}{1.5cm}
\setlength{\footnotesep}{0.4cm}
\linespread{1.3}

% additional symbols
\newcommand{\me}{\mathrm{e}}
\newcommand{\veps}{\varepsilon}
\newcommand{\E}{\mathbb{E}}
\newcommand{\1}{\mathds{1}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\Corr}{\mathrm{Corr}}
\newcommand{\vect}{\mathrm{vec}}
\newcommand{\vecht}{\mathrm{vech}}
\newcommand{\ubar}[1]{\underaccent{\bar}{#1}}
\def\chk{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}

% additional commands
\newcommand{\bi}{\begin{itemize}}
\newcommand{\ei}{\end{itemize}}
\newcommand{\bnm}{\begin{enumerate}}
\newcommand{\enm}{\end{enumerate}}

\title{IFA Comparison of Noncognitive Skills \\ BCS and MCS}
\author{Giacomo Mason \\ University College London}
\date{\today}

\begin{document}
\maketitle

<<MAIN,cache=FALSE,echo=FALSE,results='hide',message=FALSE,warning=FALSE>>=
rm(list=ls())         # Clear all objects from memory 
  opts_chunk$set(concordance=FALSE,
                 echo=FALSE,
                 tidy=FALSE,
                 autodep=TRUE,
                 cache=T, # REPLACE LATER
                 message=FALSE, 
                 warning=FALSE,
                 size='scriptsize',
                 results='hide',
                 comment=NA, 
                 fig.width=4,
                 fig.height=4, 
                 out.width='.5\\textwidth',
                 fig.show='hold',
                 tikzDefaultEngine='pdftex',
                 tikzDocumentDeclaration = "\\documentclass[10pt]{article}",
                 tikzMetricPackages = c("\\usepackage[utf8]{inputenc}","\\usepackage[T1]{fontenc}", "\\usetikzlibrary{calc}", "\\usepackage{amssymb}")
                 )

@

<<READ, cache=FALSE >>=
read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/1_load.R")
read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/2_efa.R")
read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/3_mifa_invariance.R")
read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/3_mifa_final.R")
read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/3_mifa_cog.R")
read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/4_graphs.R")

@

<<PREAMBLE, cache=F>>=
@

<<LOAD_DATA>>=
@

\section{MCS 5y vs BCS 5y Noncognitive Ability Items}

Maternal reports on child behaviours were surveyed with two different instruments in the two cohorts. For BCS, the Rutter A Scale was used \citep{Rutter1970}. The mother is presented with a list of behaviours that a child might exhibit, and is asked for each whether it 'Does not apply', it 'Somewhat applies', or it 'Certainly applies' to her child. In the MCS cohort, mothers were administered the Strengths and Difficulties Questionnaire (SDQ, \citealp{Goodman1994,Goodman1997}). The SDQ was created as an updated version of the Rutter scales, expanding it to more recent advances in child psychopathology \cite{Stone2010}. Similarly to the Rutter scale, it asks the parent to rate the child on different behaviours, with the options of 'Not true', 'Somewhat true', 'Certainly true'. Summary scores obtained from the SDQ has been shown to correlate highly with the ones from Rutter scales, but the invariance properties of the scales are sensitive to the wording of the questions \citep{Goodman2007}.

\subsection{BCS: Rutter scale}

For BCS, I augment the 20-item Rutter Scale with 4 other parent-reported questions from the parental questionnaire. 
\begin{enumerate}
\item Very restless. Often running about or jumping up and down. Hardly ever still.
\item Is squirmy or fidgety.
\item Often destroys own or others’ belongings.
\item Frequently fights other children.
\item Not much liked by other children.
\item Often worried, worries about many things
\item Tends to do things on his/her own – rather solitary.
\item Irritable. Is quick to fly off the handle.
\item Often appears miserable, unhappy, tearful or distressed.
\item Sometimes takes things belonging to others.
\item Has twitches, mannerisms or tics of the face or body.
\item Frequently sucks thumb or finger.
\item Frequently bites nails or fingers.
\item Is often disobedient.
\item Cannot settle to anything for more than a few moments.
\item Tends to be fearful or afraid of new things or new situations.
\item Is over fussy or over particular.
\item Often tells lies.
\item Bullies other children.
\item[A] Complains of headaches.
\item[B] Complains of stomach-ache or has vomited.
\item[C] Complais of biliousness.
\item[D] Has temper tantrums (that is, complete loss of temper with shouting, angry movements, etc.).
\end{enumerate}
The Rutter items are rated on three levels: Does not apply, Somewhat applies, Certainly applies. Since they are all indicators of low skills, I encode all of them in reverse, i.e. ``Certainly applies" = 0, ``Somewhat applies" = 1, ``Does not apply" = 2. The last 4 items are rated on 4 levels: Never in the last 12 months, less than once a month, at least once a month, at least once a week. I recode these into binary indicators, with \emph{Never} and \emph{Less than once a month}=1 and zero otherwise.

To increase comparability between the two cohorts, I merge together two pairs of items: 4 and 19 (to mirror the SDQ question ``Often fights with other children or bullies them"), and also A and B (to mirror the SDQ question ``Often complains of head-aches, stomach-ache or sickness"). I assign the lowest category among the two original items to the newly obtained item. I also recode items 5 and 14 to binary instead of three categories. These items are recorded with a positive phrasing in MCS, so a 3-category form would be harder to compare.

\subsection{MCS: Strengths and Difficulties Questionnare}

In MCS, I consider the 25-item strengths and difficulties questionnaire \citep{Goodman1997}.

\textbf{Hyperactivity Scale}
\begin{itemize}
\item Restless, overactive, cannot stay still for long	2
\item Constantly fidgeting or squirming	10
\item Easily distracted, concentration wanders	15
\item (+) Thinks things out before acting	21
\item (+) Sees tasks through to the end, good attention span	25
\end{itemize}
\textbf{Emotional Symptoms Scale}
\begin{itemize}
\item Often complains of head- aches, stomach-ache or sickness	3
\item Many worries, often seems worried	8
\item Often unhappy, down-hearted or tearful	13
\item Nervous or clingy in new situations, easily loses confidence	16
\item Many fears, easily scared	24
\end{itemize}
\textbf{Conduct Problems Scale}	
\begin{itemize}
\item Often has temper tantrums or hot tempers	5
\item (+) Generally obedient, usually does what adults request	7
\item Often fights with other children or bullies them	12
\item Often argumentative with adults		18
\item Can be spiteful to others		22
\end{itemize}
\textbf{Peer Problems Scale}
\begin{itemize}
\item Rather solitary, tends to play alone	6
\item (+) Has at least one good friend	11
\item (+) Generally liked by other children	14
\item Picked on or bullied by other children	19
\item Gets on better with adults than with other children	23
\end{itemize}
\textbf{Prosocial Scale}
\begin{itemize}
\item (+) Considerate of other people's feelings	1
\item (+) Shares readily with other children (treats, toys, pencils, etc.)	4
\item (+) Helpful if someone is hurt, upset or feeling ill	9
\item (+) Kind to younger children	17
\item (+) Often volunteers to help others (parents, teachers, other children)	20
\end{itemize}

All items are recorded on a 4-point scale: Not true, Somewhat true, Certainly true, Can't say. I set the latter to missing and recode the rest in ascending order of skill as for the BCS items, i.e. ``Certainly true" = 0, ``Somewhat true" = 1, ``Not true" = 2. For comparability with BCS, I retain items 3, 5, 7, and 14 in a binary format.

\subsection{Exploratory Analysis}

Our aim is to assess whether the two scales can be used to derive measures of noncognitive ability that are invariant, and can therefore be compared across cohorts. In the first step of our analysis, we consider a subset of each scale's items, selecting only those that present very similar wording across scales. The selection process results in a new 11-item scale, reported in \autoref{tab:comm} with the original wording and the prevalence of each option. As highlighted above, some of the items are coded differently in the two original scales and we thus turn them into binary dummies to ensure comparability.

The second step to our invariance analysis seeks to justify a common latent factor structure across the two cohorts. The Rutter scale is constructed to measure two latent constructs: anti-social (externalising problems), and neurotic (internalising problems) behaviours. To validate these constructs in our new scale, we carry out exploratory factor analysis based on the matrix of pairwise tetrachoric correlations between the item responses. Various methods to assess the optimal number of factors underlying these items give inconsistent results, ranging from 1 (Velicer MAP) to 7 or even 8 (BIC). I choose to conduct analyses with two factors as suggested by the theory and by the Very Simple Structure method.

For ease of interpretation, I choose to rescale the factor loadings using \emph{oblimin} rotation: this method allows the underlying factors to be correlated to each other. From the results of the exploratory FA and IRT analysis with 2 factors, the pattern that seems to emerge is consistent with the theoretical constructs of externalising (EXT) and internalising (INT) behaviour problems.

<<BCSEFA>>=
@

<<BCSEFARES, dependson='BCSEFA', echo=T, results='markup'>>=
@

The five-factor structure at the basis of the full SDQ has been validated in many contexts \citep{Stone2010}; however, lower-dimensional structures have been also suggested \citep{Dickey2004}, and recent research has shown that there are some benefits to using broader subscales that correspond to the externalising and internalising factors in Rutter \citep{Goodman2010}. To assess whether the two-factor conceptualisation suggested by the theory is valid in our ``reduced" 12-item scale, we conduct exploratory factor analysis (EFA) on the new 11-item scale separately for each cohort sample. Different methodologies for quantifying the optimal number of factors produce inconclusive results, but both Velicer MAP and very simple structure (VSS) seem to indicate that a two-factor structure is suitable. An EFA model with two factors, employing the usual cutoff of .3 for the loadings, delivers a neat and sensible separation between items, as shown in the second column of \autoref{tab:comm}. Thus, our two-factor structure seems justifiable on both theoretical and empirical grounds.

<<MCSEFA>>=
@

<<MCSEFARES, dependson='BCSEFA', echo=T, results='markup'>>=
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage

<<MEANTABLE>>=
# make table with means
@

\clearpage
\begin{landscape}
\begin{table}[ht]
\centering
\footnotesize
\caption{Item Commonalities}\label{tab:comm}
<<MEANTABLE_RES, dependson="LOAD_DATA", results='asis'>>=
require(xtable)

table <- meantab

print(xtable(table,
                                  # item num        Factor          Num cats        BCS text        BCS num           BCS Certain     BCS Somewhat    BCS Applies             MCS text          MCS num         
                    align="L{0cm}C{.04\\textwidth}C{.05\\textwidth}C{.05\\textwidth}L{.3\\textwidth}C{.03\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}L{.3\\textwidth}C{.03\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
                    digits = 1
                    ), 
             size="\\small",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0)),
                              command = c(
                              # header
                                    paste(" & & &
                                              \\multicolumn{5}{c}{\\textbf{BCS}} &
                                              \\multicolumn{5}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){4-8}  \\cmidrule(lr){9-13} 
                                            Item & Factor & N. categ &
                                              Text & & Cert. Applies & Somet. Applies & Applies &
                                              Text & & Cert. True & Somet. True & True \\\\ \\midrule
                                          "
                                            , sep="")
                                    )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabular", width = ".8\\textwidth",
             floating = FALSE
)

@
\end{table}
\end{landscape}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Measurement Invariance}

\subsubsection{Model notation}
Before any meaningful comparison across cohorts can be made, measurement invariance (MI) needs to be established. Examination of the invariance properties of a measurement system is aimed at assessing whether the latent constructs in each group are measured in a comparable way, both in terms of scale and location. In the case of continuous measures, MI analysis proceeds in a relatively straightforward way \citep{Vanderberg2000a}; MI properties in systems of categorical outcomes are instead more problematic, due to the more complex nature of identification \citep{Millsap2004, Wu2016}.

I set up a (confirmatory) item factor analysis (IFA) model for the full list of 11 common items. Given the results of the exploratory analysis -- where each measure separates neatly on one of the two factors -- I assume a \emph{dedicated} factor structure \citep{Carneiro2003}, where each measure is restricted to load on one of the two factors only.\footnote{This structure is commonly assumed in economic research to enable a more straightforward interpretation of the latent constructs in question \citep{ACFMR,Heckman2012a}. It can also be denominated \emph{congeneric} factor stucture \citep{Millsap2004}.}

To account for the ordered nature of the measures, I specify a threshold model \citep{Muthen1984}. Individual $j=1\dots N_c$ in group $c=\{A,B\}$ is assumed to have a latent continuous propensity $X^*_{ijc}$ for each item $i=1,\dots,I$. In our application, $j$ indexes the child, the group $c$ corresponds to the cohort, and $i$ indexes each behaviour in the 11-item scale summarised in \autoref{tab:comm}. Thus, $X^*_{ijc}$ is the child's propensity to exhibit each of the Rutter/SDQ behaviours.
This propensity is modelled as a function of item-specific intercepts $\nu$ and loadings $\lambda$, and the individual's level on a latent construct vector $\bm{\theta}$, plus an independent error component $u_{cij}$. In our application, the latent factor is bidimensional, and corresponds to the internalising ($\theta^{I}$) and externalising ($\theta^{E}$) traits. The dedicated factors structure implies that each item is assumed to only depend on a single latent dimensions, as delineated in the exploratory factor analysis in \autoref{tab:comm}.

The propensity for each item can be written as follows:

\begin{align}\label{eq:propens}
\begin{split}
X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{EXT}_{cj} + u_{cji} \qquad \text{for} \; i=1,\dots, 6 \\
X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{INT}_{cj} + u_{cji} \qquad \text{for} \; i=7,\dots, 11
\end{split}
\end{align}

or more compactly
$$X^*_{jc} =\bm{\nu}_c + \bm{\Lambda}_{c}\bm{\theta}_{jc} + \bm{u}_{jc}$$

with the additional assumptions
$$\bm{\Theta}_{jc} \sim N(\bm{\kappa}_c, \bm{\Phi}_c) \qquad \text{and} \qquad \bm{u}_{jc} \sim N(\bm{0}, \bm{\Psi}_c)$$

The model implies the following expression for the mean and covariance structure of the latent propensities:
$$\bm{\mu}_c=\bm{\nu}_c + \bm{\Lambda}_c\bm{\kappa}_c \qquad \text{and} \qquad \Sigma_c = \bm{\Lambda}_c \bm{\Phi}_c \bm{\Lambda}_c^\prime + \bm{\Psi}_c.$$

For each item $i$, we observe either a binary or ordinal (3-categories) manifestation $X_{cji}$. Given \eqref{eq:propens}, the observed outcome as a function of the propensities $X^*$ can be defined as a function of additional threshold parameters $\tau$, as follows:
$$X_{cji} = s \qquad \text{if} \; \tau_{s,ci} \leq X^*_{cji} < \tau_{s+1,ci} \qquad \text{for} \; s=0,1,2$$
with $\tau_{0,ci}=-\infty$ and $\tau_{3,ci}=+\infty$.
The probability of observing each category for a binary item takes values:
\begin{equation}
X_{cji}=
\begin{cases}
\qquad 0 \qquad \text{if} \; X^*_{cji}<\tau_{1,ci} \\
\qquad 1 \qquad \text{if} \; X^*_{cji}\geq\tau_{1,ci}
\end{cases}
\end{equation}

An ordinal item takes values:\footnote{As a consequence of the normality assumption on the error terms $\veps$, the probability that individual $j$ in group $c$ exhibits item $i$ with ``intensity" $s$ is:
\begin{align*}
\Pr(X_{cji}=0) &= 1- \Phi(-\tau_{1,ci} + \nu_{ci} + \lambda_{ci}\theta^{EXT}_{j}) \\
\Pr(X_{cji}=1) &= \Phi(-\tau_{1,ci} + \nu_{ci} + \lambda_{ci}\theta^{EXT}_{j})
\end{align*}
for binary items and
\begin{align*}
\Pr(X_{cji}=0) &= 1- \Phi(-\tau_{1,ci} + \nu_{ci} + \lambda_{ci}\theta^{EXT}_{j}) \\
\Pr(X_{cji}=1) &= \Phi(-\tau_{1,ci} + \nu_{ci} + \lambda_{ci}\theta^{EXT}_{j}) - \Phi(-\tau_{2,ci} + \nu_{ci} + \lambda_{ci}\theta^{EXT}_{j})\\
\Pr(X_{cji}=2) &= \Phi(-\tau_{2,ci} +  \nu_{ci} + \lambda_{ci}\theta^{EXT}_{j})
\end{align*}
for ordinal items. This IFA model can also be written as a 2-parameter Logit IRT model with item difficulty $b$ and item discrimination $a$. In the simplest case for a binary item $1$:
$$\text{Probit}(X_{cj1}=1) = a_{c1}(\theta^{EXT}_j - b_{c1})$$
such that $\tau_{c1}=a_{c1}b_{c1}$ and $a_{c1}=\lambda_{c1}$.}

\begin{equation}
X_{cji}=
\begin{cases}
\qquad 0 \qquad \text{if} \; X^*_{cji}<\tau_{1,ci} \\
\qquad 1 \qquad \text{if} \; \tau_{1,ci} \leq X^*_{cji}<\tau_{2,ci} \\
\qquad 2 \qquad \text{if} \; X^*_{cji}\geq\tau_{2,ci}
\end{cases}
\end{equation}

\subsubsection{Invariance analysis}
There are infinite ways in which the above system can be parameterised and a scale obtained -- the well-known issue of factor indeterminacy. A configural model is defined as a minimal set of parameter restrictions that ensure identification of the system across groups, without restricting the model further. Widely used parameterisations for the configural model are:
\bnm[label=\Alph*:]
\item Delta parameterisation {[WE{\textDelta}]} \citep{Wu2016}

For all groups: $$\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}, \quad \bm{\nu}=\bm{0}, \quad \text{and} \; \text{diag}(\bm{\Sigma})=\bm{I}.$$

\item Theta parameterisation {[WE{\textTheta}]} \citep{Wu2016}

For all groups: $$\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}, \quad \bm{\nu}=\bm{0}, \quad \text{and} \; \text{diag}\bm{\Psi}=\bm{I}.$$

\item Anchored parameterisation {[MT]} \citep{Millsap2004}
  \bi
  \item For all groups, normalise a reference loading to 1 for each factor
  \item Set invariant across groups one threshold per item (e.g. $\tau_{0,Ai} = \tau_{0,Bi}$), and an additional threshold in the reference items above
  \item In the first group: $\bm{\kappa_A}=\bm{0}$, $\text{diag}(\bm{\Sigma}_A)=\bm{I}$
  \item Set all intercepts $\bm{\nu}$ to zero
  \ei
\enm
The first two parameterisations (WE{\textDelta} and WE{\textTheta}) work by normalising the mean and variance of factors to the same constants in both groups, and they leave all loadings and intercepts to be freely estimated. The MT parameterisation instead proceeds by first identifying one group, and then imposing cross-group equality constraints to identify other groups \citep{Wu2016}.

To test measurement invariance, one needs to estimate nested models that place increasing restrictions on the item parameters across groups $c$. In the case of continuous measures, the usual approach is to restrict one set of parameters at a time. As an example, starting from the configural model, one would restrict all loadings to be equal across groups; if the fit of this restricted model is not worse than the configural model, \emph{metric invariance} is estabilished \citep{Vanderberg2000a}. The hierarchy of the nested models usually proceeds by testing loadings first, and then intercepts (metric and scalar invariance). Metric and scalar invariance are the minimum requirements to be able to meaningfully compare means and variances of the latent factors across groups.

A substantially identical approach is proposed for categorical measures by \cite{Millsap2004}. However, new identification results in \cite{Wu2016} indicate that, in the case of categorical measures, invariance properties cannot be examined by simply restricting one set of parameters at a time. This is because the identification conditions used in the configural baseline model, while being minimally restrictive on their own, become restrictive once certain additional restrictions are imposed. In light of this, they propose sets of restrictions that identify structures of different invariance levels. They find that some restrictions cannot be tested alone against the configural model, because they give rise to statistically equivalent models. This is true of loading invariance, and also of threshold invariance in the case when the number of categories of each ordinal item is 3 or less. Furthermore, they suggest that comparison of both latent means and variances requires invariance in loadings, thresholds, and intercepts.

Following the ordering in \cite{Wu2016}, I estimate the following increasingly restrictive models:
\bi
\item \textbf{Configural} model -- see WE{\textDelta} above
\item \textbf{Threshold invariance} model
  \bi
  \item $\tau_{1,Ai} = \tau_{1,Bi}$ for $i=1,\dots,11$, $\tau_{2,Ai} = \tau_{2,Bi}$ for $i=\{1,2,3,4,7,8,9,10\}$
  \item For all groups: $$\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}$$
  \item For the reference group: $$\bm{\nu}_A=\bm{0}, \quad \; \text{diag}(\bm{\Sigma}_A)=\bm{I}$$
  \item In our case, the number of categories in the outcomes is 3, thus this model is statistically equivalent to the configural model
  \ei
\item \textbf{Threshold and Loading invariance} model
  \bi
  \item $\tau_{1,Ai} = \tau_{1,Bi}$ for $i=1,\dots,11$, $\tau_{2,Ai} = \tau_{2,Bi}$ for $i=\{1,2,3,4,7,8,9,10\}$
  \item $\lambda_{Ai} = \lambda_{Bi}$ for $i=1,\dots,11$
  \item For all groups: $$\bm{\kappa}=\bm{0}$$
  \item For the reference group: $$\bm{\nu}_A=\bm{0}, \quad \; \text{diag}(\bm{\Sigma}_A)=\bm{I}, \quad \text{diag}(\bm{\Phi}_A) = \bm{I}$$
  \item This set of restriction allows comparison of latent factor variances
  \ei
\item \textbf{Threshold, Loading, and Intercept invariance} model
  \bi
  \item $\tau_{1,Ai} = \tau_{1,Bi}$ for $i=1,\dots,11$, $\tau_{2,Ai} = \tau_{2,Bi}$ for $i=\{1,2,3,4,7,8,9,10\}$
  \item $\lambda_{Ai} = \lambda_{Bi}$ for $i=1,\dots,11$
  \item For all groups: $$\bm{\nu}=\bm{0}$$
  \item For the reference group: $$\bm{\kappa}_A=\bm{0}, \quad \; \text{diag}(\bm{\Sigma}_A)=\bm{I}, \quad \text{diag}(\bm{\Phi}_A) = \bm{I}$$
  \item This set of restriction allows comparison of latent factor means and variances
  \ei
\ei

<<FA_INVARIANCE, dependson='LOAD_DATA'>>=
@

<<FA_FIT, dependson='FA_INVARIANCE'>>=
@

\begin{table}[ht]
\centering
\small
\caption{Fit indices}\label{tab:fit}
<<FA_FIT_RES, dependson="FA_FIT", results='asis'>>=
require(xtable)

table <- afitab

print(xtable(table,
                    align="L{0cm}L{.18\\textwidth}C{.05\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
                    digits = 4
                    ),
             size="\\small",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0)),
                              command = c(
                              # header
                                    paste("
                                            Model & Num. params & $\\chi^2$ & RMSEA & MFI & CFI & G-hat & \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\
                                          "
                                            , sep="")
                                    )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabular", width = "\\textwidth",
             floating = FALSE
)

@
\end{table}

I estimate each of the models listed above by Weighted Least Squares.\footnote{All estimates are computed using the \texttt{lavaan} package in \textbf{\textsf{R}} \citep{Rosseel2012}. Parameters of the model are estimated by mean- and variance adjusted weighted least squares (WLSMV), due to \cite{Mplus2017}. Point estimated for model parameters are obtained starting from the polychoric correlations, using diagonally weighted least squares (DWLS), and the full weight matrix is used to compute robust standard errors and test statistics. Robust WLS has proved in simulation studies to be moderately robust to small violations of the normality assumption in the latent underlying measures \citep{Flora2011}, and generally outperforms maximum likelihood in large samples \citep{Beauducel2009,Li2016}.} The fit of the increasingly restrictive models presented above is compared in \autoref{tab:fit}. For all models, a chi-squared difference would point to a lack of measurement invariance. However, this test is know to overestimate invariance in large samples. Alternative fit indices should thus be used when comparing nested models. \cite{Cheung2002} suggest relying on Comparative Fit Index (CFI, \citealp{Bentler1990}), McDonald Non-Centrality Index (MFI, \citealp{McDonald1989}), and Gamma Hat index \citep{Steiger1989} for the case of ordered measures. Commonly accepted thresholds for rejecting the restricted model are $\Delta CFI < - 0.01$, $\Delta MFI < - 0.02$, and $\Delta Gamma hat < - 0.001$.

The overall fit of the configural model is satisfactory according to widely accepted rules of thumb \citep{Maccallum1996}.\footnote{{\color{red} BETTER CITATIONS FOR OVERALL FIT}.} As discussed above, when items have less than four categories, the model with restricted thresholds is statistically equivalent to the configural model. Restricting thresholds and loadings seems to provide a satisfying fit compared to the configural model.\footnote{\cite{Meade2008}, using the results from a simulation study, suggests stricter thresholds that should apply in a variety of conditions. For CFI, a single cutoff values of $.002$ is proposed, while cutoffs for MFI depend on the problem's characteristics; in our case (2 factors, 11 items), they suggest $.0066$. \cite{Sass2014} however cast some doubts of the generalisability of these cutoffs to WLSMV estimators.} However, the model that additionally restricts intercepts -- thus allowing for latent mean comparison -- has a substantially worse fit.

<<FA_FINAL, dependson='FA_PARTINV'>>=
@

<<FA_FINAL_PARS, dependson='FA_FINAL'>>=
@

\begin{table}[ht]
\caption{Estimated loadings and thresholds} \label{tab:allpars}
\resizebox{\textwidth}{!}{
<<RES_FA_PAR, dependson="FA_FINAL_PARS", results='asis'>>=
require(xtable)

table <- allpar

# sequences to add rows
rws <- seq(2, (nrow(table)-1), by = 1) # start from row 2 (exclude model titles)
rwsl <- as.list(rws)

print.xtable(xtable(table,
                    # align = "ll|cccc|cccc"
                    align="L{0cm}L{2cm}C{2.5cm}C{2.5cm}C{2.5cm}C{2.5cm}C{1.5cm}",
                    digits = 3
                    ),
             size="\\small",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0)),
                              command = c(
                              # header
                                    paste(" & Loadings
                                            & \\multicolumn{2}{c}{\\textbf{Thresholds}}
                                            & Intercepts
                                            & \\\\ \\cmidrule(lr){3-4}

                                            & $\\bm{\\lambda}$
                                            & $\\bm{\\tau}_1$
                                            & $\\bm{\\tau}_2$
                                            & $\\bm{\\nu}$ \\\\
                                            Measure & BCS = MCS
                                            & BCS = MCS
                                            & BCS = MCS
                                            & MCS (BCS=0)
                                            & Factor
                                            \\\\", sep=""))),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabularx", width = "\\textwidth",
             floating = FALSE
)

@
}
\end{table}

\begin{table}[ht]
\caption{Estimated latent variable means and covariances} \label{tab:lvpars}
\centering
<<RES_FA_LV, dependson="FA_FINAL_PARS", results='asis'>>=
require(xtable)

table <- lvpars

print.xtable(xtable(table,
                    align = "llccc",
                    # align="L{0cm}L{.1\\textwidth}L{.2\\textwidth}L{.15\\textwidth}L{.15\\textwidth}L{.15\\textwidth}L{.15\\textwidth}",
                    digits = 3
                    ),
             size="\\small",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0,2)),
                              command = c(
                              # header
                                    paste(" & \\multicolumn{3}{c}{\\textbf{BCS}} \\\\ \\midrule
                                            & \\multicolumn{1}{c}{\\textbf{Mean}}
                                            & \\multicolumn{2}{c}{\\textbf{Covariance}} \\\\
                                            && $\\theta^{EXT}$ & $\\theta^{INT}$ \\\\", sep=""),
                                    paste(" \\midrule & \\multicolumn{3}{c}{\\textbf{MCS}} \\\\ \\midrule
                                            & \\multicolumn{1}{c}{\\textbf{Mean}}
                                            & \\multicolumn{2}{c}{\\textbf{Covariance}} \\\\
                                            && $\\theta^{EXT}$ & $\\theta^{INT}$ \\\\", sep="")
                                    )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabularx", width = ".5\\textwidth",
             floating = FALSE
)

@
\end{table}

\begin{table}[ht]
\caption{Estimated latent variable means and covariances -- model with restricted intercepts} \label{tab:lvpars2}
\centering
<<RES_FA_LV2, dependson="FA_FINAL_PARS", results='asis'>>=
require(xtable)

table <- lvpars2

print.xtable(xtable(table,
                    align = "llccc",
                    # align="L{0cm}L{.1\\textwidth}L{.2\\textwidth}L{.15\\textwidth}L{.15\\textwidth}L{.15\\textwidth}L{.15\\textwidth}",
                    digits = 3
                    ),
             size="\\small",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0,2)),
                              command = c(
                              # header
                                    paste(" & \\multicolumn{3}{c}{\\textbf{BCS}} \\\\ \\midrule
                                            & \\multicolumn{1}{c}{\\textbf{Mean}}
                                            & \\multicolumn{2}{c}{\\textbf{Covariance}} \\\\
                                            && $\\theta^{EXT}$ & $\\theta^{INT}$ \\\\", sep=""),
                                    paste(" \\midrule & \\multicolumn{3}{c}{\\textbf{MCS}} \\\\ \\midrule
                                            & \\multicolumn{1}{c}{\\textbf{Mean}}
                                            & \\multicolumn{2}{c}{\\textbf{Covariance}} \\\\
                                            && $\\theta^{EXT}$ & $\\theta^{INT}$ \\\\", sep="")
                                    )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabularx", width = ".5\\textwidth",
             floating = FALSE
)

@
\end{table}

<<CLEANUP, dependson='FA_FINAL'>>=
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
\subsection{Inequality in noncognitive ability}
Although the results of our measurement invariance analysis imply that comparison of latent means is not justified, the model with restricted loadings and thresholds can nevertheless be used to investigate how inequality of noncognitive skills has changed between the two cohorts. The variances of the $EXT$ and $INT$ factors are estimated to be larger in the MCS children by \Sexpr{round(lv.covs$MCS[1,1]-1,2)} and \Sexpr{round(lv.covs$MCS[2,2]-1,2)} of a standard deviation, respectively, indicating larger dispersion.

<<FA_SCORES, dependson='FA_FINAL'>>=
@

<<FA_REGS, dependson='FA_SCORES'>>=

reg.bcs.ext <- lm(EXT~as.factor(incq), data=subset(facandraw, cohort=="BCS"))
bcs.51.ext <- round(reg.bcs.ext$coefficients["as.factor(incq)5"],2)
reg.bcs.int <- lm(INT~as.factor(incq), data=subset(facandraw, cohort=="BCS"))
bcs.51.int <- round(reg.bcs.int$coefficients["as.factor(incq)5"],2)

reg.mcs.ext <- lm(EXT~as.factor(incq), data=subset(facandraw, cohort=="MCS"))
mcs.51.ext <- round(reg.mcs.ext$coefficients["as.factor(incq)5"],2)
reg.mcs.int <- lm(INT~as.factor(incq), data=subset(facandraw, cohort=="MCS"))
mcs.51.int <- round(reg.mcs.int$coefficients["as.factor(incq)5"],2)

@

We can derive scores for $\theta^{EXT}$ and $\theta^{INT}$ directly, and plot them against socioeconomic measures observed in the data. Our main socioeconomic indicator is a measure of weekly total family net income in the prices of the interview year.\footnote{This measure is harmonised across cohorts and derived as part of CLOSER, which is funded by the ESRC.} Using the estimated parameters for the model with invariant thresholds and loadings (but non-invariant intercepts), I compute the scores for each individual in the sample (using the Empirical Bayes modal approach). In \autoref{fig:boxscores}, we can see that the gradient in noncognitive skills along family income quintiles is more pronounced in the MCS cohort. In particular, a five-year old at the top quintile of the income distribution in 1975 has a better score by \Sexpr{bcs.51.ext} in the externalising dimension and \Sexpr{bcs.51.int} in the internalising dimension, compared to a child at the bottom quintile. In 2005, the differences amount to \Sexpr{mcs.51.ext} and \Sexpr{mcs.51.int}, respectively.

\clearpage
\begin{figure}
<<FACDENS, dev='tikz', fig.height=5, fig.width=10, out.width='1\\linewidth'>>=
@
\caption{Density of Scored Factors}
{\scriptsize \emph{Notes}: The figure shows estimated kernel densities of the scored factors from model \eqref{eq:finalmod}, separately by cohort.  \par}
\end{figure}

\begin{figure}
<<FACRAW, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
@
\caption{Raw vs Scored Factors}
{\scriptsize \emph{Notes}: The figure plots the scored factors from model \eqref{eq:finalmod} on the vertical axis against the ``raw" scores on the horizontal axis, i.e. scores obtained by simply summing the number of positive answers to each of the items. \par}
\end{figure}

\begin{figure}
<<FACSCAT, dev='tikz', fig.height=5, fig.width=10, out.width='1\\linewidth'>>=
@
\caption{Scatterplots of factor scores by cohort}
{\scriptsize \emph{Notes}: The figure plots the scored factors from model \eqref{eq:finalmod} separately by cohort. The externalising and internalising scores are plotted against one another. \par}
\end{figure}

\begin{figure}
<<FACINEQ, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
@
\caption{Distribution of scores across income quintiles}\label{fig:boxscores}
{\scriptsize \emph{Notes}: The figure plots the distribution of the scored factors from model \eqref{eq:finalmod} across the quintiles in the income distribution. \par}
\end{figure}

\begin{figure}
<<FACLOESS, dev='tikz', fig.height=5, fig.width=10, out.width='1\\linewidth'>>=
@
\caption{Scores and family income}
{\scriptsize \emph{Notes}: The figure plots LOESS of the scored factors from model \eqref{eq:finalmod} against family income. \par}
\end{figure}

\clearpage
\bibliographystyle{chicago}
\bibliography{/Users/giacomomason/Documents/BibTeX/Papers-cohorts_nc}

\end{document}

### 2018-01-24 codes/R/cohorts_nc.Rnw
- add age-sex adjustment as regression - modify plots to display by gender - separate scores from graphs into different scripts - adapt tables and graphs in Rnw NOTE: text in Rnw NOT updated

@@
 read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/3_mifa_final.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/3_mifa_cog.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/4_graphs.R")
+read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/4_scores.R")
+read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/5_graphs.R")
 
@@
                                             & $\\bm{\\tau}_2$
-                                            & $\\bm{\\nu}$ \\\\
+                                            & $\\bm{\\varepsilon}$ \\\\
                                             Measure & BCS = MCS
@@
                                             & BCS = MCS
-                                            & MCS (BCS=0)
+                                            & MCS (BCS=1)
                                             & Factor
@@
 print.xtable(xtable(table,
-                    align = "llccc",
-                    # align="L{0cm}L{.1\\textwidth}L{.2\\textwidth}L{.15\\textwidth}L{.15\\textwidth}L{.15\\textwidth}L{.15\\textwidth}",
+                    align="L{0cm}L{.07\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.07\\textwidth}",
                     digits = 3
@@
              booktabs = T,
-             add.to.row = list(pos = c(list(0,2)),
+             add.to.row = list(pos = c(list(0)),
                               command = c(
                               # header
-                                    paste(" & \\multicolumn{3}{c}{\\textbf{BCS}} \\\\ \\midrule
-                                            & \\multicolumn{1}{c}{\\textbf{Mean}}
-                                            & \\multicolumn{2}{c}{\\textbf{Covariance}} \\\\
-                                            && $\\theta^{EXT}$ & $\\theta^{INT}$ \\\\", sep=""),
-                                    paste(" \\midrule & \\multicolumn{3}{c}{\\textbf{MCS}} \\\\ \\midrule
-                                            & \\multicolumn{1}{c}{\\textbf{Mean}}
-                                            & \\multicolumn{2}{c}{\\textbf{Covariance}} \\\\
-                                            && $\\theta^{EXT}$ & $\\theta^{INT}$ \\\\", sep="")
+                                    paste(" & \\multicolumn{4}{c}{\\textbf{BCS}}
+                                            & \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
+                                            & Mean & \\multicolumn{2}{c}{Covariance} & Corr. & Mean & \\multicolumn{2}{c}{Covariance} & Corr. \\\\
+                                            ", sep="")
                                     )),
@@
              sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabularx", width = ".5\\textwidth",
-             floating = FALSE
-)
-
-@
-\end{table}
-
-\begin{table}[ht]
-\caption{Estimated latent variable means and covariances -- model with restricted intercepts} \label{tab:lvpars2}
-\centering
-<<RES_FA_LV2, dependson="FA_FINAL_PARS", results='asis'>>=
-require(xtable)
-
-table <- lvpars2
-
-print.xtable(xtable(table,
-                    align = "llccc",
-                    # align="L{0cm}L{.1\\textwidth}L{.2\\textwidth}L{.15\\textwidth}L{.15\\textwidth}L{.15\\textwidth}L{.15\\textwidth}",
-                    digits = 3
-                    ),
-             size="\\small",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = c(list(0,2)),
-                              command = c(
-                              # header
-                                    paste(" & \\multicolumn{3}{c}{\\textbf{BCS}} \\\\ \\midrule
-                                            & \\multicolumn{1}{c}{\\textbf{Mean}}
-                                            & \\multicolumn{2}{c}{\\textbf{Covariance}} \\\\
-                                            && $\\theta^{EXT}$ & $\\theta^{INT}$ \\\\", sep=""),
-                                    paste(" \\midrule & \\multicolumn{3}{c}{\\textbf{MCS}} \\\\ \\midrule
-                                            & \\multicolumn{1}{c}{\\textbf{Mean}}
-                                            & \\multicolumn{2}{c}{\\textbf{Covariance}} \\\\
-                                            && $\\theta^{EXT}$ & $\\theta^{INT}$ \\\\", sep="")
-                                    )),
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabularx", width = ".5\\textwidth",
+             tabular.environment = "tabularx", width = ".85\\textwidth",
              floating = FALSE
@@
 
-reg.bcs.ext <- lm(EXT~as.factor(incq), data=subset(facandraw, cohort=="BCS"))
+reg.bcs.ext <- lm(EXT~as.factor(incq) + age, data=subset(items.scored, cohort=="BCS"))
 bcs.51.ext <- round(reg.bcs.ext$coefficients["as.factor(incq)5"],2)
-reg.bcs.int <- lm(INT~as.factor(incq), data=subset(facandraw, cohort=="BCS"))
+reg.bcs.int <- lm(INT~as.factor(incq) + age, data=subset(items.scored, cohort=="BCS"))
 bcs.51.int <- round(reg.bcs.int$coefficients["as.factor(incq)5"],2)
 
-reg.mcs.ext <- lm(EXT~as.factor(incq), data=subset(facandraw, cohort=="MCS"))
+reg.mcs.ext <- lm(EXT~as.factor(incq) + age, data=subset(items.scored, cohort=="MCS"))
 mcs.51.ext <- round(reg.mcs.ext$coefficients["as.factor(incq)5"],2)
-reg.mcs.int <- lm(INT~as.factor(incq), data=subset(facandraw, cohort=="MCS"))
+reg.mcs.int <- lm(INT~as.factor(incq) + age, data=subset(items.scored, cohort=="MCS"))
 mcs.51.int <- round(reg.mcs.int$coefficients["as.factor(incq)5"],2)
@@
 \begin{figure}
-<<FACINEQ, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
+<<FACINEQ, dev='tikz', fig.height=12, fig.width=10, out.width='1\\linewidth'>>=
 @

### 2018-01-24 codes/R/1_load.R

@@ -71,7 +71,7 @@ for (i in 1:ncol(bcs5rutc)) bcs5rutcf[,i] <- as.ordered(bcs5rutc[,i])
 
 
 # auxiliary variables
-bcs5aux <- cbind(bcs5data[,c("bcsid", "faminc", "incq", "ysch_moth", "ysch_fath", "ageint5")],
+bcs5aux <- cbind(bcs5data[,c("bcsid", "faminc", "incq", "ysch_moth", "ysch_fath", "ageint5", "sex")],
                 cog1 = bcs5data$epvt_z,
                 cog2 = bcs5data$hfd_z,
                 cog3 = bcs5data$copy_z
@@ -114,7 +114,7 @@ mcs5sdqcf <- mcs5sdqc
 for (i in 1:ncol(mcs5sdqc)) mcs5sdqcf[,i] <- as.ordered(mcs5sdqc[,i])                    
 
 # add SES and cognitive data
-mcs5aux <- cbind(mcs5data[,c("mcsid", "faminc", "incq", "ysch_moth", "ysch_fath", "ageint5")],
+mcs5aux <- cbind(mcs5data[,c("mcsid", "faminc", "incq", "ysch_moth", "ysch_fath", "ageint5","sex")],
                 cog1 = mcs5data$nvoc_bastz,
                 cog2 = mcs5data$psim_bastz,
                 cog3 = mcs5data$patc_bastz
@@ -146,10 +146,19 @@ items.c <- X[,c(grep("X", names(X), value=T), "cohort")]   # noncog items only
 items  <- items.c                              # numeric version of noncog items only
 for (i in 1:ncol(items.c)) items[,i] <- as.numeric(items.c[,i]) - 1
 
+# with age and sex
+items.cb <- X[,c(grep("X", names(X), value=T), "cohort", "ageint5","sex")]
+colnames(items.cb)[colnames(items.cb) == "ageint5"] <- "age"
+items.cb$sex <- factor(items.cb$sex)
 
-# ADD cognitive measures
-itemsa.c <- X[,c(grep("X", names(X), value=T), grep("cog", names(X), value=T), "cohort")]   # include cognitive measures
-
+# with age, cohort+sex group
+items.cc <- X[,c(grep("X", names(X), value=T), "cohort", "ageint5", "sex")]
+colnames(items.cc)[colnames(items.cc) == "ageint5"] <- "age"
[…]

### 2018-01-24 codes/R/3_mifa_final.R

@@ -1,14 +1,13 @@
 # ############################################################################################
 ## ---- FA_FINAL
-# select final model
 
 # select final model
-finalmod <- fa.thr.load.we.th
-pta <- partable(finalmod)
+finalmod <- fa.thr.load.int.we.th
 
 ##################################################################################
 ## ---- FA_FINAL_PARS
-# for model with constrained loadings and thresholds
+# extract parameters and make table for final model estimates
+pta <- partable(finalmod)
 
 # item parameters ----------------------------------------------------------
 # extract loadings
@@ -32,17 +31,17 @@ tau2 <- BCS.tau2
 tau2$measure <- tau2$lhs
 tau2 <- tau2[ , -which(names(tau2) %in% c("rhs","lhs"))] # drop useless columns
 
-# extract intercepts (group 2 only)
-MCS.nu = pta[(pta$op == "~1" & pta$rhs == "" & pta$group == 2),c("lhs", "rhs","est")]
-nu <- MCS.nu
-nu$measure <- nu$lhs
-nu <- nu[ , -which(names(nu) %in% c("rhs","lhs"))] # drop useless columns
+# extract variances (group 2 only)
+MCS.eps = pta[(pta$op == "~~" & substr(pta$rhs,1,1) == "X" & pta$group == 2),c("lhs", "rhs","est")]
+eps <- MCS.eps
+eps$measure <- eps$lhs
+eps <- eps[ , -which(names(eps) %in% c("rhs","lhs"))] # drop useless columns
 
 
 # merge parameters together
 allpar <- merge(lambdas,tau1,by="measure", all.x = T)
 allpar <- merge(allpar,tau2,by="measure", all.x = T)
-allpar <- merge(allpar,nu,by="measure", all.x = T)
+allpar <- merge(allpar,eps,by="measure", all.x = T)
 allpar <- allpar[mixedorder(allpar$measure),]
 
 # name variables
@@ -50,7 +49,7 @@ colnames(allpar) <- c("measure",
                       "$\\lambda_{BCS} = \\lambda_{MCS}$",
[…]

### 2018-01-24 codes/R/3c_mifa_invariance.R

@@ -1,7 +1,97 @@
 
-
 ############################################################################################
 ## ---- FA_INVARIANCE
+
+# MT CONFIGURAL - THETA PARAMETERISATION
+# - one anchor loadings = 1 for each factor
+# - all intercepts = 0
+# - factor means = 0 in group 1 only
+# - thresholds: tau1 restricted equal across groups, tau2 invariant only for anchors
+
+config.mt.th <- "
+EXT =~ c(1, 1)*X2 + X1 + X3 + X4 + X5 + X6
+INT =~ c(1, 1)*X8 + X7 + X9 + X10 + X11
+X2 | c(t1_2, t1_2)*t1 + c(t2_2, t2_2)*t2
+X1 | c(t1_1, t1_1)*t1 + t2
+X3 | c(t1_3, t1_3)*t1 + t2
+X4 | c(t1_4, t1_4)*t1 + t2
+X5 | c(t1_5, t1_5)*t1
+X6 | c(t1_6, t1_6)*t1
+X8 | c(t1_8, t1_8)*t1 + c(t2_8, t2_8)*t2
+X7 | c(t1_7, t1_7)*t1 + t2
+X10 | c(t1_10, t1_10)*t1 + t2
+X11 | c(t1_11, t1_11)*t1
+X1  ~ c(0, 0)*1
+X2  ~ c(0, 0)*1
+X3  ~ c(0, 0)*1
+X4  ~ c(0, 0)*1
+X5  ~ c(0, 0)*1
+X6  ~ c(0, 0)*1
+X7  ~ c(0, 0)*1
+X8  ~ c(0, 0)*1
+X9  ~ c(0, 0)*1
+X10  ~ c(0, 0)*1
+X11  ~ c(0, 0)*1
+X1 ~~ c(1,NA)*X1
+X2 ~~ c(1,NA)*X2
+X3 ~~ c(1,NA)*X3
+X4 ~~ c(1,NA)*X4
+X5 ~~ c(1,NA)*X5
+X6 ~~ c(1,NA)*X6
+X7 ~~ c(1,NA)*X7
+X8 ~~ c(1,NA)*X8
+X9 ~~ c(1,NA)*X9
+X10 ~~ c(1,NA)*X10
+X11 ~~ c(1,NA)*X11
+EXT ~~ NA*EXT
+INT ~~ NA*INT
+EXT ~~ NA*INT
+EXT ~ c(0, NA)*1
+INT ~ c(0, NA)*1
+"
[…]

### 2018-01-24 codes/R/4_scores.R

@@ -1,0 +1,25 @@
+# FUNCTION FOR SCORES
+
+############################################################################################
+## ---- FA_SCORES
+indices <- inspect(finalmod,"case.idx")   # indices for observations used
+scores <- lavPredict(finalmod)
+items.scored <- cbind(X[,c("ID", "faminc","incq" )], items.cb, matrix(NA, nrow(items.cb), 2))
+colnames(items.scored) <- c("ID", "faminc","incq", colnames(items.cb), "EXT", "INT")
+items.scored[indices[[1]],"EXT"] <- scores[[1]][,"EXT"]
+items.scored[indices[[2]],"EXT"] <- scores[[2]][,"EXT"]
+items.scored[indices[[1]],"INT"] <- scores[[1]][,"INT"]
+items.scored[indices[[2]],"INT"] <- scores[[2]][,"INT"]
+
+# residualise scores in age
+items.scored$EXTr <- residuals(lm(EXT ~ age, data=items.scored, na.action = na.exclude))
+items.scored$INTr <- residuals(lm(INT ~ age, data=items.scored, na.action = na.exclude))
+
+# mean by sex and cohort
+aggregate(items.scored[, c("EXT", "EXTr", "INT", "INTr")], list(items.scored$sex, items.scored$cohort), mean, na.rm=T)
+
+# add raw scores
+items.scored$EXT.RAW <- rowSums(items[,paste("X", seq(1,6), sep="")], na.rm = T)
+items.scored$INT.RAW <- rowSums(items[,paste("X", seq(7,11), sep="")], na.rm = T)
+
+

### 2018-01-24 codes/R/5_graphs.R

@@ -1,50 +1,25 @@
-# FUNCTION FOR SCORES AND GRAPHS (12 ITEMS)
 
 ############################################################################################
-## ---- FA_SCORES
-indices <- inspect(finalmod,"case.idx")   # indices for observations used
-scores <- lavPredict(finalmod)
-items.scored <- cbind(X[,c("ID", "faminc","incq" )], items.c, matrix(NA, nrow(items.c), 2))
-colnames(items.scored) <- c("ID", "faminc","incq", colnames(items.c), "EXT", "INT")
-items.scored[indices[[1]],"EXT"] <- scores[[1]][,"EXT"]
-items.scored[indices[[2]],"EXT"] <- scores[[2]][,"EXT"]
-items.scored[indices[[1]],"INT"] <- scores[[1]][,"INT"]
-items.scored[indices[[2]],"INT"] <- scores[[2]][,"INT"]
-
-# add raw scores
-raw.ext <- rowSums(items[,paste("X", seq(1,6), sep="")], na.rm = T)
-raw.int <- rowSums(items[,paste("X", seq(7,11), sep="")], na.rm = T)
-facandraw <- cbind(items.scored[,c("ID","incq","faminc", "cohort", "EXT", "INT")], INT.RAW=as.matrix(raw.int), EXT.RAW=as.matrix(raw.ext))
-
 ## ---- FACDENS
-require(tikzDevice)
-require(ggplot2)
-require(cowplot)
-pdext.ebm <- ggplot(items.scored, aes(x=EXT, group=cohort, fill=cohort, colour=cohort)) + geom_density(alpha = 0.1) + ggtitle("EXT (EBM Scores)")
-pdint.ebm <- ggplot(items.scored, aes(x=INT, group=cohort, fill=cohort, colour=cohort)) + geom_density(alpha = 0.1) + ggtitle("INT (EBM Scores)")
+# densities of factor scores (residuals)
+pdext.ebm <- ggplot(items.scored, aes(x=EXTr, group=cohort, fill=cohort, colour=cohort)) + geom_density(alpha = 0.1) + ggtitle("EXT (EBM Scores)")
+pdint.ebm <- ggplot(items.scored, aes(x=INTr, group=cohort, fill=cohort, colour=cohort)) + geom_density(alpha = 0.1) + ggtitle("INT (EBM Scores)")
[…]

### 2018-01-30 codes/R/cohorts_nc.Rnw
- restricted sample to complete cases - verified correct order of groups and binding of scores with original dataset - added age histogram plot - added LOESS plot of factors on age - other fixes

@@
 
-table <- afitab
+table <- rbind(afitab[[1]],afitab[[2]],afitab[[3]],afitab[[4]])
 
@@
              booktabs = T,
-             add.to.row = list(pos = c(list(0)),
+             add.to.row = list(pos = c(list(0,4,8,12)),
                               command = c(
                               # header
-                                    paste("
-                                            Model & Num. params & $\\chi^2$ & RMSEA & MFI & CFI & G-hat & \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\
-                                          "
-                                            , sep="")
+            "Model & Num. params & $\\chi^2$ & RMSEA & MFI & CFI & G-hat & \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ \\midrule
+                      \\newline \\textbf{No gender split} \\\\",
+                      "\\newline \\textbf{Males only} \\\\ \\midrule ",
+                      "\\newline \\textbf{Females only} \\\\ \\midrule ",
+                      "\\newline \\textbf{Joint (4 groups)} \\\\ \\midrule "
                                     )),
@@
 Parameters of the model are estimated by mean- and variance adjusted weighted least squares (WLSMV), due to \cite{Mplus2017}.
-Point estimated for model parameters are obtained starting from the polychoric correlations, using diagonally weighted least squares (DWLS), and the full weight matrix is used to compute robust standard errors and test statistics.
+Point estimates for model parameters are obtained starting from the polychoric correlations, using diagonally weighted least squares (DWLS), and the full weight matrix is used to compute robust standard errors and test statistics.
 Robust WLS has proved in simulation studies to be moderately robust to small violations of the normality assumption in the latent underlying measures \citep{Flora2011}, and generally outperforms maximum likelihood in large samples \citep{Beauducel2009,Li2016}.} The fit of the increasingly restrictive models presented above is compared in \autoref{tab:fit}.
@@
 The overall fit of the configural model is satisfactory according to widely accepted rules of thumb \citep{Maccallum1996}.\footnote{{\color{red} BETTER CITATIONS FOR OVERALL FIT}.} As discussed above, when items have less than four categories, the model with restricted thresholds is statistically equivalent to the configural model.
-Restricting thresholds and loadings seems to provide a satisfying fit compared to the configural model.\footnote{\cite{Meade2008}, using the results from a simulation study, suggests stricter thresholds that should apply in a variety of conditions.
+Restricting thresholds, loadings, and intercepts at the same time seems to provide a satisfying fit compared to the configural model.\footnote{\cite{Meade2008}, using the results from a simulation study, suggests stricter thresholds that should apply in a variety of conditions.
 For CFI, a single cutoff values of $.002$ is proposed, while cutoffs for MFI depend on the problem's characteristics; in our case (2 factors, 11 items), they suggest $.0066$.
-\cite{Sass2014} however cast some doubts of the generalisability of these cutoffs to WLSMV estimators.} However, the model that additionally restricts intercepts -- thus allowing for latent mean comparison -- has a substantially worse fit.
+\cite{Sass2014} however cast some doubts of the generalisability of these cutoffs to WLSMV estimators.}
 
-<<FA_FINAL, dependson='FA_PARTINV'>>=
+<<FA_FINAL, dependson='FA_INVARIANCE'>>=
 @
@@
 
-table <- allpar
+table <- rbind(allpar[[1]],allpar[[2]],allpar[[3]])
 
@@
                     # align = "ll|cccc|cccc"
-                    align="L{0cm}L{2cm}C{2.5cm}C{2.5cm}C{2.5cm}C{2.5cm}C{1.5cm}",
+                    align="L{0cm}L{2.5cm}C{2.5cm}C{2.5cm}C{2.5cm}C{2.5cm}C{1cm}",
+                    digits = 3
+                    ),
+             size="\\small",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0,11,22)),
+                              command = c(
+                              # header
+                                          " & \\textbf{Loadings}
+                                            & \\multicolumn{2}{c}{\\textbf{Thresholds}}
+                                            & \\textbf{Variances}
+                                            & \\\\ \\cmidrule(lr){2-2} \\cmidrule(lr){3-4} \\cmidrule(lr){5-5}
+
+                                            & $\\bm{\\lambda}$
+                                            & $\\bm{\\tau}_1$
+                                            & $\\bm{\\tau}_2$
+                                            & $\\bm{\\Var(\\varepsilon)}$ \\\\
+                                            Measure & BCS = MCS
+                                            & BCS = MCS
+                                            & BCS = MCS
+                                            & MCS (BCS=1)
+                                            & Factor
+                                            \\\\ \\midrule
+                      \\newline \\textbf{No gender split} \\\\",
+                      # additional separators
+                      "\\newline \\textbf{Males only} \\\\ \\midrule ",
+                      "\\newline \\textbf{Females only} \\\\ \\midrule " 
+                      )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabularx", width = "\\textwidth",
+             floating = FALSE
+)
+
+@
+}
+\end{table}
+
+\begin{table}[ht]
+\caption{Estimated loadings and thresholds -- 4-group model} \label{tab:allpars2}
+\resizebox{\textwidth}{!}{
+<<RES_FA_PAR2, dependson="FA_FINAL_PARS", results='asis'>>=
+require(xtable)
+
+table <- allpar[[4]]
+
+# sequences to add rows
+rws <- seq(2, (nrow(table)-1), by = 1) # start from row 2 (exclude model titles)
+rwsl <- as.list(rws)
+
+print.xtable(xtable(table,
+                    # align = "ll|cccc|cccc"
+                    align="L{0cm}L{1.5cm}C{2cm}C{1.5cm}C{1.5cm}C{1.8cm}C{1.8cm}C{1.8cm}C{1cm}",
                     digits = 3
@@
                               # header
-                                    paste(" & Loadings
+                                          " & \\textbf{Loadings}
                                             & \\multicolumn{2}{c}{\\textbf{Thresholds}}
-                                            & Intercepts
-                                            & \\\\ \\cmidrule(lr){3-4}
+                                            & \\multicolumn{3}{c}{\\textbf{Variances}}
+                                            & \\\\ \\cmidrule(lr){2-2} \\cmidrule(lr){3-4} \\cmidrule(lr){5-7}
 
-                                            & $\\bm{\\lambda}$
-                                            & $\\bm{\\tau}_1$
-                                            & $\\bm{\\tau}_2$
-                                            & $\\bm{\\varepsilon}$ \\\\
-                                            Measure & BCS = MCS
-                                            & BCS = MCS
-                                            & BCS = MCS
-                                            & MCS (BCS=1)
+                                            & $\\lambda$
+                                            & $\\tau_1$
+                                            & $\\tau_2$
+                                            & \\multicolumn{3}{c}{$\\text{diag}(\\Psi)$} \\\\
+
+                                            Measure 
+                                            & All groups
+                                            & All groups
+                                            & All groups
+                                            & BCS Males
+                                            & MCS Females
+                                            & MCS Males
                                             & Factor
-                                            \\\\", sep=""))),
+                                            \\\\"
+                      )),
              sanitize.text.function = function(x){x},
@@
 
-table <- lvpars
+table <- rbind(lvpars[[1]],lvpars[[2]],lvpars[[3]],lvpars[[4]],lvpars[[5]])
 
 print.xtable(xtable(table,
-                    align="L{0cm}L{.07\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.07\\textwidth}",
+                    align="L{0cm}L{2.2cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}",
                     digits = 3
@@
              booktabs = T,
-             add.to.row = list(pos = c(list(0)),
+             add.to.row = list(pos = c(list(0,2,4,6,8,10)),
                               command = c(
                               # header
-                                    paste(" & \\multicolumn{4}{c}{\\textbf{BCS}}
+                                   "\\toprule
+                                    & \\multicolumn{4}{c}{\\textbf{BCS}}
                                     & \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
-                                            & Mean & \\multicolumn{2}{c}{Covariance} & Corr. & Mean & \\multicolumn{2}{c}{Covariance} & Corr. \\\\
-                                            ", sep="")
+                                    & Mean & \\multicolumn{2}{c}{Covariance} & Corr. & Mean & \\multicolumn{2}{c}{Covariance} & Corr. \\\\ \\midrule
+                      \\textbf{No gender split} \\\\",
+                      # additional separators
+                      "\\newline \\textbf{Males only} \\\\ ",
+                      "\\newline \\textbf{Females only} \\\\ ",
+                      "\\midrule \\newline \\textbf{Males (joint)} \\\\ ",
+                      "\\newline \\textbf{Females (joint)} \\\\ ",
+                      "\\bottomrule"
                                     )),
+             hline.after = NULL,  # remove default booktabs lines
              sanitize.text.function = function(x){x},
              sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabularx", width = ".85\\textwidth",
+             tabular.environment = "tabularx", width = "\\textwidth",
              floating = FALSE
@@
 \subsection{Inequality in noncognitive ability}
-Although the results of our measurement invariance analysis imply that comparison of latent means is not justified, the model with restricted loadings and thresholds can nevertheless be used to investigate how inequality of noncognitive skills has changed between the two cohorts.
+The model can be used to investigate how inequality of noncognitive skills has changed between the two cohorts.
 The variances of the $EXT$ and $INT$ factors are estimated to be larger in the MCS children by \Sexpr{round(lv.covs$MCS[1,1]-1,2)} and \Sexpr{round(lv.covs$MCS[2,2]-1,2)} of a standard deviation, respectively, indicating larger dispersion.
@@
 
-reg.bcs.ext <- lm(EXT~as.factor(incq) + age, data=subset(items.scored, cohort=="BCS"))
-bcs.51.ext <- round(reg.bcs.ext$coefficients["as.factor(incq)5"],2)
-reg.bcs.int <- lm(INT~as.factor(incq) + age, data=subset(items.scored, cohort=="BCS"))
-bcs.51.int <- round(reg.bcs.int$coefficients["as.factor(incq)5"],2)
+# combinations of regressions
+regcoeffs <- matrix( 
+  c(rep("BCS", 2), rep("MCS", 2), rep(c("M","F"), 2), rep(NA, 8)), # the data elements 
+  nrow=4,              # number of rows 
+  ncol=4,              # number of columns 
+  byrow = F)        # fill matrix by columns 
+colnames(regcoeffs) <- c("cohort", "sex", "EXTd", "INTd")
 
-reg.mcs.ext <- lm(EXT~as.factor(incq) + age, data=subset(items.scored, cohort=="MCS"))
-mcs.51.ext <- round(reg.mcs.ext$coefficients["as.factor(incq)5"],2)
-reg.mcs.int <- lm(INT~as.factor(incq) + age, data=subset(items.scored, cohort=="MCS"))
-mcs.51.int <- round(reg.mcs.int$coefficients["as.factor(incq)5"],2)
+# fill in differences
+regs.ext <- list()
+regs.int <- list()
+for (r in 1:nrow(regcoeffs)) {
+  regs.ext[[r]] <- lm(EXT ~ as.factor(incq) + age, data=subset(items.scored[[4]], cohort==regcoeffs[r,1] & sex==regcoeffs[r,2]))
+  regcoeffs[r,3] <- round(regs.ext[[r]]$coefficients["as.factor(incq)5"],2)
+  regs.int[[r]] <- lm(INT ~ as.factor(incq) + age, data=subset(items.scored[[4]], cohort==regcoeffs[r,1] & sex==regcoeffs[r,2]))
+  regcoeffs[r,4] <- round(regs.int[[r]]$coefficients["as.factor(incq)5"],2)
+}
 
@@
 In \autoref{fig:boxscores}, we can see that the gradient in noncognitive skills along family income quintiles is more pronounced in the MCS cohort.
-In particular, a five-year old at the top quintile of the income distribution in 1975 has a better score by \Sexpr{bcs.51.ext} in the externalising dimension and \Sexpr{bcs.51.int} in the internalising dimension, compared to a child at the bottom quintile.
-In 2005, the differences amount to \Sexpr{mcs.51.ext} and \Sexpr{mcs.51.int}, respectively.
+In particular, a male five-year old at the top quintile of the income distribution in 1975 has a better score by \Sexpr{regcoeffs[1,3]} in the externalising dimension and \Sexpr{regcoeffs[1,4]} in the internalising dimension, compared to a child at the bottom quintile.
+In 2005, the differences amount to \Sexpr{regcoeffs[3,3]} and \Sexpr{regcoeffs[3,4]}, respectively.
 
@@
 \begin{figure}
-<<FACDENS, dev='tikz', fig.height=5, fig.width=10, out.width='1\\linewidth'>>=
+\begin{center}
+<<AGEHIST, dev='tikz', fig.height=6, fig.width=6, out.width='.6\\linewidth'>>=
+@
+\end{center}
+\caption{Distribution of age in the BCS and MCS sample}
+{\scriptsize \emph{Notes}: The figure shows the density of observations at each month of age for each of the two cohorts.  \par}
+\end{figure}
+
+\clearpage
+\begin{figure}
+<<FACDENS, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
 @
 \caption{Density of Scored Factors}
-{\scriptsize \emph{Notes}: The figure shows estimated kernel densities of the scored factors from model \eqref{eq:finalmod}, separately by cohort.  \par}
+{\scriptsize \emph{Notes}: The figure shows estimated kernel densities of the scored factors separately by cohort.  \par}
+\end{figure}
+
+\begin{figure}
+<<FACLOESS_AGE, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
+@
+\caption{Scores and child age}
+{\scriptsize \emph{Notes}: The figure plots LOESS of the scored factors against child age. \par}
 \end{figure}
@@
 \caption{Raw vs Scored Factors}
-{\scriptsize \emph{Notes}: The figure plots the scored factors from model \eqref{eq:finalmod} on the vertical axis against the ``raw" scores on the horizontal axis, i.e.
+{\scriptsize \emph{Notes}: The figure plots the scored factors on the vertical axis against the ``raw" scores on the horizontal axis, i.e.
 scores obtained by simply summing the number of positive answers to each of the items.
@@
 \begin{figure}
-<<FACSCAT, dev='tikz', fig.height=5, fig.width=10, out.width='1\\linewidth'>>=
+<<FACSCAT, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
 @
 \caption{Scatterplots of factor scores by cohort}
-{\scriptsize \emph{Notes}: The figure plots the scored factors from model \eqref{eq:finalmod} separately by cohort. The externalising and internalising scores are plotted against one another. \par}
+{\scriptsize \emph{Notes}: The figure plots the scored factors separately by cohort. The externalising and internalising scores are plotted against one another. \par}
 \end{figure}
@@
 \caption{Distribution of scores across income quintiles}\label{fig:boxscores}
-{\scriptsize \emph{Notes}: The figure plots the distribution of the scored factors from model \eqref{eq:finalmod} across the quintiles in the income distribution. \par}
+{\scriptsize \emph{Notes}: The figure plots the distribution of the scored factors across the quintiles in the income distribution. \par}
 \end{figure}
@@
 \begin{figure}
-<<FACLOESS, dev='tikz', fig.height=5, fig.width=10, out.width='1\\linewidth'>>=
+<<FACLOESS_INC, dev='tikz', fig.height=5, fig.width=10, out.width='1\\linewidth'>>=
 @
 \caption{Scores and family income}
-{\scriptsize \emph{Notes}: The figure plots LOESS of the scored factors from model \eqref{eq:finalmod} against family income. \par}
+{\scriptsize \emph{Notes}: The figure plots LOESS of the scored factors against family income. \par}
 \end{figure}

### 2018-01-30 codes/R/1_load.R

@@ -78,7 +78,7 @@ bcs5aux <- cbind(bcs5data[,c("bcsid", "faminc", "incq", "ysch_moth", "ysch_fath"
                 cog2 = bcs5data$hfd_z,
                 cog3 = bcs5data$copy_z
                 )
-
+names(bcs5aux)[names(bcs5aux)=="bcsid"] <- "id"
 
 # MCS -------------------------------------------
 mcs5data <- read.dta(paste(dir_data, "mcs5yeng_rwt.dta", sep=""), convert.factors = F) # all MCS data
@@ -121,6 +121,8 @@ mcs5aux <- cbind(mcs5data[,c("mcsid", "faminc", "incq", "ysch_moth", "ysch_fath"
                 cog2 = mcs5data$psim_bastz,
                 cog3 = mcs5data$patc_bastz
 )
+names(mcs5aux)[names(mcs5aux)=="mcsid"] <- "id"
+
 
 
 # common items
@@ -143,24 +145,27 @@ colnames(Xtemp.mcs) <- nn
 X <- data.frame(rbind(Xtemp.bcs, Xtemp.mcs))
 
 # assemble final data
-cohort <- X[,"cohort"]
-items.c <- X[,c(grep("X", names(X), value=T), "cohort")]   # noncog items only
-items  <- items.c                              # numeric version of noncog items only
-for (i in 1:ncol(items.c)) items[,i] <- as.numeric(items.c[,i]) - 1
+items.c <- X[,c(grep("X", names(X), value=T), "ID", "cohort", "ageint5", "sex", "incq", "faminc")]
+colnames(items.c)[colnames(items.c) == "ageint5"] <- "age"
+items.c$sex <- factor(items.c$sex)
+levels(items.c$sex) = c("M", "F")
+items.c$cohortsex <- interaction(items.c[c("cohort","sex")]) # generate interaction
+items.c$cohortsex <- factor(items.c$cohortsex,levels(items.c$cohortsex)[c(1,3,2,4)]) # reorder
+levels(items.c$cohortsex) <- c("BCS.M", "BCS.F", "MCS.M", "MCS.F")
+items.c <- items.c[!is.na(items.c$cohortsex),] # drop missings
 
-# with age and sex
-items.cb <- X[,c(grep("X", names(X), value=T), "cohort", "ageint5","sex")]
-colnames(items.cb)[colnames(items.cb) == "ageint5"] <- "age"
-items.cb$sex <- factor(items.cb$sex)
-levels(items.cb$sex) = c("M", "F")
[…]

### 2018-01-30 codes/R/2_efa.R

@@ -6,9 +6,9 @@
 rotation <- "oblimin"
 
 # EFA with 11 item scale
-items.c.bcs <- items.c[cohort=="BCS",!(names(items.c) %in% c("cohort"))]
-nbcs <- dim(items.c.bcs)[1]
-bcscor <- hetcor(items.c.bcs, ML=TRUE, std.err = F)$correlations
+items.bcs <- items.c[items.c$cohort=="BCS",c(grep("X", names(X), value=T))]
+nbcs <- dim(items.bcs)[1]
+bcscor <- hetcor(items.bcs, ML=TRUE, std.err = F)$correlations
 bcsfa <- fa(r=bcscor, nfactors=2, rotate = rotation, fm="wls")
 
 
@@ -23,9 +23,9 @@ bcsfa$loadings
 ## ---- MCSEFA
 
 # EFA with 11 item scale
-items.c.mcs <- items.c[cohort=="MCS",!(names(items.c) %in% c("cohort"))]
-nmcs <- dim(items.c.mcs)[1]
-mcscor <- hetcor(items.c.mcs, ML=TRUE, std.err = F)$correlations
+items.mcs <- items.c[items.c$cohort=="MCS",c(grep("X", names(X), value=T))]
+nmcs <- dim(items.mcs)[1]
+mcscor <- hetcor(items.mcs, ML=TRUE, std.err = F)$correlations
 mcsfa <- fa(r=mcscor, nfactors=2, rotate = rotation, fm="wls")
 
 ## ---- MCSEFARES

### 2018-01-30 codes/R/3_mifa_final.R

@@ -122,7 +122,7 @@ for (i in 1:5) {
 rm(
    lambdas, means, pta, tau1, tau2, eps,
    Xtemp.bcs, Xtemp.mcs, bcs5_rutb419, bcs5_rutbAB, bcs5_rutc419,
-   cohort, dmfi, dcfi, dgam, indsel, mcskeepb, mnames, ncats, ncomm, nn,
+   dmfi, dcfi, dgam, indsel, mcskeepb, mnames, ncats, ncomm, nn,
    t, t2)
 
 

### 2018-01-30 codes/R/3_mifa_invariance.R

@@ -76,35 +76,35 @@ thr.load.int.we.thb <- collapse(readLines(paste0(dir_syntax, "thr_load_int_we_th
 # printfit(fa.thr.int.we.d)
 
 # WHOLE SAMPLE
-fa.c.a <- cfa(config.we.th, data = items.cb, group = "cohort", estimator="wlsmv", parameterization="theta")
+fa.c.a <- cfa(config.we.th, data = items[[1]], group = "cohort", estimator="wlsmv", parameterization="theta")
 printfit(fa.c.a)
-fa.t.a <- cfa(thr.we.th, data = items.cb, group = "cohort", estimator="wlsmv", parameterization="theta")
+fa.t.a <- cfa(thr.we.th, data = items[[1]], group = "cohort", estimator="wlsmv", parameterization="theta")
 printfit(fa.t.a)
 # first restrictive model: THRESHOLDS + LOADINGS
-fa.tl.a <- cfa(thr.load.we.th, data = items.cb, group = "cohort", estimator="wlsmv", parameterization="theta", group.equal='loadings')
+fa.tl.a <- cfa(thr.load.we.th, data = items[[1]], group = "cohort", estimator="wlsmv", parameterization="theta", group.equal='loadings')
 printfit(fa.tl.a)
 # second restrictive model: THRESHOLDS + LOADINGS + INTERCEPTS
-fa.tli.a <- cfa(thr.load.int.we.th, data = items.cb, group = "cohort", estimator="wlsmv", parameterization="theta", group.equal='loadings')
+fa.tli.a <- cfa(thr.load.int.we.th, data = items[[1]], group = "cohort", estimator="wlsmv", parameterization="theta", group.equal='loadings')
 printfit(fa.tli.a)
 
 
 # MALES
-fa.c.m   <- cfa(config.we.th,        data = subset(items.cb, sex=="M"), group = "cohort", estimator="wlsmv", parameterization="theta")
-fa.t.m   <- cfa(thr.we.th,           data = subset(items.cb, sex=="M"), group = "cohort", estimator="wlsmv", parameterization="theta")
-fa.tl.m  <- cfa(thr.load.we.th,      data = subset(items.cb, sex=="M"), group = "cohort", estimator="wlsmv", parameterization="theta", group.equal='loadings')
-fa.tli.m <- cfa(thr.load.int.we.th,  data = subset(items.cb, sex=="M"), group = "cohort", estimator="wlsmv", parameterization="theta", group.equal='loadings')
[…]

### 2018-01-30 codes/R/4_scores.R

@@ -2,24 +2,28 @@
 
 ############################################################################################
 ## ---- FA_SCORES
-indices <- inspect(finalmod,"case.idx")   # indices for observations used
-scores <- lavPredict(finalmod)
-items.scored <- cbind(X[,c("ID", "faminc","incq" )], items.cb, matrix(NA, nrow(items.cb), 2))
-colnames(items.scored) <- c("ID", "faminc","incq", colnames(items.cb), "EXT", "INT")
-items.scored[indices[[1]],"EXT"] <- scores[[1]][,"EXT"]
-items.scored[indices[[2]],"EXT"] <- scores[[2]][,"EXT"]
-items.scored[indices[[1]],"INT"] <- scores[[1]][,"INT"]
-items.scored[indices[[2]],"INT"] <- scores[[2]][,"INT"]
+scores <- list()
+scoresdf <- list()
+items.scored <- list()
+for (i in 4:4) { # only specification with separate gender groups
 
-# residualise scores in age
-items.scored$EXTr <- residuals(lm(EXT ~ age, data=items.scored, na.action = na.exclude))
-items.scored$INTr <- residuals(lm(INT ~ age, data=items.scored, na.action = na.exclude))
+  scores[[i]] <- predict(finalmod[[i]], newdata = items[[i]])
+  scoresdf[[i]] <- do.call(rbind, scores[[i]])
+  items.scored[[i]] <- cbind(items[[i]], scoresdf[[i]])
+  
+  # residualise scores in age
+  items.scored[[i]]$EXTr <- residuals(lm(EXT ~ age, data=items.scored[[i]], na.action = na.exclude))
+  items.scored[[i]]$INTr <- residuals(lm(INT ~ age, data=items.scored[[i]], na.action = na.exclude))
+  
+  # raw scores
+  items.scored[[i]]$EXT.RAW <- rowSums(apply(items.scored[[i]][,paste("X", seq(1,6), sep="")], 2, function(x) as.numeric(x)), na.rm = T)
+  items.scored[[i]]$INT.RAW <- rowSums(apply(items.scored[[i]][,paste("X", seq(7,11), sep="")], 2, function(x) as.numeric(x)), na.rm = T)
+  
+}
 
 # mean by sex and cohort
-aggregate(items.scored[, c("EXT", "EXTr", "INT", "INTr")], list(items.scored$sex, items.scored$cohort), mean, na.rm=T)
[…]

### 2018-01-30 codes/R/5_graphs.R

@@ -1,23 +1,69 @@
 
 ############################################################################################
+## ---- AGEHIST
+# histogram of age (months) by cohort
+
+# make data
+aged <- aggregate(scores2plot$cohortsex, list(age = scores2plot$age, cohort = scores2plot$cohort), FUN = length)
+# convert to densities
+aged$dens[aged$cohort == "BCS"] <-  aged[aged$cohort == "BCS", "x"]/ sum(aged[aged$cohort == "BCS", "x"])
+aged$dens[aged$cohort == "MCS"] <-  aged[aged$cohort == "MCS", "x"]/ sum(aged[aged$cohort == "MCS", "x"])
+aged[aged$cohort=="MCS", "dens"] <- - aged[aged$cohort=="MCS", "dens"] # flip MCS
+
+ggplot(aged, aes(x=age, y=dens, fill=cohort)) + 
+  ylab("Density") + scale_y_continuous(limits = c(-.5,.75)) +
+  xlab("Age (months)") +
+  geom_bar(stat="identity", position="identity") +
+  theme(legend.position = c(0.9, 0.2)) +
+  geom_hline(aes(yintercept=0))
+
 ## ---- FACDENS
-# densities of factor scores (residuals)
-pdext.ebm <- ggplot(items.scored, aes(x=EXTr, group=cohort, fill=cohort, colour=cohort)) + geom_density(alpha = 0.1) + ggtitle("EXT (EBM Scores)")
-pdint.ebm <- ggplot(items.scored, aes(x=INTr, group=cohort, fill=cohort, colour=cohort)) + geom_density(alpha = 0.1) + ggtitle("INT (EBM Scores)")
-plot_grid(pdext.ebm, pdint.ebm, ncol=2, align="h")
+# densities of factor scores
+pdext.ebm.m <- ggplot(subset(scores2plot, sex=="M"), aes(x=EXT, group=cohort, fill=cohort, colour=cohort)) + geom_density(alpha = 0.1) + ggtitle("EXT Scores (Males)")
+pdint.ebm.m <- ggplot(subset(scores2plot, sex=="M"), aes(x=INT, group=cohort, fill=cohort, colour=cohort)) + geom_density(alpha = 0.1) + ggtitle("INT Scores (Males)")
[…]

### 2018-01-31 codes/R/cohorts_nc.Rnw
- fix LOESS on income - add table with mean factors

@@
 The model can be used to investigate how inequality of noncognitive skills has changed between the two cohorts. 
-The variances of the $EXT$ and $INT$ factors are estimated to be larger in the MCS children by \Sexpr{round(lv.covs$MCS[1,1]-1,2)} and \Sexpr{round(lv.covs$MCS[2,2]-1,2)} of a standard deviation, respectively, indicating larger dispersion.
+%The variances of the $EXT$ and $INT$ factors are estimated to be larger in the MCS children by \Sexpr{round(lv.covs$MCS[1,1]-1,2)} and \Sexpr{round(lv.covs$MCS[2,2]-1,2)} of a standard deviation, respectively, indicating larger dispersion.
 
@@
 @
+
+\begin{table}[ht]
+\caption{Estimated latent variable means and covariances} \label{tab:lvpars}
+\centering
+<<FA_COMP, dependson='FA_SCORES', results='asis'>>=
+# compare mean factor scores by gender and cohort
+
+prn <- function(obj) sprintf("%.3f", round(obj,3))
+
+sctable <- aggregate(items.scored[[4]][, c("EXT", "INT", "EXT.RAW", "INT.RAW")], list(items.scored[[4]]$cohortsex), 
+                   FUN = function (x) c(mn = prn(mean(x)), med = prn(median(x))))
+sctable <- cbind(group = factor(c(1,3,2,4)), sctable)
+levels(sctable$group) <- c("BCS Males", "MCS Males", "BCS Females", "MCS Females")
+sctable <- sctable[order(sctable$group),]
+sctable <- sctable[, !names(sctable) %in% "Group.1"]
+sctable <- as.matrix(sctable,3)
+
+print.xtable(xtable(sctable,
+                    align="L{0cm}L{2.2cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}",
+                    digits = 3
+                    ),
+             size="\\small",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0,4)),
+                              command = c(
+                              # header
+                                   "\\toprule
+                                    & \\multicolumn{4}{c}{\\textbf{Scored factors}}
+                                    & \\multicolumn{4}{c}{\\textbf{Scales}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
+                                    & \\multicolumn{2}{c}{EXT} & \\multicolumn{2}{c}{INT} & \\multicolumn{2}{c}{EXT} & \\multicolumn{2}{c}{INT} \\\\ \\cmidrule(lr){2-3} \\cmidrule(lr){4-5} \\cmidrule(lr){6-7} \\cmidrule(lr){8-9}
+                      & Mean & Median & Mean & Median & Mean & Median & Mean & Median \\\\",
+                      "\\bottomrule"
+                                    )),
+             hline.after = NULL,  # remove default booktabs lines
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabularx", width = "\\textwidth",
+             floating = FALSE
+)
+
+@
+\end{table}
 
@@
 \begin{figure}
-<<FACLOESS_INC, dev='tikz', fig.height=5, fig.width=10, out.width='1\\linewidth'>>=
+<<FACLOESS_INC, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
 @

### 2018-01-31 codes/R/5_graphs.R

@@ -106,12 +106,30 @@ legend_b <- get_legend(ineqlist[[1]] + theme(legend.position="bottom"))
 p <- plot_grid( pcol, legend_b, ncol = 1, rel_heights = c(1, .1))
 p
 
-
 ## ---- FACLOESS_INC
+
 # loess plot of scores on income
-loess.ext <- ggplot(data=scores2plot, aes(x=faminc, y=EXTr, group=cohort, fill=cohort, colour=cohort) ) + geom_smooth(method = "loess") + xlab("Weekly Family Income at 10 (,000£ 2015)") + ylab("EXT")
-loess.int <- ggplot(data=scores2plot, aes(x=faminc, y=INTr, group=cohort, fill=cohort, colour=cohort) ) + geom_smooth(method = "loess") + xlab("Weekly Family Income at 10 (,000£ 2015)") + ylab("INT")
-plot_grid(loess.ext, loess.int, ncol=2, align="h")
+l.inc.ext.m <- ggplot(data=subset(scores2plot, sex=="M"), aes(x=faminc, y=EXT, group=cohort, fill=cohort, colour=cohort) ) + ylab("EXT") + ggtitle("Males Externalising")
+l.inc.ext.f <- ggplot(data=subset(scores2plot, sex=="F"), aes(x=faminc, y=EXT, group=cohort, fill=cohort, colour=cohort) ) + ylab("EXT") + ggtitle("Females Externalising")
+l.inc.int.m <- ggplot(data=subset(scores2plot, sex=="M"), aes(x=faminc, y=INT, group=cohort, fill=cohort, colour=cohort) ) + ylab("INT") + ggtitle("Males Internalising")
+l.inc.int.f <- ggplot(data=subset(scores2plot, sex=="F"), aes(x=faminc, y=INT, group=cohort, fill=cohort, colour=cohort) ) + ylab("INT") + ggtitle("Females Internalising")
+l.inc <- list(l.inc.ext.m, l.inc.ext.f, l.inc.int.m, l.inc.int.f)
+l.inc <- lapply(l.inc, 
+                 function(x) x + theme(legend.position="none") +
+                   geom_smooth(method = "loess") + 
+                   xlab("Weekly Family Income at 10 (,000£ 2015)")
+                 ) # apply options to all graphs 
+
+pcol <- plot_grid( l.inc[[1]], l.inc[[2]], l.inc[[3]], l.inc[[4]],
+                   align = 'vh',
[…]

### 2018-02-02 codes/R/cohorts_nc.Rnw
NVQ returns to INT EXT

@@
 \begin{table}[ht]
-\caption{Estimated latent variable means and covariances} \label{tab:lvpars}
+\caption{Estimated means and covariances of scores} \label{tab:scoremeans}
 \centering
@@
 
+\begin{table}[ht]
+\caption{Regression of scores on income} \label{tab:regs}
+\centering
+\resizebox{\linewidth}{!}{%
+<<FA_REGS2, dependson='FA_SCORES', results='asis'>>=
+# regress scores on income
+library(dplyr)
+library(stargazer)
+
+scores2plot$incq_infl <- factor(ntile(scores2plot$faminc_infl, 5))
+
+r.ext.f.c <- lm(EXT ~ faminc_infl*cohort, data=subset(scores2plot, sex=="F"))
+r.ext.f.q <- lm(EXT ~ incq_infl*cohort, data=subset(scores2plot, sex=="F"))
+r.int.f.c <- lm(INT ~ faminc_infl*cohort, data=subset(scores2plot, sex=="F"))
+r.int.f.q <- lm(INT ~ incq_infl*cohort, data=subset(scores2plot, sex=="F"))
+
+r.ext.m.c <- lm(EXT ~ faminc_infl*cohort, data=subset(scores2plot, sex=="M"))
+r.ext.m.q <- lm(EXT ~ incq_infl*cohort, data=subset(scores2plot, sex=="M"))
+r.int.m.c <- lm(INT ~ faminc_infl*cohort, data=subset(scores2plot, sex=="M"))
+r.int.m.q <- lm(INT ~ incq_infl*cohort, data=subset(scores2plot, sex=="M"))
+
+stargazer(r.ext.f.c, r.ext.f.q, r.int.f.c, r.int.f.q, r.ext.m.c, r.ext.m.q, r.int.m.c, r.int.m.q,
+          column.labels=c("Females","Males"), column.separate=c(4,4),
+          dep.var.caption = "", multicolumn = FALSE,
+          omit=c("Constant"),
+          covariate.labels=c("Family income (2015 CPI £)",
+                             "Family income Q2 (2015 CPI £)",
+                             "Family income Q3 (2015 CPI £)",
+                             "Family income Q4 (2015 CPI £)",
+                             "Family income Q5 (2015 CPI £)",
+                             "MCS dummy",
+                             "Family income (2015 CPI £) x MCS",
+                             "Family income Q2 (2015 CPI £) x MCS",
+                             "Family income Q3 (2015 CPI £) x MCS",
+                             "Family income Q4 (2015 CPI £) x MCS",
+                             "Family income Q5 (2015 CPI £) x MCS"
+                             ),
+          no.space=TRUE, 
+          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
+          float=FALSE    # needs begin table outside it
+)
+@
+}
+\end{table}
+
+\begin{table}[ht]
+\caption{Returns to EXT/INT in BCS} \label{tab:returns}
+\centering
+\resizebox{\linewidth}{!}{%
+<<FA_RETURNS, dependson='FA_SCORES', results='asis'>>=
+# regress scores on income
+library(dplyr)
+library(stargazer)
+
+# get BCS data and make dummy for higher education
+scoresBCS <- scores2plot[scores2plot$cohort=="BCS",]
+scoresBCS$highered <- as.numeric(scoresBCS$hinvq00>3)
+scoresBCS[is.na(scoresBCS$hinvq00), "highered"] <- NA
+
+# females
+r.nvq.f <- list()
+r.nvq.f[[1]] <- lm(highered ~ EXT + INT, data=subset(scoresBCS, sex=="F"))
+r.nvq.f[[2]] <- lm(highered ~ EXT + INT + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="F"))
+r.nvq.f[[3]] <- lm(highered ~ EXT + INT + factor(incq), data=subset(scoresBCS, sex=="F"))
+r.nvq.f[[4]] <- lm(highered ~ EXT + INT + ysch_moth + ysch_fath, data=subset(scoresBCS, sex=="F"))
+r.nvq.f[[5]] <- lm(highered ~ EXT + INT + factor(incq) + ysch_moth + ysch_fath + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="F"))
+
+# females
+r.nvq.m <- list()
+r.nvq.m[[1]] <- lm(highered ~ EXT + INT, data=subset(scoresBCS, sex=="M"))
+r.nvq.m[[2]] <- lm(highered ~ EXT + INT + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="M"))
+r.nvq.m[[3]] <- lm(highered ~ EXT + INT + factor(incq), data=subset(scoresBCS, sex=="M"))
+r.nvq.m[[4]] <- lm(highered ~ EXT + INT + ysch_moth + ysch_fath, data=subset(scoresBCS, sex=="M"))
+r.nvq.m[[5]] <- lm(highered ~ EXT + INT + factor(incq) + ysch_moth + ysch_fath + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="M"))
+
+stargazer(r.nvq.f[[1]], r.nvq.f[[2]], r.nvq.f[[3]], r.nvq.f[[4]], r.nvq.f[[5]],
+          r.nvq.m[[1]], r.nvq.m[[2]], r.nvq.m[[3]], r.nvq.m[[4]], r.nvq.m[[5]],
+          dep.var.caption = "Completed Higher Ed. at 30",
+          column.labels=c("Females","Males"), column.separate=c(5,5),
+          omit=c("Constant"),
+          covariate.labels=c("Externalising skill at 5",
+                             "Internalising skill at 5",
+                             "EPVT Z at 5",
+                             "Human figure drawing Z at 5",
+                             "Copy design Z at 5",
+                             "Family inc. at 10 Q2",
+                             "Family inc. at 10 Q3",
+                             "Family inc. at 10 Q4",
+                             "Family inc. at 10 Q5",
+                             "Mother's years of schooling",
+                             "Father's years of schooling"
+                             ),
+          no.space=TRUE, 
+          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
+          float=FALSE    # needs begin table outside it
+)
+@
+}
+\end{table}
+
+\begin{table}[ht]
+\caption{Returns to EXT/INT in BCS -- raw scores} \label{tab:returns_raw}
+\centering
+\resizebox{\linewidth}{!}{%
+<<FA_RETURNS_RAW, dependson='FA_SCORES', results='asis'>>=
+# regress scores on income
+library(dplyr)
+library(stargazer)
+
+# get BCS data and make dummy for higher education
+scoresBCS <- scores2plot[scores2plot$cohort=="BCS",]
+scoresBCS$highered <- as.numeric(scoresBCS$hinvq00>3)
+scoresBCS[is.na(scoresBCS$hinvq00), "highered"] <- NA
+
+# females
+r.nvq.f <- list()
+r.nvq.f[[1]] <- lm(highered ~ EXT.RAW + INT.RAW, data=subset(scoresBCS, sex=="F"))
+r.nvq.f[[2]] <- lm(highered ~ EXT.RAW + INT.RAW + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="F"))
+r.nvq.f[[3]] <- lm(highered ~ EXT.RAW + INT.RAW + factor(incq), data=subset(scoresBCS, sex=="F"))
+r.nvq.f[[4]] <- lm(highered ~ EXT.RAW + INT.RAW + ysch_moth + ysch_fath, data=subset(scoresBCS, sex=="F"))
+r.nvq.f[[5]] <- lm(highered ~ EXT.RAW + INT.RAW + factor(incq) + ysch_moth + ysch_fath + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="F"))
+
+# females
+r.nvq.m <- list()
+r.nvq.m[[1]] <- lm(highered ~ EXT.RAW + INT.RAW, data=subset(scoresBCS, sex=="M"))
+r.nvq.m[[2]] <- lm(highered ~ EXT.RAW + INT.RAW + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="M"))
+r.nvq.m[[3]] <- lm(highered ~ EXT.RAW + INT.RAW + factor(incq), data=subset(scoresBCS, sex=="M"))
+r.nvq.m[[4]] <- lm(highered ~ EXT.RAW + INT.RAW + ysch_moth + ysch_fath, data=subset(scoresBCS, sex=="M"))
+r.nvq.m[[5]] <- lm(highered ~ EXT.RAW + INT.RAW + factor(incq) + ysch_moth + ysch_fath + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="M"))
+
+stargazer(r.nvq.f[[1]], r.nvq.f[[2]], r.nvq.f[[3]], r.nvq.f[[4]], r.nvq.f[[5]],
+          r.nvq.m[[1]], r.nvq.m[[2]], r.nvq.m[[3]], r.nvq.m[[4]], r.nvq.m[[5]],
+          dep.var.caption = "Completed Higher Ed. at 30",
+          column.labels=c("Females","Males"), column.separate=c(5,5),
+          omit=c("Constant"),
+          covariate.labels=c("Externalising skill at 5 (raw score)",
+                             "Internalising skill at 5 (raw score)",
+                             "EPVT Z at 5",
+                             "Human figure drawing Z at 5",
+                             "Copy design Z at 5",
+                             "Family inc. at 10 Q2",
+                             "Family inc. at 10 Q3",
+                             "Family inc. at 10 Q4",
+                             "Family inc. at 10 Q5",
+                             "Mother's years of schooling",
+                             "Father's years of schooling"
+                             ),
+          no.space=TRUE, 
+          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
+          float=FALSE    # needs begin table outside it
+)
+@
+}
+\end{table}
+
 We can derive scores for $\theta^{EXT}$ and $\theta^{INT}$ directly, and plot them against socioeconomic measures observed in the data.
@@
 \begin{center}
-<<AGEHIST, dev='tikz', fig.height=6, fig.width=6, out.width='.6\\linewidth'>>=
+<<AGEHIST, dev='tikz', fig.height=6, fig.width=6, out.width='.6\\linewidth', cache=TRUE>>=
 @

### 2018-02-02 codes/R/1_load.R

@@ -21,7 +21,6 @@ library(xtable)
 library(cowplot)
 library(tikzDevice)
 library(gtools)
-library(glue)
 
 
 dir_data  <- c("/Users/giacomomason/Documents/Projects/CohortStudies/rdata/")
@@ -73,7 +72,7 @@ for (i in 1:ncol(bcs5rutc)) bcs5rutcf[,i] <- as.ordered(bcs5rutc[,i])
 
 
 # auxiliary variables
-bcs5aux <- cbind(bcs5data[,c("bcsid", "faminc_real", "faminc_infl", "incq", "ysch_moth", "ysch_fath", "ageint5", "sex")],
+bcs5aux <- cbind(bcs5data[,c("bcsid", "faminc_real", "faminc_infl", "incq", "ysch_moth", "ysch_fath", "ageint5", "sex", "hinvq00")],
                 cog1 = bcs5data$epvt_z,
                 cog2 = bcs5data$hfd_z,
                 cog3 = bcs5data$copy_z
@@ -116,14 +115,16 @@ mcs5sdqcf <- mcs5sdqc
 for (i in 1:ncol(mcs5sdqc)) mcs5sdqcf[,i] <- as.ordered(mcs5sdqc[,i])                    
 
 # add SES and cognitive data
-mcs5aux <- cbind(mcs5data[,c("mcsid", "faminc_real", "faminc_infl", "incq", "ysch_moth", "ysch_fath", "ageint5","sex")],
+mcs5aux <- cbind(mcs5data[,c("mcsid", "faminc_real", "faminc_infl", "incq", "ysch_moth", "ysch_fath", "ageint5", "sex")],
                 cog1 = mcs5data$nvoc_bastz,
                 cog2 = mcs5data$psim_bastz,
                 cog3 = mcs5data$patc_bastz
 )
 names(mcs5aux)[names(mcs5aux)=="mcsid"] <- "id"
+# add empty column with hinvq00 to match BCS
+mcs5aux$hinvq00 <- NA
 
-
+# MERGE -------------------------------------------
 
 # common items
 Xtemp.bcs <- cbind( bcs5aux,      # additional variables
@@ -138,14 +139,16 @@ Xtemp.mcs <- cbind( mcs5aux,      # additional variables
 )
 
 # append BCS and MCS
-nn <- c(colnames(bcs5aux), "cohort", "X1", "X2", "X3", "X4", "X5", "X6", "X7", "X8", "X9", "X10", "X11")
-nn[1] <- "ID"
[…]

### 2018-02-02 codes/R/2_efa.R

@@ -6,7 +6,7 @@
 rotation <- "oblimin"
 
 # EFA with 11 item scale
-items.bcs <- items.c[items.c$cohort=="BCS",c(grep("X", names(X), value=T))]
+items.bcs <- items.c[items.c$cohort=="BCS",c(grep("X", names(items.c), value=T))]
 nbcs <- dim(items.bcs)[1]
 bcscor <- hetcor(items.bcs, ML=TRUE, std.err = F)$correlations
 bcsfa <- fa(r=bcscor, nfactors=2, rotate = rotation, fm="wls")
@@ -23,7 +23,7 @@ bcsfa$loadings
 ## ---- MCSEFA
 
 # EFA with 11 item scale
-items.mcs <- items.c[items.c$cohort=="MCS",c(grep("X", names(X), value=T))]
+items.mcs <- items.c[items.c$cohort=="MCS",c(grep("X", names(items.c), value=T))]
 nmcs <- dim(items.mcs)[1]
 mcscor <- hetcor(items.mcs, ML=TRUE, std.err = F)$correlations
 mcsfa <- fa(r=mcscor, nfactors=2, rotate = rotation, fm="wls")

### 2018-02-02 codes/R/3_mifa_final.R

@@ -123,7 +123,7 @@ for (i in 1:5) {
 rm(
    lambdas, means, pta, tau1, tau2, eps,
    Xtemp.bcs, Xtemp.mcs, bcs5_rutb419, bcs5_rutbAB, bcs5_rutc419,
-   dmfi, dcfi, dgam, indsel, mcskeepb, mnames, ncats, ncomm, nn,
+   dmfi, dcfi, dgam, indsel, mcskeepb, mnames, ncats, ncomm,
    t, t2)
 
 

### 2018-02-02 codes/R/3_mifa_invariance.R

@@ -9,8 +9,8 @@
 # lv variances = 1
 # intercepts = 0
 # unique variances = 1
-config.we.th <- collapse(readLines(paste0(dir_syntax, "config_we_th.lav")), sep=" \n  ")
-config.we.thb <- collapse(readLines(paste0(dir_syntax, "config_we_thb.lav")), sep=" \n  ")
+config.we.th <- glue::collapse(readLines(paste0(dir_syntax, "config_we_th.lav")), sep=" \n  ")
+config.we.thb <- glue::collapse(readLines(paste0(dir_syntax, "config_we_thb.lav")), sep=" \n  ")
 
 # WE CONFIGURAL - DELTA PARAMETERISATION (Condition 7)
 # free loadings and thresholds
@@ -18,7 +18,7 @@ config.we.thb <- collapse(readLines(paste0(dir_syntax, "config_we_thb.lav")), se
 # lv variances = 1
 # intercepts = 0
 # scales = 1
-config.we.d <- collapse(readLines(paste0(dir_syntax, "config_we_d.lav")), sep=" \n  ")
+config.we.d <- glue::collapse(readLines(paste0(dir_syntax, "config_we_d.lav")), sep=" \n  ")
 
 
 # WE THRESHOLD invariance, theta par (Condition 15)
@@ -28,8 +28,8 @@ config.we.d <- collapse(readLines(paste0(dir_syntax, "config_we_d.lav")), sep=" 
 # lv variances = 1
 # unique variances free in group 2 (except binary items)
 # intercepts free in group 2
-thr.we.th <- collapse(readLines(paste0(dir_syntax, "thr_we_th.lav")), sep=" \n  ")
-thr.we.thb <- collapse(readLines(paste0(dir_syntax, "thr_we_thb.lav")), sep=" \n  ")
+thr.we.th <- glue::collapse(readLines(paste0(dir_syntax, "thr_we_th.lav")), sep=" \n  ")
+thr.we.thb <- glue::collapse(readLines(paste0(dir_syntax, "thr_we_thb.lav")), sep=" \n  ")
 
 
 # WE THRESHOLD+LOADINGS invariance, theta par (Condition 19)
@@ -39,8 +39,8 @@ thr.we.thb <- collapse(readLines(paste0(dir_syntax, "thr_we_thb.lav")), sep=" \n
 # lv variances free in group 2
 # unique variances free in group 2 (except binary items)
 # intercepts free in group 2
[…]

### 2018-02-08 codes/R/3_mifa_final.R
add new models and extract fit + parameters

@@ -1,15 +1,15 @@
 # ############################################################################################
 ## ---- FA_FINAL
 
-# select final model
-finalmod <- list(fa.tli.a, fa.tli.m, fa.tli.f, fa.tli.b, fa.tli.ol)
+# select final model set
+finalmod <- fa.tli
 
 ##################################################################################
 ## ---- FA_FINAL_PARS
 # extract parameters and make table for final model estimates
 
 allpar <- list()
-for (i in 1:5) {
+for (i in 1:length(finalmod)) {
   pta <- partable(finalmod[[i]])
 
   # item parameters ----------------------------------------------------------
@@ -52,7 +52,7 @@ for (i in 1:5) {
 }
 
 # additional variances for 4-group models
-for (i in 4:5) {
+for (i in 7:11) {
   pta <- partable(finalmod[[i]])
   eps3 <- pta[(pta$op == "~~" & substr(pta$rhs,1,1) == "X" & pta$group == 3),c("lhs", "rhs","est")]
   eps3$measure <- eps3$lhs
@@ -71,7 +71,9 @@ for (i in 4:5) {
 # latent variable parameters ----------------------------------------------------------
 
 lvpars <- list()
-for (i in 1:3) {
+
+# 2-group models
+for (i in 1:6) {
   # model with threshold and loading invariance
   lv.means <- inspect(finalmod[[i]], what="mean.lv") # means
   lv.covs <- inspect(finalmod[[i]], what="cov.lv")   # covariances
@@ -82,47 +84,55 @@ for (i in 1:3) {
     cbind(as.matrix(lv.means[[1]]), lv.covs[[1]], as.matrix(c(NA,lv.corrs[[1]][2,1])), as.matrix(lv.means[[2]]), lv.covs[[2]], as.matrix(c(NA,lv.corrs[[2]][2,1])))
   )
   
-}
-
-
-# 4-GROUP MODEL
-lv.means4 <- inspect(finalmod[[4]], what="mean.lv")
-lv.covs4 <- inspect(finalmod[[4]], what="cov.lv")
[…]

### 2018-02-08 codes/R/3_mifa_invariance.R

@@ -9,8 +9,13 @@
 # lv variances = 1
 # intercepts = 0
 # unique variances = 1
-config.we.th <- glue::collapse(readLines(paste0(dir_syntax, "config_we_th.lav")), sep=" \n  ")
-config.we.thb <- glue::collapse(readLines(paste0(dir_syntax, "config_we_thb.lav")), sep=" \n  ")
+c.th.2a <- glue::collapse(readLines(paste0(dir_syntax, "c_th_2a.lav")), sep=" \n  ")
+c.th.2ac <- glue::collapse(readLines(paste0(dir_syntax, "c_th_2ac.lav")), sep=" \n  ")
+c.th.2 <- glue::collapse(readLines(paste0(dir_syntax, "c_th_2.lav")), sep=" \n  ")
+c.th.4a <- glue::collapse(readLines(paste0(dir_syntax, "c_th_4a.lav")), sep=" \n  ")
+c.th.4ac <- glue::collapse(readLines(paste0(dir_syntax, "c_th_4ac.lav")), sep=" \n  ")
+c.th.4 <- glue::collapse(readLines(paste0(dir_syntax, "c_th_4.lav")), sep=" \n  ")
+c.mimic <- paste(c.th.4, "EXT ~ age", "INT ~ age",  sep=" \n  ")
 
 # WE CONFIGURAL - DELTA PARAMETERISATION (Condition 7)
 # free loadings and thresholds
@@ -18,7 +23,7 @@ config.we.thb <- glue::collapse(readLines(paste0(dir_syntax, "config_we_thb.lav"
 # lv variances = 1
 # intercepts = 0
 # scales = 1
-config.we.d <- glue::collapse(readLines(paste0(dir_syntax, "config_we_d.lav")), sep=" \n  ")
+c.dl.2a <- glue::collapse(readLines(paste0(dir_syntax, "c_dl_2a.lav")), sep=" \n  ")
 
 
 # WE THRESHOLD invariance, theta par (Condition 15)
@@ -28,8 +33,13 @@ config.we.d <- glue::collapse(readLines(paste0(dir_syntax, "config_we_d.lav")), 
 # lv variances = 1
 # unique variances free in group 2 (except binary items)
 # intercepts free in group 2
-thr.we.th <- glue::collapse(readLines(paste0(dir_syntax, "thr_we_th.lav")), sep=" \n  ")
-thr.we.thb <- glue::collapse(readLines(paste0(dir_syntax, "thr_we_thb.lav")), sep=" \n  ")
[…]

### 2018-02-08 codes/R/4_scores.R
move raw scores to 1_load

@@ -5,7 +5,7 @@
 scores <- list()
 scoresdf <- list()
 items.scored <- list()
-for (i in 4:5) { # only specifications with separate gender groups
+for (i in 7:11) { # only specifications with separate gender groups
 
   scores[[i]] <- predict(finalmod[[i]], newdata = items[[i]])
   scoresdf[[i]] <- do.call(rbind, scores[[i]])
@@ -15,15 +15,8 @@ for (i in 4:5) { # only specifications with separate gender groups
   items.scored[[i]]$EXTr <- residuals(lm(EXT ~ age, data=items.scored[[i]], na.action = na.exclude))
   items.scored[[i]]$INTr <- residuals(lm(INT ~ age, data=items.scored[[i]], na.action = na.exclude))
   
-  # raw scores
-  items.scored[[i]]$EXT.RAW <- rowSums(apply(items.scored[[i]][,paste("X", seq(1,6), sep="")], 2, function(x) as.numeric(x)), na.rm = T)
-  items.scored[[i]]$INT.RAW <- rowSums(apply(items.scored[[i]][,paste("X", seq(7,11), sep="")], 2, function(x) as.numeric(x)), na.rm = T)
-  
 }
 
-# mean by sex and cohort
-aggregate(items.scored[[4]][, c("EXT", "EXT.RAW", "INT", "INT.RAW")], list(items.scored[[4]]$cohortsex), mean, na.rm=T)
-
 # select scores to plot
-scores2plot <- items.scored[[4]]
+scores2plot <- items.scored[[8]]
 

### 2018-02-08 codes/R/4_scores_compare.R
add script to compare scores across different models

@@ -1,0 +1,61 @@
+# SCRIPT TO COMPARE SCORED FACTORS FROM DIFFERENT MODELS
+
+############################################################################################
+## ---- FA_SCORES_COMPARE
+
+# 4 group with no intercept invariance
+scores.tl <- predict(fa.tl[[7]], newdata = items[[7]]) # new scores prediction
+items.scored.tl <- cbind(items[[7]][,c("id", "cohortsex", "faminc_real", "incq")], do.call(rbind, scores.tl))
+names(items.scored.tl)[names(items.scored.tl) %in% c("EXT","INT")] <- c("EXTtl", "INTtl")
+
+# 4 group with no intercept invariance, overlapping ages (model 9)
+items.scored.tlir <- cbind(items[[9]][,c("id", "cohortsex")], do.call(rbind, scores[[9]]))
+names(items.scored.tlir)[names(items.scored.tlir) %in% c("EXT","INT")] <- c("EXTtlir", "INTtlir")
+
+# 4 group with intercept invariance, no age adjustment (model 7)
+items.scored.tli <- cbind(items[[7]][,c("id", "cohortsex")], do.call(rbind, scores[[7]]))
+names(items.scored.tli)[names(items.scored.tli) %in% c("EXT","INT")] <- c("EXTtli", "INTtli")
+
+# 4 group with intercept invariance, age adjustment (model 8)
+items.scored.tlia <- cbind(items[[8]][,c("id", "cohortsex")], do.call(rbind, scores[[8]]))
+names(items.scored.tlia)[names(items.scored.tlia) %in% c("EXT","INT")] <- c("EXTtlia", "INTtlia")
+
+# 4 group with intercept invariance, age adjustment (constrained) (model 10)
+items.scored.tliac <- cbind(items[[10]][,c("id", "cohortsex")], do.call(rbind, scores[[10]]))
+names(items.scored.tliac)[names(items.scored.tliac) %in% c("EXT","INT")] <- c("EXTtliac", "INTtliac")
+
+# 4 group MIMIC
+items.scored.mim <- cbind(items[[10]][,c("id", "cohortsex")], do.call(rbind, scores[[11]]))
+names(items.scored.mim)[names(items.scored.mim) %in% c("EXT","INT")] <- c("EXTmim", "INTmim")
+
+# merge scores
[…]

### 2018-02-08 codes/R/cohorts_nc.Rnw
restructure file to put most stuff in appendix, add temporary notes

@@
                  fig.height=4, 
-                 out.width='.5\\textwidth',
+                 out.width='\\textwidth',
                  fig.show='hold',
@@
 read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/4_scores.R")
+read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/4_scores_compare.R")
 read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/5_graphs.R")
@@
 
-<<FA_INVARIANCE, dependson='LOAD_DATA'>>=
-@
-
-<<FA_FIT, dependson='FA_INVARIANCE'>>=
-@
-
-\begin{table}[ht]
-\centering
-\small
-\caption{Fit indices}\label{tab:fit}
-<<FA_FIT_RES, dependson="FA_FIT", results='asis'>>=
-require(xtable)
-
-table <- rbind(afitab[[1]],afitab[[2]],afitab[[3]],afitab[[4]])
-
-print(xtable(table,
-                    align="L{0cm}L{.18\\textwidth}C{.05\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
-                    digits = 4
-                    ),
-             size="\\small",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = c(list(0,4,8,12)),
-                              command = c(
-                              # header
-            "Model & Num. params & $\\chi^2$ & RMSEA & MFI & CFI & G-hat & \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ \\midrule
-                      \\newline \\textbf{No gender split} \\\\",
-                      "\\newline \\textbf{Males only} \\\\ \\midrule ",
-                      "\\newline \\textbf{Females only} \\\\ \\midrule ",
-                      "\\newline \\textbf{Joint (4 groups)} \\\\ \\midrule "
-                                    )),
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabular", width = "\\textwidth",
-             floating = FALSE
-)
-
-@
-\end{table}
-
 I estimate each of the models listed above by Weighted Least Squares.\footnote{All estimates are computed using the \texttt{lavaan} package in \textbf{\textsf{R}} \citep{Rosseel2012}.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\clearpage
+\subsection{Inequality in noncognitive ability}
+The model can be used to investigate how inequality of noncognitive skills has changed between the two cohorts.
+
+\textbf{REVIEW} We can derive scores for $\theta^{EXT}$ and $\theta^{INT}$ directly, and plot them against socioeconomic measures observed in the data.
+Our main socioeconomic indicator is a measure of weekly total family net income in the prices of the interview year.\footnote{This measure is harmonised across cohorts and derived as part of CLOSER, which is funded by the ESRC.} Using the estimated parameters for the model with invariant thresholds and loadings (but non-invariant intercepts), I compute the scores for each individual in the sample (using the Empirical Bayes modal approach).
+In \autoref{fig:boxscores}, we can see that the gradient in noncognitive skills along family income quintiles is more pronounced in the MCS cohort.
+%In particular, a male five-year old at the top quintile of the income distribution in 1975 has a better score by \Sexpr{regcoeffs[1,3]} in the externalising dimension and \Sexpr{regcoeffs[1,4]} in the internalising dimension, compared to a child at the bottom quintile.
+In 2005, the differences amount to \Sexpr{regcoeffs[3,3]} and \Sexpr{regcoeffs[3,4]}, respectively.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\clearpage
+
+List of estimated models:
+\bnm
+\item no gender split, no age adjustment (2 groups, BCS MCS)
+\item no gender split, age adjustment (2 groups, BCS MCS)
+\item males only, no age adjustment (2 groups, BCS MCS)
+\item females only, no age adjustment (2 groups, BCS MCS)
+\item males only, age adjustment (2 groups, BCS MCS)
+\item females only, age adjustment (2 groups, BCS MCS)
+\item separate gender groups, no age adjustment (4 groups, BCS.M BCS.F MCS.M MCS.F)
+\item separate gender groups, age adjustment (4 groups, BCS.M BCS.F MCS.M MCS.F)
+\item separate gender groups, overlapping ages, no age adjustment (4 groups, BCS.M BCS.F MCS.M MCS.F)
+\item separate gender groups, age adjustment constrained to be the same across ages (4 groups, BCS.M BCS.F MCS.M MCS.F)
+\item MIMIC model with age as regressor (4 groups, BCS.M BCS.F MCS.M MCS.F)
+\enm
+
+Notes: 
+\bi
+\item Only the models with age adjustment seem to have invariant intercepts (see Appendix \autoref{sec:app_fit}). 
+\item But the latent means of these models don't make sense with the "raw" scores we have (see Appendix \autoref{sec:app_meancov})
+\item Also, a model on the subset of children that have very similar ages delivers results that are very similar to the age-unadjusted models, and different from the age-adjusted ones.
+So the models with age adjustment seem to be doing something else than adjusting for age, and we don't understand what this something else is...
+\item The model with free intercepts (only restricting thresholds and loadings) delivers differences in scores by group, even if the means of the latent factors are constrained to be equal (see below)
+\item The models with free and constrained intercepts deliver very similar scores (.98 correlation), and they also give sensible and comparable gradients with income
+\ei
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+<<FA_INVARIANCE, dependson='LOAD_DATA'>>=
+@
+
+<<FA_FIT, dependson='FA_INVARIANCE'>>=
+@
+
 <<FA_FINAL, dependson='FA_INVARIANCE'>>=
@@
 
+<<CLEANUP, dependson='FA_FINAL'>>=
+@
+
+<<FA_SCORES, dependson='FA_FINAL'>>=
+@
+
+\clearpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+<<FA_SCORES_COMPARE, dependson='FA_SCORES'>>=
+# assemble and compare scores across different models
+
+@
+
+When adjusting for age, mean scores do not agree with raw scores
+<<FA_SCORES_MEANS, dependson='FA_SCORES', echo=T, results='markup'>>=
+# means
+means.ext
+means.int
+@
+
+Scores that do not adjust for age are very highly correlated, regardless of whether they come from a model with free or restricted intercepts
+<<FA_SCORES_CORR, dependson='FA_SCORES', echo=T, results='markup'>>=
+cor(compare.scores[,c(grep("EXT", names(compare.scores), value=T))])
+cor(compare.scores[,c(grep("INT", names(compare.scores), value=T))])
+@
+
+The gradient of scores by income quintile are very similar
+<<FA_SCORES_INEQ, dependson='FA_SCORES', echo=T, results='markup'>>=
+
+# free intercepts
+lm(EXTtl ~ factor(incq), data=compare.scores)$coefficients
+lm(INTtl ~ factor(incq), data=compare.scores)$coefficients
+
+# constrained intercepts
+lm(EXTtli ~ factor(incq), data=compare.scores)$coefficients
+lm(INTtli ~ factor(incq), data=compare.scores)$coefficients
+
+# constrained intercepts, common ages
+lm(EXTtlir ~ factor(incq), data=compare.scores)$coefficients
+lm(INTtlir ~ factor(incq), data=compare.scores)$coefficients
+
+@
+
+<<FA_REGS, dependson='FA_SCORES'>>=
+
+# combinations of regressions
+regcoeffs <- matrix( 
+  c(rep("BCS", 2), rep("MCS", 2), rep(c("M","F"), 2), rep(NA, 8)), # the data elements 
+  nrow=4,              # number of rows 
+  ncol=4,              # number of columns 
+  byrow = F)        # fill matrix by columns 
+colnames(regcoeffs) <- c("cohort", "sex", "EXTd", "INTd")
+
+# fill in differences
+regs.ext <- list()
+regs.int <- list()
+for (r in 1:nrow(regcoeffs)) {
+  regs.ext[[r]] <- lm(EXT ~ as.factor(incq) + age, data=subset(items.scored[[8]], cohort==regcoeffs[r,1] & sex==regcoeffs[r,2]))
+  regcoeffs[r,3] <- round(regs.ext[[r]]$coefficients["as.factor(incq)5"],2)
+  regs.int[[r]] <- lm(INT ~ as.factor(incq) + age, data=subset(items.scored[[8]], cohort==regcoeffs[r,1] & sex==regcoeffs[r,2]))
+  regcoeffs[r,4] <- round(regs.int[[r]]$coefficients["as.factor(incq)5"],2)
+}
+
+@
+
+% \begin{table}[ht]
+% \caption{Estimated means of scores -- non-age adjusted 4-group model} \label{tab:scoremeans}
+% \centering
+% <<FA_COMP1, dependson='FA_SCORES', results='asis'>>=
+% # compare mean factor scores by gender and cohort
+% 
+% prn <- function(obj) sprintf("%.3f", round(obj,3))
+% toaggr <- c("EXT", "INT", "EXTr", "INTr", "EXT.RAW", "INT.RAW", "EXT.RAWr", "INT.RAWr")
+% sctable <- aggregate(items.scored[[7]][, toaggr], list(items.scored[[7]]$cohortsex), 
+%                    FUN = function (x) prn(mean(x)))
+% 
+% sctable <- cbind(group = factor(c(1,3,2,4)), sctable)
+% levels(sctable$group) <- c("BCS Males", "MCS Males", "BCS Females", "MCS Females")
+% sctable <- sctable[order(sctable$group),]
+% sctable <- sctable[, !names(sctable) %in% "Group.1"]
+% sctable <- as.matrix(sctable,3)
+% 
+% 
+% print.xtable(xtable(sctable,
+%                     align="L{0cm}L{2.2cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}",
+%                     digits = 3
+%                     ),
+%              size="\\small",
+%              include.rownames=F,
+%              include.colnames=F,
+%              booktabs = T,
+%              add.to.row = list(pos = c(list(0,4)),
+%                               command = c(
+%                               # header
+%                                    "\\toprule
+%                                     & \\multicolumn{4}{c}{\\textbf{Scored factors}}
+%                                     & \\multicolumn{4}{c}{\\textbf{Scales}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
+%                                     & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-adjusted} & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-Adjusted} \\\\ \\cmidrule(lr){2-3} \\cmidrule(lr){4-5} \\cmidrule(lr){6-7} \\cmidrule(lr){8-9}
+%                       & EXT & INT & EXT & INT & EXT & INT & EXT & INT \\\\",
+%                       "\\bottomrule"
+%                                     )),
+%              hline.after = NULL,  # remove default booktabs lines
+%              sanitize.text.function = function(x){x},
+%              sanitize.rownames.function = function(x){""}, # to not print row numbers
+%              tabular.environment = "tabularx", width = "\\textwidth",
+%              floating = FALSE
+% )
+% 
+% @
+% \end{table}
+% 
+% \begin{table}[ht]
+% \caption{Estimated means of scores -- age adjusted 4-group model} \label{tab:scoremeans}
+% \centering
+% <<FA_COMP2, dependson='FA_SCORES', results='asis'>>=
+% # compare mean factor scores by gender and cohort
+% 
+% prn <- function(obj) sprintf("%.3f", round(obj,3))
+% toaggr <- c("EXT", "INT", "EXTr", "INTr", "EXT.RAW", "INT.RAW", "EXT.RAWr", "INT.RAWr")
+% sctable <- aggregate(items.scored[[8]][, toaggr], list(items.scored[[8]]$cohortsex), 
+%                    FUN = function (x) prn(mean(x)))
+% 
+% sctable <- cbind(group = factor(c(1,3,2,4)), sctable)
+% levels(sctable$group) <- c("BCS Males", "MCS Males", "BCS Females", "MCS Females")
+% sctable <- sctable[order(sctable$group),]
+% sctable <- sctable[, !names(sctable) %in% "Group.1"]
+% sctable <- as.matrix(sctable,3)
+% 
+% 
+% print.xtable(xtable(sctable,
+%                     align="L{0cm}L{2.2cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}",
+%                     digits = 3
+%                     ),
+%              size="\\small",
+%              include.rownames=F,
+%              include.colnames=F,
+%              booktabs = T,
+%              add.to.row = list(pos = c(list(0,4)),
+%                               command = c(
+%                               # header
+%                                    "\\toprule
+%                                     & \\multicolumn{4}{c}{\\textbf{Scored factors}}
+%                                     & \\multicolumn{4}{c}{\\textbf{Scales}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
+%                                     & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-adjusted} & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-Adjusted} \\\\ \\cmidrule(lr){2-3} \\cmidrule(lr){4-5} \\cmidrule(lr){6-7} \\cmidrule(lr){8-9}
+%                       & EXT & INT & EXT & INT & EXT & INT & EXT & INT \\\\",
+%                       "\\bottomrule"
+%                                     )),
+%              hline.after = NULL,  # remove default booktabs lines
+%              sanitize.text.function = function(x){x},
+%              sanitize.rownames.function = function(x){""}, # to not print row numbers
+%              tabular.environment = "tabularx", width = "\\textwidth",
+%              floating = FALSE
+% )
+% 
+% @
+% \end{table}
+% 
+% \begin{table}[ht]
+% \caption{Estimated means of scores -- 4-group model, 59-62 months} \label{tab:scoremeans}
+% \centering
+% <<FA_COMP3, dependson='FA_SCORES', results='asis'>>=
+% # compare mean factor scores by gender and cohort
+% 
+% prn <- function(obj) sprintf("%.3f", round(obj,3))
+% toaggr <- c("EXT", "INT", "EXTr", "INTr", "EXT.RAW", "INT.RAW", "EXT.RAWr", "INT.RAWr")
+% sctable <- aggregate(items.scored[[9]][, toaggr], list(items.scored[[9]]$cohortsex), 
+%                    FUN = function (x) prn(mean(x)))
+% 
+% sctable <- cbind(group = factor(c(1,3,2,4)), sctable)
+% levels(sctable$group) <- c("BCS Males", "MCS Males", "BCS Females", "MCS Females")
+% sctable <- sctable[order(sctable$group),]
+% sctable <- sctable[, !names(sctable) %in% "Group.1"]
+% sctable <- as.matrix(sctable,3)
+% 
+% 
+% print.xtable(xtable(sctable,
+%                     align="L{0cm}L{2.2cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}",
+%                     digits = 3
+%                     ),
+%              size="\\small",
+%              include.rownames=F,
+%              include.colnames=F,
+%              booktabs = T,
+%              add.to.row = list(pos = c(list(0,4)),
+%                               command = c(
+%                               # header
+%                                    "\\toprule
+%                                     & \\multicolumn{4}{c}{\\textbf{Scored factors}}
+%                                     & \\multicolumn{4}{c}{\\textbf{Scales}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
+%                                     & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-adjusted} & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-Adjusted} \\\\ \\cmidrule(lr){2-3} \\cmidrule(lr){4-5} \\cmidrule(lr){6-7} \\cmidrule(lr){8-9}
+%                       & EXT & INT & EXT & INT & EXT & INT & EXT & INT \\\\",
+%                       "\\bottomrule"
+%                                     )),
+%              hline.after = NULL,  # remove default booktabs lines
+%              sanitize.text.function = function(x){x},
+%              sanitize.rownames.function = function(x){""}, # to not print row numbers
+%              tabular.environment = "tabularx", width = "\\textwidth",
+%              floating = FALSE
+% )
+% 
+% @
+% \end{table}
+
+% \clearpage
+% \begin{figure}
+% \begin{center}
+% <<AGEHIST, dev='tikz', fig.height=6, fig.width=6, out.width='.6\\linewidth', cache=TRUE>>=
+% @
+% \end{center}
+% \caption{Distribution of age in the BCS and MCS sample}
+% {\scriptsize \emph{Notes}: The figure shows the density of observations at each month of age for each of the two cohorts.  \par}
+% \end{figure}
+% 
+% \clearpage
+% \begin{figure}
+% <<FACDENS, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
+% @
+% \caption{Density of Scored Factors}
+% {\scriptsize \emph{Notes}: The figure shows estimated kernel densities of the scored factors separately by cohort.  \par}
+% \end{figure}
+% 
+% \begin{figure}
+% <<FACLOESS_AGE, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
+% @
+% \caption{Scores and child age}
+% {\scriptsize \emph{Notes}: The figure plots LOESS of the scored factors against child age. \par}
+% \end{figure}
+% 
+% 
+% \begin{figure}
+% <<FACRAW, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
+% @
+% \caption{Raw vs Scored Factors}
+% {\scriptsize \emph{Notes}: The figure plots the scored factors on the vertical axis against the ``raw" scores on the horizontal axis, i.e.
+scores obtained by simply summing the number of positive answers to each of the items.
+\par}
+% \end{figure}
+% 
+% \begin{figure}
+% <<FACSCAT, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
+% @
+% \caption{Scatterplots of factor scores by cohort}
+% {\scriptsize \emph{Notes}: The figure plots the scored factors separately by cohort. The externalising and internalising scores are plotted against one another. \par}
+% \end{figure}
+% 
+% \begin{figure}
+% <<FACINEQ, dev='tikz', fig.height=12, fig.width=10, out.width='1\\linewidth'>>=
+% @
+% \caption{Distribution of scores across income quintiles}\label{fig:boxscores}
+% {\scriptsize \emph{Notes}: The figure plots the distribution of the scored factors across the quintiles in the income distribution. \par}
+% \end{figure}
+% 
+% \begin{figure}
+% <<FACLOESS_INC, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
+% @
+% \caption{Scores and family income}
+% {\scriptsize \emph{Notes}: The figure plots LOESS of the scored factors against family income. \par}
+% \end{figure}
+
+\clearpage
+\bibliographystyle{chicago}
+\bibliography{/Users/giacomomason/Documents/BibTeX/Papers-cohorts_nc}
+
+\clearpage
+\begin{appendices}
+
+\section{Additional fit comparisons}\label{sec:app_fit}
+
 \begin{table}[ht]
-\caption{Estimated loadings and thresholds} \label{tab:allpars}
+\centering
+\small
+\caption{Fit indices -- 2-group models}\label{tab:fit2}
+<<FA_FIT_RES2, dependson="FA_FIT", results='asis'>>=
+require(xtable)
+
+table <- rbind(afitab[[1]],afitab[[2]],afitab[[3]],afitab[[4]],afitab[[5]],afitab[[6]])
+
+print(xtable(table,
+                    align="L{0cm}L{.20\\textwidth}C{.03\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
+                    digits = 4
+                    ),
+             size="\\small",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0,3,6,9,12,15)),
+                              command = c(
+                              # header
+            "Model & Num. params & $\\chi^2$ & RMSEA & MFI & CFI & G-hat & \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ \\midrule
+                       \\textbf{(1) No gender split, not age adjusted} \\\\",
+                      "\\newline \\textbf{(2) No gender split, age adjusted} \\\\ \\midrule ",
+                      "\\newline \\textbf{(3) Males only, not age adjusted} \\\\ \\midrule ",
+                      "\\newline \\textbf{(4) Females only, not age adjusted} \\\\ \\midrule ",
+                      "\\newline \\textbf{(5) Males only, age adjusted} \\\\ \\midrule ",
+                      "\\newline \\textbf{(6) Females only, age adjusted} \\\\ \\midrule "
+                                    )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = "\\textwidth",
+             floating = FALSE
+)
+
+@
+\end{table}
+
+\begin{table}[ht]
+\centering
+\small
+\caption{Fit indices -- 4-group models}\label{tab:fit4}
+<<FA_FIT_RES4, dependson="FA_FIT", results='asis'>>=
+require(xtable)
+
+table <- rbind(afitab[[7]],afitab[[8]],afitab[[9]],afitab[[11]])
+
+print(xtable(table,
+                    align="L{0cm}L{.20\\textwidth}C{.03\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
+                    digits = 4
+                    ),
+             size="\\small",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0,3,6,9)),
+                              command = c(
+                              # header
+            "Model & Num. params & $\\chi^2$ & RMSEA & MFI & CFI & G-hat & \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ \\midrule
+                      \\newline \\textbf{(7) not age adjusted} \\\\",
+                      "\\newline \\textbf{(8) age adjusted} \\\\\\midrule ",
+                      "\\newline \\textbf{(9) 59-62m range} \\\\ \\midrule ",
+                      "\\newline \\textbf{(11) MIMIC in age} \\\\ \\midrule "
+                                    )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = "\\textwidth",
+             floating = FALSE
+)
+
+@
+\end{table}
+
+\clearpage
+\section{Latent variable mean structures}\label{sec:app_meancov}
+
+\begin{table}[ht]
+\caption{Estimated latent variable means and covariances -- 4 group models} \label{tab:lvpars4}
+\centering
+<<RES_FA_LV4, dependson="FA_FINAL_PARS", results='asis'>>=
+require(xtable)
+
+table <- rbind(lvpars[[7]][[1]],lvpars[[7]][[2]],
+               lvpars[[8]][[1]],lvpars[[8]][[2]],
+               lvpars[[9]][[1]],lvpars[[9]][[2]]
+)
+
+print.xtable(xtable(table,
+                    align="L{0cm}L{3cm}C{1.2cm}C{1.2cm}C{1.2cm}C{1.2cm}C{1.2cm}C{1.2cm}C{1.2cm}C{1.2cm}",
+                    digits = 3
+                    ),
+             size="\\small",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0,2,4,6,8,10,12)),
+                              command = c(
+                              # header
+                                   "\\toprule
+                                    & \\multicolumn{4}{c}{\\textbf{BCS}}
+                                    & \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
+                                    & Mean & \\multicolumn{2}{c}{Covariance} & Corr. & Mean & \\multicolumn{2}{c}{Covariance} & Corr. \\\\ \\midrule
+                      \\textbf{(7) Males (joint, not age adj.)} \\\\ ",
+                      "\\newline \\textbf{(7) Females (joint, not age adj.)} \\\\ ",
+                      "\\midrule \\newline \\textbf{(8) Males (joint, age adj.)} \\\\ ",
+                      "\\newline \\textbf{(8) Females (joint, age adj.)} \\\\ ",
+                      "\\midrule \\newline \\textbf{(9) Males (joint, 59-62m)} \\\\ ",
+                      "\\newline \\textbf{(9) Females (joint, 59-62m)} \\\\ ",
+                      "\\bottomrule"
+                                    )),
+             hline.after = NULL,  # remove default booktabs lines
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabularx", width = "\\textwidth",
+             floating = FALSE
+)
+
+@
+\end{table}
+
+\begin{table}[ht]
+\caption{Estimated latent variable means and covariances -- 2 group models} \label{tab:lvpars2}
+\centering
+<<RES_FA_LV2, dependson="FA_FINAL_PARS", results='asis'>>=
+require(xtable)
+
+table <- rbind(lvpars[[1]],lvpars[[2]],
+               lvpars[[3]],lvpars[[4]],
+               lvpars[[4]],lvpars[[6]]
+)
+
+print.xtable(xtable(table,
+                    align="L{0cm}L{3cm}C{1.2cm}C{1.2cm}C{1.2cm}C{1.2cm}C{1.2cm}C{1.2cm}C{1.2cm}C{1.2cm}",
+                    digits = 3
+                    ),
+             size="\\small",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0,2,4,6,8,10,12)),
+                              command = c(
+                              # header
+                                   "\\toprule
+                                    & \\multicolumn{4}{c}{\\textbf{BCS}}
+                                    & \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
+                                    & Mean & \\multicolumn{2}{c}{Covariance} & Corr. & Mean & \\multicolumn{2}{c}{Covariance} & Corr. \\\\ \\midrule
+                      \\textbf{(1) No gender split, no age adj.} \\\\",
+                      "\\textbf{(2) No gender split, age adj.} \\\\",
+                      "\\midrule \\newline \\textbf{(3) Males, no age adj.} \\\\ ",
+                      "\\newline \\textbf{(4) Females, no age adj.} \\\\ ",
+                      "\\midrule \\newline \\textbf{(5) Males, age adj.} \\\\ ",
+                      "\\newline \\textbf{(6) Females, age adj.} \\\\ ",
+                      "\\bottomrule"
+                                    )),
+             hline.after = NULL,  # remove default booktabs lines
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabularx", width = "\\textwidth",
+             floating = FALSE
+)
+
+@
+\end{table}
+
+\clearpage
+\section{Estimated model parameters}
+
+\begin{table}[ht]
+\caption{Estimated loadings and thresholds -- 2-group models with no age adjustment} \label{tab:allpars2a}
 \resizebox{\textwidth}{!}{
-<<RES_FA_PAR, dependson="FA_FINAL_PARS", results='asis'>>=
+<<RES_FA_PAR2A, dependson="FA_FINAL_PARS", results='asis'>>=
 require(xtable)
 
-table <- rbind(allpar[[1]],allpar[[2]],allpar[[3]])
+table <- rbind(allpar[[1]],allpar[[3]],allpar[[4]])
 
@@
                                             \\\\ \\midrule
-                      \\newline \\textbf{No gender split} \\\\",
+                      \\newline \\textbf{(1) No gender split} \\\\",
                       # additional separators
-                      "\\newline \\textbf{Males only} \\\\ \\midrule ",
-                      "\\newline \\textbf{Females only} \\\\ \\midrule " 
+                      "\\newline \\textbf{(3) Males only} \\\\ \\midrule ",
+                      "\\newline \\textbf{(4) Females only} \\\\ \\midrule "
                       )),
@@
 \begin{table}[ht]
-\caption{Estimated loadings and thresholds -- 4-group model} \label{tab:allpars2}
+\caption{Estimated loadings and thresholds -- 2-group models with age adjustment} \label{tab:allpars2b}
 \resizebox{\textwidth}{!}{
-<<RES_FA_PAR2, dependson="FA_FINAL_PARS", results='asis'>>=
+<<RES_FA_PAR2B, dependson="FA_FINAL_PARS", results='asis'>>=
 require(xtable)
 
-table <- allpar[[4]]
+table <- rbind(allpar[[2]],allpar[[5]],allpar[[6]])
 
@@
                     # align = "ll|cccc|cccc"
-                    align="L{0cm}L{1.5cm}C{2cm}C{1.5cm}C{1.5cm}C{1.8cm}C{1.8cm}C{1.8cm}C{1cm}",
+                    align="L{0cm}L{2.5cm}C{2.5cm}C{2.5cm}C{2.5cm}C{2.5cm}C{1cm}",
                     digits = 3
@@
              booktabs = T,
-             add.to.row = list(pos = c(list(0)),
+             add.to.row = list(pos = c(list(0,11,22)),
+                              command = c(
+                              # header
+                                          " & \\textbf{Loadings}
+                                            & \\multicolumn{2}{c}{\\textbf{Thresholds}}
+                                            & \\textbf{Variances}
+                                            & \\\\ \\cmidrule(lr){2-2} \\cmidrule(lr){3-4} \\cmidrule(lr){5-5}
+
+                                            & $\\bm{\\lambda}$
+                                            & $\\bm{\\tau}_1$
+                                            & $\\bm{\\tau}_2$
+                                            & $\\bm{\\Var(\\varepsilon)}$ \\\\
+                                            Measure & BCS = MCS
+                                            & BCS = MCS
+                                            & BCS = MCS
+                                            & MCS (BCS=1)
+                                            & Factor
+                                            \\\\ \\midrule
+                      \\newline \\textbf{(2) No gender split} \\\\",
+                      # additional separators
+                      "\\newline \\textbf{(5) Males only} \\\\ \\midrule ",
+                      "\\newline \\textbf{(6) Females only} \\\\ \\midrule "
+                      )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabularx", width = "\\textwidth",
+             floating = FALSE
+)
+
+@
+}
+\end{table}
+
+\begin{table}[ht]
+\caption{Estimated loadings and thresholds -- 4-group models} \label{tab:allpars4}
+\resizebox{\textwidth}{!}{
+<<RES_FA_PAR4, dependson="FA_FINAL_PARS", results='asis'>>=
+require(xtable)
+
+table <- rbind(allpar[[7]], allpar[[8]], allpar[[9]])
+
+# sequences to add rows
+rws <- seq(2, (nrow(table)-1), by = 1) # start from row 2 (exclude model titles)
+rwsl <- as.list(rws)
+
+print.xtable(xtable(table,
+                    # align = "ll|cccc|cccc"
+                    align="L{0cm}L{2.2cm}C{1.8cm}C{1.4cm}C{1.4cm}C{1.7cm}C{1.7cm}C{1.7cm}C{1cm}",
+                    digits = 3
+                    ),
+             size="\\small",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0,11,22)),
                               command = c(
@@
                                             Measure
-                                            & All groups
-                                            & All groups
-                                            & All groups
-                                            & BCS Males
-                                            & MCS Females
-                                            & MCS Males
+                                            & All
+                                            & All
+                                            & All
+                                            & BCS M
+                                            & MCS F
+                                            & MCS M
                                             & Factor
-                                            \\\\"
+                                            \\\\ \\midrule 
+                                          \\newline \\textbf{(7) not age adjusted} \\\\"
+                                          ,
+                                          "\\newline \\textbf{(8) age adjusted} \\\\ \\midrule ",
+                                          "\\newline \\textbf{(9) 59-62m} \\\\ \\midrule "
+
                       )),
@@
 
-\begin{table}[ht]
-\caption{Estimated latent variable means and covariances} \label{tab:lvpars}
-\centering
-<<RES_FA_LV, dependson="FA_FINAL_PARS", results='asis'>>=
-require(xtable)
-
-table <- rbind(lvpars[[1]],lvpars[[2]],lvpars[[3]],lvpars[[4]],lvpars[[5]])
-
-print.xtable(xtable(table,
-                    align="L{0cm}L{2.2cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}",
-                    digits = 3
-                    ),
-             size="\\small",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = c(list(0,2,4,6,8,10)),
-                              command = c(
-                              # header
-                                   "\\toprule
-                                    & \\multicolumn{4}{c}{\\textbf{BCS}}
-                                    & \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
-                                    & Mean & \\multicolumn{2}{c}{Covariance} & Corr. & Mean & \\multicolumn{2}{c}{Covariance} & Corr. \\\\ \\midrule
-                      \\textbf{No gender split} \\\\",
-                      # additional separators
-                      "\\newline \\textbf{Males only} \\\\ ",
-                      "\\newline \\textbf{Females only} \\\\ ",
-                      "\\midrule \\newline \\textbf{Males (joint)} \\\\ ",
-                      "\\newline \\textbf{Females (joint)} \\\\ ",
-                      "\\bottomrule"
-                                    )),
-             hline.after = NULL,  # remove default booktabs lines
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabularx", width = "\\textwidth",
-             floating = FALSE
-)
-
-@
-\end{table}
-
-<<CLEANUP, dependson='FA_FINAL'>>=
-@
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \clearpage
-\subsection{Inequality in noncognitive ability}
-The model can be used to investigate how inequality of noncognitive skills has changed between the two cohorts. 
-%The variances of the $EXT$ and $INT$ factors are estimated to be larger in the MCS children by \Sexpr{round(lv.covs$MCS[1,1]-1,2)} and \Sexpr{round(lv.covs$MCS[2,2]-1,2)} of a standard deviation, respectively, indicating larger dispersion.
-
-<<FA_SCORES, dependson='FA_FINAL'>>=
-@
-
-<<FA_REGS, dependson='FA_SCORES'>>=
-
-# combinations of regressions
-regcoeffs <- matrix( 
-  c(rep("BCS", 2), rep("MCS", 2), rep(c("M","F"), 2), rep(NA, 8)), # the data elements 
-  nrow=4,              # number of rows 
-  ncol=4,              # number of columns 
-  byrow = F)        # fill matrix by columns 
-colnames(regcoeffs) <- c("cohort", "sex", "EXTd", "INTd")
-
-# fill in differences
-regs.ext <- list()
-regs.int <- list()
-for (r in 1:nrow(regcoeffs)) {
-  regs.ext[[r]] <- lm(EXT ~ as.factor(incq) + age, data=subset(items.scored[[4]], cohort==regcoeffs[r,1] & sex==regcoeffs[r,2]))
-  regcoeffs[r,3] <- round(regs.ext[[r]]$coefficients["as.factor(incq)5"],2)
-  regs.int[[r]] <- lm(INT ~ as.factor(incq) + age, data=subset(items.scored[[4]], cohort==regcoeffs[r,1] & sex==regcoeffs[r,2]))
-  regcoeffs[r,4] <- round(regs.int[[r]]$coefficients["as.factor(incq)5"],2)
-}
-
-@
-
-\begin{table}[ht]
-\caption{Estimated means and covariances of scores} \label{tab:scoremeans}
-\centering
-<<FA_COMP, dependson='FA_SCORES', results='asis'>>=
-# compare mean factor scores by gender and cohort
-
-prn <- function(obj) sprintf("%.3f", round(obj,3))
-
-sctable <- aggregate(items.scored[[4]][, c("EXT", "INT", "EXT.RAW", "INT.RAW")], list(items.scored[[4]]$cohortsex), 
-                   FUN = function (x) c(mn = prn(mean(x)), med = prn(median(x))))
-sctable <- cbind(group = factor(c(1,3,2,4)), sctable)
-levels(sctable$group) <- c("BCS Males", "MCS Males", "BCS Females", "MCS Females")
-sctable <- sctable[order(sctable$group),]
-sctable <- sctable[, !names(sctable) %in% "Group.1"]
-sctable <- as.matrix(sctable,3)
-
-print.xtable(xtable(sctable,
-                    align="L{0cm}L{2.2cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}",
-                    digits = 3
-                    ),
-             size="\\small",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = c(list(0,4)),
-                              command = c(
-                              # header
-                                   "\\toprule
-                                    & \\multicolumn{4}{c}{\\textbf{Scored factors}}
-                                    & \\multicolumn{4}{c}{\\textbf{Scales}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
-                                    & \\multicolumn{2}{c}{EXT} & \\multicolumn{2}{c}{INT} & \\multicolumn{2}{c}{EXT} & \\multicolumn{2}{c}{INT} \\\\ \\cmidrule(lr){2-3} \\cmidrule(lr){4-5} \\cmidrule(lr){6-7} \\cmidrule(lr){8-9}
-                      & Mean & Median & Mean & Median & Mean & Median & Mean & Median \\\\",
-                      "\\bottomrule"
-                                    )),
-             hline.after = NULL,  # remove default booktabs lines
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabularx", width = "\\textwidth",
-             floating = FALSE
-)
-
-@
-\end{table}
+\section{Inequality}
 
@@
 
-We can derive scores for $\theta^{EXT}$ and $\theta^{INT}$ directly, and plot them against socioeconomic measures observed in the data.
-Our main socioeconomic indicator is a measure of weekly total family net income in the prices of the interview year.\footnote{This measure is harmonised across cohorts and derived as part of CLOSER, which is funded by the ESRC.} Using the estimated parameters for the model with invariant thresholds and loadings (but non-invariant intercepts), I compute the scores for each individual in the sample (using the Empirical Bayes modal approach).
-In \autoref{fig:boxscores}, we can see that the gradient in noncognitive skills along family income quintiles is more pronounced in the MCS cohort.
-In particular, a male five-year old at the top quintile of the income distribution in 1975 has a better score by \Sexpr{regcoeffs[1,3]} in the externalising dimension and \Sexpr{regcoeffs[1,4]} in the internalising dimension, compared to a child at the bottom quintile.
-In 2005, the differences amount to \Sexpr{regcoeffs[3,3]} and \Sexpr{regcoeffs[3,4]}, respectively.
-
-\clearpage
-\begin{figure}
-\begin{center}
-<<AGEHIST, dev='tikz', fig.height=6, fig.width=6, out.width='.6\\linewidth', cache=TRUE>>=
-@
-\end{center}
-\caption{Distribution of age in the BCS and MCS sample}
-{\scriptsize \emph{Notes}: The figure shows the density of observations at each month of age for each of the two cohorts.  \par}
-\end{figure}
-
-\clearpage
-\begin{figure}
-<<FACDENS, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
-@
-\caption{Density of Scored Factors}
-{\scriptsize \emph{Notes}: The figure shows estimated kernel densities of the scored factors separately by cohort.  \par}
-\end{figure}
-
-\begin{figure}
-<<FACLOESS_AGE, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
-@
-\caption{Scores and child age}
-{\scriptsize \emph{Notes}: The figure plots LOESS of the scored factors against child age. \par}
-\end{figure}
-
-\begin{figure}
-<<FACRAW, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
-@
-\caption{Raw vs Scored Factors}
-{\scriptsize \emph{Notes}: The figure plots the scored factors on the vertical axis against the ``raw" scores on the horizontal axis, i.e.
-scores obtained by simply summing the number of positive answers to each of the items.
-\par}
-\end{figure}
-
-\begin{figure}
-<<FACSCAT, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
-@
-\caption{Scatterplots of factor scores by cohort}
-{\scriptsize \emph{Notes}: The figure plots the scored factors separately by cohort. The externalising and internalising scores are plotted against one another. \par}
-\end{figure}
-
-\begin{figure}
-<<FACINEQ, dev='tikz', fig.height=12, fig.width=10, out.width='1\\linewidth'>>=
-@
-\caption{Distribution of scores across income quintiles}\label{fig:boxscores}
-{\scriptsize \emph{Notes}: The figure plots the distribution of the scored factors across the quintiles in the income distribution. \par}
-\end{figure}
-
-\begin{figure}
-<<FACLOESS_INC, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
-@
-\caption{Scores and family income}
-{\scriptsize \emph{Notes}: The figure plots LOESS of the scored factors against family income. \par}
-\end{figure}
-
-\clearpage
-\bibliographystyle{chicago}
-\bibliography{/Users/giacomomason/Documents/BibTeX/Papers-cohorts_nc}
+\end{appendices}
 

### 2018-02-23 codes/R/2_efa.R
added internal consistency checks

@@ -13,12 +13,26 @@ bcsfa <- fa(r=bcscor, nfactors=2, rotate = rotation, fm="wls")
 
 
 ## ---- BCSEFARES
+# Scree
+ns.bcs <- nScree(bcscor)$Components
 # VSS for number of factors
-nScree(bcscor)
-VSS(bcscor, plot=F, rotate=rotation, n.obs = nbcs)
+vss.bcs <- VSS(bcscor, plot=F, rotate=rotation, n.obs = nbcs)
+#combine
+efatab.bcs <- rbind(t(ns.bcs), which.max(vss.bcs$cfit.1), which.max(vss.bcs$cfit.2))
+rownames(efatab.bcs) <- c("Optimal Coordinates", "Acceleration Factor", "Parallel Analysis", "Kaiser", "VSS Compl. 1", "VSS Compl. 2")
+efatab.bcs
+
 # FACTOR ANALYSIS loadings
 bcsfa$loadings
 
+## ---- BCSREL
+# reliability 
+fz.bcs <- fisherz(bcscor)
+fz.bcs[fz.bcs==Inf] <- NA 
+fz.bcs.ext <- fz.bcs[1:6,1:6]
+mr.bcs.ext <- fisherz2r(mean(fz.bcs.ext, na.rm=T))
+fz.bcs.int <- fz.bcs[7:11,7:11]
+mr.bcs.int <- fisherz2r(mean(fz.bcs.int, na.rm=T))
 
 ## ---- MCSEFA
 
@@ -30,9 +44,25 @@ mcsfa <- fa(r=mcscor, nfactors=2, rotate = rotation, fm="wls")
 
 ## ---- MCSEFARES
 # VSS for number of factors
-nScree(mcscor)
-VSS(mcscor, plot=F, rotate=rotation, n.obs = nmcs)
+ns.mcs <- nScree(mcscor)$Components
+ns.mcs
+# VSS for number of factors
+vss.mcs <- VSS(mcscor, plot=F, rotate=rotation, n.obs = nmcs)
+
+#combine
+efatab.mcs <- rbind(t(ns.mcs), which.max(vss.mcs$cfit.1), which.max(vss.mcs$cfit.2))
[…]

### 2018-02-23 codes/R/4_scores.R
moved selection of score model to graphs file

@@ -17,6 +17,4 @@ for (i in 7:11) { # only specifications with separate gender groups
   
 }
 
-# select scores to plot
-scores2plot <- items.scored[[8]]
 

### 2018-02-23 codes/R/5_graphs.R

@@ -1,3 +1,7 @@
+############################################################################################
+## ---- FA_SCORES_SELECT
+# select scores to plot
+scores2plot <- items.scored[[8]]
 
 ############################################################################################
 ## ---- AGEHIST

### 2018-02-23 codes/R/cohorts_nc.Rnw
selection of score model in separate chunk

@@
 
-For BCS, I augment the 20-item Rutter Scale with 4 other parent-reported questions from the parental questionnaire. 
+For BCS, I augment the 19-item Rutter Scale with 4 other parent-reported questions from the parental questionnaire. 
 \begin{enumerate}
@@
 The five-factor structure at the basis of the full SDQ has been validated in many contexts \citep{Stone2010}; however, lower-dimensional structures have been also suggested \citep{Dickey2004}, and recent research has shown that there are some benefits to using broader subscales that correspond to the externalising and internalising factors in Rutter \citep{Goodman2010}.
-To assess whether the two-factor conceptualisation suggested by the theory is valid in our ``reduced" 12-item scale, we conduct exploratory factor analysis (EFA) on the new 11-item scale separately for each cohort sample.
+To assess whether the two-factor conceptualisation suggested by the theory is valid in our ``reduced" 11-item scale, we conduct exploratory factor analysis (EFA) on the new 11-item scale separately for each cohort sample.
 Different methodologies for quantifying the optimal number of factors produce inconclusive results, but both Velicer MAP and very simple structure (VSS) seem to indicate that a two-factor structure is suitable.
@@
 \begin{split}
-X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{EXT}_{cj} + u_{cji} \qquad \text{for} \; i=1,\dots, 6 \\
-X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{INT}_{cj} + u_{cji} \qquad \text{for} \; i=7,\dots, 11
+X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{EXT}_{jc} + u_{ijc} \qquad \text{for} \; i=1,\dots, 6 \\
+X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{INT}_{jc} + u_{ijc} \qquad \text{for} \; i=7,\dots, 11
 \end{split}
@@
 Given \eqref{eq:propens}, the observed outcome as a function of the propensities $X^*$ can be defined as a function of additional threshold parameters $\tau$, as follows:
-$$X_{cji} = s \qquad \text{if} \; \tau_{s,ci} \leq X^*_{cji} < \tau_{s+1,ci} \qquad \text{for} \; s=0,1,2$$
-with $\tau_{0,ci}=-\infty$ and $\tau_{3,ci}=+\infty$.
+$$X_{ijc} = s \qquad \text{if} \; \tau_{s,ic} \leq X^*_{ijc} < \tau_{s+1,ic} \qquad \text{for} \; s=0,1,2$$
+with $\tau_{0,ic}=-\infty$ and $\tau_{3,ic}=+\infty$.
 The probability of observing each category for a binary item takes values:
@@
 
-For all groups: $$\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}, \quad \bm{\nu}=\bm{0}, \quad \text{and} \; \text{diag}\bm{\Psi}=\bm{I}.$$
+For all groups: $$\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}, \quad \bm{\nu}=\bm{0}, \quad \text{and} \; \text{diag}(\bm{\Psi})=\bm{I}.$$
 
@@
 <<FA_INVARIANCE, dependson='LOAD_DATA'>>=
+# estimate models of different levels of invariance
 @
@@
 <<FA_FIT, dependson='FA_INVARIANCE'>>=
+# assemble fit indices into matrices
 @
@@
 <<FA_FINAL, dependson='FA_INVARIANCE'>>=
+# select final model set
 @
@@
 <<FA_FINAL_PARS, dependson='FA_FINAL'>>=
+# get parameters of final model
 @
@@
 <<FA_SCORES, dependson='FA_FINAL'>>=
+# produce scores for different models
+@
+
+<<FA_SCORES_COMPARE, dependson='FA_SCORES'>>=
+# assemble and compare scores across different models
+@
+
+<<FA_SCORES_SELECT, dependson='FA_FINAL'>>=
+# select model to use for inequality analysis
 @

### 2018-02-27 codes/R/slides/cohorts_nc_slides.Rnw
slides for NCDS conf

<<MAIN,cache=FALSE,echo=FALSE,results='hide',message=FALSE,warning=FALSE>>=
rm(list=ls())         # Clear all objects from memory 
  opts_chunk$set(concordance=FALSE,
                 echo=FALSE,
                 tidy=FALSE,
                 autodep=TRUE,
                 cache=T, # REPLACE LATER
                 message=FALSE, 
                 warning=FALSE,
                 size='scriptsize',
                 results='hide',
                 comment=NA, 
                 fig.width=4,
                 fig.height=4, 
                 out.width='\\textwidth',
                 fig.show='hold',
                 tikzDefaultEngine='pdftex',
                 tikzDocumentDeclaration = "\\documentclass[10pt]{article}",
                 tikzMetricPackages = c("\\usepackage[utf8]{inputenc}","\\usepackage[T1]{fontenc}", "\\usetikzlibrary{calc}", "\\usepackage{amssymb}")
                 )

@

<<READ, cache=FALSE >>=
read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/1_load.R")
read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/2_efa.R")
read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/3_mifa_invariance.R")
read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/3_mifa_final.R")
read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/4_scores.R")
read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/4_scores_compare.R")
read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/5_graphs.R")

@

<<PREAMBLE>>=
@

<<LOAD_DATA>>=
@

\documentclass[10pt,dvipsnames]{beamer}
\usepackage{mathpazo}  % serif math
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage[absolute,overlay]{textpos}
\usepackage{booktabs}
\usepackage{bm}        % bold maths
\usepackage{dsfont}
\usepackage{ragged2e}
\usepackage{multirow}
\usepackage{natbib}
\usepackage[outdir=./]{epstopdf}
\usepackage{accents}  % accents in math mode
\usepackage{tabularx}
\usepackage{adjustbox}
\usepackage{caption}
\usefonttheme[onlymath]{serif}
\usepackage{textgreek}

%tikz library
\usepackage{tikz}
\usetikzlibrary{intersections,calc,arrows,shapes}
\tikzset{>=latex}

% stata folder path
\newcommand{\statapics}{/Users/giacomomason/Dropbox/UCL/Whitehall/stata/w_graphs}

% additional symbols
\newcommand{\me}{\mathrm{e}}
\newcommand{\veps}{\varepsilon}
\newcommand{\E}{\mathbb{E}}
\newcommand{\1}{\mathds{1}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\Corr}{\mathrm{Corr}}
\newcommand{\vect}{\mathrm{vec}}
\newcommand{\vecht}{\mathrm{vech}}
\newcommand{\ubar}[1]{\underaccent{\bar}{#1}}

% additional commands
\newcommand{\be}{\begin{eqnarray}}
\newcommand{\ee}{\end{eqnarray}}
\newcommand{\bi}{\begin{itemize}}
\newcommand{\ei}{\end{itemize}}
\newcommand{\bnm}{\begin{enumerate}}
\newcommand{\enm}{\end{enumerate}}

% UCL beamer template
\definecolor{steelblue}{RGB}{70,130,180}
\definecolor{orangeucl}{RGB}{255,170,40}
\usetheme{ucl}
\setbeamertemplate{footline}[author title date]
\useoutertheme[small]{ucltitlebanner}
\setbeamertemplate{banner}[ucl][1]
\setbeamercolor{banner stripe}{bg=steelblue!70,fg=black}
\setbeamercolor{block body}{bg=steelblue!30,fg=black}
% override default square sizes to make smaller
\defbeamertemplate*{itemize item}{mysquare}{\tiny\raise0.5pt\hbox{$\blacksquare$}}
\defbeamertemplate*{itemize subitem}{mysquare}{\tiny\raise0.5pt\hbox{$\blacksquare$}}
\defbeamertemplate*{itemize subsubitem}{mysquare}{\tiny\raise0.5pt\hbox{$\blacksquare$}}
\setbeamertemplate{itemize item}[mysquare]               % override default square sizes
\setbeamertemplate{itemize subitem}[mysquare]
\setbeamertemplate{itemize subsubitem}[mysquare]

% define new columns
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\renewcommand{\arraystretch}{1.2}%

\setlength{\footnotesep}{0.4cm}

% reduced font size
\newcommand\Fontvi{\fontsize{8}{8}\selectfont} % reduce font size on slide
\newcommand\Fontvii{\fontsize{6}{6}\selectfont} % reduce font size on slide

% print TOC at every section break
\AtBeginSection[] { 
   \begin{frame}[noframenumbering]
   \frametitle{Outline}
   \tableofcontents[currentsection]
\end{frame}}

\title[Noncognitive skills in cohorts]{\bf{Comparability of child noncognitive skills across UK birth cohorts}}

\author[Mason]{Orazio Attanasio \and Richard Blundell \\ Gabriella Conti \and Giacomo Mason  }
\institute[UCL Economics]{UCL Department of Economics}

\date[HCRG]{Human Capital Reading Group \\ \today}

\begin{document}

\frame{\titlepage}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% INTRO
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Motivation}
\framesubtitle{The role of noncognitive skills}
\bi
\item Traditional labour literature used to focus mainly on cognitive skills
    \bi 
    \item returns to ability in adulthood (labour market, health, \dots)
    \item process of child development
    \ei
\item Attention broadened to include \emph{non}-cognitive skills \citep{Cunha2007a,Cunha2010}
    \bi
    \item Socio-emotional, behavioural problems, character, personality traits, locus of control, grit, \dots
    \item Found to be very important in shaping adult outcomes \citep{Heckman2006,Carneiro2007,Heckman2012b}
    \item Malleable until later childhood
    \item Primary mechanism of transmission of effects from early life programmes to adult outcomes \citep{Heckman2013,Kautz2014a}
    \ei
\item Today, focus on behaviour problems and mental health
\ei
\end{frame}

\begin{frame}
\frametitle{Motivation}
\framesubtitle{Behavioural problems}

... something more on how behavioural problems are defined and fit within the literature on noncognitive skills

\end{frame}

\begin{frame}
\frametitle{Motivation}
\framesubtitle{Inequality in skills}

\bi
\item Socioeconomic gradients in behaviour problems are documented, and are persistent across childhood/adolescence \citep{Campbell1995,Segal2008}
\item Still not clear 
    \bi
    \item how these gradients change across time
    \item whether the returns to behavioural skills evolve as well
    \ei
\item \emph{Measurement} is a key (often overlooked) issue
    \bi
    \item Usually, behavioural constructs are measured via scales
    \item Most research uses (normalised) aggregate scores from scales \citep{Blanden2007}
    \item But does the same scale measure the same constructs in the same way:
      \bi 
      \item across different populations?
      \item in the same population, at different points in time?
      \ei
    \item And what if only different scales are available?
    \ei
\ei

\end{frame}

\begin{frame}
\frametitle{This study}

\bi
\item Focus on behavioural problems in two British cohorts
  \bi \item British Cohort Study (BCS, b. 1970)
  \item Millennium Cohort Study (MCS, b. 2000/1) \ei
\item Construct a common 11-item, 2-dimensional scale from different instruments
\item Investigate its psychometric properties (measurement invariance)
\item Use scores from the new scale to document changes in socioeconomic gradients between cohorts 30 years apart
\ei

\end{frame}

\begin{frame}
\frametitle{Outline}
  \tableofcontents
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% DATA
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Data}

\begin{frame}
\frametitle{Data}
\framesubtitle{British Cohort Study}

\begin{columns}

\begin{column}{0.6\textwidth}
\bi
\item Data from 5 year follow-up, maternal questionnaire
\item Rutter A scale \citep{Rutter1970}
  \bi
  \item 19 items (behaviours)
  \item Options: \textit{Doesn't apply, Applies somewhat, Certainly applies}
  \item Usually conceptualised as measuring two dimensions
  \bi \item \emph{externalising} behaviour problems (anti-social) \item \emph{internalising} behaviour problems (neurotic) \ei
  \ei
  \item Additional questions from maternal module
\ei
\end{column}

\begin{column}{0.4\textwidth}
\includegraphics[width=\textwidth]{figs/bcsrut} 
\includegraphics[width=\textwidth]{figs/bcsrut2} 
\end{column}

\end{columns}

\end{frame}

\begin{frame}
\frametitle{Data}
\framesubtitle{Millennium Cohort Study}

\begin{columns}

\begin{column}{0.5\textwidth}

\bi
\item Data from 5 year follow-up (MCS3), maternal questionnaire
\item Complex survey design: need to rebalance data to make representative
\item Strengths and Difficulties Questionnaire (SDQ, \citealp{Goodman1994})
  \bi
  \item ``Updated" and expanded version of the Rutter scale 
  \item 25 items (behaviours)
  \item Options: \textit{Not true, Somewhat true, Certainly true}
  \item Designed with 5 subscales
  \item Broader subscales are sometimes justifiable \citep{Goodman2010}
  \ei
\ei

\end{column}

\begin{column}{0.5\textwidth}
\includegraphics[width=\textwidth]{figs/mcssdq} 
\end{column}

\end{columns}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% ITEM CHOICE and MEAN TABLE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Harmonisation}

\begin{frame}
\frametitle{Harmonisation}
\framesubtitle{Choice of items}

\bi
\item Aim: develop a new scale using only items that have similar wording across the two cohorts
\item Try to preserve three categories to maximise information
\item Combine items when needed (e.g. when SDQ items are positively coded)
\item Recode all items to higher categories for ``better" behaviours
\ei

\end{frame}

<<MEANTABLE>>=
# make table with means
@

\begin{frame}
\frametitle{Harmonisation}
\framesubtitle{Items}
\centering
\resizebox{!}{4cm}{
<<ITEM_TEXT, dependson="MEANTABLE", results='asis'>>=
require(xtable)

table <- meantab
table <- table[,!names(meantab) %in% c("cats", "fac", "bcs.num", "mcs.num")] # drop if unneeded

print(xtable(table,
                    align="L{0cm}C{.04\\textwidth}L{.4\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}L{.4\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
                    digits = 1
                    ), 
             size="\\small",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0)),
                              command = c(
                              # header
                                    paste(" & 
                                              \\multicolumn{4}{c}{\\textbf{BCS}} &
                                              \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9} 
                                            Item &
                                              Text & Cert. Appl. & Smt. Appl. & Appl. &
                                              Text & Cert. True & Smt. True & True \\\\ \\midrule
                                          "
                                            , sep="")
                                    )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabular", width = ".8\\textwidth",
             floating = FALSE
)

@
} %scalebox
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% EXPLORATORY FACTOR ANALYSIS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

<<BCSEFA>>=
@

<<BCSEFARES, dependson='BCSEFA'>>=
@

<<MCSEFA>>=
@

<<MCSEFARES, dependson='MCSEFA'>>=
@

\section{Exploratory Analysis}

\begin{frame}
\frametitle{EFA}
\framesubtitle{Number of factors}

\bi
\item Exploratory factor analysis on the two separate cohorts
\item Two-factor structure seems ok according to various approaches
\ei

\begin{center}
<<EFATAB, dependson='MCSEFA', results='asis'>>=
# combine results
efatab <- cbind(efatab.bcs,efatab.mcs)
colnames(efatab) <- c("BCS", "MCS")
rownames(efatab) <- c("Optimal Coordinates", "Acceleration Factor", "Parallel Analysis", "Kaiser", "VSS Compl. 1", "VSS Compl. 2")

print(xtable(efatab,
                                  # item num        Factor          Num cats        BCS text        BCS num           BCS Certain     BCS Somewhat    BCS Applies             MCS text          MCS num         
                    align="L{.4\\textwidth}C{.2\\textwidth}C{.2\\textwidth}",
                    digits = 0
                    ), 
             size="\\small",
             include.rownames=T,
             include.colnames=T,
             booktabs = T,
             sanitize.text.function = function(x){x},
             tabular.environment = "tabular", width = ".8\\textwidth",
             floating = FALSE
)
@
\end{center}
\end{frame}

\begin{frame}
\frametitle{EFA}
\framesubtitle{Factor structure}
\bi
\item In both cohorts, delivers a clear-cut split between two factors
\item Factors can be coherently interpreted as internalising and internalising behaviour problems
\ei 
\vspace{12pt}
\centering
\begin{tabular}{C{.05\textwidth}L{.3\textwidth}|C{.05\textwidth}L{.3\textwidth}}
  & \textbf{Externalising}  &    & \textbf{Internalising} \\
1 & Restless                & 7  & Worried                \\
2 & Squirmy/fidgety         & 8  & Afraid of new things   \\
3 & Fights/bullies children & 9  & Solitary               \\
4 & Cannot settle           & 10 & Unhappy/tearful        \\
5 & Temper tantrums         & 11 & Head/stomachaches \\
6 & Disobedient             &    &                       
\end{tabular}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% MEASUREMENT INVARIANCE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Measurement Invariance}

\begin{frame}
\frametitle{Measurement Invariance}

\bi
\item Two constructs isolated in the new scale, separately by cohort
\item But are externalising and internalising problems measured in a comparable way \emph{across cohorts}?
  \bi
  \item Slightly different wording of question and options
  \item Different ordering in questionnaire
  \ei 
\vspace{12pt}\pause
\item \textbf{Measurement Invariance (MI)} analysis
  \bi
  \item Factor analysis -- but can also use IRT
  \item Pool the two cohorts together
  \item Start from minimally restrictive model, where parameters are free to vary across cohorts
  \item Place increasing restrictions on parameters across cohorts and test fit
  \ei 
\ei

\end{frame}

\begin{frame}
\frametitle{MI}\label{mimain}
\framesubtitle{Invariance steps}

\bi
\item Set up categorical Item Factor Analysis (IFA) model \hyperlink{miapp1}{\beamerbutton{Det.}}
\item Specify and compare model of different invariance levels

\bi
\item Approach in \citep{Wu2016} \hyperlink{miapp2}{\beamerbutton{Det.}}
\item \textbf{Configural} model
  \bi
  \item Minimally restrictive (all parameters free across cohorts)
  \ei
\item \textbf{Threshold and Loading invariance} model
  \bi
  \item Thresholds and loadings are restricted to be invariant across cohorts
  \item Means set to zero in both grups, variances set to 1 in reference cohort only
  \item Allows comparison of latent factor variances
  \ei
\item \textbf{Threshold, Loading, and Intercept invariance} model
  \bi
  \item Thresholds, loadings, intercepts restricted to be invariant across cohorts
  \item Mean and variances unrestricted in target cohort
  \item Allows comparison of latent factor means and variances
  \ei
\ei
\ei

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

<<FA_INVARIANCE, dependson='LOAD_DATA'>>=
# estimate models of different levels of invariance
@

<<FA_FIT, dependson='FA_INVARIANCE'>>=
# assemble fit indices into matrices
@

<<FA_FINAL, dependson='FA_INVARIANCE'>>=
# select final model set
@

<<FA_FINAL_PARS, dependson='FA_FINAL'>>=
# get parameters of final model
@

<<CLEANUP, dependson='FA_FINAL'>>=
@

<<FA_SCORES, dependson='FA_FINAL'>>=
# produce scores for different models
@

% # <<FA_SCORES_COMPARE, dependson='FA_SCORES'>>=
% # # assemble and compare scores across different models
% # @

<<FA_SCORES_SELECT, dependson='FA_FINAL'>>=
# select model to use for inequality analysis
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% MEASUREMENT INVARIANCE RESULTS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}[noframenumbering]
\frametitle{MI}
\framesubtitle{Fit comparison}
\begin{table}[ht]
\centering
\small
<<FA_FIT_RES, dependson="FA_FIT", results='asis'>>=
require(xtable)

table <- rbind(afitab[[3]],afitab[[4]],afitab[[7]])

print(xtable(table,
                    align="L{0cm}L{.20\\textwidth}C{.03\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}",
                    digits = 4
                    ),
             size="\\tiny",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0,3,6)),
                              command = c(
                              # header
            "Model & Num. params & $\\chi^2$ & RMSEA & MFI & CFI & G-hat & \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ \\midrule
                       \\textbf{Males only} \\\\ ",
                      "\\newline \\textbf{Females only} \\\\ \\midrule ",
                      "\\newline \\textbf{Males and females} \\\\ \\midrule "
                                    )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabular", width = "\\textwidth",
             floating = FALSE
)

@
\end{table}
\end{frame}

\begin{frame}
\frametitle{MI}\label{mires}
\framesubtitle{Results}

\bi
\item Models with restricted intercepts do not pass MI test -- although the absolute fit is not terrible
  \bi
  \item $\Rightarrow$ cannot compare latent means according to \cite{Wu2016}
  \item but more traditional approaches with item anchoring would disagree \cite{Millsap2004}
  \ei
\item There are age differences between BCS and MCS due to survey design \hyperlink{agehist}{\beamerbutton{Graph}}
\item But restricting the age range to maximise overlap does not give significantly different results \hyperlink{fit4}{\beamerbutton{Graph}}
\item {\color{red} Can still use the model with non-invariant intercepts to score the latent factors and investigate inequality}
\ei

\end{frame}

\begin{frame}
\frametitle{Factor scores vs. Sum scores}
\begin{figure}
<<FACRAW, dev='tikz', fig.height=7, fig.width=7, out.width='.65\\linewidth'>>=
@
\end{figure}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% INEQUALITY
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Inequality in behaviour problems}

\begin{frame}
\frametitle{Inequality}

\bi
\item Has inequality in behavioural problems increased between 1975 and 2005?
\item Investigate income gradient in externalising and internalising scores
  \bi
  \item Merge harmonised income data from CLOSER Work Package 2(a)
  \item Weekly family income when child is 10
  \item Would be more interesting with harmonised social class...
  \ei
\ei

\end{frame}

\begin{frame}
\frametitle{Inequality}
\framesubtitle{Visualising the gradient}
\begin{figure}
<<FACINEQ, dev='tikz', fig.height=6.7, fig.width=10, out.width='\\linewidth'>>=
@
\end{figure}
\end{frame}

<<LOESS_REAL, dependson="FA_SCORES_SELECT">>=
@

\begin{frame}
\frametitle{Inequality}
\framesubtitle{Visualising the gradient -- inflation- and growth-adjusted income}
\begin{figure}
<<LOESS_REAL_EXT, dev='tikz', fig.height=8, fig.width=12, out.width='.95\\linewidth'>>=
ploess.real.ext
@
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Inequality}
\framesubtitle{Visualising the gradient -- inflation- and growth-adjusted income}
\begin{figure}
<<LOESS_REAL_INT, dev='tikz', fig.height=8, fig.width=12, out.width='.95\\linewidth'>>=
ploess.real.int
@
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Inequality}
\framesubtitle{Quantifying the gradient -- inflation- and growth-adjusted income}
\centering
\resizebox{!}{4cm}{%
<<FA_REGS_REAL, dependson='FA_SCORES', results='asis'>>=
# regress scores on income
library(dplyr)
library(stargazer)

scores2plot$incq_real <- factor(ntile(scores2plot$faminc_real, 5))

# get min, max
minmax.real <-aggregate(scores2plot[,c("faminc_real")],list(scores2plot$incq_real), 
            FUN = function(x) c(min = min(x, na.rm=T), max = max(x, na.rm=T)))
minmax.real <- round(minmax.real[,-1],2)*1000

r.ext.f.q <- lm(EXT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="F"))
r.int.f.q <- lm(INT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="F"))

r.ext.m.q <- lm(EXT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="M"))
r.int.m.q <- lm(INT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="M"))

stargazer(r.ext.m.q, r.int.m.q, r.ext.f.q, r.int.f.q,
          column.labels=c("Males","Females"), column.separate=c(2,2),
          dep.var.caption = "", multicolumn = FALSE,
          omit=c("Constant"),
          covariate.labels=c(
                paste0("Income Q2 [",toString(minmax.real[2,1]),"-",toString(minmax.real[2,2])," £/w]"),
                paste0("Income Q3 [",toString(minmax.real[3,1]),"-",toString(minmax.real[3,2])," £/w]"),
                paste0("Income Q4 [",toString(minmax.real[4,1]),"-",toString(minmax.real[4,2])," £/w]"),
                paste0("Income Q5 [",toString(minmax.real[5,1]),"-",toString(minmax.real[5,2])," £/w]"),
                             "MCS dummy",
                             "Income Q2 x MCS",
                             "Income Q3 x MCS",
                             "Income Q4 x MCS",
                             "Income Q5 x MCS"
                             ),
          no.space=TRUE, 
          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
          float=FALSE    # needs begin table outside it
)
@
}
\end{frame}

\begin{frame}
\frametitle{Inequality}

\bi
  \item The difference in scores between top and bottom quintiles has increased
  \item Scores at top quintiles have drifted upwards
\ei

\end{frame}

\begin{frame}
\frametitle{Returns}
\framesubtitle{Education returns to behaviour skills in BCS}
\centering
\resizebox{!}{4cm}{%
<<FA_RETURNS, dependson='FA_SCORES', results='asis'>>=
# regress scores on income
library(dplyr)
library(stargazer)

# get BCS data and make dummy for higher education
scoresBCS <- scores2plot[scores2plot$cohort=="BCS",]
scoresBCS$highered <- as.numeric(scoresBCS$hinvq00>3)
scoresBCS[is.na(scoresBCS$hinvq00), "highered"] <- NA

# females
r.nvq.f <- list()
r.nvq.f[[1]] <- lm(highered ~ EXT + INT, data=subset(scoresBCS, sex=="F"))
r.nvq.f[[2]] <- lm(highered ~ EXT + INT + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="F"))
r.nvq.f[[3]] <- lm(highered ~ EXT + INT + factor(incq), data=subset(scoresBCS, sex=="F"))
r.nvq.f[[4]] <- lm(highered ~ EXT + INT + ysch_moth + ysch_fath, data=subset(scoresBCS, sex=="F"))
r.nvq.f[[5]] <- lm(highered ~ EXT + INT + factor(incq) + ysch_moth + ysch_fath + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="F"))

# females
r.nvq.m <- list()
r.nvq.m[[1]] <- lm(highered ~ EXT + INT, data=subset(scoresBCS, sex=="M"))
r.nvq.m[[2]] <- lm(highered ~ EXT + INT + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="M"))
r.nvq.m[[3]] <- lm(highered ~ EXT + INT + factor(incq), data=subset(scoresBCS, sex=="M"))
r.nvq.m[[4]] <- lm(highered ~ EXT + INT + ysch_moth + ysch_fath, data=subset(scoresBCS, sex=="M"))
r.nvq.m[[5]] <- lm(highered ~ EXT + INT + factor(incq) + ysch_moth + ysch_fath + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="M"))

stargazer(r.nvq.m[[1]], r.nvq.m[[2]], r.nvq.m[[3]], r.nvq.m[[4]], r.nvq.m[[5]],
          r.nvq.f[[1]], r.nvq.f[[2]], r.nvq.f[[3]], r.nvq.f[[4]], r.nvq.f[[5]],
          dep.var.caption = "Completed Higher Ed. at 30",
          column.labels=c("Males","Females"), column.separate=c(5,5),
          omit=c("Constant"),
          covariate.labels=c("Externalising skill at 5",
                             "Internalising skill at 5",
                             "EPVT Z at 5",
                             "Human figure drawing Z at 5",
                             "Copy design Z at 5",
                             "Family inc. at 10 Q2",
                             "Family inc. at 10 Q3",
                             "Family inc. at 10 Q4",
                             "Family inc. at 10 Q5",
                             "Mother's years of schooling",
                             "Father's years of schooling"
                             ),
          no.space=TRUE, 
          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
          float=FALSE    # needs begin table outside it
)
@
}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% MEASUREMENT INVARIANCE APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}[noframenumbering]
\centering
{\Large Appendix}
\end{frame}

\begin{frame}[noframenumbering]
\frametitle{MI}\label{miapp1}
\framesubtitle{IFA model}
Categorical Item Factor Analysis (IFA) model \hyperlink{mimain}{\beamerreturnbutton{Back}}

\bi
\item Threshold model \citep{Muthen1984}
\item Latent propensity to endorse each item $i$ for child $j$ in cohort $c$
\begin{align}\label{eq:propens}
\begin{split}
X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{EXT}_{jc} + u_{ijc} \qquad \text{for} \; i=1,\dots, 6 \\
X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{INT}_{jc} + u_{ijc} \qquad \text{for} \; i=7,\dots, 11
\end{split}
\end{align}

\item Mean/covariance structure
$$\bm{\mu}_c=\bm{\nu}_c + \bm{\Lambda}_c\bm{\kappa}_c \qquad \text{and} \qquad \Sigma_c = \bm{\Lambda}_c \bm{\Phi}_c \bm{\Lambda}_c^\prime + \bm{\Psi}_c.$$
with $$(\theta^{EXT}_{jc}, \theta^{INT}_{jc}) = \bm{\Theta}_{jc} \sim N(\bm{\kappa}_c, \bm{\Phi}_c) \qquad \text{and} \qquad \bm{u}_{jc} \sim N(\bm{0}, \bm{\Psi}_c)$$

\item Discreteness
$$X_{ijc} = s \qquad \text{if} \; \tau_{s,ic} \leq X^*_{ijc} < \tau_{s+1,ic} \qquad \text{for} \; s=0,1,2$$
with $\tau_{0,ic}=-\infty$ and $\tau_{3,ic}=+\infty$.

\ei
\end{frame}

\begin{frame}[noframenumbering]
\frametitle{MI}
\framesubtitle{Key Parameters for Invariance}

\bi
\item Loadings $\lambda_{ci}$
\item Thresholds $\tau_{1,ci}$, $\tau_{2,ci}$
\item Intercepts $\nu_{ci}$
\item Unique variances $\Var(u_{ci})$
\item Latent mean vector $\bm{\kappa}_c$
\item Latent covariance matrix $\bm{\Phi}_c$
\ei

\end{frame}

\begin{frame}[noframenumbering]
\frametitle{MI}\label{miapp2}
\framesubtitle{Invariance steps \hyperlink{mimain}{\beamerreturnbutton{Back}}}
\Fontvi
\bi
\item \textbf{Configural} model
$$\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}, \quad \bm{\nu}=\bm{0}, \quad \text{and} \; \text{diag}(\bm{\Psi})=\bm{I}.$$

\item \textbf{Threshold invariance} model
  \bi
  \item $\tau_{1,Ai} = \tau_{1,Bi}$ for $i=1,\dots,11$, $\tau_{2,Ai} = \tau_{2,Bi}$ for $i=\{1,2,3,4,7,8,9,10\}$
  \item For all groups: $\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}$
  \item For the reference group: $\bm{\nu}_A=\bm{0}, \quad \; \text{diag}(\bm{\Sigma}_A)=\bm{I}$
  \ei
\item \textbf{Threshold and Loading invariance} model
  \bi
  \item $\tau_{1,Ai} = \tau_{1,Bi}$ for $i=1,\dots,11$, $\tau_{2,Ai} = \tau_{2,Bi}$ for $i=\{1,2,3,4,7,8,9,10\}$
  \item $\lambda_{Ai} = \lambda_{Bi}$ for $i=1,\dots,11$
  \item For all groups: $\bm{\kappa}=\bm{0}$
  \item For the reference group: $\bm{\nu}_A=\bm{0}, \quad \; \text{diag}(\bm{\Sigma}_A)=\bm{I}, \quad \text{diag}(\bm{\Phi}_A) = \bm{I}$
  \ei
\item \textbf{Threshold, Loading, and Intercept invariance} model
  \bi
  \item $\tau_{1,Ai} = \tau_{1,Bi}$ for $i=1,\dots,11$, $\tau_{2,Ai} = \tau_{2,Bi}$ for $i=\{1,2,3,4,7,8,9,10\}$
  \item $\lambda_{Ai} = \lambda_{Bi}$ for $i=1,\dots,11$
  \item For all groups: $\bm{\nu}=\bm{0}$
  \item For the reference group: $\bm{\kappa}_A=\bm{0}, \quad \; \text{diag}(\bm{\Sigma}_A)=\bm{I}, \quad \text{diag}(\bm{\Phi}_A) = \bm{I}$
  \ei
\ei

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% ROBUSTNESS APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}[noframenumbering]
\frametitle{Age distribution \hyperlink{mires}{\beamerreturnbutton{Back}}}\label{agehist}
\begin{figure}
\begin{center}
<<AGEHIST, dev='tikz', fig.height=6, fig.width=6, out.width='.6\\linewidth', cache=TRUE>>=
@
\end{center}
\caption{Distribution of age in the BCS and MCS sample}
\end{figure}
\end{frame}

\begin{frame}[noframenumbering]
\frametitle{MI}
\framesubtitle{Fit comparison -- Robustness}
List of estimated models:
\bnm
\item no gender split, no age adjustment (2 groups, BCS MCS)
\item no gender split, age adjustment (2 groups, BCS MCS)
\item males only, no age adjustment (2 groups, BCS MCS)
\item females only, no age adjustment (2 groups, BCS MCS)
\item males only, age adjustment (2 groups, BCS MCS)
\item females only, age adjustment (2 groups, BCS MCS)
\item separate gender groups, no age adjustment (4 groups, BCS.M BCS.F MCS.M MCS.F)
\item separate gender groups, age adjustment (4 groups, BCS.M BCS.F MCS.M MCS.F)
\item separate gender groups, overlapping ages, no age adjustment (4 groups, BCS.M BCS.F MCS.M MCS.F)
\item separate gender groups, age adjustment constrained to be the same across ages (4 groups, BCS.M BCS.F MCS.M MCS.F)
\item MIMIC model with age as regressor (4 groups, BCS.M BCS.F MCS.M MCS.F)
\enm

\end{frame}

\begin{frame}[noframenumbering]
\frametitle{MI}
\framesubtitle{Fit comparison -- 2 group models, age unadjusted}

\begin{table}[ht]
\centering
\small
<<FA_FIT_RES2a, dependson="FA_FIT", results='asis'>>=
require(xtable)

table <- rbind(afitab[[1]],afitab[[3]],afitab[[4]])

print(xtable(table,
                    align="L{0cm}L{.20\\textwidth}C{.03\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}",
                    digits = 4
                    ),
             size="\\tiny",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0,3,6)),
                              command = c(
                              # header
            "Model & Num. params & $\\chi^2$ & RMSEA & MFI & CFI & G-hat & \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ \\midrule
                       \\textbf{(1) No gender split, not age adjusted} \\\\",
                      "\\newline \\textbf{(3) Males only, not age adjusted} \\\\ \\midrule ",
                      "\\newline \\textbf{(4) Females only, not age adjusted} \\\\ \\midrule "
                                    )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabular", width = "\\textwidth",
             floating = FALSE
)

@
\end{table}

\end{frame}

\begin{frame}[noframenumbering]
\frametitle{MI}
\framesubtitle{Fit comparison -- 2 group models, age adjusted}
\begin{table}[ht]
\centering
\small
<<FA_FIT_RES2b, dependson="FA_FIT", results='asis'>>=
require(xtable)

table <- rbind(afitab[[2]],afitab[[5]],afitab[[6]])

print(xtable(table,
                    align="L{0cm}L{.20\\textwidth}C{.03\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}",
                    digits = 4
                    ),
             size="\\tiny",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0,3,6)),
                              command = c(
                              # header
            "Model & Num. params & $\\chi^2$ & RMSEA & MFI & CFI & G-hat & \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ \\midrule
                       \\textbf{(2) No gender split, age adjusted} \\\\ ",
                      "\\newline \\textbf{(5) Males only, age adjusted} \\\\ \\midrule ",
                      "\\newline \\textbf{(6) Females only, age adjusted} \\\\ \\midrule "
                                    )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabular", width = "\\textwidth",
             floating = FALSE
)

@
\end{table}
\end{frame}

\begin{frame}[noframenumbering]
\frametitle{MI}\label{fit4}
\framesubtitle{Fit comparison -- 4 group models \hyperlink{mires}{\beamerreturnbutton{Back}}}

\begin{table}[ht]
\centering
\small
<<FA_FIT_RES4, dependson="FA_FIT", results='asis'>>=
require(xtable)

table <- rbind(afitab[[7]],afitab[[8]],afitab[[9]])

print(xtable(table,
                    align="L{0cm}L{.20\\textwidth}C{.03\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}",
                    digits = 4
                    ),
             size="\\tiny",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0,3,6)),
                              command = c(
                              # header
            "Model & Num. params & $\\chi^2$ & RMSEA & MFI & CFI & G-hat & \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ \\midrule
                      \\newline \\textbf{(7) not age adjusted} \\\\",
                      "\\newline \\textbf{(8) age adjusted} \\\\\\midrule ",
                      "\\newline \\textbf{(9) 59-62m range} \\\\ \\midrule "
                                    )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabular", width = "\\textwidth",
             floating = FALSE
)

@
\end{table}

\end{frame}

<<LOESS_INFL, dependson="FA_SCORES_SELECT">>=
@

\begin{frame}
\frametitle{Inequality}
\framesubtitle{Visualising the gradient -- inflation-adjusted income}
\begin{figure}
<<LOESS_INFL_EXT, dev='tikz', fig.height=8, fig.width=12, out.width='.95\\linewidth'>>=
ploess.infl.ext
@
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Inequality}
\framesubtitle{Visualising the gradient -- inflation-adjusted income}
\begin{figure}
<<LOESS_INFL_INT, dev='tikz', fig.height=8, fig.width=12, out.width='.95\\linewidth'>>=
ploess.infl.int
@
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Inequality}
\framesubtitle{Quantifying the gradient -- inflation-adjusted income}
\centering
\resizebox{!}{4cm}{%
<<FA_REGS_INFL, dependson='FA_SCORES', results='asis'>>=
# regress scores on income
library(dplyr)
library(stargazer)

scores2plot$incq_infl <- factor(ntile(scores2plot$faminc_infl, 5))

r.ext.f.c <- lm(EXT ~ faminc_infl*cohort, data=subset(scores2plot, sex=="F"))
r.ext.f.q <- lm(EXT ~ incq_infl*cohort, data=subset(scores2plot, sex=="F"))
r.int.f.c <- lm(INT ~ faminc_infl*cohort, data=subset(scores2plot, sex=="F"))
r.int.f.q <- lm(INT ~ incq_infl*cohort, data=subset(scores2plot, sex=="F"))

r.ext.m.c <- lm(EXT ~ faminc_infl*cohort, data=subset(scores2plot, sex=="M"))
r.ext.m.q <- lm(EXT ~ incq_infl*cohort, data=subset(scores2plot, sex=="M"))
r.int.m.c <- lm(INT ~ faminc_infl*cohort, data=subset(scores2plot, sex=="M"))
r.int.m.q <- lm(INT ~ incq_infl*cohort, data=subset(scores2plot, sex=="M"))

stargazer(r.ext.m.q, r.int.m.q, r.ext.f.q, r.int.f.q,
          column.labels=c("Males","Females"), column.separate=c(2,2),
          dep.var.caption = "", multicolumn = FALSE,
          omit=c("Constant"),
          covariate.labels=c(
                             "Income Q2",
                             "Income Q3",
                             "Income Q4",
                             "Income Q5",
                             "MCS dummy",
                             "Income Q2 x MCS",
                             "Income Q3 x MCS",
                             "Income Q4 x MCS",
                             "Income Q5 x MCS"
                             ),
          no.space=TRUE, 
          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
          float=FALSE    # needs begin table outside it
)
@
}
\end{frame}

\begin{frame}[noframenumbering]
\bibliographystyle{chicago}
\def\bibfont{\tiny}
\bibliography{/Users/giacomomason/Documents/BibTex/Papers-cohorts_nc}
\end{frame}

\end{document}

### 2018-02-27 codes/R/cohorts_nc.Rnw
minor changes to text tables

@@
 for (r in 1:nrow(regcoeffs)) {
-  regs.ext[[r]] <- lm(EXT ~ as.factor(incq) + age, data=subset(items.scored[[8]], cohort==regcoeffs[r,1] & sex==regcoeffs[r,2]))
+  regs.ext[[r]] <- lm(EXT ~ as.factor(incq) + age, data=subset(scores2plot, cohort==regcoeffs[r,1] & sex==regcoeffs[r,2]))
   regcoeffs[r,3] <- round(regs.ext[[r]]$coefficients["as.factor(incq)5"],2)
-  regs.int[[r]] <- lm(INT ~ as.factor(incq) + age, data=subset(items.scored[[8]], cohort==regcoeffs[r,1] & sex==regcoeffs[r,2]))
+  regs.int[[r]] <- lm(INT ~ as.factor(incq) + age, data=subset(scores2plot, cohort==regcoeffs[r,1] & sex==regcoeffs[r,2]))
   regcoeffs[r,4] <- round(regs.int[[r]]$coefficients["as.factor(incq)5"],2)

### 2018-02-27 codes/R/5_graphs.R
beautify graphs, separate unused ones

@@ -47,50 +47,29 @@ p
 
 ## ---- FACRAW
 # see how scored factors compare with raw scores
-box.ext.bcs <- ggplot(data=subset(scores2plot, cohort=="BCS"), aes(x=as.factor(EXT.RAW), y=EXT)) + geom_boxplot() + ggtitle("BCS EXT")
-box.int.bcs <- ggplot(data=subset(scores2plot, cohort=="BCS"), aes(x=as.factor(INT.RAW), y=INT)) + geom_boxplot() + ggtitle("BCS INT")
-box.ext.mcs <- ggplot(data=subset(scores2plot, cohort=="MCS"), aes(x=as.factor(EXT.RAW), y=EXT)) + geom_boxplot() + ggtitle("MCS EXT")
-box.int.mcs <- ggplot(data=subset(scores2plot, cohort=="MCS"), aes(x=as.factor(INT.RAW), y=INT)) + geom_boxplot() + ggtitle("MCS INT")
+box.ext.bcs <- ggplot(data=subset(scores2plot, cohort=="BCS"), aes(x=as.factor(EXT.RAW), y=EXT)) + geom_boxplot() + ggtitle("BCS EXT") + xlab("Sum score") + ylab("Factor score")
+box.int.bcs <- ggplot(data=subset(scores2plot, cohort=="BCS"), aes(x=as.factor(INT.RAW), y=INT)) + geom_boxplot() + ggtitle("BCS INT") + xlab("Sum score") + ylab("Factor score")
+box.ext.mcs <- ggplot(data=subset(scores2plot, cohort=="MCS"), aes(x=as.factor(EXT.RAW), y=EXT)) + geom_boxplot() + ggtitle("MCS EXT") + xlab("Sum score") + ylab("Factor score")
+box.int.mcs <- ggplot(data=subset(scores2plot, cohort=="MCS"), aes(x=as.factor(INT.RAW), y=INT)) + geom_boxplot() + ggtitle("MCS INT") + xlab("Sum score") + ylab("Factor score")
 plot_grid(box.ext.bcs, box.int.bcs, box.ext.mcs, box.int.mcs, ncol=2, align="h")
 
-## ---- FACSCAT
[…]

### 2018-02-27 codes/R/5_graphs_alt.R

@@ -1,0 +1,107 @@
+## ---- FACINEQ2
+ineq.ext.bcs <- ggplot(data=subset(facandraw, cohort=="BCS" & !is.na(facandraw$incq)), aes(x=as.factor(incq), y=EXT)) + ggtitle("BCS EXT") +
+  geom_boxplot() + scale_x_discrete("Family Income Quintile at 10") + scale_y_continuous(limits = c(-3, 2))
+ineq.int.bcs <- ggplot(data=subset(facandraw, cohort=="BCS" & !is.na(facandraw$incq)), aes(x=as.factor(incq), y=INT)) + ggtitle("BCS INT") +
+  geom_boxplot() + scale_x_discrete("Family Income Quintile at 10") + scale_y_continuous(limits = c(-3, 2))
+ineq.ext.mcs <- ggplot(data=subset(facandraw, cohort=="MCS" & !is.na(facandraw$incq)), aes(x=as.factor(incq), y=EXT)) + ggtitle("MCS EXT") +
+  geom_boxplot() + scale_x_discrete("Family Income Quintile at 10") + scale_y_continuous(limits = c(-3, 2))
+ineq.int.mcs <- ggplot(data=subset(facandraw, cohort=="MCS" & !is.na(facandraw$incq)), aes(x=as.factor(incq), y=INT)) + ggtitle("MCS INT") +
+  geom_boxplot() + scale_x_discrete("Family Income Quintile at 10") + scale_y_continuous(limits = c(-3, 2))
+
+plot_grid(ineq.ext.bcs, ineq.ext.mcs, ineq.int.bcs, ineq.int.mcs, ncol=2, align="h")
+
+## ---- FACINEQ3
+# boxplots of scores by gender and cohort
+
+#common options
+addopts <- function(x) {
+  x <- x + geom_boxplot() + scale_x_discrete("Family Income Quintile at 10") + scale_y_continuous(name = "Factor score", limits = c(-3.5, 2.2)) +
+    stat_summary(fun.y=mean, geom="point", size=4, fill="white", color="black", aes(shape=cohort),position=position_dodge(.8))+
+    labs(list(fill="", shape="")) + theme(legend.position="none") +
+    guides(fill = guide_legend(override.aes = list(size=2)))
+  return(x)
+} 
[…]

### 2018-02-27 codes/R/4_scores.R
minor

@@ -5,7 +5,7 @@
 scores <- list()
 scoresdf <- list()
 items.scored <- list()
-for (i in 7:11) { # only specifications with separate gender groups
+for (i in 7:7) { # only specifications with separate gender groups
 
   scores[[i]] <- predict(finalmod[[i]], newdata = items[[i]])
   scoresdf[[i]] <- do.call(rbind, scores[[i]])
@@ -16,9 +16,4 @@ for (i in 7:11) { # only specifications with separate gender groups
   items.scored[[i]]$INTr <- residuals(lm(INT ~ age, data=items.scored[[i]], na.action = na.exclude))
   
 }
-
-############################################################################################
-## ---- FA_SCORES_SELECT
-# select scores to plot
-scores2plot <- items.scored[[8]]
 

### 2018-02-27 codes/R/4_scores_compare.R

@@ -5,7 +5,7 @@
 
 # 4 group with no intercept invariance
 scores.tl <- predict(fa.tl[[7]], newdata = items[[7]]) # new scores prediction
-items.scored.tl <- cbind(items[[7]][,c("id", "cohortsex", "faminc_real", "incq")], do.call(rbind, scores.tl))
+items.scored.tl <- cbind(items[[7]], do.call(rbind, scores.tl))
 names(items.scored.tl)[names(items.scored.tl) %in% c("EXT","INT")] <- c("EXTtl", "INTtl")
 
 # 4 group with no intercept invariance, overlapping ages (model 9)
@@ -47,6 +47,12 @@ means.int <- cbind(group = factor(c(1,3,2,4)), means.int)
 levels(means.int$group) <- c("BCS Males", "MCS Males", "BCS Females", "MCS Females")
 means.int <- means.int[order(means.int$group),]
 means.int <- as.matrix(means.int[, !names(means.int) %in% "Group.1"],3)
+
+############################################################################################
+## ---- FA_SCORES_SELECT
+# select scores to plot
+scores2plot <- items.scored.tl
+names(scores2plot)[names(scores2plot) %in% c("EXTtl","INTtl")] <- c("EXT", "INT")
 
 
 # ggplot(compare.scores, aes(x=EXTtl, y=EXTtli, color=cohortsex)) + geom_point(size=2, alpha=.2) + coord_fixed()

### 2018-02-28 codes/R/2_efa.R
EFA by gender

@@ -11,19 +11,42 @@ nbcs <- dim(items.bcs)[1]
 bcscor <- hetcor(items.bcs, ML=TRUE, std.err = F)$correlations
 bcsfa <- fa(r=bcscor, nfactors=2, rotate = rotation, fm="wls")
 
+items.bcs.m <- items.c[items.c$cohort=="BCS" & items.c$sex=="M",c(grep("X[0-9]", names(items.c), value=T))]
+nbcs.m <- dim(items.bcs.m)[1]
+bcscor.m <- hetcor(items.bcs.m, ML=TRUE, std.err = F)$correlations
+bcsfa.m <- fa(r=bcscor.m, nfactors=2, rotate = rotation, fm="wls")
+
+items.bcs.f <- items.c[items.c$cohort=="BCS" & items.c$sex=="F",c(grep("X[0-9]", names(items.c), value=T))]
+nbcs.f <- dim(items.bcs.f)[1]
+bcscor.f <- hetcor(items.bcs.f, ML=TRUE, std.err = F)$correlations
+bcsfa.f <- fa(r=bcscor.f, nfactors=2, rotate = rotation, fm="wls")
+
+
 
 ## ---- BCSEFARES
 # Scree
 ns.bcs <- nScree(bcscor)$Components
+ns.bcs.m <- nScree(bcscor.m)$Components
+ns.bcs.f <- nScree(bcscor.f)$Components
+
 # VSS for number of factors
 vss.bcs <- VSS(bcscor, plot=F, rotate=rotation, n.obs = nbcs)
+vss.bcs.m <- VSS(bcscor.m, plot=F, rotate=rotation, n.obs = nbcs.m)
+vss.bcs.f <- VSS(bcscor.f, plot=F, rotate=rotation, n.obs = nbcs.f)
 #combine
-efatab.bcs <- rbind(t(ns.bcs), which.max(vss.bcs$cfit.1), which.max(vss.bcs$cfit.2))
-rownames(efatab.bcs) <- c("Optimal Coordinates", "Acceleration Factor", "Parallel Analysis", "Kaiser", "VSS Compl. 1", "VSS Compl. 2")
+efatab.bcs <- cbind(
+  rbind(t(ns.bcs), which.max(vss.bcs$cfit.1), which.max(vss.bcs$cfit.2), which.min(vss.bcs$map)),
[…]

### 2018-02-28 codes/R/1_load.R
separate scoring from assembling data with scores

@@ -33,6 +33,10 @@ printfit <- function(m) {
 
 ## ---- LOAD_DATA
 
+# auxiliary variables to retain
+auxvars <- c("faminc_real", "faminc_infl", "incq", "ysch_moth", "ysch_fath", 
+             "ageint5", "sex", "bwt", "smkpr", "pscl")
+
 # BCS -------------------------------------------
 bcs5data <- read.dta(paste(dir_data, "bcs5yeng.dta", sep=""), convert.factors = F) # all BCS data
 bcs5rutb <- bcs5data[,grep("bcs5_rutb", names(bcs5data), value=TRUE)]                 # BINARY Rutter items only
@@ -72,7 +76,7 @@ for (i in 1:ncol(bcs5rutc)) bcs5rutcf[,i] <- as.ordered(bcs5rutc[,i])
 
 
 # auxiliary variables
-bcs5aux <- cbind(bcs5data[,c("bcsid", "faminc_real", "faminc_infl", "incq", "ysch_moth", "ysch_fath", "ageint5", "sex", "hinvq00")],
+bcs5aux <- cbind(bcs5data[,c("bcsid", auxvars, "hinvq00")],
                 cog1 = bcs5data$epvt_z,
                 cog2 = bcs5data$hfd_z,
                 cog3 = bcs5data$copy_z
@@ -115,7 +119,7 @@ mcs5sdqcf <- mcs5sdqc
 for (i in 1:ncol(mcs5sdqc)) mcs5sdqcf[,i] <- as.ordered(mcs5sdqc[,i])                    
 
 # add SES and cognitive data
-mcs5aux <- cbind(mcs5data[,c("mcsid", "faminc_real", "faminc_infl", "incq", "ysch_moth", "ysch_fath", "ageint5", "sex")],
+mcs5aux <- cbind(mcs5data[,c("mcsid", auxvars)],
                 cog1 = mcs5data$nvoc_bastz,
                 cog2 = mcs5data$psim_bastz,
                 cog3 = mcs5data$patc_bastz
@@ -145,10 +149,8 @@ X.all <- data.frame(rbind(Xtemp.bcs, Xtemp.mcs))
 
 # assemble final data
 items.c <- X.all[,c(grep("X", names(X.all), value=T), 
-                "id", "cohort", "ageint5", "sex", 
-                "cog1", "cog2", "cog3",
-                "incq", "faminc_real", "faminc_infl", "ysch_moth", "ysch_fath",
[…]

### 2018-02-28 codes/R/4_scores.R

@@ -3,17 +3,19 @@
 ############################################################################################
 ## ---- FA_SCORES
 scores <- list()
-scoresdf <- list()
-items.scored <- list()
-for (i in 7:7) { # only specifications with separate gender groups
-
+for (i in 7:11) { # only specifications with separate gender groups
   scores[[i]] <- predict(finalmod[[i]], newdata = items[[i]])
+}
+
+############################################################################################
+## ---- FA_SCORES_MERGE
+items.scored <- list()
+scoresdf <- list()
+for (i in 7:11) { # only specifications with separate gender groups
   scoresdf[[i]] <- do.call(rbind, scores[[i]])
   items.scored[[i]] <- cbind(items[[i]], scoresdf[[i]])
   
   # residualise scores in age
   items.scored[[i]]$EXTr <- residuals(lm(EXT ~ age, data=items.scored[[i]], na.action = na.exclude))
   items.scored[[i]]$INTr <- residuals(lm(INT ~ age, data=items.scored[[i]], na.action = na.exclude))
-  
 }
-

### 2018-02-28 codes/R/slides/cohorts_nc_slides.Rnw
slides discussed 28/2/18

@@
 \bi
-\item Data from 5 year follow-up, maternal questionnaire
+\item Data from 5 year follow-up (England only), maternal questionnaire
 \item Rutter A scale \citep{Rutter1970}
@@
 \bi
-\item Data from 5 year follow-up (MCS3), maternal questionnaire
-\item Complex survey design: need to rebalance data to make representative
+\item Data from 5 year follow-up (England only), maternal questionnaire
+\item Complex survey design: rebalanced data to make representative
 \item Strengths and Difficulties Questionnaire (SDQ, \citealp{Goodman1994})
@@
 efatab <- cbind(efatab.bcs,efatab.mcs)
-colnames(efatab) <- c("BCS", "MCS")
-rownames(efatab) <- c("Optimal Coordinates", "Acceleration Factor", "Parallel Analysis", "Kaiser", "VSS Compl. 1", "VSS Compl. 2")
+colnames(efatab) <- c("BCS All", "BCS M", "BCS F", "MCS All", "MCS M", "MCS F")
 
 print(xtable(efatab,
-                                  # item num        Factor          Num cats        BCS text        BCS num           BCS Certain     BCS Somewhat    BCS Applies             MCS text          MCS num
-                    align="L{.4\\textwidth}C{.2\\textwidth}C{.2\\textwidth}",
+                    align="L{.25\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}",
                     digits = 0
@@
 
-<<FA_INVARIANCE, dependson='LOAD_DATA'>>=
+<<FA_INVARIANCE>>=
 # estimate models of different levels of invariance
@@
 
-% # <<FA_SCORES_COMPARE, dependson='FA_SCORES'>>=
-% # # assemble and compare scores across different models
-% # @
-
-<<FA_SCORES_SELECT, dependson='FA_FINAL'>>=
+<<FA_SCORES_MERGE, dependson='LOAD_DATA'>>=
+# merge scores with other data
+@
+
+<<FA_SCORES_COMPARE, dependson='FA_SCORES'>>=
+# assemble and compare scores across different models
+@
+
+<<FA_SCORES_SELECT, dependson='FA_SCORES_COMPARE'>>=
 # select model to use for inequality analysis
@@
 
-\begin{frame}
-\frametitle{Factor scores vs. Sum scores}
-\begin{figure}
-<<FACRAW, dev='tikz', fig.height=7, fig.width=7, out.width='.65\\linewidth'>>=
-@
-\end{figure}
-\end{frame}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-<<LOESS_REAL, dependson="FA_SCORES_SELECT">>=
+<<LOESS_REAL, dependson='FA_SCORES_SELECT'>>=
 @
@@
 \begin{figure}
-<<LOESS_REAL_EXT, dev='tikz', fig.height=8, fig.width=12, out.width='.95\\linewidth'>>=
+<<LOESS_REAL_EXT, dependson='LOESS_REAL', dev='tikz', fig.height=8, fig.width=12, out.width='.95\\linewidth'>>=
 ploess.real.ext
@@
 \begin{figure}
-<<LOESS_REAL_INT, dev='tikz', fig.height=8, fig.width=12, out.width='.95\\linewidth'>>=
+<<LOESS_REAL_INT, dependson='LOESS_REAL', dev='tikz', fig.height=8, fig.width=12, out.width='.95\\linewidth'>>=
 ploess.real.int
@@
 \resizebox{!}{4cm}{%
-<<FA_REGS_REAL, dependson='FA_SCORES', results='asis'>>=
+<<FA_REGS_REAL, dependson='FA_SCORES_SELECT', results='asis'>>=
 # regress scores on income
@@
                 paste0("Income Q4 [",toString(minmax.real[4,1]),"-",toString(minmax.real[4,2])," £/w]"),
-                paste0("Income Q5 [",toString(minmax.real[5,1]),"-",toString(minmax.real[5,2])," £/w]"),
+                paste0("Income Q5 [",toString(minmax.real[5,1]),"+ £/w]"),
                              "MCS dummy",
@@
 \frametitle{Inequality}
+\framesubtitle{Quantifying the gradient -- males}
+\centering
+\resizebox{12cm}{4cm}{%
+<<FA_REGS_REAL_M, dependson='FA_SCORES', results='asis'>>=
+# regress scores on income
+library(dplyr)
+library(stargazer)
+
+r.ext.m.bcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="BCS"))
+r.ext2.m.bcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="BCS"))
+r.ext.m.mcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="MCS"))
+r.ext2.m.mcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="MCS"))
+r.int.m.bcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="BCS"))
+r.int2.m.bcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="BCS"))
+r.int.m.mcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="MCS"))
+r.int2.m.mcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="MCS"))
+
+stargazer(r.ext.m.bcs, r.ext2.m.bcs, r.int.m.bcs, r.int2.m.bcs, r.ext.m.mcs, r.ext2.m.mcs, r.int.m.mcs, r.int2.m.mcs,
+          column.labels=c("BCS","MCS"), column.separate=c(4,4),
+          dep.var.caption = "", multicolumn = FALSE,
+          omit=c("Constant"),
+          covariate.labels=c(
+                             "Income Q2",
+                             "Income Q3",
+                             "Income Q4",
+                             "Income Q5",
+                             "Birthweight",
+                             "Mat. Smoking in preg.",
+                             "Father years of schooling",
+                             "Mother years of schooling"
+                             ),
+          no.space=TRUE, 
+          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
+          float=FALSE    # needs begin table outside it
+)
+@
+}
+\end{frame}
+
+\begin{frame}
+\frametitle{Inequality}
+\framesubtitle{Quantifying the gradient -- females}
+\centering
+\resizebox{12cm}{4cm}{%
+<<FA_REGS_REAL_F, dependson='FA_SCORES', results='asis'>>=
+# regress scores on income
+library(dplyr)
+library(stargazer)
+
+r.ext.f.bcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="BCS"))
+r.ext2.f.bcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="BCS"))
+r.ext.f.mcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="MCS"))
+r.ext2.f.mcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="MCS"))
+r.int.f.bcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="BCS"))
+r.int2.f.bcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="BCS"))
+r.int.f.mcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="MCS"))
+r.int2.f.mcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="MCS"))
+
+stargazer(r.ext.f.bcs, r.ext2.f.bcs, r.int.f.bcs, r.int2.f.bcs, r.ext.f.mcs, r.ext2.f.mcs, r.int.f.mcs, r.int2.f.mcs,
+          column.labels=c("BCS","MCS"), column.separate=c(4,4),
+          dep.var.caption = "", multicolumn = FALSE,
+          omit=c("Constant"),
+          covariate.labels=c(
+                             "Income Q2",
+                             "Income Q3",
+                             "Income Q4",
+                             "Income Q5",
+                             "Birthweight",
+                             "Mat. Smoking in preg.",
+                             "Father years of schooling",
+                             "Mother years of schooling"
+                             ),
+          no.space=TRUE, 
+          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
+          float=FALSE    # needs begin table outside it
+)
+@
+}
+\end{frame}
+
+\begin{frame}
+\frametitle{Inequality}
 
@@
 \resizebox{!}{4cm}{%
-<<FA_RETURNS, dependson='FA_SCORES', results='asis'>>=
+<<FA_RETURNS, dependson='FA_SCORES_SELECT', results='asis'>>=
 # regress scores on income
@@
 
+\begin{frame}[noframenumbering]
+\frametitle{Factor scores vs. Sum scores}
+\begin{figure}
+<<FACRAW, dev='tikz', fig.height=7, fig.width=7, out.width='.65\\linewidth'>>=
+@
+\end{figure}
+\end{frame}
+
 <<LOESS_INFL, dependson="FA_SCORES_SELECT">>=
@@
 \begin{figure}
-<<LOESS_INFL_EXT, dev='tikz', fig.height=8, fig.width=12, out.width='.95\\linewidth'>>=
+<<LOESS_INFL_EXT, dependson='LOESS_INFL', dev='tikz', fig.height=8, fig.width=12, out.width='.95\\linewidth'>>=
 ploess.infl.ext
@@
 \begin{figure}
-<<LOESS_INFL_INT, dev='tikz', fig.height=8, fig.width=12, out.width='.95\\linewidth'>>=
+<<LOESS_INFL_INT, dependson='LOESS_INFL', dev='tikz', fig.height=8, fig.width=12, out.width='.95\\linewidth'>>=
 ploess.infl.int
@@
 \resizebox{!}{4cm}{%
-<<FA_REGS_INFL, dependson='FA_SCORES', results='asis'>>=
+<<FA_REGS_INFL, dependson='FA_SCORES_SELECT', results='asis'>>=
 # regress scores on income

### 2018-03-01 codes/R/5_graphs.R
axis label

@@ -147,7 +147,7 @@ l.inc <- lapply(l.inc,
                 function(x) x + theme(legend.position="none") + labs(list(fill="", colour = "")) +
                   geom_smooth(method = "loess") + 
                   coord_cartesian(xlim = inc.lims, ylim = c(-.6, .6)) + #axis limits withouth affecting sample
-                  xlab("Weekly Family Income at 10 (,000£ 2015 CPI+GDP)") +
+                  xlab("Weekly Family Income at 10 (,000£ 2015 CPI)") +
                   ylab("Factor Score")
 ) # apply options to all graphs 
 

### 2018-03-01 codes/R/slides/cohorts_nc_slides.Rnw
slides 1/3/18

@@
 % UCL beamer template
-\definecolor{steelblue}{RGB}{70,130,180}
+\definecolor{deepskyblue}{RGB}{0,133,178}
 \definecolor{orangeucl}{RGB}{255,170,40}
@@
 \setbeamertemplate{banner}[ucl][1]
-\setbeamercolor{banner stripe}{bg=steelblue!70,fg=black}
+\setbeamercolor{banner stripe}{bg=deepskyblue!70,fg=black}
 \setbeamercolor{block body}{bg=steelblue!30,fg=black}
@@
 
-\title[Noncognitive skills in cohorts]{\bf{Comparability of child noncognitive skills across UK birth cohorts}}
+\title[Noncognitive skills in cohorts]{\bf{Inequality in non-cognitive skills: \\ a cross-cohort comparison}}
 
@@
 
-\date[HCRG]{Human Capital Reading Group \\ \today}
+\date[NCDS]{NCDS 60 Years of Our Lives \\ \today}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{frame}
-\frametitle{Motivation}
-\framesubtitle{The role of noncognitive skills}
-\bi
-\item Traditional labour literature used to focus mainly on cognitive skills
-    \bi 
-    \item returns to ability in adulthood (labour market, health, \dots)
-    \item process of child development
-    \ei
-\item Attention broadened to include \emph{non}-cognitive skills \citep{Cunha2007a,Cunha2010}
-    \bi
-    \item Socio-emotional, behavioural problems, character, personality traits, locus of control, grit, \dots
-    \item Found to be very important in shaping adult outcomes \citep{Heckman2006,Carneiro2007,Heckman2012b}
-    \item Malleable until later childhood
-    \item Primary mechanism of transmission of effects from early life programmes to adult outcomes \citep{Heckman2013,Kautz2014a}
-    \ei
-\item Today, focus on behaviour problems and mental health
-\ei
-\end{frame}
-
-\begin{frame}
-\frametitle{Motivation}
-\framesubtitle{Behavioural problems}
-
-... something more on how behavioural problems are defined and fit within the literature on noncognitive skills
-
-\end{frame}
 
@@
 \bi
-\item Socioeconomic gradients in behaviour problems are documented, and are persistent across childhood/adolescence \citep{Campbell1995,Segal2008}
-\item Still not clear 
-    \bi
-    \item how these gradients change across time
-    \item whether the returns to behavioural skills evolve as well
-    \ei
-\item \emph{Measurement} is a key (often overlooked) issue
+\item Non-cognitive (socio-emotional, behavioural) skills are important in childhood
+\item Socioeconomic gradients in these skills are persistent 
+\pause 
+\item \textbf{Q1}: Can we derive \textbf{comparable measures} across populations?
     \bi
     \item Usually, behavioural constructs are measured via scales
-    \item Most research uses (normalised) aggregate scores from scales \citep{Blanden2007}
     \item But does the same scale measure the same constructs in the same way:
@@
     \ei
+\pause
+\item \textbf{Q2}: Can we use these measures to document \textbf{changes in socioeconomic inequality} in skills across populations?
 \ei
@@
   \item Millennium Cohort Study (MCS, b. 2000/1) \ei
-\item Construct a common 11-item, 2-dimensional scale from different instruments
-\item Investigate its psychometric properties (measurement invariance)
+\item Construct a common 11-item scale, from different instruments
+\item Investigate some psychometric properties
 \item Use scores from the new scale to document changes in socioeconomic gradients between cohorts 30 years apart
@@
 
-\section{Data}
+\section{Harmonised scale}
+
+\subsection{Data}
 
@@
 \item Data from 5 year follow-up (England only), maternal questionnaire
-\item Complex survey design: rebalanced data to make representative
+  \bi \item Complex survey design: rebalance sample to make representative \ei
 \item Strengths and Difficulties Questionnaire (SDQ, \citealp{Goodman1994})
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% ITEM CHOICE and MEAN TABLE
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Harmonisation}
-
-\begin{frame}
-\frametitle{Harmonisation}
-\framesubtitle{Choice of items}
-
-\bi
-\item Aim: develop a new scale using only items that have similar wording across the two cohorts
-\item Try to preserve three categories to maximise information
-\item Combine items when needed (e.g. when SDQ items are positively coded)
-\item Recode all items to higher categories for ``better" behaviours
-\ei
-
-\end{frame}
+\subsection{Common items}
 
@@
 \frametitle{Harmonisation}
-\framesubtitle{Items}
+\framesubtitle{Choice of common items}
 \centering
@@
 
-\section{Exploratory Analysis}
-
-\begin{frame}
-\frametitle{EFA}
-\framesubtitle{Number of factors}
-
-\bi
-\item Exploratory factor analysis on the two separate cohorts
-\item Two-factor structure seems ok according to various approaches
-\ei
-
+\subsection{Exploratory Analysis}
+
+\begin{frame}
+\frametitle{EFA}\label{efa}
+\framesubtitle{Factor structure}
+
+\bi
+\item Different methods for assessing number of factors give varied results \hyperlink{numfac}{\beamerbutton{Det.}}
+\item We choose two factors to maximise interpretability
+\item In both cohorts, EFA delivers a clear-cut split between two factors
+\item Factors can be coherently interpreted as internalising and internalising skills
+\ei 
+\vspace{12pt}
+\centering
+\begin{tabular}{C{.05\textwidth}L{.3\textwidth}|C{.05\textwidth}L{.3\textwidth}}
+  & \textbf{Externalising}  &    & \textbf{Internalising} \\
+1 & Restless                & 7  & Worried                \\
+2 & Squirmy/fidgety         & 8  & Afraid of new things   \\
+3 & Fights/bullies children & 9  & Solitary               \\
+4 & Cannot settle           & 10 & Unhappy/tearful        \\
+5 & Temper tantrums         & 11 & Head/stomachaches \\
+6 & Disobedient             &    &                       
+\end{tabular}
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% MEASUREMENT INVARIANCE
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Measurement Invariance}
+
+\begin{frame}
+\frametitle{Measurement Invariance}\label{mimain}
+\framesubtitle{Model setup}
+
+\bi
+\item Two constructs isolated in the new scale, separately by cohort
+\item But are externalising and internalising problems measured in a comparable way \emph{across cohorts}?
+\vspace{12pt}\pause
+\item \textbf{Measurement Invariance (MI)} analysis
+  \bi
+  \item Formal test of the invariance hypothesis
+  \item Set up categorical Item Factor Analysis (IFA) model \hyperlink{miapp1}{\beamerbutton{Det.}}
+  \item Pool the two cohorts together, and estimate models of different invariance level -- following \cite{Wu2016}:\hyperlink{miapp2}{\beamerbutton{Det.}}
+    \bnm
+    \item Start from minimally restrictive model, where parameters are free to vary
+    \item Restrict threshold and loading parameters to be equal across cohorts
+    \item Restrict intercept parameters as well
+    \enm
+  \item Test relative fit of restricted models compared to initial model
+  \ei 
+\ei
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+<<FA_INVARIANCE>>=
+# estimate models of different levels of invariance
+@
+
+<<FA_FIT, dependson='FA_INVARIANCE'>>=
+# assemble fit indices into matrices
+@
+
+<<FA_FINAL, dependson='FA_INVARIANCE'>>=
+# select final model set
+@
+
+<<FA_FINAL_PARS, dependson='FA_FINAL'>>=
+# get parameters of final model
+@
+
+<<CLEANUP, dependson='FA_FINAL'>>=
+@
+
+<<FA_SCORES, dependson='FA_FINAL'>>=
+# produce scores for different models
+@
+
+<<FA_SCORES_MERGE, dependson='LOAD_DATA'>>=
+# merge scores with other data
+@
+
+<<FA_SCORES_COMPARE, dependson='FA_SCORES'>>=
+# assemble and compare scores across different models
+@
+
+<<FA_SCORES_SELECT, dependson='FA_SCORES_COMPARE'>>=
+# select model to use for inequality analysis
+@
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% MEASUREMENT INVARIANCE RESULTS
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\frametitle{Measurement Invariance}\label{mires}
+\framesubtitle{Results}
+
+\bi
+\item Models with restricted intercepts do \emph{not} pass MI test
+  \bi
+  \item \cite{Wu2016} suggests we cannot compare latent means
+  \item Traditional approaches would disagree... \citep{Millsap2004}
+  \ei
+\item Might \emph{age differences} in the sample be playing a role?
+  \bi
+  \item BCS and MCS samples differ due to survey design \hyperlink{agehist}{\beamerbutton{Graph}}
+  \item But restricting the age range to maximise overlap does not give significantly different results \hyperlink{fit4}{\beamerbutton{Det.}}
+  \ei
+  \pause
+\item \textbf{Q1}: Can we derive comparable measures of noncognitive skills across cohorts?
+\item Yes, we can the model with non-invariant intercepts to score the latent factors (externalising and internalising skills)
+\ei
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% INEQUALITY
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Inequality in behaviour problems}
+
+\begin{frame}
+\frametitle{Inequality}
+
+\bi
+\item \textbf{Q2}: Has socioeconomic inequality in behavioural problems increased between 1975 and 2005?
+\item Investigate income gradient in externalising and internalising scores
+  \bi
+  \item Merge harmonised income data from CLOSER Work Package 2(a)
+  \item Weekly family income when child is 10 (not contemporaneous)
+  \item Working on access to harmonised social class data
+  \ei
+\ei
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Inequality}\label{meanci}
+\framesubtitle{Visualising the gradient \hyperlink{loessext}{\beamerbutton{EXT}}\hyperlink{loessint}{\beamerbutton{INT}}}
+\begin{figure}
+<<FACINEQ, dev='tikz', fig.height=6.7, fig.width=10, out.width='\\linewidth'>>=
+@
+\end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{Inequality}
+\framesubtitle{Quantifying the gradient -- inflation- and growth-adjusted income}
+\centering
+\resizebox{!}{4cm}{%
+<<FA_REGS_REAL, dependson='FA_SCORES_SELECT', results='asis'>>=
+# regress scores on income
+library(dplyr)
+library(stargazer)
+
+scores2plot$incq_real <- factor(ntile(scores2plot$faminc_real, 5))
+
+# get min, max
+minmax.real <-aggregate(scores2plot[,c("faminc_real")],list(scores2plot$incq_real), 
+            FUN = function(x) c(min = min(x, na.rm=T), max = max(x, na.rm=T)))
+minmax.real <- round(minmax.real[,-1],2)*1000
+
+r.ext.f.q <- lm(EXT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="F"))
+r.int.f.q <- lm(INT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="F"))
+
+r.ext.m.q <- lm(EXT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="M"))
+r.int.m.q <- lm(INT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="M"))
+
+stargazer(r.ext.m.q, r.int.m.q, r.ext.f.q, r.int.f.q,
+          column.labels=c("Males","Females"), column.separate=c(2,2),
+          dep.var.caption = "", multicolumn = FALSE,
+          omit=c("Constant"),
+          covariate.labels=c(
+                paste0("Income Q2 [",toString(minmax.real[2,1]),"-",toString(minmax.real[2,2])," £/w]"),
+                paste0("Income Q3 [",toString(minmax.real[3,1]),"-",toString(minmax.real[3,2])," £/w]"),
+                paste0("Income Q4 [",toString(minmax.real[4,1]),"-",toString(minmax.real[4,2])," £/w]"),
+                paste0("Income Q5 [",toString(minmax.real[5,1]),"+ £/w]"),
+                             "MCS dummy",
+                             "Income Q2 x MCS",
+                             "Income Q3 x MCS",
+                             "Income Q4 x MCS",
+                             "Income Q5 x MCS"
+                             ),
+          no.space=TRUE, 
+          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
+          float=FALSE    # needs begin table outside it
+)
+@
+}
+\end{frame}
+
+\begin{frame}
+\frametitle{Inequality}
+\framesubtitle{Quantifying the gradient -- males}
+\centering
+\resizebox{12cm}{4cm}{%
+<<FA_REGS_REAL_M, dependson='FA_SCORES', results='asis'>>=
+# regress scores on income
+library(dplyr)
+library(stargazer)
+
+r.ext.m.bcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="BCS"))
+r.ext2.m.bcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="BCS"))
+r.ext.m.mcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="MCS"))
+r.ext2.m.mcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="MCS"))
+r.int.m.bcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="BCS"))
+r.int2.m.bcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="BCS"))
+r.int.m.mcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="MCS"))
+r.int2.m.mcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="MCS"))
+
+stargazer(r.ext.m.bcs, r.ext2.m.bcs, r.int.m.bcs, r.int2.m.bcs, r.ext.m.mcs, r.ext2.m.mcs, r.int.m.mcs, r.int2.m.mcs,
+          column.labels=c("BCS","MCS"), column.separate=c(4,4),
+          dep.var.caption = "", multicolumn = FALSE,
+          omit=c("Constant"),
+          covariate.labels=c(
+                             "Income Q2",
+                             "Income Q3",
+                             "Income Q4",
+                             "Income Q5",
+                             "Birthweight",
+                             "Mat. Smoking in preg.",
+                             "Father years of schooling",
+                             "Mother years of schooling"
+                             ),
+          no.space=TRUE, 
+          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
+          float=FALSE    # needs begin table outside it
+)
+@
+}
+\end{frame}
+
+\begin{frame}
+\frametitle{Inequality}
+\framesubtitle{Quantifying the gradient -- females}
+\centering
+\resizebox{12cm}{4cm}{%
+<<FA_REGS_REAL_F, dependson='FA_SCORES', results='asis'>>=
+# regress scores on income
+library(dplyr)
+library(stargazer)
+
+r.ext.f.bcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="BCS"))
+r.ext2.f.bcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="BCS"))
+r.ext.f.mcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="MCS"))
+r.ext2.f.mcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="MCS"))
+r.int.f.bcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="BCS"))
+r.int2.f.bcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="BCS"))
+r.int.f.mcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="MCS"))
+r.int2.f.mcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="MCS"))
+
+stargazer(r.ext.f.bcs, r.ext2.f.bcs, r.int.f.bcs, r.int2.f.bcs, r.ext.f.mcs, r.ext2.f.mcs, r.int.f.mcs, r.int2.f.mcs,
+          column.labels=c("BCS","MCS"), column.separate=c(4,4),
+          dep.var.caption = "", multicolumn = FALSE,
+          omit=c("Constant"),
+          covariate.labels=c(
+                             "Income Q2",
+                             "Income Q3",
+                             "Income Q4",
+                             "Income Q5",
+                             "Birthweight",
+                             "Mat. Smoking in preg.",
+                             "Father years of schooling",
+                             "Mother years of schooling"
+                             ),
+          no.space=TRUE, 
+          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
+          float=FALSE    # needs begin table outside it
+)
+@
+}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% CONCLUSION
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Conclusions and next steps}
+
+\begin{frame}
+\frametitle{Conclusions and next steps}
+
+\bi
+\item We derived a comparable scale of non-cognitive skills / behaviour problems
+\item We have shown that inequality (across the income dimension) has increased
+  \bi
+  \item Profile has become steeper (particularly for males)
+  \item Scores at upper quantiles have drifted upwards
+  \ei
+\pause
+\vspace{12pt}
+\item Next steps
+  \bi
+  \item Harmonise more cohorts (NCDS) and ages (10)
+  \item Extend analysis to the \emph{returns} to non-cognitive skills -- are they remunerated differently on the labour market?
+  \ei
+\ei
+
+\end{frame}
+
+\begin{frame}[noframenumbering]
+\centering
+{\Large Thank you!}
+\end{frame}
+
+% \begin{frame}
+% \frametitle{Returns}
+% \framesubtitle{Education returns to behaviour skills in BCS}
+% \centering
+% \resizebox{!}{4cm}{%
+% <<FA_RETURNS, dependson='FA_SCORES_SELECT', results='asis'>>=
+% # regress scores on income
+% library(dplyr)
+% library(stargazer)
+% 
+% # get BCS data and make dummy for higher education
+% scoresBCS <- scores2plot[scores2plot$cohort=="BCS",]
+% scoresBCS$highered <- as.numeric(scoresBCS$hinvq00>3)
+% scoresBCS[is.na(scoresBCS$hinvq00), "highered"] <- NA
+% 
+% # females
+% r.nvq.f <- list()
+% r.nvq.f[[1]] <- lm(highered ~ EXT + INT, data=subset(scoresBCS, sex=="F"))
+% r.nvq.f[[2]] <- lm(highered ~ EXT + INT + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="F"))
+% r.nvq.f[[3]] <- lm(highered ~ EXT + INT + factor(incq), data=subset(scoresBCS, sex=="F"))
+% r.nvq.f[[4]] <- lm(highered ~ EXT + INT + ysch_moth + ysch_fath, data=subset(scoresBCS, sex=="F"))
+% r.nvq.f[[5]] <- lm(highered ~ EXT + INT + factor(incq) + ysch_moth + ysch_fath + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="F"))
+% 
+% # females
+% r.nvq.m <- list()
+% r.nvq.m[[1]] <- lm(highered ~ EXT + INT, data=subset(scoresBCS, sex=="M"))
+% r.nvq.m[[2]] <- lm(highered ~ EXT + INT + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="M"))
+% r.nvq.m[[3]] <- lm(highered ~ EXT + INT + factor(incq), data=subset(scoresBCS, sex=="M"))
+% r.nvq.m[[4]] <- lm(highered ~ EXT + INT + ysch_moth + ysch_fath, data=subset(scoresBCS, sex=="M"))
+% r.nvq.m[[5]] <- lm(highered ~ EXT + INT + factor(incq) + ysch_moth + ysch_fath + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="M"))
+% 
+% stargazer(r.nvq.m[[1]], r.nvq.m[[2]], r.nvq.m[[3]], r.nvq.m[[4]], r.nvq.m[[5]],
+%           r.nvq.f[[1]], r.nvq.f[[2]], r.nvq.f[[3]], r.nvq.f[[4]], r.nvq.f[[5]],
+%           dep.var.caption = "Completed Higher Ed. at 30",
+%           column.labels=c("Males","Females"), column.separate=c(5,5),
+%           omit=c("Constant"),
+%           covariate.labels=c("Externalising skill at 5",
+%                              "Internalising skill at 5",
+%                              "EPVT Z at 5",
+%                              "Human figure drawing Z at 5",
+%                              "Copy design Z at 5",
+%                              "Family inc. at 10 Q2",
+%                              "Family inc. at 10 Q3",
+%                              "Family inc. at 10 Q4",
+%                              "Family inc. at 10 Q5",
+%                              "Mother's years of schooling",
+%                              "Father's years of schooling"
+%                              ),
+%           no.space=TRUE, 
+%           omit.stat= c("rsq", "f", "ser", "aic", "ll"),
+%           float=FALSE    # needs begin table outside it
+% )
+% @
+% }
+% \end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% MEASUREMENT INVARIANCE APPENDIX
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}[noframenumbering]
+\centering
+{\Large Appendix}
+\end{frame}
+
+\begin{frame}[noframenumbering]
+\frametitle{EFA}\label{numfac}
+\framesubtitle{Number of factors \hyperlink{efa}{\beamerreturnbutton{Back}}}
 \begin{center}
@@
 
-\begin{frame}
-\frametitle{EFA}
-\framesubtitle{Factor structure}
-\bi
-\item In both cohorts, delivers a clear-cut split between two factors
-\item Factors can be coherently interpreted as internalising and internalising behaviour problems
-\ei 
-\vspace{12pt}
-\centering
-\begin{tabular}{C{.05\textwidth}L{.3\textwidth}|C{.05\textwidth}L{.3\textwidth}}
-  & \textbf{Externalising}  &    & \textbf{Internalising} \\
-1 & Restless                & 7  & Worried                \\
-2 & Squirmy/fidgety         & 8  & Afraid of new things   \\
-3 & Fights/bullies children & 9  & Solitary               \\
-4 & Cannot settle           & 10 & Unhappy/tearful        \\
-5 & Temper tantrums         & 11 & Head/stomachaches \\
-6 & Disobedient             &    &                       
-\end{tabular}
-
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% MEASUREMENT INVARIANCE
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Measurement Invariance}
-
-\begin{frame}
-\frametitle{Measurement Invariance}
-
-\bi
-\item Two constructs isolated in the new scale, separately by cohort
-\item But are externalising and internalising problems measured in a comparable way \emph{across cohorts}?
-  \bi
-  \item Slightly different wording of question and options
-  \item Different ordering in questionnaire
-  \ei 
-\vspace{12pt}\pause
-\item \textbf{Measurement Invariance (MI)} analysis
-  \bi
-  \item Factor analysis -- but can also use IRT
-  \item Pool the two cohorts together
-  \item Start from minimally restrictive model, where parameters are free to vary across cohorts
-  \item Place increasing restrictions on parameters across cohorts and test fit
-  \ei 
-\ei
-
-\end{frame}
-
-\begin{frame}
-\frametitle{MI}\label{mimain}
-\framesubtitle{Invariance steps}
-
-\bi
-\item Set up categorical Item Factor Analysis (IFA) model \hyperlink{miapp1}{\beamerbutton{Det.}}
-\item Specify and compare model of different invariance levels
-
-\bi
-\item Approach in \citep{Wu2016} \hyperlink{miapp2}{\beamerbutton{Det.}}
+\begin{frame}[noframenumbering]
+\frametitle{MI}\label{miapp1}
+\framesubtitle{IFA model}
+Categorical Item Factor Analysis (IFA) model \hyperlink{mimain}{\beamerreturnbutton{Back}}
+
+\bi
+\item Threshold model \citep{Muthen1984}
+\item Latent propensity to endorse each item $i$ for child $j$ in cohort $c$
+\begin{align}\label{eq:propens}
+\begin{split}
+X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{EXT}_{jc} + u_{ijc} \qquad \text{for} \; i=1,\dots, 6 \\
+X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{INT}_{jc} + u_{ijc} \qquad \text{for} \; i=7,\dots, 11
+\end{split}
+\end{align}
+
+\item Mean/covariance structure
+$$\bm{\mu}_c=\bm{\nu}_c + \bm{\Lambda}_c\bm{\kappa}_c \qquad \text{and} \qquad \Sigma_c = \bm{\Lambda}_c \bm{\Phi}_c \bm{\Lambda}_c^\prime + \bm{\Psi}_c.$$
+with $$(\theta^{EXT}_{jc}, \theta^{INT}_{jc}) = \bm{\Theta}_{jc} \sim N(\bm{\kappa}_c, \bm{\Phi}_c) \qquad \text{and} \qquad \bm{u}_{jc} \sim N(\bm{0}, \bm{\Psi}_c)$$
+
+\item Discreteness
+$$X_{ijc} = s \qquad \text{if} \; \tau_{s,ic} \leq X^*_{ijc} < \tau_{s+1,ic} \qquad \text{for} \; s=0,1,2$$
+with $\tau_{0,ic}=-\infty$ and $\tau_{3,ic}=+\infty$.
+
+\ei
+\end{frame}
+
+\begin{frame}[noframenumbering]
+\frametitle{MI}
+\framesubtitle{Key Parameters for Invariance}
+
+\bi
+\item Loadings $\lambda_{ci}$
+\item Thresholds $\tau_{1,ci}$, $\tau_{2,ci}$
+\item Intercepts $\nu_{ci}$
+\item Unique variances $\Var(u_{ci})$
+\item Latent mean vector $\bm{\kappa}_c$
+\item Latent covariance matrix $\bm{\Phi}_c$
+\ei
+
+\end{frame}
+
+\begin{frame}[noframenumbering]
+\frametitle{MI}\label{miapp2}
+\framesubtitle{Invariance steps \hyperlink{mimain}{\beamerreturnbutton{Back}}}
+\Fontvi
+\bi
 \item \textbf{Configural} model
@@
   \item Minimally restrictive (all parameters free across cohorts)
+  \item $\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}, \quad \bm{\nu}=\bm{0}, \quad \text{and} \; \text{diag}(\bm{\Psi})=\bm{I}.$
+  \ei
+
+\item \textbf{Threshold invariance} model
+  \bi
+  \item $\tau_{1,Ai} = \tau_{1,Bi}$ for $i=1,\dots,11$, $\tau_{2,Ai} = \tau_{2,Bi}$ for $i=\{1,2,3,4,7,8,9,10\}$
+  \item For all groups: $\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}$
+  \item For the reference group: $\bm{\nu}_A=\bm{0}, \quad \; \text{diag}(\bm{\Sigma}_A)=\bm{I}$
   \ei
@@
   \bi
-  \item Thresholds and loadings are restricted to be invariant across cohorts
-  \item Means set to zero in both grups, variances set to 1 in reference cohort only
+  \item $\tau_{1,Ai} = \tau_{1,Bi}$ for $i=1,\dots,11$, $\tau_{2,Ai} = \tau_{2,Bi}$ for $i=\{1,2,3,4,7,8,9,10\}$
+  \item $\lambda_{Ai} = \lambda_{Bi}$ for $i=1,\dots,11$
+  \item For all groups: $\bm{\kappa}=\bm{0}$
+  \item For the reference group: $\bm{\nu}_A=\bm{0}, \quad \; \text{diag}(\bm{\Sigma}_A)=\bm{I}, \quad \text{diag}(\bm{\Phi}_A) = \bm{I}$
   \item Allows comparison of latent factor variances
@@
   \bi
-  \item Thresholds, loadings, intercepts restricted to be invariant across cohorts
-  \item Mean and variances unrestricted in target cohort
+  \item $\tau_{1,Ai} = \tau_{1,Bi}$ for $i=1,\dots,11$, $\tau_{2,Ai} = \tau_{2,Bi}$ for $i=\{1,2,3,4,7,8,9,10\}$
+  \item $\lambda_{Ai} = \lambda_{Bi}$ for $i=1,\dots,11$
+  \item For all groups: $\bm{\nu}=\bm{0}$
+  \item For the reference group: $\bm{\kappa}_A=\bm{0}, \quad \; \text{diag}(\bm{\Sigma}_A)=\bm{I}, \quad \text{diag}(\bm{\Phi}_A) = \bm{I}$
   \item Allows comparison of latent factor means and variances
@@
 \ei
-\ei
-
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-<<FA_INVARIANCE>>=
-# estimate models of different levels of invariance
-@
-
-<<FA_FIT, dependson='FA_INVARIANCE'>>=
-# assemble fit indices into matrices
-@
-
-<<FA_FINAL, dependson='FA_INVARIANCE'>>=
-# select final model set
-@
-
-<<FA_FINAL_PARS, dependson='FA_FINAL'>>=
-# get parameters of final model
-@
-
-<<CLEANUP, dependson='FA_FINAL'>>=
-@
-
-<<FA_SCORES, dependson='FA_FINAL'>>=
-# produce scores for different models
-@
-
-<<FA_SCORES_MERGE, dependson='LOAD_DATA'>>=
-# merge scores with other data
-@
-
-<<FA_SCORES_COMPARE, dependson='FA_SCORES'>>=
-# assemble and compare scores across different models
-@
-
-<<FA_SCORES_SELECT, dependson='FA_SCORES_COMPARE'>>=
-# select model to use for inequality analysis
-@
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% MEASUREMENT INVARIANCE RESULTS
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\end{frame}
 
@@
 
-\begin{frame}
-\frametitle{MI}\label{mires}
-\framesubtitle{Results}
-
-\bi
-\item Models with restricted intercepts do not pass MI test -- although the absolute fit is not terrible
-  \bi
-  \item $\Rightarrow$ cannot compare latent means according to \cite{Wu2016}
-  \item but more traditional approaches with item anchoring would disagree \cite{Millsap2004}
-  \ei
-\item There are age differences between BCS and MCS due to survey design \hyperlink{agehist}{\beamerbutton{Graph}}
-\item But restricting the age range to maximise overlap does not give significantly different results \hyperlink{fit4}{\beamerbutton{Graph}}
-\item {\color{red} Can still use the model with non-invariant intercepts to score the latent factors and investigate inequality}
-\ei
-
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% INEQUALITY
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Inequality in behaviour problems}
-
-\begin{frame}
-\frametitle{Inequality}
-
-\bi
-\item Has inequality in behavioural problems increased between 1975 and 2005?
-\item Investigate income gradient in externalising and internalising scores
-  \bi
-  \item Merge harmonised income data from CLOSER Work Package 2(a)
-  \item Weekly family income when child is 10
-  \item Would be more interesting with harmonised social class...
-  \ei
-\ei
-
-\end{frame}
-
-\begin{frame}
-\frametitle{Inequality}
-\framesubtitle{Visualising the gradient}
-\begin{figure}
-<<FACINEQ, dev='tikz', fig.height=6.7, fig.width=10, out.width='\\linewidth'>>=
-@
-\end{figure}
-\end{frame}
-
-<<LOESS_REAL, dependson='FA_SCORES_SELECT'>>=
-@
-
-\begin{frame}
-\frametitle{Inequality}
-\framesubtitle{Visualising the gradient -- inflation- and growth-adjusted income}
-\begin{figure}
-<<LOESS_REAL_EXT, dependson='LOESS_REAL', dev='tikz', fig.height=8, fig.width=12, out.width='.95\\linewidth'>>=
-ploess.real.ext
-@
-\end{figure}
-\end{frame}
-
-\begin{frame}
-\frametitle{Inequality}
-\framesubtitle{Visualising the gradient -- inflation- and growth-adjusted income}
-\begin{figure}
-<<LOESS_REAL_INT, dependson='LOESS_REAL', dev='tikz', fig.height=8, fig.width=12, out.width='.95\\linewidth'>>=
-ploess.real.int
-@
-\end{figure}
-\end{frame}
-
-\begin{frame}
-\frametitle{Inequality}
-\framesubtitle{Quantifying the gradient -- inflation- and growth-adjusted income}
-\centering
-\resizebox{!}{4cm}{%
-<<FA_REGS_REAL, dependson='FA_SCORES_SELECT', results='asis'>>=
-# regress scores on income
-library(dplyr)
-library(stargazer)
-
-scores2plot$incq_real <- factor(ntile(scores2plot$faminc_real, 5))
-
-# get min, max
-minmax.real <-aggregate(scores2plot[,c("faminc_real")],list(scores2plot$incq_real), 
-            FUN = function(x) c(min = min(x, na.rm=T), max = max(x, na.rm=T)))
-minmax.real <- round(minmax.real[,-1],2)*1000
-
-r.ext.f.q <- lm(EXT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="F"))
-r.int.f.q <- lm(INT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="F"))
-
-r.ext.m.q <- lm(EXT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="M"))
-r.int.m.q <- lm(INT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="M"))
-
-stargazer(r.ext.m.q, r.int.m.q, r.ext.f.q, r.int.f.q,
-          column.labels=c("Males","Females"), column.separate=c(2,2),
-          dep.var.caption = "", multicolumn = FALSE,
-          omit=c("Constant"),
-          covariate.labels=c(
-                paste0("Income Q2 [",toString(minmax.real[2,1]),"-",toString(minmax.real[2,2])," £/w]"),
-                paste0("Income Q3 [",toString(minmax.real[3,1]),"-",toString(minmax.real[3,2])," £/w]"),
-                paste0("Income Q4 [",toString(minmax.real[4,1]),"-",toString(minmax.real[4,2])," £/w]"),
-                paste0("Income Q5 [",toString(minmax.real[5,1]),"+ £/w]"),
-                             "MCS dummy",
-                             "Income Q2 x MCS",
-                             "Income Q3 x MCS",
-                             "Income Q4 x MCS",
-                             "Income Q5 x MCS"
-                             ),
-          no.space=TRUE, 
-          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
-          float=FALSE    # needs begin table outside it
-)
-@
-}
-\end{frame}
-
-\begin{frame}
-\frametitle{Inequality}
-\framesubtitle{Quantifying the gradient -- males}
-\centering
-\resizebox{12cm}{4cm}{%
-<<FA_REGS_REAL_M, dependson='FA_SCORES', results='asis'>>=
-# regress scores on income
-library(dplyr)
-library(stargazer)
-
-r.ext.m.bcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="BCS"))
-r.ext2.m.bcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="BCS"))
-r.ext.m.mcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="MCS"))
-r.ext2.m.mcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="MCS"))
-r.int.m.bcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="BCS"))
-r.int2.m.bcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="BCS"))
-r.int.m.mcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="MCS"))
-r.int2.m.mcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="MCS"))
-
-stargazer(r.ext.m.bcs, r.ext2.m.bcs, r.int.m.bcs, r.int2.m.bcs, r.ext.m.mcs, r.ext2.m.mcs, r.int.m.mcs, r.int2.m.mcs,
-          column.labels=c("BCS","MCS"), column.separate=c(4,4),
-          dep.var.caption = "", multicolumn = FALSE,
-          omit=c("Constant"),
-          covariate.labels=c(
-                             "Income Q2",
-                             "Income Q3",
-                             "Income Q4",
-                             "Income Q5",
-                             "Birthweight",
-                             "Mat. Smoking in preg.",
-                             "Father years of schooling",
-                             "Mother years of schooling"
-                             ),
-          no.space=TRUE, 
-          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
-          float=FALSE    # needs begin table outside it
-)
-@
-}
-\end{frame}
-
-\begin{frame}
-\frametitle{Inequality}
-\framesubtitle{Quantifying the gradient -- females}
-\centering
-\resizebox{12cm}{4cm}{%
-<<FA_REGS_REAL_F, dependson='FA_SCORES', results='asis'>>=
-# regress scores on income
-library(dplyr)
-library(stargazer)
-
-r.ext.f.bcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="BCS"))
-r.ext2.f.bcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="BCS"))
-r.ext.f.mcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="MCS"))
-r.ext2.f.mcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="MCS"))
-r.int.f.bcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="BCS"))
-r.int2.f.bcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="BCS"))
-r.int.f.mcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="MCS"))
-r.int2.f.mcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="MCS"))
-
-stargazer(r.ext.f.bcs, r.ext2.f.bcs, r.int.f.bcs, r.int2.f.bcs, r.ext.f.mcs, r.ext2.f.mcs, r.int.f.mcs, r.int2.f.mcs,
-          column.labels=c("BCS","MCS"), column.separate=c(4,4),
-          dep.var.caption = "", multicolumn = FALSE,
-          omit=c("Constant"),
-          covariate.labels=c(
-                             "Income Q2",
-                             "Income Q3",
-                             "Income Q4",
-                             "Income Q5",
-                             "Birthweight",
-                             "Mat. Smoking in preg.",
-                             "Father years of schooling",
-                             "Mother years of schooling"
-                             ),
-          no.space=TRUE, 
-          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
-          float=FALSE    # needs begin table outside it
-)
-@
-}
-\end{frame}
-
-\begin{frame}
-\frametitle{Inequality}
-
-\bi
-  \item The difference in scores between top and bottom quintiles has increased
-  \item Scores at top quintiles have drifted upwards
-\ei
-
-\end{frame}
-
-\begin{frame}
-\frametitle{Returns}
-\framesubtitle{Education returns to behaviour skills in BCS}
-\centering
-\resizebox{!}{4cm}{%
-<<FA_RETURNS, dependson='FA_SCORES_SELECT', results='asis'>>=
-# regress scores on income
-library(dplyr)
-library(stargazer)
-
-# get BCS data and make dummy for higher education
-scoresBCS <- scores2plot[scores2plot$cohort=="BCS",]
-scoresBCS$highered <- as.numeric(scoresBCS$hinvq00>3)
-scoresBCS[is.na(scoresBCS$hinvq00), "highered"] <- NA
-
-# females
-r.nvq.f <- list()
-r.nvq.f[[1]] <- lm(highered ~ EXT + INT, data=subset(scoresBCS, sex=="F"))
-r.nvq.f[[2]] <- lm(highered ~ EXT + INT + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="F"))
-r.nvq.f[[3]] <- lm(highered ~ EXT + INT + factor(incq), data=subset(scoresBCS, sex=="F"))
-r.nvq.f[[4]] <- lm(highered ~ EXT + INT + ysch_moth + ysch_fath, data=subset(scoresBCS, sex=="F"))
-r.nvq.f[[5]] <- lm(highered ~ EXT + INT + factor(incq) + ysch_moth + ysch_fath + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="F"))
-
-# females
-r.nvq.m <- list()
-r.nvq.m[[1]] <- lm(highered ~ EXT + INT, data=subset(scoresBCS, sex=="M"))
-r.nvq.m[[2]] <- lm(highered ~ EXT + INT + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="M"))
-r.nvq.m[[3]] <- lm(highered ~ EXT + INT + factor(incq), data=subset(scoresBCS, sex=="M"))
-r.nvq.m[[4]] <- lm(highered ~ EXT + INT + ysch_moth + ysch_fath, data=subset(scoresBCS, sex=="M"))
-r.nvq.m[[5]] <- lm(highered ~ EXT + INT + factor(incq) + ysch_moth + ysch_fath + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="M"))
-
-stargazer(r.nvq.m[[1]], r.nvq.m[[2]], r.nvq.m[[3]], r.nvq.m[[4]], r.nvq.m[[5]],
-          r.nvq.f[[1]], r.nvq.f[[2]], r.nvq.f[[3]], r.nvq.f[[4]], r.nvq.f[[5]],
-          dep.var.caption = "Completed Higher Ed. at 30",
-          column.labels=c("Males","Females"), column.separate=c(5,5),
-          omit=c("Constant"),
-          covariate.labels=c("Externalising skill at 5",
-                             "Internalising skill at 5",
-                             "EPVT Z at 5",
-                             "Human figure drawing Z at 5",
-                             "Copy design Z at 5",
-                             "Family inc. at 10 Q2",
-                             "Family inc. at 10 Q3",
-                             "Family inc. at 10 Q4",
-                             "Family inc. at 10 Q5",
-                             "Mother's years of schooling",
-                             "Father's years of schooling"
-                             ),
-          no.space=TRUE, 
-          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
-          float=FALSE    # needs begin table outside it
-)
-@
-}
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% MEASUREMENT INVARIANCE APPENDIX
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{frame}[noframenumbering]
-\centering
-{\Large Appendix}
-\end{frame}
-
-\begin{frame}[noframenumbering]
-\frametitle{MI}\label{miapp1}
-\framesubtitle{IFA model}
-Categorical Item Factor Analysis (IFA) model \hyperlink{mimain}{\beamerreturnbutton{Back}}
-
-\bi
-\item Threshold model \citep{Muthen1984}
-\item Latent propensity to endorse each item $i$ for child $j$ in cohort $c$
-\begin{align}\label{eq:propens}
-\begin{split}
-X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{EXT}_{jc} + u_{ijc} \qquad \text{for} \; i=1,\dots, 6 \\
-X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{INT}_{jc} + u_{ijc} \qquad \text{for} \; i=7,\dots, 11
-\end{split}
-\end{align}
-
-\item Mean/covariance structure
-$$\bm{\mu}_c=\bm{\nu}_c + \bm{\Lambda}_c\bm{\kappa}_c \qquad \text{and} \qquad \Sigma_c = \bm{\Lambda}_c \bm{\Phi}_c \bm{\Lambda}_c^\prime + \bm{\Psi}_c.$$
-with $$(\theta^{EXT}_{jc}, \theta^{INT}_{jc}) = \bm{\Theta}_{jc} \sim N(\bm{\kappa}_c, \bm{\Phi}_c) \qquad \text{and} \qquad \bm{u}_{jc} \sim N(\bm{0}, \bm{\Psi}_c)$$
-
-\item Discreteness
-$$X_{ijc} = s \qquad \text{if} \; \tau_{s,ic} \leq X^*_{ijc} < \tau_{s+1,ic} \qquad \text{for} \; s=0,1,2$$
-with $\tau_{0,ic}=-\infty$ and $\tau_{3,ic}=+\infty$.
-
-\ei
-\end{frame}
-
-\begin{frame}[noframenumbering]
-\frametitle{MI}
-\framesubtitle{Key Parameters for Invariance}
-
-\bi
-\item Loadings $\lambda_{ci}$
-\item Thresholds $\tau_{1,ci}$, $\tau_{2,ci}$
-\item Intercepts $\nu_{ci}$
-\item Unique variances $\Var(u_{ci})$
-\item Latent mean vector $\bm{\kappa}_c$
-\item Latent covariance matrix $\bm{\Phi}_c$
-\ei
-
-\end{frame}
-
-\begin{frame}[noframenumbering]
-\frametitle{MI}\label{miapp2}
-\framesubtitle{Invariance steps \hyperlink{mimain}{\beamerreturnbutton{Back}}}
-\Fontvi
-\bi
-\item \textbf{Configural} model
-$$\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}, \quad \bm{\nu}=\bm{0}, \quad \text{and} \; \text{diag}(\bm{\Psi})=\bm{I}.$$
-
-\item \textbf{Threshold invariance} model
-  \bi
-  \item $\tau_{1,Ai} = \tau_{1,Bi}$ for $i=1,\dots,11$, $\tau_{2,Ai} = \tau_{2,Bi}$ for $i=\{1,2,3,4,7,8,9,10\}$
-  \item For all groups: $\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}$
-  \item For the reference group: $\bm{\nu}_A=\bm{0}, \quad \; \text{diag}(\bm{\Sigma}_A)=\bm{I}$
-  \ei
-\item \textbf{Threshold and Loading invariance} model
-  \bi
-  \item $\tau_{1,Ai} = \tau_{1,Bi}$ for $i=1,\dots,11$, $\tau_{2,Ai} = \tau_{2,Bi}$ for $i=\{1,2,3,4,7,8,9,10\}$
-  \item $\lambda_{Ai} = \lambda_{Bi}$ for $i=1,\dots,11$
-  \item For all groups: $\bm{\kappa}=\bm{0}$
-  \item For the reference group: $\bm{\nu}_A=\bm{0}, \quad \; \text{diag}(\bm{\Sigma}_A)=\bm{I}, \quad \text{diag}(\bm{\Phi}_A) = \bm{I}$
-  \ei
-\item \textbf{Threshold, Loading, and Intercept invariance} model
-  \bi
-  \item $\tau_{1,Ai} = \tau_{1,Bi}$ for $i=1,\dots,11$, $\tau_{2,Ai} = \tau_{2,Bi}$ for $i=\{1,2,3,4,7,8,9,10\}$
-  \item $\lambda_{Ai} = \lambda_{Bi}$ for $i=1,\dots,11$
-  \item For all groups: $\bm{\nu}=\bm{0}$
-  \item For the reference group: $\bm{\kappa}_A=\bm{0}, \quad \; \text{diag}(\bm{\Sigma}_A)=\bm{I}, \quad \text{diag}(\bm{\Phi}_A) = \bm{I}$
-  \ei
-\ei
-
-\end{frame}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \end{figure}
-\end{frame}
-
-\begin{frame}[noframenumbering]
-\frametitle{MI}
-\framesubtitle{Fit comparison -- Robustness}
-List of estimated models:
-\bnm
-\item no gender split, no age adjustment (2 groups, BCS MCS)
-\item no gender split, age adjustment (2 groups, BCS MCS)
-\item males only, no age adjustment (2 groups, BCS MCS)
-\item females only, no age adjustment (2 groups, BCS MCS)
-\item males only, age adjustment (2 groups, BCS MCS)
-\item females only, age adjustment (2 groups, BCS MCS)
-\item separate gender groups, no age adjustment (4 groups, BCS.M BCS.F MCS.M MCS.F)
-\item separate gender groups, age adjustment (4 groups, BCS.M BCS.F MCS.M MCS.F)
-\item separate gender groups, overlapping ages, no age adjustment (4 groups, BCS.M BCS.F MCS.M MCS.F)
-\item separate gender groups, age adjustment constrained to be the same across ages (4 groups, BCS.M BCS.F MCS.M MCS.F)
-\item MIMIC model with age as regressor (4 groups, BCS.M BCS.F MCS.M MCS.F)
-\enm
-
-\end{frame}
-
-\begin{frame}[noframenumbering]
-\frametitle{MI}
-\framesubtitle{Fit comparison -- 2 group models, age unadjusted}
-
-\begin{table}[ht]
-\centering
-\small
-<<FA_FIT_RES2a, dependson="FA_FIT", results='asis'>>=
-require(xtable)
-
-table <- rbind(afitab[[1]],afitab[[3]],afitab[[4]])
-
-print(xtable(table,
-                    align="L{0cm}L{.20\\textwidth}C{.03\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}",
-                    digits = 4
-                    ),
-             size="\\tiny",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = c(list(0,3,6)),
-                              command = c(
-                              # header
-            "Model & Num. params & $\\chi^2$ & RMSEA & MFI & CFI & G-hat & \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ \\midrule
-                       \\textbf{(1) No gender split, not age adjusted} \\\\",
-                      "\\newline \\textbf{(3) Males only, not age adjusted} \\\\ \\midrule ",
-                      "\\newline \\textbf{(4) Females only, not age adjusted} \\\\ \\midrule "
-                                    )),
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabular", width = "\\textwidth",
-             floating = FALSE
-)
-
-@
-\end{table}
-
-\end{frame}
-
-\begin{frame}[noframenumbering]
-\frametitle{MI}
-\framesubtitle{Fit comparison -- 2 group models, age adjusted}
-\begin{table}[ht]
-\centering
-\small
-<<FA_FIT_RES2b, dependson="FA_FIT", results='asis'>>=
-require(xtable)
-
-table <- rbind(afitab[[2]],afitab[[5]],afitab[[6]])
-
-print(xtable(table,
-                    align="L{0cm}L{.20\\textwidth}C{.03\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}",
-                    digits = 4
-                    ),
-             size="\\tiny",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = c(list(0,3,6)),
-                              command = c(
-                              # header
-            "Model & Num. params & $\\chi^2$ & RMSEA & MFI & CFI & G-hat & \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ \\midrule
-                       \\textbf{(2) No gender split, age adjusted} \\\\ ",
-                      "\\newline \\textbf{(5) Males only, age adjusted} \\\\ \\midrule ",
-                      "\\newline \\textbf{(6) Females only, age adjusted} \\\\ \\midrule "
-                                    )),
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabular", width = "\\textwidth",
-             floating = FALSE
-)
-
-@
-\end{table}
 \end{frame}
@@
 
+<<LOESS_REAL, dependson='FA_SCORES_SELECT'>>=
+@
+
+\begin{frame}
+\frametitle{Inequality \hyperlink{meanci}{\beamerreturnbutton{Back}}}\label{loessext}
+\framesubtitle{Visualising the gradient -- inflation- and growth-adjusted income}
+\begin{figure}
+<<LOESS_REAL_EXT, dependson='LOESS_REAL', dev='tikz', fig.height=8, fig.width=12, out.width='.95\\linewidth'>>=
+ploess.real.ext
+@
+\end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{Inequality \hyperlink{meanci}{\beamerreturnbutton{Back}}}\label{loessint}
+\framesubtitle{Visualising the gradient -- inflation- and growth-adjusted income}
+\begin{figure}
+<<LOESS_REAL_INT, dependson='LOESS_REAL', dev='tikz', fig.height=8, fig.width=12, out.width='.95\\linewidth'>>=
+ploess.real.int
+@
+\end{figure}
+\end{frame}
+
 <<LOESS_INFL, dependson="FA_SCORES_SELECT">>=

### 2018-03-27 codes/R/6_ineq.R
inequality decompositions (JMP)

@@ -1,0 +1,140 @@
+# INEQUALITY DECOMPOSITION
+
+#######################################################################################################
+## ---- IQRDEC_FUN
+# functions
+
+# function to get centiles
+getcent <- function(var){
+  pc <- c()
+  for (p in seq(1:99)/100) {
+    pc <- c(pc,quantile(var,p, na.rm=T))
+  }
+  return(pc)
+}
+
+# function for quantile differences
+qdiff <- function(data, q2=.75, q1=.25) {
+  pc1 <- quantile(data, q1)
+  pc2 <- quantile(data, q2)
+  iqr <- pc2-pc1 
+  return(iqr)
+}
+
+# function for quantile differences table
+iqtab <- function(var, by) {
+  tab <- aggregate(var, by=by, 
+                   FUN = function(x) c(
+                     iq9010 = round(qdiff(x, .9, .1) ,3),
+                     iq7525 = round(qdiff(x, .75, .25) ,3),
+                     iq9050 = round(qdiff(x, .9, .5) ,3),
+                     iq5010 = round(qdiff(x, .5, .1) ,3)
+                   ))
+  rownames(tab) <- paste(as.character(tab[,1]),as.character(tab[,2]), sep = ".")
+  tab[,c(1,2)] <- NULL
+  return(tab)
+}
+
+#######################################################################################################
+## ---- IQRDEC
+
+# data
+df.final2 <- scores2plot
+
+# ADD RESIDUALISED SCORES
+df.final$EXTres <- NA
+df.final$INTres <- NA
+
+df.final <- df.final2[FALSE,]
+for (g in c("M","F")) {
+  for (c in c("BCS", "MCS")) {
+    subs <- df.final2[df.final2$cohort==c & df.final2$sex==g,]
+    subs$EXTres <- residuals(lm(EXT ~ as.factor(incq) + ysch_moth + ysch_fath, data=subs, na.action=na.exclude))
+    subs$INTres <- residuals(lm(INT ~ as.factor(incq) + ysch_moth + ysch_fath, data=subs, na.action=na.exclude))
+    df.final <- rbind(df.final,subs)
+  }
+}
+# KEEP ONLY OBS WITH COMPLETE DATA
[…]

### 2018-03-27 codes/R/1_load.R
add harmonised social class, and graph

@@ -35,12 +35,58 @@ printfit <- function(m) {
 
 # auxiliary variables to retain
 auxvars <- c("faminc_real", "faminc_infl", "incq", "ysch_moth", "ysch_fath", 
-             "ageint5", "sex", "bwt", "smkpr", "pscl")
+             "ageint5", "sex", "bwt", "smkpr", "scl10")
 
 # BCS -------------------------------------------
 bcs5data <- read.dta(paste(dir_data, "bcs5yeng.dta", sep=""), convert.factors = F) # all BCS data
 bcs5rutb <- bcs5data[,grep("bcs5_rutb", names(bcs5data), value=TRUE)]                 # BINARY Rutter items only
 bcs5rutc <- bcs5data[,grep("bcs5_rutc", names(bcs5data), value=TRUE)]                 # 3CAT Rutter items only
+## ---- PREAMBLE
+rm(list=ls())         # Clear all objects from memory 
+set.seed(42)
+
+library(nFactors)
+library(polycor)
+library(random.polychor.pa)
+library(foreign)
+library(ggplot2)
+library(psych)
+library(mirt)
+library(cowplot)
+library(tikzDevice)
+library(lavaan)
+library(semTools)
+library(difR)
+library(np)
+library(gdata)
+library(knitr)
+library(xtable)
+library(cowplot)
+library(tikzDevice)
+library(gtools)
+
+
+dir_data  <- c("/Users/giacomomason/Documents/Projects/CohortStudies/rdata/")
+dir_syntax  <- c("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/syntax/")
+
+# define printfit function
+printfit <- function(m) {
+  c( fitMeasures(m, c("npar", "df", "rmsea", "mfi", "cfi.scaled")), moreFitIndices(m)["gammaHat"] )
+}
+
+## ---- LOAD_DATA
+
+# auxiliary variables to retain
+auxvars <- c("faminc_real", "faminc_infl", "incq", "ysch_moth", "ysch_fath", 
+             "ageint5", "sex", "bwt", "smkpr", "scl10")
+
+# BCS -------------------------------------------
+bcs5data <- read.dta(paste(dir_data, "bcs5yeng.dta", sep=""), convert.factors = F) # all BCS data
[…]

### 2018-03-27 codes/R/5_graphs.R

@@ -85,6 +85,46 @@ legend_b <- get_legend(ineqlist[[1]] + theme(legend.position="bottom"))
 p <- plot_grid( pcol, legend_b, ncol = 1, rel_heights = c(1, .1))
 p
 
+## ---- FACINEQ_SC
+# MEAN/CI plot of scores by social class
+
+scores2plot$scl10b <- NA
+scores2plot$scl10b[scores2plot$scl10 %in% c("I", "II")] <- 5
+scores2plot$scl10b[scores2plot$scl10 %in% c("IIINM")] <- 4
+scores2plot$scl10b[scores2plot$scl10 %in% c("IIIM")] <- 3
+scores2plot$scl10b[scores2plot$scl10 %in% c("IV", "V")] <- 2
+scores2plot$scl10b[scores2plot$scl10 %in% c("other")] <- 1
+scores2plot$scl10b <- factor(scores2plot$scl10b, labels = c("oth","IV-V","IIIM","IIINM","I-II"))
+
+#common options
+addopts <- function(x) {
+  x <- x + xlab("Parental Social Class at 10") + ylab("Factor score") +
+    coord_cartesian(ylim = c(-.4, .5)) +
+    stat_summary(geom="errorbar", fun.data=mean_cl_normal, width=.2, position=position_dodge(.5)) +
+    stat_summary(fun.y=mean, geom="point", size=4, aes(colour=cohort), position=position_dodge(.5)) +
+    labs(list(colour="")) + theme(legend.position="none")
+  return(x)
+}
+
+ineq.ext.m <- ggplot(data=subset(scores2plot, sex=="M" & !is.na(scores2plot$scl10b)), aes(x=as.factor(scl10b), y=EXT, colour=cohort)) + ggtitle("Males Externalising")
+ineq.ext.f <- ggplot(data=subset(scores2plot, sex=="F" & !is.na(scores2plot$scl10b)), aes(x=as.factor(scl10b), y=EXT, colour=cohort)) + ggtitle("Females Externalising")
+ineq.int.m <- ggplot(data=subset(scores2plot, sex=="M" & !is.na(scores2plot$scl10b)), aes(x=as.factor(scl10b), y=INT, colour=cohort)) + ggtitle("Males Internalising")
+ineq.int.f <- ggplot(data=subset(scores2plot, sex=="F" & !is.na(scores2plot$scl10b)), aes(x=as.factor(scl10b), y=INT, colour=cohort)) + ggtitle("Females Internalising")
+
[…]

### 2018-03-27 codes/R/backupcode.Rnw
old knitr dump

--- 2018-02-27 codes/R/cohorts_nc.Rnw
@@
-\documentclass[pdftex,10pt,twoside,a4paper]{article}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsfonts}
-\usepackage{amstext}% for resizing the symbol in math
-\usepackage{amsthm}
-\usepackage[utf8]{inputenc}
-\usepackage[english]{babel}
-\usepackage[babel]{csquotes}
-\usepackage{booktabs}
-\usepackage{tocloft}
-\usepackage{dsfont}
-\usepackage{bbm}
-\usepackage{xcolor}
-\usepackage{accents}
-\usepackage{setspace} % for spacing
-\onehalfspacing
-\usepackage{rotating}
-\usepackage{fullpage}
-\usepackage{pdflscape}  % for landscape page
-\usepackage{lmodern}  %font
-\usepackage{natbib}
-\usepackage{array}
-\usepackage{graphicx}
-\usepackage{adjustbox}
-\usepackage{tikz}
-\usepackage{bm}
-\usepackage{tabularx}
-\usepackage{marvosym}
-\usepackage{calc}
-\usepackage{array}
-\usepackage{verbatim}
-\usepackage{multirow}
-\usepackage[multiple]{footmisc}   % for multiple footnotes
-\usepackage{textgreek}            % for greek letters in text
-
-\setlength{\jot}{11pt} % increase vertical spacing in equations
-\usepackage{txfonts}
-\usepackage[T1]{fontenc}
-
-% define new columns
-\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\renewcommand{\arraystretch}{1.2}%
-
-% appendices and numeration
-\usepackage[nottoc,numbib]{tocbibind} 		% bibliography in TOC
-\usepackage[toc,page]{appendix} %appendix
-\newcommand{\tocless}[2]{\bgroup\let\addcontentsline=\nocontentsline#1{#2}\egroup} % for empty section
-\numberwithin{equation}{section} %for section eq numeration
-
-\usepackage{hyperref}                            % references
-\AtBeginDocument{%
-\renewcommand\sectionautorefname{Section}        % custom autoref names
-\renewcommand\subsectionautorefname{Section}
-}
-\newcommand{\aref}[1]{\hyperref[#1]{Appendix~\ref{#1}}} % for appendix sections
-\usepackage{cleveref}
-
-%tikz
-\usepackage{tikz}
-\usetikzlibrary{shapes,intersections}
-\usetikzlibrary{positioning,arrows, arrows.meta}
-
-%lists
-\usepackage[inline]{enumitem}
-\setlist{noitemsep}
-\setlist[enumerate]{labelsep=*, leftmargin=*}
-\setlist[itemize]{labelsep=*, leftmargin=*}
-\usepackage[textfont={small},labelfont={bf,small},justification=centering]{caption}     % to allow multiple lines in captions, and customize the look
-
-% page setup (from Racine)
-\setlength{\topmargin}{0cm}
-\setlength{\textheight}{24cm}
-\setlength{\oddsidemargin}{.0cm}
-\setlength{\evensidemargin}{.0cm}
-\setlength{\textwidth}{16.5cm}
-\setlength{\footskip}{1.5cm}
-\setlength{\footnotesep}{0.4cm}
-\linespread{1.3}
-
-% additional symbols
-\newcommand{\me}{\mathrm{e}}
-\newcommand{\veps}{\varepsilon}
-\newcommand{\E}{\mathbb{E}}
-\newcommand{\1}{\mathds{1}}
-\newcommand{\Cov}{\mathrm{Cov}}
-\newcommand{\Var}{\mathrm{Var}}
-\newcommand{\Corr}{\mathrm{Corr}}
-\newcommand{\vect}{\mathrm{vec}}
-\newcommand{\vecht}{\mathrm{vech}}
-\newcommand{\ubar}[1]{\underaccent{\bar}{#1}}
-\def\chk{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}
-
-% additional commands
-\newcommand{\bi}{\begin{itemize}}
-\newcommand{\ei}{\end{itemize}}
-\newcommand{\bnm}{\begin{enumerate}}
-\newcommand{\enm}{\end{enumerate}}
-
-\title{IFA Comparison of Noncognitive Skills \\ BCS and MCS}
-\author{Giacomo Mason \\ University College London}
-\date{\today}
-
-\begin{document}
-\maketitle
-
-<<MAIN,cache=FALSE,echo=FALSE,results='hide',message=FALSE,warning=FALSE>>=
-rm(list=ls())         # Clear all objects from memory 
-  opts_chunk$set(concordance=FALSE,
-                 echo=FALSE,
-                 tidy=FALSE,
-                 autodep=TRUE,
-                 cache=T, # REPLACE LATER
-                 message=FALSE, 
-                 warning=FALSE,
-                 size='scriptsize',
-                 results='hide',
-                 comment=NA, 
-                 fig.width=4,
-                 fig.height=4, 
-                 out.width='\\textwidth',
-                 fig.show='hold',
-                 tikzDefaultEngine='pdftex',
-                 tikzDocumentDeclaration = "\\documentclass[10pt]{article}",
-                 tikzMetricPackages = c("\\usepackage[utf8]{inputenc}","\\usepackage[T1]{fontenc}", "\\usetikzlibrary{calc}", "\\usepackage{amssymb}")
-                 )
-
-@
-
-<<READ, cache=FALSE >>=
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/1_load.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/2_efa.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/3_mifa_invariance.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/3_mifa_final.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/4_scores.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/4_scores_compare.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/5_graphs.R")
-
-@
-
-<<PREAMBLE, cache=F>>=
-@
-
-<<LOAD_DATA>>=
-@
-
-\section{MCS 5y vs BCS 5y Noncognitive Ability Items}
-
-Maternal reports on child behaviours were surveyed with two different instruments in the two cohorts.
-For BCS, the Rutter A Scale was used \citep{Rutter1970}.
-The mother is presented with a list of behaviours that a child might exhibit, and is asked for each whether it 'Does not apply', it 'Somewhat applies', or it 'Certainly applies' to her child.
-In the MCS cohort, mothers were administered the Strengths and Difficulties Questionnaire (SDQ, \citealp{Goodman1994,Goodman1997}).
-The SDQ was created as an updated version of the Rutter scales, expanding it to more recent advances in child psychopathology \cite{Stone2010}.
-Similarly to the Rutter scale, it asks the parent to rate the child on different behaviours, with the options of 'Not true', 'Somewhat true', 'Certainly true'.
-Summary scores obtained from the SDQ has been shown to correlate highly with the ones from Rutter scales, but the invariance properties of the scales are sensitive to the wording of the questions \citep{Goodman2007}.
-
-\subsection{BCS: Rutter scale}
-
-For BCS, I augment the 19-item Rutter Scale with 4 other parent-reported questions from the parental questionnaire. 
-\begin{enumerate}
-\item Very restless. Often running about or jumping up and down. Hardly ever still.
-\item Is squirmy or fidgety.
-\item Often destroys own or others’ belongings.
-\item Frequently fights other children.
-\item Not much liked by other children.
-\item Often worried, worries about many things
-\item Tends to do things on his/her own – rather solitary.
-\item Irritable. Is quick to fly off the handle.
-\item Often appears miserable, unhappy, tearful or distressed.
-\item Sometimes takes things belonging to others.
-\item Has twitches, mannerisms or tics of the face or body.
-\item Frequently sucks thumb or finger.
-\item Frequently bites nails or fingers.
-\item Is often disobedient.
-\item Cannot settle to anything for more than a few moments.
-\item Tends to be fearful or afraid of new things or new situations.
-\item Is over fussy or over particular.
-\item Often tells lies.
-\item Bullies other children.
-\item[A] Complains of headaches.
-\item[B] Complains of stomach-ache or has vomited.
-\item[C] Complais of biliousness.
-\item[D] Has temper tantrums (that is, complete loss of temper with shouting, angry movements, etc.).
-\end{enumerate}
-The Rutter items are rated on three levels: Does not apply, Somewhat applies, Certainly applies.
-Since they are all indicators of low skills, I encode all of them in reverse, i.e.
-``Certainly applies" = 0, ``Somewhat applies" = 1, ``Does not apply" = 2.
-The last 4 items are rated on 4 levels: Never in the last 12 months, less than once a month, at least once a month, at least once a week.
-I recode these into binary indicators, with \emph{Never} and \emph{Less than once a month}=1 and zero otherwise.
-
-To increase comparability between the two cohorts, I merge together two pairs of items: 4 and 19 (to mirror the SDQ question ``Often fights with other children or bullies them"), and also A and B (to mirror the SDQ question ``Often complains of head-aches, stomach-ache or sickness").
-I assign the lowest category among the two original items to the newly obtained item.
-I also recode items 5 and 14 to binary instead of three categories.
-These items are recorded with a positive phrasing in MCS, so a 3-category form would be harder to compare.
-
-\subsection{MCS: Strengths and Difficulties Questionnare}
-
-In MCS, I consider the 25-item strengths and difficulties questionnaire \citep{Goodman1997}.
-
-\textbf{Hyperactivity Scale}
-\begin{itemize}
-\item Restless, overactive, cannot stay still for long	2
-\item Constantly fidgeting or squirming	10
-\item Easily distracted, concentration wanders	15
-\item (+) Thinks things out before acting	21
-\item (+) Sees tasks through to the end, good attention span	25
-\end{itemize}
-\textbf{Emotional Symptoms Scale}
-\begin{itemize}
-\item Often complains of head- aches, stomach-ache or sickness	3
-\item Many worries, often seems worried	8
-\item Often unhappy, down-hearted or tearful	13
-\item Nervous or clingy in new situations, easily loses confidence	16
-\item Many fears, easily scared	24
-\end{itemize}
-\textbf{Conduct Problems Scale}	
-\begin{itemize}
-\item Often has temper tantrums or hot tempers	5
-\item (+) Generally obedient, usually does what adults request	7
-\item Often fights with other children or bullies them	12
-\item Often argumentative with adults		18
-\item Can be spiteful to others		22
-\end{itemize}
-\textbf{Peer Problems Scale}
-\begin{itemize}
-\item Rather solitary, tends to play alone	6
-\item (+) Has at least one good friend	11
-\item (+) Generally liked by other children	14
-\item Picked on or bullied by other children	19
-\item Gets on better with adults than with other children	23
-\end{itemize}
-\textbf{Prosocial Scale}
-\begin{itemize}
-\item (+) Considerate of other people's feelings	1
-\item (+) Shares readily with other children (treats, toys, pencils, etc.)	4
-\item (+) Helpful if someone is hurt, upset or feeling ill	9
-\item (+) Kind to younger children	17
-\item (+) Often volunteers to help others (parents, teachers, other children)	20
-\end{itemize}
-
-All items are recorded on a 4-point scale: Not true, Somewhat true, Certainly true, Can't say.
-I set the latter to missing and recode the rest in ascending order of skill as for the BCS items, i.e.
-``Certainly true" = 0, ``Somewhat true" = 1, ``Not true" = 2.
-For comparability with BCS, I retain items 3, 5, 7, and 14 in a binary format.
-
-\subsection{Exploratory Analysis}
-
-Our aim is to assess whether the two scales can be used to derive measures of noncognitive ability that are invariant, and can therefore be compared across cohorts.
-In the first step of our analysis, we consider a subset of each scale's items, selecting only those that present very similar wording across scales.
-The selection process results in a new 11-item scale, reported in \autoref{tab:comm} with the original wording and the prevalence of each option.
-As highlighted above, some of the items are coded differently in the two original scales and we thus turn them into binary dummies to ensure comparability.
-
-The second step to our invariance analysis seeks to justify a common latent factor structure across the two cohorts.
-The Rutter scale is constructed to measure two latent constructs: anti-social (externalising problems), and neurotic (internalising problems) behaviours.
-To validate these constructs in our new scale, we carry out exploratory factor analysis based on the matrix of pairwise tetrachoric correlations between the item responses.
-Various methods to assess the optimal number of factors underlying these items give inconsistent results, ranging from 1 (Velicer MAP) to 7 or even 8 (BIC).
-I choose to conduct analyses with two factors as suggested by the theory and by the Very Simple Structure method.
-
-For ease of interpretation, I choose to rescale the factor loadings using \emph{oblimin} rotation: this method allows the underlying factors to be correlated to each other.
-From the results of the exploratory FA and IRT analysis with 2 factors, the pattern that seems to emerge is consistent with the theoretical constructs of externalising (EXT) and internalising (INT) behaviour problems.
-
-<<BCSEFA>>=
-@
-
-<<BCSEFARES, dependson='BCSEFA', echo=T, results='markup'>>=
-@
-
-The five-factor structure at the basis of the full SDQ has been validated in many contexts \citep{Stone2010}; however, lower-dimensional structures have been also suggested \citep{Dickey2004}, and recent research has shown that there are some benefits to using broader subscales that correspond to the externalising and internalising factors in Rutter \citep{Goodman2010}.
-To assess whether the two-factor conceptualisation suggested by the theory is valid in our ``reduced" 11-item scale, we conduct exploratory factor analysis (EFA) on the new 11-item scale separately for each cohort sample.
-Different methodologies for quantifying the optimal number of factors produce inconclusive results, but both Velicer MAP and very simple structure (VSS) seem to indicate that a two-factor structure is suitable.
-An EFA model with two factors, employing the usual cutoff of .3 for the loadings, delivers a neat and sensible separation between items, as shown in the second column of \autoref{tab:comm}.
-Thus, our two-factor structure seems justifiable on both theoretical and empirical grounds.
-
-<<MCSEFA>>=
-@
-
-<<MCSEFARES, dependson='BCSEFA', echo=T, results='markup'>>=
-@
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\clearpage
-
-<<MEANTABLE>>=
-# make table with means
-@
-
-\clearpage
-\begin{landscape}
-\begin{table}[ht]
-\centering
-\footnotesize
-\caption{Item Commonalities}\label{tab:comm}
-<<MEANTABLE_RES, dependson="LOAD_DATA", results='asis'>>=
-require(xtable)
-
-table <- meantab
-
-print(xtable(table,
-                                  # item num        Factor          Num cats        BCS text        BCS num           BCS Certain     BCS Somewhat    BCS Applies             MCS text          MCS num
-                    align="L{0cm}C{.04\\textwidth}C{.05\\textwidth}C{.05\\textwidth}L{.3\\textwidth}C{.03\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}L{.3\\textwidth}C{.03\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
-                    digits = 1
-                    ), 
-             size="\\small",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = c(list(0)),
-                              command = c(
-                              # header
-                                    paste(" & & &
-                                              \\multicolumn{5}{c}{\\textbf{BCS}} &
-                                              \\multicolumn{5}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){4-8}  \\cmidrule(lr){9-13} 
-                                            Item & Factor & N. categ &
-                                              Text & & Cert. Applies & Somet. Applies & Applies &
-                                              Text & & Cert. True & Somet. True & True \\\\ \\midrule
-                                          "
-                                            , sep="")
-                                    )),
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabular", width = ".8\\textwidth",
-             floating = FALSE
-)
-
-@
-\end{table}
-\end{landscape}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Measurement Invariance}
-
-\subsubsection{Model notation}
-Before any meaningful comparison across cohorts can be made, measurement invariance (MI) needs to be established.
-Examination of the invariance properties of a measurement system is aimed at assessing whether the latent constructs in each group are measured in a comparable way, both in terms of scale and location.
-In the case of continuous measures, MI analysis proceeds in a relatively straightforward way \citep{Vanderberg2000a}; MI properties in systems of categorical outcomes are instead more problematic, due to the more complex nature of identification \citep{Millsap2004, Wu2016}.
-
-I set up a (confirmatory) item factor analysis (IFA) model for the full list of 11 common items.
-Given the results of the exploratory analysis -- where each measure separates neatly on one of the two factors -- I assume a \emph{dedicated} factor structure \citep{Carneiro2003}, where each measure is restricted to load on one of the two factors only.\footnote{This structure is commonly assumed in economic research to enable a more straightforward interpretation of the latent constructs in question \citep{ACFMR,Heckman2012a}.
-It can also be denominated \emph{congeneric} factor stucture \citep{Millsap2004}.}
-
-To account for the ordered nature of the measures, I specify a threshold model \citep{Muthen1984}.
-Individual $j=1\dots N_c$ in group $c=\{A,B\}$ is assumed to have a latent continuous propensity $X^*_{ijc}$ for each item $i=1,\dots,I$.
-In our application, $j$ indexes the child, the group $c$ corresponds to the cohort, and $i$ indexes each behaviour in the 11-item scale summarised in \autoref{tab:comm}.
-Thus, $X^*_{ijc}$ is the child's propensity to exhibit each of the Rutter/SDQ behaviours.
-This propensity is modelled as a function of item-specific intercepts $\nu$ and loadings $\lambda$, and the individual's level on a latent construct vector $\bm{\theta}$, plus an independent error component $u_{cij}$.
-In our application, the latent factor is bidimensional, and corresponds to the internalising ($\theta^{I}$) and externalising ($\theta^{E}$) traits.
-The dedicated factors structure implies that each item is assumed to only depend on a single latent dimensions, as delineated in the exploratory factor analysis in \autoref{tab:comm}.
-
-The propensity for each item can be written as follows:
-
-\begin{align}\label{eq:propens}
-\begin{split}
-X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{EXT}_{jc} + u_{ijc} \qquad \text{for} \; i=1,\dots, 6 \\
-X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{INT}_{jc} + u_{ijc} \qquad \text{for} \; i=7,\dots, 11
-\end{split}
-\end{align}
-
-or more compactly
-$$X^*_{jc} =\bm{\nu}_c + \bm{\Lambda}_{c}\bm{\theta}_{jc} + \bm{u}_{jc}$$
-
-with the additional assumptions
-$$\bm{\Theta}_{jc} \sim N(\bm{\kappa}_c, \bm{\Phi}_c) \qquad \text{and} \qquad \bm{u}_{jc} \sim N(\bm{0}, \bm{\Psi}_c)$$
-
-The model implies the following expression for the mean and covariance structure of the latent propensities:
-$$\bm{\mu}_c=\bm{\nu}_c + \bm{\Lambda}_c\bm{\kappa}_c \qquad \text{and} \qquad \Sigma_c = \bm{\Lambda}_c \bm{\Phi}_c \bm{\Lambda}_c^\prime + \bm{\Psi}_c.$$
-
-For each item $i$, we observe either a binary or ordinal (3-categories) manifestation $X_{cji}$.
-Given \eqref{eq:propens}, the observed outcome as a function of the propensities $X^*$ can be defined as a function of additional threshold parameters $\tau$, as follows:
-$$X_{ijc} = s \qquad \text{if} \; \tau_{s,ic} \leq X^*_{ijc} < \tau_{s+1,ic} \qquad \text{for} \; s=0,1,2$$
-with $\tau_{0,ic}=-\infty$ and $\tau_{3,ic}=+\infty$.
-The probability of observing each category for a binary item takes values:
-\begin{equation}
-X_{cji}=
-\begin{cases}
-\qquad 0 \qquad \text{if} \; X^*_{cji}<\tau_{1,ci} \\
-\qquad 1 \qquad \text{if} \; X^*_{cji}\geq\tau_{1,ci}
-\end{cases}
-\end{equation}
-
-An ordinal item takes values:\footnote{As a consequence of the normality assumption on the error terms $\veps$, the probability that individual $j$ in group $c$ exhibits item $i$ with ``intensity" $s$ is:
-\begin{align*}
-\Pr(X_{cji}=0) &= 1- \Phi(-\tau_{1,ci} + \nu_{ci} + \lambda_{ci}\theta^{EXT}_{j}) \\
-\Pr(X_{cji}=1) &= \Phi(-\tau_{1,ci} + \nu_{ci} + \lambda_{ci}\theta^{EXT}_{j})
-\end{align*}
-for binary items and
-\begin{align*}
-\Pr(X_{cji}=0) &= 1- \Phi(-\tau_{1,ci} + \nu_{ci} + \lambda_{ci}\theta^{EXT}_{j}) \\
-\Pr(X_{cji}=1) &= \Phi(-\tau_{1,ci} + \nu_{ci} + \lambda_{ci}\theta^{EXT}_{j}) - \Phi(-\tau_{2,ci} + \nu_{ci} + \lambda_{ci}\theta^{EXT}_{j})\\
-\Pr(X_{cji}=2) &= \Phi(-\tau_{2,ci} +  \nu_{ci} + \lambda_{ci}\theta^{EXT}_{j})
-\end{align*}
-for ordinal items. This IFA model can also be written as a 2-parameter Logit IRT model with item difficulty $b$ and item discrimination $a$. In the simplest case for a binary item $1$:
-$$\text{Probit}(X_{cj1}=1) = a_{c1}(\theta^{EXT}_j - b_{c1})$$
-such that $\tau_{c1}=a_{c1}b_{c1}$ and $a_{c1}=\lambda_{c1}$.}
-
-\begin{equation}
-X_{cji}=
-\begin{cases}
-\qquad 0 \qquad \text{if} \; X^*_{cji}<\tau_{1,ci} \\
-\qquad 1 \qquad \text{if} \; \tau_{1,ci} \leq X^*_{cji}<\tau_{2,ci} \\
-\qquad 2 \qquad \text{if} \; X^*_{cji}\geq\tau_{2,ci}
-\end{cases}
-\end{equation}
-
-\subsubsection{Invariance analysis}
-There are infinite ways in which the above system can be parameterised and a scale obtained -- the well-known issue of factor indeterminacy.
-A configural model is defined as a minimal set of parameter restrictions that ensure identification of the system across groups, without restricting the model further.
-Widely used parameterisations for the configural model are:
-\bnm[label=\Alph*:]
-\item Delta parameterisation {[WE{\textDelta}]} \citep{Wu2016}
-
-For all groups: $$\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}, \quad \bm{\nu}=\bm{0}, \quad \text{and} \; \text{diag}(\bm{\Sigma})=\bm{I}.$$
-
-\item Theta parameterisation {[WE{\textTheta}]} \citep{Wu2016}
-
-For all groups: $$\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}, \quad \bm{\nu}=\bm{0}, \quad \text{and} \; \text{diag}(\bm{\Psi})=\bm{I}.$$
-
-\item Anchored parameterisation {[MT]} \citep{Millsap2004}
-  \bi
-  \item For all groups, normalise a reference loading to 1 for each factor
-  \item Set invariant across groups one threshold per item (e.g. $\tau_{0,Ai} = \tau_{0,Bi}$), and an additional threshold in the reference items above
-  \item In the first group: $\bm{\kappa_A}=\bm{0}$, $\text{diag}(\bm{\Sigma}_A)=\bm{I}$
-  \item Set all intercepts $\bm{\nu}$ to zero
-  \ei
-\enm
-The first two parameterisations (WE{\textDelta} and WE{\textTheta}) work by normalising the mean and variance of factors to the same constants in both groups, and they leave all loadings and intercepts to be freely estimated.
-The MT parameterisation instead proceeds by first identifying one group, and then imposing cross-group equality constraints to identify other groups \citep{Wu2016}.
-
-To test measurement invariance, one needs to estimate nested models that place increasing restrictions on the item parameters across groups $c$.
-In the case of continuous measures, the usual approach is to restrict one set of parameters at a time.
-As an example, starting from the configural model, one would restrict all loadings to be equal across groups; if the fit of this restricted model is not worse than the configural model, \emph{metric invariance} is estabilished \citep{Vanderberg2000a}.
-The hierarchy of the nested models usually proceeds by testing loadings first, and then intercepts (metric and scalar invariance).
-Metric and scalar invariance are the minimum requirements to be able to meaningfully compare means and variances of the latent factors across groups.
-
-A substantially identical approach is proposed for categorical measures by \cite{Millsap2004}.
-However, new identification results in \cite{Wu2016} indicate that, in the case of categorical measures, invariance properties cannot be examined by simply restricting one set of parameters at a time.
-This is because the identification conditions used in the configural baseline model, while being minimally restrictive on their own, become restrictive once certain additional restrictions are imposed.
-In light of this, they propose sets of restrictions that identify structures of different invariance levels.
-They find that some restrictions cannot be tested alone against the configural model, because they give rise to statistically equivalent models.
-This is true of loading invariance, and also of threshold invariance in the case when the number of categories of each ordinal item is 3 or less.
-Furthermore, they suggest that comparison of both latent means and variances requires invariance in loadings, thresholds, and intercepts.
-
-Following the ordering in \cite{Wu2016}, I estimate the following increasingly restrictive models:
-\bi
-\item \textbf{Configural} model -- see WE{\textDelta} above
-\item \textbf{Threshold invariance} model
-  \bi
-  \item $\tau_{1,Ai} = \tau_{1,Bi}$ for $i=1,\dots,11$, $\tau_{2,Ai} = \tau_{2,Bi}$ for $i=\{1,2,3,4,7,8,9,10\}$
-  \item For all groups: $$\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}$$
-  \item For the reference group: $$\bm{\nu}_A=\bm{0}, \quad \; \text{diag}(\bm{\Sigma}_A)=\bm{I}$$
-  \item In our case, the number of categories in the outcomes is 3, thus this model is statistically equivalent to the configural model
-  \ei
-\item \textbf{Threshold and Loading invariance} model
-  \bi
-  \item $\tau_{1,Ai} = \tau_{1,Bi}$ for $i=1,\dots,11$, $\tau_{2,Ai} = \tau_{2,Bi}$ for $i=\{1,2,3,4,7,8,9,10\}$
-  \item $\lambda_{Ai} = \lambda_{Bi}$ for $i=1,\dots,11$
-  \item For all groups: $$\bm{\kappa}=\bm{0}$$
-  \item For the reference group: $$\bm{\nu}_A=\bm{0}, \quad \; \text{diag}(\bm{\Sigma}_A)=\bm{I}, \quad \text{diag}(\bm{\Phi}_A) = \bm{I}$$
-  \item This set of restriction allows comparison of latent factor variances
-  \ei
-\item \textbf{Threshold, Loading, and Intercept invariance} model
-  \bi
-  \item $\tau_{1,Ai} = \tau_{1,Bi}$ for $i=1,\dots,11$, $\tau_{2,Ai} = \tau_{2,Bi}$ for $i=\{1,2,3,4,7,8,9,10\}$
-  \item $\lambda_{Ai} = \lambda_{Bi}$ for $i=1,\dots,11$
-  \item For all groups: $$\bm{\nu}=\bm{0}$$
-  \item For the reference group: $$\bm{\kappa}_A=\bm{0}, \quad \; \text{diag}(\bm{\Sigma}_A)=\bm{I}, \quad \text{diag}(\bm{\Phi}_A) = \bm{I}$$
-  \item This set of restriction allows comparison of latent factor means and variances
-  \ei
-\ei
-
-I estimate each of the models listed above by Weighted Least Squares.\footnote{All estimates are computed using the \texttt{lavaan} package in \textbf{\textsf{R}} \citep{Rosseel2012}.
-Parameters of the model are estimated by mean- and variance adjusted weighted least squares (WLSMV), due to \cite{Mplus2017}.
-Point estimates for model parameters are obtained starting from the polychoric correlations, using diagonally weighted least squares (DWLS), and the full weight matrix is used to compute robust standard errors and test statistics.
-Robust WLS has proved in simulation studies to be moderately robust to small violations of the normality assumption in the latent underlying measures \citep{Flora2011}, and generally outperforms maximum likelihood in large samples \citep{Beauducel2009,Li2016}.} The fit of the increasingly restrictive models presented above is compared in \autoref{tab:fit}.
-For all models, a chi-squared difference would point to a lack of measurement invariance.
-However, this test is know to overestimate invariance in large samples.
-Alternative fit indices should thus be used when comparing nested models.
-\cite{Cheung2002} suggest relying on Comparative Fit Index (CFI, \citealp{Bentler1990}), McDonald Non-Centrality Index (MFI, \citealp{McDonald1989}), and Gamma Hat index \citep{Steiger1989} for the case of ordered measures.
-Commonly accepted thresholds for rejecting the restricted model are $\Delta CFI < - 0.01$, $\Delta MFI < - 0.02$, and $\Delta Gamma hat < - 0.001$.
-
-The overall fit of the configural model is satisfactory according to widely accepted rules of thumb \citep{Maccallum1996}.\footnote{{\color{red} BETTER CITATIONS FOR OVERALL FIT}.} As discussed above, when items have less than four categories, the model with restricted thresholds is statistically equivalent to the configural model.
-Restricting thresholds, loadings, and intercepts at the same time seems to provide a satisfying fit compared to the configural model.\footnote{\cite{Meade2008}, using the results from a simulation study, suggests stricter thresholds that should apply in a variety of conditions.
-For CFI, a single cutoff values of $.002$ is proposed, while cutoffs for MFI depend on the problem's characteristics; in our case (2 factors, 11 items), they suggest $.0066$.
-\cite{Sass2014} however cast some doubts of the generalisability of these cutoffs to WLSMV estimators.}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\clearpage
-\subsection{Inequality in noncognitive ability}
-The model can be used to investigate how inequality of noncognitive skills has changed between the two cohorts.
-
-\textbf{REVIEW} We can derive scores for $\theta^{EXT}$ and $\theta^{INT}$ directly, and plot them against socioeconomic measures observed in the data.
-Our main socioeconomic indicator is a measure of weekly total family net income in the prices of the interview year.\footnote{This measure is harmonised across cohorts and derived as part of CLOSER, which is funded by the ESRC.} Using the estimated parameters for the model with invariant thresholds and loadings (but non-invariant intercepts), I compute the scores for each individual in the sample (using the Empirical Bayes modal approach).
-In \autoref{fig:boxscores}, we can see that the gradient in noncognitive skills along family income quintiles is more pronounced in the MCS cohort.
-%In particular, a male five-year old at the top quintile of the income distribution in 1975 has a better score by \Sexpr{regcoeffs[1,3]} in the externalising dimension and \Sexpr{regcoeffs[1,4]} in the internalising dimension, compared to a child at the bottom quintile.
-In 2005, the differences amount to \Sexpr{regcoeffs[3,3]} and \Sexpr{regcoeffs[3,4]}, respectively.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
-
-List of estimated models:
-\bnm
-\item no gender split, no age adjustment (2 groups, BCS MCS)
-\item no gender split, age adjustment (2 groups, BCS MCS)
-\item males only, no age adjustment (2 groups, BCS MCS)
-\item females only, no age adjustment (2 groups, BCS MCS)
-\item males only, age adjustment (2 groups, BCS MCS)
-\item females only, age adjustment (2 groups, BCS MCS)
-\item separate gender groups, no age adjustment (4 groups, BCS.M BCS.F MCS.M MCS.F)
-\item separate gender groups, age adjustment (4 groups, BCS.M BCS.F MCS.M MCS.F)
-\item separate gender groups, overlapping ages, no age adjustment (4 groups, BCS.M BCS.F MCS.M MCS.F)
-\item separate gender groups, age adjustment constrained to be the same across ages (4 groups, BCS.M BCS.F MCS.M MCS.F)
-\item MIMIC model with age as regressor (4 groups, BCS.M BCS.F MCS.M MCS.F)
-\enm
-
-Notes: 
-\bi
-\item Only the models with age adjustment seem to have invariant intercepts (see Appendix \autoref{sec:app_fit}). 
-\item But the latent means of these models don't make sense with the "raw" scores we have (see Appendix \autoref{sec:app_meancov})
-\item Also, a model on the subset of children that have very similar ages delivers results that are very similar to the age-unadjusted models, and different from the age-adjusted ones.
-So the models with age adjustment seem to be doing something else than adjusting for age, and we don't understand what this something else is...
-\item The model with free intercepts (only restricting thresholds and loadings) delivers differences in scores by group, even if the means of the latent factors are constrained to be equal (see below)
-\item The models with free and constrained intercepts deliver very similar scores (.98 correlation), and they also give sensible and comparable gradients with income
-\ei
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-<<FA_INVARIANCE, dependson='LOAD_DATA'>>=
-# estimate models of different levels of invariance
-@
-
-<<FA_FIT, dependson='FA_INVARIANCE'>>=
-# assemble fit indices into matrices
-@
-
-<<FA_FINAL, dependson='FA_INVARIANCE'>>=
-# select final model set
-@
-
-<<FA_FINAL_PARS, dependson='FA_FINAL'>>=
-# get parameters of final model
-@
-
-<<CLEANUP, dependson='FA_FINAL'>>=
-@
-
-<<FA_SCORES, dependson='FA_FINAL'>>=
-# produce scores for different models
-@
-
-<<FA_SCORES_COMPARE, dependson='FA_SCORES'>>=
-# assemble and compare scores across different models
-@
-
-<<FA_SCORES_SELECT, dependson='FA_FINAL'>>=
-# select model to use for inequality analysis
-@
-
-\clearpage
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-<<FA_SCORES_COMPARE, dependson='FA_SCORES'>>=
-# assemble and compare scores across different models
-
-@
-
-When adjusting for age, mean scores do not agree with raw scores
-<<FA_SCORES_MEANS, dependson='FA_SCORES', echo=T, results='markup'>>=
-# means
-means.ext
-means.int
-@
-
-Scores that do not adjust for age are very highly correlated, regardless of whether they come from a model with free or restricted intercepts
-<<FA_SCORES_CORR, dependson='FA_SCORES', echo=T, results='markup'>>=
-cor(compare.scores[,c(grep("EXT", names(compare.scores), value=T))])
-cor(compare.scores[,c(grep("INT", names(compare.scores), value=T))])
-@
-
-The gradient of scores by income quintile are very similar
-<<FA_SCORES_INEQ, dependson='FA_SCORES', echo=T, results='markup'>>=
-
-# free intercepts
-lm(EXTtl ~ factor(incq), data=compare.scores)$coefficients
-lm(INTtl ~ factor(incq), data=compare.scores)$coefficients
-
-# constrained intercepts
-lm(EXTtli ~ factor(incq), data=compare.scores)$coefficients
-lm(INTtli ~ factor(incq), data=compare.scores)$coefficients
-
-# constrained intercepts, common ages
-lm(EXTtlir ~ factor(incq), data=compare.scores)$coefficients
-lm(INTtlir ~ factor(incq), data=compare.scores)$coefficients
-
-@
-
-<<FA_REGS, dependson='FA_SCORES'>>=
-
-# combinations of regressions
-regcoeffs <- matrix( 
-  c(rep("BCS", 2), rep("MCS", 2), rep(c("M","F"), 2), rep(NA, 8)), # the data elements 
-  nrow=4,              # number of rows 
-  ncol=4,              # number of columns 
-  byrow = F)        # fill matrix by columns 
-colnames(regcoeffs) <- c("cohort", "sex", "EXTd", "INTd")
-
-# fill in differences
-regs.ext <- list()
-regs.int <- list()
-for (r in 1:nrow(regcoeffs)) {
-  regs.ext[[r]] <- lm(EXT ~ as.factor(incq) + age, data=subset(scores2plot, cohort==regcoeffs[r,1] & sex==regcoeffs[r,2]))
-  regcoeffs[r,3] <- round(regs.ext[[r]]$coefficients["as.factor(incq)5"],2)
-  regs.int[[r]] <- lm(INT ~ as.factor(incq) + age, data=subset(scores2plot, cohort==regcoeffs[r,1] & sex==regcoeffs[r,2]))
-  regcoeffs[r,4] <- round(regs.int[[r]]$coefficients["as.factor(incq)5"],2)
-}
-
-@
-
-% \begin{table}[ht]
-% \caption{Estimated means of scores -- non-age adjusted 4-group model} \label{tab:scoremeans}
-% \centering
-% <<FA_COMP1, dependson='FA_SCORES', results='asis'>>=
-% # compare mean factor scores by gender and cohort
-% 
-% prn <- function(obj) sprintf("%.3f", round(obj,3))
-% toaggr <- c("EXT", "INT", "EXTr", "INTr", "EXT.RAW", "INT.RAW", "EXT.RAWr", "INT.RAWr")
-% sctable <- aggregate(items.scored[[7]][, toaggr], list(items.scored[[7]]$cohortsex), 
-%                    FUN = function (x) prn(mean(x)))
-% 
-% sctable <- cbind(group = factor(c(1,3,2,4)), sctable)
-% levels(sctable$group) <- c("BCS Males", "MCS Males", "BCS Females", "MCS Females")
-% sctable <- sctable[order(sctable$group),]
-% sctable <- sctable[, !names(sctable) %in% "Group.1"]
-% sctable <- as.matrix(sctable,3)
-% 
-% 
-% print.xtable(xtable(sctable,
-%                     align="L{0cm}L{2.2cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}",
-%                     digits = 3
-%                     ),
-%              size="\\small",
-%              include.rownames=F,
-%              include.colnames=F,
-%              booktabs = T,
-%              add.to.row = list(pos = c(list(0,4)),
-%                               command = c(
-%                               # header
-%                                    "\\toprule
-%                                     & \\multicolumn{4}{c}{\\textbf{Scored factors}}
-%                                     & \\multicolumn{4}{c}{\\textbf{Scales}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
-%                                     & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-adjusted} & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-Adjusted} \\\\ \\cmidrule(lr){2-3} \\cmidrule(lr){4-5} \\cmidrule(lr){6-7} \\cmidrule(lr){8-9}
-%                       & EXT & INT & EXT & INT & EXT & INT & EXT & INT \\\\",
-%                       "\\bottomrule"
-%                                     )),
-%              hline.after = NULL,  # remove default booktabs lines
-%              sanitize.text.function = function(x){x},
-%              sanitize.rownames.function = function(x){""}, # to not print row numbers
-%              tabular.environment = "tabularx", width = "\\textwidth",
-%              floating = FALSE
-% )
-% 
-% @
-% \end{table}
-% 
-% \begin{table}[ht]
-% \caption{Estimated means of scores -- age adjusted 4-group model} \label{tab:scoremeans}
-% \centering
-% <<FA_COMP2, dependson='FA_SCORES', results='asis'>>=
-% # compare mean factor scores by gender and cohort
-% 
-% prn <- function(obj) sprintf("%.3f", round(obj,3))
-% toaggr <- c("EXT", "INT", "EXTr", "INTr", "EXT.RAW", "INT.RAW", "EXT.RAWr", "INT.RAWr")
-% sctable <- aggregate(items.scored[[8]][, toaggr], list(items.scored[[8]]$cohortsex), 
-%                    FUN = function (x) prn(mean(x)))
-% 
-% sctable <- cbind(group = factor(c(1,3,2,4)), sctable)
-% levels(sctable$group) <- c("BCS Males", "MCS Males", "BCS Females", "MCS Females")
-% sctable <- sctable[order(sctable$group),]
-% sctable <- sctable[, !names(sctable) %in% "Group.1"]
-% sctable <- as.matrix(sctable,3)
-% 
-% 
-% print.xtable(xtable(sctable,
-%                     align="L{0cm}L{2.2cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}",
-%                     digits = 3
-%                     ),
-%              size="\\small",
-%              include.rownames=F,
-%              include.colnames=F,
-%              booktabs = T,
-%              add.to.row = list(pos = c(list(0,4)),
-%                               command = c(
-%                               # header
-%                                    "\\toprule
-%                                     & \\multicolumn{4}{c}{\\textbf{Scored factors}}
-%                                     & \\multicolumn{4}{c}{\\textbf{Scales}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
-%                                     & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-adjusted} & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-Adjusted} \\\\ \\cmidrule(lr){2-3} \\cmidrule(lr){4-5} \\cmidrule(lr){6-7} \\cmidrule(lr){8-9}
-%                       & EXT & INT & EXT & INT & EXT & INT & EXT & INT \\\\",
-%                       "\\bottomrule"
-%                                     )),
-%              hline.after = NULL,  # remove default booktabs lines
-%              sanitize.text.function = function(x){x},
-%              sanitize.rownames.function = function(x){""}, # to not print row numbers
-%              tabular.environment = "tabularx", width = "\\textwidth",
-%              floating = FALSE
-% )
-% 
-% @
-% \end{table}
-% 
-% \begin{table}[ht]
-% \caption{Estimated means of scores -- 4-group model, 59-62 months} \label{tab:scoremeans}
-% \centering
-% <<FA_COMP3, dependson='FA_SCORES', results='asis'>>=
-% # compare mean factor scores by gender and cohort
-% 
-% prn <- function(obj) sprintf("%.3f", round(obj,3))
-% toaggr <- c("EXT", "INT", "EXTr", "INTr", "EXT.RAW", "INT.RAW", "EXT.RAWr", "INT.RAWr")
-% sctable <- aggregate(items.scored[[9]][, toaggr], list(items.scored[[9]]$cohortsex), 
-%                    FUN = function (x) prn(mean(x)))
-% 
-% sctable <- cbind(group = factor(c(1,3,2,4)), sctable)
-% levels(sctable$group) <- c("BCS Males", "MCS Males", "BCS Females", "MCS Females")
-% sctable <- sctable[order(sctable$group),]
-% sctable <- sctable[, !names(sctable) %in% "Group.1"]
-% sctable <- as.matrix(sctable,3)
-% 
-% 
-% print.xtable(xtable(sctable,
-%                     align="L{0cm}L{2.2cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}",
-%                     digits = 3
-%                     ),
-%              size="\\small",
-%              include.rownames=F,
-%              include.colnames=F,
-%              booktabs = T,
-%              add.to.row = list(pos = c(list(0,4)),
-%                               command = c(
-%                               # header
-%                                    "\\toprule
-%                                     & \\multicolumn{4}{c}{\\textbf{Scored factors}}
-%                                     & \\multicolumn{4}{c}{\\textbf{Scales}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
-%                                     & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-adjusted} & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-Adjusted} \\\\ \\cmidrule(lr){2-3} \\cmidrule(lr){4-5} \\cmidrule(lr){6-7} \\cmidrule(lr){8-9}
-%                       & EXT & INT & EXT & INT & EXT & INT & EXT & INT \\\\",
-%                       "\\bottomrule"
-%                                     )),
-%              hline.after = NULL,  # remove default booktabs lines
-%              sanitize.text.function = function(x){x},
-%              sanitize.rownames.function = function(x){""}, # to not print row numbers
-%              tabular.environment = "tabularx", width = "\\textwidth",
-%              floating = FALSE
-% )
-% 
-% @
-% \end{table}
-
-% \clearpage
-% \begin{figure}
-% \begin{center}
-% <<AGEHIST, dev='tikz', fig.height=6, fig.width=6, out.width='.6\\linewidth', cache=TRUE>>=
-% @
-% \end{center}
-% \caption{Distribution of age in the BCS and MCS sample}
-% {\scriptsize \emph{Notes}: The figure shows the density of observations at each month of age for each of the two cohorts.  \par}
-% \end{figure}
-% 
-% \clearpage
-% \begin{figure}
-% <<FACDENS, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
-% @
-% \caption{Density of Scored Factors}
-% {\scriptsize \emph{Notes}: The figure shows estimated kernel densities of the scored factors separately by cohort.  \par}
-% \end{figure}
-% 
-% \begin{figure}
-% <<FACLOESS_AGE, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
-% @
-% \caption{Scores and child age}
-% {\scriptsize \emph{Notes}: The figure plots LOESS of the scored factors against child age. \par}
-% \end{figure}
-% 
-% 
-% \begin{figure}
-% <<FACRAW, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
-% @
-% \caption{Raw vs Scored Factors}
-% {\scriptsize \emph{Notes}: The figure plots the scored factors on the vertical axis against the ``raw" scores on the horizontal axis, i.e.
-scores obtained by simply summing the number of positive answers to each of the items.
-\par}
-% \end{figure}
-% 
-% \begin{figure}
-% <<FACSCAT, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
-% @
-% \caption{Scatterplots of factor scores by cohort}
-% {\scriptsize \emph{Notes}: The figure plots the scored factors separately by cohort. The externalising and internalising scores are plotted against one another. \par}
-% \end{figure}
-% 
-% \begin{figure}
-% <<FACINEQ, dev='tikz', fig.height=12, fig.width=10, out.width='1\\linewidth'>>=
-% @
-% \caption{Distribution of scores across income quintiles}\label{fig:boxscores}
-% {\scriptsize \emph{Notes}: The figure plots the distribution of the scored factors across the quintiles in the income distribution. \par}
-% \end{figure}
-% 
-% \begin{figure}
-% <<FACLOESS_INC, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
-% @
-% \caption{Scores and family income}
-% {\scriptsize \emph{Notes}: The figure plots LOESS of the scored factors against family income. \par}
-% \end{figure}
-
-\clearpage
-\bibliographystyle{chicago}
-\bibliography{/Users/giacomomason/Documents/BibTeX/Papers-cohorts_nc}
-
-\clearpage
-\begin{appendices}
-
-\section{Additional fit comparisons}\label{sec:app_fit}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Tables for additional fit indices
 
@@
 
-\clearpage
-\section{Latent variable mean structures}\label{sec:app_meancov}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Tables for estimated latent variable structures
 
@@
 
-\clearpage
-\section{Estimated model parameters}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Tables for estimated model parameters
 
@@
 
-\clearpage
-\section{Inequality}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Tables for estimated means of scores
+
+\begin{table}[ht]
+\caption{Estimated means of scores -- non-age adjusted 4-group model} \label{tab:scoremeans}
+\centering
+<<FA_COMP1, dependson='FA_SCORES', results='asis'>>=
+# compare mean factor scores by gender and cohort
+
+prn <- function(obj) sprintf("%.3f", round(obj,3))
+toaggr <- c("EXT", "INT", "EXTr", "INTr", "EXT.RAW", "INT.RAW", "EXT.RAWr", "INT.RAWr")
+sctable <- aggregate(items.scored[[7]][, toaggr], list(items.scored[[7]]$cohortsex),
+                   FUN = function (x) prn(mean(x)))
+
+sctable <- cbind(group = factor(c(1,3,2,4)), sctable)
+levels(sctable$group) <- c("BCS Males", "MCS Males", "BCS Females", "MCS Females")
+sctable <- sctable[order(sctable$group),]
+sctable <- sctable[, !names(sctable) %in% "Group.1"]
+sctable <- as.matrix(sctable,3)
+
+print.xtable(xtable(sctable,
+                    align="L{0cm}L{2.2cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}",
+                    digits = 3
+                    ),
+             size="\\small",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0,4)),
+                              command = c(
+                              # header
+                                   "\\toprule
+                                    & \\multicolumn{4}{c}{\\textbf{Scored factors}}
+                                    & \\multicolumn{4}{c}{\\textbf{Scales}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
+                                    & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-adjusted} & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-Adjusted} \\\\ \\cmidrule(lr){2-3} \\cmidrule(lr){4-5} \\cmidrule(lr){6-7} \\cmidrule(lr){8-9}
+                      & EXT & INT & EXT & INT & EXT & INT & EXT & INT \\\\",
+                      "\\bottomrule"
+                                    )),
+             hline.after = NULL,  # remove default booktabs lines
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabularx", width = "\\textwidth",
+             floating = FALSE
+)
+
+@
+\end{table}
+
+\begin{table}[ht]
+\caption{Estimated means of scores -- age adjusted 4-group model} \label{tab:scoremeans}
+\centering
+<<FA_COMP2, dependson='FA_SCORES', results='asis'>>=
+# compare mean factor scores by gender and cohort
+
+prn <- function(obj) sprintf("%.3f", round(obj,3))
+toaggr <- c("EXT", "INT", "EXTr", "INTr", "EXT.RAW", "INT.RAW", "EXT.RAWr", "INT.RAWr")
+sctable <- aggregate(items.scored[[8]][, toaggr], list(items.scored[[8]]$cohortsex),
+                   FUN = function (x) prn(mean(x)))
+
+sctable <- cbind(group = factor(c(1,3,2,4)), sctable)
+levels(sctable$group) <- c("BCS Males", "MCS Males", "BCS Females", "MCS Females")
+sctable <- sctable[order(sctable$group),]
+sctable <- sctable[, !names(sctable) %in% "Group.1"]
+sctable <- as.matrix(sctable,3)
+
+print.xtable(xtable(sctable,
+                    align="L{0cm}L{2.2cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}",
+                    digits = 3
+                    ),
+             size="\\small",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0,4)),
+                              command = c(
+                              # header
+                                   "\\toprule
+                                    & \\multicolumn{4}{c}{\\textbf{Scored factors}}
+                                    & \\multicolumn{4}{c}{\\textbf{Scales}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
+                                    & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-adjusted} & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-Adjusted} \\\\ \\cmidrule(lr){2-3} \\cmidrule(lr){4-5} \\cmidrule(lr){6-7} \\cmidrule(lr){8-9}
+                      & EXT & INT & EXT & INT & EXT & INT & EXT & INT \\\\",
+                      "\\bottomrule"
+                                    )),
+             hline.after = NULL,  # remove default booktabs lines
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabularx", width = "\\textwidth",
+             floating = FALSE
+)
+
+@
+\end{table}
+
+\begin{table}[ht]
+\caption{Estimated means of scores -- 4-group model, 59-62 months} \label{tab:scoremeans}
+\centering
+<<FA_COMP3, dependson='FA_SCORES', results='asis'>>=
+# compare mean factor scores by gender and cohort
+
+prn <- function(obj) sprintf("%.3f", round(obj,3))
+toaggr <- c("EXT", "INT", "EXTr", "INTr", "EXT.RAW", "INT.RAW", "EXT.RAWr", "INT.RAWr")
+sctable <- aggregate(items.scored[[9]][, toaggr], list(items.scored[[9]]$cohortsex),
+                   FUN = function (x) prn(mean(x)))
+
+sctable <- cbind(group = factor(c(1,3,2,4)), sctable)
+levels(sctable$group) <- c("BCS Males", "MCS Males", "BCS Females", "MCS Females")
+sctable <- sctable[order(sctable$group),]
+sctable <- sctable[, !names(sctable) %in% "Group.1"]
+sctable <- as.matrix(sctable,3)
+
+print.xtable(xtable(sctable,
+                    align="L{0cm}L{2.2cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}",
+                    digits = 3
+                    ),
+             size="\\small",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0,4)),
+                              command = c(
+                              # header
+                                   "\\toprule
+                                    & \\multicolumn{4}{c}{\\textbf{Scored factors}}
+                                    & \\multicolumn{4}{c}{\\textbf{Scales}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
+                                    & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-adjusted} & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-Adjusted} \\\\ \\cmidrule(lr){2-3} \\cmidrule(lr){4-5} \\cmidrule(lr){6-7} \\cmidrule(lr){8-9}
+                      & EXT & INT & EXT & INT & EXT & INT & EXT & INT \\\\",
+                      "\\bottomrule"
+                                    )),
+             hline.after = NULL,  # remove default booktabs lines
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabularx", width = "\\textwidth",
+             floating = FALSE
+)
+
+@
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Tables for inequality regressions
 
@@
 \end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Tables for returns to skills
 
@@
 
-\end{appendices}
-
-\end{document}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% FIGURES
+
+\clearpage
+\begin{figure}
+\begin{center}
+<<AGEHIST, dev='tikz', fig.height=6, fig.width=6, out.width='.6\\linewidth', cache=TRUE>>=
+@
+\end{center}
+\caption{Distribution of age in the BCS and MCS sample}
+{\scriptsize \emph{Notes}: The figure shows the density of observations at each month of age for each of the two cohorts.  \par}
+\end{figure}
+
+\clearpage
+\begin{figure}
+<<FACDENS, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
+@
+\caption{Density of Scored Factors}
+{\scriptsize \emph{Notes}: The figure shows estimated kernel densities of the scored factors separately by cohort.  \par}
+\end{figure}
+
+\begin{figure}
+<<FACLOESS_AGE, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
+@
+\caption{Scores and child age}
+{\scriptsize \emph{Notes}: The figure plots LOESS of the scored factors against child age. \par}
+\end{figure}
+
+\begin{figure}
+<<FACRAW, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
+@
+\caption{Raw vs Scored Factors}
+{\scriptsize \emph{Notes}: The figure plots the scored factors on the vertical axis against the ``raw" scores on the horizontal axis, i.e.
+scores obtained by simply summing the number of positive answers to each of the items.
+\par}
+\end{figure}
+
+\begin{figure}
+<<FACSCAT, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
+@
+\caption{Scatterplots of factor scores by cohort}
+{\scriptsize \emph{Notes}: The figure plots the scored factors separately by cohort. The externalising and internalising scores are plotted against one another. \par}
+\end{figure}
+
+\begin{figure}
+<<FACINEQ, dev='tikz', fig.height=12, fig.width=10, out.width='1\\linewidth'>>=
+@
+\caption{Distribution of scores across income quintiles}\label{fig:boxscores}
+{\scriptsize \emph{Notes}: The figure plots the distribution of the scored factors across the quintiles in the income distribution. \par}
+\end{figure}
+
+\begin{figure}
+<<FACLOESS_INC, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
+@
+\caption{Scores and family income}
+{\scriptsize \emph{Notes}: The figure plots LOESS of the scored factors against family income. \par}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+<<FA_REGS, dependson='FA_SCORES'>>=
+
+# combinations of regressions
+regcoeffs <- matrix(
+  c(rep("BCS", 2), rep("MCS", 2), rep(c("M","F"), 2), rep(NA, 8)), # the data elements
+  nrow=4,              # number of rows
+  ncol=4,              # number of columns
+  byrow = F)        # fill matrix by columns
+colnames(regcoeffs) <- c("cohort", "sex", "EXTd", "INTd")
+
+# fill in differences
+regs.ext <- list()
+regs.int <- list()
+for (r in 1:nrow(regcoeffs)) {
+  regs.ext[[r]] <- lm(EXT ~ as.factor(incq) + age, data=subset(scores2plot, cohort==regcoeffs[r,1] & sex==regcoeffs[r,2]))
+  regcoeffs[r,3] <- round(regs.ext[[r]]$coefficients["as.factor(incq)5"],2)
+  regs.int[[r]] <- lm(INT ~ as.factor(incq) + age, data=subset(scores2plot, cohort==regcoeffs[r,1] & sex==regcoeffs[r,2]))
+  regcoeffs[r,4] <- round(regs.int[[r]]$coefficients["as.factor(incq)5"],2)
+}
+
+@
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\clearpage
+\section{Robustness}
+
+List of estimated models:
+\bnm
+\item[1 MAIN] Separate gender groups, no age adjustment
+\item Separate gender groups, overlapping ages, no age adjustment
+\item no gender split, no age adjustment (2 groups, BCS MCS)
+\item males only, no age adjustment (2 groups, BCS MCS)
+\item females only, no age adjustment (2 groups, BCS MCS)
+\item no gender split, age adjustment (2 groups, BCS MCS)
+\item males only, age adjustment (2 groups, BCS MCS)
+\item females only, age adjustment (2 groups, BCS MCS)
+\item separate gender groups, age adjustment (4 groups, BCS.M BCS.F MCS.M MCS.F)
+\item separate gender groups, age adjustment constrained to be the same across ages (4 groups, BCS.M BCS.F MCS.M MCS.F)
+\item MIMIC model with age as regressor (4 groups, BCS.M BCS.F MCS.M MCS.F)
+\enm
+
+Notes:
+\bi
+\item Only the models with age adjustment seem to have invariant intercepts (see Appendix \autoref{sec:app_fit}).
+\item But the latent means of these models don't make sense with the "raw" scores we have (see Appendix \autoref{sec:app_meancov})
+\item Also, a model on the subset of children that have very similar ages delivers results that are very similar to the age-unadjusted models, and different from the age-adjusted ones.
+So the models with age adjustment seem to be doing something else than adjusting for age, and we don't understand what this something else is...
+\item The model with free intercepts (only restricting thresholds and loadings) delivers differences in scores by group, even if the means of the latent factors are constrained to be equal (see below)
+\item The models with free and constrained intercepts deliver very similar scores (.98 correlation), and they also give sensible and comparable gradients with income
+\ei
+
+<<FA_SCORES_COMPARE, dependson='FA_SCORES'>>=
+# assemble and compare scores across different models
+
+@
+
+When adjusting for age, mean scores do not agree with raw scores
+<<FA_SCORES_MEANS, dependson='FA_SCORES', echo=T, results='markup'>>=
+# means
+means.ext
+means.int
+@
+
+Scores that do not adjust for age are very highly correlated, regardless of whether they come from a model with free or restricted intercepts
+<<FA_SCORES_CORR, dependson='FA_SCORES', echo=T, results='markup'>>=
+cor(compare.scores[,c(grep("EXT", names(compare.scores), value=T))])
+cor(compare.scores[,c(grep("INT", names(compare.scores), value=T))])
+@
+
+The gradient of scores by income quintile are very similar
+<<FA_SCORES_INEQ, dependson='FA_SCORES', echo=T, results='markup'>>=
+
+# free intercepts
+lm(EXTtl ~ factor(incq), data=compare.scores)$coefficients
+lm(INTtl ~ factor(incq), data=compare.scores)$coefficients
+
+# constrained intercepts
+lm(EXTtli ~ factor(incq), data=compare.scores)$coefficients
+lm(INTtli ~ factor(incq), data=compare.scores)$coefficients
+
+# constrained intercepts, common ages
+lm(EXTtlir ~ factor(incq), data=compare.scores)$coefficients
+lm(INTtlir ~ factor(incq), data=compare.scores)$coefficients
+
+@

### 2018-03-28 codes/R/1_load.R
recode scores for plots

@@ -214,6 +214,15 @@ items.c$INT.RAW <- rowSums(apply(items.c[,paste("X", seq(7,11), sep="")], 2, fun
 items.c$EXT.RAWr <- residuals(lm(EXT.RAW ~ age, data=items.c, na.action = na.exclude))
 items.c$INT.RAWr <- residuals(lm(INT.RAW ~ age, data=items.c, na.action = na.exclude))
 
+# recode social class for plots
+items.c$scl10b <- NA
+items.c$scl10b[items.c$scl10 %in% c("I", "II")] <- 5
+items.c$scl10b[items.c$scl10 %in% c("IIINM")] <- 4
+items.c$scl10b[items.c$scl10 %in% c("IIIM")] <- 3
+items.c$scl10b[items.c$scl10 %in% c("IV", "V")] <- 2
+items.c$scl10b[items.c$scl10 %in% c("other")] <- 1
+items.c$scl10b <- factor(items.c$scl10b, labels = c("oth","IV-V","IIIM","IIINM","I-II"))
+
 # MODEL LIST
 # 1 (MAIN) : separate gender groups, no age adjustment (4 groups, BCS.M BCS.F MCS.M MCS.F)
 # 2 (AGECHECK): separate gender groups, overlapping ages, no age adjustment (4 groups, BCS.M BCS.F MCS.M MCS.F)

### 2018-03-28 codes/R/4_scores.R

@@ -5,4 +5,3 @@
 # score model selected as final
 scores.final <- predict(finalmod, newdata = items[[1]])
 scores2plot <- cbind(items[[1]], do.call(rbind, scores.final))
-

### 2018-03-28 codes/R/cohorts_nc.Rnw
graphs for CF decomposition

@@
                  autodep=TRUE,
-                 cache=T, # REPLACE LATER
+                 cache=TRUE,
                  message=FALSE, 
@@
 read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/2_efa.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/3_mifa_invariance.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/3_mifa_final.R")
+read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/3_inv_main.R")
+read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/3_inv_aux.R")
+read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/3_inv_fit.R")
 read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/4_scores.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/4_scores_compare.R")
+read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/4_scores_aux.R")
 read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/5_graphs.R")
-
-@
-
-<<PREAMBLE, cache=F>>=
+read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/6_ineq.R")
+read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/7_counterf.R")
+read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/7_counterf_graphs.R")
+
+@
+
+<<PREAMBLE, cache=FALSE >>=
 @
@@
 @
-
-<<BCSEFARES, dependson='BCSEFA', echo=T, results='markup'>>=
-@
+% 
+% <<BCSEFARES, dependson='BCSEFA', echo=T, results='markup'>>=
+% @
 
@@
 @
-
-<<MCSEFARES, dependson='BCSEFA', echo=T, results='markup'>>=
-@
+% 
+% # <<MCSEFARES, dependson='BCSEFA', echo=T, results='markup'>>=
+% # @
+%
 
@@
 \caption{Item Commonalities}\label{tab:comm}
-<<MEANTABLE_RES, dependson="LOAD_DATA", results='asis'>>=
+<<MEANTABLE_RES, dependson="MEANTABLE", results='asis'>>=
 require(xtable)
@@
 
-A substantially identical approach is proposed for categorical measures by \cite{Millsap2004}.
+A substantially similar approach is proposed for categorical measures by \cite{Millsap2004}.
 However, new identification results in \cite{Wu2016} indicate that, in the case of categorical measures, invariance properties cannot be examined by simply restricting one set of parameters at a time.
@@
 I estimate each of the models listed above by Weighted Least Squares.\footnote{All estimates are computed using the \texttt{lavaan} package in \textbf{\textsf{R}} \citep{Rosseel2012}.
-Parameters of the model are estimated by mean- and variance adjusted weighted least squares (WLSMV), due to \cite{Mplus2017}.
+Parameters of the model are estimated by mean- and variance adjusted weighted least squares (WLSMV) (see \citealp{Mplus2017}).
 Point estimates for model parameters are obtained starting from the polychoric correlations, using diagonally weighted least squares (DWLS), and the full weight matrix is used to compute robust standard errors and test statistics.
@@
 
-The overall fit of the configural model is satisfactory according to widely accepted rules of thumb \citep{Maccallum1996}.\footnote{{\color{red} BETTER CITATIONS FOR OVERALL FIT}.} As discussed above, when items have less than four categories, the model with restricted thresholds is statistically equivalent to the configural model.
-Restricting thresholds, loadings, and intercepts at the same time seems to provide a satisfying fit compared to the configural model.\footnote{\cite{Meade2008}, using the results from a simulation study, suggests stricter thresholds that should apply in a variety of conditions.
+The overall fit of the configural model is satisfactory according to widely accepted rules of thumb \citep{Maccallum1996}.\footnote{{\color{red} BETTER CITATIONS FOR OVERALL FIT}.} As discussed above, when items have less than four categories, the model with restricted thresholds is statistically equivalent to the configural model.\footnote{\cite{Meade2008}, using the results from a simulation study, suggests stricter thresholds that should apply in a variety of conditions.
 For CFI, a single cutoff values of $.002$ is proposed, while cutoffs for MFI depend on the problem's characteristics; in our case (2 factors, 11 items), they suggest $.0066$.
@@
 
+{\color{red} COMMENT FIT RESULTS}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\clearpage
-\subsection{Inequality in noncognitive ability}
-The model can be used to investigate how inequality of noncognitive skills has changed between the two cohorts.
-
-\textbf{REVIEW} We can derive scores for $\theta^{EXT}$ and $\theta^{INT}$ directly, and plot them against socioeconomic measures observed in the data.
-Our main socioeconomic indicator is a measure of weekly total family net income in the prices of the interview year.\footnote{This measure is harmonised across cohorts and derived as part of CLOSER, which is funded by the ESRC.} Using the estimated parameters for the model with invariant thresholds and loadings (but non-invariant intercepts), I compute the scores for each individual in the sample (using the Empirical Bayes modal approach).
-In \autoref{fig:boxscores}, we can see that the gradient in noncognitive skills along family income quintiles is more pronounced in the MCS cohort.
-%In particular, a male five-year old at the top quintile of the income distribution in 1975 has a better score by \Sexpr{regcoeffs[1,3]} in the externalising dimension and \Sexpr{regcoeffs[1,4]} in the internalising dimension, compared to a child at the bottom quintile.
-In 2005, the differences amount to \Sexpr{regcoeffs[3,3]} and \Sexpr{regcoeffs[3,4]}, respectively.
+
+<<FA_INV_MAIN, dependson="LOAD_DATA">>=
+# main models only
+@
+
+% % <<FA_INV_AUX, dependson='LOAD_DATA'>>=
+% % # robustness models
+% % @
+
+<<FA_FIT, dependson='FA_INV_MAIN'>>=
+# assemble fit indices into matrices
+@
+
+<<FA_SCORES_MAIN, dependson='FA_INV_MAIN'>>=
+# produce scores for main models
+@
+
+% % <<FA_SCORES_AUX, dependson='FA_INV_AUX'>>=
+% % # produce scores for robustness models
+% % @
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
-
-List of estimated models:
-\bnm
-\item no gender split, no age adjustment (2 groups, BCS MCS)
-\item no gender split, age adjustment (2 groups, BCS MCS)
-\item males only, no age adjustment (2 groups, BCS MCS)
-\item females only, no age adjustment (2 groups, BCS MCS)
-\item males only, age adjustment (2 groups, BCS MCS)
-\item females only, age adjustment (2 groups, BCS MCS)
-\item separate gender groups, no age adjustment (4 groups, BCS.M BCS.F MCS.M MCS.F)
-\item separate gender groups, age adjustment (4 groups, BCS.M BCS.F MCS.M MCS.F)
-\item separate gender groups, overlapping ages, no age adjustment (4 groups, BCS.M BCS.F MCS.M MCS.F)
-\item separate gender groups, age adjustment constrained to be the same across ages (4 groups, BCS.M BCS.F MCS.M MCS.F)
-\item MIMIC model with age as regressor (4 groups, BCS.M BCS.F MCS.M MCS.F)
-\enm
-
-Notes: 
-\bi
-\item Only the models with age adjustment seem to have invariant intercepts (see Appendix \autoref{sec:app_fit}). 
-\item But the latent means of these models don't make sense with the "raw" scores we have (see Appendix \autoref{sec:app_meancov})
-\item Also, a model on the subset of children that have very similar ages delivers results that are very similar to the age-unadjusted models, and different from the age-adjusted ones.
-So the models with age adjustment seem to be doing something else than adjusting for age, and we don't understand what this something else is...
-\item The model with free intercepts (only restricting thresholds and loadings) delivers differences in scores by group, even if the means of the latent factors are constrained to be equal (see below)
-\item The models with free and constrained intercepts deliver very similar scores (.98 correlation), and they also give sensible and comparable gradients with income
-\ei
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-<<FA_INVARIANCE, dependson='LOAD_DATA'>>=
-# estimate models of different levels of invariance
-@
-
-<<FA_FIT, dependson='FA_INVARIANCE'>>=
-# assemble fit indices into matrices
-@
-
-<<FA_FINAL, dependson='FA_INVARIANCE'>>=
-# select final model set
-@
-
-<<FA_FINAL_PARS, dependson='FA_FINAL'>>=
-# get parameters of final model
-@
-
-<<CLEANUP, dependson='FA_FINAL'>>=
-@
-
-<<FA_SCORES, dependson='FA_FINAL'>>=
-# produce scores for different models
-@
-
-<<FA_SCORES_COMPARE, dependson='FA_SCORES'>>=
-# assemble and compare scores across different models
-@
-
-<<FA_SCORES_SELECT, dependson='FA_FINAL'>>=
-# select model to use for inequality analysis
-@
-
-\clearpage
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-<<FA_SCORES_COMPARE, dependson='FA_SCORES'>>=
-# assemble and compare scores across different models
-
-@
-
-When adjusting for age, mean scores do not agree with raw scores
-<<FA_SCORES_MEANS, dependson='FA_SCORES', echo=T, results='markup'>>=
-# means
-means.ext
-means.int
-@
-
-Scores that do not adjust for age are very highly correlated, regardless of whether they come from a model with free or restricted intercepts
-<<FA_SCORES_CORR, dependson='FA_SCORES', echo=T, results='markup'>>=
-cor(compare.scores[,c(grep("EXT", names(compare.scores), value=T))])
-cor(compare.scores[,c(grep("INT", names(compare.scores), value=T))])
-@
-
-The gradient of scores by income quintile are very similar
-<<FA_SCORES_INEQ, dependson='FA_SCORES', echo=T, results='markup'>>=
-
-# free intercepts
-lm(EXTtl ~ factor(incq), data=compare.scores)$coefficients
-lm(INTtl ~ factor(incq), data=compare.scores)$coefficients
-
-# constrained intercepts
-lm(EXTtli ~ factor(incq), data=compare.scores)$coefficients
-lm(INTtli ~ factor(incq), data=compare.scores)$coefficients
-
-# constrained intercepts, common ages
-lm(EXTtlir ~ factor(incq), data=compare.scores)$coefficients
-lm(INTtlir ~ factor(incq), data=compare.scores)$coefficients
-
-@
-
-<<FA_REGS, dependson='FA_SCORES'>>=
-
-# combinations of regressions
-regcoeffs <- matrix( 
-  c(rep("BCS", 2), rep("MCS", 2), rep(c("M","F"), 2), rep(NA, 8)), # the data elements 
-  nrow=4,              # number of rows 
-  ncol=4,              # number of columns 
-  byrow = F)        # fill matrix by columns 
-colnames(regcoeffs) <- c("cohort", "sex", "EXTd", "INTd")
-
-# fill in differences
-regs.ext <- list()
-regs.int <- list()
-for (r in 1:nrow(regcoeffs)) {
-  regs.ext[[r]] <- lm(EXT ~ as.factor(incq) + age, data=subset(scores2plot, cohort==regcoeffs[r,1] & sex==regcoeffs[r,2]))
-  regcoeffs[r,3] <- round(regs.ext[[r]]$coefficients["as.factor(incq)5"],2)
-  regs.int[[r]] <- lm(INT ~ as.factor(incq) + age, data=subset(scores2plot, cohort==regcoeffs[r,1] & sex==regcoeffs[r,2]))
-  regcoeffs[r,4] <- round(regs.int[[r]]$coefficients["as.factor(incq)5"],2)
-}
-
-@
-
-% \begin{table}[ht]
-% \caption{Estimated means of scores -- non-age adjusted 4-group model} \label{tab:scoremeans}
-% \centering
-% <<FA_COMP1, dependson='FA_SCORES', results='asis'>>=
-% # compare mean factor scores by gender and cohort
-% 
-% prn <- function(obj) sprintf("%.3f", round(obj,3))
-% toaggr <- c("EXT", "INT", "EXTr", "INTr", "EXT.RAW", "INT.RAW", "EXT.RAWr", "INT.RAWr")
-% sctable <- aggregate(items.scored[[7]][, toaggr], list(items.scored[[7]]$cohortsex), 
-%                    FUN = function (x) prn(mean(x)))
-% 
-% sctable <- cbind(group = factor(c(1,3,2,4)), sctable)
-% levels(sctable$group) <- c("BCS Males", "MCS Males", "BCS Females", "MCS Females")
-% sctable <- sctable[order(sctable$group),]
-% sctable <- sctable[, !names(sctable) %in% "Group.1"]
-% sctable <- as.matrix(sctable,3)
-% 
-% 
-% print.xtable(xtable(sctable,
-%                     align="L{0cm}L{2.2cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}",
-%                     digits = 3
-%                     ),
-%              size="\\small",
-%              include.rownames=F,
-%              include.colnames=F,
-%              booktabs = T,
-%              add.to.row = list(pos = c(list(0,4)),
-%                               command = c(
-%                               # header
-%                                    "\\toprule
-%                                     & \\multicolumn{4}{c}{\\textbf{Scored factors}}
-%                                     & \\multicolumn{4}{c}{\\textbf{Scales}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
-%                                     & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-adjusted} & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-Adjusted} \\\\ \\cmidrule(lr){2-3} \\cmidrule(lr){4-5} \\cmidrule(lr){6-7} \\cmidrule(lr){8-9}
-%                       & EXT & INT & EXT & INT & EXT & INT & EXT & INT \\\\",
-%                       "\\bottomrule"
-%                                     )),
-%              hline.after = NULL,  # remove default booktabs lines
-%              sanitize.text.function = function(x){x},
-%              sanitize.rownames.function = function(x){""}, # to not print row numbers
-%              tabular.environment = "tabularx", width = "\\textwidth",
-%              floating = FALSE
-% )
-% 
-% @
-% \end{table}
-% 
-% \begin{table}[ht]
-% \caption{Estimated means of scores -- age adjusted 4-group model} \label{tab:scoremeans}
-% \centering
-% <<FA_COMP2, dependson='FA_SCORES', results='asis'>>=
-% # compare mean factor scores by gender and cohort
-% 
-% prn <- function(obj) sprintf("%.3f", round(obj,3))
-% toaggr <- c("EXT", "INT", "EXTr", "INTr", "EXT.RAW", "INT.RAW", "EXT.RAWr", "INT.RAWr")
-% sctable <- aggregate(items.scored[[8]][, toaggr], list(items.scored[[8]]$cohortsex), 
-%                    FUN = function (x) prn(mean(x)))
-% 
-% sctable <- cbind(group = factor(c(1,3,2,4)), sctable)
-% levels(sctable$group) <- c("BCS Males", "MCS Males", "BCS Females", "MCS Females")
-% sctable <- sctable[order(sctable$group),]
-% sctable <- sctable[, !names(sctable) %in% "Group.1"]
-% sctable <- as.matrix(sctable,3)
-% 
-% 
-% print.xtable(xtable(sctable,
-%                     align="L{0cm}L{2.2cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}",
-%                     digits = 3
-%                     ),
-%              size="\\small",
-%              include.rownames=F,
-%              include.colnames=F,
-%              booktabs = T,
-%              add.to.row = list(pos = c(list(0,4)),
-%                               command = c(
-%                               # header
-%                                    "\\toprule
-%                                     & \\multicolumn{4}{c}{\\textbf{Scored factors}}
-%                                     & \\multicolumn{4}{c}{\\textbf{Scales}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
-%                                     & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-adjusted} & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-Adjusted} \\\\ \\cmidrule(lr){2-3} \\cmidrule(lr){4-5} \\cmidrule(lr){6-7} \\cmidrule(lr){8-9}
-%                       & EXT & INT & EXT & INT & EXT & INT & EXT & INT \\\\",
-%                       "\\bottomrule"
-%                                     )),
-%              hline.after = NULL,  # remove default booktabs lines
-%              sanitize.text.function = function(x){x},
-%              sanitize.rownames.function = function(x){""}, # to not print row numbers
-%              tabular.environment = "tabularx", width = "\\textwidth",
-%              floating = FALSE
-% )
-% 
-% @
-% \end{table}
-% 
-% \begin{table}[ht]
-% \caption{Estimated means of scores -- 4-group model, 59-62 months} \label{tab:scoremeans}
-% \centering
-% <<FA_COMP3, dependson='FA_SCORES', results='asis'>>=
-% # compare mean factor scores by gender and cohort
-% 
-% prn <- function(obj) sprintf("%.3f", round(obj,3))
-% toaggr <- c("EXT", "INT", "EXTr", "INTr", "EXT.RAW", "INT.RAW", "EXT.RAWr", "INT.RAWr")
-% sctable <- aggregate(items.scored[[9]][, toaggr], list(items.scored[[9]]$cohortsex), 
-%                    FUN = function (x) prn(mean(x)))
-% 
-% sctable <- cbind(group = factor(c(1,3,2,4)), sctable)
-% levels(sctable$group) <- c("BCS Males", "MCS Males", "BCS Females", "MCS Females")
-% sctable <- sctable[order(sctable$group),]
-% sctable <- sctable[, !names(sctable) %in% "Group.1"]
-% sctable <- as.matrix(sctable,3)
-% 
-% 
-% print.xtable(xtable(sctable,
-%                     align="L{0cm}L{2.2cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}C{1.3cm}",
-%                     digits = 3
-%                     ),
-%              size="\\small",
-%              include.rownames=F,
-%              include.colnames=F,
-%              booktabs = T,
-%              add.to.row = list(pos = c(list(0,4)),
-%                               command = c(
-%                               # header
-%                                    "\\toprule
-%                                     & \\multicolumn{4}{c}{\\textbf{Scored factors}}
-%                                     & \\multicolumn{4}{c}{\\textbf{Scales}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
-%                                     & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-adjusted} & \\multicolumn{2}{c}{Normal} & \\multicolumn{2}{c}{Age-Adjusted} \\\\ \\cmidrule(lr){2-3} \\cmidrule(lr){4-5} \\cmidrule(lr){6-7} \\cmidrule(lr){8-9}
-%                       & EXT & INT & EXT & INT & EXT & INT & EXT & INT \\\\",
-%                       "\\bottomrule"
-%                                     )),
-%              hline.after = NULL,  # remove default booktabs lines
-%              sanitize.text.function = function(x){x},
-%              sanitize.rownames.function = function(x){""}, # to not print row numbers
-%              tabular.environment = "tabularx", width = "\\textwidth",
-%              floating = FALSE
-% )
-% 
-% @
-% \end{table}
-
-% \clearpage
-% \begin{figure}
-% \begin{center}
-% <<AGEHIST, dev='tikz', fig.height=6, fig.width=6, out.width='.6\\linewidth', cache=TRUE>>=
-% @
-% \end{center}
-% \caption{Distribution of age in the BCS and MCS sample}
-% {\scriptsize \emph{Notes}: The figure shows the density of observations at each month of age for each of the two cohorts.  \par}
-% \end{figure}
-% 
-% \clearpage
-% \begin{figure}
-% <<FACDENS, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
-% @
-% \caption{Density of Scored Factors}
-% {\scriptsize \emph{Notes}: The figure shows estimated kernel densities of the scored factors separately by cohort.  \par}
-% \end{figure}
-% 
-% \begin{figure}
-% <<FACLOESS_AGE, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
-% @
-% \caption{Scores and child age}
-% {\scriptsize \emph{Notes}: The figure plots LOESS of the scored factors against child age. \par}
-% \end{figure}
-% 
-% 
-% \begin{figure}
-% <<FACRAW, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
-% @
-% \caption{Raw vs Scored Factors}
-% {\scriptsize \emph{Notes}: The figure plots the scored factors on the vertical axis against the ``raw" scores on the horizontal axis, i.e.
-scores obtained by simply summing the number of positive answers to each of the items.
-\par}
-% \end{figure}
-% 
-% \begin{figure}
-% <<FACSCAT, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
-% @
-% \caption{Scatterplots of factor scores by cohort}
-% {\scriptsize \emph{Notes}: The figure plots the scored factors separately by cohort. The externalising and internalising scores are plotted against one another. \par}
-% \end{figure}
-% 
-% \begin{figure}
-% <<FACINEQ, dev='tikz', fig.height=12, fig.width=10, out.width='1\\linewidth'>>=
-% @
-% \caption{Distribution of scores across income quintiles}\label{fig:boxscores}
-% {\scriptsize \emph{Notes}: The figure plots the distribution of the scored factors across the quintiles in the income distribution. \par}
-% \end{figure}
-% 
-% \begin{figure}
-% <<FACLOESS_INC, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
-% @
-% \caption{Scores and family income}
-% {\scriptsize \emph{Notes}: The figure plots LOESS of the scored factors against family income. \par}
-% \end{figure}
-
-\clearpage
-\bibliographystyle{chicago}
-\bibliography{/Users/giacomomason/Documents/BibTeX/Papers-cohorts_nc}
-
-\clearpage
-\begin{appendices}
-
-\section{Additional fit comparisons}\label{sec:app_fit}
 
@@
 \small
-\caption{Fit indices -- 2-group models}\label{tab:fit2}
-<<FA_FIT_RES2, dependson="FA_FIT", results='asis'>>=
+\caption{Measurement invariance fit comparison}\label{tab:fit}
+<<FA_FIT_RES, dependson="FA_FIT", results='asis'>>=
 require(xtable)
 
-table <- rbind(afitab[[1]],afitab[[2]],afitab[[3]],afitab[[4]],afitab[[5]],afitab[[6]])
+table <- rbind(afitab[[1]],afitab[[2]])
 
@@
              booktabs = T,
-             add.to.row = list(pos = c(list(0,3,6,9,12,15)),
+             add.to.row = list(pos = c(list(0,3)),
                               command = c(
@@
             "Model & Num. params & $\\chi^2$ & RMSEA & MFI & CFI & G-hat & \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ \\midrule
-                       \\textbf{(1) No gender split, not age adjusted} \\\\",
-                      "\\newline \\textbf{(2) No gender split, age adjusted} \\\\ \\midrule ",
-                      "\\newline \\textbf{(3) Males only, not age adjusted} \\\\ \\midrule ",
-                      "\\newline \\textbf{(4) Females only, not age adjusted} \\\\ \\midrule ",
-                      "\\newline \\textbf{(5) Males only, age adjusted} \\\\ \\midrule ",
-                      "\\newline \\textbf{(6) Females only, age adjusted} \\\\ \\midrule "
+                       \\textbf{All ages} \\\\",
+                      "\\newline \\textbf{59-61 months sample} \\\\ \\midrule "
                                     )),
@@
 
+\clearpage
+\begin{figure}
+<<FACDENS, dependson="FA_SCORES_MAIN", dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
+@
+\caption{Density of Scored Factors}
+{\scriptsize \emph{Notes}: The figure shows estimated kernel densities of the scored factors separately by cohort.  \par}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\clearpage
+\subsection{Inequality in noncognitive ability}
+The model can be used to investigate how inequality of noncognitive skills has changed between the two cohorts.
+
+\textbf{REVIEW} We can derive scores for $\theta^{EXT}$ and $\theta^{INT}$ directly, and plot them against socioeconomic measures observed in the data.
+Our main socioeconomic indicator is a measure of weekly total family net income in the prices of the interview year.\footnote{This measure is harmonised across cohorts and derived as part of CLOSER, which is funded by the ESRC.} Using the estimated parameters for the model with invariant thresholds and loadings (but non-invariant intercepts), I compute the scores for each individual in the sample (using the Empirical Bayes modal approach).
+In \autoref{fig:meanci}, we can see that the gradient in noncognitive skills along family income quintiles is more pronounced in the MCS cohort.
+%In particular, a male five-year old at the top quintile of the income distribution in 1975 has a better score by \Sexpr{regcoeffs[1,3]} in the externalising dimension and \Sexpr{regcoeffs[1,4]} in the internalising dimension, compared to a child at the bottom quintile.
+In 2005, the differences amount to \Sexpr{regcoeffs[3,3]} and \Sexpr{regcoeffs[3,4]}, respectively.
+
+\begin{figure}
+<<FACINEQ, dependson="FA_SCORES_MAIN", dev='tikz', fig.height=12, fig.width=10, out.width='1\\linewidth'>>=
+@
+\caption{Mean scores across income quintiles}\label{fig:meanci}
+{\scriptsize \emph{Notes}: The figure plots the mean and its 95\% confidence interval across the quintiles in the income distribution. \par}
+\end{figure}
+
+\begin{figure}
+<<FACINEQ_SC, dependson="FA_SCORES_MAIN", dev='tikz', fig.height=12, fig.width=10, out.width='1\\linewidth'>>=
+@
+\caption{Mean scores across social class}\label{fig:meanci_sc}
+{\scriptsize \emph{Notes}: The figure plots the mean and its 95\% confidence interval across the quintiles in the income distribution. \par}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \clearpage
+% THIS BREAKS THE CACHING FOR SOME REASON
+% <<IQRDEC_FUN>>=
+% # load functions
+% @
+% 
+% <<IQRDEC, dependson="FA_SCORES_MAIN">>=
+% # decompose
+% @
+% 
+% % Compare inequality statistics of original vs residualised scores
+% % <<IQRDEC_RES, dependson='IQRDEC', echo=T, results='markup'>>=
+% % iqs.orig
+% % iqs.resid
+% % @
+
+\clearpage
+\subsection{Decomposition of inequality}
+
+We have populations $k \in \mathcal{K}$.
+For each population, we observe covariates $X_k$.
+For some $j \subseteq \mathcal{K}$, we observe outcome $Y_j$.
+We can identify the covariate distributions $F_{X_k}$ and the conditional distributions $F_{Y_j \mid X_j}$ and quantile functions $Q_{Y_j \mid X_j}$.
+Define the support of $X_k$ as $\mathcal{X}_k$ and the region of interest for the outcome as $\mathcal{Y}_j$.
+
+The interest is in the counterfactual distribution of the outcome:
+$$ F_{Y(j \mid k)}(y) = \int_{\mathcal{X}_k} F_{Y_j \mid X_j}(y \mid x) d F_{X_k}(x), \qquad y \in \mathcal{Y}_j $$
+and the corresponding quantiles:
+$$ Q_{Y(j \mid k)}(\tau) = F^\leftarrow_{Y(j \mid k)}(\tau) $$
+
+Given this, we can define counterfactual effects (CE):
+\bi
+\item Type 1 (Structure effect): CE of changing the conditional distribution: $Q_{Y(j \mid k)}(\tau) - Q_{Y(l \mid k)}(\tau)$
+\item Type 2 (Composition effect): CE of changing the covariate distribution: $Q_{Y(j \mid k)}(\tau) - Q_{Y(j \mid m)}(\tau)$
+\item Type 3: CE of changing both conditional and covariate distribution: $Q_{Y(j \mid k)}(\tau) - Q_{Y(l \mid m)}(\tau)$
+\ei
+
+In our application, we have $k=BCS,MCS$.
+We can thus decompose the observed differences in the distribution of non-cognitive skills into structure and composition effects.
+For example, the difference in the distribution of externalising skills can be written as:
+$$F_{(\theta^{EXT}_{MCS} \mid X_{MCS})} - F_{(\theta^{EXT}_{BCS} \mid X_{BCS})} =
+  \left[ F_{(\theta^{EXT}_{MCS} \mid X_{MCS})} - F_{(\theta^{EXT}_{BCS} \mid X_{MCS})} \right] +
+  \left[ F_{(\theta^{EXT}_{BCS} \mid X_{MCS})} - F_{(\theta^{EXT}_{BCS} \mid X_{BCS})} \right]$$
+
+<<COUNTERF, dependson="FA_SCORES_MAIN">>=
+# perform counterfactuals
+@
+
+<<COUNTERF_PREP_PLOTS, dependson='COUNTERF'>>=
+# prepare counterfactual plots
+@
+
+\begin{landscape}
+\begin{figure}
+<<COUNTERF_EXT, dependson='COUNTERF_PREP_PLOTS', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
+cfp.ext
+@
+\caption{Counterfactual decomposition -- externalising skills}\label{fig:counterf}
+{\scriptsize \emph{Notes}: \par}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+<<COUNTERF_INT, dependson='COUNTERF_PREP_PLOTS', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
+cfp.int
+@
+\caption{Counterfactual decomposition -- internalising skills}\label{fig:counterf}
+{\scriptsize \emph{Notes}: \par}
+\end{figure}
+\end{landscape}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\clearpage
+\bibliographystyle{chicago}
+\bibliography{/Users/giacomomason/Documents/BibTeX/Papers-cohorts_nc}
+%\bibliography{/Users/giacomomason/Dropbox/Zotero/cohorts.bib}
+
+\clearpage
+\begin{appendices}
+\section{Estimated model parameters}
+
+<<FUN_GETPARS>>=
+# load function that gets parameters
+@
+
 \begin{table}[ht]
-\centering
-\small
-\caption{Fit indices -- 4-group models}\label{tab:fit4}
-<<FA_FIT_RES4, dependson="FA_FIT", results='asis'>>=
-require(xtable)
-
-table <- rbind(afitab[[7]],afitab[[8]],afitab[[9]],afitab[[11]])
-
-print(xtable(table,
-                    align="L{0cm}L{.20\\textwidth}C{.03\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
-                    digits = 4
+\caption{Estimated loadings and thresholds} \label{tab:allpars}
+\resizebox{\textwidth}{!}{
+<<RES_FA_PAR, dependson="FA_FIT", results='asis'>>=
+source("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/getpars.R")
+
+table <- getmeaspars(finalmod, groups = 4, mode = "tl")
+
+# sequences to add rows
+rws <- seq(2, (nrow(table)-1), by = 1) # start from row 2 (exclude model titles)
+rwsl <- as.list(rws)
+
+print.xtable(xtable(table,
+                    # align = "ll|cccc|cccc"
+              align="C{0cm}C{1.2cm}C{1cm}C{1.3cm}C{1.1cm}C{1.1cm}C{1.1cm}C{1.1cm}C{1.1cm}C{1.1cm}C{1.1cm}C{1.1cm}",
+              digits = 3
                     ),
@@
              booktabs = T,
-             add.to.row = list(pos = c(list(0,3,6,9)),
+             add.to.row = list(pos = c(list(0)),
                               command = c(
                               # header
-            "Model & Num. params & $\\chi^2$ & RMSEA & MFI & CFI & G-hat & \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ \\midrule
-                      \\newline \\textbf{(7) not age adjusted} \\\\",
-                      "\\newline \\textbf{(8) age adjusted} \\\\\\midrule ",
-                      "\\newline \\textbf{(9) 59-62m range} \\\\ \\midrule ",
-                      "\\newline \\textbf{(11) MIMIC in age} \\\\ \\midrule "
+                                          " &
+                                            & \\textbf{Loadings}
+                                            & \\multicolumn{2}{c}{\\textbf{Thresholds}}
+                                            & \\multicolumn{3}{c}{\\textbf{Intercepts}}
+                                            & \\multicolumn{3}{c}{\\textbf{Variances}}
+                                            \\\\ \\cmidrule(lr){3-3} \\cmidrule(lr){4-5} \\cmidrule(lr){6-8} \\cmidrule(lr){9-11}
+
+                                            &
+                                            & $\\lambda$
+                                            & $\\tau_1$
+                                            & $\\tau_2$
+                                            & \\multicolumn{3}{c}{$\\nu$}
+                                            & \\multicolumn{3}{c}{$\\text{diag}(\\Psi)$} \\\\
+
+                                            Measure
+                                            & Factor
+                                            & All
+                                            & All
+                                            & All
+                                            & BCS F
+                                            & MCS M
+                                            & MCS F
+                                            & BCS F
+                                            & MCS M
+                                            & MCS F
+                                            \\\\ "
                       )),
@@
              sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabular", width = "\\textwidth",
+             tabular.environment = "tabularx", width = "\\textwidth",
              floating = FALSE
 )
-
-@
+@
+}
 \end{table}
 
-\clearpage
-\section{Latent variable mean structures}\label{sec:app_meancov}
-
 \begin{table}[ht]
-\caption{Estimated latent variable means and covariances -- 4 group models} \label{tab:lvpars4}
+\caption{Estimated latent variable means and covariances} \label{tab:lvpars}
 \centering
-<<RES_FA_LV4, dependson="FA_FINAL_PARS", results='asis'>>=
+<<RES_FA_LV, dependson="FA_FIT", results='asis'>>=
 require(xtable)
-
-table <- rbind(lvpars[[7]][[1]],lvpars[[7]][[2]],
-               lvpars[[8]][[1]],lvpars[[8]][[2]],
-               lvpars[[9]][[1]],lvpars[[9]][[2]]
-)
+source("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/getpars.R")
+
+lvtab <- getlvpars(finalmod, groups = 4)
+table <- rbind(lvtab$Males,lvtab$Females)
 
@@
              booktabs = T,
-             add.to.row = list(pos = c(list(0,2,4,6,8,10,12)),
+             add.to.row = list(pos = c(list(0,2,4)),
                               command = c(
@@
                                     & Mean & \\multicolumn{2}{c}{Covariance} & Corr. & Mean & \\multicolumn{2}{c}{Covariance} & Corr. \\\\ \\midrule
-                      \\textbf{(7) Males (joint, not age adj.)} \\\\ ",
-                      "\\newline \\textbf{(7) Females (joint, not age adj.)} \\\\ ",
-                      "\\midrule \\newline \\textbf{(8) Males (joint, age adj.)} \\\\ ",
-                      "\\newline \\textbf{(8) Females (joint, age adj.)} \\\\ ",
-                      "\\midrule \\newline \\textbf{(9) Males (joint, 59-62m)} \\\\ ",
-                      "\\newline \\textbf{(9) Females (joint, 59-62m)} \\\\ ",
+                      \\textbf{Males} \\\\",
+                      "\\textbf{Females} \\\\",
                       "\\bottomrule"
@@
 
-\begin{table}[ht]
-\caption{Estimated latent variable means and covariances -- 2 group models} \label{tab:lvpars2}
-\centering
-<<RES_FA_LV2, dependson="FA_FINAL_PARS", results='asis'>>=
-require(xtable)
-
-table <- rbind(lvpars[[1]],lvpars[[2]],
-               lvpars[[3]],lvpars[[4]],
-               lvpars[[4]],lvpars[[6]]
-)
-
-print.xtable(xtable(table,
-                    align="L{0cm}L{3cm}C{1.2cm}C{1.2cm}C{1.2cm}C{1.2cm}C{1.2cm}C{1.2cm}C{1.2cm}C{1.2cm}",
-                    digits = 3
-                    ),
-             size="\\small",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = c(list(0,2,4,6,8,10,12)),
-                              command = c(
-                              # header
-                                   "\\toprule
-                                    & \\multicolumn{4}{c}{\\textbf{BCS}}
-                                    & \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
-                                    & Mean & \\multicolumn{2}{c}{Covariance} & Corr. & Mean & \\multicolumn{2}{c}{Covariance} & Corr. \\\\ \\midrule
-                      \\textbf{(1) No gender split, no age adj.} \\\\",
-                      "\\textbf{(2) No gender split, age adj.} \\\\",
-                      "\\midrule \\newline \\textbf{(3) Males, no age adj.} \\\\ ",
-                      "\\newline \\textbf{(4) Females, no age adj.} \\\\ ",
-                      "\\midrule \\newline \\textbf{(5) Males, age adj.} \\\\ ",
-                      "\\newline \\textbf{(6) Females, age adj.} \\\\ ",
-                      "\\bottomrule"
-                                    )),
-             hline.after = NULL,  # remove default booktabs lines
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabularx", width = "\\textwidth",
-             floating = FALSE
-)
-
-@
-\end{table}
-
-\clearpage
-\section{Estimated model parameters}
-
-\begin{table}[ht]
-\caption{Estimated loadings and thresholds -- 2-group models with no age adjustment} \label{tab:allpars2a}
-\resizebox{\textwidth}{!}{
-<<RES_FA_PAR2A, dependson="FA_FINAL_PARS", results='asis'>>=
-require(xtable)
-
-table <- rbind(allpar[[1]],allpar[[3]],allpar[[4]])
-
-# sequences to add rows
-rws <- seq(2, (nrow(table)-1), by = 1) # start from row 2 (exclude model titles)
-rwsl <- as.list(rws)
-
-print.xtable(xtable(table,
-                    # align = "ll|cccc|cccc"
-                    align="L{0cm}L{2.5cm}C{2.5cm}C{2.5cm}C{2.5cm}C{2.5cm}C{1cm}",
-                    digits = 3
-                    ),
-             size="\\small",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = c(list(0,11,22)),
-                              command = c(
-                              # header
-                                          " & \\textbf{Loadings}
-                                            & \\multicolumn{2}{c}{\\textbf{Thresholds}}
-                                            & \\textbf{Variances}
-                                            & \\\\ \\cmidrule(lr){2-2} \\cmidrule(lr){3-4} \\cmidrule(lr){5-5}
-
-                                            & $\\bm{\\lambda}$
-                                            & $\\bm{\\tau}_1$
-                                            & $\\bm{\\tau}_2$
-                                            & $\\bm{\\Var(\\varepsilon)}$ \\\\
-                                            Measure & BCS = MCS
-                                            & BCS = MCS
-                                            & BCS = MCS
-                                            & MCS (BCS=1)
-                                            & Factor
-                                            \\\\ \\midrule
-                      \\newline \\textbf{(1) No gender split} \\\\",
-                      # additional separators
-                      "\\newline \\textbf{(3) Males only} \\\\ \\midrule ",
-                      "\\newline \\textbf{(4) Females only} \\\\ \\midrule "
-                      )),
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabularx", width = "\\textwidth",
-             floating = FALSE
-)
-
-@
-}
-\end{table}
-
-\begin{table}[ht]
-\caption{Estimated loadings and thresholds -- 2-group models with age adjustment} \label{tab:allpars2b}
-\resizebox{\textwidth}{!}{
-<<RES_FA_PAR2B, dependson="FA_FINAL_PARS", results='asis'>>=
-require(xtable)
-
-table <- rbind(allpar[[2]],allpar[[5]],allpar[[6]])
-
-# sequences to add rows
-rws <- seq(2, (nrow(table)-1), by = 1) # start from row 2 (exclude model titles)
-rwsl <- as.list(rws)
-
-print.xtable(xtable(table,
-                    # align = "ll|cccc|cccc"
-                    align="L{0cm}L{2.5cm}C{2.5cm}C{2.5cm}C{2.5cm}C{2.5cm}C{1cm}",
-                    digits = 3
-                    ),
-             size="\\small",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = c(list(0,11,22)),
-                              command = c(
-                              # header
-                                          " & \\textbf{Loadings}
-                                            & \\multicolumn{2}{c}{\\textbf{Thresholds}}
-                                            & \\textbf{Variances}
-                                            & \\\\ \\cmidrule(lr){2-2} \\cmidrule(lr){3-4} \\cmidrule(lr){5-5}
-
-                                            & $\\bm{\\lambda}$
-                                            & $\\bm{\\tau}_1$
-                                            & $\\bm{\\tau}_2$
-                                            & $\\bm{\\Var(\\varepsilon)}$ \\\\
-                                            Measure & BCS = MCS
-                                            & BCS = MCS
-                                            & BCS = MCS
-                                            & MCS (BCS=1)
-                                            & Factor
-                                            \\\\ \\midrule
-                      \\newline \\textbf{(2) No gender split} \\\\",
-                      # additional separators
-                      "\\newline \\textbf{(5) Males only} \\\\ \\midrule ",
-                      "\\newline \\textbf{(6) Females only} \\\\ \\midrule "
-                      )),
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabularx", width = "\\textwidth",
-             floating = FALSE
-)
-
-@
-}
-\end{table}
-
-\begin{table}[ht]
-\caption{Estimated loadings and thresholds -- 4-group models} \label{tab:allpars4}
-\resizebox{\textwidth}{!}{
-<<RES_FA_PAR4, dependson="FA_FINAL_PARS", results='asis'>>=
-require(xtable)
-
-table <- rbind(allpar[[7]], allpar[[8]], allpar[[9]])
-
-# sequences to add rows
-rws <- seq(2, (nrow(table)-1), by = 1) # start from row 2 (exclude model titles)
-rwsl <- as.list(rws)
-
-print.xtable(xtable(table,
-                    # align = "ll|cccc|cccc"
-                    align="L{0cm}L{2.2cm}C{1.8cm}C{1.4cm}C{1.4cm}C{1.7cm}C{1.7cm}C{1.7cm}C{1cm}",
-                    digits = 3
-                    ),
-             size="\\small",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = c(list(0,11,22)),
-                              command = c(
-                              # header
-                                          " & \\textbf{Loadings}
-                                            & \\multicolumn{2}{c}{\\textbf{Thresholds}}
-                                            & \\multicolumn{3}{c}{\\textbf{Variances}}
-                                            & \\\\ \\cmidrule(lr){2-2} \\cmidrule(lr){3-4} \\cmidrule(lr){5-7}
-
-                                            & $\\lambda$
-                                            & $\\tau_1$
-                                            & $\\tau_2$
-                                            & \\multicolumn{3}{c}{$\\text{diag}(\\Psi)$} \\\\
-
-                                            Measure
-                                            & All
-                                            & All
-                                            & All
-                                            & BCS M
-                                            & MCS F
-                                            & MCS M
-                                            & Factor
-                                            \\\\ \\midrule 
-                                          \\newline \\textbf{(7) not age adjusted} \\\\"
-                                          ,
-                                          "\\newline \\textbf{(8) age adjusted} \\\\ \\midrule ",
-                                          "\\newline \\textbf{(9) 59-62m} \\\\ \\midrule "
-
-                      )),
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabularx", width = "\\textwidth",
-             floating = FALSE
-)
-
-@
-}
-\end{table}
-
-\clearpage
-\section{Inequality}
-
-\begin{table}[ht]
-\caption{Regression of scores on income} \label{tab:regs}
-\centering
-\resizebox{\linewidth}{!}{%
-<<FA_REGS2, dependson='FA_SCORES', results='asis'>>=
-# regress scores on income
-library(dplyr)
-library(stargazer)
-
-scores2plot$incq_infl <- factor(ntile(scores2plot$faminc_infl, 5))
-
-r.ext.f.c <- lm(EXT ~ faminc_infl*cohort, data=subset(scores2plot, sex=="F"))
-r.ext.f.q <- lm(EXT ~ incq_infl*cohort, data=subset(scores2plot, sex=="F"))
-r.int.f.c <- lm(INT ~ faminc_infl*cohort, data=subset(scores2plot, sex=="F"))
-r.int.f.q <- lm(INT ~ incq_infl*cohort, data=subset(scores2plot, sex=="F"))
-
-r.ext.m.c <- lm(EXT ~ faminc_infl*cohort, data=subset(scores2plot, sex=="M"))
-r.ext.m.q <- lm(EXT ~ incq_infl*cohort, data=subset(scores2plot, sex=="M"))
-r.int.m.c <- lm(INT ~ faminc_infl*cohort, data=subset(scores2plot, sex=="M"))
-r.int.m.q <- lm(INT ~ incq_infl*cohort, data=subset(scores2plot, sex=="M"))
-
-stargazer(r.ext.f.c, r.ext.f.q, r.int.f.c, r.int.f.q, r.ext.m.c, r.ext.m.q, r.int.m.c, r.int.m.q,
-          column.labels=c("Females","Males"), column.separate=c(4,4),
-          dep.var.caption = "", multicolumn = FALSE,
-          omit=c("Constant"),
-          covariate.labels=c("Family income (2015 CPI £)",
-                             "Family income Q2 (2015 CPI £)",
-                             "Family income Q3 (2015 CPI £)",
-                             "Family income Q4 (2015 CPI £)",
-                             "Family income Q5 (2015 CPI £)",
-                             "MCS dummy",
-                             "Family income (2015 CPI £) x MCS",
-                             "Family income Q2 (2015 CPI £) x MCS",
-                             "Family income Q3 (2015 CPI £) x MCS",
-                             "Family income Q4 (2015 CPI £) x MCS",
-                             "Family income Q5 (2015 CPI £) x MCS"
-                             ),
-          no.space=TRUE, 
-          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
-          float=FALSE    # needs begin table outside it
-)
-@
-}
-\end{table}
-
-\begin{table}[ht]
-\caption{Returns to EXT/INT in BCS} \label{tab:returns}
-\centering
-\resizebox{\linewidth}{!}{%
-<<FA_RETURNS, dependson='FA_SCORES', results='asis'>>=
-# regress scores on income
-library(dplyr)
-library(stargazer)
-
-# get BCS data and make dummy for higher education
-scoresBCS <- scores2plot[scores2plot$cohort=="BCS",]
-scoresBCS$highered <- as.numeric(scoresBCS$hinvq00>3)
-scoresBCS[is.na(scoresBCS$hinvq00), "highered"] <- NA
-
-# females
-r.nvq.f <- list()
-r.nvq.f[[1]] <- lm(highered ~ EXT + INT, data=subset(scoresBCS, sex=="F"))
-r.nvq.f[[2]] <- lm(highered ~ EXT + INT + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="F"))
-r.nvq.f[[3]] <- lm(highered ~ EXT + INT + factor(incq), data=subset(scoresBCS, sex=="F"))
-r.nvq.f[[4]] <- lm(highered ~ EXT + INT + ysch_moth + ysch_fath, data=subset(scoresBCS, sex=="F"))
-r.nvq.f[[5]] <- lm(highered ~ EXT + INT + factor(incq) + ysch_moth + ysch_fath + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="F"))
-
-# females
-r.nvq.m <- list()
-r.nvq.m[[1]] <- lm(highered ~ EXT + INT, data=subset(scoresBCS, sex=="M"))
-r.nvq.m[[2]] <- lm(highered ~ EXT + INT + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="M"))
-r.nvq.m[[3]] <- lm(highered ~ EXT + INT + factor(incq), data=subset(scoresBCS, sex=="M"))
-r.nvq.m[[4]] <- lm(highered ~ EXT + INT + ysch_moth + ysch_fath, data=subset(scoresBCS, sex=="M"))
-r.nvq.m[[5]] <- lm(highered ~ EXT + INT + factor(incq) + ysch_moth + ysch_fath + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="M"))
-
-stargazer(r.nvq.f[[1]], r.nvq.f[[2]], r.nvq.f[[3]], r.nvq.f[[4]], r.nvq.f[[5]],
-          r.nvq.m[[1]], r.nvq.m[[2]], r.nvq.m[[3]], r.nvq.m[[4]], r.nvq.m[[5]],
-          dep.var.caption = "Completed Higher Ed. at 30",
-          column.labels=c("Females","Males"), column.separate=c(5,5),
-          omit=c("Constant"),
-          covariate.labels=c("Externalising skill at 5",
-                             "Internalising skill at 5",
-                             "EPVT Z at 5",
-                             "Human figure drawing Z at 5",
-                             "Copy design Z at 5",
-                             "Family inc. at 10 Q2",
-                             "Family inc. at 10 Q3",
-                             "Family inc. at 10 Q4",
-                             "Family inc. at 10 Q5",
-                             "Mother's years of schooling",
-                             "Father's years of schooling"
-                             ),
-          no.space=TRUE, 
-          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
-          float=FALSE    # needs begin table outside it
-)
-@
-}
-\end{table}
-
-\begin{table}[ht]
-\caption{Returns to EXT/INT in BCS -- raw scores} \label{tab:returns_raw}
-\centering
-\resizebox{\linewidth}{!}{%
-<<FA_RETURNS_RAW, dependson='FA_SCORES', results='asis'>>=
-# regress scores on income
-library(dplyr)
-library(stargazer)
-
-# get BCS data and make dummy for higher education
-scoresBCS <- scores2plot[scores2plot$cohort=="BCS",]
-scoresBCS$highered <- as.numeric(scoresBCS$hinvq00>3)
-scoresBCS[is.na(scoresBCS$hinvq00), "highered"] <- NA
-
-# females
-r.nvq.f <- list()
-r.nvq.f[[1]] <- lm(highered ~ EXT.RAW + INT.RAW, data=subset(scoresBCS, sex=="F"))
-r.nvq.f[[2]] <- lm(highered ~ EXT.RAW + INT.RAW + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="F"))
-r.nvq.f[[3]] <- lm(highered ~ EXT.RAW + INT.RAW + factor(incq), data=subset(scoresBCS, sex=="F"))
-r.nvq.f[[4]] <- lm(highered ~ EXT.RAW + INT.RAW + ysch_moth + ysch_fath, data=subset(scoresBCS, sex=="F"))
-r.nvq.f[[5]] <- lm(highered ~ EXT.RAW + INT.RAW + factor(incq) + ysch_moth + ysch_fath + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="F"))
-
-# females
-r.nvq.m <- list()
-r.nvq.m[[1]] <- lm(highered ~ EXT.RAW + INT.RAW, data=subset(scoresBCS, sex=="M"))
-r.nvq.m[[2]] <- lm(highered ~ EXT.RAW + INT.RAW + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="M"))
-r.nvq.m[[3]] <- lm(highered ~ EXT.RAW + INT.RAW + factor(incq), data=subset(scoresBCS, sex=="M"))
-r.nvq.m[[4]] <- lm(highered ~ EXT.RAW + INT.RAW + ysch_moth + ysch_fath, data=subset(scoresBCS, sex=="M"))
-r.nvq.m[[5]] <- lm(highered ~ EXT.RAW + INT.RAW + factor(incq) + ysch_moth + ysch_fath + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="M"))
-
-stargazer(r.nvq.f[[1]], r.nvq.f[[2]], r.nvq.f[[3]], r.nvq.f[[4]], r.nvq.f[[5]],
-          r.nvq.m[[1]], r.nvq.m[[2]], r.nvq.m[[3]], r.nvq.m[[4]], r.nvq.m[[5]],
-          dep.var.caption = "Completed Higher Ed. at 30",
-          column.labels=c("Females","Males"), column.separate=c(5,5),
-          omit=c("Constant"),
-          covariate.labels=c("Externalising skill at 5 (raw score)",
-                             "Internalising skill at 5 (raw score)",
-                             "EPVT Z at 5",
-                             "Human figure drawing Z at 5",
-                             "Copy design Z at 5",
-                             "Family inc. at 10 Q2",
-                             "Family inc. at 10 Q3",
-                             "Family inc. at 10 Q4",
-                             "Family inc. at 10 Q5",
-                             "Mother's years of schooling",
-                             "Father's years of schooling"
-                             ),
-          no.space=TRUE, 
-          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
-          float=FALSE    # needs begin table outside it
-)
-@
-}
-\end{table}
-
 \end{appendices}

### 2018-03-28 codes/R/5_graphs.R

@@ -88,14 +88,6 @@ p
 ## ---- FACINEQ_SC
 # MEAN/CI plot of scores by social class
 
-scores2plot$scl10b <- NA
-scores2plot$scl10b[scores2plot$scl10 %in% c("I", "II")] <- 5
-scores2plot$scl10b[scores2plot$scl10 %in% c("IIINM")] <- 4
-scores2plot$scl10b[scores2plot$scl10 %in% c("IIIM")] <- 3
-scores2plot$scl10b[scores2plot$scl10 %in% c("IV", "V")] <- 2
-scores2plot$scl10b[scores2plot$scl10 %in% c("other")] <- 1
-scores2plot$scl10b <- factor(scores2plot$scl10b, labels = c("oth","IV-V","IIIM","IIINM","I-II"))
-
 #common options
 addopts <- function(x) {
   x <- x + xlab("Parental Social Class at 10") + ylab("Factor score") +

### 2018-03-28 codes/R/7_counterf.R

@@ -1,5 +1,6 @@
 #######################################################################################################
 ## ---- COUNTERF
+library(Counterfactual)
 
 # numeric cohort
 scores2plot$cohn <- as.numeric(scores2plot$cohort)-1

### 2018-03-28 codes/R/7_counterf_graphs.R

@@ -1,5 +1,5 @@
 #######################################################################################################
-## ---- COUNTERF_PLOT
+## ---- COUNTERF_PREP_PLOTS
 
 # function to get data table out of counterfactual result
 getcfdat <- function(cffit) {
@@ -13,50 +13,89 @@ getcfdat <- function(cffit) {
     com.u = cffit$resCE[,4],
     tot = cffit$resTE[,1],
     tot.l = cffit$resTE[,3],
-    tot.u = cffit$resTE[,4]
+    tot.u = cffit$resTE[,4],
+    qf.cf = cffit$model_quantile_counter[,1],
+    qf.ref1 = cffit$model_quantile_ref1[,1],
+    qf.ref0 = cffit$model_quantile_ref0[,1]
   ))
   return(plotdata)
 }
 
-cfpdata.ext.m <- getcfdat(logitcf.ext.m)
-cfpdata.ext.f <- getcfdat(logitcf.ext.f)
-
-# common options
-addopts <- function(x) {
-  x <- x + ylab("Quantile Effect") +
-    scale_x_continuous(name = "", breaks = seq(.1,.9,.1)) +
-    coord_cartesian(ylim = c(-.4, .5)) +
-    labs(list(colour="")) + theme(legend.position="none")
-  return(x)
+# function to make plots
+makecfplots <- function(cffit) {
+  
+  # make data
+  cfdata <- getcfdat(cffit)
+  
+  # common options
+  addopts.all <- function(x) {
+    x <- x +
+      theme(
+        axis.title.y=element_blank(),
+        plot.title = element_text(hjust = 1, face = "italic")
+        ) +
+      scale_x_continuous(name = "Quantile", breaks = seq(.1,.9,.1))
+    return(x)
+  }
+  # options for effects only
+  addopts.ce <- function(x) {
+    x <- x +
+      coord_cartesian(ylim = c(-.4, .5)) +
+      labs(list(colour="")) + theme(legend.position="none")
+    return(x)
+  }
+  
+  # plots
+  p.qfs <- ggplot(cfdata, aes(quant))+ ggtitle("Quantile functions") +
+    geom_line(aes(y=qf.ref0, colour = "BCS")) + 
+    geom_line(aes(y=qf.ref1, colour = "MCS")) +
[…]

### 2018-03-28 codes/R/cohorts_nc.Rnw
minor change to text

@@
 Examination of the invariance properties of a measurement system is aimed at assessing whether the latent constructs in each group are measured in a comparable way, both in terms of scale and location.
-In the case of continuous measures, MI analysis proceeds in a relatively straightforward way \citep{Vanderberg2000a}; MI properties in systems of categorical outcomes are instead more problematic, due to the more complex nature of identification \citep{Millsap2004, Wu2016}.
+In the case of continuous measures, MI analysis proceeds in a relatively straightforward way \citep{Vanderberg2000}; MI properties in systems of categorical outcomes are instead more problematic, due to the more complex nature of identification \citep{Millsap2004, Wu2016a}.
 
 I set up a (confirmatory) item factor analysis (IFA) model for the full list of 11 common items.
-Given the results of the exploratory analysis -- where each measure separates neatly on one of the two factors -- I assume a \emph{dedicated} factor structure \citep{Carneiro2003}, where each measure is restricted to load on one of the two factors only.\footnote{This structure is commonly assumed in economic research to enable a more straightforward interpretation of the latent constructs in question \citep{ACFMR,Heckman2012a}.
+Given the results of the exploratory analysis -- where each measure separates neatly on one of the two factors -- I assume a \emph{dedicated} factor structure, where each measure is restricted to load on one of the two factors only.\footnote{This structure is commonly assumed in economic research to enable a more straightforward interpretation of the latent constructs in question \citep{Attanasio2017c,Heckman2013}.
 It can also be denominated \emph{congeneric} factor stucture \citep{Millsap2004}.}
@@
 \bnm[label=\Alph*:]
-\item Delta parameterisation {[WE{\textDelta}]} \citep{Wu2016}
+\item Delta parameterisation {[WE{\textDelta}]} \citep{Wu2016a}
 
@@
 
-\item Theta parameterisation {[WE{\textTheta}]} \citep{Wu2016}
+\item Theta parameterisation {[WE{\textTheta}]} \citep{Wu2016a}
 
@@
 The first two parameterisations (WE{\textDelta} and WE{\textTheta}) work by normalising the mean and variance of factors to the same constants in both groups, and they leave all loadings and intercepts to be freely estimated.
-The MT parameterisation instead proceeds by first identifying one group, and then imposing cross-group equality constraints to identify other groups \citep{Wu2016}.
+The MT parameterisation instead proceeds by first identifying one group, and then imposing cross-group equality constraints to identify other groups \citep{Wu2016a}.
 
@@
 In the case of continuous measures, the usual approach is to restrict one set of parameters at a time.
-As an example, starting from the configural model, one would restrict all loadings to be equal across groups; if the fit of this restricted model is not worse than the configural model, \emph{metric invariance} is estabilished \citep{Vanderberg2000a}.
+As an example, starting from the configural model, one would restrict all loadings to be equal across groups; if the fit of this restricted model is not worse than the configural model, \emph{metric invariance} is estabilished \citep{Vanderberg2000}.
 The hierarchy of the nested models usually proceeds by testing loadings first, and then intercepts (metric and scalar invariance).
@@
 A substantially similar approach is proposed for categorical measures by \cite{Millsap2004}.
-However, new identification results in \cite{Wu2016} indicate that, in the case of categorical measures, invariance properties cannot be examined by simply restricting one set of parameters at a time.
+However, new identification results in \cite{Wu2016a} indicate that, in the case of categorical measures, invariance properties cannot be examined by simply restricting one set of parameters at a time.
 This is because the identification conditions used in the configural baseline model, while being minimally restrictive on their own, become restrictive once certain additional restrictions are imposed.
@@
 
-Following the ordering in \cite{Wu2016}, I estimate the following increasingly restrictive models:
+Following the ordering in \cite{Wu2016a}, I estimate the following increasingly restrictive models:
 \bi
@@
 I estimate each of the models listed above by Weighted Least Squares.\footnote{All estimates are computed using the \texttt{lavaan} package in \textbf{\textsf{R}} \citep{Rosseel2012}.
-Parameters of the model are estimated by mean- and variance adjusted weighted least squares (WLSMV) (see \citealp{Mplus2017}).
+Parameters of the model are estimated by mean- and variance adjusted weighted least squares (WLSMV) (see \citealp{Muthen2017a}).
 Point estimates for model parameters are obtained starting from the polychoric correlations, using diagonally weighted least squares (DWLS), and the full weight matrix is used to compute robust standard errors and test statistics.
-Robust WLS has proved in simulation studies to be moderately robust to small violations of the normality assumption in the latent underlying measures \citep{Flora2011}, and generally outperforms maximum likelihood in large samples \citep{Beauducel2009,Li2016}.} The fit of the increasingly restrictive models presented above is compared in \autoref{tab:fit}.
+Robust WLS has proved in simulation studies to be moderately robust to small violations of the normality assumption in the latent underlying measures \citep{Flora2004}, and generally outperforms maximum likelihood in large samples \citep{Beauducel2006,Li2016}.} The fit of the increasingly restrictive models presented above is compared in \autoref{tab:fit}.
 For all models, a chi-squared difference would point to a lack of measurement invariance.
@@
 For CFI, a single cutoff values of $.002$ is proposed, while cutoffs for MFI depend on the problem's characteristics; in our case (2 factors, 11 items), they suggest $.0066$.
-\cite{Sass2014} however cast some doubts of the generalisability of these cutoffs to WLSMV estimators.}
+\cite{Sass2014a} however cast some doubts of the generalisability of these cutoffs to WLSMV estimators.}
 
@@
 \bibliographystyle{chicago}
-\bibliography{/Users/giacomomason/Documents/BibTeX/Papers-cohorts_nc}
-%\bibliography{/Users/giacomomason/Dropbox/Zotero/cohorts.bib}
+%\bibliography{/Users/giacomomason/Documents/BibTeX/Papers-cohorts_nc}
+\bibliography{/Users/giacomomason/Dropbox/Zotero/cohorts.bib}
 

### 2018-04-10 codes/R/cohorts_nc.Rnw
add tables to quantify gradients (M and F)

@@
 From the results of the exploratory FA and IRT analysis with 2 factors, the pattern that seems to emerge is consistent with the theoretical constructs of externalising (EXT) and internalising (INT) behaviour problems.
-
-<<BCSEFA>>=
-@
+% 
+% <<BCSEFA>>=
+% @
 % 
@@
 
-<<MCSEFA>>=
-@
+% 
+% <<MCSEFA>>=
+% @
 % 
@@
 
+\clearpage
+\begin{figure}
+\begin{center}
+<<INCINEQ, dependson="FA_SCORES_MAIN", dev='tikz', fig.height=4, fig.width=4, out.width='.5\\linewidth'>>=
+@
+\end{center}
+\caption{Evolution of income inequality }
+{\scriptsize \emph{Notes}:  \par}
+\end{figure}
+
 \begin{figure}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \clearpage
-% THIS BREAKS THE CACHING FOR SOME REASON
-% <<IQRDEC_FUN>>=
-% # load functions
+\clearpage
+
+<<DEC_VARS>>=
+# define variables for decomposition
+@
+% 
+% !!!!!!!!!! THIS BREAKS CACHING !!!!!!!!!!!!!!!!!!!!!
+% <<DEC_IQR, dependson="DEC_VARS">>=
+% # decompose SD and IQRs
 % @
 % 
-% <<IQRDEC, dependson="FA_SCORES_MAIN">>=
-% # decompose
+% Compare inequality statistics of original vs residualised scores
+% <<DEC_RES, dependson='DEC_IQR', echo=T, results='markup'>>=
+% iqs.orig
+% iqs.resid
 % @
-% 
-% % Compare inequality statistics of original vs residualised scores
-% % <<IQRDEC_RES, dependson='IQRDEC', echo=T, results='markup'>>=
-% % iqs.orig
-% % iqs.resid
-% % @
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\caption{Quantifying the gradient -- males}
+\centering
+<<FA_REGS_REAL_M, dependson='DEC_VARS', results='asis'>>=
+# regress scores on income
+library(dplyr)
+library(stargazer)
+
+r.ext.m.bcs <- lm(form.ext, data=subset(scores2plot, sex=="M" & cohort=="BCS"))
+r.ext.m.mcs <- lm(form.ext, data=subset(scores2plot, sex=="M" & cohort=="MCS"))
+r.int.m.bcs <- lm(form.int, data=subset(scores2plot, sex=="M" & cohort=="BCS"))
+r.int.m.mcs <- lm(form.int, data=subset(scores2plot, sex=="M" & cohort=="MCS"))
+
+stargazer(r.ext.m.bcs, r.ext.m.mcs , r.int.m.bcs, r.int.m.mcs, #r.ext.f.bcs, r.ext.f.mcs , r.int.f.bcs, r.int.f.mcs,
+          column.labels=c("BCS M","MCS M","BCS M","MCS M"), #,"BCS F","MCS F","BCS F","MCS F"),
+          dep.var.caption = "", multicolumn = FALSE,
+          omit=c("Constant", paste0("region", c(2:9))),
+          # covariate.labels=c(
+          #                    "Income Q2",
+          #                    "Income Q3",
+          #                    "Income Q4",
+          #                    "Income Q5"
+          #                    ),
+          no.space=TRUE,
+          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
+          notes = "All estimates control for Government Office Region of birth.",
+          float=FALSE    # needs begin table outside it
+)
+@
+\end{table}
+
+\begin{table}
+\caption{Quantifying the gradient -- females}
+\centering
+<<FA_REGS_REAL_F, dependson='DEC_VARS', results='asis'>>=
+# regress scores on income
+library(dplyr)
+library(stargazer)
+
+r.ext.m.bcs <- lm(form.ext, data=subset(scores2plot, sex=="F" & cohort=="BCS"))
+r.ext.m.mcs <- lm(form.ext, data=subset(scores2plot, sex=="F" & cohort=="MCS"))
+r.int.m.bcs <- lm(form.int, data=subset(scores2plot, sex=="F" & cohort=="BCS"))
+r.int.m.mcs <- lm(form.int, data=subset(scores2plot, sex=="F" & cohort=="MCS"))
+
+stargazer(r.ext.m.bcs, r.ext.m.mcs , r.int.m.bcs, r.int.m.mcs, #r.ext.f.bcs, r.ext.f.mcs , r.int.f.bcs, r.int.f.mcs,
+          column.labels=c("BCS F","MCS F","BCS F","MCS F"), #,"BCS F","MCS F","BCS F","MCS F"),
+          dep.var.caption = "", multicolumn = FALSE,
+          omit=c("Constant", paste0("region", c(2:9))),
+          # covariate.labels=c(
+          #                    "Income Q2",
+          #                    "Income Q3",
+          #                    "Income Q4",
+          #                    "Income Q5"
+          #                    ),
+          no.space=TRUE,
+          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
+          notes = "All estimates control for Government Office Region of birth.",
+          float=FALSE    # needs begin table outside it
+)
+@
+\end{table}
 
@@
 
-<<COUNTERF, dependson="FA_SCORES_MAIN">>=
-# perform counterfactuals
-@
-
-<<COUNTERF_PREP_PLOTS, dependson='COUNTERF'>>=
-# prepare counterfactual plots
-@
-
-\begin{landscape}
-\begin{figure}
-<<COUNTERF_EXT, dependson='COUNTERF_PREP_PLOTS', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
-cfp.ext
-@
-\caption{Counterfactual decomposition -- externalising skills}\label{fig:counterf}
-{\scriptsize \emph{Notes}: \par}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-<<COUNTERF_INT, dependson='COUNTERF_PREP_PLOTS', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
-cfp.int
-@
-\caption{Counterfactual decomposition -- internalising skills}\label{fig:counterf}
-{\scriptsize \emph{Notes}: \par}
-\end{figure}
-\end{landscape}
+% 
+% <<COUNTERF, dependson="FA_SCORES_MAIN">>=
+% # perform counterfactuals
+% @
+% 
+% <<COUNTERF_PREP_PLOTS, dependson='COUNTERF'>>=
+% # prepare counterfactual plots
+% @
+% 
+% \begin{landscape}
+% \begin{figure}
+% <<COUNTERF_EXT, dependson='COUNTERF_PREP_PLOTS', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
+% cfp.ext
+% @
+% \caption{Counterfactual decomposition -- externalising skills}\label{fig:counterf}
+% {\scriptsize \emph{Notes}: \par}
+% \end{figure}
+% \end{landscape}
+% 
+% \begin{landscape}
+% \begin{figure}
+% <<COUNTERF_INT, dependson='COUNTERF_PREP_PLOTS', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
+% cfp.int
+% @
+% \caption{Counterfactual decomposition -- internalising skills}\label{fig:counterf}
+% {\scriptsize \emph{Notes}: \par}
+% \end{figure}
+% \end{landscape}
 

### 2018-04-10 codes/R/getpars.R
function to extract parameters and make tables

@@ -1,0 +1,158 @@
+
+## ---- FUN_GETPARS
+
+########################################################################
+# GETMEASPARS
+# Function to extract measurement parameters from lavaan parameter table
+
+getmeaspars <- function(fit, groups = 2, mode = "tli") {
+  
+  if (groups!=2 & groups!=4) {
+    stop("Number of groups is 2 or 4")
+  }
+  
+  if (!mode %in% c("tl", "tli")) {
+    stop("\"mode\" is either \"tl\" or \"tli\" ")
+  }
+  
+  pta <- partable(fit)
+  
+  # item parameters ----------------------------------------------------------
+  # extract loadings
+  lambdas <- pta[(pta$op == "=~" & pta$group == 1),c("lhs", "rhs","est")]
+  lambdas$measure <- lambdas$rhs   # column for merging
+  lambdas <- lambdas[ , -which(names(lambdas) %in% c("rhs","lhs"))] # drop useless columns
+  
+  # extract tau1
+  tau1 <- pta[(pta$op == "|" & pta$rhs == "t1" & pta$group == 1),c("lhs", "rhs","est")]
+  tau1$measure <- tau1$lhs
+  tau1 <- tau1[ , -which(names(tau1) %in% c("rhs","lhs"))] # drop useless columns
+  
+  # extract tau2
+  tau2 <- pta[(pta$op == "|" & pta$rhs == "t2" & pta$group == 1),c("lhs", "rhs","est")]
+  tau2$measure <- tau2$lhs
+  tau2 <- tau2[ , -which(names(tau2) %in% c("rhs","lhs"))] # drop useless columns
+  
+  # extract variances (group 2 only)
+  eps <- pta[(pta$op == "~~" & substr(pta$rhs,1,1) == "X" & pta$group == 2),c("lhs", "rhs","est")]
+  eps$measure <- eps$lhs
+  eps <- eps[ , -which(names(eps) %in% c("rhs","lhs"))] # drop useless columns
+  
+  # extract intercepts (group 2 only)
+  nus <- pta[(pta$op == "~1" & substr(pta$lhs,1,1) == "X" & pta$group == 2),c("lhs", "rhs","est")]
+  nus$measure <- nus$lhs   # column for merging
+  nus <- nus[ , -which(names(nus) %in% c("rhs","lhs"))] # drop useless columns
+  
+  # merge parameters together
[…]

### 2018-04-10 codes/R/cohorts_nc.Rnw
change _ to . in social class labels to avoid plot conflicts in latex

@@
 
-% 
-% <<COUNTERF, dependson="FA_SCORES_MAIN">>=
-% # perform counterfactuals
-% @
-% 
-% <<COUNTERF_PREP_PLOTS, dependson='COUNTERF'>>=
-% # prepare counterfactual plots
-% @
-% 
-% \begin{landscape}
-% \begin{figure}
-% <<COUNTERF_EXT, dependson='COUNTERF_PREP_PLOTS', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
-% cfp.ext
-% @
-% \caption{Counterfactual decomposition -- externalising skills}\label{fig:counterf}
-% {\scriptsize \emph{Notes}: \par}
-% \end{figure}
-% \end{landscape}
-% 
-% \begin{landscape}
-% \begin{figure}
-% <<COUNTERF_INT, dependson='COUNTERF_PREP_PLOTS', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
-% cfp.int
-% @
-% \caption{Counterfactual decomposition -- internalising skills}\label{fig:counterf}
-% {\scriptsize \emph{Notes}: \par}
-% \end{figure}
-% \end{landscape}
+<<COUNTERF, dependson="FA_SCORES_MAIN">>=
+# perform counterfactuals
+@
+
+<<COUNTERF_PREP_PLOTS, dependson='COUNTERF'>>=
+# prepare counterfactual plots
+@
+
+\begin{landscape}
+\begin{figure}
+<<COUNTERF_EXT, dependson='COUNTERF_PREP_PLOTS', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
+cfp.ext
+@
+\caption{Counterfactual decomposition -- externalising skills}\label{fig:counterf_ext}
+{\scriptsize \emph{Notes}: \par}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+<<COUNTERF_INT, dependson='COUNTERF_PREP_PLOTS', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
+cfp.int
+@
+\caption{Counterfactual decomposition -- internalising skills}\label{fig:counterf_int}
+{\scriptsize \emph{Notes}: \par}
+\end{figure}
+\end{landscape}
 

### 2018-04-10 codes/R/1_load.R

@@ -183,7 +183,7 @@ items.c$scl10b[items.c$scl10 %in% c("IIINM")] <- 4
 items.c$scl10b[items.c$scl10 %in% c("IIIM")] <- 3
 items.c$scl10b[items.c$scl10 %in% c("IV", "V")] <- 2
 items.c$scl10b[items.c$scl10 %in% c("other")] <- 1
-items.c$scl10b <- factor(items.c$scl10b, labels = c("oth","IV_V","IIIM","IIINM","I_II"))
+items.c$scl10b <- factor(items.c$scl10b, labels = c("oth","IV V","IIIM","IIINM","I II"))
 
 # convert to factor
 items.c$smkpr <- factor(items.c$smkpr)

### 2018-04-10 codes/R/5_graphs.R

@@ -154,6 +154,7 @@ legend_b <- get_legend(ineqlist[[1]] + theme(legend.position="bottom"))
 p <- plot_grid( pcol, legend_b, ncol = 1, rel_heights = c(1, .1))
 p
 
+
 ############################################################################################
 ## ---- LOESS_REAL
 require(grid)

### 2018-04-11 codes/R/7_counterf.R
fix formula

@@ -10,7 +10,7 @@ quants <- seq(.1,.9,.025)
 
 # remove region from formulas
 form.ext.cf <- as.formula(paste("EXT ~ ", paste0(decvars[decvars!="region"], collapse = '+')))
-form.int.cf <- as.formula(paste("EXT ~ ", paste0(decvars[decvars!="region"], collapse = '+')))
+form.int.cf <- as.formula(paste("INT ~ ", paste0(decvars[decvars!="region"], collapse = '+')))
 
 # Logit regression
 logitcf.ext.m <- counterfactual(form.ext.cf, data = subset(scores2plot, sex=='M'),

### 2018-04-11 codes/R/7_counterf_graphs.R
extract more features from fitted counterfactual

@@ -6,17 +6,23 @@ getcfdat <- function(cffit) {
   plotdata <- data.frame(cbind(
     quant = cffit$quantiles,
     str = cffit$resSE[,1],
+    str.se = cffit$resSE[,2],
     str.l = cffit$resSE[,3],
     str.u = cffit$resSE[,4],
     com = cffit$resCE[,1],
+    com.se = cffit$resCE[,2],
     com.l = cffit$resCE[,3],
     com.u = cffit$resCE[,4],
     tot = cffit$resTE[,1],
+    tot.se = cffit$resTE[,2],
     tot.l = cffit$resTE[,3],
     tot.u = cffit$resTE[,4],
     qf.cf = cffit$model_quantile_counter[,1],
+    qf.cf.se = cffit$model_quantile_counter[,2],
     qf.ref1 = cffit$model_quantile_ref1[,1],
-    qf.ref0 = cffit$model_quantile_ref0[,1]
+    qf.ref1.se = cffit$model_quantile_ref1[,2],
+    qf.ref0 = cffit$model_quantile_ref0[,1],
+    qf.ref0.se = cffit$model_quantile_ref0[,2]
   ))
   return(plotdata)
 }

### 2018-04-11 codes/R/cohorts_nc.Rnw
add quantile differences decomposition

@@
 read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/7_counterf_graphs.R")
+read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/7_counterf_tabs.R")
 
@@
 
+<<COUNTERF_PREP_TAB, dependson='COUNTERF'>>=
+# prepare counterfactual tables
+@
+
+<<COUNTERF_INEQ, dependson='COUNTERF_PREP_TABS', results='markup'>>=
+cat("Males EXT")
+print(iqdecomp(logitcf.ext.m))
+
+cat("Males INT")
+print(iqdecomp(logitcf.int.m))
+
+cat("Females EXT")
+print(iqdecomp(logitcf.ext.f))
+
+cat("Females INT")
+print(iqdecomp(logitcf.int.f))
+@
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-04-11 codes/R/7_counterf_tabs.R

@@ -1,0 +1,27 @@
+## ---- COUNTERF_PREP_TAB
+# prepare tables with summary inequality statistics
+
+iqdecomp <- function(cffit) {
+  cfd <- getcfdat(cffit) # get output from counterfactual
+  
+  # matrix of quantile differences of interest
+  qdiffmat <- matrix(c(.9,.1,
+                       .5,.1,
+                       .9,.5,
+                       .75,.25), nrow=4, byrow=T)
+  difftab <- matrix(NA, nrow=4, ncol=3)
+  rownames(difftab) <- c("90-10", "50-10", "90-50", "75-25")
+  colnames(difftab) <- c("Total Diff", "Structure", "Composition")
+  for (r in 1:nrow(qdiffmat)) {
+    attach(cfd)
+    
+    changes <- c(
+      tot = (qf.ref1[quant==qdiffmat[r,1]] - qf.ref1[quant==qdiffmat[r,2]]) - (qf.ref0[quant==qdiffmat[r,1]] - qf.ref0[quant==qdiffmat[r,2]]),
+      str = (qf.ref1[quant==qdiffmat[r,1]] - qf.ref1[quant==qdiffmat[r,2]]) - (qf.cf[quant==qdiffmat[r,1]] - qf.cf[quant==qdiffmat[r,2]]),
+      com = (qf.cf[quant==qdiffmat[r,1]] - qf.cf[quant==qdiffmat[r,2]]) - (qf.ref0[quant==qdiffmat[r,1]] - qf.ref0[quant==qdiffmat[r,2]])
+    )
+    detach(cfd)
+    difftab[r,] <- changes
+  }
+  return(difftab)
+}

### 2018-04-12 codes/R/6_ineq.R
move list of regressors

@@ -1,14 +1,5 @@
 # INEQUALITY DECOMPOSITION
 
-## ---- DEC_VARS
-# select variables for decomposition
-decvars <- c("region","faminc_infl","scl10b","ysch_moth5","smkpr","mothageb","numch5")
-
-# formulas
-form.ext <- as.formula(paste("EXT ~ ", paste0(decvars, collapse = '+')))
-form.int <- as.formula(paste("INT ~ ", paste0(decvars, collapse = '+')))
-
-
 #######################################################################################################
 ## ---- DEC_IQR
 # functions

### 2018-04-12 codes/R/7_counterf.R

@@ -1,4 +1,13 @@
 #######################################################################################################
+## ---- DEC_VARS
+# select variables for decomposition
+decvars <- c("region","faminc_infl","scl10b","ysch_moth5","mothageb","numch5")
+
+# formulas
+form.ext <- as.formula(paste("EXT ~ ", paste0(decvars, collapse = '+')))
+form.int <- as.formula(paste("INT ~ ", paste0(decvars, collapse = '+')))
+
+
 ## ---- COUNTERF
 library(Counterfactual)
 

### 2018-04-12 codes/R/7_counterf_graphs.R
composition before structure

@@ -62,14 +62,14 @@ makecfplots <- function(cffit) {
   p.tot <- ggplot(cfdata, aes(quant, tot))+
     geom_line() + ggtitle("Total difference") +
     geom_ribbon(aes(ymin=tot.l,ymax=tot.u),alpha=0.3)
-  p.str <- ggplot(cfdata, aes(quant, str))+
-    geom_line() + ggtitle("Structure") +
-    geom_ribbon(aes(ymin=str.l,ymax=str.u),alpha=0.3)
   p.com <- ggplot(cfdata, aes(quant, com))+
     geom_line() + ggtitle("Composition") +
     geom_ribbon(aes(ymin=com.l,ymax=com.u),alpha=0.3)
+  p.str <- ggplot(cfdata, aes(quant, str))+
+    geom_line() + ggtitle("Structure") +
+    geom_ribbon(aes(ymin=str.l,ymax=str.u),alpha=0.3)
   
-  cfplist <- list(p.qfs, p.tot, p.str, p.com) 
+  cfplist <- list(p.qfs, p.tot, p.com, p.str) 
   cfplist <- lapply(cfplist, addopts.all) # apply options to all graphs
   cfplist[2:4] <- lapply(cfplist[2:4], addopts.ce) # apply options to all graphs
   

### 2018-04-12 codes/R/cohorts_nc.Rnw
add sequential decomposition graphs

@@
 read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/7_counterf_tabs.R")
+read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/7_counterf_seqdec.R")
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% COUNTERFACTUAL DECOMPOSITIONS
+
+<<COUNTERF_SEQDEC_PREP>>=
+# data for decomposition plots
+@
+
+<<COUNTERF_SEQDEC_PLOT>>=
+# decomposition plots
+@
+
+\begin{landscape}
+\begin{figure}
+<<COUNTERF_SEQDEC_EXT, dependson='COUNTERF_SEQDEC_PREP', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
+cfp.seqdec.ext
+@
+\caption{Counterfactual sequential decomposition -- externalising skills}\label{fig:counterf_seq_ext}
+{\scriptsize \emph{Notes}: \par}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+<<COUNTERF_SEQDEC_INT, dependson='COUNTERF_SEQDEC_PREP', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
+cfp.seqdec.int
+@
+\caption{Counterfactual sequential decomposition -- internalising skills}\label{fig:counterf_seq_int}
+{\scriptsize \emph{Notes}: \par}
+\end{figure}
+\end{landscape}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \bibliographystyle{chicago}
-%\bibliography{/Users/giacomomason/Documents/BibTeX/Papers-cohorts_nc}
 \bibliography{/Users/giacomomason/Dropbox/Zotero/cohorts.bib}

### 2018-04-12 codes/R/7_counterf_seqdec.R

@@ -1,0 +1,122 @@
+## ---- COUNTERF_SEQDEC_PREP
+
+# function to make sequential counterfactual decomposition plots
+# arguments: 
+# 1) list of CF estimates with increasing num of covariates
+# 2) list of composition plot titles (defaults to X1-XK)
+makecfdecplots <- function(flst, titles = NA) {
+  
+  numdec <- length(flst) # number of decompositions
+  
+  if (is.na(titles)) {
+    titles <- paste0("X",seq(1:numdec))
+  } 
+  else {
+    if (length(titles) != numdec) stop("List of titles must be same as number of decompositions ")
+  }
+  
+  # assemble decomposition data
+  cfdb <- data.frame(cbind(flst[[1]]$quantiles, flst[[1]]$total_effect)) # get quantiles and total effect
+  colnames(cfdb) <- c("quant","tot")
+  for (i in 1:numdec) cfdb <- cbind(cfdb, flst[[i]]$composition_effect)     # get composition effects
+  colnames(cfdb)[3:(3+numdec-1)] <- paste0("com", seq(1:numdec))
+  cfdb <- cbind(cfdb, flst[[numdec]]$structral_effect)                    # get last structure effect
+  colnames(cfdb)[3+numdec] <- "str"
+  cfdb$comd1 <- cfdb$com1                                   # compute differences in composition effects
+  for (i in 2:numdec) cfdb[,paste0("comd",i)] <- cfdb[,paste0("com",i)] - cfdb[,paste0("com",i-1)]
+  
+  # get axis boundaries
+  maxy <- max(cfdb[! names(cfdb) %in% "quant"]) +.05
+  miny <- min(cfdb[! names(cfdb) %in% "quant"]) -.05
+  
+  # common options
+  addopts.all <- function(x) {
+    x <- x +
+      theme(
+        axis.title.y=element_blank(),
+        plot.title = element_text(hjust = 1, face = "italic"),
+        legend.position="none"
+      ) +
+      scale_x_continuous(name = "Quantile", breaks = seq(.1,.9,.1)) +
+      coord_cartesian(ylim = c(miny, maxy)) +
+      labs(list(colour="")) +
+      geom_smooth(se=F) + geom_line() # PLOTS
+    return(x)
+  }
+  
+  # plots
+  cfplist <- list()
[…]

### 2018-04-13 codes/R/paper/cohorts_nc.Rnw
here, folder structure

--- 2018-04-12 codes/R/cohorts_nc.Rnw
@@
 <<READ, cache=FALSE >>=
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/1_load.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/2_efa.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/3_inv_main.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/3_inv_aux.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/3_inv_fit.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/4_scores.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/4_scores_aux.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/5_graphs.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/6_ineq.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/7_counterf.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/7_counterf_graphs.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/7_counterf_tabs.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/7_counterf_seqdec.R")
+library(here)
+
+read_chunk(here("codes/R/1_load.R"))
+read_chunk(here("codes/R/2_efa.R"))
+read_chunk(here("codes/R/3_inv_main.R"))
+read_chunk(here("codes/R/3_inv_aux.R"))
+read_chunk(here("codes/R/3_inv_fit.R"))
+read_chunk(here("codes/R/4_scores.R"))
+read_chunk(here("codes/R/4_scores_aux.R"))
+read_chunk(here("codes/R/5_graphs.R"))
+read_chunk(here("codes/R/6_ineq.R"))
+read_chunk(here("codes/R/7_counterf.R"))
+read_chunk(here("codes/R/7_counterf_graphs.R"))
+read_chunk(here("codes/R/7_counterf_tabs.R"))
+read_chunk(here("codes/R/7_counterf_seqdec.R"))
 

### 2018-04-25 codes/R/paper/cohorts_nc.Rnw
commonalities table by gender

@@
 \footnotesize
-\caption{Item Commonalities}\label{tab:comm}
-<<MEANTABLE_RES, dependson="MEANTABLE", results='asis'>>=
+\caption{Item Commonalities -- Males}\label{tab:comm_m}
+<<MEANTABLE_RES_M, dependson="MEANTABLE", results='asis'>>=
 require(xtable)
 
-table <- meantab
+table <- meantab[[1]]
+
+print(xtable(table,
+                                  # item num        Factor          Num cats        BCS text        BCS num           BCS Certain     BCS Somewhat    BCS Applies             MCS text          MCS num
+                    align="L{0cm}C{.04\\textwidth}C{.05\\textwidth}C{.05\\textwidth}L{.3\\textwidth}C{.03\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}L{.3\\textwidth}C{.03\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
+                    digits = 1
+                    ), 
+             size="\\small",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0)),
+                              command = c(
+                              # header
+                                    paste(" & & &
+                                              \\multicolumn{5}{c}{\\textbf{BCS}} &
+                                              \\multicolumn{5}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){4-8}  \\cmidrule(lr){9-13} 
+                                            Item & Factor & N. categ &
+                                              Text & & Cert. Applies & Somet. Applies & Applies &
+                                              Text & & Cert. True & Somet. True & True \\\\ \\midrule
+                                          "
+                                            , sep="")
+                                    )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = ".8\\textwidth",
+             floating = FALSE
+)
+
+@
+\end{table}
+\end{landscape}
+
+\clearpage
+\begin{landscape}
+\begin{table}[ht]
+\centering
+\footnotesize
+\caption{Item Commonalities -- Females}\label{tab:comm_f}
+<<MEANTABLE_RES_F, dependson="MEANTABLE", results='asis'>>=
+require(xtable)
+
+table <- meantab[[2]]
 

### 2018-04-25 codes/R/1_load.R

@@ -225,68 +225,78 @@ for (i in 1:length(items)) items[[i]] <- items[[i]][order(items[[i]][,"cohortsex
 ## ---- MEANTABLE
 # table of mean values of items
 
-ncomm <- 11                             # number of common items
+ncomm <- 11                           # number of common items
 ncats <- c(3,3,3,3,2,2,3,3,3,3,2)     # number of categories per item
 
-means <- matrix(NA, ncomm, 7)
-colnames(means) <- c("num", "BCS_ca", "BCS_sa", "BCS_a", "MCS_ca", "MCS_sa", "MCS_a")
+means <- list()
+X.allg <- list()
+meantab <- list()
+means[[1]] <- matrix(NA, ncomm, 7)    # males
+means[[2]] <- matrix(NA, ncomm, 7)    # females
+X.allg[[1]] <- X.all[X.all$sex==1,]
+X.allg[[2]] <- X.all[X.all$sex==2,]
 
-for (i in 1:ncomm) {
-  means[i,1] <- i
-  t <- table(X.all$cohort, X.all[,grep("X[0-9]", names(X.all))][,i]) # ith column of item-only matrix
-  t2 <- 100*prop.table(t, 1)                           # convert to row percentages
+for (g in c(1,2)) {                   # males/females
+  colnames(means[[g]]) <- c("num", "BCS_ca", "BCS_sa", "BCS_a", "MCS_ca", "MCS_sa", "MCS_a")
   
-  if (ncats[i] == 3) { # 3-category items
-    means[i,2] <- t2[1,1]
-    means[i,3] <- t2[1,2]
-    means[i,5] <- t2[2,1]
-    means[i,6] <- t2[2,2]
+  for (i in 1:ncomm) {                # items
+    means[[g]][i,1] <- i
+    t <- table(X.allg[[g]]$cohort, X.allg[[g]][,grep("X[0-9]", names(X.allg[[g]]))][,i]) # ith column of item-only matrix
+    t2 <- 100*prop.table(t, 1)                           # convert to row percentages
+    
+    if (ncats[i] == 3) { # 3-category items
+      means[[g]][i,2] <- t2[1,1]
+      means[[g]][i,3] <- t2[1,2]
+      means[[g]][i,5] <- t2[2,1]
+      means[[g]][i,6] <- t2[2,2]
+    }
+    if (ncats[i] == 2) { # binary items
[…]

### 2018-04-25 codes/R/paper/cohorts_nc.Rnw
add EFA in appendix

@@
 From the results of the exploratory FA and IRT analysis with 2 factors, the pattern that seems to emerge is consistent with the theoretical constructs of externalising (EXT) and internalising (INT) behaviour problems.
-% 
-% <<BCSEFA>>=
-% @
-% 
-% <<BCSEFARES, dependson='BCSEFA', echo=T, results='markup'>>=
-% @
 
@@
 Thus, our two-factor structure seems justifiable on both theoretical and empirical grounds.
-
-% 
-% <<MCSEFA>>=
-% @
-% 
-% # <<MCSEFARES, dependson='BCSEFA', echo=T, results='markup'>>=
-% # @
-%
 
@@
 \begin{appendices}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Exploratory Factor Analysis}
+
+<<BCSEFA>>=
+@
+
+<<MCSEFA>>=
+@
+
+<<BCSEFARES, dependson='BCSEFA', results='markup'>>=
+@
+
+<<MCSEFARES, dependson='BCSEFA', results='markup'>>=
+@
+
+\begin{table}[ht]
+\caption{Number of factors} \label{tab:nfac}
+<<EFATAB, dependson='MCSEFARES', results='asis'>>=
+# combine results
+efatab <- cbind(efatab.bcs,efatab.mcs)
+colnames(efatab) <- c("BCS All", "BCS M", "BCS F", "MCS All", "MCS M", "MCS F")
+
+print(xtable(efatab,
+                    align="L{.25\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}",
+                    digits = 0
+                    ), 
+             size="\\small",
+             include.rownames=T,
+             include.colnames=T,
+             booktabs = T,
+             sanitize.text.function = function(x){x},
+             tabular.environment = "tabular", width = ".8\\textwidth",
+             floating = FALSE
+)
+@
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\clearpage
 \section{Estimated model parameters}

### 2018-04-25 codes/R/1_load.R

@@ -297,7 +297,7 @@ for (g in c(1,2)) {                   # males/females
 }
 
 
-
+rm(mcskeepb, ncats, ncomm, t, t2, X.all, X.allg, means)
 
 
 

### 2018-04-25 codes/R/2_efa.R

@@ -41,11 +41,13 @@ efatab.bcs <- cbind(
 )
 rownames(efatab.bcs) <- c("Optimal Coordinates", "Acceleration Factor", "Parallel Analysis", "Kaiser", 
                           "VSS Compl. 1", "VSS Compl. 2", "Velicer MAP")
-efatab.bcs
 
 # FACTOR ANALYSIS loadings
+cat("\n\n ----------------------- \n BCS Loadings - Males and Females")
 bcsfa$loadings
+cat("\n\n ----------------------- \n BCS Loadings - Males")
 bcsfa.m$loadings
+cat("\n\n ----------------------- \n BCS Loadings - Females")
 bcsfa.f$loadings
 
 ## ---- BCSREL
@@ -95,11 +97,13 @@ efatab.mcs <- cbind(
   
 rownames(efatab.mcs) <- c("Optimal Coordinates", "Acceleration Factor", "Parallel Analysis", "Kaiser", 
                           "VSS Compl. 1", "VSS Compl. 2", "Velicer MAP")
-efatab.mcs
 
 # FACTOR ANALYSIS
+cat("\n\n ----------------------- \n MCS Loadings - Males and Females")
 mcsfa$loadings
+cat("\n\n ----------------------- \n MCS Loadings - Males")
 mcsfa.m$loadings
+cat("\n\n ----------------------- \n MCS Loadings - Females")
 mcsfa.f$loadings
 
 ## ---- MCSREL

### 2018-04-25 codes/R/paper/cohorts_nc.Rnw
add mini table of quantile income differences

@@
 
+<<INCINEQ_STATS, dependson="FA_SCORES_MAIN", results="markup">>=
+tb <- data.frame(
+  aggregate(scores2plot$faminc_infl, by=list(scores2plot$cohort), FUN = function(x) quantile(x, c(.1, .25, .75, .9), na.rm=T))
+)
+tb$d9010 <- tb$x[,4]/tb$x[,1]
+tb$d7525 <- tb$x[,3]/tb$x[,2]
+
+cat("Quantile differences in income by cohort")
+tb
+@
+
 \clearpage

### 2018-04-27 codes/R/1_load.R
packages

@@ -22,6 +22,7 @@ library(tikzDevice)
 library(gtools)
 library(here)
 library(rio)
+library(Counterfactual)
 
 
 dir_data  <- here("rdata/")
@@ -36,7 +37,7 @@ printfit <- function(m) {
 
 # auxiliary variables to retain
 auxvars <- c("sex", "region",
-             "bwt", "lowbwt", "parity", "nprevst", "caesbirth", "smkpr", "gestaw", "preterm", 
+             "bwt", "lowbwt", "parity", "firstb", "nprevst", "caesbirth", "smkpr", "gestaw", "preterm", 
              "mothageb", "teenm", "singlem", "mheight", "mempl",
              "mysch5", "fysch5", "numch5", "ageint5", "mhied5",
              "faminc10_real", "faminc10_infl", "incq10", "scl10"
@@ -191,9 +192,9 @@ items.c$scl10b[items.c$scl10 %in% c("other")] <- 5
 items.c$scl10b <- factor(items.c$scl10b, labels = c("IV V","IIIM","IIINM","I II","oth"))
 
 # convert to factor
-facnms <- c("smkpr", "region", "lowbwt", "caesbirth", "preterm", "teenm", "singlem", "mempl", "mhied5")
+facnms <- c("smkpr", "region", "lowbwt", "caesbirth", "preterm", "firstb", "teenm", "singlem", "mempl", "mhied5")
 items.c[,facnms] <- lapply(items.c[,facnms] , factor)
-items.c$incq <- ordered(items.c$incq)
+items.c$incq10 <- ordered(items.c$incq10)
 
 # save dataset
 export(items.c[,!names(items.c) %in% c("INT.RAW", "EXT.RAW", "INT.RAWr", "EXT.RAWr")], paste0(dir_data,"cohorts_all.dta"))

### 2018-05-03 codes/R/7_counterf_seqdec.R
less ticks on x axis

@@ -1,4 +1,5 @@
 ## ---- COUNTERF_SEQDEC_PREP
+require(Counterfactual)
 
 # function to make sequential counterfactual decomposition plots
 # arguments: 
@@ -37,7 +38,7 @@ makecfdecplots <- function(flst, titles = NA) {
         plot.title = element_text(hjust = 1, face = "italic"),
         legend.position="none"
       ) +
-      scale_x_continuous(name = "Quantile", breaks = seq(.1,.9,.1)) +
+      scale_x_continuous(name = "Quantile", breaks = seq(.1,.9,.2)) +
       coord_cartesian(ylim = c(miny, maxy)) +
       labs(list(colour="")) +
       geom_smooth(se=F) + geom_line() # PLOTS
@@ -59,12 +60,12 @@ makecfdecplots <- function(flst, titles = NA) {
 }
 
 
+# ESTIMATE COUNTERFACTUALS (without inference)
 
-
-# MAKE COUNTERFACTUALS (without inference)
-
-# sequence of decomposition variables
-decseq <- list("faminc10_real", "scl10b", "mysch5", paste0(c("mothageb", "numch5"), collapse = "+"))
+# numeric cohort
+finaldata$cohn <- as.numeric(finaldata$cohort)-1
+# decide how many quantiles
+quants <- seq(.04,.9,.02)
 
 # allocate
 lcf.ext.m <- list()
@@ -74,22 +75,22 @@ lcf.int.f <- list()
 
 for (j in 1:length(decseq)) {
   lcf.ext.m[[j]] <- counterfactual(as.formula(paste("EXT ~ ", paste0(decseq[1:j], collapse = '+')))
-                                   , data = subset(scores2plot, sex=='M'),
+                                   , data = subset(finaldata, sex=='M'),
                                    group = cohn, treatment=TRUE, decomposition=TRUE, 
                                    method = "logit", quantiles = quants,
                                    nreg=100, noboot = T)
   lcf.ext.f[[j]] <- counterfactual(as.formula(paste("EXT ~ ", paste0(decseq[1:j], collapse = '+')))
-                                   , data = subset(scores2plot, sex=='F'),
+                                   , data = subset(finaldata, sex=='F'),
                                    group = cohn, treatment=TRUE, decomposition=TRUE, 
                                    method = "logit", quantiles = quants,
                                    nreg=100, noboot = T)
[…]

### 2018-05-03 codes/R/paper/cohorts_nc.Rnw
add formula chunk, add 90-10 dec for EXT/INT

--- 2018-04-25 codes/R/paper/cohorts_nc.Rnw
@@
 read_chunk(here("codes/R/5_graphs.R"))
-read_chunk(here("codes/R/6_ineq.R"))
+read_chunk(here("codes/R/6_decvars.R"))
 read_chunk(here("codes/R/7_counterf.R"))
@@
 read_chunk(here("codes/R/7_counterf_seqdec.R"))
+read_chunk(here("codes/R/getpars.R"))
 
@@
 
-<<FA_INV_MAIN, dependson="LOAD_DATA">>=
+<<FA_INV_MAIN>>=
 # main models only
@@
 
-% % <<FA_INV_AUX, dependson='LOAD_DATA'>>=
-% % # robustness models
-% % @
-
-<<FA_FIT, dependson='FA_INV_MAIN'>>=
+<<FA_FIT>>=
 # assemble fit indices into matrices
@@
 
-<<FA_SCORES_MAIN, dependson='FA_INV_MAIN'>>=
+<<FA_SCORES_MAIN>>=
 # produce scores for main models
 @
-
-% % <<FA_SCORES_AUX, dependson='FA_INV_AUX'>>=
-% % # produce scores for robustness models
-% % @
 
@@
 \begin{figure}
-<<FACDENS, dependson="FA_SCORES_MAIN", dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
+<<FACDENS, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
 @
@@
 In \autoref{fig:meanci}, we can see that the gradient in noncognitive skills along family income quintiles is more pronounced in the MCS cohort.
-%In particular, a male five-year old at the top quintile of the income distribution in 1975 has a better score by \Sexpr{regcoeffs[1,3]} in the externalising dimension and \Sexpr{regcoeffs[1,4]} in the internalising dimension, compared to a child at the bottom quintile.
-In 2005, the differences amount to \Sexpr{regcoeffs[3,3]} and \Sexpr{regcoeffs[3,4]}, respectively.
+
+<<INEQ_STATS, dependson="FA_SCORES_MAIN", results="markup">>=
+
+# income 
+tb <- data.frame(
+  aggregate(scores2plot$faminc10_infl, by=list(scores2plot$cohort), FUN = function(x) quantile(x, c(.1, .25, .75, .9), na.rm=T))
+)
+tb$d9010 <- tb$x[,4]/tb$x[,1]
+tb$d7525 <- tb$x[,3]/tb$x[,2]
+cat("Quantile differences in income by cohort")
+tb
+
+tb.ext <- data.frame(
+  aggregate(scores2plot$EXT, by=list(scores2plot$cohortsex), FUN = function(x) quantile(x, c(.1, .25, .5, .75, .9), na.rm=T))
+)
+tb.ext$d9010 <- abs(tb.ext$x[,"90%"]-tb.ext$x[,"10%"])
+tb.ext$d9010 <- abs(tb.ext$x[,"75%"]-tb.ext$x[,"25%"])
+tb.ext$d5010 <- abs(tb.ext$x[,"50%"]-tb.ext$x[,"10%"])
+tb.ext$d9050 <- abs(tb.ext$x[,"90%"]-tb.ext$x[,"50%"])
+cat("\n Quantile differences in EXT by cohort/gender")
+tb.ext
+
+tb.int <- data.frame(
+  aggregate(scores2plot$INT, by=list(scores2plot$cohortsex), FUN = function(x) quantile(x, c(.1, .25, .5, .75, .9), na.rm=T))
+)
+tb.int$d9010 <- abs(tb.int$x[,"90%"]-tb.int$x[,"10%"])
+tb.int$d9010 <- abs(tb.int$x[,"75%"]-tb.int$x[,"25%"])
+tb.int$d5010 <- abs(tb.int$x[,"50%"]-tb.int$x[,"10%"])
+tb.int$d9050 <- abs(tb.int$x[,"90%"]-tb.int$x[,"50%"])
+cat("\n Quantile differences in INT by cohort/gender")
+tb.int
+
+@
 
@@
 \end{figure}
-
-\begin{figure}
-<<FACINEQ, dependson="FA_SCORES_MAIN", dev='tikz', fig.height=12, fig.width=10, out.width='1\\linewidth'>>=
-@
-\caption{Mean scores across income quintiles}\label{fig:meanci}
-{\scriptsize \emph{Notes}: The figure plots the mean and its 95\% confidence interval across the quintiles in the income distribution. \par}
-\end{figure}
+% 
+% \begin{figure}
+% <<FACINEQ, dependson="FA_SCORES_MAIN", dev='tikz', fig.height=12, fig.width=10, out.width='1\\linewidth'>>=
+% @
+% \caption{Mean scores across income quintiles}\label{fig:meanci}
+% {\scriptsize \emph{Notes}: The figure plots the mean and its 95\% confidence interval across the quintiles in the income distribution. \par}
+% \end{figure}
 
@@
 
-<<DEC_VARS>>=
-# define variables for decomposition
+<<FORMULAS>>=
+# define variables and formulas for regression and decomposition
 @
-% 
-% !!!!!!!!!! THIS BREAKS CACHING !!!!!!!!!!!!!!!!!!!!!
-% <<DEC_IQR, dependson="DEC_VARS">>=
-% # decompose SD and IQRs
-% @
-% 
-% Compare inequality statistics of original vs residualised scores
-% <<DEC_RES, dependson='DEC_IQR', echo=T, results='markup'>>=
-% iqs.orig
-% iqs.resid
-% @
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{table}
-\caption{Quantifying the gradient -- males}
-\centering
-<<FA_REGS_REAL_M, dependson='DEC_VARS', results='asis'>>=
-# regress scores on income
-library(dplyr)
-library(stargazer)
-
-r.ext.m.bcs <- lm(form.ext, data=subset(scores2plot, sex=="M" & cohort=="BCS"))
-r.ext.m.mcs <- lm(form.ext, data=subset(scores2plot, sex=="M" & cohort=="MCS"))
-r.int.m.bcs <- lm(form.int, data=subset(scores2plot, sex=="M" & cohort=="BCS"))
-r.int.m.mcs <- lm(form.int, data=subset(scores2plot, sex=="M" & cohort=="MCS"))
-
-stargazer(r.ext.m.bcs, r.ext.m.mcs , r.int.m.bcs, r.int.m.mcs, #r.ext.f.bcs, r.ext.f.mcs , r.int.f.bcs, r.int.f.mcs,
-          column.labels=c("BCS M","MCS M","BCS M","MCS M"), #,"BCS F","MCS F","BCS F","MCS F"),
-          dep.var.caption = "", multicolumn = FALSE,
-          omit=c("Constant", paste0("region", c(2:9))),
-          # covariate.labels=c(
-          #                    "Income Q2",
-          #                    "Income Q3",
-          #                    "Income Q4",
-          #                    "Income Q5"
-          #                    ),
-          no.space=TRUE,
-          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
-          notes = "All estimates control for Government Office Region of birth.",
-          float=FALSE    # needs begin table outside it
-)
-@
-\end{table}
-
-\begin{table}
-\caption{Quantifying the gradient -- females}
-\centering
-<<FA_REGS_REAL_F, dependson='DEC_VARS', results='asis'>>=
-# regress scores on income
-library(dplyr)
-library(stargazer)
-
-r.ext.m.bcs <- lm(form.ext, data=subset(scores2plot, sex=="F" & cohort=="BCS"))
-r.ext.m.mcs <- lm(form.ext, data=subset(scores2plot, sex=="F" & cohort=="MCS"))
-r.int.m.bcs <- lm(form.int, data=subset(scores2plot, sex=="F" & cohort=="BCS"))
-r.int.m.mcs <- lm(form.int, data=subset(scores2plot, sex=="F" & cohort=="MCS"))
-
-stargazer(r.ext.m.bcs, r.ext.m.mcs , r.int.m.bcs, r.int.m.mcs, #r.ext.f.bcs, r.ext.f.mcs , r.int.f.bcs, r.int.f.mcs,
-          column.labels=c("BCS F","MCS F","BCS F","MCS F"), #,"BCS F","MCS F","BCS F","MCS F"),
-          dep.var.caption = "", multicolumn = FALSE,
-          omit=c("Constant", paste0("region", c(2:9))),
-          # covariate.labels=c(
-          #                    "Income Q2",
-          #                    "Income Q3",
-          #                    "Income Q4",
-          #                    "Income Q5"
-          #                    ),
-          no.space=TRUE,
-          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
-          notes = "All estimates control for Government Office Region of birth.",
-          float=FALSE    # needs begin table outside it
-)
-@
-\end{table}
+% 
+% \begin{table}
+% \caption{Quantifying the gradient}
+% \centering
+% \resizebox{\textwidth}{!}{
+% <<FA_REGS_REAL_M, results='asis'>>=
+% # regress scores on income
+% library(stargazer)
+% 
+% decvars_reg <- c(decvars, "cohort") # add cohort
+% interact_reg <- paste0(decvars_reg[! decvars_reg %in% c("cohort")], "*cohort")
+% 
+% form_ext_reg <- as.formula(paste("EXT ~ ", paste0(decvars2, collapse = '+')))
+% form.ext2i <- as.formula(paste("EXT ~ ", paste0(c(decvars2, interact2), collapse = '+')))
+% form.int2 <- as.formula(paste("INT ~ ", paste0(decvars2, collapse = '+')))
+% form.int2i <- as.formula(paste("INT ~ ", paste0(c(decvars2, interact2), collapse = '+')))
+% 
+% r.ext.m  <- lm(form.ext2,  data=subset(scores2plot, sex=="M"))
+% r.exti.m <- lm(form.ext2i, data=subset(scores2plot, sex=="M"))
+% r.ext.f  <- lm(form.ext2,  data=subset(scores2plot, sex=="F"))
+% r.exti.f <- lm(form.ext2i, data=subset(scores2plot, sex=="F"))
+% r.int.m  <- lm(form.int2,  data=subset(scores2plot, sex=="M"))
+% r.inti.m <- lm(form.int2i, data=subset(scores2plot, sex=="M"))
+% r.int.f  <- lm(form.int2,  data=subset(scores2plot, sex=="F"))
+% r.inti.f <- lm(form.int2i, data=subset(scores2plot, sex=="F"))
+% 
+% stargazer(r.ext.m, r.exti.m, r.ext.f, r.exti.f, r.int.m, r.inti.m, r.int.f, r.inti.f,
+%           column.labels=c("M","M","F","F","M","M","F","F"), #,"BCS F","MCS F","BCS F","MCS F"),
+%           dep.var.caption = "", multicolumn = FALSE,
+%           #omit=c("Constant", paste0("region", c(2:9))),
+%           no.space=TRUE,
+%           omit.stat= c("rsq", "f", "ser", "aic", "ll"),
+%           notes = "All estimates control for Government Office Region of birth.",
+%           single.row=TRUE,
+%           float=FALSE    # needs begin table outside it
+% )
+% @
+% }
+% \end{table}
 
@@
 
-<<COUNTERF, dependson="FA_SCORES_MAIN">>=
-# perform counterfactuals
-@
-
-<<COUNTERF_PREP_PLOTS, dependson='COUNTERF'>>=
-# prepare counterfactual plots
-@
-
-\begin{landscape}
-\begin{figure}
-<<COUNTERF_EXT, dependson='COUNTERF_PREP_PLOTS', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
-cfp.ext
-@
-\caption{Counterfactual decomposition -- externalising skills}\label{fig:counterf_ext}
-{\scriptsize \emph{Notes}: \par}
-\end{figure}
-\end{landscape}
-
-\begin{landscape}
-\begin{figure}
-<<COUNTERF_INT, dependson='COUNTERF_PREP_PLOTS', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
-cfp.int
-@
-\caption{Counterfactual decomposition -- internalising skills}\label{fig:counterf_int}
-{\scriptsize \emph{Notes}: \par}
-\end{figure}
-\end{landscape}
-
-<<COUNTERF_PREP_TAB, dependson='COUNTERF'>>=
-# prepare counterfactual tables
-@
-
-<<COUNTERF_INEQ, dependson='COUNTERF_PREP_TABS', results='markup'>>=
-cat("Males EXT")
-print(iqdecomp(logitcf.ext.m))
-
-cat("Males INT")
-print(iqdecomp(logitcf.int.m))
-
-cat("Females EXT")
-print(iqdecomp(logitcf.ext.f))
-
-cat("Females INT")
-print(iqdecomp(logitcf.int.f))
-@
+% 
+% <<COUNTERF, dependson="FA_SCORES_MAIN">>=
+% # perform counterfactuals
+% @
+% 
+% <<COUNTERF_PREP_PLOTS, dependson='COUNTERF'>>=
+% # prepare counterfactual plots
+% @
+% 
+% \begin{landscape}
+% \begin{figure}
+% <<COUNTERF_EXT, dependson='COUNTERF_PREP_PLOTS', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
+% cfp.ext
+% @
+% \caption{Counterfactual decomposition -- externalising skills}\label{fig:counterf_ext}
+% {\scriptsize \emph{Notes}: \par}
+% \end{figure}
+% \end{landscape}
+% 
+% \begin{landscape}
+% \begin{figure}
+% <<COUNTERF_INT, dependson='COUNTERF_PREP_PLOTS', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
+% cfp.int
+% @
+% \caption{Counterfactual decomposition -- internalising skills}\label{fig:counterf_int}
+% {\scriptsize \emph{Notes}: \par}
+% \end{figure}
+% \end{landscape}
+% 
+% <<COUNTERF_PREP_TAB, dependson='COUNTERF'>>=
+% # prepare counterfactual tables
+% @
+% 
+% <<COUNTERF_INEQ, dependson='COUNTERF_PREP_TABS', results='markup'>>=
+% cat("Males EXT")
+% print(iqdecomp(logitcf.ext.m))
+% 
+% cat("Males INT")
+% print(iqdecomp(logitcf.int.m))
+% 
+% cat("Females EXT")
+% print(iqdecomp(logitcf.ext.f))
+% 
+% cat("Females INT")
+% print(iqdecomp(logitcf.int.f))
+% @
 
@@
 <<RES_FA_PAR, dependson="FA_FIT", results='asis'>>=
-source("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/getpars.R")
 
@@
 require(xtable)
-source("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/getpars.R")
 
@@
 
+\clearpage
+\section{Summary statistics}
+\begin{table}[ht]
+\caption{Summary statistics} \label{tab:lvpars}
+\centering
+<<SUMSTATS, results='asis'>>=
+require(dummies)
+
+datacompl <- scores2plot[complete.cases(scores2plot[,decvars2]),] # complete cases
+# expand dummies
+datasum <- cbind(
+  datacompl[,c("EXT","INT","cohortsex")], 
+  dummy.data.frame(datacompl[,c(decvars2)], sep = "")
+)
+
+# statistics by group
+sumtab <- data.frame(
+  aggregate(datasum, by=list(datasum$cohortsex), FUN = function(x) c(mn = mean(x, na.rm=T)))
+)
+
+# drop unused factors and variables
+sumtab <- sumtab[, substring(names(sumtab), nchar(names(sumtab))) != "0"]
+sumtab <- sumtab[, ! names(sumtab) %in% c("EXT","INT","cohortsex","cohortBCS","cohortMCS")]
+group <- sumtab$Group.1
+sumtabt <- as.data.frame(t(sumtab[,-1]))
+colnames(sumtabt) <- group
+# rownames to var
+sumtabt$var <- rownames(sumtabt)
+# reorder columns
+sumtabt <- sumtabt[c("var", "BCS.M", "MCS.M", "BCS.F", "MCS.F")]
+
+# table
+print.xtable(xtable(sumtabt,
+                    align="L{0cm}L{3cm}C{1.2cm}C{2cm}C{2cm}C{2cm}",
+                    digits = 3
+                    ),
+             size="\\small",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0)),
+                              command = c(
+                              # header
+                                   "& \\multicolumn{2}{c}{\\textbf{Males}} & \\multicolumn{2}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-3} \\cmidrule(lr){4-5}
+                                    & BCS & MCS & BCS & MCS \\\\"
+                                    )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabularx", width = "\\textwidth",
+             floating = FALSE
+)
+
+@
+\end{table}
+
 \end{appendices}

### 2018-05-03 codes/R/5_graphs.R
add raw score histograms

@@ -17,8 +17,48 @@ ggplot(aged, aes(x=age, y=dens, fill=cohort)) +
   theme(legend.position = c(0.9, 0.2)) +
   geom_hline(aes(yintercept=0))
 
+
+############################################################################################
+## ---- RAWHIST
+# histograms for raw scores
+
+# common options
+addopts.raw <- function(x) {
+  x <- x +
+    theme(
+      axis.title.y=element_blank(),
+      legend.position="none"
+    ) +
+    scale_x_continuous(name = "Raw score", breaks = seq(0,10,1)) +
+    scale_fill_discrete("") + # remove fill guide title
+    scale_colour_discrete(guide=FALSE) +  # remove colour legend
+    geom_bar(position="dodge", aes(y = ..prop..)) # PLOTS
+  return(x)
+}
+
+rawext.m <- ggplot(subset(scores2plot, sex=="M"),aes(x=EXT.RAW, fill=cohort)) + ggtitle("EXT Scores (Males)") + coord_cartesian(ylim = c(0,.25))
+rawint.m <- ggplot(subset(scores2plot, sex=="M"),aes(x=INT.RAW, fill=cohort)) + ggtitle("INT Scores (Males)") + coord_cartesian(ylim = c(0,.4))
+rawext.f <- ggplot(subset(scores2plot, sex=="F"),aes(x=EXT.RAW, fill=cohort)) + ggtitle("EXT Scores (Females)") + coord_cartesian(ylim = c(0,.25))
+rawint.f <- ggplot(subset(scores2plot, sex=="F"),aes(x=INT.RAW, fill=cohort)) + ggtitle("INT Scores (Females)") + coord_cartesian(ylim = c(0,.4))
+rawlist <- list(rawext.m, rawint.m, rawext.f, rawint.f) 
+rawlist <- lapply(rawlist, addopts.raw) # apply options to all graphs
+
+pcol <- plot_grid( rawlist[[1]], rawlist[[2]], rawlist[[3]], rawlist[[4]],
+                   align = 'vh',
+                   hjust = -1,
+                   nrow = 2
+)
+
+# add legend
+legend_b <- get_legend(rawlist[[1]] + theme(legend.position="bottom"))
+p <- plot_grid( pcol, legend_b, ncol = 1, rel_heights = c(1, .1))
+p
+
+
 ############################################################################################
 ## ---- FACDENS
+# densities of scored factors
+
 # Kolmogorov-Smirnov test
 ksp <- list()
 for (g in c("M", "F")) {

### 2018-05-09 codes/R/paper/cohorts_nc.Rnw
add raw score histograms
add new graph by mother employment, add histograms

@@
 
+\clearpage
+\begin{figure}
+<<RAWHIST, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
+@
+\caption{Frequency of Raw Scores}
+{\scriptsize \emph{Notes}: The figure shows frequencies of raw (sum) scores separately by cohort.
+Raw scores are obtained using the same 11 items in the CFA model, building a categorical index.
+Children for which each behaviour ``certainly applies/true" are assigned zero points, one point for ``somewhat applies/true" and two points for ``does not apply/not true".
+For binary items, one point is assigned for ``true".
+The raw scores are the sum of the points across the 11 behaviours considered, six for EXT and five for INT.
+\par}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \end{figure}
+
+<<FACINEQ_DEMEAN>>=
+# function for demeaning
+@
+
 % 
@@
 \caption{Mean scores across social class}\label{fig:meanci_sc}
-{\scriptsize \emph{Notes}: The figure plots the mean and its 95\% confidence interval across the quintiles in the income distribution. \par}
+{\scriptsize \emph{Notes}: The figure plots the mean and its 95\% confidence interval across harmonised social class of the cohort member's family at 10 years of age. \par}
+\end{figure}
+
+\begin{figure}
+<<FACINEQ_YS, dependson="FA_SCORES_MAIN", dev='tikz', fig.height=12, fig.width=10, out.width='1\\linewidth'>>=
+@
+\caption{Mean scores across maternal schooling}\label{fig:meanci_ys}
+{\scriptsize \emph{Notes}: The figure plots the mean and its 95\% confidence interval across grouped years of schooling of the cohort member's mother at 5 years of age. \par}
+\end{figure}
+
+\begin{figure}
+<<FACINEQ_ME, dependson="FA_SCORES_MAIN", dev='tikz', fig.height=12, fig.width=10, out.width='1\\linewidth'>>=
+@
+\caption{Mean scores across maternal employment}\label{fig:meanci_me}
+{\scriptsize \emph{Notes}: The figure plots the mean and its 95\% confidence interval by employment status of the cohort member's mother at 5 years of age. \par}
 \end{figure}
@@
 <<SUMSTATS, results='asis'>>=
-require(dummies)
-
-datacompl <- scores2plot[complete.cases(scores2plot[,decvars2]),] # complete cases
+
+datacompl <- scores2plot[complete.cases(scores2plot[,decvars]),] # complete cases
 # expand dummies
@@
   datacompl[,c("EXT","INT","cohortsex")], 
-  dummy.data.frame(datacompl[,c(decvars2)], sep = "")
+  dummy.data.frame(datacompl[,c(decvars)], sep = "")
 )

### 2018-05-09 codes/R/1_load.R

@@ -199,7 +199,7 @@ levels(items.c$mysch5b) <- c("Compuls.", "C - C+2", "C+3 - C+5", ">C+5")
 facnms <- c("smkpr", "region", "lowbwt", "caesbirth", "preterm", "firstb", "teenm", "singlem", "mempl", "mempl5", "mhied5")
 items.c[,facnms] <- lapply(items.c[,facnms] , factor)
 items.c$incq10 <- ordered(items.c$incq10)
-levels(items.c$mempl5) <- c("Unemployed", "Empl./Education")
+levels(items.c$mempl5) <- c("Unempl./At home", "Empl./Education")
 
 # save dataset
 export(items.c[,!names(items.c) %in% c("INT.RAW", "EXT.RAW", "INT.RAWr", "EXT.RAWr")], paste0(dir_data,"cohorts_all.dta"))

### 2018-05-09 codes/R/5_graphs.R

@@ -177,114 +177,225 @@ legend_b <- get_legend(ineqlist[[1]] + theme(legend.position="bottom"))
 p <- plot_grid( pcol, legend_b, ncol = 1, rel_heights = c(1, .1))
 p
 
+############################################################################################
+## ---- FACINEQ_DEMEAN
+# function to demean variables in data to have mean zero in a given level
+# data = dataset
+# refvar = reference variable (factor)
+# reflev = level of refvar at which to demean
+# suffix = string to append to demeaned variable
+demean <- function(data, refvar, reflev, suffix) {
+  
+  # get mean by cohortsex to rescale
+  tempdata <- data[data[,refvar]==reflev,]
+  meansfac <- aggregate(tempdata[,c("EXT","INT")], by=list(tempdata$cohortsex), FUN = function(x) mean(x, na.rm=T))
+  rownames(meansfac) <- meansfac$Group.1
+  meansfac <- meansfac[,-1]
+  # rescale scores so that mean in lowest level is 0
+  for (f in c("EXT","INT")) {
+    fd <- paste0(f,suffix)
+    data[,fd] <- NA
+    for (c in c("BCS","MCS")) {
+      for (g in c("M","F")) {
+        cs <- paste0(c,".",g)
+        data[data$cohortsex==cs,fd] <- data[data$cohortsex==cs,f]-meansfac[cs,f]
+      }
+    }
+  }
+  return(data)
+}
 
 ############################################################################################
 ## ---- FACINEQ_SC
 # MEAN/CI plot of scores by social class
+library(dplyr)
+library(grid)
+library(gridExtra)
+library(ggpubr)
 
-# get mean by cohortsex to rescale
-scoreslowclass <- scores2plot[scores2plot$scl10b=="IV V",]
-meansfac <- aggregate(scoreslowclass[,c("EXT","INT")], by=list(scoreslowclass$cohortsex), FUN = function(x) mean(x, na.rm=T))
-rownames(meansfac) <- meansfac$Group.1
-meansfac <- meansfac[,-1]
+# demean scores for lowest class
+scores2plot <- demean(scores2plot, "scl10b", "IV V", "dsc")
 
-# rescale scores so that mean in lowest level is 0
-scores2plot$EXTdsc <- NA
[…]

### 2018-05-10 codes/R/6_selectvars.R
change name

@@ -1,5 +1,5 @@
 #######################################################################################################
-## ---- FORMULAS
+## ---- SELECTVARS
 
 # select variables for decomposition, in groups
 decvarslist <- list()

### 2018-05-10 codes/R/7_counterf_seqdec.R
minor graph edit

@@ -94,7 +94,7 @@ makecfdecplots <- function(flst, titles = NA, kbwt = 0.1) {
     x <- x +
       theme(
         axis.title.y=element_blank(),axis.line.y=element_blank(),axis.ticks.y=element_blank(),axis.text.y=element_blank(),
-        plot.title = element_text(margin=margin(t=10), face = "italic"),
+        plot.title = element_text(margin=margin(t=15), face = "italic"), # italic title inside graph
         legend.position="none"
       ) +
       scale_x_continuous(name = "Quantile", breaks = seq(.1,.9,.2)) +

### 2018-05-10 codes/R/8_regs.R
separate file with all regressions

@@ -1,0 +1,79 @@
+## ---- REGRESSIONS
+
+# function to collapse regressors in formula
+pplus <- function(x) paste0(x, collapse="+")
+
+###################################################################
+# REGRESSIONS OF EXT/INT ON CHILDHOOD VARIABLES
+# add region to vector of regressors
+decvars_reg <- c(decvars, c("region"))
+decvars_int_reg <- c(decvars_int, c("region"))
+
+# formulas
+form_ext <- list()
+form_ext[[1]] <- as.formula("EXT ~ mysch5b")
+form_ext[[2]] <- as.formula(paste("EXT ~ ", pplus(decvars_reg)))
+form_ext[[3]] <- as.formula(paste("EXT ~ ", pplus(decvars_int_reg)))
+form_int <- list()
+form_int[[1]] <- as.formula("INT ~ mysch5b")
+form_int[[2]] <- as.formula(paste("INT ~ ", pplus(decvars_reg)))
+form_int[[3]] <- as.formula(paste("INT ~ ", pplus(decvars_int_reg)))
+
+# regressions
+r.ext <- list()
+r.int <- list()
+for (cs in c("BCS.M", "MCS.M", "BCS.F", "MCS.F")) {
+  r.ext[[cs]] <- list()
+  r.int[[cs]] <- list()
+  for (i in 1:3) {
+    r.ext[[cs]][[i]]  <- lm(form_ext[[i]],  data=subset(scores2plot, cohortsex==cs))
+    r.int[[cs]][[i]]  <- lm(form_int[[i]],  data=subset(scores2plot, cohortsex==cs))
+  }
+}
+
+
+###################################################################
+# REGRESSIONS OF ADOLESCENT OUTCOMES ON EXT/INT & CHILDHOOD VARIABLES
+
+# assemble outcomes data
+bcs16outc <- read.dta(paste0(dir_data, "bcs16outc.dta"), convert.factors = F) # all BCS data
+bcs16outc$id <- bcs16outc$bcsid
+bcs16outc$age16 <- factor(bcs16outc$age16)
+bcsoutc <- merge(finaldata, bcs16outc, by="id")
+
[…]

### 2018-05-10 codes/R/paper/cohorts_nc.Rnw
add all tables

@@
 read_chunk(here("codes/R/5_graphs.R"))
-read_chunk(here("codes/R/6_decvars.R"))
+read_chunk(here("codes/R/6_selectvars.R"))
 read_chunk(here("codes/R/7_counterf.R"))
@@
 read_chunk(here("codes/R/7_counterf_seqdec.R"))
+read_chunk(here("codes/R/8_regs.R"))
 read_chunk(here("codes/R/getpars.R"))
@@
 
-% 
-% \begin{figure}
-% <<FACINEQ, dependson="FA_SCORES_MAIN", dev='tikz', fig.height=12, fig.width=10, out.width='1\\linewidth'>>=
-% @
-% \caption{Mean scores across income quintiles}\label{fig:meanci}
-% {\scriptsize \emph{Notes}: The figure plots the mean and its 95\% confidence interval across the quintiles in the income distribution. \par}
-% \end{figure}
-
 \begin{figure}
@@
 
-<<FORMULAS>>=
+<<SELECTVARS>>=
 # define variables and formulas for regression and decomposition
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% 
-% \begin{table}
-% \caption{Quantifying the gradient}
-% \centering
-% \resizebox{\textwidth}{!}{
-% <<FA_REGS_REAL_M, results='asis'>>=
-% # regress scores on income
-% library(stargazer)
-% 
-% decvars_reg <- c(decvars, "cohort") # add cohort
-% interact_reg <- paste0(decvars_reg[! decvars_reg %in% c("cohort")], "*cohort")
-% 
-% form_ext_reg <- as.formula(paste("EXT ~ ", paste0(decvars2, collapse = '+')))
-% form.ext2i <- as.formula(paste("EXT ~ ", paste0(c(decvars2, interact2), collapse = '+')))
-% form.int2 <- as.formula(paste("INT ~ ", paste0(decvars2, collapse = '+')))
-% form.int2i <- as.formula(paste("INT ~ ", paste0(c(decvars2, interact2), collapse = '+')))
-% 
-% r.ext.m  <- lm(form.ext2,  data=subset(scores2plot, sex=="M"))
-% r.exti.m <- lm(form.ext2i, data=subset(scores2plot, sex=="M"))
-% r.ext.f  <- lm(form.ext2,  data=subset(scores2plot, sex=="F"))
-% r.exti.f <- lm(form.ext2i, data=subset(scores2plot, sex=="F"))
-% r.int.m  <- lm(form.int2,  data=subset(scores2plot, sex=="M"))
-% r.inti.m <- lm(form.int2i, data=subset(scores2plot, sex=="M"))
-% r.int.f  <- lm(form.int2,  data=subset(scores2plot, sex=="F"))
-% r.inti.f <- lm(form.int2i, data=subset(scores2plot, sex=="F"))
-% 
-% stargazer(r.ext.m, r.exti.m, r.ext.f, r.exti.f, r.int.m, r.inti.m, r.int.f, r.inti.f,
-%           column.labels=c("M","M","F","F","M","M","F","F"), #,"BCS F","MCS F","BCS F","MCS F"),
-%           dep.var.caption = "", multicolumn = FALSE,
-%           #omit=c("Constant", paste0("region", c(2:9))),
-%           no.space=TRUE,
-%           omit.stat= c("rsq", "f", "ser", "aic", "ll"),
-%           notes = "All estimates control for Government Office Region of birth.",
-%           single.row=TRUE,
-%           float=FALSE    # needs begin table outside it
-% )
-% @
-% }
-% \end{table}
+
+<<REGRESSIONS, dependson="SELECTVARS">>=
+# run all regression models
+@
+
+\begin{table}
+\caption{Quantifying the gradient -- Males externalising}
+\centering
+\resizebox{\textwidth}{!}{
+<<REGS_EXT_M, results='asis', dependson="REGRESSIONS">>=
+library(stargazer)
+stargazer(r.ext$BCS.M[[1]], r.ext$BCS.M[[2]], r.ext$BCS.M[[3]], r.ext$MCS.M[[1]], r.ext$MCS.M[[2]], r.ext$MCS.M[[3]],
+          column.labels=c("BCS","BCS","BCS","MCS","MCS","MCS"),
+          dep.var.caption = "", multicolumn = FALSE,
+          #omit=c("Constant", paste0("region", c(2:9))),
+          no.space=TRUE,
+          omit.stat= c("f", "ser", "aic", "ll"),
+          #notes = "All estimates control for Government Office Region of birth.",
+          single.row=TRUE,
+          float=FALSE    # needs begin table outside it
+)
+@
+}
+\end{table}
+
+\begin{table}
+\caption{Quantifying the gradient -- Females externalising}
+\centering
+\resizebox{\textwidth}{!}{
+<<REGS_EXT_F, results='asis', dependson="REGRESSIONS">>=
+library(stargazer)
+stargazer(r.ext$BCS.F[[1]], r.ext$BCS.F[[2]], r.ext$BCS.F[[3]], r.ext$MCS.F[[1]], r.ext$MCS.F[[2]], r.ext$MCS.F[[3]],
+          column.labels=c("BCS","BCS","BCS","MCS","MCS","MCS"),
+          dep.var.caption = "", multicolumn = FALSE,
+          #omit=c("Constant", paste0("region", c(2:9))),
+          no.space=TRUE,
+          omit.stat= c("f", "ser", "aic", "ll"),
+          #notes = "All estimates control for Government Office Region of birth.",
+          single.row=TRUE,
+          float=FALSE    # needs begin table outside it
+)
+@
+}
+\end{table}
+
+\begin{table}
+\caption{Quantifying the gradient -- Males internalising}
+\centering
+\resizebox{\textwidth}{!}{
+<<REGS_INT_M, results='asis', dependson="REGRESSIONS">>=
+library(stargazer)
+stargazer(r.int$BCS.M[[1]], r.int$BCS.M[[2]], r.int$BCS.M[[3]], r.int$MCS.M[[1]], r.int$MCS.M[[2]], r.ext$MCS.M[[3]],
+          column.labels=c("BCS","BCS","BCS","MCS","MCS","MCS"),
+          dep.var.caption = "", multicolumn = FALSE,
+          #omit=c("Constant", paste0("region", c(2:9))),
+          no.space=TRUE,
+          omit.stat= c("f", "ser", "aic", "ll"),
+          #notes = "All estimates control for Government Office Region of birth.",
+          single.row=TRUE,
+          float=FALSE    # needs begin table outside it
+)
+@
+}
+\end{table}
+
+\begin{table}
+\caption{Quantifying the gradient -- Females internalising}
+\centering
+\resizebox{\textwidth}{!}{
+<<REGS_INT_F, results='asis', dependson="REGRESSIONS">>=
+library(stargazer)
+stargazer(r.int$BCS.F[[1]], r.int$BCS.F[[2]], r.int$BCS.F[[3]], r.int$MCS.F[[1]], r.int$MCS.F[[2]], r.int$MCS.F[[3]],
+          column.labels=c("BCS","BCS","BCS","MCS","MCS","MCS"),
+          dep.var.caption = "", multicolumn = FALSE,
+          #omit=c("Constant", paste0("region", c(2:9))),
+          no.space=TRUE,
+          omit.stat= c("f", "ser", "aic", "ll"),
+          #notes = "All estimates control for Government Office Region of birth.",
+          single.row=TRUE,
+          float=FALSE    # needs begin table outside it
+)
+@
+}
+\end{table}
+
+\begin{landscape}
+\begin{table}
+\caption{Effect of behavioural skills on adolescent outcomes -- BCS males}
+\centering
+\resizebox{24cm}{!}{
+<<REGS_OUTC_BCS_M, results='asis', dependson="REGRESSIONS">>=
+library(stargazer)
+# get mean dependent vars
+dvmeans <- colMeans(bcsoutc[bcsoutc$sex=="F",bcsoutcvars], na.rm = T)
+dvmeans <- round(unlist(lapply(dvmeans, function(x) rep(x,2))),3)
+
+stargazer(bcsoutcmod_m_uc[[1]], bcsoutcmod_m_c[[1]],
+          bcsoutcmod_m_uc[[2]], bcsoutcmod_m_c[[2]],
+          bcsoutcmod_m_uc[[3]], bcsoutcmod_m_c[[3]],
+          bcsoutcmod_m_uc[[4]], bcsoutcmod_m_c[[4]],
+          bcsoutcmod_m_uc[[5]], bcsoutcmod_m_c[[5]],
+          bcsoutcmod_m_uc[[6]], bcsoutcmod_m_c[[6]],
+          add.lines = list(c("Mean Dep Var", dvmeans)),
+          no.space=TRUE,
+          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
+          single.row=T,
+          float=FALSE    # needs begin table outside it
+)
+@
+}
+\end{table}
+\end{landscape}
+
+\begin{landscape}
+\begin{table}
+\caption{Effect of behavioural skills on adolescent outcomes -- MCS males}
+\centering
+\resizebox{24cm}{!}{
+<<REGS_OUTC_MCS_M, results='asis', dependson="REGRESSIONS">>=
+library(stargazer)
+# get mean dependent vars
+dvmeans <- colMeans(mcsoutc[mcsoutc$sex=="F",mcsoutcvars], na.rm = T)
+dvmeans <- round(unlist(lapply(dvmeans, function(x) rep(x,2))),3)
+
+stargazer(mcsoutcmod_m_uc[[1]], mcsoutcmod_m_c[[1]],
+          mcsoutcmod_m_uc[[2]], mcsoutcmod_m_c[[2]],
+          mcsoutcmod_m_uc[[3]], mcsoutcmod_m_c[[3]],
+          mcsoutcmod_m_uc[[4]], mcsoutcmod_m_c[[4]],
+          mcsoutcmod_m_uc[[5]], mcsoutcmod_m_c[[5]],
+          add.lines = list(c("Mean Dep Var", dvmeans)),
+          no.space=TRUE,
+          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
+          single.row=T,
+          float=FALSE    # needs begin table outside it
+)
+@
+}
+\end{table}
+\end{landscape}
+
+\begin{landscape}
+\begin{table}
+\caption{Effect of behavioural skills on adolescent outcomes -- BCS females}
+\centering
+\resizebox{24cm}{!}{
+<<REGS_OUTC_BCS_F, results='asis', dependson="REGRESSIONS">>=
+library(stargazer)
+# get mean dependent vars
+dvmeans <- colMeans(bcsoutc[bcsoutc$sex=="F",bcsoutcvars], na.rm = T)
+dvmeans <- round(unlist(lapply(dvmeans, function(x) rep(x,2))),3)
+
+stargazer(bcsoutcmod_f_uc[[1]], bcsoutcmod_f_c[[1]],
+          bcsoutcmod_f_uc[[2]], bcsoutcmod_f_c[[2]],
+          bcsoutcmod_f_uc[[3]], bcsoutcmod_f_c[[3]],
+          bcsoutcmod_f_uc[[4]], bcsoutcmod_f_c[[4]],
+          bcsoutcmod_f_uc[[5]], bcsoutcmod_f_c[[5]],
+          bcsoutcmod_f_uc[[6]], bcsoutcmod_f_c[[6]],
+          add.lines = list(c("Mean Dep Var", dvmeans)),
+          no.space=TRUE,
+          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
+          single.row=T,
+          float=FALSE    # needs begin table outside it
+)
+@
+}
+\end{table}
+\end{landscape}
+
+\begin{landscape}
+\begin{table}
+\caption{Effect of behavioural skills on adolescent outcomes -- MCS females}
+\centering
+\resizebox{24cm}{!}{
+<<REGS_OUTC_MCS_F, results='asis', dependson="REGRESSIONS">>=
+library(stargazer)
+# get mean dependent vars
+dvmeans <- colMeans(mcsoutc[mcsoutc$sex=="F",mcsoutcvars], na.rm = T)
+dvmeans <- round(unlist(lapply(dvmeans, function(x) rep(x,2))),3)
+
+stargazer(mcsoutcmod_f_uc[[1]], mcsoutcmod_f_c[[1]],
+          mcsoutcmod_f_uc[[2]], mcsoutcmod_f_c[[2]],
+          mcsoutcmod_f_uc[[3]], mcsoutcmod_f_c[[3]],
+          mcsoutcmod_f_uc[[4]], mcsoutcmod_f_c[[4]],
+          mcsoutcmod_f_uc[[5]], mcsoutcmod_f_c[[5]],
+          add.lines = list(c("Mean Dep Var", dvmeans)),
+          no.space=TRUE,
+          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
+          single.row=T,
+          float=FALSE    # needs begin table outside it
+)
+@
+}
+\end{table}
+\end{landscape}
 
@@
 
+<<COUNTERF_SEQDEC_ESTIM, dependson="SELECTVARS">>=
+# estimate sequential decompositions
+@
+
 <<COUNTERF_SEQDEC_PREP>>=
-# data for decomposition plots
+# extract data
 @
@@
 \begin{figure}
-<<COUNTERF_SEQDEC_EXT, dependson='COUNTERF_SEQDEC_PREP', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
+<<COUNTERF_SEQDEC_EXT, dependson='COUNTERF_SEQDEC_PLOT', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
 cfp.seqdec.ext
@@
 \begin{figure}
-<<COUNTERF_SEQDEC_INT, dependson='COUNTERF_SEQDEC_PREP', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
+<<COUNTERF_SEQDEC_INT, dependson='COUNTERF_SEQDEC_PLOT', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
 cfp.seqdec.int

### 2018-05-11 codes/R/5_graphs.R
delete incineq, reorder SCL in plot

@@ -213,6 +213,9 @@ library(grid)
 library(gridExtra)
 library(ggpubr)
 
+# reorder social class to have "other" on the LHS (scl10c)
+scores2plot$scl10c <- factor(scores2plot$scl10b, levels(scores2plot$scl10b)[c(5,1:4)])
+
 # demean scores for lowest class
 scores2plot <- demean(scores2plot, "scl10b", "IV V", "dsc")
 
@@ -227,20 +230,21 @@ addopts_main <- function(x) {
     labs(list(colour="")) + theme(legend.position="none")
   return(x)
 }
-ineq.ext.m <- ggplot(data=subset(scores2plot, sex=="M" & !is.na(scores2plot$scl10b)), aes(x=scl10b, y=EXTdsc, colour=cohort)) + ggtitle("Males Externalising")
-ineq.ext.f <- ggplot(data=subset(scores2plot, sex=="F" & !is.na(scores2plot$scl10b)), aes(x=scl10b, y=EXTdsc, colour=cohort)) + ggtitle("Females Externalising")
-ineq.int.m <- ggplot(data=subset(scores2plot, sex=="M" & !is.na(scores2plot$scl10b)), aes(x=scl10b, y=INTdsc, colour=cohort)) + ggtitle("Males Internalising")
-ineq.int.f <- ggplot(data=subset(scores2plot, sex=="F" & !is.na(scores2plot$scl10b)), aes(x=scl10b, y=INTdsc, colour=cohort)) + ggtitle("Females Internalising")
+ineq.ext.m <- ggplot(data=subset(scores2plot, sex=="M" & !is.na(scores2plot$scl10c)), aes(x=scl10c, y=EXTdsc, colour=cohort)) + ggtitle("Males Externalising")
+ineq.ext.f <- ggplot(data=subset(scores2plot, sex=="F" & !is.na(scores2plot$scl10c)), aes(x=scl10c, y=EXTdsc, colour=cohort)) + ggtitle("Females Externalising")
+ineq.int.m <- ggplot(data=subset(scores2plot, sex=="M" & !is.na(scores2plot$scl10c)), aes(x=scl10c, y=INTdsc, colour=cohort)) + ggtitle("Males Internalising")
+ineq.int.f <- ggplot(data=subset(scores2plot, sex=="F" & !is.na(scores2plot$scl10c)), aes(x=scl10c, y=INTdsc, colour=cohort)) + ggtitle("Females Internalising")
[…]

### 2018-05-19 codes/R/slides/cohorts_tapes.Rnw
removed some variables

--- 2018-03-01 codes/R/slides/cohorts_nc_slides.Rnw
@@
 \bi
-\item Non-cognitive (socio-emotional, behavioural) skills are important in childhood
+\item Non-cognitive (socio-emotional, behavioural) skills developed during childhood are important for lifetime outcomes
 \item Socioeconomic gradients in these skills are persistent 
@@
   \item Millennium Cohort Study (MCS, b. 2000/1) \ei
-\item Construct a common 11-item scale, from different instruments
-\item Investigate some psychometric properties
+\ei
+\vspace{6pt} \pause
+\bnm
+\item Construct a common 11-item scale of non-cognitive skills, starting from different instruments
+\item Investigate its psychometric properties
 \item Use scores from the new scale to document changes in socioeconomic gradients between cohorts 30 years apart
+\enm
+\vspace{6pt} \pause
+\bi
+\item \textbf{Key findings}:
+  \bi
+  \item Inequality in non-cognitive skills has increased in 1975-2005
+  \item Sharpest change for boys rather than girls
+  \item Mostly driven by improvements at the top of the distribution
+  \ei
 \ei
@@
 \bi
+\item Restricting thresholds and loadings provides a good fit
+  \bi \item We can compare the scale (variance) of the latent factors \ei
 \item Models with restricted intercepts do \emph{not} pass MI test
   \bi
-  \item \cite{Wu2016} suggests we cannot compare latent means
-  \item Traditional approaches would disagree... \citep{Millsap2004}
-  \ei
-\item Might \emph{age differences} in the sample be playing a role?
-  \bi
-  \item BCS and MCS samples differ due to survey design \hyperlink{agehist}{\beamerbutton{Graph}}
-  \item But restricting the age range to maximise overlap does not give significantly different results \hyperlink{fit4}{\beamerbutton{Det.}}
-  \ei
+  \item Unclear whether mean comparisons can be done
+  \item \cite{Wu2016} vs \cite{Millsap2004}
+  \ei
+% \item Might \emph{age differences} in the sample be playing a role?
+%   \bi
+%   \item BCS and MCS samples differ due to survey design \hyperlink{agehist}{\beamerbutton{Graph}}
+%   \item But restricting the age range to maximise overlap does not give significantly different results \hyperlink{fit4}{\beamerbutton{Det.}}
+%   \ei
   \pause
 \item \textbf{Q1}: Can we derive comparable measures of noncognitive skills across cohorts?
-\item Yes, we can the model with non-invariant intercepts to score the latent factors (externalising and internalising skills)
+\item Yes, we can use the model with non-invariant intercepts to score the latent factors (externalising and internalising skills)
 \ei
@@
 
+% \begin{frame}
+% \frametitle{Inequality}
+% \framesubtitle{Quantifying the gradient -- inflation- and growth-adjusted income}
+% \centering
+% \resizebox{!}{4cm}{%
+% <<FA_REGS_REAL, dependson='FA_SCORES_SELECT', results='asis'>>=
+% # regress scores on income
+% library(dplyr)
+% library(stargazer)
+% 
+% scores2plot$incq_real <- factor(ntile(scores2plot$faminc_real, 5))
+% 
+% # get min, max
+% minmax.real <-aggregate(scores2plot[,c("faminc_real")],list(scores2plot$incq_real), 
+%             FUN = function(x) c(min = min(x, na.rm=T), max = max(x, na.rm=T)))
+% minmax.real <- round(minmax.real[,-1],2)*1000
+% 
+% r.ext.f.q <- lm(EXT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="F"))
+% r.int.f.q <- lm(INT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="F"))
+% 
+% r.ext.m.q <- lm(EXT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="M"))
+% r.int.m.q <- lm(INT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="M"))
+% 
+% stargazer(r.ext.m.q, r.int.m.q, r.ext.f.q, r.int.f.q,
+%           column.labels=c("Males","Females"), column.separate=c(2,2),
+%           dep.var.caption = "", multicolumn = FALSE,
+%           omit=c("Constant"),
+%           covariate.labels=c(
+%                 paste0("Income Q2 [",toString(minmax.real[2,1]),"-",toString(minmax.real[2,2])," £/w]"),
+%                 paste0("Income Q3 [",toString(minmax.real[3,1]),"-",toString(minmax.real[3,2])," £/w]"),
+%                 paste0("Income Q4 [",toString(minmax.real[4,1]),"-",toString(minmax.real[4,2])," £/w]"),
+%                 paste0("Income Q5 [",toString(minmax.real[5,1]),"+ £/w]"),
+%                              "MCS dummy",
+%                              "Income Q2 x MCS",
+%                              "Income Q3 x MCS",
+%                              "Income Q4 x MCS",
+%                              "Income Q5 x MCS"
+%                              ),
+%           no.space=TRUE, 
+%           omit.stat= c("rsq", "f", "ser", "aic", "ll"),
+%           float=FALSE    # needs begin table outside it
+% )
+% @
+% }
+% \end{frame}
+
 \begin{frame}
 \frametitle{Inequality}
-\framesubtitle{Quantifying the gradient -- inflation- and growth-adjusted income}
+\framesubtitle{Quantifying the gradient}
 \centering
-\resizebox{!}{4cm}{%
-<<FA_REGS_REAL, dependson='FA_SCORES_SELECT', results='asis'>>=
+\resizebox{12cm}{!}{%
+<<FA_REGS_REAL_MF, dependson='FA_SCORES', results='asis'>>=
 # regress scores on income
@@
 
-scores2plot$incq_real <- factor(ntile(scores2plot$faminc_real, 5))
-
-# get min, max
-minmax.real <-aggregate(scores2plot[,c("faminc_real")],list(scores2plot$incq_real), 
-            FUN = function(x) c(min = min(x, na.rm=T), max = max(x, na.rm=T)))
-minmax.real <- round(minmax.real[,-1],2)*1000
-
-r.ext.f.q <- lm(EXT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="F"))
-r.int.f.q <- lm(INT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="F"))
-
-r.ext.m.q <- lm(EXT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="M"))
-r.int.m.q <- lm(INT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="M"))
-
-stargazer(r.ext.m.q, r.int.m.q, r.ext.f.q, r.int.f.q,
-          column.labels=c("Males","Females"), column.separate=c(2,2),
-          dep.var.caption = "", multicolumn = FALSE,
-          omit=c("Constant"),
-          covariate.labels=c(
-                paste0("Income Q2 [",toString(minmax.real[2,1]),"-",toString(minmax.real[2,2])," £/w]"),
-                paste0("Income Q3 [",toString(minmax.real[3,1]),"-",toString(minmax.real[3,2])," £/w]"),
-                paste0("Income Q4 [",toString(minmax.real[4,1]),"-",toString(minmax.real[4,2])," £/w]"),
-                paste0("Income Q5 [",toString(minmax.real[5,1]),"+ £/w]"),
-                             "MCS dummy",
-                             "Income Q2 x MCS",
-                             "Income Q3 x MCS",
-                             "Income Q4 x MCS",
-                             "Income Q5 x MCS"
-                             ),
-          no.space=TRUE, 
-          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
-          float=FALSE    # needs begin table outside it
-)
-@
-}
-\end{frame}
-
-\begin{frame}
-\frametitle{Inequality}
-\framesubtitle{Quantifying the gradient -- males}
-\centering
-\resizebox{12cm}{4cm}{%
-<<FA_REGS_REAL_M, dependson='FA_SCORES', results='asis'>>=
-# regress scores on income
-library(dplyr)
-library(stargazer)
-
 r.ext.m.bcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="BCS"))
-r.ext2.m.bcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="BCS"))
 r.ext.m.mcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="MCS"))
-r.ext2.m.mcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="MCS"))
 r.int.m.bcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="BCS"))
-r.int2.m.bcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="BCS"))
 r.int.m.mcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="MCS"))
-r.int2.m.mcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="MCS"))
-
-stargazer(r.ext.m.bcs, r.ext2.m.bcs, r.int.m.bcs, r.int2.m.bcs, r.ext.m.mcs, r.ext2.m.mcs, r.int.m.mcs, r.int2.m.mcs,
-          column.labels=c("BCS","MCS"), column.separate=c(4,4),
+
+r.ext.f.bcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="BCS"))
+r.ext.f.mcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="MCS"))
+r.int.f.bcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="BCS"))
+r.int.f.mcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="MCS"))
+
+stargazer(r.ext.m.bcs, r.ext.m.mcs , r.int.m.bcs, r.int.m.mcs, r.ext.f.bcs, r.ext.f.mcs , r.int.f.bcs, r.int.f.mcs,
+          column.labels=c("BCS M","MCS M","BCS M","MCS M","BCS F","MCS F","BCS F","MCS F"),
           dep.var.caption = "", multicolumn = FALSE,
@@
                              "Income Q4",
-                             "Income Q5",
-                             "Birthweight",
-                             "Mat. Smoking in preg.",
-                             "Father years of schooling",
-                             "Mother years of schooling"
+                             "Income Q5"
                              ),
@@
 
-\begin{frame}
-\frametitle{Inequality}
-\framesubtitle{Quantifying the gradient -- females}
-\centering
-\resizebox{12cm}{4cm}{%
-<<FA_REGS_REAL_F, dependson='FA_SCORES', results='asis'>>=
-# regress scores on income
-library(dplyr)
-library(stargazer)
-
-r.ext.f.bcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="BCS"))
-r.ext2.f.bcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="BCS"))
-r.ext.f.mcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="MCS"))
-r.ext2.f.mcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="MCS"))
-r.int.f.bcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="BCS"))
-r.int2.f.bcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="BCS"))
-r.int.f.mcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="MCS"))
-r.int2.f.mcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="MCS"))
-
-stargazer(r.ext.f.bcs, r.ext2.f.bcs, r.int.f.bcs, r.int2.f.bcs, r.ext.f.mcs, r.ext2.f.mcs, r.int.f.mcs, r.int2.f.mcs,
-          column.labels=c("BCS","MCS"), column.separate=c(4,4),
-          dep.var.caption = "", multicolumn = FALSE,
-          omit=c("Constant"),
-          covariate.labels=c(
-                             "Income Q2",
-                             "Income Q3",
-                             "Income Q4",
-                             "Income Q5",
-                             "Birthweight",
-                             "Mat. Smoking in preg.",
-                             "Father years of schooling",
-                             "Mother years of schooling"
-                             ),
-          no.space=TRUE, 
-          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
-          float=FALSE    # needs begin table outside it
-)
-@
-}
-\end{frame}
+% 
+% \begin{frame}
+% \frametitle{Inequality}
+% \framesubtitle{Quantifying the gradient -- males}
+% \centering
+% \resizebox{12cm}{4cm}{%
+% <<FA_REGS_REAL_M, dependson='FA_SCORES', results='asis'>>=
+% # regress scores on income
+% library(dplyr)
+% library(stargazer)
+% 
+% 
+% r.ext.m.bcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="BCS"))
+% r.ext2.m.bcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="BCS"))
+% r.ext.m.mcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="MCS"))
+% r.ext2.m.mcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="MCS"))
+% r.int.m.bcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="BCS"))
+% r.int2.m.bcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="BCS"))
+% r.int.m.mcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="MCS"))
+% r.int2.m.mcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="MCS"))
+% 
+% stargazer(r.ext.m.bcs, r.ext2.m.bcs, r.int.m.bcs, r.int2.m.bcs, r.ext.m.mcs, r.ext2.m.mcs, r.int.m.mcs, r.int2.m.mcs,
+%           column.labels=c("BCS","MCS"), column.separate=c(4,4),
+%           dep.var.caption = "", multicolumn = FALSE,
+%           omit=c("Constant"),
+%           covariate.labels=c(
+%                              "Income Q2",
+%                              "Income Q3",
+%                              "Income Q4",
+%                              "Income Q5",
+%                              "Birthweight",
+%                              "Mat. Smoking in preg.",
+%                              "Father years of schooling",
+%                              "Mother years of schooling"
+%                              ),
+%           no.space=TRUE, 
+%           omit.stat= c("rsq", "f", "ser", "aic", "ll"),
+%           float=FALSE    # needs begin table outside it
+% )
+% @
+% }
+% \end{frame}
+% 
+% 
+% 
+% \begin{frame}
+% \frametitle{Inequality}
+% \framesubtitle{Quantifying the gradient -- females}
+% \centering
+% \resizebox{12cm}{4cm}{%
+% <<FA_REGS_REAL_F, dependson='FA_SCORES', results='asis'>>=
+% # regress scores on income
+% library(dplyr)
+% library(stargazer)
+% 
+% 
+% r.ext.f.bcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="BCS"))
+% r.ext2.f.bcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="BCS"))
+% r.ext.f.mcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="MCS"))
+% r.ext2.f.mcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="MCS"))
+% r.int.f.bcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="BCS"))
+% r.int2.f.bcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="BCS"))
+% r.int.f.mcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="MCS"))
+% r.int2.f.mcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="MCS"))
+% 
+% 
+% 
+% stargazer(r.ext.f.bcs, r.ext2.f.bcs, r.int.f.bcs, r.int2.f.bcs, r.ext.f.mcs, r.ext2.f.mcs, r.int.f.mcs, r.int2.f.mcs,
+%           column.labels=c("BCS","MCS"), column.separate=c(4,4),
+%           dep.var.caption = "", multicolumn = FALSE,
+%           omit=c("Constant"),
+%           covariate.labels=c(
+%                              "Income Q2",
+%                              "Income Q3",
+%                              "Income Q4",
+%                              "Income Q5",
+%                              "Birthweight",
+%                              "Mat. Smoking in preg.",
+%                              "Father years of schooling",
+%                              "Mother years of schooling"
+%                              ),
+%           no.space=TRUE, 
+%           omit.stat= c("rsq", "f", "ser", "aic", "ll"),
+%           float=FALSE    # needs begin table outside it
+% )
+% @
+% }
+% \end{frame}
 
@@
 \bi
-\item We derived a comparable scale of non-cognitive skills / behaviour problems
-\item We have shown that inequality (across the income dimension) has increased
+\item We derived a comparable scale of non-cognitive skills, based on maternal reports of behaviour problems
+\item We have shown that inequality in these skills at five years of age has increased between children born in 1970 and 2000
   \bi
@@
   \item Harmonise more cohorts (NCDS) and ages (10)
-  \item Extend analysis to the \emph{returns} to non-cognitive skills -- are they remunerated differently on the labour market?
+  \item Broaden analysis to the \emph{returns} to non-cognitive skills -- are these remunerated differently on the labour market for people born in different years?
   \ei
@@
 
-\begin{frame}
+\begin{frame}[noframenumbering]
 \frametitle{Inequality \hyperlink{meanci}{\beamerreturnbutton{Back}}}\label{loessext}
@@
 
-\begin{frame}
+\begin{frame}[noframenumbering]
 \frametitle{Inequality \hyperlink{meanci}{\beamerreturnbutton{Back}}}\label{loessint}
@@
 
-\begin{frame}
+\begin{frame}[noframenumbering]
 \frametitle{Inequality}

### 2018-05-19 codes/R/6_selectvars.R

@@ -7,7 +7,7 @@ decvarslist$ses <- c("scl10b")
 decvarslist$med <- c("mpsla5")
 decvarslist$mem <- c("mempl5")
 decvarslist$mch <- c("mothageb", "teenm", "mheight", "singlem", "ethn")
-decvarslist$prg <- c("parity", "firstb", "nprevst", "smkpr", "caesbirth", "gestaw", "preterm", "bwt", "lowbwt")
+decvarslist$prg <- c("parity", "firstb", "nprevst", "smkpr", "caesbirth", "preterm", "lbwt")
 
 # interaction between mother characteristics and initial conditions
 mchxbth <-  list(apply(
@@ -30,3 +30,22 @@ finaldata <- scores2plot[complete.cases(scores2plot[,decvars]),]
 
 # export to stata for Gelbach
 export(finaldata, paste0(dir_data, "finaldata.dta"))
+
+# labels
+declabs <- c(scl10b = "Parental social class (10)",
+             mpsla5 = "Mother post-compuls. education (5)",
+             mempl5 = "Mother employment status (5)",
+             mothageb = "Mother age (0)",
+             teenm = "Teen mother (0)",
+             mheight = "Mother height (0)",
+             singlem = "Unmarried mother (0)",
+             ethn = "Nonwhite ethnicity (0)",
+             parity = "Parity (0)",
+             firstb = "Firstborn child (0)",
+             nprevst = "Num previous stillbirths (0)",
+             smkpr = "Mother smoked in pregnancy (0)",
+             caesbirth = "Caesarean birth (0)",
+             preterm = "Preterm birth (0)",
+             lbwt = "(log) Birthweight (0)"
+)
+

### 2018-05-19 codes/R/slides/cohorts_tapes.Rnw
slides file

@@
 <<READ, cache=FALSE >>=
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/1_load.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/2_efa.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/3_mifa_invariance.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/3_mifa_final.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/4_scores.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/4_scores_compare.R")
-read_chunk("/Users/giacomomason/Documents/Projects/CohortStudies/codes/R/5_graphs.R")
-
-@
-
-<<PREAMBLE>>=
+library(here)
+read_chunk(here("codes/R/1_load.R"))
+read_chunk(here("codes/R/2_efa.R"))
+read_chunk(here("codes/R/3_inv_main.R"))
+read_chunk(here("codes/R/3_inv_aux.R"))
+read_chunk(here("codes/R/3_inv_fit.R"))
+read_chunk(here("codes/R/4_scores.R"))
+read_chunk(here("codes/R/4_scores_aux.R"))
+read_chunk(here("codes/R/5_graphs.R"))
+read_chunk(here("codes/R/6_selectvars.R"))
+read_chunk(here("codes/R/8_regs.R"))
+read_chunk(here("codes/R/getpars.R"))
+
+@
+
+<<PREAMBLE >>=
 @
@@
 
-\date[NCDS]{NCDS 60 Years of Our Lives \\ \today}
+\date[TAPES]{Trans-Atlantic Public Economics Seminar \\ \today}
 
@@
 \begin{frame}
-\frametitle{Harmonisation}
-\framesubtitle{Choice of common items}
+\frametitle{Harmonisation}\label{meantab_f}
+\framesubtitle{Choice of common items (females) \hyperlink{meantab_m}{\beamerbutton{Males}}}
 \centering
 \resizebox{!}{4cm}{
-<<ITEM_TEXT, dependson="MEANTABLE", results='asis'>>=
+<<MEANTAB_F, dependson="MEANTABLE", results='asis'>>=
 require(xtable)
 
-table <- meantab
-table <- table[,!names(meantab) %in% c("cats", "fac", "bcs.num", "mcs.num")] # drop if unneeded
+table <- meantab[[2]] # females (but just need item names)
+table <- table[,!names(table) %in% c("cats", "fac", "bcs.num", "mcs.num")] # drop if unneeded
 
@@
 
-<<BCSEFA>>=
-@
-
-<<BCSEFARES, dependson='BCSEFA'>>=
-@
-
-<<MCSEFA>>=
-@
-
-<<MCSEFARES, dependson='MCSEFA'>>=
-@
-
 \subsection{Exploratory Analysis}
@@
 
-<<FA_INVARIANCE>>=
-# estimate models of different levels of invariance
-@
-
-<<FA_FIT, dependson='FA_INVARIANCE'>>=
+<<FA_INV_MAIN>>=
+# main models only
+@
+
+<<FA_FIT>>=
 # assemble fit indices into matrices
@@
 
-<<FA_FINAL, dependson='FA_INVARIANCE'>>=
-# select final model set
-@
-
-<<FA_FINAL_PARS, dependson='FA_FINAL'>>=
-# get parameters of final model
-@
-
-<<CLEANUP, dependson='FA_FINAL'>>=
-@
-
-<<FA_SCORES, dependson='FA_FINAL'>>=
-# produce scores for different models
-@
-
-<<FA_SCORES_MERGE, dependson='LOAD_DATA'>>=
-# merge scores with other data
-@
-
-<<FA_SCORES_COMPARE, dependson='FA_SCORES'>>=
-# assemble and compare scores across different models
-@
-
-<<FA_SCORES_SELECT, dependson='FA_SCORES_COMPARE'>>=
-# select model to use for inequality analysis
+<<FA_SCORES_MAIN>>=
+# produce scores for main models
+@
+
+<<FA_SCORES_BIND>>=
+# bind scores with data
 @
@@
 \frametitle{Measurement Invariance}\label{mires}
-\framesubtitle{Results}
-
-\bi
-\item Restricting thresholds and loadings provides a good fit
+\framesubtitle{Results \hyperlink{measpars}{\beamerbutton{Measurement}}\hyperlink{lvpars}{\beamerbutton{Latent variable}}}
+
+\bi
+\item Restricting thresholds and loadings provides a good fit \hyperlink{fit}{\beamerbutton{Fit}}
   \bi \item We can compare the scale (variance) of the latent factors \ei
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% SCORES
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\frametitle{Factor scores}\label{density}
+\framesubtitle{Densities \hyperlink{rawhist}{\beamerbutton{Raw scores}}}
+\begin{figure}
+<<FACDENS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
+@
+\end{figure}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%% INEQUALITY
@@
 \begin{frame}
-\frametitle{Inequality}\label{meanci}
-\framesubtitle{Visualising the gradient \hyperlink{loessext}{\beamerbutton{EXT}}\hyperlink{loessint}{\beamerbutton{INT}}}
+\frametitle{Inequality}\label{ineq_uc}
+\framesubtitle{Unconditional inequality}
+\centering
+<<INEQ_STATS, dependson="FA_SCORES_BIND", results='asis'>>=
+
+# income 
+tb <- data.frame(
+  aggregate(scores2plot$faminc10_infl, by=list(scores2plot$cohort), FUN = function(x) quantile(x, c(.1, .25, .75, .9), na.rm=T))
+)
+tb$d9010 <- tb$x[,4]/tb$x[,1]
+tb$d7525 <- tb$x[,3]/tb$x[,2]
+
+tb.ext <- data.frame(
+  aggregate(scores2plot$EXT, by=list(scores2plot$cohortsex), FUN = function(x) quantile(x, c(.1, .25, .5, .75, .9), na.rm=T))
+)
+tb.ext$d9010 <- abs(tb.ext$x[,"90%"]-tb.ext$x[,"10%"])
+tb.ext$d7525 <- abs(tb.ext$x[,"75%"]-tb.ext$x[,"25%"])
+tb.ext$d9050 <- abs(tb.ext$x[,"90%"]-tb.ext$x[,"50%"])
+tb.ext$d5010 <- abs(tb.ext$x[,"50%"]-tb.ext$x[,"10%"])
+rownames(tb.ext) <- tb.ext$Group.1
+tb.ext <- t(tb.ext[,-c(1,2)]) # remove aggregated quantiles and transpose
+tb.ext <- tb.ext[,c("BCS.M","MCS.M","BCS.F","MCS.F")]
+
+tb.int <- data.frame(
+  aggregate(scores2plot$INT, by=list(scores2plot$cohortsex), FUN = function(x) quantile(x, c(.1, .25, .5, .75, .9), na.rm=T))
+)
+tb.int$d9010 <- abs(tb.int$x[,"90%"]-tb.int$x[,"10%"])
+tb.int$d7525 <- abs(tb.int$x[,"75%"]-tb.int$x[,"25%"])
+tb.int$d9050 <- abs(tb.int$x[,"90%"]-tb.int$x[,"50%"])
+tb.int$d5010 <- abs(tb.int$x[,"50%"]-tb.int$x[,"10%"])
+rownames(tb.int) <- tb.int$Group.1
+tb.int <- t(tb.int[,-c(1,2)]) # remove aggregated quantiles and transpose
+tb.int <- tb.int[,c("BCS.M","MCS.M","BCS.F","MCS.F")]
+
+# assemble and print
+table <- rbind(tb.ext,tb.int)
+table <- cbind(data.frame(as.matrix(rep(c("90-10", "75-25", "90-50", "50-10"),2))),
+               table)
+
+print(xtable(table,
+                    align="L{0cm}L{.15\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}",
+                    digits = 4
+                    ),
+             size="\\scriptsize",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0,4)),
+                              command = c(
+                              # header
+            "Measure & \\multicolumn{2}{c}{\\textbf{Males}} & \\multicolumn{2}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-3}  \\cmidrule(lr){4-5} 
+                       & BCS & MCS & BCS & MCS  \\\\
+                      \\textbf{Externalising} \\\\ ",
+                      "\\textbf{Internalising} \\\\ \\midrule "
+                                    )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = "\\textwidth",
+             floating = FALSE
+)
+
+@
+\end{frame}
+
+<<FACINEQ_MAKE>>=
+# make factor inequality graphs
+@
+
+\begin{frame}
+\frametitle{Inequality}\label{ineq_sc}
+\framesubtitle{Social class}
+\vspace{-.2cm}
 \begin{figure}
-<<FACINEQ, dev='tikz', fig.height=6.7, fig.width=10, out.width='\\linewidth'>>=
+<<FACINEQ_SC, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
+# plot
+fi <- fi_sc # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
 @
@@
 
-% \begin{frame}
-% \frametitle{Inequality}
-% \framesubtitle{Quantifying the gradient -- inflation- and growth-adjusted income}
-% \centering
-% \resizebox{!}{4cm}{%
-% <<FA_REGS_REAL, dependson='FA_SCORES_SELECT', results='asis'>>=
-% # regress scores on income
-% library(dplyr)
-% library(stargazer)
-% 
-% scores2plot$incq_real <- factor(ntile(scores2plot$faminc_real, 5))
-% 
-% # get min, max
-% minmax.real <-aggregate(scores2plot[,c("faminc_real")],list(scores2plot$incq_real), 
-%             FUN = function(x) c(min = min(x, na.rm=T), max = max(x, na.rm=T)))
-% minmax.real <- round(minmax.real[,-1],2)*1000
-% 
-% r.ext.f.q <- lm(EXT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="F"))
-% r.int.f.q <- lm(INT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="F"))
-% 
-% r.ext.m.q <- lm(EXT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="M"))
-% r.int.m.q <- lm(INT ~ as.factor(incq_real)*cohort, data=subset(scores2plot, sex=="M"))
-% 
-% stargazer(r.ext.m.q, r.int.m.q, r.ext.f.q, r.int.f.q,
-%           column.labels=c("Males","Females"), column.separate=c(2,2),
-%           dep.var.caption = "", multicolumn = FALSE,
-%           omit=c("Constant"),
-%           covariate.labels=c(
-%                 paste0("Income Q2 [",toString(minmax.real[2,1]),"-",toString(minmax.real[2,2])," £/w]"),
-%                 paste0("Income Q3 [",toString(minmax.real[3,1]),"-",toString(minmax.real[3,2])," £/w]"),
-%                 paste0("Income Q4 [",toString(minmax.real[4,1]),"-",toString(minmax.real[4,2])," £/w]"),
-%                 paste0("Income Q5 [",toString(minmax.real[5,1]),"+ £/w]"),
-%                              "MCS dummy",
-%                              "Income Q2 x MCS",
-%                              "Income Q3 x MCS",
-%                              "Income Q4 x MCS",
-%                              "Income Q5 x MCS"
-%                              ),
-%           no.space=TRUE, 
-%           omit.stat= c("rsq", "f", "ser", "aic", "ll"),
-%           float=FALSE    # needs begin table outside it
-% )
-% @
-% }
-% \end{frame}
-
-\begin{frame}
-\frametitle{Inequality}
-\framesubtitle{Quantifying the gradient}
-\centering
-\resizebox{12cm}{!}{%
-<<FA_REGS_REAL_MF, dependson='FA_SCORES', results='asis'>>=
-# regress scores on income
-library(dplyr)
-library(stargazer)
-
-r.ext.m.bcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="BCS"))
-r.ext.m.mcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="MCS"))
-r.int.m.bcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="BCS"))
-r.int.m.mcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="MCS"))
-
-r.ext.f.bcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="BCS"))
-r.ext.f.mcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="MCS"))
-r.int.f.bcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="BCS"))
-r.int.f.mcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="MCS"))
-
-stargazer(r.ext.m.bcs, r.ext.m.mcs , r.int.m.bcs, r.int.m.mcs, r.ext.f.bcs, r.ext.f.mcs , r.int.f.bcs, r.int.f.mcs,
-          column.labels=c("BCS M","MCS M","BCS M","MCS M","BCS F","MCS F","BCS F","MCS F"),
-          dep.var.caption = "", multicolumn = FALSE,
-          omit=c("Constant"),
-          covariate.labels=c(
-                             "Income Q2",
-                             "Income Q3",
-                             "Income Q4",
-                             "Income Q5"
+\begin{frame}
+\frametitle{Inequality}\label{ineq_sc}
+\framesubtitle{Age mother left education}
+\vspace{-.2cm}
+\begin{figure}
+<<FACINEQ_YS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
+# plot
+fi <- fi_ys # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{Inequality}\label{ineq_sc}
+\framesubtitle{Maternal education and employment}
+\vspace{-.2cm}
+\begin{figure}
+<<FACINEQ_MECS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=8.5, fig.width=13.5, out.width='\\linewidth'>>=
+# plot
+fi <- fi_mecs # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\end{figure}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% EXPLAINING EXT and INT
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+<<SELECTVARS>>=
+# define variables and formulas for regression and decomposition
+@
+
+<<REGS_DEP, dependson="SELECTVARS">>=
+# run regressions explaining EXT INT
+@
+
+\begin{frame}
+\frametitle{What explains behavioural skills?}
+
+write about variables used and groups
+
+\end{frame}
+
+\begin{frame}
+\frametitle{What explains behavioural skills?}
+\framesubtitle{Males}
+\begin{table}
+\centering
+\resizebox{!}{4cm}{
+<<REGS_M, results='asis', dependson="REGRESSIONS">>=
+
+# select models
+regs <- list(r_ext$BCS.M[["OLS"]],r_ext$MCS.M[["OLS"]],r_int$BCS.M[["OLS"]],r_int$MCS.M[["OLS"]])
+
+# robust SEs
+robust_se <- list()
+robust_p  <- list()
+for (i in 1:length(regs)) {
+  cov1              <- sandwich::vcovHC(regs[[i]], type = "HC1")
+  ct                <- lmtest::coeftest(regs[[i]], vcov = cov1)
+  robust_se[[i]] <- ct[, 2]
+  robust_p[[i]] <- ct[, 4]
+}
+
+texreg::texreg(regs,
+               stars = c(0.1, 0.05, 0.01), digits = 3,
+               override.se = robust_se, override.pvalues = robust_p,
+               single.row = TRUE, scriptsize = TRUE,
+               caption = "", label = "",
+               omit.coef = "region|Inter",
+               groups = grouplabs,
+               custom.coef.names = varlabs,
+               custom.model.names = c("EXT BCS", "EXT MCS", "INT BCS", "INT MCS"),
+               table = FALSE
+               )
+
+@
+}
+\end{table}
+\end{frame}
+
+\begin{frame}
+\frametitle{What explains behavioural skills?}
+\framesubtitle{Females}
+\begin{table}
+\centering
+\resizebox{!}{4cm}{
+<<REGS_F, results='asis', dependson="REGRESSIONS">>=
+
+# select models
+regs <- list(r_ext$BCS.F[["OLS"]],r_ext$MCS.F[["OLS"]],r_int$BCS.F[["OLS"]],r_int$MCS.F[["OLS"]])
+
+# robust SEs
+robust_se <- list()
+robust_p  <- list()
+for (i in 1:length(regs)) {
+  cov1              <- sandwich::vcovHC(regs[[i]], type = "HC1")
+  ct                <- lmtest::coeftest(regs[[i]], vcov = cov1)
+  robust_se[[i]] <- ct[, 2]
+  robust_p[[i]] <- ct[, 4]
+}
+
+texreg::texreg(regs,
+               stars = c(0.1, 0.05, 0.01), digits = 3,
+               override.se = robust_se, override.pvalues = robust_p,
+               single.row = TRUE, scriptsize = TRUE,
+               caption = "", label = "",
+               omit.coef = "region|Inter",
+               groups = grouplabs,
+               custom.coef.names = varlabs,
+               custom.model.names = c("EXT BCS", "EXT MCS", "INT BCS", "INT MCS"),
+               table = FALSE
+               )
+
+@
+}
+\end{table}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% GELBACH DECOMPOSITION
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\frametitle{Decomposition (EXT)}
+\framesubtitle{Explaining the education gradient}
+\centering
+\scriptsize
+\resizebox{\textwidth}{!}{
+\input{../../../tables/med_gelb_ext}
+}
+\end{frame}
+
+\begin{frame}
+\frametitle{Decomposition (INT)}
+\framesubtitle{Explaining the education gradient}
+\centering
+\scriptsize
+\resizebox{\textwidth}{!}{
+\input{../../../tables/med_gelb_int}
+}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% EFFECTS ON LATER OUTCOMES
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+<<REGS_OUTC, dependson="SELECTVARS">>=
+# run regressions explaining adult outcomes
+@
+
+<<REGS_OUTC_TAB, dependson="REGS_OUTC">>=
+# make tables
+@
+
+\begin{frame}
+\frametitle{Outcomes}
+\framesubtitle{Adolescence (BCS)}
+\centering
+\resizebox{\textwidth}{!}{
+<<REGS_OUTC_BCS1, results='asis', dependson="REGRESSIONS">>=
+
+# collapse to matrix (first 4 elements of bcs_outctab are adolescence)
+bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab[c(1:4)],matrix,ncol=11,byrow=FALSE))
+table <- bcs_outctab_mat
+
+print(xtable(table,
+                    align="R{0cm}L{.22\\textwidth}C{.05\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.05\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}",
+                    digits = 3
                     ),
-          no.space=TRUE,
-          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
-          float=FALSE    # needs begin table outside it
+             size="\\scriptsize",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0)),
+                              command = c(
+                              # header
+            " & \\multicolumn{5}{c}{\\textbf{Males}} & \\multicolumn{5}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-6}  \\cmidrule(lr){7-11}
+& Mean & \\multicolumn{4}{c}{Coefficients} & Mean & \\multicolumn{4}{c}{Coefficients} \\\\ \\cmidrule(lr){3-6}  \\cmidrule(lr){8-11}"
+            )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = "\\textwidth",
+             floating = FALSE
 )
@@
 
-% 
-% \begin{frame}
-% \frametitle{Inequality}
-% \framesubtitle{Quantifying the gradient -- males}
-% \centering
-% \resizebox{12cm}{4cm}{%
-% <<FA_REGS_REAL_M, dependson='FA_SCORES', results='asis'>>=
-% # regress scores on income
-% library(dplyr)
-% library(stargazer)
-% 
-% 
-% r.ext.m.bcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="BCS"))
-% r.ext2.m.bcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="BCS"))
-% r.ext.m.mcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="MCS"))
-% r.ext2.m.mcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="MCS"))
-% r.int.m.bcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="BCS"))
-% r.int2.m.bcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="BCS"))
-% r.int.m.mcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="M" & cohort=="MCS"))
-% r.int2.m.mcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="M" & cohort=="MCS"))
-% 
-% stargazer(r.ext.m.bcs, r.ext2.m.bcs, r.int.m.bcs, r.int2.m.bcs, r.ext.m.mcs, r.ext2.m.mcs, r.int.m.mcs, r.int2.m.mcs,
-%           column.labels=c("BCS","MCS"), column.separate=c(4,4),
-%           dep.var.caption = "", multicolumn = FALSE,
-%           omit=c("Constant"),
-%           covariate.labels=c(
-%                              "Income Q2",
-%                              "Income Q3",
-%                              "Income Q4",
-%                              "Income Q5",
-%                              "Birthweight",
-%                              "Mat. Smoking in preg.",
-%                              "Father years of schooling",
-%                              "Mother years of schooling"
-%                              ),
-%           no.space=TRUE, 
-%           omit.stat= c("rsq", "f", "ser", "aic", "ll"),
-%           float=FALSE    # needs begin table outside it
-% )
-% @
-% }
-% \end{frame}
-% 
-% 
-% 
-% \begin{frame}
-% \frametitle{Inequality}
-% \framesubtitle{Quantifying the gradient -- females}
-% \centering
-% \resizebox{12cm}{4cm}{%
-% <<FA_REGS_REAL_F, dependson='FA_SCORES', results='asis'>>=
-% # regress scores on income
-% library(dplyr)
-% library(stargazer)
-% 
-% 
-% r.ext.f.bcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="BCS"))
-% r.ext2.f.bcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="BCS"))
-% r.ext.f.mcs <- lm(EXT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="MCS"))
-% r.ext2.f.mcs <- lm(EXT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="MCS"))
-% r.int.f.bcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="BCS"))
-% r.int2.f.bcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="BCS"))
-% r.int.f.mcs <- lm(INT ~ as.factor(incq), data=subset(scores2plot, sex=="F" & cohort=="MCS"))
-% r.int2.f.mcs <- lm(INT ~ as.factor(incq) + bwt + as.factor(smkpr) + ysch_fath + ysch_moth, data=subset(scores2plot, sex=="F" & cohort=="MCS"))
-% 
-% 
-% 
-% stargazer(r.ext.f.bcs, r.ext2.f.bcs, r.int.f.bcs, r.int2.f.bcs, r.ext.f.mcs, r.ext2.f.mcs, r.int.f.mcs, r.int2.f.mcs,
-%           column.labels=c("BCS","MCS"), column.separate=c(4,4),
-%           dep.var.caption = "", multicolumn = FALSE,
-%           omit=c("Constant"),
-%           covariate.labels=c(
-%                              "Income Q2",
-%                              "Income Q3",
-%                              "Income Q4",
-%                              "Income Q5",
-%                              "Birthweight",
-%                              "Mat. Smoking in preg.",
-%                              "Father years of schooling",
-%                              "Mother years of schooling"
-%                              ),
-%           no.space=TRUE, 
-%           omit.stat= c("rsq", "f", "ser", "aic", "ll"),
-%           float=FALSE    # needs begin table outside it
-% )
-% @
-% }
-% \end{frame}
+\begin{frame}
+\frametitle{Outcomes}
+\framesubtitle{Adulthood (BCS)}
+\centering
+\resizebox{\textwidth}{!}{
+<<REGS_OUTC_BCS2, results='asis', dependson="REGRESSIONS">>=
+
+# collapse to matrix (last 4 elements of bcs_outctab are adulthood)
+bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab[c(5:8)],matrix,ncol=11,byrow=FALSE))
+table <- bcs_outctab_mat
+
+print(xtable(table,
+                    align="R{0cm}L{.22\\textwidth}C{.05\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.05\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}",
+                    digits = 3
+                    ),
+             size="\\scriptsize",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0)),
+                              command = c(
+                              # header
+            " & \\multicolumn{5}{c}{\\textbf{Males}} & \\multicolumn{5}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-6}  \\cmidrule(lr){7-11}
+& Mean & \\multicolumn{4}{c}{Coefficients} & Mean & \\multicolumn{4}{c}{Coefficients} \\\\ \\cmidrule(lr){3-6}  \\cmidrule(lr){8-11}"
+            )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = "\\textwidth",
+             floating = FALSE
+)
+@
+}
+\end{frame}
+
+\begin{frame}
+\frametitle{Outcomes}
+\framesubtitle{Adolescence (MCS)}
+\centering
+\resizebox{\textwidth}{!}{
+<<REGS_OUTC_MCS, results='asis', dependson="REGRESSIONS">>=
+
+# collapse to matrix
+mcs_outctab_mat <- do.call(rbind,lapply(mcs_outctab,matrix,ncol=11,byrow=FALSE))
+table <- mcs_outctab_mat
+
+print(xtable(table,
+                    align="R{0cm}L{.22\\textwidth}C{.05\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.05\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}",
+                    digits = 3
+                    ),
+             size="\\scriptsize",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0)),
+                              command = c(
+                              # header
+            " & \\multicolumn{5}{c}{\\textbf{Males}} & \\multicolumn{5}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-6}  \\cmidrule(lr){7-11}
+& Mean & \\multicolumn{4}{c}{Coefficients} & Mean & \\multicolumn{4}{c}{Coefficients} \\\\ \\cmidrule(lr){3-6}  \\cmidrule(lr){8-11}"
+            )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = "\\textwidth",
+             floating = FALSE
+)
+@
+}
+\end{frame}
 
@@
 
-% \begin{frame}
-% \frametitle{Returns}
-% \framesubtitle{Education returns to behaviour skills in BCS}
-% \centering
-% \resizebox{!}{4cm}{%
-% <<FA_RETURNS, dependson='FA_SCORES_SELECT', results='asis'>>=
-% # regress scores on income
-% library(dplyr)
-% library(stargazer)
-% 
-% # get BCS data and make dummy for higher education
-% scoresBCS <- scores2plot[scores2plot$cohort=="BCS",]
-% scoresBCS$highered <- as.numeric(scoresBCS$hinvq00>3)
-% scoresBCS[is.na(scoresBCS$hinvq00), "highered"] <- NA
-% 
-% # females
-% r.nvq.f <- list()
-% r.nvq.f[[1]] <- lm(highered ~ EXT + INT, data=subset(scoresBCS, sex=="F"))
-% r.nvq.f[[2]] <- lm(highered ~ EXT + INT + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="F"))
-% r.nvq.f[[3]] <- lm(highered ~ EXT + INT + factor(incq), data=subset(scoresBCS, sex=="F"))
-% r.nvq.f[[4]] <- lm(highered ~ EXT + INT + ysch_moth + ysch_fath, data=subset(scoresBCS, sex=="F"))
-% r.nvq.f[[5]] <- lm(highered ~ EXT + INT + factor(incq) + ysch_moth + ysch_fath + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="F"))
-% 
-% # females
-% r.nvq.m <- list()
-% r.nvq.m[[1]] <- lm(highered ~ EXT + INT, data=subset(scoresBCS, sex=="M"))
-% r.nvq.m[[2]] <- lm(highered ~ EXT + INT + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="M"))
-% r.nvq.m[[3]] <- lm(highered ~ EXT + INT + factor(incq), data=subset(scoresBCS, sex=="M"))
-% r.nvq.m[[4]] <- lm(highered ~ EXT + INT + ysch_moth + ysch_fath, data=subset(scoresBCS, sex=="M"))
-% r.nvq.m[[5]] <- lm(highered ~ EXT + INT + factor(incq) + ysch_moth + ysch_fath + cog1 + cog2 + cog3, data=subset(scoresBCS, sex=="M"))
-% 
-% stargazer(r.nvq.m[[1]], r.nvq.m[[2]], r.nvq.m[[3]], r.nvq.m[[4]], r.nvq.m[[5]],
-%           r.nvq.f[[1]], r.nvq.f[[2]], r.nvq.f[[3]], r.nvq.f[[4]], r.nvq.f[[5]],
-%           dep.var.caption = "Completed Higher Ed. at 30",
-%           column.labels=c("Males","Females"), column.separate=c(5,5),
-%           omit=c("Constant"),
-%           covariate.labels=c("Externalising skill at 5",
-%                              "Internalising skill at 5",
-%                              "EPVT Z at 5",
-%                              "Human figure drawing Z at 5",
-%                              "Copy design Z at 5",
-%                              "Family inc. at 10 Q2",
-%                              "Family inc. at 10 Q3",
-%                              "Family inc. at 10 Q4",
-%                              "Family inc. at 10 Q5",
-%                              "Mother's years of schooling",
-%                              "Father's years of schooling"
-%                              ),
-%           no.space=TRUE, 
-%           omit.stat= c("rsq", "f", "ser", "aic", "ll"),
-%           float=FALSE    # needs begin table outside it
-% )
-% @
-% }
-% \end{frame}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \end{frame}
+
+\begin{frame}
+\frametitle{Harmonisation}\label{meantab_m}
+\framesubtitle{Choice of common items (males) \hyperlink{meantab_f}{\beamerreturnbutton{Females}}}
+\centering
+\resizebox{!}{4cm}{
+<<MEANTAB_M, dependson="MEANTABLE", results='asis'>>=
+require(xtable)
+
+table <- meantab[[1]] # males
+table <- table[,!names(table) %in% c("cats", "fac", "bcs.num", "mcs.num")] # drop if unneeded
+
+print(xtable(table,
+                    align="L{0cm}C{.04\\textwidth}L{.4\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}L{.4\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
+                    digits = 1
+                    ), 
+             size="\\small",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0)),
+                              command = c(
+                              # header
+                                    paste(" & 
+                                              \\multicolumn{4}{c}{\\textbf{BCS}} &
+                                              \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9} 
+                                            Item &
+                                              Text & Cert. Appl. & Smt. Appl. & Appl. &
+                                              Text & Cert. True & Smt. True & True \\\\ \\midrule
+                                          "
+                                            , sep="")
+                                    )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = ".8\\textwidth",
+             floating = FALSE
+)
+
+@
+} %scalebox
+\end{frame}
+
+<<BCSEFA>>=
+@
+
+<<BCSEFARES, dependson='BCSEFA'>>=
+@
+
+<<MCSEFA>>=
+@
+
+<<MCSEFARES, dependson='MCSEFA'>>=
+@
 
@@
 \ei
-
 \end{frame}
@@
 \begin{frame}[noframenumbering]
-\frametitle{MI}
-\framesubtitle{Fit comparison}
-\begin{table}[ht]
-\centering
-\small
+\frametitle{MI}\label{fit}
+\framesubtitle{Fit comparison \hyperlink{mires}{\beamerreturnbutton{Back}}}
+\centering
+\resizebox{\textwidth}{!}{
 <<FA_FIT_RES, dependson="FA_FIT", results='asis'>>=
@@
 
-table <- rbind(afitab[[3]],afitab[[4]],afitab[[7]])
+table <- rbind(afitab[[1]],afitab[[2]])
 
 print(xtable(table,
-                    align="L{0cm}L{.20\\textwidth}C{.03\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}",
+                    align="L{0cm}L{.20\\textwidth}C{.03\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}",
                     digits = 4
@@
              booktabs = T,
-             add.to.row = list(pos = c(list(0,3,6)),
+             add.to.row = list(pos = c(list(0,3)),
                               command = c(
@@
             "Model & Num. params & $\\chi^2$ & RMSEA & MFI & CFI & G-hat & \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ \\midrule
-                       \\textbf{Males only} \\\\ ",
-                      "\\newline \\textbf{Females only} \\\\ \\midrule ",
-                      "\\newline \\textbf{Males and females} \\\\ \\midrule "
+                       \\textbf{All ages} \\\\",
+                      "\\newline \\textbf{59-61 months sample} \\\\ \\midrule "
                                     )),
@@
 @
-\end{table}
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% ROBUSTNESS APPENDIX
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+}
+\end{frame}
+
+<<FUN_GETPARS>>=
+# load function that gets parameters
+@
 
 \begin{frame}[noframenumbering]
-\frametitle{Age distribution \hyperlink{mires}{\beamerreturnbutton{Back}}}\label{agehist}
-\begin{figure}
-\begin{center}
-<<AGEHIST, dev='tikz', fig.height=6, fig.width=6, out.width='.6\\linewidth', cache=TRUE>>=
-@
-\end{center}
-\caption{Distribution of age in the BCS and MCS sample}
-\end{figure}
-\end{frame}
-
-\begin{frame}[noframenumbering]
-\frametitle{MI}\label{fit4}
-\framesubtitle{Fit comparison -- 4 group models \hyperlink{mires}{\beamerreturnbutton{Back}}}
-
-\begin{table}[ht]
-\centering
-\small
-<<FA_FIT_RES4, dependson="FA_FIT", results='asis'>>=
-require(xtable)
-
-table <- rbind(afitab[[7]],afitab[[8]],afitab[[9]])
-
-print(xtable(table,
-                    align="L{0cm}L{.20\\textwidth}C{.03\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}C{.05\\textwidth}",
-                    digits = 4
+\frametitle{MI}\label{measpars}
+\framesubtitle{Estimated measurement parameters \hyperlink{mires}{\beamerreturnbutton{Back}}}
+<<RES_FA_PAR, dependson="FA_FIT", results='asis'>>=
+
+table <- getmeaspars(finalmod, groups = 4, mode = "tl")
+
+# sequences to add rows
+rws <- seq(2, (nrow(table)-1), by = 1) # start from row 2 (exclude model titles)
+rwsl <- as.list(rws)
+
+print.xtable(xtable(table,
+              align="C{0cm}C{.04\\textwidth}C{.04\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}",
+              digits = 3
                     ),
-             size="\\tiny",
+             size="\\scriptsize",
              include.rownames=F,
@@
              booktabs = T,
-             add.to.row = list(pos = c(list(0,3,6)),
+             add.to.row = list(pos = c(list(0)),
                               command = c(
                               # header
-            "Model & Num. params & $\\chi^2$ & RMSEA & MFI & CFI & G-hat & \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ \\midrule
-                      \\newline \\textbf{(7) not age adjusted} \\\\",
-                      "\\newline \\textbf{(8) age adjusted} \\\\\\midrule ",
-                      "\\newline \\textbf{(9) 59-62m range} \\\\ \\midrule "
+                                          " &
+                                            & \\textbf{Loadings}
+                                            & \\multicolumn{2}{c}{\\textbf{Thresholds}}
+                                            & \\multicolumn{3}{c}{\\textbf{Intercepts}}
+                                            & \\multicolumn{3}{c}{\\textbf{Variances}}
+                                            \\\\ \\cmidrule(lr){3-3} \\cmidrule(lr){4-5} \\cmidrule(lr){6-8} \\cmidrule(lr){9-11}
+
+                                            &
+                                            & $\\lambda$
+                                            & $\\tau_1$
+                                            & $\\tau_2$
+                                            & \\multicolumn{3}{c}{$\\nu$}
+                                            & \\multicolumn{3}{c}{$\\text{diag}(\\Psi)$} \\\\
+
+                                            Measure
+                                            & Factor
+                                            & All
+                                            & All
+                                            & All
+                                            & BCS F
+                                            & MCS M
+                                            & MCS F
+                                            & BCS F
+                                            & MCS M
+                                            & MCS F
+                                            \\\\ "
                       )),
@@
              sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabular", width = "\\textwidth",
+             tabular.environment = "tabularx", width = "\\textwidth",
              floating = FALSE
 )
+@
+\end{frame}
+
+\begin{frame}[noframenumbering]
+\frametitle{MI}\label{lvpars}
+\framesubtitle{Estimated latent means and variances \hyperlink{mires}{\beamerreturnbutton{Back}}}
+\begin{table}[ht]
+\small
+\centering
+<<RES_FA_LV, dependson="FA_FIT", results='asis'>>=
+require(xtable)
+
+lvtab <- getlvpars(finalmod, groups = 4)
+table <- rbind(lvtab$Males,lvtab$Females)
+
+print.xtable(xtable(table,
+                     align="C{0cm}L{.13\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.07\\textwidth}",
+                    digits = 3
+                    ),
+             size="\\small",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0,2,4)),
+                              command = c(
+                              # header
+                                   "\\toprule
+                                    & \\multicolumn{4}{c}{\\textbf{BCS}}
+                                    & \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
+                                    & Mean & \\multicolumn{2}{c}{Covariance} & Corr. & Mean & \\multicolumn{2}{c}{Covariance} & Corr. \\\\ \\midrule
+                      \\textbf{Males} \\\\",
+                      "\\textbf{Females} \\\\",
+                      "\\bottomrule"
+                                    )),
+             hline.after = NULL,  # remove default booktabs lines
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabularx", width = "\\textwidth",
+             floating = FALSE
+)
 
@@
 \end{table}
-
-\end{frame}
-
-\begin{frame}[noframenumbering]
-\frametitle{Factor scores vs. Sum scores}
+\end{frame}
+
+\begin{frame}
+\frametitle{Factor scores}\label{rawhist}
+\framesubtitle{Raw scores \hyperlink{density}{\beamerreturnbutton{Back}}}
 \begin{figure}
-<<FACRAW, dev='tikz', fig.height=7, fig.width=7, out.width='.65\\linewidth'>>=
+<<RAWHIST, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
 @
@@
 
-<<LOESS_REAL, dependson='FA_SCORES_SELECT'>>=
-@
-
-\begin{frame}[noframenumbering]
-\frametitle{Inequality \hyperlink{meanci}{\beamerreturnbutton{Back}}}\label{loessext}
-\framesubtitle{Visualising the gradient -- inflation- and growth-adjusted income}
-\begin{figure}
-<<LOESS_REAL_EXT, dependson='LOESS_REAL', dev='tikz', fig.height=8, fig.width=12, out.width='.95\\linewidth'>>=
-ploess.real.ext
-@
-\end{figure}
-\end{frame}
-
-\begin{frame}[noframenumbering]
-\frametitle{Inequality \hyperlink{meanci}{\beamerreturnbutton{Back}}}\label{loessint}
-\framesubtitle{Visualising the gradient -- inflation- and growth-adjusted income}
-\begin{figure}
-<<LOESS_REAL_INT, dependson='LOESS_REAL', dev='tikz', fig.height=8, fig.width=12, out.width='.95\\linewidth'>>=
-ploess.real.int
-@
-\end{figure}
-\end{frame}
-
-<<LOESS_INFL, dependson="FA_SCORES_SELECT">>=
-@
-
-\begin{frame}[noframenumbering]
-\frametitle{Inequality}
-\framesubtitle{Visualising the gradient -- inflation-adjusted income}
-\begin{figure}
-<<LOESS_INFL_EXT, dependson='LOESS_INFL', dev='tikz', fig.height=8, fig.width=12, out.width='.95\\linewidth'>>=
-ploess.infl.ext
-@
-\end{figure}
-\end{frame}
-
-\begin{frame}[noframenumbering]
-\frametitle{Inequality}
-\framesubtitle{Visualising the gradient -- inflation-adjusted income}
-\begin{figure}
-<<LOESS_INFL_INT, dependson='LOESS_INFL', dev='tikz', fig.height=8, fig.width=12, out.width='.95\\linewidth'>>=
-ploess.infl.int
-@
-\end{figure}
-\end{frame}
-
-\begin{frame}[noframenumbering]
-\frametitle{Inequality}
-\framesubtitle{Quantifying the gradient -- inflation-adjusted income}
-\centering
-\resizebox{!}{4cm}{%
-<<FA_REGS_INFL, dependson='FA_SCORES_SELECT', results='asis'>>=
-# regress scores on income
-library(dplyr)
-library(stargazer)
-
-scores2plot$incq_infl <- factor(ntile(scores2plot$faminc_infl, 5))
-
-r.ext.f.c <- lm(EXT ~ faminc_infl*cohort, data=subset(scores2plot, sex=="F"))
-r.ext.f.q <- lm(EXT ~ incq_infl*cohort, data=subset(scores2plot, sex=="F"))
-r.int.f.c <- lm(INT ~ faminc_infl*cohort, data=subset(scores2plot, sex=="F"))
-r.int.f.q <- lm(INT ~ incq_infl*cohort, data=subset(scores2plot, sex=="F"))
-
-r.ext.m.c <- lm(EXT ~ faminc_infl*cohort, data=subset(scores2plot, sex=="M"))
-r.ext.m.q <- lm(EXT ~ incq_infl*cohort, data=subset(scores2plot, sex=="M"))
-r.int.m.c <- lm(INT ~ faminc_infl*cohort, data=subset(scores2plot, sex=="M"))
-r.int.m.q <- lm(INT ~ incq_infl*cohort, data=subset(scores2plot, sex=="M"))
-
-stargazer(r.ext.m.q, r.int.m.q, r.ext.f.q, r.int.f.q,
-          column.labels=c("Males","Females"), column.separate=c(2,2),
-          dep.var.caption = "", multicolumn = FALSE,
-          omit=c("Constant"),
-          covariate.labels=c(
-                             "Income Q2",
-                             "Income Q3",
-                             "Income Q4",
-                             "Income Q5",
-                             "MCS dummy",
-                             "Income Q2 x MCS",
-                             "Income Q3 x MCS",
-                             "Income Q4 x MCS",
-                             "Income Q5 x MCS"
-                             ),
-          no.space=TRUE, 
-          omit.stat= c("rsq", "f", "ser", "aic", "ll"),
-          float=FALSE    # needs begin table outside it
-)
-@
-}
-\end{frame}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% ROBUSTNESS APPENDIX
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% \begin{frame}[noframenumbering]
+% \frametitle{Age distribution \hyperlink{mires}{\beamerreturnbutton{Back}}}\label{agehist}
+% \begin{figure}
+% \begin{center}
+% <<AGEHIST, dev='tikz', fig.height=6, fig.width=6, out.width='.6\\linewidth', cache=TRUE>>=
+% @
+% \end{center}
+% \caption{Distribution of age in the BCS and MCS sample}
+% \end{figure}
+% \end{frame}
 

### 2018-05-19 codes/R/8_regs.R
add tobit, format outcomes tables

@@ -9,26 +9,41 @@ pplus <- function(x) paste0(x, collapse="+")
 decvars_reg <- c(decvars, c("region"))
 decvars_int_reg <- c(decvars_int, c("region"))
 
+# labels for groups (texreg)
+grouplabs <- list("Social class (10)" = 1:4, 
+                  "Maternal education (5)" = 5, 
+                  "Maternal employment (5)" = 6:7,
+                  "Maternal background (birth)" = 8:12,
+                  "Pregnancy" = 13:18)
+# labels for variables (texreg)
+varlabs <- c("IIIM", "IIINM", "I/II", "Other",
+             "Post-compulsory",
+             "Part time", "Full time",
+             "Age", "Teen", "Height", "Unmarried", "Nonwhite",
+             "Parity", "Firstborn", "Num prev. stillbirths",
+             "Mother smoked in pregnancy","Caesarean birth","Preterm birth",
+             "(log) Birthweight (0)"
+)
+
 # formulas
-form_ext <- list()
-form_ext[[1]] <- as.formula("EXT ~ mpsla5")
-form_ext[[2]] <- as.formula(paste("EXT ~ ", pplus(decvars_reg)))
-#form_ext[[3]] <- as.formula(paste("EXT ~ ", pplus(decvars_int_reg)))
-form_int <- list()
-form_int[[1]] <- as.formula("INT ~ mpsla5")
-form_int[[2]] <- as.formula(paste("INT ~ ", pplus(decvars_reg)))
-#form_int[[3]] <- as.formula(paste("INT ~ ", pplus(decvars_int_reg)))
+form_ext <- as.formula(paste("EXT ~ ", pplus(decvars_reg)))
+form_int <- as.formula(paste("INT ~ ", pplus(decvars_reg)))
 
 # regressions
-r.ext <- list()
-r.int <- list()
+r_ext <- list()
+r_int <- list()
 for (cs in c("BCS.M", "MCS.M", "BCS.F", "MCS.F")) {
-  r.ext[[cs]] <- list()
-  r.int[[cs]] <- list()
-  for (i in 1:2) {
-    r.ext[[cs]][[i]]  <- lm(form_ext[[i]],  data=subset(scores2plot, cohortsex==cs))
[…]

### 2018-05-22 codes/R/6_selectvars.R
trimmed some regressors

@@ -3,11 +3,10 @@
 
 # select variables for decomposition, in groups
 decvarslist <- list()
-decvarslist$ses <- c("scl10b")
 decvarslist$med <- c("mpsla5")
 decvarslist$mem <- c("mempl5")
-decvarslist$mch <- c("mothageb", "teenm", "mheight", "singlem", "ethn")
-decvarslist$prg <- c("parity", "firstb", "nprevst", "smkpr", "caesbirth", "preterm", "lbwt")
+decvarslist$mch <- c("mothageb", "mheight", "singlem", "ethn")
+decvarslist$prg <- c("parity", "firstb", "nprevst", "smkpr", "preterm", "lbwt")
 
 # interaction between mother characteristics and initial conditions
 mchxbth <-  list(apply(
@@ -16,7 +15,7 @@ mchxbth <-  list(apply(
 decvarslist_int <- c(decvarslist, mchxbth)
 
 # names of the groups
-decvarsgroups <- c("Social Class", "Mat. Education", "Mat. Employment", "Mat. Charact.", "Pregnancy/Birth")
+decvarsgroups <- c("Mat. Education", "Mat. Employment", "Mat. Charact.", "Pregnancy/Birth")
 
 # make into vector
 decvars <- unname(unlist(decvarslist))

### 2018-05-22 codes/R/slides/cohorts_tapes.Rnw
slide update

@@
 read_chunk(here("codes/R/6_selectvars.R"))
-read_chunk(here("codes/R/8_regs.R"))
+read_chunk(here("codes/R/8_regs_dep.R"))
+read_chunk(here("codes/R/8_regs_outc.R"))
 read_chunk(here("codes/R/getpars.R"))
@@
 \bi
-\item Non-cognitive (socio-emotional, behavioural) skills developed during childhood are important for lifetime outcomes
+\item Socio-emotional skills developed during childhood are important for lifetime outcomes
 \item Socioeconomic gradients in these skills are persistent 
+\item {\color{red} IMPROVE INTRO...}
+\ei
+
+\end{frame}
+
+\begin{frame}
+\frametitle{This study}
+\framesubtitle{Research questions}
+\bi
+\item Usually, socioemotional constructs are measured using validated scales
+    \bi \item e.g. Rutter, BSAG, SDQ \ei
 \pause
-\item \textbf{Q1}: Can we derive \textbf{comparable measures} across populations?
-    \bi
-    \item Usually, behavioural constructs are measured via scales
-    \item But does the same scale measure the same constructs in the same way:
+~\\    
+\item \textbf{Q1}: Can we derive \textbf{comparable measures} of socioemotional skills across populations?
+    \bi
+    \item Does the same instrument measure the same constructs in the same way:
       \bi 
@@
       \ei
-    \item And what if only different scales are available?
+    \item And what if only different instruments are available?
     \ei
 \pause
-\item \textbf{Q2}: Can we use these measures to document \textbf{changes in socioeconomic inequality} in skills across populations?
-\ei
-
+~\\    
+\item \textbf{Q2}: \textbf{How many latent dimensions} are underlying these instruments?
+    \bi
+    \item Can they be simply summarised into a single score?
+    \ei
+\ei
 \end{frame}
@@
 \frametitle{This study}
-
-\bi
-\item Focus on behavioural problems in two British cohorts
+\framesubtitle{Overview}
+
+\bi
+\item Focus on socioemotional skills in two British cohorts
   \bi \item British Cohort Study (BCS, b. 1970)
@@
 \ei
-\vspace{6pt} \pause
+~\\ \pause
+\textbf{Contribution}
+\bi
+\item Construct a common 11-item scale, starting from different survey instruments
+\item Investigate its psychometric properties, and find
   \bnm 
-\item Construct a common 11-item scale of non-cognitive skills, starting from different instruments
-\item Investigate its psychometric properties
-\item Use scores from the new scale to document changes in socioeconomic gradients between cohorts 30 years apart
+  \item \textbf{\textit{Comparability}} in terms of scale (but not location)
+  \item \textbf{\textit{Bidimensionality}} -- Externalising and Internalising
   \enm
-\vspace{6pt} \pause
-\bi
-\item \textbf{Key findings}:
+\ei
+~\\ \pause
+
+\textbf{Key findings}
   \bi
@@
   \item Sharpest change for boys rather than girls
-  \item Mostly driven by improvements at the top of the distribution
-  \ei
+  \item {\color{red} OTHER RESULTS}
   \ei
@@
 
-\section{Harmonised scale}
-
-\subsection{Data}
+\section{Data}
+
+\begin{frame}
+\frametitle{Data}
+\framesubtitle{The UK birth cohort studies}
+
+\bi
+\item Four main cohorts:
+  \bi
+  \item \textit{National Survey of Health and Development} (NSHD) -- b. March 1946
+  \item \textit{National Child Development Study} (NCDS) -- b. March 1958
+  \item \textit{British Cohort Study} (BCS) -- b. April 1970
+  \item \textit{Millennium Cohort Study} (MCS) -- b. 2000-2001
+  \ei
+~\\
+\item Follow 15,000+ individuals from birth onwards
+\item Include extensive data on
+  \bi
+  \item Perinatal period (pregnancy, birth outcomes)
+  \item Skill formation -- cognitive, socioemotional
+  \item Health and health behaviours
+  \item Family background and composition
+  \item Schooling and labour market outcomes
+  \ei
+~\\
+\pause
+\item Not always consistent
+  \bi
+  \item Different design, timing, questions, survey instruments...
+  \item Harmonisation efforts (CLOSER)
+  \ei
+\ei
+
+\end{frame}
 
@@
 \bi
-\item Data from 5 year follow-up (England only), maternal questionnaire
-\item Rutter A scale \citep{Rutter1970}
-  \bi
-  \item 19 items (behaviours)
+\item Data from 5 year follow-up
+\item Sample restricted to England
+\item Socioemotional skills: \textbf{Rutter A scale} \citep{Rutter1970}
+  \bi
+  \item 19 items (descriptions of behaviours) from maternal questionnaire
   \item Options: \textit{Doesn't apply, Applies somewhat, Certainly applies}
-  \item Usually conceptualised as measuring two dimensions
-  \bi \item \emph{externalising} behaviour problems (anti-social) \item \emph{internalising} behaviour problems (neurotic) \ei
   \ei
@@
 \bi
-\item Data from 5 year follow-up (England only), maternal questionnaire
-  \bi \item Complex survey design: rebalance sample to make representative \ei
-\item Strengths and Difficulties Questionnaire (SDQ, \citealp{Goodman1994})
+\item Data from 5 year follow-up
+\item Sample restricted to England
+\item Complex survey design: rebalance sample to make representative
+\item Socioemotional skills: \textbf{Strengths and Difficulties Questionnaire} (SDQ, \citealp{Goodman1994})
   \bi
   \item ``Updated" and expanded version of the Rutter scale 
-  \item 25 items (behaviours)
+  \item 25 items (descriptions of behaviours)
   \item Options: \textit{Not true, Somewhat true, Certainly true}
-  \item Designed with 5 subscales
-  \item Broader subscales are sometimes justifiable \citep{Goodman2010}
   \ei
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% HARMONISATION
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Harmonised scale}
+
 \subsection{Common items}
+
+\begin{frame}
+\frametitle{Harmonisation}
+\framesubtitle{Choice of common items}
+
+\bi
+\item Rutter and SDQ are different instruments
+\item Focus on subset of items that have similar wording between the two instruments
+\item New 11-item subscale
+\ei
+
+\end{frame}
 
@@
 
-table <- meantab[[2]] # females (but just need item names)
+table <- meantab[[2]] # females
 table <- table[,!names(table) %in% c("cats", "fac", "bcs.num", "mcs.num")] # drop if unneeded
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% EXPLORATORY FACTOR ANALYSIS
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \subsection{Exploratory Analysis}
@@
 \begin{frame}
+\frametitle{Factor structure}
+\framesubtitle{Existing evidence}
+
+\end{frame}
+
+\begin{frame}
 \frametitle{EFA}\label{efa}
-\framesubtitle{Factor structure}
+\framesubtitle{Dimensionality}
 
@@
 \begin{tabular}{C{.05\textwidth}L{.3\textwidth}|C{.05\textwidth}L{.3\textwidth}}
-  & \textbf{Externalising}  &    & \textbf{Internalising} \\
+  & \textbf{{\color{OliveGreen}Externalising}}  &    & \textbf{{\color{Orange}Internalising}}  \\
 1 & Restless                & 7  & Worried                \\
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% MEASUREMENT INVARIANCE
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Measurement Invariance}
-
-\begin{frame}
-\frametitle{Measurement Invariance}\label{mimain}
-\framesubtitle{Model setup}
-
-\bi
-\item Two constructs isolated in the new scale, separately by cohort
-\item But are externalising and internalising problems measured in a comparable way \emph{across cohorts}?
-\vspace{12pt}\pause
-\item \textbf{Measurement Invariance (MI)} analysis
-  \bi
-  \item Formal test of the invariance hypothesis
-  \item Set up categorical Item Factor Analysis (IFA) model \hyperlink{miapp1}{\beamerbutton{Det.}}
-  \item Pool the two cohorts together, and estimate models of different invariance level -- following \cite{Wu2016}:\hyperlink{miapp2}{\beamerbutton{Det.}}
-    \bnm
-    \item Start from minimally restrictive model, where parameters are free to vary
-    \item Restrict threshold and loading parameters to be equal across cohorts
-    \item Restrict intercept parameters as well
-    \enm
-  \item Test relative fit of restricted models compared to initial model
-  \ei 
-\ei
-
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-<<FA_INV_MAIN>>=
-# main models only
-@
-
-<<FA_FIT>>=
-# assemble fit indices into matrices
-@
-
-<<FA_SCORES_MAIN>>=
-# produce scores for main models
-@
-
-<<FA_SCORES_BIND>>=
-# bind scores with data
-@
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% MEASUREMENT INVARIANCE RESULTS
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{frame}
-\frametitle{Measurement Invariance}\label{mires}
-\framesubtitle{Results \hyperlink{measpars}{\beamerbutton{Measurement}}\hyperlink{lvpars}{\beamerbutton{Latent variable}}}
-
-\bi
-\item Restricting thresholds and loadings provides a good fit \hyperlink{fit}{\beamerbutton{Fit}}
-  \bi \item We can compare the scale (variance) of the latent factors \ei
-\item Models with restricted intercepts do \emph{not} pass MI test
-  \bi
-  \item Unclear whether mean comparisons can be done
-  \item \cite{Wu2016} vs \cite{Millsap2004}
-  \ei
-% \item Might \emph{age differences} in the sample be playing a role?
-%   \bi
-%   \item BCS and MCS samples differ due to survey design \hyperlink{agehist}{\beamerbutton{Graph}}
-%   \item But restricting the age range to maximise overlap does not give significantly different results \hyperlink{fit4}{\beamerbutton{Det.}}
-%   \ei
-  \pause
-\item \textbf{Q1}: Can we derive comparable measures of noncognitive skills across cohorts?
-\item Yes, we can use the model with non-invariant intercepts to score the latent factors (externalising and internalising skills)
-\ei
-
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% SCORES
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{frame}
-\frametitle{Factor scores}\label{density}
-\framesubtitle{Densities \hyperlink{rawhist}{\beamerbutton{Raw scores}}}
-\begin{figure}
-<<FACDENS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
-@
-\end{figure}
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% INEQUALITY
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Inequality in behaviour problems}
-
-\begin{frame}
-\frametitle{Inequality}
-
-\bi
-\item \textbf{Q2}: Has socioeconomic inequality in behavioural problems increased between 1975 and 2005?
-\item Investigate income gradient in externalising and internalising scores
-  \bi
-  \item Merge harmonised income data from CLOSER Work Package 2(a)
-  \item Weekly family income when child is 10 (not contemporaneous)
-  \item Working on access to harmonised social class data
-  \ei
-\ei
-
-\end{frame}
-
-\begin{frame}
-\frametitle{Inequality}\label{ineq_uc}
-\framesubtitle{Unconditional inequality}
-\centering
-<<INEQ_STATS, dependson="FA_SCORES_BIND", results='asis'>>=
-
-# income 
-tb <- data.frame(
-  aggregate(scores2plot$faminc10_infl, by=list(scores2plot$cohort), FUN = function(x) quantile(x, c(.1, .25, .75, .9), na.rm=T))
-)
-tb$d9010 <- tb$x[,4]/tb$x[,1]
-tb$d7525 <- tb$x[,3]/tb$x[,2]
-
-tb.ext <- data.frame(
-  aggregate(scores2plot$EXT, by=list(scores2plot$cohortsex), FUN = function(x) quantile(x, c(.1, .25, .5, .75, .9), na.rm=T))
-)
-tb.ext$d9010 <- abs(tb.ext$x[,"90%"]-tb.ext$x[,"10%"])
-tb.ext$d7525 <- abs(tb.ext$x[,"75%"]-tb.ext$x[,"25%"])
-tb.ext$d9050 <- abs(tb.ext$x[,"90%"]-tb.ext$x[,"50%"])
-tb.ext$d5010 <- abs(tb.ext$x[,"50%"]-tb.ext$x[,"10%"])
-rownames(tb.ext) <- tb.ext$Group.1
-tb.ext <- t(tb.ext[,-c(1,2)]) # remove aggregated quantiles and transpose
-tb.ext <- tb.ext[,c("BCS.M","MCS.M","BCS.F","MCS.F")]
-
-tb.int <- data.frame(
-  aggregate(scores2plot$INT, by=list(scores2plot$cohortsex), FUN = function(x) quantile(x, c(.1, .25, .5, .75, .9), na.rm=T))
-)
-tb.int$d9010 <- abs(tb.int$x[,"90%"]-tb.int$x[,"10%"])
-tb.int$d7525 <- abs(tb.int$x[,"75%"]-tb.int$x[,"25%"])
-tb.int$d9050 <- abs(tb.int$x[,"90%"]-tb.int$x[,"50%"])
-tb.int$d5010 <- abs(tb.int$x[,"50%"]-tb.int$x[,"10%"])
-rownames(tb.int) <- tb.int$Group.1
-tb.int <- t(tb.int[,-c(1,2)]) # remove aggregated quantiles and transpose
-tb.int <- tb.int[,c("BCS.M","MCS.M","BCS.F","MCS.F")]
-
-# assemble and print
-table <- rbind(tb.ext,tb.int)
-table <- cbind(data.frame(as.matrix(rep(c("90-10", "75-25", "90-50", "50-10"),2))),
-               table)
-
-print(xtable(table,
-                    align="L{0cm}L{.15\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}",
-                    digits = 4
-                    ),
-             size="\\scriptsize",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = c(list(0,4)),
-                              command = c(
-                              # header
-            "Measure & \\multicolumn{2}{c}{\\textbf{Males}} & \\multicolumn{2}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-3}  \\cmidrule(lr){4-5} 
-                       & BCS & MCS & BCS & MCS  \\\\
-                      \\textbf{Externalising} \\\\ ",
-                      "\\textbf{Internalising} \\\\ \\midrule "
-                                    )),
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabular", width = "\\textwidth",
-             floating = FALSE
-)
-
-@
-\end{frame}
-
-<<FACINEQ_MAKE>>=
-# make factor inequality graphs
-@
-
-\begin{frame}
-\frametitle{Inequality}\label{ineq_sc}
-\framesubtitle{Social class}
-\vspace{-.2cm}
-\begin{figure}
-<<FACINEQ_SC, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
-# plot
-fi <- fi_sc # select result
-gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
-                        fi$ineqlist[[3]],fi$ineqlist[[4]],
-                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
-                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
-@
-\end{figure}
-\end{frame}
-
-\begin{frame}
-\frametitle{Inequality}\label{ineq_sc}
-\framesubtitle{Age mother left education}
-\vspace{-.2cm}
-\begin{figure}
-<<FACINEQ_YS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
-# plot
-fi <- fi_ys # select result
-gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
-                        fi$ineqlist[[3]],fi$ineqlist[[4]],
-                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
-                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
-@
-\end{figure}
-\end{frame}
-
-\begin{frame}
-\frametitle{Inequality}\label{ineq_sc}
-\framesubtitle{Maternal education and employment}
-\vspace{-.2cm}
-\begin{figure}
-<<FACINEQ_MECS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=8.5, fig.width=13.5, out.width='\\linewidth'>>=
-# plot
-fi <- fi_mecs # select result
-gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
-                        fi$ineqlist[[3]],fi$ineqlist[[4]],
-                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
-                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
-@
-\end{figure}
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% EXPLAINING EXT and INT
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-<<SELECTVARS>>=
-# define variables and formulas for regression and decomposition
-@
-
-<<REGS_DEP, dependson="SELECTVARS">>=
-# run regressions explaining EXT INT
-@
-
-\begin{frame}
-\frametitle{What explains behavioural skills?}
-
-write about variables used and groups
-
-\end{frame}
-
-\begin{frame}
-\frametitle{What explains behavioural skills?}
-\framesubtitle{Males}
-\begin{table}
+\begin{frame}
+\frametitle{EFA}\label{meantab_fcc2}
+\framesubtitle{Choice of common items (females) \hyperlink{meantab_fcc3}{\beamerbutton{3 factors}}}
 \centering
 \resizebox{!}{4cm}{
-<<REGS_M, results='asis', dependson="REGRESSIONS">>=
-
-# select models
-regs <- list(r_ext$BCS.M[["OLS"]],r_ext$MCS.M[["OLS"]],r_int$BCS.M[["OLS"]],r_int$MCS.M[["OLS"]])
-
-# robust SEs
-robust_se <- list()
-robust_p  <- list()
-for (i in 1:length(regs)) {
-  cov1              <- sandwich::vcovHC(regs[[i]], type = "HC1")
-  ct                <- lmtest::coeftest(regs[[i]], vcov = cov1)
-  robust_se[[i]] <- ct[, 2]
-  robust_p[[i]] <- ct[, 4]
-}
-
-texreg::texreg(regs,
-               stars = c(0.1, 0.05, 0.01), digits = 3,
-               override.se = robust_se, override.pvalues = robust_p,
-               single.row = TRUE, scriptsize = TRUE,
-               caption = "", label = "",
-               omit.coef = "region|Inter",
-               groups = grouplabs,
-               custom.coef.names = varlabs,
-               custom.model.names = c("EXT BCS", "EXT MCS", "INT BCS", "INT MCS"),
-               table = FALSE
-               )
-
-@
-}
-\end{table}
-\end{frame}
-
-\begin{frame}
-\frametitle{What explains behavioural skills?}
-\framesubtitle{Females}
-\begin{table}
-\centering
-\resizebox{!}{4cm}{
-<<REGS_F, results='asis', dependson="REGRESSIONS">>=
-
-# select models
-regs <- list(r_ext$BCS.F[["OLS"]],r_ext$MCS.F[["OLS"]],r_int$BCS.F[["OLS"]],r_int$MCS.F[["OLS"]])
-
-# robust SEs
-robust_se <- list()
-robust_p  <- list()
-for (i in 1:length(regs)) {
-  cov1              <- sandwich::vcovHC(regs[[i]], type = "HC1")
-  ct                <- lmtest::coeftest(regs[[i]], vcov = cov1)
-  robust_se[[i]] <- ct[, 2]
-  robust_p[[i]] <- ct[, 4]
-}
-
-texreg::texreg(regs,
-               stars = c(0.1, 0.05, 0.01), digits = 3,
-               override.se = robust_se, override.pvalues = robust_p,
-               single.row = TRUE, scriptsize = TRUE,
-               caption = "", label = "",
-               omit.coef = "region|Inter",
-               groups = grouplabs,
-               custom.coef.names = varlabs,
-               custom.model.names = c("EXT BCS", "EXT MCS", "INT BCS", "INT MCS"),
-               table = FALSE
-               )
-
-@
-}
-\end{table}
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% GELBACH DECOMPOSITION
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{frame}
-\frametitle{Decomposition (EXT)}
-\framesubtitle{Explaining the education gradient}
-\centering
-\scriptsize
-\resizebox{\textwidth}{!}{
-\input{../../../tables/med_gelb_ext}
-}
-\end{frame}
-
-\begin{frame}
-\frametitle{Decomposition (INT)}
-\framesubtitle{Explaining the education gradient}
-\centering
-\scriptsize
-\resizebox{\textwidth}{!}{
-\input{../../../tables/med_gelb_int}
-}
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% EFFECTS ON LATER OUTCOMES
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-<<REGS_OUTC, dependson="SELECTVARS">>=
-# run regressions explaining adult outcomes
-@
-
-<<REGS_OUTC_TAB, dependson="REGS_OUTC">>=
-# make tables
-@
-
-\begin{frame}
-\frametitle{Outcomes}
-\framesubtitle{Adolescence (BCS)}
-\centering
-\resizebox{\textwidth}{!}{
-<<REGS_OUTC_BCS1, results='asis', dependson="REGRESSIONS">>=
-
-# collapse to matrix (first 4 elements of bcs_outctab are adolescence)
-bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab[c(1:4)],matrix,ncol=11,byrow=FALSE))
-table <- bcs_outctab_mat
-
-print(xtable(table,
-                    align="R{0cm}L{.22\\textwidth}C{.05\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.05\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}",
-                    digits = 3
-                    ),
-             size="\\scriptsize",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = c(list(0)),
-                              command = c(
-                              # header
-            " & \\multicolumn{5}{c}{\\textbf{Males}} & \\multicolumn{5}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-6}  \\cmidrule(lr){7-11}
-& Mean & \\multicolumn{4}{c}{Coefficients} & Mean & \\multicolumn{4}{c}{Coefficients} \\\\ \\cmidrule(lr){3-6}  \\cmidrule(lr){8-11}"
-            )),
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabular", width = "\\textwidth",
-             floating = FALSE
-)
-@
-}
-\end{frame}
-
-\begin{frame}
-\frametitle{Outcomes}
-\framesubtitle{Adulthood (BCS)}
-\centering
-\resizebox{\textwidth}{!}{
-<<REGS_OUTC_BCS2, results='asis', dependson="REGRESSIONS">>=
-
-# collapse to matrix (last 4 elements of bcs_outctab are adulthood)
-bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab[c(5:8)],matrix,ncol=11,byrow=FALSE))
-table <- bcs_outctab_mat
-
-print(xtable(table,
-                    align="R{0cm}L{.22\\textwidth}C{.05\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.05\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}",
-                    digits = 3
-                    ),
-             size="\\scriptsize",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = c(list(0)),
-                              command = c(
-                              # header
-            " & \\multicolumn{5}{c}{\\textbf{Males}} & \\multicolumn{5}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-6}  \\cmidrule(lr){7-11}
-& Mean & \\multicolumn{4}{c}{Coefficients} & Mean & \\multicolumn{4}{c}{Coefficients} \\\\ \\cmidrule(lr){3-6}  \\cmidrule(lr){8-11}"
-            )),
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabular", width = "\\textwidth",
-             floating = FALSE
-)
-@
-}
-\end{frame}
-
-\begin{frame}
-\frametitle{Outcomes}
-\framesubtitle{Adolescence (MCS)}
-\centering
-\resizebox{\textwidth}{!}{
-<<REGS_OUTC_MCS, results='asis', dependson="REGRESSIONS">>=
-
-# collapse to matrix
-mcs_outctab_mat <- do.call(rbind,lapply(mcs_outctab,matrix,ncol=11,byrow=FALSE))
-table <- mcs_outctab_mat
-
-print(xtable(table,
-                    align="R{0cm}L{.22\\textwidth}C{.05\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.05\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}",
-                    digits = 3
-                    ),
-             size="\\scriptsize",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = c(list(0)),
-                              command = c(
-                              # header
-            " & \\multicolumn{5}{c}{\\textbf{Males}} & \\multicolumn{5}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-6}  \\cmidrule(lr){7-11}
-& Mean & \\multicolumn{4}{c}{Coefficients} & Mean & \\multicolumn{4}{c}{Coefficients} \\\\ \\cmidrule(lr){3-6}  \\cmidrule(lr){8-11}"
-            )),
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabular", width = "\\textwidth",
-             floating = FALSE
-)
-@
-}
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% CONCLUSION
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Conclusions and next steps}
-
-\begin{frame}
-\frametitle{Conclusions and next steps}
-
-\bi
-\item We derived a comparable scale of non-cognitive skills, based on maternal reports of behaviour problems
-\item We have shown that inequality in these skills at five years of age has increased between children born in 1970 and 2000
-  \bi
-  \item Profile has become steeper (particularly for males)
-  \item Scores at upper quantiles have drifted upwards
-  \ei
-\pause
-\vspace{12pt}
-\item Next steps
-  \bi
-  \item Harmonise more cohorts (NCDS) and ages (10)
-  \item Broaden analysis to the \emph{returns} to non-cognitive skills -- are these remunerated differently on the labour market for people born in different years?
-  \ei
-\ei
-
-\end{frame}
-
-\begin{frame}[noframenumbering]
-\centering
-{\Large Thank you!}
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% MEASUREMENT INVARIANCE APPENDIX
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{frame}[noframenumbering]
-\centering
-{\Large Appendix}
-\end{frame}
-
-\begin{frame}
-\frametitle{Harmonisation}\label{meantab_m}
-\framesubtitle{Choice of common items (males) \hyperlink{meantab_f}{\beamerreturnbutton{Females}}}
-\centering
-\resizebox{!}{4cm}{
-<<MEANTAB_M, dependson="MEANTABLE", results='asis'>>=
+<<MEANTAB_FCC, dependson="MEANTABLE", results='asis'>>=
 require(xtable)
 
-table <- meantab[[1]] # males
+table <- meantab_cc2[[2]] # females
 table <- table[,!names(table) %in% c("cats", "fac", "bcs.num", "mcs.num")] # drop if unneeded
@@
 
-<<BCSEFA>>=
-@
-
-<<BCSEFARES, dependson='BCSEFA'>>=
-@
-
-<<MCSEFA>>=
-@
-
-<<MCSEFARES, dependson='MCSEFA'>>=
-@
-
-\begin{frame}[noframenumbering]
-\frametitle{EFA}\label{numfac}
-\framesubtitle{Number of factors \hyperlink{efa}{\beamerreturnbutton{Back}}}
-\begin{center}
-<<EFATAB, dependson='MCSEFA', results='asis'>>=
-# combine results
-efatab <- cbind(efatab.bcs,efatab.mcs)
-colnames(efatab) <- c("BCS All", "BCS M", "BCS F", "MCS All", "MCS M", "MCS F")
-
-print(xtable(efatab,
-                    align="L{.25\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}",
-                    digits = 0
-                    ),
-             size="\\small",
-             include.rownames=T,
-             include.colnames=T,
-             booktabs = T,
-             sanitize.text.function = function(x){x},
-             tabular.environment = "tabular", width = ".8\\textwidth",
-             floating = FALSE
-)
-@
-\end{center}
-\end{frame}
-
-\begin{frame}[noframenumbering]
-\frametitle{MI}\label{miapp1}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% MEASUREMENT INVARIANCE
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Measurement Invariance}
+
+\begin{frame}
+\frametitle{Measurement Invariance}\label{mimain}
+\framesubtitle{Model setup}
+
+\bi
+\item Two constructs isolated in the new scale, separately by cohort
+\item But are externalising and internalising problems measured in a comparable way \emph{across cohorts}?
+\vspace{12pt}\pause
+\item \textbf{Measurement Invariance (MI)} analysis
+  \bi
+  \item Formal test of the invariance hypothesis
+  \item Set up categorical Item Factor Analysis (IFA) model
+  \item Pool the two cohorts together, and estimate models of different invariance level -- following \cite{Wu2016}
+  \item Test relative fit of restricted models compared to initial model
+  \ei 
+\ei
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Measurement Invariance}\label{midet}
 \framesubtitle{IFA model}
-Categorical Item Factor Analysis (IFA) model \hyperlink{mimain}{\beamerreturnbutton{Back}}
+Categorical Item Factor Analysis (IFA) model
 
@@
 \item Threshold model \citep{Muthen1984}
-\item Latent propensity to endorse each item $i$ for child $j$ in cohort $c$
+\item Latent propensity to endorse each item $i$ for child $j$ in group $c$
 \begin{align}\label{eq:propens}
@@
 
+\begin{frame}
+\frametitle{MI}\label{misteps}
+\framesubtitle{Invariance steps \hyperlink{miapp1}{\beamerbutton{Details}}}
+\bnm
+\item \textbf{Configural} model
+  \bi
+  \item Minimally restrictive (all parameters free across groups)
+  \item Normalised on latent means $\bm{\kappa}$ and variances $\bm{\Phi}$
+  \ei
+\item \textbf{Threshold and Loading invariance} model
+  \bi
+  \item Restrict $\bm{\tau}_{1}$, $\bm{\tau}_{2}$, and $\bm{\Lambda}$ to be equal across groups
+  \item Allows comparison of latent factor variances
+  \ei
+\item \textbf{Threshold, Loading, and Intercept invariance} model
+  \bi
+  \item Restrict $\bm{\tau}_{1}$, $\bm{\tau}_{2}$, $\bm{\Lambda}$, and $\bm{\nu}$ to be equal across groups
+  \item Allows comparison of latent factor variances \emph{and} means
+  \ei
+\enm
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+<<FA_INV_MAIN>>=
+# main models only
+@
+
+<<FA_FIT>>=
+# assemble fit indices into matrices
+@
+
+<<FA_SCORES_MAIN>>=
+# produce scores for main models
+@
+
+<<FA_SCORES_BIND>>=
+# bind scores with data
+@
+
+\begin{frame}
+\frametitle{Measurement Invariance}\label{mires}
+\framesubtitle{Results \hyperlink{measpars}{\beamerbutton{Measurement}}\hyperlink{lvpars}{\beamerbutton{Latent variable}}}
+
+\bi
+\item We use RMSE to compare fit between models of different invariance level
+\item Restricting thresholds and loadings provides a good fit \hyperlink{fit}{\beamerbutton{Fit details}}
+  \bi \item We can compare the scale (variance) of the latent factors \ei
+\item Models with restricted intercepts do \emph{not} pass MI test
+  \bi
+  \item Unclear whether mean comparisons can be done
+  \item \cite{Wu2016} vs \cite{Millsap2004}
+  \ei
+\ei
+\begin{center}
+<<FA_FIT_RES, dependson="FA_FIT", results='asis'>>=
+require(xtable)
+
+table <- rbind(afitab[[1]])
+table <- as.matrix(table[,c("mnames","npar","rmsea")])
+
+for (c in 1:ncol(table)) {
+  table[2,c] <- paste0("{\\color{green}", table[2,c], "}")
+  table[3,c] <- paste0("{\\color{red}", table[3,c], "}")
+}
+
+print(xtable(table,
+                    align="L{0cm}L{.3\\textwidth}C{.1\\textwidth}C{.15\\textwidth}",
+                    digits = 4
+                    ),
+             size="\\footnotesize",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0)),
+                              command = c(
+                              # header
+            "Model & Num. params & RMSEA \\\\"
+                                    )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = "5\\textwidth",
+             floating = FALSE
+)
+
+@
+\end{center}
+\end{frame}
+
+\begin{frame}
+\frametitle{Measurement Invariance}
+\framesubtitle{Results \hyperlink{measpars}{\beamerbutton{Measurement}}\hyperlink{lvpars}{\beamerbutton{Latent variable}}}
+\bi
+\item \textbf{Q1}: Can we derive comparable measures of noncognitive skills across cohorts?
+\item Yes, we can use the model with non-invariant intercepts to score the latent factors (externalising and internalising skills)
+\item All scores are coded so that \emph{higher is better}
+\ei
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% SCORES
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\frametitle{Factor scores}\label{density}
+\framesubtitle{Densities \hyperlink{rawhist}{\beamerbutton{Raw scores}}}
+\begin{figure}
+<<FACDENS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
+@
+\end{figure}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% INEQUALITY
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Inequality in behaviour problems}
+
+\begin{frame}
+\frametitle{Inequality}
+
+\bi
+\item \textbf{Q2}: Has socioeconomic inequality in behavioural problems increased between 1975 and 2005?
+\item Investigate income gradient in externalising and internalising scores
+  \bi
+  \item Merge harmonised income data from CLOSER Work Package 2(a)
+  \item Weekly family income when child is 10 (not contemporaneous)
+  \item Working on access to harmonised social class data
+  \ei
+\ei
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Inequality}\label{ineq_uc}
+\framesubtitle{Unconditional inequality}
+\centering
+<<INEQ_STATS, dependson="FA_SCORES_BIND", results='asis'>>=
+
+# income 
+tb <- data.frame(
+  aggregate(scores2plot$faminc10_infl, by=list(scores2plot$cohort), FUN = function(x) quantile(x, c(.1, .25, .75, .9), na.rm=T))
+)
+tb$d9010 <- tb$x[,4]/tb$x[,1]
+tb$d7525 <- tb$x[,3]/tb$x[,2]
+
+tb.ext <- data.frame(
+  aggregate(scores2plot$EXT, by=list(scores2plot$cohortsex), FUN = function(x) quantile(x, c(.1, .25, .5, .75, .9), na.rm=T))
+)
+tb.ext$d9010 <- abs(tb.ext$x[,"90%"]-tb.ext$x[,"10%"])
+tb.ext$d7525 <- abs(tb.ext$x[,"75%"]-tb.ext$x[,"25%"])
+tb.ext$d9050 <- abs(tb.ext$x[,"90%"]-tb.ext$x[,"50%"])
+tb.ext$d5010 <- abs(tb.ext$x[,"50%"]-tb.ext$x[,"10%"])
+rownames(tb.ext) <- tb.ext$Group.1
+tb.ext <- t(tb.ext[,-c(1,2)]) # remove aggregated quantiles and transpose
+tb.ext <- tb.ext[,c("BCS.M","MCS.M","BCS.F","MCS.F")]
+
+tb.int <- data.frame(
+  aggregate(scores2plot$INT, by=list(scores2plot$cohortsex), FUN = function(x) quantile(x, c(.1, .25, .5, .75, .9), na.rm=T))
+)
+tb.int$d9010 <- abs(tb.int$x[,"90%"]-tb.int$x[,"10%"])
+tb.int$d7525 <- abs(tb.int$x[,"75%"]-tb.int$x[,"25%"])
+tb.int$d9050 <- abs(tb.int$x[,"90%"]-tb.int$x[,"50%"])
+tb.int$d5010 <- abs(tb.int$x[,"50%"]-tb.int$x[,"10%"])
+rownames(tb.int) <- tb.int$Group.1
+tb.int <- t(tb.int[,-c(1,2)]) # remove aggregated quantiles and transpose
+tb.int <- tb.int[,c("BCS.M","MCS.M","BCS.F","MCS.F")]
+
+# assemble and print
+table <- rbind(tb.ext,tb.int)
+table <- cbind(data.frame(as.matrix(rep(c("90-10", "75-25", "90-50", "50-10"),2))),
+               table)
+
+print(xtable(table,
+                    align="L{0cm}L{.15\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}",
+                    digits = 4
+                    ),
+             size="\\scriptsize",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0,4)),
+                              command = c(
+                              # header
+            "Measure & \\multicolumn{2}{c}{\\textbf{Males}} & \\multicolumn{2}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-3}  \\cmidrule(lr){4-5} 
+                       & BCS & MCS & BCS & MCS  \\\\
+                      \\textbf{Externalising} \\\\ ",
+                      "\\textbf{Internalising} \\\\ \\midrule "
+                                    )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = "\\textwidth",
+             floating = FALSE
+)
+
+@
+\end{frame}
+
+<<FACINEQ_MAKE>>=
+# make factor inequality graphs
+@
+
+\begin{frame}
+\frametitle{Inequality}\label{ineq_sc}
+\framesubtitle{Social class}
+\vspace{-.2cm}
+\begin{figure}
+<<FACINEQ_SC, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
+# plot
+fi <- fi_sc # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{Inequality}\label{ineq_ys}
+\framesubtitle{Age mother left education}
+\vspace{-.2cm}
+\begin{figure}
+<<FACINEQ_YS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
+# plot
+fi <- fi_ys # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{Inequality}\label{ineq_mecs}
+\framesubtitle{Maternal education and employment}
+\vspace{-.2cm}
+\begin{figure}
+<<FACINEQ_MECS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=11, out.width='\\linewidth'>>=
+# plot
+fi <- fi_mecs # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\end{figure}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% EXPLAINING EXT and INT
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+<<SELECTVARS>>=
+# define variables and formulas for regression and decomposition
+@
+
+<<REGS_DEP, dependson="SELECTVARS">>=
+# run regressions explaining EXT INT
+@
+
+\begin{frame}
+\frametitle{What explains behavioural skills?}
+
+write about variables used and groups
+
+\end{frame}
+
+\begin{frame}
+\frametitle{What explains behavioural skills?}
+\framesubtitle{Males}
+\begin{table}
+\centering
+\resizebox{\textwidth}{!}{
+<<REGS_M, results='asis', dependson="REGRESSIONS">>=
+
+# select models
+regs <- list(r_ext$BCS.M[["OLS"]],r_ext$MCS.M[["OLS"]],r_int$BCS.M[["OLS"]],r_int$MCS.M[["OLS"]])
+
+# robust SEs
+robust_se <- list()
+robust_p  <- list()
+for (i in 1:length(regs)) {
+  cov1              <- sandwich::vcovHC(regs[[i]], type = "HC1")
+  ct                <- lmtest::coeftest(regs[[i]], vcov = cov1)
+  robust_se[[i]] <- ct[, 2]
+  robust_p[[i]] <- ct[, 4]
+}
+
+texreg::texreg(regs,
+               stars = c(0.1, 0.05, 0.01), digits = 3,
+               override.se = robust_se, override.pvalues = robust_p,
+               single.row = TRUE, scriptsize = TRUE,
+               caption = "", label = "",
+               omit.coef = "region|Inter",
+               groups = grouplabs,
+               custom.coef.names = varlabs,
+               custom.model.names = c("EXT BCS", "EXT MCS", "INT BCS", "INT MCS"),
+               include.rmse = FALSE,
+               table = FALSE
+               )
+
+@
+}
+\end{table}
+\end{frame}
+
+\begin{frame}
+\frametitle{What explains behavioural skills?}
+\framesubtitle{Females}
+\begin{table}
+\centering
+\resizebox{\textwidth}{!}{
+<<REGS_F, results='asis', dependson="REGRESSIONS">>=
+
+# select models
+regs <- list(r_ext$BCS.F[["OLS"]],r_ext$MCS.F[["OLS"]],r_int$BCS.F[["OLS"]],r_int$MCS.F[["OLS"]])
+
+# robust SEs
+robust_se <- list()
+robust_p  <- list()
+for (i in 1:length(regs)) {
+  cov1              <- sandwich::vcovHC(regs[[i]], type = "HC1")
+  ct                <- lmtest::coeftest(regs[[i]], vcov = cov1)
+  robust_se[[i]] <- ct[, 2]
+  robust_p[[i]] <- ct[, 4]
+}
+
+texreg::texreg(regs,
+               stars = c(0.1, 0.05, 0.01), digits = 3,
+               override.se = robust_se, override.pvalues = robust_p,
+               single.row = TRUE, scriptsize = TRUE,
+               caption = "", label = "",
+               omit.coef = "region|Inter",
+               groups = grouplabs,
+               custom.coef.names = varlabs,
+               custom.model.names = c("EXT BCS", "EXT MCS", "INT BCS", "INT MCS"),
+               include.rmse = FALSE,
+               table = FALSE
+               )
+
+@
+}
+\end{table}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% GELBACH DECOMPOSITION
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}
+\frametitle{Decomposition (EXT)}\label{gelb_ext}
+\framesubtitle{Explaining the education gradient \hyperlink{gelbsc_ext}{\beamerbutton{Social class}}}
+\centering
+\scriptsize
+\resizebox{\textwidth}{!}{
+\input{../../../tables/med_gelb_ext}
+}
+\end{frame}
+
+\begin{frame}
+\frametitle{Decomposition (INT)}\label{gelb_int}
+\framesubtitle{Explaining the education gradient \hyperlink{gelbsc_int}{\beamerbutton{Social class}}}
+\centering
+\scriptsize
+\resizebox{\textwidth}{!}{
+\input{../../../tables/med_gelb_int}
+}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% EFFECTS ON LATER OUTCOMES
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+<<REGS_OUTC, dependson="SELECTVARS">>=
+# run regressions explaining adult outcomes
+@
+
+<<REGS_OUTC_TAB, dependson="REGS_OUTC">>=
+# make tables
+@
+
+<<REGS_OUTC_TABPR, dependson="REGS_OUTC">>=
+# function to display outcome regression tables
+printoutctab <- function(tab) {
+print(xtable(tab,
+                    align="R{0cm}L{.2\\textwidth}C{.05\\textwidth}C{.09\\textwidth}C{.09\\textwidth}C{.09\\textwidth}C{.09\\textwidth}C{.05\\textwidth}C{.09\\textwidth}C{.09\\textwidth}C{.09\\textwidth}C{.09\\textwidth}",
+                    digits = 3
+                    ),
+             size="\\scriptsize",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0)),
+                              command = c(
+                              # header
+            " & \\multicolumn{5}{c}{\\textbf{Males}} & \\multicolumn{5}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-6}  \\cmidrule(lr){7-11}
+& Mean & \\multicolumn{4}{c}{Coefficients} & Mean & \\multicolumn{4}{c}{Coefficients} \\\\ \\cmidrule(lr){3-6}  \\cmidrule(lr){8-11}"
+            )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = "\\textwidth",
+             floating = FALSE
+)
+}
+@
+
+\begin{frame}
+\frametitle{Outcomes}
+\framesubtitle{Adolescence (MCS)}
+\centering
+\resizebox{\textwidth}{!}{
+<<REGS_OUTC_MCS, results='asis', dependson="REGRESSIONS">>=
+
+# collapse to matrix
+mcs_outctab_mat <- do.call(rbind,lapply(mcs_outctab,matrix,ncol=11,byrow=FALSE))
+table <- mcs_outctab_mat
+
+# NAs in columns and rows with COG
+table[,c(6,11)] <- NA
+table[seq(4,16,4),] <- NA
+
+printoutctab(table)
+
+@
+}
+\end{frame}
+
 \begin{frame}[noframenumbering]
-\frametitle{MI}
-\framesubtitle{Key Parameters for Invariance}
-
-\bi
-\item Loadings $\lambda_{ci}$
-\item Thresholds $\tau_{1,ci}$, $\tau_{2,ci}$
-\item Intercepts $\nu_{ci}$
-\item Unique variances $\Var(u_{ci})$
-\item Latent mean vector $\bm{\kappa}_c$
-\item Latent covariance matrix $\bm{\Phi}_c$
-\ei
-
+\frametitle{Outcomes}
+\framesubtitle{Adolescence (MCS)}
+\centering
+\resizebox{\textwidth}{!}{
+<<REGS_OUTC_MCS_COG, results='asis', dependson="REGRESSIONS">>=
+
+# collapse to matrix
+mcs_outctab_mat <- do.call(rbind,lapply(mcs_outctab,matrix,ncol=11,byrow=FALSE))
+table <- mcs_outctab_mat
+
+printoutctab(table)
+
+@
+}
+\end{frame}
+
+\begin{frame}
+\frametitle{Outcomes}
+\framesubtitle{Adolescence (BCS)}
+\centering
+\resizebox{\textwidth}{!}{
+<<REGS_OUTC_BCS1, results='asis', dependson="REGRESSIONS">>=
+
+# collapse to matrix
+bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab[c("smktry16", "alcoh16", "bmi16")],matrix,ncol=11,byrow=FALSE))
+table <- bcs_outctab_mat
+
+# NAs in columns and rows with COG
+table[,c(6,11)] <- NA
+table[seq(4,12,4),] <- NA
+
+printoutctab(table)
+
+@
+}
 \end{frame}
@@
 \begin{frame}[noframenumbering]
-\frametitle{MI}\label{miapp2}
-\framesubtitle{Invariance steps \hyperlink{mimain}{\beamerreturnbutton{Back}}}
+\frametitle{Outcomes}
+\framesubtitle{Adolescence (BCS)}
+\centering
+\resizebox{\textwidth}{!}{
+<<REGS_OUTC_BCS1_COG, results='asis', dependson="REGRESSIONS">>=
+
+# collapse to matrix
+bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab[c("smktry16", "alcoh16", "bmi16")],matrix,ncol=11,byrow=FALSE))
+table <- bcs_outctab_mat
+
+printoutctab(table)
+
+@
+}
+\end{frame}
+
+\begin{frame}
+\frametitle{Outcomes}
+\framesubtitle{Adulthood (BCS)}
+\centering
+\resizebox{\textwidth}{!}{
+<<REGS_OUTC_BCS2, results='asis', dependson="REGRESSIONS">>=
+
+# collapse to matrix (last 4 elements of bcs_outctab are adulthood)
+bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab[c(5:8)],matrix,ncol=11,byrow=FALSE))
+table <- bcs_outctab_mat
+
+# NAs in columns and rows with COG
+table[,c(6,11)] <- NA
+table[seq(4,16,4),] <- NA
+
+printoutctab(table)
+
+@
+}
+\end{frame}
+
+\begin{frame}[noframenumbering]
+\frametitle{Outcomes}
+\framesubtitle{Adulthood (BCS)}
+\centering
+\resizebox{\textwidth}{!}{
+<<REGS_OUTC_BCS2_COG, results='asis', dependson="REGRESSIONS">>=
+
+# collapse to matrix (last 4 elements of bcs_outctab are adulthood)
+bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab[c(5:8)],matrix,ncol=11,byrow=FALSE))
+table <- bcs_outctab_mat
+
+printoutctab(table)
+
+@
+}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% CONCLUSION
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Conclusions and next steps}
+
+\begin{frame}
+\frametitle{Conclusions and next steps}
+
+\bi
+\item We derived a comparable scale of non-cognitive skills, based on maternal reports of behaviour problems
+\item We have shown that inequality in these skills at five years of age has increased between children born in 1970 and 2000
+  \bi
+  \item Profile has become steeper (particularly for males)
+  \item Scores at upper quantiles have drifted upwards
+  \ei
+\pause
+\vspace{12pt}
+\item Next steps
+  \bi
+  \item Harmonise more cohorts (NCDS) and ages (10)
+  \item Broaden analysis to the \emph{returns} to non-cognitive skills -- are these remunerated differently on the labour market for people born in different years?
+  \ei
+\ei
+
+\end{frame}
+
+\begin{frame}[noframenumbering]
+\centering
+{\Large Thank you!}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% MEASUREMENT INVARIANCE APPENDIX
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}[noframenumbering]
+\centering
+{\Large Appendix}
+\end{frame}
+
+\begin{frame}
+\frametitle{Harmonisation}\label{meantab_m}
+\framesubtitle{Choice of common items (males) \hyperlink{meantab_f}{\beamerreturnbutton{Females}}}
+\centering
+\resizebox{!}{4cm}{
+<<MEANTAB_M, dependson="MEANTABLE", results='asis'>>=
+require(xtable)
+
+table <- meantab[[1]] # males
+table <- table[,!names(table) %in% c("cats", "fac", "bcs.num", "mcs.num")] # drop if unneeded
+
+print(xtable(table,
+                    align="L{0cm}C{.04\\textwidth}L{.4\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}L{.4\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
+                    digits = 1
+                    ), 
+             size="\\small",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0)),
+                              command = c(
+                              # header
+                                    paste(" & 
+                                              \\multicolumn{4}{c}{\\textbf{BCS}} &
+                                              \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9} 
+                                            Item &
+                                              Text & Cert. Appl. & Smt. Appl. & Appl. &
+                                              Text & Cert. True & Smt. True & True \\\\ \\midrule
+                                          "
+                                            , sep="")
+                                    )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = ".8\\textwidth",
+             floating = FALSE
+)
+
+@
+} %scalebox
+\end{frame}
+% 
+% <<BCSEFA>>=
+% @
+% 
+% <<BCSEFARES, dependson='BCSEFA'>>=
+% @
+% 
+% <<MCSEFA>>=
+% @
+% 
+% <<MCSEFARES, dependson='MCSEFA'>>=
+% @
+% 
+% 
+% \begin{frame}[noframenumbering]
+% \frametitle{EFA}\label{numfac}
+% \framesubtitle{Number of factors \hyperlink{efa}{\beamerreturnbutton{Back}}}
+% \begin{center}
+% <<EFATAB, dependson='MCSEFARES', results='asis'>>=
+% # combine results
+% efatab <- cbind(efatab.bcs,efatab.mcs)
+% colnames(efatab) <- c("BCS All", "BCS M", "BCS F", "MCS All", "MCS M", "MCS F")
+% 
+% 
+% print(xtable(efatab,
+%                     align="L{.25\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}",
+%                     digits = 0
+%                     ),
+%              size="\\small",
+%              include.rownames=T,
+%              include.colnames=T,
+%              booktabs = T,
+%              sanitize.text.function = function(x){x},
+%              tabular.environment = "tabular", width = ".8\\textwidth",
+%              floating = FALSE
+% )
+% @
+% \end{center}
+% \end{frame}
+
+\begin{frame}
+\frametitle{EFA}\label{meantab_fcc3}
+\framesubtitle{Choice of common items (females) \hyperlink{meantab_fcc2}{\beamerreturnbutton{2 factors}}}
+\centering
+\resizebox{!}{4cm}{
+<<MEANTAB_FCC3, dependson="MEANTABLE", results='asis'>>=
+require(xtable)
+
+table <- meantab_cc3[[2]] # females
+table <- table[,!names(table) %in% c("cats", "fac", "bcs.num", "mcs.num")] # drop if unneeded
+
+print(xtable(table,
+                    align="L{0cm}C{.04\\textwidth}L{.4\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}L{.4\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
+                    digits = 1
+                    ), 
+             size="\\small",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0)),
+                              command = c(
+                              # header
+                                    paste(" & 
+                                              \\multicolumn{4}{c}{\\textbf{BCS}} &
+                                              \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9} 
+                                            Item &
+                                              Text & Cert. Appl. & Smt. Appl. & Appl. &
+                                              Text & Cert. True & Smt. True & True \\\\ \\midrule
+                                          "
+                                            , sep="")
+                                    )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = ".8\\textwidth",
+             floating = FALSE
+)
+
+@
+} %scalebox
+\end{frame}
+
+\begin{frame}[noframenumbering]
+\frametitle{MI}\label{miapp1}
+\framesubtitle{Invariance steps \hyperlink{misteps}{\beamerreturnbutton{Back}}}
 \Fontvi
@@
 \resizebox{\textwidth}{!}{
-<<FA_FIT_RES, dependson="FA_FIT", results='asis'>>=
+<<FA_FIT_RES_APP, dependson="FA_FIT", results='asis'>>=
 require(xtable)
@@
 
-\begin{frame}
+\begin{frame}[noframenumbering]
 \frametitle{Factor scores}\label{rawhist}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% ROBUSTNESS APPENDIX
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% \begin{frame}[noframenumbering]
-% \frametitle{Age distribution \hyperlink{mires}{\beamerreturnbutton{Back}}}\label{agehist}
-% \begin{figure}
-% \begin{center}
-% <<AGEHIST, dev='tikz', fig.height=6, fig.width=6, out.width='.6\\linewidth', cache=TRUE>>=
-% @
-% \end{center}
-% \caption{Distribution of age in the BCS and MCS sample}
-% \end{figure}
-% \end{frame}
+%%%%%%%% GELBACH DECOMPOSITION WITH SOCIAL CLASS
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}[noframenumbering]
+\frametitle{Decomposition (EXT)}\label{gelbsc_ext}
+\framesubtitle{Explaining the education gradient \hyperlink{gelb_ext}{\beamerreturnbutton{Back}}}
+\centering
+\scriptsize
+\resizebox{\textwidth}{!}{
+\input{../../../tables/med_gelb_sc_ext}
+}
+\end{frame}
+
+\begin{frame}[noframenumbering]
+\frametitle{Decomposition (INT)}\label{gelbsc_int}
+\framesubtitle{Explaining the education gradient \hyperlink{gelb_int}{\beamerreturnbutton{Back}}}
+\centering
+\scriptsize
+\resizebox{\textwidth}{!}{
+\input{../../../tables/med_gelb_sc_int}
+}
+\end{frame}
 
@@
 \def\bibfont{\tiny}
-\bibliography{/Users/giacomomason/Documents/BibTex/Papers-cohorts_nc}
+\bibliography{/Users/giacomomason/Dropbox/Zotero/cohorts.bib}
 \end{frame}

### 2018-05-22 codes/R/paper/cohorts_nc.Rnw
delete incineq, reorder SCL in plot
paper update DO NOT RUN, to be checked

--- 2018-05-09 codes/R/paper/cohorts_nc.Rnw
@@
 \usepackage{textgreek}            % for greek letters in text
+\usepackage{ragged2e}             % for table notes justification
 
@@
 
-@
-
-<<READ, cache=FALSE >>=
-library(here)
-
-read_chunk(here("codes/R/1_load.R"))
-read_chunk(here("codes/R/2_efa.R"))
-read_chunk(here("codes/R/3_inv_main.R"))
-read_chunk(here("codes/R/3_inv_aux.R"))
-read_chunk(here("codes/R/3_inv_fit.R"))
-read_chunk(here("codes/R/4_scores.R"))
-read_chunk(here("codes/R/4_scores_aux.R"))
-read_chunk(here("codes/R/5_graphs.R"))
-read_chunk(here("codes/R/6_decvars.R"))
-read_chunk(here("codes/R/7_counterf.R"))
-read_chunk(here("codes/R/7_counterf_graphs.R"))
-read_chunk(here("codes/R/7_counterf_tabs.R"))
-read_chunk(here("codes/R/7_counterf_seqdec.R"))
-read_chunk(here("codes/R/getpars.R"))
-
-@
-
-<<PREAMBLE, cache=FALSE >>=
-@
-
-<<LOAD_DATA>>=
 @
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\clearpage
-
-<<MEANTABLE>>=
-# make table with means
-@
-
-\clearpage
-\begin{landscape}
-\begin{table}[ht]
-\centering
-\footnotesize
-\caption{Item Commonalities -- Males}\label{tab:comm_m}
-<<MEANTABLE_RES_M, dependson="MEANTABLE", results='asis'>>=
-require(xtable)
-
-table <- meantab[[1]]
-
-print(xtable(table,
-                                  # item num        Factor          Num cats        BCS text        BCS num           BCS Certain     BCS Somewhat    BCS Applies             MCS text          MCS num
-                    align="L{0cm}C{.04\\textwidth}C{.05\\textwidth}C{.05\\textwidth}L{.3\\textwidth}C{.03\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}L{.3\\textwidth}C{.03\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
-                    digits = 1
-                    ), 
-             size="\\small",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = c(list(0)),
-                              command = c(
-                              # header
-                                    paste(" & & &
-                                              \\multicolumn{5}{c}{\\textbf{BCS}} &
-                                              \\multicolumn{5}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){4-8}  \\cmidrule(lr){9-13} 
-                                            Item & Factor & N. categ &
-                                              Text & & Cert. Applies & Somet. Applies & Applies &
-                                              Text & & Cert. True & Somet. True & True \\\\ \\midrule
-                                          "
-                                            , sep="")
-                                    )),
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabular", width = ".8\\textwidth",
-             floating = FALSE
-)
-
-@
-\end{table}
-\end{landscape}
-
-\clearpage
-\begin{landscape}
-\begin{table}[ht]
-\centering
-\footnotesize
-\caption{Item Commonalities -- Females}\label{tab:comm_f}
-<<MEANTABLE_RES_F, dependson="MEANTABLE", results='asis'>>=
-require(xtable)
-
-table <- meantab[[2]]
-
-print(xtable(table,
-                                  # item num        Factor          Num cats        BCS text        BCS num           BCS Certain     BCS Somewhat    BCS Applies             MCS text          MCS num
-                    align="L{0cm}C{.04\\textwidth}C{.05\\textwidth}C{.05\\textwidth}L{.3\\textwidth}C{.03\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}L{.3\\textwidth}C{.03\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
-                    digits = 1
-                    ), 
-             size="\\small",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = c(list(0)),
-                              command = c(
-                              # header
-                                    paste(" & & &
-                                              \\multicolumn{5}{c}{\\textbf{BCS}} &
-                                              \\multicolumn{5}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){4-8}  \\cmidrule(lr){9-13} 
-                                            Item & Factor & N. categ &
-                                              Text & & Cert. Applies & Somet. Applies & Applies &
-                                              Text & & Cert. True & Somet. True & True \\\\ \\midrule
-                                          "
-                                            , sep="")
-                                    )),
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabular", width = ".8\\textwidth",
-             floating = FALSE
-)
-
-@
-\end{table}
-\end{landscape}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\clearpage
+\subsection{Inequality in noncognitive ability}
+The model can be used to investigate how inequality of noncognitive skills has changed between the two cohorts.
+
+\textbf{REVIEW} We can derive scores for $\theta^{EXT}$ and $\theta^{INT}$ directly, and plot them against socioeconomic measures observed in the data.
+Our main socioeconomic indicator is a measure of weekly total family net income in the prices of the interview year.\footnote{This measure is harmonised across cohorts and derived as part of CLOSER, which is funded by the ESRC.} Using the estimated parameters for the model with invariant thresholds and loadings (but non-invariant intercepts), I compute the scores for each individual in the sample (using the Empirical Bayes modal approach).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+<<READ, cache=FALSE >>=
+library(here)
+
+read_chunk(here("codes/R/1_load.R"))
+read_chunk(here("codes/R/2_efa.R"))
+read_chunk(here("codes/R/3_inv_main.R"))
+read_chunk(here("codes/R/3_inv_aux.R"))
+read_chunk(here("codes/R/3_inv_fit.R"))
+read_chunk(here("codes/R/4_scores.R"))
+read_chunk(here("codes/R/4_scores_aux.R"))
+read_chunk(here("codes/R/5_graphs.R"))
+read_chunk(here("codes/R/6_selectvars.R"))
+read_chunk(here("codes/R/7_counterf.R"))
+read_chunk(here("codes/R/7_counterf_graphs.R"))
+read_chunk(here("codes/R/7_counterf_tabs.R"))
+read_chunk(here("codes/R/7_counterf_seqdec.R"))
+read_chunk(here("codes/R/8_regs_dep.R"))
+read_chunk(here("codes/R/8_regs_outc.R"))
+read_chunk(here("codes/R/getpars.R"))
+
+@
+
+<<PREAMBLE >>=
+@
+
+<<LOAD_DATA>>=
+@
+
+<<MEANTABLE>>=
+# make table with means
+@
+
+\clearpage
+\begin{landscape}
+\begin{table}[ht]
+\centering
+\footnotesize
+\caption{Item Commonalities -- Males}\label{tab:comm_m}
+<<MEANTABLE_RES_M, dependson="MEANTABLE", results='asis'>>=
+require(xtable)
+
+table <- meantab[[1]]
+
+print(xtable(table,
+                                  # item num        Factor          Num cats        BCS text        BCS num           BCS Certain     BCS Somewhat    BCS Applies             MCS text          MCS num
+                    align="L{0cm}C{.04\\textwidth}C{.05\\textwidth}C{.05\\textwidth}L{.3\\textwidth}C{.03\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}L{.3\\textwidth}C{.03\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
+                    digits = 1
+                    ), 
+             size="\\small",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0)),
+                              command = c(
+                              # header
+                                    paste(" & & &
+                                              \\multicolumn{5}{c}{\\textbf{BCS}} &
+                                              \\multicolumn{5}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){4-8}  \\cmidrule(lr){9-13} 
+                                            Item & Factor & N. categ &
+                                              Text & & Cert. Applies & Somet. Applies & Applies &
+                                              Text & & Cert. True & Somet. True & True \\\\ \\midrule
+                                          "
+                                            , sep="")
+                                    )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = ".8\\textwidth",
+             floating = FALSE
+)
+
+@
+\end{table}
+\end{landscape}
+
+\clearpage
+\begin{landscape}
+\begin{table}[ht]
+\centering
+\footnotesize
+\caption{Item Commonalities -- Females}\label{tab:comm_f}
+<<MEANTABLE_RES_F, dependson="MEANTABLE", results='asis'>>=
+require(xtable)
+
+table <- meantab[[2]]
+
+print(xtable(table,
+                                  # item num        Factor          Num cats        BCS text        BCS num           BCS Certain     BCS Somewhat    BCS Applies             MCS text          MCS num
+                    align="L{0cm}C{.04\\textwidth}C{.05\\textwidth}C{.05\\textwidth}L{.3\\textwidth}C{.03\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}L{.3\\textwidth}C{.03\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
+                    digits = 1
+                    ), 
+             size="\\small",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0)),
+                              command = c(
+                              # header
+                                    paste(" & & &
+                                              \\multicolumn{5}{c}{\\textbf{BCS}} &
+                                              \\multicolumn{5}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){4-8}  \\cmidrule(lr){9-13} 
+                                            Item & Factor & N. categ &
+                                              Text & & Cert. Applies & Somet. Applies & Applies &
+                                              Text & & Cert. True & Somet. True & True \\\\ \\midrule
+                                          "
+                                            , sep="")
+                                    )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = ".8\\textwidth",
+             floating = FALSE
+)
+
+@
+\end{table}
+\end{landscape}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 # produce scores for main models
+@
+
+<<FA_SCORES_BIND>>=
+# bind scores with data
 @
@@
 \begin{figure}
-<<FACDENS, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
+<<FACDENS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
 @
@@
 \begin{figure}
-<<RAWHIST, dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
+<<RAWHIST, dependson="FA_SCORES_BIND", dev='tikz', fig.height=10, fig.width=10, out.width='1\\linewidth'>>=
 @
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\clearpage
-\subsection{Inequality in noncognitive ability}
-The model can be used to investigate how inequality of noncognitive skills has changed between the two cohorts.
-
-\textbf{REVIEW} We can derive scores for $\theta^{EXT}$ and $\theta^{INT}$ directly, and plot them against socioeconomic measures observed in the data.
-Our main socioeconomic indicator is a measure of weekly total family net income in the prices of the interview year.\footnote{This measure is harmonised across cohorts and derived as part of CLOSER, which is funded by the ESRC.} Using the estimated parameters for the model with invariant thresholds and loadings (but non-invariant intercepts), I compute the scores for each individual in the sample (using the Empirical Bayes modal approach).
-In \autoref{fig:meanci}, we can see that the gradient in noncognitive skills along family income quintiles is more pronounced in the MCS cohort.
-
-<<INEQ_STATS, dependson="FA_SCORES_MAIN", results="markup">>=
+\begin{table}[ht]
+\centering
+\small
+\caption{Inequality measures for behavioural skills}\label{tab:ineqscores}
+<<INEQ_STATS, dependson="FA_SCORES_BIND", results='asis'>>=
 
@@
 tb$d7525 <- tb$x[,3]/tb$x[,2]
-cat("Quantile differences in income by cohort")
-tb
 
@@
 tb.ext$d9010 <- abs(tb.ext$x[,"90%"]-tb.ext$x[,"10%"])
-tb.ext$d9010 <- abs(tb.ext$x[,"75%"]-tb.ext$x[,"25%"])
+tb.ext$d7525 <- abs(tb.ext$x[,"75%"]-tb.ext$x[,"25%"])
+tb.ext$d9050 <- abs(tb.ext$x[,"90%"]-tb.ext$x[,"50%"])
 tb.ext$d5010 <- abs(tb.ext$x[,"50%"]-tb.ext$x[,"10%"])
-tb.ext$d9050 <- abs(tb.ext$x[,"90%"]-tb.ext$x[,"50%"])
-cat("\n Quantile differences in EXT by cohort/gender")
-tb.ext
+rownames(tb.ext) <- tb.ext$Group.1
+tb.ext <- t(tb.ext[,-c(1,2)]) # remove aggregated quantiles and transpose
+tb.ext <- tb.ext[,c("BCS.M","MCS.M","BCS.F","MCS.F")]
 
@@
 tb.int$d9010 <- abs(tb.int$x[,"90%"]-tb.int$x[,"10%"])
-tb.int$d9010 <- abs(tb.int$x[,"75%"]-tb.int$x[,"25%"])
+tb.int$d7525 <- abs(tb.int$x[,"75%"]-tb.int$x[,"25%"])
+tb.int$d9050 <- abs(tb.int$x[,"90%"]-tb.int$x[,"50%"])
 tb.int$d5010 <- abs(tb.int$x[,"50%"]-tb.int$x[,"10%"])
-tb.int$d9050 <- abs(tb.int$x[,"90%"]-tb.int$x[,"50%"])
-cat("\n Quantile differences in INT by cohort/gender")
-tb.int
-
+rownames(tb.int) <- tb.int$Group.1
+tb.int <- t(tb.int[,-c(1,2)]) # remove aggregated quantiles and transpose
+tb.int <- tb.int[,c("BCS.M","MCS.M","BCS.F","MCS.F")]
+
+# assemble and print
+table <- rbind(tb.ext,tb.int)
+table <- cbind(data.frame(as.matrix(rep(c("90-10", "75-25", "90-50", "50-10"),2))),
+               table)
+
+print(xtable(table,
+                    align="L{0cm}L{.20\\textwidth}C{.15\\textwidth}C{.15\\textwidth}C{.15\\textwidth}C{.15\\textwidth}",
+                    digits = 4
+                    ),
+             size="\\small",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0,4)),
+                              command = c(
+                              # header
+            "Measure & \\multicolumn{2}{c}{\\textbf{Males}} & \\multicolumn{2}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-3}  \\cmidrule(lr){4-5} 
+                       & BCS & MCS & BCS & MCS  \\\\
+                      \\textbf{Externalising} \\\\ \\midrule ",
+                      "\\textbf{Internalising} \\\\ \\midrule "
+                                    )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = "\\textwidth",
+             floating = FALSE
+)
+
+@
+\end{table}
+
+<<FACINEQ_MAKE>>=
+# make factor inequality graphs
 @
@@
 \begin{figure}
-\begin{center}
-<<INCINEQ, dependson="FA_SCORES_MAIN", dev='tikz', fig.height=4, fig.width=4, out.width='.5\\linewidth'>>=
-@
-\end{center}
-\caption{Evolution of income inequality }
-{\scriptsize \emph{Notes}:  \par}
-\end{figure}
-
-<<FACINEQ_DEMEAN>>=
-# function for demeaning
-@
-
-% 
-% \begin{figure}
-% <<FACINEQ, dependson="FA_SCORES_MAIN", dev='tikz', fig.height=12, fig.width=10, out.width='1\\linewidth'>>=
-% @
-% \caption{Mean scores across income quintiles}\label{fig:meanci}
-% {\scriptsize \emph{Notes}: The figure plots the mean and its 95\% confidence interval across the quintiles in the income distribution. \par}
-% \end{figure}
-
-\begin{figure}
-<<FACINEQ_SC, dependson="FA_SCORES_MAIN", dev='tikz', fig.height=12, fig.width=10, out.width='1\\linewidth'>>=
+<<FACINEQ_SC, dependson="FA_SCORES_BIND", dev='tikz', fig.height=12, fig.width=10, out.width='1\\linewidth'>>=
+# plot
+fi <- fi_sc # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
 @
@@
 
+\clearpage
 \begin{figure}
-<<FACINEQ_YS, dependson="FA_SCORES_MAIN", dev='tikz', fig.height=12, fig.width=10, out.width='1\\linewidth'>>=
+<<FACINEQ_YS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=12, fig.width=10, out.width='1\\linewidth'>>=
+fi <- fi_ys # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
 @
@@
 
-\begin{figure}
-<<FACINEQ_ME, dependson="FA_SCORES_MAIN", dev='tikz', fig.height=12, fig.width=10, out.width='1\\linewidth'>>=
-@
-\caption{Mean scores across maternal employment}\label{fig:meanci_me}
-{\scriptsize \emph{Notes}: The figure plots the mean and its 95\% confidence interval by employment status of the cohort member's mother at 5 years of age. \par}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\clearpage
-
-<<FORMULAS>>=
-# define variables and formulas for regression and decomposition
-@
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% 
-% \begin{table}
-% \caption{Quantifying the gradient}
-% \centering
-% \resizebox{\textwidth}{!}{
-% <<FA_REGS_REAL_M, results='asis'>>=
-% # regress scores on income
-% library(stargazer)
-% 
-% decvars_reg <- c(decvars, "cohort") # add cohort
-% interact_reg <- paste0(decvars_reg[! decvars_reg %in% c("cohort")], "*cohort")
-% 
-% form_ext_reg <- as.formula(paste("EXT ~ ", paste0(decvars2, collapse = '+')))
-% form.ext2i <- as.formula(paste("EXT ~ ", paste0(c(decvars2, interact2), collapse = '+')))
-% form.int2 <- as.formula(paste("INT ~ ", paste0(decvars2, collapse = '+')))
-% form.int2i <- as.formula(paste("INT ~ ", paste0(c(decvars2, interact2), collapse = '+')))
-% 
-% r.ext.m  <- lm(form.ext2,  data=subset(scores2plot, sex=="M"))
-% r.exti.m <- lm(form.ext2i, data=subset(scores2plot, sex=="M"))
-% r.ext.f  <- lm(form.ext2,  data=subset(scores2plot, sex=="F"))
-% r.exti.f <- lm(form.ext2i, data=subset(scores2plot, sex=="F"))
-% r.int.m  <- lm(form.int2,  data=subset(scores2plot, sex=="M"))
-% r.inti.m <- lm(form.int2i, data=subset(scores2plot, sex=="M"))
-% r.int.f  <- lm(form.int2,  data=subset(scores2plot, sex=="F"))
-% r.inti.f <- lm(form.int2i, data=subset(scores2plot, sex=="F"))
-% 
-% stargazer(r.ext.m, r.exti.m, r.ext.f, r.exti.f, r.int.m, r.inti.m, r.int.f, r.inti.f,
-%           column.labels=c("M","M","F","F","M","M","F","F"), #,"BCS F","MCS F","BCS F","MCS F"),
-%           dep.var.caption = "", multicolumn = FALSE,
-%           #omit=c("Constant", paste0("region", c(2:9))),
-%           no.space=TRUE,
-%           omit.stat= c("rsq", "f", "ser", "aic", "ll"),
-%           notes = "All estimates control for Government Office Region of birth.",
-%           single.row=TRUE,
-%           float=FALSE    # needs begin table outside it
-% )
-% @
-% }
-% \end{table}
-
-\clearpage
-\subsection{Decomposition of inequality}
-
-We have populations $k \in \mathcal{K}$.
-For each population, we observe covariates $X_k$.
-For some $j \subseteq \mathcal{K}$, we observe outcome $Y_j$.
-We can identify the covariate distributions $F_{X_k}$ and the conditional distributions $F_{Y_j \mid X_j}$ and quantile functions $Q_{Y_j \mid X_j}$.
-Define the support of $X_k$ as $\mathcal{X}_k$ and the region of interest for the outcome as $\mathcal{Y}_j$.
-
-The interest is in the counterfactual distribution of the outcome:
-$$ F_{Y(j \mid k)}(y) = \int_{\mathcal{X}_k} F_{Y_j \mid X_j}(y \mid x) d F_{X_k}(x), \qquad y \in \mathcal{Y}_j $$
-and the corresponding quantiles:
-$$ Q_{Y(j \mid k)}(\tau) = F^\leftarrow_{Y(j \mid k)}(\tau) $$
-
-Given this, we can define counterfactual effects (CE):
-\bi
-\item Type 1 (Structure effect): CE of changing the conditional distribution: $Q_{Y(j \mid k)}(\tau) - Q_{Y(l \mid k)}(\tau)$
-\item Type 2 (Composition effect): CE of changing the covariate distribution: $Q_{Y(j \mid k)}(\tau) - Q_{Y(j \mid m)}(\tau)$
-\item Type 3: CE of changing both conditional and covariate distribution: $Q_{Y(j \mid k)}(\tau) - Q_{Y(l \mid m)}(\tau)$
-\ei
-
-In our application, we have $k=BCS,MCS$.
-We can thus decompose the observed differences in the distribution of non-cognitive skills into structure and composition effects.
-For example, the difference in the distribution of externalising skills can be written as:
-$$F_{(\theta^{EXT}_{MCS} \mid X_{MCS})} - F_{(\theta^{EXT}_{BCS} \mid X_{BCS})} =
-  \left[ F_{(\theta^{EXT}_{MCS} \mid X_{MCS})} - F_{(\theta^{EXT}_{BCS} \mid X_{MCS})} \right] +
-  \left[ F_{(\theta^{EXT}_{BCS} \mid X_{MCS})} - F_{(\theta^{EXT}_{BCS} \mid X_{BCS})} \right]$$
-
-% 
-% <<COUNTERF, dependson="FA_SCORES_MAIN">>=
-% # perform counterfactuals
-% @
-% 
-% <<COUNTERF_PREP_PLOTS, dependson='COUNTERF'>>=
-% # prepare counterfactual plots
-% @
-% 
-% \begin{landscape}
-% \begin{figure}
-% <<COUNTERF_EXT, dependson='COUNTERF_PREP_PLOTS', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
-% cfp.ext
-% @
-% \caption{Counterfactual decomposition -- externalising skills}\label{fig:counterf_ext}
-% {\scriptsize \emph{Notes}: \par}
-% \end{figure}
-% \end{landscape}
-% 
-% \begin{landscape}
-% \begin{figure}
-% <<COUNTERF_INT, dependson='COUNTERF_PREP_PLOTS', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
-% cfp.int
-% @
-% \caption{Counterfactual decomposition -- internalising skills}\label{fig:counterf_int}
-% {\scriptsize \emph{Notes}: \par}
-% \end{figure}
-% \end{landscape}
-% 
-% <<COUNTERF_PREP_TAB, dependson='COUNTERF'>>=
-% # prepare counterfactual tables
-% @
-% 
-% <<COUNTERF_INEQ, dependson='COUNTERF_PREP_TABS', results='markup'>>=
-% cat("Males EXT")
-% print(iqdecomp(logitcf.ext.m))
-% 
-% cat("Males INT")
-% print(iqdecomp(logitcf.int.m))
-% 
-% cat("Females EXT")
-% print(iqdecomp(logitcf.ext.f))
-% 
-% cat("Females INT")
-% print(iqdecomp(logitcf.int.f))
-% @
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% COUNTERFACTUAL DECOMPOSITIONS
-
-<<COUNTERF_SEQDEC_PREP>>=
-# data for decomposition plots
-@
-
-<<COUNTERF_SEQDEC_PLOT>>=
-# decomposition plots
-@
-
+\clearpage
+\par\vfill\break % break page to change margins
+\advance\hoffset by -1cm % Shift top margin
 \begin{landscape}
 \begin{figure}
-<<COUNTERF_SEQDEC_EXT, dependson='COUNTERF_SEQDEC_PREP', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
-cfp.seqdec.ext
-@
-\caption{Counterfactual sequential decomposition -- externalising skills}\label{fig:counterf_seq_ext}
-{\scriptsize \emph{Notes}: \par}
+<<FACINEQ_MECS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=10, fig.width=14, out.width='1\\linewidth'>>=
+fi <- fi_mecs # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,1.5,0.3))
+@
+\caption{Mean scores across maternal employment and schooling}\label{fig:meanci_me}
+{\scriptsize \emph{Notes}: The figure plots the mean and its 95\% confidence interval across interactions between a dummy for whether the cohort member's mother continued education past compulsory schooling age, and employment status of the cohort member's mother at 5 years of age.
+\par}
 \end{figure}
 \end{landscape}
+\par\vfill\break
+\advance\hoffset by 1cm % Shift top margin back to normal
+
+\clearpage
+\begin{figure}
+<<FACINEQ_SM, dependson="FA_SCORES_BIND", dev='tikz', fig.height=12, fig.width=10, out.width='1\\linewidth'>>=
+fi <- fi_sm # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\caption{Mean scores across maternal marital status}\label{fig:meanci_sm}
+{\scriptsize \emph{Notes}: The figure plots the mean and its 95\% confidence interval by marital status of the cohort member's mother at 5 years of age. \par}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\clearpage
+
+<<SELECTVARS>>=
+# define variables and formulas for regression and decomposition
+@
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+<<REGS_DEP, dependson="SELECTVARS">>=
+# run regressions explaining EXT INT
+@
 
 \begin{landscape}
-\begin{figure}
-<<COUNTERF_SEQDEC_INT, dependson='COUNTERF_SEQDEC_PREP', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
-cfp.seqdec.int
-@
-\caption{Counterfactual sequential decomposition -- internalising skills}\label{fig:counterf_seq_int}
-{\scriptsize \emph{Notes}: \par}
-\end{figure}
+\begin{table}
+\caption{Quantifying the gradient -- Males}
+\centering
+\resizebox{24cm}{!}{
+<<REGS_M, results='asis', dependson="REGRESSIONS">>=
+library(stargazer)
+stargazer(r.ext$BCS.M[[1]], r.ext$BCS.M[[2]], r.ext$MCS.M[[1]], r.ext$MCS.M[[2]], r.int$BCS.M[[1]], r.int$BCS.M[[2]], r.int$MCS.M[[1]], r.int$MCS.M[[2]],
+          column.labels=c("BCS","BCS","MCS","MCS","BCS","BCS","MCS","MCS"),
+          dep.var.caption = "", multicolumn = FALSE,
+          #omit=c("Constant", paste0("region", c(2:9))),
+          no.space=TRUE,
+          omit.stat= c("f", "ser", "aic", "ll"),
+          omit.table.layout = "n",
+          single.row=TRUE,
+          float=FALSE    # needs begin table outside it
+)
+@
+}
+\\[-12pt] \justify {\scriptsize \emph{Notes}: This table shows coefficients from regressions of externalising and internalising skills on potential determinants.
+\par}
+\end{table}
 \end{landscape}
+
+\begin{landscape}
+\begin{table}
+\caption{Quantifying the gradient -- Females}
+\centering
+\resizebox{24cm}{!}{
+<<REGS_F, results='asis', dependson="REGRESSIONS">>=
+library(stargazer)
+stargazer(r.ext$BCS.F[[1]], r.ext$BCS.F[[2]], r.ext$MCS.F[[1]], r.ext$MCS.F[[2]], r.int$BCS.F[[1]], r.int$BCS.F[[2]], r.int$MCS.F[[1]], r.int$MCS.F[[2]],
+          column.labels=c("BCS","BCS","MCS","MCS","BCS","BCS","MCS","MCS"),
+          dep.var.caption = "", multicolumn = FALSE,
+          #omit=c("Constant", paste0("region", c(2:9))),
+          no.space=TRUE,
+          omit.stat= c("f", "ser", "aic", "ll"),
+          omit.table.layout = "n",
+          single.row=TRUE,
+          float=FALSE    # needs begin table outside it
+)
+@
+}
+\\[-12pt] \justify {\scriptsize \emph{Notes}: This table shows coefficients from regressions of externalising and internalising skills on potential determinants.
+\par}
+\end{table}
+\end{landscape}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\par\vfill\break % break page to change margins
+\advance\hoffset by -1cm % Shift top margin
+\begin{landscape}
+\begin{table}[!htp]
+\caption{Gelbach decomposition of the EXT/INT differential by maternal education}
+\centering
+\input{../../../tables/med_gelb}
+\\[-12pt] \justify {\scriptsize \emph{Notes}: This table shows the decomposition of the differential in externalising and internalising skills between children whose mothers are educated past the compulsory schooling age, and children whose mothers left school at the minimum school leaving age.
+Education of mothers is observed when the child is aged around 5 (1975 for BCS, 2005 for MCS).
+Col.
+(1) shows: the unconditional difference, i.e.
+the coefficient on maternal education in a regression that controls only for region of birth; the conditional difference, i.e.
+the same coefficient after adding explanatory variables; and the portion of the unconditional difference explained by each group of variables.
+Col.
+(2) shows the fraction of the total difference due to each group of variables.
+The decomposition follows Gelbach (2016), and is invariant to the order in which the explanatory variables are added to the conditional model.
+\par}
+\end{table}
+\end{landscape}
+\par\vfill\break
+\advance\hoffset by 1cm % Shift top margin back to normal
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+<<REGS_OUTC, dependson="SELECTVARS">>=
+# run regressions explaining adult outcomes
+@
+
+<<REGS_OUTC_TAB, dependson="REGS_OUTC">>=
+# make tables
+@
+
+\begin{table}
+\caption{Effect of behavioural skills on adolescent outcomes -- BCS}
+\centering
+\resizebox{\linewidth}{!}{
+<<REGS_OUTC_BCS, results='asis', dependson="REGRESSIONS">>=
+
+# collapse to matrix
+bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab,matrix,ncol=11,byrow=FALSE))
+table <- bcs_outctab_mat
+
+print(xtable(table,
+                    align="R{0cm}L{.25\\textwidth}C{.06\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.06\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}",
+                    digits = 3
+                    ),
+             size="\\small",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0)),
+                              command = c(
+                              # header
+            " & \\multicolumn{5}{c}{\\textbf{Males}} & \\multicolumn{5}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-6}  \\cmidrule(lr){7-11}
+& Mean & \\multicolumn{4}{c}{Coefficients} & Mean & \\multicolumn{4}{c}{Coefficients} \\\\ \\cmidrule(lr){3-6}  \\cmidrule(lr){8-11}"
+            )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = "\\textwidth",
+             floating = FALSE
+)
+@
+}
+\\[-12pt] \justify {\scriptsize \emph{Notes}: This table shows coefficients from regressions of adolescent and adult outcomes on externalising and internalising skills, also conditioning on early childhood variables.
+\par}
+\end{table}
+
+\begin{table}
+\caption{Effect of behavioural skills on adolescent outcomes -- MCS}
+\centering
+\resizebox{\linewidth}{!}{
+<<REGS_OUTC_MCS, results='asis', dependson="REGRESSIONS">>=
+
+# collapse to matrix
+mcs_outctab <- do.call(rbind,lapply(mcs_outctab,matrix,ncol=11,byrow=FALSE))
+table <- mcs_outctab
+
+print(xtable(table,
+                    align="R{0cm}L{.25\\textwidth}C{.06\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.06\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}",
+                    digits = 3
+                    ),
+             size="\\small",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0)),
+                              command = c(
+                              # header
+            " & \\multicolumn{5}{c}{\\textbf{Males}} & \\multicolumn{5}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-6}  \\cmidrule(lr){7-11}
+& Mean & \\multicolumn{4}{c}{Coefficients} & Mean & \\multicolumn{4}{c}{Coefficients} \\\\ \\cmidrule(lr){3-6}  \\cmidrule(lr){8-11}"
+            )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = "\\textwidth",
+             floating = FALSE
+)
+@
+}
+\\[-12pt] \justify {\scriptsize \emph{Notes}: This table shows coefficients from regressions of adolescent and adult outcomes on externalising and internalising skills, also conditioning on early childhood variables.
+\par}
+\end{table}
 
@@
 
-<<BCSEFA>>=
-@
-
-<<MCSEFA>>=
+<<BCSEFA, dependson="LOAD_DATA">>=
+@
+
+<<MCSEFA, dependson="LOAD_DATA">>=
 @
@@
 
-datacompl <- scores2plot[complete.cases(scores2plot[,decvars]),] # complete cases
 # expand dummies
 datasum <- cbind(
-  datacompl[,c("EXT","INT","cohortsex")], 
-  dummy.data.frame(datacompl[,c(decvars)], sep = "")
+  finaldata[,c("EXT","INT","cohortsex")], 
+  dummy.data.frame(finaldata[,c(decvars)], sep = "")
 )

### 2018-05-23 codes/R/6_selectvars.R
add social class/occupation (birth and 5)

@@ -3,8 +3,9 @@
 
 # select variables for decomposition, in groups
 decvarslist <- list()
+decvarslist$ses <- c("fscl5wb")
 decvarslist$med <- c("mpsla5")
-decvarslist$mem <- c("mempl5")
+decvarslist$mem <- c("mempl5b")
 decvarslist$mch <- c("mothageb", "mheight", "singlem", "ethn")
 decvarslist$prg <- c("parity", "firstb", "nprevst", "smkpr", "preterm", "lbwt")
 
@@ -15,7 +16,7 @@ mchxbth <-  list(apply(
 decvarslist_int <- c(decvarslist, mchxbth)
 
 # names of the groups
-decvarsgroups <- c("Mat. Education", "Mat. Employment", "Mat. Charact.", "Pregnancy/Birth")
+decvarsgroups <- c("Father Occupation", "Mat. Education", "Mat. Employment", "Mat. Charact.", "Pregnancy/Birth")
 
 # make into vector
 decvars <- unname(unlist(decvarslist))
@@ -29,22 +30,3 @@ finaldata <- scores2plot[complete.cases(scores2plot[,decvars]),]
 
 # export to stata for Gelbach
 export(finaldata, paste0(dir_data, "finaldata.dta"))
-
-# labels
-declabs <- c(scl10b = "Parental social class (10)",
-             mpsla5 = "Mother post-compuls. education (5)",
-             mempl5 = "Mother employment status (5)",
-             mothageb = "Mother age (0)",
-             teenm = "Teen mother (0)",
-             mheight = "Mother height (0)",
-             singlem = "Unmarried mother (0)",
-             ethn = "Nonwhite ethnicity (0)",
-             parity = "Parity (0)",
-             firstb = "Firstborn child (0)",
-             nprevst = "Num previous stillbirths (0)",
-             smkpr = "Mother smoked in pregnancy (0)",
-             caesbirth = "Caesarean birth (0)",
-             preterm = "Preterm birth (0)",
-             lbwt = "(log) Birthweight (0)"
-)
-

### 2018-05-23 codes/R/8_regs_dep.R
update regression groups

@@ -10,18 +10,18 @@ decvars_reg <- c(decvars, c("region"))
 decvars_int_reg <- c(decvars_int, c("region"))
 
 # labels for groups (texreg)
-grouplabs <- list(#"Social class (10)" = 1:4, 
-                  "Maternal education (5)" = 1, 
-                  "Maternal employment (5)" = 2:3,
-                  "Maternal background (birth)" = 4:7,
-                  "Pregnancy" = 8:13)
+grouplabs <- list("Father occupation (5)" = 1:2, 
+                  "Maternal education (5)" = 3, 
+                  "Maternal employment (5)" = 4,
+                  "Maternal background (birth)" = 5:8,
+                  "Pregnancy" = 9:14)
 # labels for variables (texreg)
-varlabs <- c(#"IIIM", "IIINM", "I/II", "Other",
+varlabs <- c("Blue collar", "No father",
              "Post-compulsory",
-             "Part time", "Full time",
+             "Employed",
              "Age", "Height", "Unmarried", "Nonwhite child",
              "Parity", "Firstborn", "Num prev. stillbirths",
-             "Mother smoked in pregnancy","Preterm birth",
+             "Mother smoked in pregnancy", "Preterm birth",
              "(log) Birthweight"
 )
 

### 2018-05-23 codes/R/slides/cohorts_tapes.Rnw
add facineq by father occupation

@@
     \bi
-    \item Does the same instrument measure the same constructs in the same way:
+    \item Most economic literature uses simple sum scores
+    \item But does the same instrument measure the same constructs in the same way:
       \bi 
@@
 
+How many dimensions underlie the new 11-item scale?
+\bi
+\item Rutter A scale (BCS) 
+  \bi 
+  \item Traditionally conceptualised as bidimensional \citep{Blanden2007}
+    \bi
+    \item \emph{anti-social} -- or aggressive-hyperactive
+    \item \emph{neurotic} -- or anxious-fearful
+    \ei
+  \item Sometimes the former is split into aggressive-antisocial and hyperkinetic \citep{McGee1985}
+  \ei
+\item Strengths and difficulties (MCS) 
+  \bi
+  \item Five subscales (emotional, hyperactivity, prosocial, peer, conduct) \citep{Stone2010}
+  \item Evidence that lower-dimensional structures can work \citep{Goodman2010}
+  \ei
+\ei
+
 \end{frame}
@@
 
+On our data:
 \bi
@@
 \item In both cohorts, EFA delivers a clear-cut split between two factors
-\item Factors can be coherently interpreted as internalising and internalising skills
+\item Factors can be coherently interpreted as externalising ($\sim$ anti-social) and internalising ($\sim$ neurotic)
 \ei 
@@
 \item Two constructs isolated in the new scale, separately by cohort
-\item But are externalising and internalising problems measured in a comparable way \emph{across cohorts}?
+\item But are externalising and internalising measured in a comparable way \emph{across cohorts}?
+\ei
 \vspace{12pt}\pause
-\item \textbf{Measurement Invariance (MI)} analysis
-  \bi
-  \item Formal test of the invariance hypothesis
-  \item Set up categorical Item Factor Analysis (IFA) model
-  \item Pool the two cohorts together, and estimate models of different invariance level -- following \cite{Wu2016}
+\textbf{Measurement Invariance (MI)} analysis \citep{Wu2016a}
+  \bi
+  \item Formal test of the invariance hypothesis using factor analysis
+  \item Aim: assess if constructs are measured in the same way in two or more groups
+  \item Procedure:
+    \bnm
+    \item Start from the most general model -- parameters allowed differ across groups
+    \item Impose restrictions on sets of parameters to be equal across groups
     \item Test relative fit of restricted models compared to initial model
-  \ei 
+    \enm
   \ei
@@
 \begin{frame}
-\frametitle{MI}\label{misteps}
-\framesubtitle{Invariance steps \hyperlink{miapp1}{\beamerbutton{Details}}}
+\frametitle{Measurement Invariance}\label{misteps}
+\framesubtitle{Steps \hyperlink{miapp1}{\beamerbutton{Details}}}
+
+We follow the approach in \cite{Wu2016a}
+\bi
+\item Define group ($c$) as cohort-gender cell
+  \bi \item Invariance results do not vary if separating by gender \ei
+\ei
+~\\
 \bnm
@@
 \frametitle{Measurement Invariance}\label{mires}
-\framesubtitle{Results \hyperlink{measpars}{\beamerbutton{Measurement}}\hyperlink{lvpars}{\beamerbutton{Latent variable}}}
+\framesubtitle{Results}
 
@@
 \item Models with restricted intercepts do \emph{not} pass MI test
-  \bi
-  \item Unclear whether mean comparisons can be done
-  \item \cite{Wu2016} vs \cite{Millsap2004}
-  \ei
+  \bi \item We cannot compare the location of latent factors in different groups \ei
 \ei
@@
 print(xtable(table,
-                    align="L{0cm}L{.3\\textwidth}C{.1\\textwidth}C{.15\\textwidth}",
+                    align="L{0cm}L{.5\\textwidth}C{.1\\textwidth}C{.15\\textwidth}",
                     digits = 4
@@
              sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabular", width = "5\\textwidth",
+             tabular.environment = "tabular", width = "\\textwidth",
              floating = FALSE
@@
 \frametitle{Measurement Invariance}
-\framesubtitle{Results \hyperlink{measpars}{\beamerbutton{Measurement}}\hyperlink{lvpars}{\beamerbutton{Latent variable}}}
-\bi
-\item \textbf{Q1}: Can we derive comparable measures of noncognitive skills across cohorts?
-\item Yes, we can use the model with non-invariant intercepts to score the latent factors (externalising and internalising skills)
+\framesubtitle{Factor scores}
+\bi
+\item Nevertheless, we can use the model with non-invariant intercepts to score the latent factors (externalising and internalising dimensions)
+\item Estimated parameters: \hyperlink{measpars}{\beamerbutton{Measurement}}\hyperlink{lvpars}{\beamerbutton{Latent variable}}
+\item Use Empirical Bayes approach to predict individual scores
 \item All scores are coded so that \emph{higher is better}
+\item \textbf{CAUTION}: location differences by cohort are not interpretable!
+~\\
+\pause
+\item Is this an improvement on using simple sum scores?
+  \bi
+  \item Scores from the model exploit the covariance structure of the whole 11-item scale
+  \item Every pattern of response is assigned a score
+  \item Substantial increase in variability ({\color{red} could add boxplot of factor scores vs sum scores})
+  \ei
 \ei
@@
 \begin{frame}
+\frametitle{Inequality}\label{ineq_fsc}
+\framesubtitle{Father occupation}
+\vspace{-.2cm}
+\begin{figure}
+<<FACINEQ_FSC, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
+# plot
+fi <- fi_fsc # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\end{figure}
+\end{frame}
+
+\begin{frame}
 \frametitle{Inequality}\label{ineq_ys}
@@
 \begin{frame}
-\frametitle{What explains behavioural skills?}
-\framesubtitle{Males}
+\frametitle{What explains socioemotional skills?}
+\framesubtitle{Externalising}
 \begin{table}
@@
 # select models
-regs <- list(r_ext$BCS.M[["OLS"]],r_ext$MCS.M[["OLS"]],r_int$BCS.M[["OLS"]],r_int$MCS.M[["OLS"]])
+regs <- list(r_ext$BCS.M[["OLS"]],r_ext$MCS.M[["OLS"]],r_ext$BCS.F[["OLS"]],r_ext$MCS.F[["OLS"]])
 
@@
                custom.coef.names = varlabs,
-               custom.model.names = c("EXT BCS", "EXT MCS", "INT BCS", "INT MCS"),
+               custom.model.names = c("Males BCS", "Males MCS", "Females BCS", "Females MCS"),
                include.rmse = FALSE,
@@
 \begin{frame}
-\frametitle{What explains behavioural skills?}
-\framesubtitle{Females}
+\frametitle{What explains socioemotional skills?}
+\framesubtitle{Internalising}
 \begin{table}
@@
 # select models
-regs <- list(r_ext$BCS.F[["OLS"]],r_ext$MCS.F[["OLS"]],r_int$BCS.F[["OLS"]],r_int$MCS.F[["OLS"]])
+regs <- list(r_int$BCS.M[["OLS"]],r_int$MCS.M[["OLS"]],r_int$BCS.F[["OLS"]],r_int$MCS.F[["OLS"]])
 
@@
 \end{frame}
-% 
-% <<BCSEFA>>=
-% @
-% 
-% <<BCSEFARES, dependson='BCSEFA'>>=
-% @
-% 
-% <<MCSEFA>>=
-% @
-% 
-% <<MCSEFARES, dependson='MCSEFA'>>=
-% @
-% 
-% 
-% \begin{frame}[noframenumbering]
-% \frametitle{EFA}\label{numfac}
-% \framesubtitle{Number of factors \hyperlink{efa}{\beamerreturnbutton{Back}}}
-% \begin{center}
-% <<EFATAB, dependson='MCSEFARES', results='asis'>>=
-% # combine results
-% efatab <- cbind(efatab.bcs,efatab.mcs)
-% colnames(efatab) <- c("BCS All", "BCS M", "BCS F", "MCS All", "MCS M", "MCS F")
-% 
-% 
-% print(xtable(efatab,
-%                     align="L{.25\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}",
-%                     digits = 0
-%                     ),
-%              size="\\small",
-%              include.rownames=T,
-%              include.colnames=T,
-%              booktabs = T,
-%              sanitize.text.function = function(x){x},
-%              tabular.environment = "tabular", width = ".8\\textwidth",
-%              floating = FALSE
-% )
-% @
-% \end{center}
-% \end{frame}
+
+<<BCSEFA>>=
+@
+
+<<BCSEFARES, dependson='BCSEFA'>>=
+@
+
+<<MCSEFA>>=
+@
+
+<<MCSEFARES, dependson='MCSEFA'>>=
+@
+
+\begin{frame}[noframenumbering]
+\frametitle{EFA}\label{numfac}
+\framesubtitle{Number of factors \hyperlink{efa}{\beamerreturnbutton{Back}}}
+\begin{center}
+<<EFATAB, dependson='MCSEFARES', results='asis'>>=
+# combine results
+efatab <- cbind(efatab.bcs,efatab.mcs)
+colnames(efatab) <- c("BCS All", "BCS M", "BCS F", "MCS All", "MCS M", "MCS F")
+
+print(xtable(efatab,
+                    align="L{.25\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}",
+                    digits = 0
+                    ),
+             size="\\small",
+             include.rownames=T,
+             include.colnames=T,
+             booktabs = T,
+             sanitize.text.function = function(x){x},
+             tabular.environment = "tabular", width = ".8\\textwidth",
+             floating = FALSE
+)
+@
+\end{center}
+\end{frame}
 

### 2018-05-23 codes/R/5_graphs.R

@@ -225,7 +225,7 @@ plotineq <- function(data, categ, suffix,
   return(out)
 }
 
-# SOCIAL CLASS
+# SOCIAL CLASS at 10 -------------------------
 # reorder social class to have "other" on the LHS (scl10c)
 scores2plot$scl10c <- factor(scores2plot$scl10b, levels(scores2plot$scl10b)[c(5,1:4)])
 # demean scores for lowest class
@@ -234,20 +234,26 @@ scores2plot <- demean(scores2plot, "scl10b", "IIINM", "dsc")
 fi_sc <- plotineq(scores2plot, "scl10b", "dsc", ylab = "Score (IIINM = 0)", xlab = "Family Social Class at 10",
                   ylim = c(-.6,.4), vline=4.5)
 
+# FATHER SOCIAL CLASS at 5 --------------------
+# demean scores for lowest class
+scores2plot <- demean(scores2plot, "fscl5wb", "Blue collar", "dfsc")
+# make plots
+fi_fsc <- plotineq(scores2plot, "fscl5wb", "dfsc", ylab = "Score (Blue c. = 0)", xlab = "Father Occupation at 5",
+                  ylim = c(-.6,.4), vline=4.5)
 
-# MATERNAL SCHOOLING
+
+# MATERNAL SCHOOLING -------------------------
 # demean scores for lowest level
 scores2plot <- demean(scores2plot, "mysch5b", "17-18", "dys")
 fi_ys <- plotineq(scores2plot, "mysch5b", "dys", ylab = "Score (17-18 = 0)", xlab = "Age mother left FTE",
                   ylim = c(-.6,.4))
 
-# MATERNAL EMPLOYMENT
+# MATERNAL EMPLOYMENT  -------------------------
 scores2plot <- demean(scores2plot, "mempl5", "Unempl./At home", "dme")
 fi_me <- plotineq(scores2plot, "mempl5", "dme", ylab = "Score (Unempl./At home = 0)", xlab = "Maternal employment at 5")
 
-# EMPLOYMENT*EDUCATION
+# EMPLOYMENT*EDUCATION  -------------------------
 # generate interaction
-scores2plot$mempl5b <- dplyr::recode(scores2plot$mempl5, "Full time" = "Employed", "Part time" = "Employed")
 scores2plot$mscem5 <- interaction(scores2plot[c("mempl5b","mpsla5")]) # generate interaction
 levels(scores2plot$mscem5) <- c("Compuls. \n Unempl./Home",
                                 "Compuls. \n Employed",

### 2018-05-24 codes/R/8_regs_outc.R
split load data from regs

@@ -1,7 +1,7 @@
 
 ###################################################################
-## ---- REGS_OUTC
-# REGRESSIONS OF ADOLESCENT OUTCOMES ON EXT/INT & CHILDHOOD VARIABLES
+## ---- LOAD_DATA_OUTC
+# LOAD OUTCOMES DATA
 
 # assemble outcomes data (BCS)
 bcs16outc <- read.dta(paste0(dir_data, "bcsoutc.dta"), convert.factors = F) # all BCS data
@@ -57,6 +57,10 @@ mcsoutclabs <- c(smktry14 = "Tried smoking (14)",
 )
 mcsoutcvars <- names(mcsoutclabs)
 
+
+###################################################################
+## ---- REGS_OUTC
+# REGRESSIONS OF ADOLESCENT OUTCOMES ON EXT/INT & CHILDHOOD VARIABLES
 
 # estimate BCS models
 bcsoutcmod_m_ext <- list()

### 2018-05-24 codes/R/slides/cohorts_tapes.Rnw
update slides

@@
 
+% spacing
+\usepackage{setspace}
+\setstretch{1.3}
+\setbeamerfont{frametitle}{size=\linespread{1.0}\Large}
+
 % define new columns
@@
 \author[Mason]{Orazio Attanasio \and Richard Blundell \\ Gabriella Conti \and Giacomo Mason  }
-\institute[UCL Economics]{UCL Department of Economics}
+\institute[UCL \& IFS]{UCL and IFS}
 
@@
 \framesubtitle{Inequality in skills}
-
 \bi
@@
 \ei
-
+\end{frame}
+
+\begin{frame}
+\frametitle{Motivation}
+\framesubtitle{Measuring socioemotional skills}
+
+\bi
+  \item Usually, socioemotional constructs are measured using validated scales
+    \bi \item e.g. Rutter, BSAG, SDQ \ei
+  \item Most economic literature uses simple sum scores of these scales
+  \item Sometimes useful to consider different populations/groups
+    \bi
+    \item Different countries
+    \item Different subgroups (gender, race, ...)
+    \item Same population, at different points in time
+    \ei
+  \item Do these scales measure the same constructs?
+  \item Are the constructs measured in the same way?
+  \item And what if only info from different scales is available?
+\ei
 \end{frame}
@@
 \framesubtitle{Research questions}
-\bi
-\item Usually, socioemotional constructs are measured using validated scales
-    \bi \item e.g. Rutter, BSAG, SDQ \ei
+We compare the evolution of socioemotional skills in the UK
+\bi
+\item \textbf{Q1}: \textbf{How many latent dimensions} do we need to capture socioemotional skills in the data?
+~\\
 \pause
+\item \textbf{Q2}: Can we derive \textbf{comparable measures} of socioemotional skills across cohorts of children born in different years?
 ~\\
-\item \textbf{Q1}: Can we derive \textbf{comparable measures} of socioemotional skills across populations?
-    \bi
-    \item Most economic literature uses simple sum scores
-    \item But does the same instrument measure the same constructs in the same way:
-      \bi 
-      \item across different populations?
-      \item in the same population, at different points in time?
-      \ei
-    \item And what if only different instruments are available?
-    \ei
 \pause
+\item \textbf{Q3}: How has \textbf{inequality} in early-life socioemotional skills evolved in the UK between 1975 and 2005?
 ~\\
-\item \textbf{Q2}: \textbf{How many latent dimensions} are underlying these instruments?
-    \bi
-    \item Can they be simply summarised into a single score?
-    \ei
+\pause
+\item \textbf{Q4}: Have the \textbf{determinants} of socioemotional skills evolved?
+~\\
+\pause
+\item \textbf{Q5}: How do early socioemotional skills map into \textbf{adolescent/adult outcomes}?
 \ei
@@
   \item Millennium Cohort Study (MCS, b. 2000/1) \ei
-\ei
 ~\\ \pause
-\textbf{Contribution}
+\item \textbf{Contribution}
 \bi
@@
   \bnm 
+  \item \textbf{\textit{Bidimensionality}} -- Externalising and Internalising
   \item \textbf{\textit{Comparability}} in terms of scale (but not location)
-  \item \textbf{\textit{Bidimensionality}} -- Externalising and Internalising
   \enm
@@
 
-\textbf{Key findings}
+\item \textbf{Key findings}
   \bi
@@
   \ei
+\ei
 
@@
 \bi
-\item Four main cohorts:
+\item Four main cohort studies:
   \bi
@@
   \ei
-~\\
-\item Follow 15,000+ individuals from birth onwards
-\item Include extensive data on
+\item Nationally representative samples of 10,000+ individuals from birth
   \bi
@@
   \ei
-~\\
 \pause
-\item Not always consistent
+\vspace{0.2cm}
+\item Not always \textbf{consistent}
   \bi
   \item Different design, timing, questions, survey instruments...
-  \item Harmonisation efforts (CLOSER)
+  \item Harmonisation effort underway (CLOSER)
   \ei
@@
 \item Data from 5 year follow-up
-\item Sample restricted to England
+  \bi \item Sample restricted to England \ei
 \item Socioemotional skills: \textbf{Rutter A scale} \citep{Rutter1970}
@@
 \item Data from 5 year follow-up
+  \bi
   \item Sample restricted to England
   \item Complex survey design: rebalance sample to make representative
+  \ei
 \item Socioemotional skills: \textbf{Strengths and Difficulties Questionnaire} (SDQ, \citealp{Goodman1994})
@@
 
-\section{Harmonised scale}
-
-\subsection{Common items}
+\section{Deriving a harmonised scale}
 
@@
 \item Rutter and SDQ are different instruments
+  \bi \item Not ideal to just use sum scores and assume they return comparable information... \ei
 \item Focus on subset of items that have similar wording between the two instruments
-\item New 11-item subscale
+\item $\Rightarrow$ New 11-item subscale
 \ei
@@
 
-\begin{frame}
-\frametitle{Harmonisation}\label{meantab_f}
-\framesubtitle{Choice of common items (females) \hyperlink{meantab_m}{\beamerbutton{Males}}}
-\centering
-\resizebox{!}{4cm}{
-<<MEANTAB_F, dependson="MEANTABLE", results='asis'>>=
-require(xtable)
-
-table <- meantab[[2]] # females
-table <- table[,!names(table) %in% c("cats", "fac", "bcs.num", "mcs.num")] # drop if unneeded
-
-print(xtable(table,
-                    align="L{0cm}C{.04\\textwidth}L{.4\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}L{.4\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
+<<MEANTABLE_PRINT>>=
+# function to print mean table
+printmeantab <- function(tab) {
+  print(xtable(tab,
+               align="L{0cm}C{.04\\textwidth}L{.45\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}L{.45\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
                digits = 1
@@
   )
+}
+
+@
+
+\begin{frame}
+\frametitle{Harmonisation}\label{meantab_f}
+\framesubtitle{Choice of common items (females) \hyperlink{meantab_m}{\beamerbutton{Males}}}
+\begin{spacing}{1.1}
+\centering
+\resizebox{!}{4cm}{
+<<MEANTAB_F, dependson="MEANTABLE", results='asis'>>=
+require(xtable)
+
+table <- meantab[[2]] # females
+table <- table[,!names(table) %in% c("cats", "fac", "bcs.num", "mcs.num")] # drop if unneeded
+
+printmeantab(table)
 
@@
 } %scalebox
-\end{frame}
-
-\subsection{Exploratory Analysis}
+\end{spacing}
+\end{frame}
 
@@
 
-How many dimensions underlie the new 11-item scale?
+\textbf{Q1}: How many dimensions underlie the new 11-item scale?
 \bi
@@
     \ei
-  \item Sometimes the former is split into aggressive-antisocial and hyperkinetic \citep{McGee1985}
+  \item Sometimes the former is split into aggressive/conduct vs hyperkinetic \citep{McGee1985,Collishaw2004}
   \ei
@@
 
-On our data:
 \bi
@@
 \centering
+\begin{spacing}{1.1}
 \begin{tabular}{C{.05\textwidth}L{.3\textwidth}|C{.05\textwidth}L{.3\textwidth}}
@@
 \end{tabular}
+\end{spacing}
 
@@
 \framesubtitle{Choice of common items (females) \hyperlink{meantab_fcc3}{\beamerbutton{3 factors}}}
+\begin{spacing}{1.1}
 \centering
@@
 
-print(xtable(table,
-                    align="L{0cm}C{.04\\textwidth}L{.4\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}L{.4\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
-                    digits = 1
-                    ), 
-             size="\\small",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = c(list(0)),
-                              command = c(
-                              # header
-                                    paste(" & 
-                                              \\multicolumn{4}{c}{\\textbf{BCS}} &
-                                              \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9} 
-                                            Item &
-                                              Text & Cert. Appl. & Smt. Appl. & Appl. &
-                                              Text & Cert. True & Smt. True & True \\\\ \\midrule
-                                          "
-                                            , sep="")
-                                    )),
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabular", width = ".8\\textwidth",
-             floating = FALSE
-)
+printmeantab(table)
 
@@
 } %scalebox
+\end{spacing}
 \end{frame}
@@
   \item Formal test of the invariance hypothesis using factor analysis
-  \item Aim: assess if constructs are measured in the same way in two or more groups
+  \item Aim: assess if constructs are measured in the same way in 2+ groups
   \item Procedure:
@@
 \framesubtitle{IFA model}
+\setstretch{1.1}
+
 Categorical Item Factor Analysis (IFA) model
-
-\bi
-\item Threshold model \citep{Muthen1984}
-\item Latent propensity to endorse each item $i$ for child $j$ in group $c$
+\bi
+\item Latent propensity to endorse each item $i$ for child $j$ in group $c$ \citep{Muthen1984}
 \begin{align}\label{eq:propens}
@@
 \ei
-~\\
+\vspace{.2cm}
 \bnm
@@
 \begin{center}
+\resizebox{\textwidth}{!}{
 <<FA_FIT_RES, dependson="FA_FIT", results='asis'>>=
@@
 table <- rbind(afitab[[1]])
-table <- as.matrix(table[,c("mnames","npar","rmsea")])
+table <- merge(table,afitab[[2]], by="mnames") # merge additional result
+table <- as.matrix(table[,c("mnames","npar.x","chisq.x","rmsea.x","chisq.y","rmsea.y")])
 
@@
 print(xtable(table,
-                    align="L{0cm}L{.5\\textwidth}C{.1\\textwidth}C{.15\\textwidth}",
+                    align="L{0cm}L{.35\\textwidth}C{.12\\textwidth}C{.15\\textwidth}C{.1\\textwidth}C{.15\\textwidth}C{.1\\textwidth}",
                     digits = 4
@@
                               # header
-            "Model & Num. params & RMSEA \\\\"
+            " & & \\multicolumn{2}{c}{\\textbf{Full sample}} & \\multicolumn{2}{c}{\\textbf{59-61 months}} \\\\ \\cmidrule(lr){3-4}  \\cmidrule(lr){5-6}
+            Model & Num. par. & $\\chi^2$ & RMSEA & $\\chi^2$ & RMSEA \\\\"
                                     )),
@@
 @
+}
 \end{center}
 \end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% SCORES
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
   \bi
-  \item Scores from the model exploit the covariance structure of the whole 11-item scale
-  \item Every pattern of response is assigned a score
+  \item Sum scores assume that each item conveys the same information
+  \item Scores obtained from the model exploit the covariance structure of the whole 11-item scale
+  \item Every 11-dimensional pattern of response in the data can be assigned a different score
   \item Substantial increase in variability ({\color{red} could add boxplot of factor scores vs sum scores})
@@
 \end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% SCORES
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\section{Inequality in behaviour problems}
+\section{Inequality in socioemotional skills}
 
@@
 \bi
-\item \textbf{Q2}: Has socioeconomic inequality in behavioural problems increased between 1975 and 2005?
-\item Investigate income gradient in externalising and internalising scores
-  \bi
-  \item Merge harmonised income data from CLOSER Work Package 2(a)
-  \item Weekly family income when child is 10 (not contemporaneous)
-  \item Working on access to harmonised social class data
-  \ei
+\item \textbf{Q3}: How has socioeconomic inequality in socioemotional skills increased between 1975 (BCS) and 2005 (MCS)?
+\item 
 \ei
@@
 \frametitle{Inequality}\label{ineq_uc}
-\framesubtitle{Unconditional inequality}
+\framesubtitle{Unconditional (quantile differences)}
 \centering
@@
                     align="L{0cm}L{.15\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}",
-                    digits = 4
+                    digits = 3
                     ),
@@
 \begin{frame}
-\frametitle{Inequality}\label{ineq_sc}
-\framesubtitle{Social class}
-\vspace{-.2cm}
-\begin{figure}
-<<FACINEQ_SC, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
-# plot
-fi <- fi_sc # select result
-gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
-                        fi$ineqlist[[3]],fi$ineqlist[[4]],
-                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
-                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
-@
-\end{figure}
-\end{frame}
-
-\begin{frame}
 \frametitle{Inequality}\label{ineq_fsc}
-\framesubtitle{Father occupation}
+\framesubtitle{Father occupation \hyperlink{ineq_sc}{\beamerbutton{Social class (10)}}}
 \vspace{-.2cm}
@@
 \begin{frame}
-\frametitle{Inequality}\label{ineq_ys}
-\framesubtitle{Age mother left education}
+\frametitle{Inequality}\label{ineq_ps}
+\framesubtitle{Mother compulsory education \hyperlink{ineq_ys}{\beamerbutton{Age left education}}}
 \vspace{-.2cm}
 \begin{figure}
-<<FACINEQ_YS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
+<<FACINEQ_PS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
 # plot
-fi <- fi_ys # select result
+fi <- fi_ps # select result
 gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
@@
 
+\begin{frame}
+\frametitle{Inequality}\label{ineq_sm}
+\framesubtitle{Smoking in pregnancy}
+\vspace{-.2cm}
+\begin{figure}
+<<FACINEQ_SM, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=11, out.width='\\linewidth'>>=
+# plot
+fi <- fi_sm # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{Inequality}
+
+\textbf{Q3}: How has \textbf{inequality} in early-life socioemotional skills evolved in the UK between 1975 and 2005?
+
+\bi
+\item Unconditionally, the distribution is more spread out (except for girls' internalising)
+\item Inequality across socioeconomic indicators has increased, especially
+  \bi
+  \item for boys rather than girls
+  \item considering inequality across maternal schooling and employment, rather than paternal occupation
+  \ei
+\ei
+
+\end{frame}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Determinants of socioemotional skills}
 
@@
 \begin{frame}
-\frametitle{What explains behavioural skills?}
-
-write about variables used and groups
-
-\end{frame}
-
-\begin{frame}
-\frametitle{What explains socioemotional skills?}
+\frametitle{Determinants of socioemotional skills}
+
+We explain externalising and internalising skills using information from the birth and 5 year surveys \\
+
+\begin{table}
+\footnotesize
+\begin{tabular}{L{.25\textwidth}C{.1\textwidth}L{.5\textwidth}}
+Group & Age &  \\
+\midrule
+Father occupation & 5 & White collar (SC I-IIINM) \newline Blue collar (SC IIIM-V-other) \newline No father figure \\
+Maternal education & 5 & Post-compulsory schooling \\
+Maternal employment & 5 & Employed \\
+Maternal background & 0/5 & Age, height \newline Nonwhite ethnicity (child) \newline Number of children in HH \\
+Pregnancy & 0 & Firstborn child, number of previous stillbirths \newline Mother smoked in pregnancy \newline Preterm birth, (log) birth weight \\
+\end{tabular}
+\end{table}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Determinants of socioemotional skills}
 \framesubtitle{Externalising}
@@
 \centering
+\setstretch{1.0}
 \resizebox{\textwidth}{!}{
@@
                caption = "", label = "",
-               omit.coef = "region|Inter",
+               omit.coef = "region|Inter|numch5|mheight|nprevst|preterm",
                groups = grouplabs,
@@
                custom.model.names = c("Males BCS", "Males MCS", "Females BCS", "Females MCS"),
+               custom.note = "",
                include.rmse = FALSE,
@@
 }
+\\[-12pt] \justify {\scriptsize \emph{Notes}: Robust standard errors in parentheses.
+All estimates additionally control for region of birth, mother height, number of previous stillbirths at child's birth, preterm birth, a dummy for missing gestational age, and number of other children in the household at child age 5.
+\par}
 \end{table}
@@
 \begin{frame}
-\frametitle{What explains socioemotional skills?}
+\frametitle{Determinants of socioemotional skills}
 \framesubtitle{Internalising}
+\vspace{-.3cm}
 \begin{table}
+\setstretch{1.0}
 \centering
@@
 \begin{frame}
-\frametitle{Decomposition (EXT)}\label{gelb_ext}
-\framesubtitle{Explaining the education gradient \hyperlink{gelbsc_ext}{\beamerbutton{Social class}}}
-\centering
-\scriptsize
-\resizebox{\textwidth}{!}{
-\input{../../../tables/med_gelb_ext}
-}
-\end{frame}
-
-\begin{frame}
-\frametitle{Decomposition (INT)}\label{gelb_int}
-\framesubtitle{Explaining the education gradient \hyperlink{gelbsc_int}{\beamerbutton{Social class}}}
-\centering
-\scriptsize
-\resizebox{\textwidth}{!}{
-\input{../../../tables/med_gelb_int}
-}
+\frametitle{Decomposition}
+\framesubtitle{Gelbach (2016)}
+
+write up Gelbach approach...
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Decomposition (EXT)}\label{gelb_ext_f}
+\framesubtitle{Explaining the gradient by maternal compulsory schooling \hyperlink{gelb_ext}{\beamerbutton{Table}}}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Decomposition (INT)}\label{gelb_int_f}
+\framesubtitle{Explaining the gradient by maternal compulsory schooling \hyperlink{gelb_int}{\beamerbutton{Table}}}
+
 \end{frame}
@@
 
-<<REGS_OUTC, dependson="SELECTVARS">>=
+\section{Mapping skills to later outcomes}
+
+<<LOAD_DATA_OUTC, dependson="SELECTVARS">>=
+# rload outcomes data
+@
+
+<<REGS_OUTC, dependson="LOAD_DATA_OUTC">>=
 # run regressions explaining adult outcomes
@@
 \begin{frame}
-\frametitle{Outcomes}
+\frametitle{Mapping skills into behaviours}
 \framesubtitle{Adolescence (MCS)}
 \centering
+\setstretch{1.0}
 \resizebox{\textwidth}{!}{
@@
 \begin{frame}[noframenumbering]
-\frametitle{Outcomes}
+\frametitle{Mapping skills into behaviours}
 \framesubtitle{Adolescence (MCS)}
 \centering
+\setstretch{1.0}
 \resizebox{\textwidth}{!}{
@@
 \begin{frame}
-\frametitle{Outcomes}
+\frametitle{Mapping skills into behaviours}
 \framesubtitle{Adolescence (BCS)}
 \centering
+\setstretch{1.0}
 \resizebox{\textwidth}{!}{
@@
 \begin{frame}[noframenumbering]
-\frametitle{Outcomes}
+\frametitle{Mapping skills into behaviours}
 \framesubtitle{Adolescence (BCS)}
 \centering
+\setstretch{1.0}
 \resizebox{\textwidth}{!}{
@@
 \begin{frame}
-\frametitle{Outcomes}
+\frametitle{Mapping skills into labour market}
 \framesubtitle{Adulthood (BCS)}
 \centering
+\setstretch{1.0}
 \resizebox{\textwidth}{!}{
@@
 \begin{frame}[noframenumbering]
-\frametitle{Outcomes}
+\frametitle{Mapping skills into labour market}
 \framesubtitle{Adulthood (BCS)}
 \centering
+\setstretch{1.0}
 \resizebox{\textwidth}{!}{
@@
 \framesubtitle{Choice of common items (males) \hyperlink{meantab_f}{\beamerreturnbutton{Females}}}
+\begin{spacing}{1.1}
 \centering
@@
 } %scalebox
-\end{frame}
-
-<<BCSEFA>>=
+\end{spacing}
+\end{frame}
+
+<<BCSEFA, dependson='LOAD_DATA'>>=
 @
@@
 
-<<MCSEFA>>=
+<<MCSEFA, dependson='LOAD_DATA'>>=
 @
@@
 \framesubtitle{Number of factors \hyperlink{efa}{\beamerreturnbutton{Back}}}
+\setstretch{1.1}
 \begin{center}
@@
 \framesubtitle{Choice of common items (females) \hyperlink{meantab_fcc2}{\beamerreturnbutton{2 factors}}}
+\begin{spacing}{1.1}
 \centering
@@
 \framesubtitle{Invariance steps \hyperlink{misteps}{\beamerreturnbutton{Back}}}
+\setstretch{1.0}
 \Fontvi
@@
 \framesubtitle{Estimated measurement parameters \hyperlink{mires}{\beamerreturnbutton{Back}}}
+\setstretch{1.1}
 <<RES_FA_PAR, dependson="FA_FIT", results='asis'>>=
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% GELBACH DECOMPOSITION WITH SOCIAL CLASS
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{frame}[noframenumbering]
-\frametitle{Decomposition (EXT)}\label{gelbsc_ext}
-\framesubtitle{Explaining the education gradient \hyperlink{gelb_ext}{\beamerreturnbutton{Back}}}
+%%%%%%%% ADDITIONAL INEQ GRAPHS
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}[noframenumbering]
+\frametitle{Inequality}\label{ineq_sc}
+\framesubtitle{Social class \hyperlink{ineq_fsc}{\beamerreturnbutton{Back}}}
+\vspace{-.2cm}
+\begin{figure}
+<<FACINEQ_SC, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
+# plot
+fi <- fi_sc # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\end{figure}
+\end{frame}
+
+\begin{frame}[noframenumbering]
+\frametitle{Inequality}\label{ineq_ys}
+\framesubtitle{Age mother left education \hyperlink{ineq_ps}{\beamerreturnbutton{Back}}}
+\vspace{-.2cm}
+\begin{figure}
+<<FACINEQ_YS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
+# plot
+fi <- fi_ys # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\end{figure}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% GELBACH DECOMPOSITION
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}[noframenumbering]
+\frametitle{Decomposition (EXT)}\label{gelb_ext}
+\framesubtitle{Explaining the education gradient \hyperlink{gelb_ext_f}{\beamerreturnbutton{Back}}}
 \centering
 \scriptsize
+\setstretch{1.1}
 \resizebox{\textwidth}{!}{
-\input{../../../tables/med_gelb_sc_ext}
-}
-\end{frame}
-
-\begin{frame}[noframenumbering]
-\frametitle{Decomposition (INT)}\label{gelbsc_int}
-\framesubtitle{Explaining the education gradient \hyperlink{gelb_int}{\beamerreturnbutton{Back}}}
+\input{../../../tables/med_gelb_ext}
+}
+\end{frame}
+
+\begin{frame}[noframenumbering]
+\frametitle{Decomposition (INT)}\label{gelb_int}
+\framesubtitle{Explaining the education gradient \hyperlink{gelb_int_f}{\beamerreturnbutton{Back}}}
 \centering
 \scriptsize
+\setstretch{1.1}
 \resizebox{\textwidth}{!}{
-\input{../../../tables/med_gelb_sc_int}
-}
-\end{frame}
+\input{../../../tables/med_gelb_int}
+}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% BIBLIOGRAPHY
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2018-05-24 codes/R/9_gelbfig.R
sketch gelbach figure

@@ -1,0 +1,18 @@
+
+
+gelbdata <- read.dta(paste0(dir_data, "gelbres.dta"), convert.factors = F) # all BCS data
+
+# keep only conditional and explained 
+gelbdata <- gelbdata[gelbdata$varname %in% c("fulleff","med1","med2","med3","med4"),]
+gelbdata$varname[gelbdata$varname=="fulleff"] <- "med0"
+
+gelbdata$M_BCS <- as.numeric(substr(gelbdata$coef_1_1,2,7))
+gelbdata$M_MCS <- as.numeric(substr(gelbdata$coef_2_1,2,7))
+gelbdata$F_BCS <- as.numeric(substr(gelbdata$coef_1_2,2,7))
+gelbdata$F_MCS <- as.numeric(substr(gelbdata$coef_2_2,2,7))
+
+gelbdata_ext <- gelbdata[gelbdata$outcome=="EXT",names(gelbdata) %in% c("varname", "M_BCS", "M_MCS", "F_BCS", "F_MCS")]
+
+gelbdata_ext <- reshape2::melt(gelbdata_ext, id=c("varname"))
+colnames(gelbdata_ext) <- c("vgroup", "cohortsex", "value")
+ggplot(data = gelbdata_ext, aes(x = cohortsex, y= value, fill=vgroup)) + geom_bar(stat = "identity")

### 2018-05-31 codes/R/4_scores.R
lavaan update new lavpredict syntax

@@ -3,7 +3,7 @@
 ############################################################################################
 ## ---- FA_SCORES_MAIN
 # score model selected as final
-scores.final <- predict(finalmod, newdata = items[[1]])
+scores.final <- lavPredict(finalmod, newdata = items[[1]])
 
 ############################################################################################
 ## ---- FA_SCORES_BIND

### 2018-05-31 codes/R/slides/cohorts_tapes.Rnw
slides update after Hoynes questions

@@
 read_chunk(here("codes/R/8_regs_outc.R"))
+read_chunk(here("codes/R/9_gelbfig.R"))
 read_chunk(here("codes/R/getpars.R"))
@@
 \frametitle{Motivation}
-\framesubtitle{Inequality in skills}
-\bi
-\item Socio-emotional skills developed during childhood are important for lifetime outcomes
-\item Socioeconomic gradients in these skills are persistent 
-\item {\color{red} IMPROVE INTRO...}
+\framesubtitle{Inequality in socioemotional skills}
+\bi
+\item Socioemotional skills developed during childhood are important for lifetime outcomes \citep{Lundberg2017}
+\item The UK has witnessed sharp increases in income inequality and decreases in social mobility since the 1970s
+\item What happened to inequalities in early childhood development in this period?
 \ei
@@
 \bi
+  \item Accurate, consistent measurement of skills is fundamental
   \item Usually, socioemotional constructs are measured using validated scales
-    \bi \item e.g. Rutter, BSAG, SDQ \ei
-  \item Most economic literature uses simple sum scores of these scales
-  \item Sometimes useful to consider different populations/groups
-    \bi
-    \item Different countries
-    \item Different subgroups (gender, race, ...)
-    \item Same population, at different points in time
-    \ei
-  \item Do these scales measure the same constructs?
-  \item Are the constructs measured in the same way?
+    \bi 
+    \item e.g. Rutter, BSAG, SDQ 
+    \item Obtained by aggregating different items
+    \ei
+  \item Can we use these scales to compare different cohorts, by gender?
   \item And what if only info from different scales is available?
+    \bi
+    \item Can different scales measure the \textbf{same constructs}?
+    \item Can the constructs be measured \textbf{in comparable ways}?
+    \ei
 \ei
@@
 \framesubtitle{Research questions}
-We compare the evolution of socioemotional skills in the UK
-\bi
-\item \textbf{Q1}: \textbf{How many latent dimensions} do we need to capture socioemotional skills in the data?
+We compare the evolution of inequality in socioemotional skills at age 5 across two UK cohorts born 30 years apart
+\bi
+\item \textbf{Q1}: \textbf{How many latent dimensions} underlie the available survey instruments?
 ~\\
 \pause
-\item \textbf{Q2}: Can we derive \textbf{comparable measures} of socioemotional skills across cohorts of children born in different years?
+\item \textbf{Q2}: Can we derive \textbf{comparable measures} of socioemotional skills?
 ~\\
 \pause
-\item \textbf{Q3}: How has \textbf{inequality} in early-life socioemotional skills evolved in the UK between 1975 and 2005?
+\item \textbf{Q3}: How has \textbf{inequality} in early-life socioemotional skills evolved?
 ~\\
 \pause
-\item \textbf{Q4}: Have the \textbf{determinants} of socioemotional skills evolved?
+\item \textbf{Q4}: Have the \textbf{determinants} of socioemotional skills changed?
 ~\\
 \pause
-\item \textbf{Q5}: How do early socioemotional skills map into \textbf{adolescent/adult outcomes}?
+\item \textbf{Q5}: Has the mapping of socioemotional skills into \textbf{adolescent/adult outcomes} changed?
 \ei
@@
 \item Focus on socioemotional skills in two British cohorts 
-  \bi \item British Cohort Study (BCS, b. 1970)
-  \item Millennium Cohort Study (MCS, b. 2000/1) \ei
-~\\ \pause
-\item \textbf{Contribution}
-\bi
-\item Construct a common 11-item scale, starting from different survey instruments
-\item Investigate its psychometric properties, and find
+  \bi \item British Cohort Study (b. 1970)
+  \item Millennium Cohort Study (b. 2000/1) \ei
+\item \textbf{Methodological contribution}
+  \bi
+  \item Construct a common 11-item scale from different survey instruments
     \bnm 
     \item \textbf{\textit{Bidimensionality}} -- Externalising and Internalising
-  \item \textbf{\textit{Comparability}} in terms of scale (but not location)
+    \item \textbf{\textit{Comparability}} in terms of variance (but not mean)
     \enm
   \ei
-~\\ \pause
-
 \item \textbf{Key findings}
@@
   \item Sharpest change for boys rather than girls
-  \item {\color{red} OTHER RESULTS}
-  \ei
-\ei
-
+  \item Increased importance of pregnancy factors, decreased importance of paternal education, in explaining skill inequality by maternal education
+  \item In both cohorts, socioemotional skills (especially externalising) are significant predictors of social and health outcomes in adolescence
+  \ei
+\ei
 \end{frame}
@@
   \item Perinatal period (pregnancy, birth outcomes)
-  \item Skill formation -- cognitive, socioemotional
+  \item Skill formation -- cognitive, socioemotional \citep{Joshi2014}
   \item Health and health behaviours
@@
 \item Focus on subset of items that have similar wording between the two instruments
+\item Recode items to enhance comparability when needed
+  \bi \item E.g. ``Generally obedient" dichotomised to match ``Often disobedient"  \ei
 \item $\Rightarrow$ New 11-item subscale
@@
 \begin{frame}
-\frametitle{Factor structure}
-\framesubtitle{Existing evidence}
+\frametitle{Exploratory FA}\label{efa1}
+\framesubtitle{Dimensionality}
 
 \textbf{Q1}: How many dimensions underlie the new 11-item scale?
-\bi
-\item Rutter A scale (BCS) 
-  \bi 
-  \item Traditionally conceptualised as bidimensional \citep{Blanden2007}
-    \bi
-    \item \emph{anti-social} -- or aggressive-hyperactive
-    \item \emph{neurotic} -- or anxious-fearful
-    \ei
-  \item Sometimes the former is split into aggressive/conduct vs hyperkinetic \citep{McGee1985,Collishaw2004}
-  \ei
-\item Strengths and difficulties (MCS) 
-  \bi
-  \item Five subscales (emotional, hyperactivity, prosocial, peer, conduct) \citep{Stone2010}
-  \item Evidence that lower-dimensional structures can work \citep{Goodman2010}
-  \ei
-\ei
-
-\end{frame}
-
-\begin{frame}
-\frametitle{EFA}\label{efa}
+
+\bi
+\item Factor analysis routinely used to construct measures of socioemotional skills \citep{Conti2014,Attanasio2017c}
+\item Starts with \emph{exploratory} step
+\item Decompose the covariance matrix of items to assess:
+  \bi \item the dimensionality of the underlying factor structure (how many latent factors)
+  \item which items measure which latent factor \ei
+\item Psychometric literature often finds \textbf{two latent factors} provide good synthesis for Rutter and SDQ \hyperlink{efalit}{\beamerbutton{Det.}}
+\item This can be tested in our data
+\ei
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Exploratory FA}\label{efa2}
 \framesubtitle{Dimensionality}
 
-\bi
-\item Different methods for assessing number of factors give varied results \hyperlink{numfac}{\beamerbutton{Det.}}
-\item We choose two factors to maximise interpretability
-\item In both cohorts, EFA delivers a clear-cut split between two factors
-\item Factors can be coherently interpreted as externalising ($\sim$ anti-social) and internalising ($\sim$ neurotic)
-\ei 
-\vspace{12pt}
-\centering
-\begin{spacing}{1.1}
+\textbf{Q1}: How many dimensions underlie the new 11-item scale?
+
+\bi
+\item Dimensionality
+  \bi
+  \item Methods for assessing dimensionality imply 1-3 factors \hyperlink{numfac}{\beamerbutton{Det.}}
+  \item We choose \textbf{two factors} to maximise interpretability
+  \ei
+\item Structure
+  \bi
+  \item Exploratory FA using weighted least squares and \textit{oblimin} rotation
+  \item Factors can be coherently interpreted as \textbf{externalising} ($\sim$ anti-social) and \textbf{internalising} ($\sim$ neurotic)
+  \ei
+\ei 
+\pause
+\vspace{6pt}
+\centering
+\small
+\begin{spacing}{1.0}
 \begin{tabular}{C{.05\textwidth}L{.3\textwidth}|C{.05\textwidth}L{.3\textwidth}}
@@
 \bi
-\item Two constructs isolated in the new scale, separately by cohort
+\item Two constructs uncovered in the new scale, separately by cohort
 \item But are externalising and internalising measured in a comparable way \emph{across cohorts}?
@@
 \vspace{12pt}\pause
-\textbf{Measurement Invariance (MI)} analysis \citep{Wu2016a}
+\textbf{Measurement Invariance (MI)} analysis
   \bi
@@
     \bnm
-    \item Start from the most general model -- parameters allowed differ across groups
+    \item Start from the most general model -- parameters allowed to differ across groups
     \item Impose restrictions on sets of parameters to be equal across groups
@@
 \frametitle{Measurement Invariance}\label{midet}
-\framesubtitle{IFA model}
+\framesubtitle{IFA model \hyperlink{midet_app}{\beamerbutton{Details}}}
 \setstretch{1.1}
@@
 \bi
+\item Observed categorical items $X_1$-$X_{11}$ measuring latent factor $\theta$
 \item Latent propensity to endorse each item $i$ for child $j$ in group $c$ \citep{Muthen1984}
-\begin{align}\label{eq:propens}
-\begin{split}
-X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{EXT}_{jc} + u_{ijc} \qquad \text{for} \; i=1,\dots, 6 \\
-X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{INT}_{jc} + u_{ijc} \qquad \text{for} \; i=7,\dots, 11
-\end{split}
-\end{align}
-
-\item Mean/covariance structure
-$$\bm{\mu}_c=\bm{\nu}_c + \bm{\Lambda}_c\bm{\kappa}_c \qquad \text{and} \qquad \Sigma_c = \bm{\Lambda}_c \bm{\Phi}_c \bm{\Lambda}_c^\prime + \bm{\Psi}_c.$$
-with $$(\theta^{EXT}_{jc}, \theta^{INT}_{jc}) = \bm{\Theta}_{jc} \sim N(\bm{\kappa}_c, \bm{\Phi}_c) \qquad \text{and} \qquad \bm{u}_{jc} \sim N(\bm{0}, \bm{\Psi}_c)$$
-
-\item Discreteness
+$$ X^*_{ijc} =\nu_{ci} + \lambda_{ci}\theta_{jc} + u_{ijc} $$
+\item Discreteness \citep{Piatek2016}
 $$X_{ijc} = s \qquad \text{if} \; \tau_{s,ic} \leq X^*_{ijc} < \tau_{s+1,ic} \qquad \text{for} \; s=0,1,2$$
 with $\tau_{0,ic}=-\infty$ and $\tau_{3,ic}=+\infty$.
-
-\ei
-\end{frame}
-
-\begin{frame}
+\vspace{.2cm}
+\item We follow the approach in \cite{Wu2016a}
+  \bi
+  \item Define group ($c$) as cohort-gender cell
+  \ei
+\ei
+\end{frame}
+
+\begin{frame}[t]
 \frametitle{Measurement Invariance}\label{misteps}
 \framesubtitle{Steps \hyperlink{miapp1}{\beamerbutton{Details}}}
-
-We follow the approach in \cite{Wu2016a}
-\bi
-\item Define group ($c$) as cohort-gender cell
-  \bi \item Invariance results do not vary if separating by gender \ei
-\ei
-\vspace{.2cm}
+$$ X^*_{ijc} =
+\only<1>{\textcolor{red}{\nu_{ci}} + \textcolor{red}{\lambda_{ci}}\theta_{jc} + u_{ijc}} 
+\only<2>{\textcolor{red}{\nu_{ci}} + \textcolor{green}{\lambda_{i}}\theta_{jc} + u_{ijc}} 
+\only<3>{\textcolor{green}{\nu_{i}} + \textcolor{green}{\lambda_{i}}\theta_{jc} + u_{ijc}} 
+$$
+$$X_{ijc} = s \qquad \text{if} \; 
+\only<1>{\textcolor{red}{\tau_{s,ic}} \leq X^*_{ijc} < \textcolor{red}{\tau_{s+1,ic}}}
+\only<2>{\textcolor{green}{\tau_{s,i}} \leq X^*_{ijc} < \textcolor{green}{\tau_{s+1,i}}}
+\only<3>{\textcolor{green}{\tau_{s,i}} \leq X^*_{ijc} < \textcolor{green}{\tau_{s+1,i}}}
+\qquad \text{for} \; s=0,1,2$$
+\vspace{-.2cm}
 \bnm
@@
   \item Minimally restrictive (all parameters free across groups)
-  \item Normalised on latent means $\bm{\kappa}$ and variances $\bm{\Phi}$
-  \ei
+  \ei
+  \pause
 \item \textbf{Threshold and Loading invariance} model
@@
   \ei
+  \pause
 \item \textbf{Threshold, Loading, and Intercept invariance} model
@@
 \bi
-\item We use RMSE to compare fit between models of different invariance level
-\item Restricting thresholds and loadings provides a good fit \hyperlink{fit}{\beamerbutton{Fit details}}
+\item We use RMSE to compare fit between models with different invariance restriction
+\item Restricting only thresholds and loadings provides a good fit \hyperlink{fit}{\beamerbutton{Fit details}}
   \bi \item We can compare the scale (variance) of the latent factors \ei
@@
                               # header
-            " & & \\multicolumn{2}{c}{\\textbf{Full sample}} & \\multicolumn{2}{c}{\\textbf{59-61 months}} \\\\ \\cmidrule(lr){3-4}  \\cmidrule(lr){5-6}
+            " & & \\multicolumn{2}{c}{\\textbf{Full sample}} & \\multicolumn{2}{c}{\\textbf{59-61 months} \\hyperlink{agehist}{\\beamerbutton{Hist.}}} \\\\ \\cmidrule(lr){3-4}  \\cmidrule(lr){5-6}
             Model & Num. par. & $\\chi^2$ & RMSEA & $\\chi^2$ & RMSEA \\\\"
@@
 \begin{frame}
-\frametitle{Measurement Invariance}
+\frametitle{Measurement Invariance}\label{miscores}
 \framesubtitle{Factor scores}
-\bi
-\item Nevertheless, we can use the model with non-invariant intercepts to score the latent factors (externalising and internalising dimensions)
+
+\textbf{Q2}: Can we derive \textbf{comparable measures} of socioemotional skills?
+\bi
+\item We can use the model with non-invariant intercepts to score the latent factors (externalising and internalising dimensions)
 \item Estimated parameters: \hyperlink{measpars}{\beamerbutton{Measurement}}\hyperlink{lvpars}{\beamerbutton{Latent variable}}
-\item Use Empirical Bayes approach to predict individual scores
-\item All scores are coded so that \emph{higher is better}
-\item \textbf{CAUTION}: location differences by cohort are not interpretable!
-~\\
-\pause
+\item Use Empirical Bayes approach to predict individual scores \citep{Wu2013}
+\item All scores are coded so that \emph{\textbf{a higher value means better socioemotional skills}}
+\item \textbf{CAUTION}: location differences by cohort/gender are not interpretable!
+\ei
+\end{frame}
+
+\begin{frame}
+\frametitle{Measurement Invariance}\label{miscores2}
+\framesubtitle{Factor scores vs. Sum scores}
+\bi
 \item Is this an improvement on using simple sum scores?
   \bi
-  \item Sum scores assume that each item conveys the same information
+  \item Sum scores assume that each item conveys the same information, and that everything is invariant
   \item Scores obtained from the model exploit the covariance structure of the whole 11-item scale
   \item Every 11-dimensional pattern of response in the data can be assigned a different score
-  \item Substantial increase in variability ({\color{red} could add boxplot of factor scores vs sum scores})
+  \item Substantial increase in variability \hyperlink{facraw}{\beamerbutton{Graph}}
   \ei
@@
 
-\bi
-\item \textbf{Q3}: How has socioeconomic inequality in socioemotional skills increased between 1975 (BCS) and 2005 (MCS)?
-\item 
+\textbf{Q3}: How has \textbf{inequality} in early-life socioemotional skills evolved in the UK between 1975 and 2005?
+
+\bi
+\item Previous work in the mental health literature has mostly looked at secular trends in socioemotional skills
+  \bi 
+  \item Focus on adolescent period \citep{Collishaw2004,Collishaw2010a}
+  \item Mixed findings -- often stable for boys, increasing for girls \citep{Maughan2005,Bor2014}
+  \ei
+\item Less attention devoted to secular changes in 
+  \bi
+  \item socioemotional skills in the pre-school period
+  \item socioeconomic inequality in skills
+  \item early-life determinants of skills
+  \ei
+\item We look at
+  \bi 
+  \item Distribution (quantile differences)
+  \item Mean skills by parents' socioeconomic background
+  \ei
 \ei
@@
                     align="L{0cm}L{.15\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}",
-                    digits = 3
+                    digits = 2
                     ),
@@
 \begin{figure}
-<<FACINEQ_SM, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=11, out.width='\\linewidth'>>=
+<<FACINEQ_SM, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
 # plot
@@
   \item for boys rather than girls
-  \item considering inequality across maternal schooling and employment, rather than paternal occupation
+  \item considering inequality by maternal schooling and employment, rather than paternal occupation
   \ei
@@
 \begin{frame}
-\frametitle{Determinants of socioemotional skills}
-
-We explain externalising and internalising skills using information from the birth and 5 year surveys \\
-
-\begin{table}
-\footnotesize
-\begin{tabular}{L{.25\textwidth}C{.1\textwidth}L{.5\textwidth}}
-Group & Age &  \\
-\midrule
-Father occupation & 5 & White collar (SC I-IIINM) \newline Blue collar (SC IIIM-V-other) \newline No father figure \\
-Maternal education & 5 & Post-compulsory schooling \\
-Maternal employment & 5 & Employed \\
-Maternal background & 0/5 & Age, height \newline Nonwhite ethnicity (child) \newline Number of children in HH \\
-Pregnancy & 0 & Firstborn child, number of previous stillbirths \newline Mother smoked in pregnancy \newline Preterm birth, (log) birth weight \\
-\end{tabular}
-\end{table}
-
+\frametitle{Determinants of socioemotional skills}\label{detintro}
+
+\textbf{Q4}: Have the \textbf{determinants} of socioemotional skills changed across cohorts?
+\bi 
+\item Previous work has mostly focused on single cohorts, and not estabilished invariance
+\item \cite{McCulloch2000,Blanden2008,Fitzsimons2017}
+\item Linear regression (Tobit results are very similar)
+\item Merge information from the birth and 5 year surveys \hyperlink{detvars}{\beamerbutton{Variables}}
+\ei
 \end{frame}
@@
 
+\begin{frame}
+\frametitle{Determinants of socioemotional skills}
+
+\textbf{Q4}: Have the \textbf{determinants} of socioemotional skills changed across cohorts?
+
+\bi
+\item Diminishing importance of father's occupation in determining skills
+\item Increased importance of
+  \bi 
+  \item Maternal education and employment
+  \item Smoking in pregnancy
+  \item Birthweight
+  \ei
+\ei
+
+\end{frame}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-write up Gelbach approach...
-
-\end{frame}
+\bi
+\item We decompose the difference between children of mothers with compulsory education vs. post-compulsory
+\item Use the same groups of covariates as above
+\item Approach by \cite{Gelbach2016}
+  \bi
+  \item Based on simple omitted variable bias expression
+  \item Invariant to the ordering of variables
+  \ei
+\ei
+
+\end{frame}
+
+<<GELB_FIG>>=
+# make gelbach graphs
+@
 
@@
 \frametitle{Decomposition (EXT)}\label{gelb_ext_f}
-\framesubtitle{Explaining the gradient by maternal compulsory schooling \hyperlink{gelb_ext}{\beamerbutton{Table}}}
-
+\framesubtitle{Maternal compulsory schooling \hyperlink{gelb_ext}{\beamerbutton{Table}}}
+\begin{figure}
+<<GELB_FIG_EXT, dependson="GELB_FIG", dev='tikz', fig.height=4.5, fig.width=9, out.width='\\linewidth'>>=
+# plot
+gelbplot_ext
+@
+\end{figure}
 \end{frame}
@@
 \frametitle{Decomposition (INT)}\label{gelb_int_f}
-\framesubtitle{Explaining the gradient by maternal compulsory schooling \hyperlink{gelb_int}{\beamerbutton{Table}}}
+\framesubtitle{Maternal compulsory schooling \hyperlink{gelb_int}{\beamerbutton{Table}}}
+\begin{figure}
+<<GELB_FIG_INT, dependson="GELB_FIG", dev='tikz', fig.height=4.5, fig.width=9, out.width='\\linewidth'>>=
+# plot
+gelbplot_int
+@
+\end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{Decomposition}
+\bi
+\item Differences in age 5 skills by mother's education have increased between 1975 and 2005
+\item Observed variables can explain around half of the differentials
+\item Father's occupation explains less in 2005 than 1975, while pregnancy factors explain more
+\ei
 
@@
 
-\section{Mapping skills to later outcomes}
+\section{Mapping skills to adolescent/adult outcomes}
+
+\begin{frame}
+\frametitle{Mapping skills to adolescent/adult outcomes}
+
+\textbf{Q5}: Has the mapping of socioemotional skills into \textbf{adolescent/adult outcomes} changed?
+
+\bi
+\item We examine how sociemotional skills at age 5 map into outcomes later in the life course
+  \bi 
+  \item BCS: Age 16, 30, 38, 42
+  \item MCS: Age 14 (last available survey)
+  \ei
+\item Previous work has mostly focused on single cohorts\cite{Carneiro2007,Conti2010a,Joshi2014}
+\item Focus on behaviours and labour market outcomes
+\pause
+\item Include a \textbf{cognitive factor} at age 5, using available tests
+  \bi 
+  \item BCS: English Picture Vocabulary Test, Human Figure Drawing, Copy Designs
+  \item MCS: British Ability Scales (Naming Vocabulary, Picture Similarities, Pattern Construction)
+  \item Bartlett scoring
+  \ei
+\ei
+\end{frame}
 
 <<LOAD_DATA_OUTC, dependson="SELECTVARS">>=
-# rload outcomes data
+# load outcomes data
 @
@@
 \begin{frame}
-\frametitle{Mapping skills into behaviours}
-\framesubtitle{Adolescence (MCS)}
+\frametitle{Mapping skills}
+\framesubtitle{Adolescent behaviours (MCS)}
+\begin{table}
 \centering
@@
 # collapse to matrix
-mcs_outctab_mat <- do.call(rbind,lapply(mcs_outctab,matrix,ncol=11,byrow=FALSE))
+mcs_outctab_mat <- do.call(rbind,lapply(mcs_outctab[c("smktry14", "bmi14")],matrix,ncol=11,byrow=FALSE))
 table <- mcs_outctab_mat
@@
 # NAs in columns and rows with COG
-table[,c(6,11)] <- NA
-table[seq(4,16,4),] <- NA
+table[,c(6,11)] <- paste0("{\\color{white}", table[,c(6,11)],"}")
+table[seq(4,nrow(table),4),] <- paste0("{\\color{white}", table[seq(4,nrow(table),4),],"}")
 
@@
 }
-\end{frame}
-
-\begin{frame}[noframenumbering]
-\frametitle{Mapping skills into behaviours}
-\framesubtitle{Adolescence (MCS)}
+\\[-12pt] \justify {\scriptsize \emph{Notes}: Robust standard errors in parentheses.
+All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
+\par}
+\end{table}
+\end{frame}
+
+\begin{frame}[noframenumbering]
+\frametitle{Mapping skills}\label{map_mcs}
+\framesubtitle{Adolescent behaviours (MCS) \hyperlink{map_mcs_app}{\beamerbutton{More}}}
+\begin{table}
 \centering
@@
 }
-\end{frame}
-
-\begin{frame}
-\frametitle{Mapping skills into behaviours}
-\framesubtitle{Adolescence (BCS)}
+\\[-12pt] \justify {\scriptsize \emph{Notes}: Robust standard errors in parentheses.
+All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
+\par}
+\end{table}
+\end{frame}
+
+\begin{frame}
+\frametitle{Mapping skills}
+\framesubtitle{Adolescent behaviours (BCS)}
+\begin{table}
 \centering
@@
 # collapse to matrix
-bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab[c("smktry16", "alcoh16", "bmi16")],matrix,ncol=11,byrow=FALSE))
+bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab[c("smktry16", "bmi16")],matrix,ncol=11,byrow=FALSE))
 table <- bcs_outctab_mat
@@
 # NAs in columns and rows with COG
-table[,c(6,11)] <- NA
-table[seq(4,12,4),] <- NA
+table[,c(6,11)] <- paste0("{\\color{white}", table[,c(6,11)],"}")
+table[seq(4,nrow(table),4),] <- paste0("{\\color{white}", table[seq(4,nrow(table),4),],"}")
 
@@
 }
-\end{frame}
-
-\begin{frame}[noframenumbering]
-\frametitle{Mapping skills into behaviours}
-\framesubtitle{Adolescence (BCS)}
+\\[-12pt] \justify {\scriptsize \emph{Notes}: Robust standard errors in parentheses.
+All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
+\par}
+\end{table}
+\end{frame}
+
+\begin{frame}[noframenumbering]
+\frametitle{Mapping skills}\label{map_bcs1}
+\framesubtitle{Adolescent behaviours (BCS) \hyperlink{map_bcs1_app}{\beamerbutton{More}}}
+\begin{table}
 \centering
@@
 }
-\end{frame}
-
-\begin{frame}
-\frametitle{Mapping skills into labour market}
-\framesubtitle{Adulthood (BCS)}
+\\[-12pt] \justify {\scriptsize \emph{Notes}: Robust standard errors in parentheses.
+All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
+\par}
+\end{table}
+\end{frame}
+
+\begin{frame}
+\frametitle{Mapping skills}
+\framesubtitle{Adult labour market (BCS)}
+\begin{table}
 \centering
@@
 # collapse to matrix (last 4 elements of bcs_outctab are adulthood)
-bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab[c(5:8)],matrix,ncol=11,byrow=FALSE))
-table <- bcs_outctab_mat
+table <- do.call(rbind,lapply(bcs_outctab[c("hnvq30", "lhgrpay38")],matrix,ncol=11,byrow=FALSE))
 
 # NAs in columns and rows with COG
-table[,c(6,11)] <- NA
-table[seq(4,16,4),] <- NA
+table[,c(6,11)] <- paste0("{\\color{white}", table[,c(6,11)],"}")
+table[seq(4,nrow(table),4),] <- paste0("{\\color{white}", table[seq(4,nrow(table),4),],"}")
 
@@
 }
-\end{frame}
-
-\begin{frame}[noframenumbering]
-\frametitle{Mapping skills into labour market}
-\framesubtitle{Adulthood (BCS)}
+\\[-12pt] \justify {\scriptsize \emph{Notes}: Robust standard errors in parentheses.
+All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
+\par}
+\end{table}
+\end{frame}
+
+\begin{frame}[noframenumbering]
+\frametitle{Mapping skills}\label{map_bcs2}
+\framesubtitle{Adult labour market (BCS) \hyperlink{map_bcs2_app}{\beamerbutton{More}}}
+\begin{table}
 \centering
@@
 # collapse to matrix (last 4 elements of bcs_outctab are adulthood)
-bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab[c(5:8)],matrix,ncol=11,byrow=FALSE))
-table <- bcs_outctab_mat
+table <- do.call(rbind,lapply(bcs_outctab[c("hnvq30", "lhgrpay38")],matrix,ncol=11,byrow=FALSE))
 
@@
 }
+\\[-12pt] \justify {\scriptsize \emph{Notes}: Robust standard errors in parentheses.
+All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
+\par}
+\end{table}
+\end{frame}
+
+\begin{frame}
+\frametitle{Mapping skills into adolescent/adult outcomes}
+
+\textbf{Q5}: Has the mapping of socioemotional skills into \textbf{adolescent/adult outcomes} changed?
+
+\bi
+\item Worse externalising skills are significant predictors of \textbf{adolescent smoking and BMI}
+  \bi \item (even controlling for cognitive skills) \ei
+\item Internalising skills have opposite sign
+  \bi \item Better INT can imply more sociable, outgoing child \ei
+\item For \textbf{higher educational achievement}
+  \bi 
+  \item Cognitive skills dominate
+  \item Higher EXT and lower INT achieve better
+  \ei
+\item For \textbf{wages}, cognitive skills are predominant
+\ei
+
 \end{frame}
@@
 
-\section{Conclusions and next steps}
-
-\begin{frame}
-\frametitle{Conclusions and next steps}
-
-\bi
-\item We derived a comparable scale of non-cognitive skills, based on maternal reports of behaviour problems
-\item We have shown that inequality in these skills at five years of age has increased between children born in 1970 and 2000
+\section{Conclusion}
+
+\begin{frame}
+\frametitle{Conclusion}
+
+\bi
+\item Using a novel psychometric method, we derived two measures of non-cognitive skills at age five that is comparable across two UK cohorts
+\item We have shown that inequality in these skills at five years of age by parental background has increased between children born in 1970 and 2000
   \bi
   \item Profile has become steeper (particularly for males)
-  \item Scores at upper quantiles have drifted upwards
-  \ei
-\pause
-\vspace{12pt}
-\item Next steps
-  \bi
-  \item Harmonise more cohorts (NCDS) and ages (10)
-  \item Broaden analysis to the \emph{returns} to non-cognitive skills -- are these remunerated differently on the labour market for people born in different years?
-  \ei
+  \item Inequality across maternal education is starker
+  \ei
+\item The determinants of socioemotional skills have changed
+  \bi
+  \item Father's occupation is less important
+  \item Maternal and pregnancy factors are predominant
+  \ei
+\item For both cohorts, socioemotional skills predict adolescent outcomes, for health outcomes over and above cognitive skills
 \ei
@@
 
-\begin{frame}
+\begin{frame}[noframenumbering]
 \frametitle{Harmonisation}\label{meantab_m}
@@
 \begin{frame}[noframenumbering]
+\frametitle{EFA}\label{efalit}
+\framesubtitle{Literature \hyperlink{efa1}{\beamerreturnbutton{Back}}}
+Rutter A scale (BCS) 
+  \bi 
+  \item Traditionally conceptualised as bidimensional \citep{Blanden2007}
+    \bi
+    \item \emph{anti-social} -- or aggressive-hyperactive
+    \item \emph{neurotic} -- or anxious/fearful
+    \ei
+  \item Sometimes three-dimensional \citep{McGee1985,Collishaw2004}
+    \bi
+    \item \emph{hyperactive}
+    \item \emph{conduct} -- or aggressive/ant-isocial
+    \item \emph{neurotic} -- or anxious/fearful
+    \ei
+  \ei
+Strengths and Difficulties Questionnaire (MCS) 
+  \bi
+  \item Five subscales with full 25 items -- peer problems, emotional, conduct, hyperactive, prosocial \citep{Stone2010}
+  \item Evidence that lower-dimensional structures can work \citep{Goodman2010}
+  \ei
+
+\end{frame}
+
+\begin{frame}[noframenumbering]
 \frametitle{EFA}\label{numfac}
-\framesubtitle{Number of factors \hyperlink{efa}{\beamerreturnbutton{Back}}}
+\framesubtitle{Number of factors \hyperlink{efa2}{\beamerreturnbutton{Back}}}
 \setstretch{1.1}
@@
 
-\begin{frame}
+\begin{frame}[noframenumbering]
 \frametitle{EFA}\label{meantab_fcc3}
@@
 \end{spacing}
+\end{frame}
+
+\begin{frame}[noframenumbering]
+\frametitle{MI}\label{midet_app}
+\framesubtitle{IFA model \hyperlink{midet}{\beamerreturnbutton{Back}}}
+\setstretch{1.1}
+Categorical Item Factor Analysis (IFA) model
+\bi
+\item Latent propensity to endorse each item $i$ for child $j$ in group $c$ \citep{Muthen1984}
+\begin{align}\label{eq:propens}
+\begin{split}
+X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{EXT}_{jc} + u_{ijc} \qquad \text{for} \; i=1,\dots, 6 \\
+X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{INT}_{jc} + u_{ijc} \qquad \text{for} \; i=7,\dots, 11
+\end{split}
+\end{align}
+\item Mean/covariance structure
+$$\bm{\mu}_c=\bm{\nu}_c + \bm{\Lambda}_c\bm{\kappa}_c \qquad \text{and} \qquad \Sigma_c = \bm{\Lambda}_c \bm{\Phi}_c \bm{\Lambda}_c^\prime + \bm{\Psi}_c.$$
+with $$(\theta^{EXT}_{jc}, \theta^{INT}_{jc}) = \bm{\Theta}_{jc} \sim N(\bm{\kappa}_c, \bm{\Phi}_c) \qquad \text{and} \qquad \bm{u}_{jc} \sim N(\bm{0}, \bm{\Psi}_c)$$
+\item Discreteness \citep{Piatek2016}
+$$X_{ijc} = s \qquad \text{if} \; \tau_{s,ic} \leq X^*_{ijc} < \tau_{s+1,ic} \qquad \text{for} \; s=0,1,2$$
+with $\tau_{0,ic}=-\infty$ and $\tau_{3,ic}=+\infty$.
+\ei
 \end{frame}
@@
 \begin{frame}[noframenumbering]
+\frametitle{Age}\label{agehist}
+\framesubtitle{Child age at maternal report \hyperlink{mires}{\beamerreturnbutton{Back}}}
+\begin{figure}
+<<AGEHIST, dependson="FA_SCORES_BIND", dev='tikz', fig.height=5, fig.width=8, out.width='.7\\linewidth'>>=
+@
+\end{figure}
+\end{frame}
+
+\begin{frame}[noframenumbering]
 \frametitle{Factor scores}\label{rawhist}
@@
 <<RAWHIST, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
+@
+\end{figure}
+\end{frame}
+
+\begin{frame}[noframenumbering]
+\frametitle{Factor scores}\label{facraw}
+\framesubtitle{Factor scores vs. sum scores \hyperlink{miscores}{\beamerreturnbutton{Back}}}
+\begin{figure}
+<<FACRAW, dependson="FA_SCORES_BIND", dev='tikz', fig.height=6, fig.width=9, out.width='.9\\linewidth'>>=
 @
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%% DETERMINANTS
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}[noframenumbering]
+\frametitle{Determinants of socioemotional skills} \label{detvars}
+\framesubtitle{Variables \hyperlink{detintro}{\beamerreturnbutton{Back}}}
+
+Variable groupings
+
+\begin{table}
+\footnotesize
+\begin{tabular}{L{.25\textwidth}C{.1\textwidth}L{.5\textwidth}}
+Group & Age &  \\
+\midrule
+Maternal education & 5 & Post-compulsory schooling \\
+Maternal employment & 5 & Employed \\
+Father occupation & 5 & White collar (SC I-IIINM) \newline Blue collar (SC IIIM-V-other) \newline No father figure \\
+Maternal background & 0/5 & Age, height \newline Nonwhite ethnicity (child) \newline Number of children in HH \\
+Pregnancy & 0 & Firstborn child, number of previous stillbirths \newline Mother smoked in pregnancy \newline Preterm birth, (log) birth weight \\
+\end{tabular}
+\end{table}
+
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%% GELBACH DECOMPOSITION
@@
 
+\begin{frame}[noframenumbering]
+\frametitle{Mapping skills}\label{map_mcs_app}
+\framesubtitle{Adolescent behaviours (MCS) \hyperlink{map_mcs}{\beamerreturnbutton{Back}}}
+\centering
+\setstretch{1.0}
+\resizebox{\textwidth}{!}{
+<<REGS_OUTC_MCS_COG_APP, results='asis', dependson="REGRESSIONS">>=
+
+# collapse to matrix
+mcs_outctab_mat <- do.call(rbind,lapply(mcs_outctab[c("alctry14", "canntry14", "selfharm14")],matrix,ncol=11,byrow=FALSE))
+table <- mcs_outctab_mat
+
+printoutctab(table)
+
+@
+}
+\end{frame}
+
+\begin{frame}[noframenumbering]
+\frametitle{Mapping skills}\label{map_bcs1_app}
+\framesubtitle{Adolescent behaviours (BCS) \hyperlink{map_bcs1}{\beamerreturnbutton{Back}}}
+\centering
+\setstretch{1.0}
+\resizebox{\textwidth}{!}{
+<<REGS_OUTC_BCS1_COG_APP, results='asis', dependson="REGRESSIONS">>=
+
+# collapse to matrix
+table <- do.call(rbind,lapply(bcs_outctab[c("alcoh16", "canntry16")],matrix,ncol=11,byrow=FALSE))
+
+printoutctab(table)
+
+@
+}
+\end{frame}
+
+\begin{frame}[noframenumbering]
+\frametitle{Mapping skills}\label{map_bcs2_app}
+\framesubtitle{Adult labour market (BCS) \hyperlink{map_bcs2}{\beamerreturnbutton{Back}}}
+\centering
+\setstretch{1.0}
+\resizebox{\textwidth}{!}{
+<<REGS_OUTC_BCS2_COG_APP, results='asis', dependson="REGRESSIONS">>=
+
+# collapse to matrix
+table <- do.call(rbind,lapply(bcs_outctab[c("hscl42", "bmi42")],matrix,ncol=11,byrow=FALSE))
+
+printoutctab(table)
+
+@
+}
+\end{frame}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\begin{frame}[noframenumbering]
+\begin{frame}[allowframebreaks]
 \bibliographystyle{chicago}

### 2018-06-04 codes/R/slides/cohorts_tapes.Rnw
version presented at TAPES

@@
     \item e.g. Rutter, BSAG, SDQ 
-    \item Obtained by aggregating different items
-    \ei
-  \item Can we use these scales to compare different cohorts, by gender?
+    \item Obtained by aggregating different questions (items)
+    \ei
+  \item Can we use information from these scales to compare different populations (cohorts, gender)?
   \item And what if only info from different scales is available?
@@
 We compare the evolution of inequality in socioemotional skills at age 5 across two UK cohorts born 30 years apart
+\pause
 \bi
@@
   \item Sharpest change for boys rather than girls
-  \item Increased importance of pregnancy factors, decreased importance of paternal education, in explaining skill inequality by maternal education
+  \item Increased importance of pregnancy factors, decreased importance of paternal occupation, in explaining skill inequality by maternal education
   \item In both cohorts, socioemotional skills (especially externalising) are significant predictors of social and health outcomes in adolescence
@@
   \item Different design, timing, questions, survey instruments...
-  \item Harmonisation effort underway (CLOSER)
   \ei
@@
   \bi \item E.g. ``Generally obedient" dichotomised to match ``Often disobedient"  \ei
-\item $\Rightarrow$ New 11-item subscale
+\item $\Rightarrow$ \textbf{Novel 11-item scale}
 \ei
@@
 \bi
-\item Factor analysis routinely used to construct measures of socioemotional skills \citep{Conti2014,Attanasio2017c}
-\item Starts with \emph{exploratory} step
-\item Decompose the covariance matrix of items to assess:
-  \bi \item the dimensionality of the underlying factor structure (how many latent factors)
-  \item which items measure which latent factor \ei
+\item Factor analysis routinely used to examine socioemotional skills \citep{Conti2014,Attanasio2017c}
+\item Starts with \emph{exploratory factor analysis} step
+\item Objective is to assess
+  \bi
+  \item the dimensionality of the underlying factor structure (\emph{how many} latent factors)
+  \item \emph{which items} measure which latent factor 
+  \ei
+\vspace{.3cm}
 \item Psychometric literature often finds \textbf{two latent factors} provide good synthesis for Rutter and SDQ \hyperlink{efalit}{\beamerbutton{Det.}}
-\item This can be tested in our data
+\item We can test this in our data
 \ei
@@
 \bi
-\item Dimensionality
-  \bi
-  \item Methods for assessing dimensionality imply 1-3 factors \hyperlink{numfac}{\beamerbutton{Det.}}
+\item Exploratory FA: weighted least squares and \textit{oblimin} factor rotation
+\item Separately by cohort (BCS, MCS) and gender
+\item We find
+  \bi
+  \item Dimensionality methods imply 1-3 factors \hyperlink{numfac}{\beamerbutton{Det.}}
   \item We choose \textbf{two factors} to maximise interpretability
-  \ei
-\item Structure
-  \bi
-  \item Exploratory FA using weighted least squares and \textit{oblimin} rotation
-  \item Factors can be coherently interpreted as \textbf{externalising} ($\sim$ anti-social) and \textbf{internalising} ($\sim$ neurotic)
   \ei
@@
 \begin{tabular}{C{.05\textwidth}L{.3\textwidth}|C{.05\textwidth}L{.3\textwidth}}
-  & \textbf{{\color{OliveGreen}Externalising}}  &    & \textbf{{\color{Orange}Internalising}}  \\
+  & \textbf{{\color{OliveGreen}Externalising SE skills}}  &    & \textbf{{\color{Orange}Internalising SE skills}}  \\
 1 & Restless                & 7  & Worried                \\
@@
     \bnm
-    \item Start from the most general model -- parameters allowed to differ across groups
+    \item Start from most general model -- parameters allowed to differ across groups
     \item Impose restrictions on sets of parameters to be equal across groups
@@
 \bi
-\item Observed categorical items $X_1$-$X_{11}$ measuring latent factor $\theta$
-\item Latent propensity to endorse each item $i$ for child $j$ in group $c$ \citep{Muthen1984}
+\item We observe categorical items $X_1$-$X_{11}$ measuring latent factor $\theta$ (externalising/internalising socioemotional skills)
+\pause
+\item Different groups of individuals ($c$) -- children in cohort-gender cell
+\pause
+\item Define propensity to endorse item $i$ for child $j$ in group $c$ \citep{Muthen1984}
 $$ X^*_{ijc} =\nu_{ci} + \lambda_{ci}\theta_{jc} + u_{ijc} $$
+\pause
 \item Discreteness \citep{Piatek2016}
@@
 with $\tau_{0,ic}=-\infty$ and $\tau_{3,ic}=+\infty$.
-\vspace{.2cm}
-\item We follow the approach in \cite{Wu2016a}
-  \bi
-  \item Define group ($c$) as cohort-gender cell
-  \ei
 \ei
@@
 \framesubtitle{Steps \hyperlink{miapp1}{\beamerbutton{Details}}}
+\setstretch{1.1}
+\cite{Wu2016a}
 $$ X^*_{ijc} =
@@
 \qquad \text{for} \; s=0,1,2$$
-\vspace{-.2cm}
+\vspace{-.4cm}
+\setstretch{1.3}
 \bnm
@@
 \bi
-\item We use RMSE to compare fit between models with different invariance restriction
-\item Restricting only thresholds and loadings provides a good fit \hyperlink{fit}{\beamerbutton{Fit details}}
+\item We use RMSE to compare fit between models with different invariance restriction \hyperlink{fit}{\beamerbutton{Alternative fit measures}}
+\item Restricting only thresholds and loadings provides a good fit
   \bi \item We can compare the scale (variance) of the latent factors \ei
-\item Models with restricted intercepts do \emph{not} pass MI test
+\item Model with restricted intercepts do \emph{not} pass MI test
   \bi \item We cannot compare the location of latent factors in different groups \ei
@@
 \bi
-\item We can use the model with non-invariant intercepts to score the latent factors (externalising and internalising dimensions)
-\item Estimated parameters: \hyperlink{measpars}{\beamerbutton{Measurement}}\hyperlink{lvpars}{\beamerbutton{Latent variable}}
-\item Use Empirical Bayes approach to predict individual scores \citep{Wu2013}
-\item All scores are coded so that \emph{\textbf{a higher value means better socioemotional skills}}
-\item \textbf{CAUTION}: location differences by cohort/gender are not interpretable!
+\item We can use the model with non-invariant intercepts to \emph{score} the latent factors
+  \bi
+  \item Assign a predicted value to each child's externalising and internalising skills
+  \item Empirical Bayes approach method \citep{Wu2013}
+  \ei
+\item Estimated parameters \hyperlink{measpars}{\beamerbutton{Measurement}}\hyperlink{lvpars}{\beamerbutton{Latent variable}}
+\pause
+\item \textbf{CAUTION}:
+  \bi
+  \item Scores coded so that \emph{\textbf{a higher value means better socioemotional skills}}
+  \item \textbf{Location differences} by cohort/gender are not interpretable! 
+  \ei
 \ei
@@
 \bi
-\item Is this an improvement on using simple sum scores?
+\item Is this an improvement on using simple \textbf{sum scores}?
   \bi
@@
   \item Scores obtained from the model exploit the covariance structure of the whole 11-item scale
-  \item Every 11-dimensional pattern of response in the data can be assigned a different score
+  \item Every 11-dimensional pattern of response in the data is assigned a different score
   \item Substantial increase in variability \hyperlink{facraw}{\beamerbutton{Graph}}
@@
 \frametitle{Inequality}\label{ineq_ps}
-\framesubtitle{Mother compulsory education \hyperlink{ineq_ys}{\beamerbutton{Age left education}}}
+\framesubtitle{Mother compulsory education 
+\hyperlink{ineq_ys}{\beamerbutton{Age left education}}
+\hyperlink{ineq_mecs}{\beamerbutton{Education + employment}}
+}
 \vspace{-.2cm}
@@
 fi <- fi_ps # select result
-gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
-                        fi$ineqlist[[3]],fi$ineqlist[[4]],
-                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
-                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
-@
-\end{figure}
-\end{frame}
-
-\begin{frame}
-\frametitle{Inequality}\label{ineq_mecs}
-\framesubtitle{Maternal education and employment}
-\vspace{-.2cm}
-\begin{figure}
-<<FACINEQ_MECS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=11, out.width='\\linewidth'>>=
-# plot
-fi <- fi_mecs # select result
 gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
@@
 \bi
-\item Unconditionally, the distribution is more spread out (except for girls' internalising)
+\item Unconditionally, the distribution is more spread out (except for girls' internalising skills)
 \item Inequality across socioeconomic indicators has increased, especially
@@
 \item Previous work has mostly focused on single cohorts, and not estabilished invariance
-\item \cite{McCulloch2000,Blanden2008,Fitzsimons2017}
+  \bi \item \cite{McCulloch2000,Blanden2008,Fitzsimons2017} \ei
+\item We regress externalising and internalising socioemotional skills on contemporaneous and past determinants
+  \bi
+  \item Merge information from the surveys at birth and 5 years
   \item Linear regression (Tobit results are very similar)
-\item Merge information from the birth and 5 year surveys \hyperlink{detvars}{\beamerbutton{Variables}}
-\ei
+  \ei
+\ei
+\end{frame}
+
+\begin{frame}
+\frametitle{Determinants of socioemotional skills} \label{detvars}
+\framesubtitle{Variable groups}
+
+\begin{table}
+\footnotesize
+\begin{tabular}{L{.25\textwidth}C{.1\textwidth}L{.5\textwidth}}
+Group & Age &  \\
+\midrule
+Maternal education & 5 & Post-compulsory schooling \\ \\[-1.5em]
+Maternal employment & 5 & Employed \\ \\[-1.5em]
+Father occupation & 5 & White collar (SC I-IIINM) \newline Blue collar (SC IIIM-V-other) \newline No father figure \\ \\[-1.5em]
+Maternal background & 0/5 & Age, height \newline Nonwhite ethnicity (child) \newline Unmarried at birth \newline Number of children in HH \\ \\[-1.5em]
+Pregnancy & 0 & Firstborn child, number of previous stillbirths \newline Mother smoked in pregnancy \newline Preterm birth, (log) birth weight \\ \\[-1.5em]
+\end{tabular}
+\end{table}
+
 \end{frame}
@@
 \begin{frame}
-\frametitle{Decomposition}
-\framesubtitle{Gelbach (2016)}
-
-\bi
-\item We decompose the difference between children of mothers with compulsory education vs. post-compulsory
-\item Use the same groups of covariates as above
-\item Approach by \cite{Gelbach2016}
+\frametitle{Decomposition}\label{gelb}
+
+\bi
+\item Stark differences in skills between children of high vs low educated mothers \hyperlink{ineq_ps}{\beamerbutton{Graph}}
+\item We decompose these difference into components attributable to different groups of factors (using the same groups as above)
+\item Approach by \cite{Gelbach2016} \hyperlink{gelb_det}{\beamerbutton{Details}}
   \bi
@@
   \item Invariant to the ordering of variables
+  \item No causal interpretation
   \ei
@@
 \begin{frame}
-\frametitle{Decomposition (EXT)}\label{gelb_ext_f}
-\framesubtitle{Maternal compulsory schooling \hyperlink{gelb_ext}{\beamerbutton{Table}}}
+\frametitle{Decomposition}\label{gelb_ext_f}
+\framesubtitle{Externalising \hyperlink{gelb_ext}{\beamerbutton{Table}}}
 \begin{figure}
@@
 \begin{frame}
-\frametitle{Decomposition (INT)}\label{gelb_int_f}
-\framesubtitle{Maternal compulsory schooling \hyperlink{gelb_int}{\beamerbutton{Table}}}
+\frametitle{Decomposition}\label{gelb_int_f}
+\framesubtitle{Internalising \hyperlink{gelb_int}{\beamerbutton{Table}}}
 \begin{figure}
@@
 \bi
-\item We examine how sociemotional skills at age 5 map into outcomes later in the life course
+\item Previous work has mostly looked at single cohorts \citep{Carneiro2007,Conti2010a,Joshi2014}
+\item We examine how sociemotional skills at age 5 map into later outcomes
   \bi 
@@
   \ei
-\item Previous work has mostly focused on single cohorts\cite{Carneiro2007,Conti2010a,Joshi2014}
-\item Focus on behaviours and labour market outcomes
+\item Regress behaviours and labour market outcomes on socioemotional skills and same groups of early-life covariates as above
 \pause
@@
   \item BCS: English Picture Vocabulary Test, Human Figure Drawing, Copy Designs
-  \item MCS: British Ability Scales (Naming Vocabulary, Picture Similarities, Pattern Construction)
+  \item MCS: British Ability Scales (Naming Vocab., Picture Simil., Pattern Constr.)
   \item Bartlett scoring
@@
 \bi
-\item Using a novel psychometric method, we derived two measures of non-cognitive skills at age five that is comparable across two UK cohorts
+\item Using a novel psychometric method, we derived two measures of socioemotional skills at age five that are comparable across two UK cohorts
 \item We have shown that inequality in these skills at five years of age by parental background has increased between children born in 1970 and 2000
@@
   \ei
-\item For both cohorts, socioemotional skills predict adolescent outcomes, for health outcomes over and above cognitive skills
+\item In both cohorts, socioemotional skills predict adolescent outcomes -- for health outcomes, this is true over and above cognitive skills
 \ei
@@
 \bi
-\item Latent propensity to endorse each item $i$ for child $j$ in group $c$ \citep{Muthen1984}
+\item Latent propensity to endorse item $i$ for child $j$ in group $c$ \citep{Muthen1984}
 \begin{align}\label{eq:propens}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% DETERMINANTS
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{frame}[noframenumbering]
-\frametitle{Determinants of socioemotional skills} \label{detvars}
-\framesubtitle{Variables \hyperlink{detintro}{\beamerreturnbutton{Back}}}
-
-Variable groupings
-
-\begin{table}
-\footnotesize
-\begin{tabular}{L{.25\textwidth}C{.1\textwidth}L{.5\textwidth}}
-Group & Age &  \\
-\midrule
-Maternal education & 5 & Post-compulsory schooling \\
-Maternal employment & 5 & Employed \\
-Father occupation & 5 & White collar (SC I-IIINM) \newline Blue collar (SC IIIM-V-other) \newline No father figure \\
-Maternal background & 0/5 & Age, height \newline Nonwhite ethnicity (child) \newline Number of children in HH \\
-Pregnancy & 0 & Firstborn child, number of previous stillbirths \newline Mother smoked in pregnancy \newline Preterm birth, (log) birth weight \\
-\end{tabular}
-\end{table}
-
+\begin{frame}[noframenumbering]
+\frametitle{Inequality}\label{ineq_mecs}
+\framesubtitle{Maternal education and employment \hyperlink{ineq_ps}{\beamerreturnbutton{Back}}}
+\vspace{-.2cm}
+\begin{figure}
+<<FACINEQ_MECS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=11, out.width='\\linewidth'>>=
+# plot
+fi <- fi_mecs # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\end{figure}
 \end{frame}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{frame}[noframenumbering]
+\frametitle{Gelbach (2016) \hyperlink{gelb}{\beamerreturnbutton{Back}}}\label{gelb_det}
+\setstretch{1.1}
+
+\textbf{Base} model -- conditional on covariates $\bm{X}$: \hfill $Y = \alpha + \beta_1 T + \veps$
+
+\textbf{Full} model: \hfill $Y = \alpha + \beta_1 T + \bm{M} \bm{\beta}_2 + \veps $
+\vspace{.2cm}
+
+OLS estimate of $\beta_1$ in the base model: $$ \beta_1^{\text{base}} = \beta_1 + \bm{\Gamma} \bm{\beta}_2 = \beta_1 + \delta $$
+
+where $\delta$ is \textbf{omitted variable bias} due to omission of mediators $\bm{M}$
+
+and $\bm{\Gamma}$ are coefficients from a linear projection of $\bm{M}$ on $T$: \hfill $\bm{M} = T\bm{\Gamma} + \bm{u} $
+
+with \hfill
+$M^1 = \gamma_0^1 + \gamma_1^1 T + u^1  \qquad \dots \qquad M^k = \gamma_0^{k_m} + \gamma_1^{k_m} T + u^{k_m} $
+
+Decomposition of the bias term across mediators
+$$\delta = \beta_1^{\text{base}} - \beta_1 = \gamma_1^1 \beta_2^1 + \dots +  \gamma_1^{k_m} \beta_2^{k_m} = \sum_{j=1}^{k_m} \gamma_1^j \beta_2^j $$
+where $\delta^j =  \gamma_1^j \beta_2^j$ is the component due to each mediator.
+
+\end{frame}
 
@@
 
-\begin{frame}[allowframebreaks]
+\begin{frame}[allowframebreaks,noframenumbering]
 \bibliographystyle{chicago}

### 2018-06-04 codes/R/9_gelbfig.R

@@ -53,7 +53,7 @@ addopts_gelb <- function(x) {
     geom_bar(stat = "identity", alpha=.7, width = 0.5) + 
     geom_vline(xintercept = 2.5) +
     coord_flip(ylim = c(0.0,0.3)) +
-    scale_y_continuous(name = "Difference") + xlab("") +
+    scale_y_continuous(name = "Difference (high vs low educated mothers)") + xlab("") +
     scale_fill_manual(values = gcols, name = "") +
     geom_text(aes(label=ifelse(value >= .02, paste0(sprintf("%.1f", perc),"\\%"),"")), position=position_stack(vjust=0.5)) +
     guides(fill=guide_legend(ncol=1,byrow=TRUE)) +

### 2018-07-02 codes/R/3_inv_fit.R
additional AFIs

@@ -1,7 +1,7 @@
 ## ---- FA_FIT
 
 # assemble AFIs indices for table
-indsel <- c("npar", "chisq", "rmsea", "mfi", "cfi.scaled")
+indsel <- c("npar", "chisq", "rmsea", "srmr", "mfi", "cfi.scaled")
 mnames <- c("Configural", "Threshold + Loading Invariance", "Threshold, Loading, + Intercept Invariance")
 
 afitab <- list()
@@ -14,15 +14,26 @@ for (i in 1:length(fa.c)) {
   afitab[[i]]$npar  <- as.character(afitab[[i]]$npar)
   afitab[[i]]$chisq <- as.character(round(afitab[[i]]$chisq,1))
   
+  # anova to get pval of chisquare
+  pv.tl <- anova(fa.c[[i]], fa.tl[[i]])$`Pr(>Chisq)`[2]
+  pv.tli <- anova(fa.c[[i]], fa.tli[[i]])$`Pr(>Chisq)`[2]
+  
   # add deltas
+  drmsea <- NA
+  dsrmr <- NA
   dmfi <- NA
   dcfi <- NA
   dgam <- NA
   for (r in 2:nrow(afitab[[i]])) {
+    drmsea <- c(drmsea, afitab[[i]][r,"rmsea"] -  afitab[[i]][1,"rmsea"])
+    dsrmr <- c(dsrmr, afitab[[i]][r,"srmr"] -  afitab[[i]][1,"srmr"])
     dmfi <- c(dmfi, afitab[[i]][r,"mfi"] -  afitab[[i]][1,"mfi"])
     dcfi <- c(dcfi, afitab[[i]][r,"cfi.scaled"] -  afitab[[i]][1,"cfi.scaled"])
     dgam <- c(dgam, afitab[[i]][r,"gammaHat"] -  afitab[[i]][1,"gammaHat"])
   }
+  afitab[[i]]$pchisq <- as.matrix(c(NA,pv.tl,pv.tli))
+  afitab[[i]]$drmsea <- as.matrix(drmsea)
+  afitab[[i]]$dsrmr <- as.matrix(dsrmr)
   afitab[[i]]$dmfi <- as.matrix(dmfi)
   afitab[[i]]$dcfi <- as.matrix(dcfi)
   afitab[[i]]$dgam <- as.matrix(dgam)

### 2018-07-02 codes/R/paper/paper_jpube.Rnw
new paper version

<<MAIN,cache=FALSE,echo=FALSE,results='hide',message=FALSE,warning=FALSE>>=
rm(list=ls())         # Clear all objects from memory 
  opts_chunk$set(concordance=FALSE,
                 echo=FALSE,
                 tidy=FALSE,
                 autodep=TRUE,
                 cache=T, # REPLACE LATER
                 message=FALSE, 
                 warning=FALSE,
                 size='scriptsize',
                 results='hide',
                 comment=NA, 
                 fig.width=4,
                 fig.height=4, 
                 out.width='\\textwidth',
                 fig.show='hold',
                 tikzDefaultEngine='pdftex',
                 tikzDocumentDeclaration = "\\documentclass[10pt]{article}",
                 tikzMetricPackages = c("\\usepackage[utf8]{inputenc}","\\usepackage[T1]{fontenc}", "\\usetikzlibrary{calc}", "\\usepackage{amssymb}")
                 )

@

<<READ, cache=FALSE >>=
library(here)
read_chunk(here("codes/R/1_load.R"))
read_chunk(here("codes/R/2_efa.R"))
read_chunk(here("codes/R/3_inv_main.R"))
read_chunk(here("codes/R/3_inv_aux.R"))
read_chunk(here("codes/R/3_inv_fit.R"))
read_chunk(here("codes/R/4_scores.R"))
read_chunk(here("codes/R/4_scores_aux.R"))
read_chunk(here("codes/R/5_graphs.R"))
read_chunk(here("codes/R/6_selectvars.R"))
read_chunk(here("codes/R/8_regs_dep.R"))
read_chunk(here("codes/R/8_regs_outc.R"))
read_chunk(here("codes/R/9_gelbfig.R"))
read_chunk(here("codes/R/getpars.R"))

@

<<PREAMBLE >>=
@

<<LOAD_DATA>>=
@

<<MEANTABLE>>=
# make table with means
@

<<MEANTABLE_PRINT>>=
# function to print mean table
printmeantab <- function(tab) {
  print(xtable(tab,
               align="L{0cm}C{.04\\textwidth}C{.05\\textwidth}C{.05\\textwidth}L{.4\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}L{.4\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
               digits = 1
  ), 
  size="\\small",
  include.rownames=F,
  include.colnames=F,
  booktabs = T,
  add.to.row = list(pos = c(list(0)),
                    command = c(
                      # header
                      paste(" & & &
                            \\multicolumn{4}{c}{\\textbf{BCS}} &
                            \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){4-7} \\cmidrule(lr){8-11} 
                            Item & Factor & Num. Categ. &
                            Wording & Cert. Appl. (\\%) & Smtm. Appl. (\\%) & Appl. (\\%) &
                            Wording & Cert. True (\\%) & Smwt. True (\\%) & True (\\%) \\\\ \\midrule
                            "
                            , sep="")
                    )),
  sanitize.text.function = function(x){x},
  sanitize.rownames.function = function(x){""}, # to not print row numbers
  tabular.environment = "tabular", width = ".8\\textwidth",
  floating = FALSE
  )
}

@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

<<FA_INV_MAIN>>=
# main models only
@

<<FA_FIT>>=
# assemble fit indices into matrices
@

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[pdftex,10pt,twoside,a4paper]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amstext}% for resizing the symbol in math
\usepackage{amsthm}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[babel]{csquotes}
\usepackage{booktabs}
\usepackage{tocloft}
\usepackage{dsfont}
\usepackage{xcolor}
\usepackage{accents}
\usepackage{rotating}
\usepackage{pdflscape}  % for landscape page
\usepackage{lmodern}  %font
\usepackage{natbib}
\usepackage{array}
\usepackage{graphicx}
\usepackage{adjustbox}
\usepackage{tikz}
\usepackage{bm}
\usepackage{array}
\usepackage{multirow}
\usepackage[multiple]{footmisc}   % for multiple footnotes
\usepackage{textgreek}            % for greek letters in text
\usepackage{ragged2e}             % for table notes justification
\usepackage{setspace} % for spacing
\onehalfspacing

\setlength{\jot}{11pt} % increase vertical spacing in equations
\usepackage{txfonts}
\usepackage[T1]{fontenc}

% define new columns for tables
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\renewcommand{\arraystretch}{1.2}%

% appendices/numeration/references
\usepackage[nottoc,numbib]{tocbibind} 		% bibliography in TOC
\usepackage[toc,page]{appendix} %appendix
\newcommand{\tocless}[2]{\bgroup\let\addcontentsline=\nocontentsline#1{#2}\egroup} % for empty section
\numberwithin{equation}{section} %for section eq numeration
\usepackage{hyperref}                            % references
\AtBeginDocument{%
\renewcommand\sectionautorefname{Section}        % custom autoref names
\renewcommand\subsectionautorefname{Section}
}
\newcommand{\aref}[1]{\hyperref[#1]{Appendix~\ref{#1}}} % for appendix sections
\usepackage{cleveref}

%tikz
\usepackage{tikz}
\usetikzlibrary{shapes,intersections}
\usetikzlibrary{positioning,arrows, arrows.meta}

%lists
\usepackage[inline]{enumitem}
\setlist{noitemsep}
\setlist[enumerate]{labelsep=*, leftmargin=*}
\setlist[itemize]{labelsep=*, leftmargin=*}
\usepackage[textfont={small},labelfont={bf,small},justification=centering]{caption}     % to allow multiple lines in captions, and customize the look

% page setup (from Racine)
\setlength{\topmargin}{0cm}
\setlength{\textheight}{24cm}
\setlength{\oddsidemargin}{.0cm}
\setlength{\evensidemargin}{.0cm}
\setlength{\textwidth}{16.5cm}
\setlength{\footskip}{1.5cm}
\setlength{\footnotesep}{0.4cm}
\linespread{1.3}

% additional symbols
\newcommand{\me}{\mathrm{e}}
\newcommand{\veps}{\varepsilon}
\newcommand{\E}{\mathbb{E}}
\newcommand{\1}{\mathds{1}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\Corr}{\mathrm{Corr}}
\newcommand{\vect}{\mathrm{vec}}
\newcommand{\vecht}{\mathrm{vech}}
\newcommand{\ubar}[1]{\underaccent{\bar}{#1}}
\def\chk{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}

% additional commands
\newcommand{\bi}{\begin{itemize}}
\newcommand{\ei}{\end{itemize}}
\newcommand{\bnm}{\begin{enumerate}}
\newcommand{\enm}{\end{enumerate}}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}

\usepackage{authblk}
\title{Inequality in non-cognitive skills: a cross-cohort comparison}
\author[1,2]{Orazio Attanasio}
\author[1,2]{Richard Blundell}
\author[1,2]{Gabriella Conti}
\author[1,2]{Giacomo Mason}
\affil[1]{\emph{Institute for Fiscal Studies}}
\affil[1]{\emph{UCL Department of Economics}}
%\institute{UCL and IFS}
\date{\today}

\begin{document}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
\section{Measurement Invariance}

Meaningful comparison of latent constructs across different groups of individuals and/or measurement occasions (in our case, cohorts) requires that the available measures have the same relationship with the latent factors. This property is denominated measurement invariance \citep{Vandenberg2000,Putnick2016}.

Measurement invariance (MI) is a \emph{formally testable} property. Two main frameworks have been used for assessing MI: item-response theory (IRT), or structural equation modeling (SEM). This paper focuses on the latter, as it is the most commonly employed.\footnote{For an introduction to MI in an IRT framework, see for example \cite{Tay2015}.} In general, MI analyses in a SEM framework are carried out by specifying and estimating factor models of different invariance levels, and comparing their relative fit. It starts with the estimation of a minimally restrictive model, where all measurement parameters are allowed to vary across groups -- net of the necessary normalisations -- and proceeds with progressively restricting sets of parameters to be equal across groups. If the fit of more restrictive models is comparable to the configural model, different levels of invariance can be established.

In accordance with the SEM approach, we set up a (confirmatory) item factor analysis (IFA) model to examine measurement invariance. In our application, we observe 11 categorical items ($X_1$-$X_{11}$). Our exploratory analysis above shows that the correlation structure of these measures can be captured by a two-dimensional latent construct (factor) $\bm{\theta}$ -- externalising and internalising socioemotional skills. We make the commonly adopted assumption of a dedicated (or congeneric) factor structure, where each measure is assumed to load on only one latent dimension \citep{Attanasio2017c,Heckman2013}.

To account for the ordered nature of the measures, we specify a threshold model \citep{Muthen1984}. Individual $j=1\dots N_c$ in group $c=\{A,B\}$ is assumed to have a latent continuous propensity $X^*_{ijc}$ for each item $i=1,\dots,I$. Thus, $X^*_{ijc}$ is the child's propensity to exhibit each of the behaviours described in \autoref{tab:comm}, as reported by the mother. This propensity is modelled as a function of item- and group-specific intercepts $\nu_{ic}$ and loadings $\bm{\lambda}_{ic}$, and the individual's level on the latent construct vector $\bm{\theta}_{jc}$, plus an independent error component $u_{ijc}$. In our application, the latent factor is bidimensional, and corresponds to the externalising ($\theta_{jc}^{EXT}$) and internalising ($\theta_{jc}^{INT}$) skills.

The propensity for each item can be written as follows:\footnote{The dedicated factor structure is ensured by the sparsity of the loading matrix, i.e.: $$\bm{\Lambda}_{c} \coloneqq \begin{bmatrix} \lambda_{1c}, \dots, \lambda_{6c} & \bm{0} \\ \bm{0} & \lambda_{7c}, \dots, \lambda_{11c} \end{bmatrix}.$$}
\be\label{eq:propens}
X^*_{ijc} =\nu_{ic} + \bm{\lambda}_{ic}\bm{\theta}_{jc} + u_{ijc} \qquad \text{for} \quad i = 1,\dots,11 \\
\ee

or more compactly:
$$\bm{X}^*_{jc} =\bm{\nu}_c + \bm{\Lambda}_{c}\bm{\theta}_{jc} + \bm{u}_{jc}$$

Additional standard distributional assumptions are placed on the latent constructs and error terms:
$$\bm{\theta}_{jc} \sim N(\bm{\kappa}_c, \bm{\Phi}_c) \qquad \text{and} \qquad \bm{u}_{jc} \sim N(\bm{0}, \bm{\Psi}_c)$$

The model implies the following expression for the mean and covariance structure of the latent propensities:
$$\bm{\mu}_c=\bm{\nu}_c + \bm{\Lambda}_c\bm{\kappa}_c \qquad \text{and} \qquad \Sigma_c = \bm{\Lambda}_c \bm{\Phi}_c \bm{\Lambda}_c^\prime + \bm{\Psi}_c.$$

Finally, discreteness in the observed measures $X_{cji}$ is incorporated by introducing item- and group-specific threshold parameters $\tau_{ic}$. The observed measures as a function of the propensities $X^*$ can be then written as follows:
$$X_{ijc} = s \qquad \text{if} \; \tau_{s,ic} \leq X^*_{ijc} < \tau_{s+1,ic} \qquad \text{for} \; s=0,1,2$$
with $\tau_{0,ic}=-\infty$ and $\tau_{3,ic}=+\infty$.

% FOOTNOTE\footnote{As a consequence of the normality assumption on the error terms $\veps$, the probability that individual $j$ in group $c$ exhibits item $i$ with ``intensity" $s$ is:
% \begin{align*}
% \Pr(X_{ijc}=0) &= 1- \Phi(-\tau_{1,ic} + \nu_{ic} + \lambda_{ic}\bm{\theta}_{jc}) \\
% \Pr(X_{ijc}=1) &= \Phi(-\tau_{1,ic} + \nu_{ic} + \lambda_{ic}\bm{\theta}_{jc})
% \end{align*}
% for binary items and
% \begin{align*}
% \Pr(X_{ijc}=0) &= 1- \Phi(-\tau_{1,ic} + \nu_{ic} + \lambda_{ic}\bm{\theta}_{jc}) \\
% \Pr(X_{ijc}=1) &= \Phi(-\tau_{1,ic} + \nu_{ic} + \lambda_{ic}\bm{\theta}_{jc}) - \Phi(-\tau_{2,ic} + \nu_{ic} + \lambda_{ic}\bm{\theta}_{jc})\\
% \Pr(X_{ijc}=2) &= \Phi(-\tau_{2,ic} +  \nu_{ic} + \lambda_{ic}\bm{\theta}_{jc})
% \end{align*}
% for ordinal items. This IFA model can also be written as a 2-parameter Logit IRT model with item difficulty $b$ and item discrimination $a$. In the simplest case for a binary item $1$:
% $$\text{Probit}(X_{1jc}=1) = a_{1c}(\bm{\theta}_{jc} - b_{1c})$$
% such that $\tau_{1c}=a_{1c}b_{1c}$ and $a_{1c}=\lambda_{1c}$.}

% Binary items take values:
% \begin{equation}
% X_{cji}=
% \begin{cases}
% \qquad 0 \qquad \text{if} \; X^*_{cji}<\tau_{1,ci} \\
% \qquad 1 \qquad \text{if} \; X^*_{cji}\geq\tau_{1,ci}
% \end{cases}
% \end{equation}
% 
% An ordinal item takes values:
% 
% \begin{equation}
% X_{cji}=
% \begin{cases}
% \qquad 0 \qquad \text{if} \; X^*_{cji}<\tau_{1,ci} \\
% \qquad 1 \qquad \text{if} \; \tau_{1,ci} \leq X^*_{cji}<\tau_{2,ci} \\
% \qquad 2 \qquad \text{if} \; X^*_{cji}\geq\tau_{2,ci}
% \end{cases}
% \end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The model specification above cannot identify the parameters unless additional normalisations are imposed. A configural model is defined as a minimal set of restrictions that ensure identification of parameters across groups. There are infinite, equivalent ways in which the configural model can be parameterised -- the well-known issue of factor indeterminacy. Widely used parameterisations for the configural model are:
\bi[label=$\diamond$]
\item \underline{Delta parameterisation} {[WE{\textDelta}]} \citep{Wu2016a}

For all groups: $$\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}, \quad \bm{\nu}=\bm{0}, \quad \text{and} \; \text{diag}(\bm{\Sigma})=\bm{I}.$$

\item \underline{Theta parameterisation} {[WE{\textTheta}]} \citep{Wu2016a}

For all groups: $$\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}, \quad \bm{\nu}=\bm{0}, \quad \text{and} \; \text{diag}(\bm{\Psi})=\bm{I}.$$

\item \underline{Anchored parameterisation} {[MT]} \citep{Millsap2004}
  \bi
  \item For all groups, normalise a reference loading to 1 for each factor
  \item Set invariant across groups one threshold per item (e.g. $\tau_{0,Ai} = \tau_{0,Bi}$), and an additional threshold in the reference items above
  \item In the first group: $\bm{\kappa_A}=\bm{0}$, $\text{diag}(\bm{\Sigma}_A)=\bm{I}$
  \item Set all intercepts $\bm{\nu}$ to zero
  \ei
\ei
The first two parameterisations (WE{\textDelta} and WE{\textTheta}) normalise the mean and variance of factors to the same constants in both groups, and they leave all loadings and intercepts to be freely estimated; they only differ in whether the additional required normalisation is imposed on the variances of the error terms ($\bm{\Psi}$) or on the diagonal of the covariance matrix of the measures ($\bm{\Sigma}$). The MT parameterisation instead proceeds by identifying parameters in one group first, and then imposing cross-group equality constraints to identify parameters in other groups \citep{Wu2016a}.

Once the configural model has been estimated, measurement invariance analysis proceeds by estimating nested models that place increasing restrictions on the item parameters, by constraining them to be equal across groups $c$. In the case where available measures are continuous, MI analysis is straightforward \citep{vandeSchoot2012}: one set of parameters is restricted at a time, and the fit of the resulting models are compared to the configural model. The hierarchy of the nested models usually proceeds by testing loadings first, and then intercepts (to establish \emph{metric} and \emph{scalar} invariance -- see \citealp{Vandenberg2000}).

Invariance of systems with categorical measures is less well understood. In particular, the lack of explicit location and scale in the measures introduces an additional set of parameters (thresholds $\tau$) and makes identification reliant on more stringent normalisations. The first comprehensive approach for categorical measures is proposed by \cite{Millsap2004}. New identification results in \cite{Wu2016a} indicate that, in the categorical case, invariance properties cannot be examined by simply restricting one set of parameters at a time. This is because the identification conditions used in the configural baseline model, while being minimally restrictive on their own, become binding once certain additional restrictions are imposed. In light of this, they propose models that identify structures of different invariance levels. They find that some restrictions cannot be tested alone against the configural model, because the models they generate are statistically equivalent. This is true of loading invariance, and also of threshold invariance in the case when the number of categories of each ordinal item is 3 or less. Furthermore, they suggest that comparison of both latent means and variances requires invariance in loadings, thresholds, and intercepts.

{\color{red} ADD SOME MORE COMMENTS ON IDENTIFICATION HERE, to make everything more palatable to economics audience.}

We estimate the sequence of models detailed in \autoref{tab:invparam} by Weighted Least Squares.\footnote{Parameters are estimated by mean- and variance-adjusted weighted least squares (WLSMV) (see \citealp{Muthen2017a}); estimation starts from the items' polychoric correlation matrix, uses diagonally weighted least squares (DWLS), and exploits the full weight matrix to compute robust standard errors and test statistics. Robust WLS has proved in simulation studies to be moderately robust to small violations of the normality assumption in the latent underlying measures \citep{Flora2004}, and generally outperforms maximum likelihood in large samples \citep{Beauducel2006,Li2016}. All estimates are computed using the \texttt{lavaan} package in \textbf{\textsf{R}} \citep{Rosseel2012}.}
For the purposes of the analysis, we define groups $c$ as cohort-gender cells, with the reference group being males in the BCS (1970) cohort. We then compare the fit of each model against the configural model (i.e. the least restrictive). According to recommended practice, we diplay a range of fit indices. Comparison of $\chi^2$ values across models is the most common strategy. However, tests based on $\Delta \chi^2$ are known to display increasing Type I error rates with sample size and model complexity \citep{Sass2014a}. The use of approximate fit indices (AFIs) has been proposed alongside $\chi^2$ to overcome these limitation. These indices do not have a known sampling distribution, thus making it necessary to rely on rules of thumb to assess what level of \textDelta AFI indicates invariance. Nevertheless, AFIs provide valuable corrections for model parsimony.\footnote{Commonly adopted AFIs in studies of measurement invariance are the root mean squared error of approximation (RMSEA) and the Tucker-Lewis index (TLI). Simulation evidence by \cite{Cheung2002} shows that these indices can show correlation between overall and relative fit, and suggest relying on additional indices, such as the comparative fit index (CFI, \citealp{Bentler1990}), McDonald non-centrality index (MFI, \citealp{McDonald1989}), and Gamma-hat index \citep{Steiger1989} for the case of ordered measures. Commonly accepted thresholds for rejection are $\Delta CFI < - 0.01$, $\Delta MFI < - 0.02$, and $\Delta Gamma hat < - 0.001$. \cite{Meade2008}, using the results from a simulation study, suggests stricter thresholds that should apply in a variety of conditions. For CFI, a single cutoff value of $.002$ is proposed, while cutoffs for MFI depend on the problem's characteristics; in our case (2 factors, 11 items), they suggest $.0066$. \cite{Sass2014a} however cast some doubts of the generalisability of these cutoffs to WLSMV estimators.}

{\color{red} Motivation for choosing RMSEA (does not seem supported in literature, see above).}

The fit of each estimated model using our full sample is compared in Panel A of \autoref{tab:fit}.\footnote{We do not present fit results for the threshold invariance model, as it is statistically equivalent to the configural model and thus its fit is mathematically the same.} The overall fit of the configural model is satisfactory according to widely accepted rules of thumb, with RMSEA close to $0.05$ and CFI close to $.95$ \citep{Hu1999}. For all invariance levels, a chi-squared difference would point to a lack of measurement invariance. However, as detailed above, this test is expected to lead to substantial Type I error with a relatively large sample size such as ours. For the model with restricted thresholds and loadings, most of the \textDelta AFIs indicate that this level of invariance is justified in our data. However, further restricting intercepts results in a model where invariance is rejected across the board.

\begin{table}[]
\centering\setstretch{1.2} \footnotesize
\caption{Parameterisations for measurement invariance}\label{tab:invparam}
\begin{tabular}{L{.18\textwidth}L{.45\textwidth}L{.3\textwidth}}
\toprule
\textbf{Invariance level}     & \textbf{Description} & \textbf{Restrictions} \\ \midrule \\
Configural (WE{\textTheta})   
    & \bi[label=$\cdot$] \item Minimally restrictive model for identification \ei
    & For all groups: $\begin{array}{|l} \text{diag}(\bm{\Phi}) = \bm{I} \\
                      \bm{\kappa}=\bm{0} \\
                      \bm{\nu}=\bm{0} \\
                      \text{diag}(\bm{\Psi})=\bm{I} \end{array}$
                      \\ \\
Threshold invariance
    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ to be equal across groups
                  \item Statistically equivalent to configural (when measures have 3 categories or less)\ei
    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
                      \newline For all groups: $\begin{array}{|l} \text{diag}(\bm{\Phi}) = \bm{I} \\ 
                                                \bm{\kappa}=\bm{0} \end{array}$
                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\nu}_A=\bm{0} \\
                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \end{array}$
                   \\ \\
Threshold and Loading invariance
    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ and loadings $\lambda$ to be equal across groups
                  \item Allows comparison of latent factor variances \ei
    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
                      \newline $\lambda_{ci} = \lambda_{c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline For all groups: $\bm{\kappa}=\bm{0}$
                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\nu}_A=\bm{0} \\
                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \\
                                                  \text{diag}(\bm{\Phi}_A) = \bm{I} \end{array}$ 
                   \\ \\
Threshold, Loading, and Intercept invariance
    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ and loadings $\lambda$ to be equal across groups
                  \item Restricts intercepts $\nu$ to zero in both groups
                  \item Allows comparison of latent factor variances \emph{and} means \ei
    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
                      \newline $\lambda_{ci} = \lambda_{c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline For all groups: $\bm{\nu}=\bm{0}$
                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\kappa}_A=\bm{0} \\
                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \\
                                                  \text{diag}(\bm{\Phi}_A) = \bm{I} \end{array}$ 
\\ \\ \bottomrule
\end{tabular}
\end{table}

\clearpage
\bibliographystyle{chicago}
\bibliography{/Users/giacomomason/Dropbox/Zotero/cohorts.bib}

\clearpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage

\begin{landscape}
\begin{table}[]
\caption{Item wording and prevalence, by cohort -- females}\label{tab:meantab_f}
\centering\setstretch{1.2} \footnotesize
<<MEANTAB_F, dependson="MEANTABLE", results='asis'>>=
require(xtable)
table <- meantab[[2]] # females
table <- table[,!names(table) %in% c("bcs.num", "mcs.num")] # drop if unneeded
printmeantab(table)
@
\end{table}
\end{landscape}

\begin{landscape}
\begin{table}[]
\caption{Item wording and prevalence, by cohort -- males}\label{tab:meantab_m}
\centering\setstretch{1.2} \footnotesize
<<MEANTAB_M, dependson="MEANTABLE", results='asis'>>=
require(xtable)
table <- meantab[[1]] # males
table <- table[,!names(table) %in% c("bcs.num", "mcs.num")] # drop if unneeded
printmeantab(table)
@
\end{table}
\end{landscape}

\begin{landscape}
\begin{table}[ht]
\caption{Measurement invariance fit comparison}\label{tab:fit}
\centering\setstretch{1.2} \footnotesize
<<FA_FIT_RES, dependson="FA_FIT", results='asis'>>=
require(xtable)

table <- rbind(afitab[[1]],afitab[[2]])

print(xtable(table,
                    align="L{0cm}L{.20\\textwidth}C{.05\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}",
                    digits = 4
                    ),
             size="\\small",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0,3)),
                              command = c(
                              # header
            "& & \\multicolumn{6}{c}{\\textbf{Absolute fit}} & \\multicolumn{6}{c}{\\textbf{Relative fit}} 
                  \\\\ \\cmidrule(lr){3-8} \\cmidrule(lr){9-14}
            Model & Num. params & $\\chi^2$ & RMSEA & SRMR & MFI & CFI & G-hat & $\\chi^2$ $p$ & \\textDelta RMSEA & \\textDelta SRMR &  \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ 
& (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8) & (9) & (10) & (11) & (12) & (13) \\\\ \\midrule
                       \\textbf{A: All ages} \\\\",
                      "\\newline \\textbf{B: 59-61 months sample} \\\\ \\midrule "
                                    )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabular", width = "\\textwidth",
             floating = FALSE
)

@
\justify {\scriptsize \emph{Notes}: The table presents fit indices for models of different invariance levels, following \cite{Wu2016a} Col. (1) displays the number of estimated parameters for each model. Col. (2) and (8) present the value of the $\chi^2$ statistic and the pvalue of the test of equality with respect to the configural model. Col. (3)-(7) and (9)-(13) present alternative fit indices (AFIs), in absolute values and differences from the configural model respectively. RMSEA = Root mean squared error of approximation; SRMR = standardised root mean residual; MFI = McDonald non-centrality index; CFI = comparative fit index; G-hat = gamma-hat. Panel A shows results for the whole sample of children in the BCS and MCS cohorts. Panel B is restricted to a subsample of children in the age range of maximum overlap between the two cohorts -- 59-61 months.
\par}
\end{table}
\end{landscape}

\end{document}

### 2018-07-10 codes/R/notes/backupcode.Rnw
move notes to separate folder

--- 2018-03-27 codes/R/backupcode.Rnw
@@
+% DECOMPOSITION GRAPHS AND TEXT
+
+\clearpage
+\subsection{Decomposition of inequality}
+
+We have populations $k \in \mathcal{K}$.
+For each population, we observe covariates $X_k$.
+For some $j \subseteq \mathcal{K}$, we observe outcome $Y_j$.
+We can identify the covariate distributions $F_{X_k}$ and the conditional distributions $F_{Y_j \mid X_j}$ and quantile functions $Q_{Y_j \mid X_j}$.
+Define the support of $X_k$ as $\mathcal{X}_k$ and the region of interest for the outcome as $\mathcal{Y}_j$.
+
+The interest is in the counterfactual distribution of the outcome:
+$$ F_{Y(j \mid k)}(y) = \int_{\mathcal{X}_k} F_{Y_j \mid X_j}(y \mid x) d F_{X_k}(x), \qquad y \in \mathcal{Y}_j $$
+and the corresponding quantiles:
+$$ Q_{Y(j \mid k)}(\tau) = F^\leftarrow_{Y(j \mid k)}(\tau) $$
+
+Given this, we can define counterfactual effects (CE):
+\bi
+\item Type 1 (Structure effect): CE of changing the conditional distribution: $Q_{Y(j \mid k)}(\tau) - Q_{Y(l \mid k)}(\tau)$
+\item Type 2 (Composition effect): CE of changing the covariate distribution: $Q_{Y(j \mid k)}(\tau) - Q_{Y(j \mid m)}(\tau)$
+\item Type 3: CE of changing both conditional and covariate distribution: $Q_{Y(j \mid k)}(\tau) - Q_{Y(l \mid m)}(\tau)$
+\ei
+
+In our application, we have $k=BCS,MCS$.
+We can thus decompose the observed differences in the distribution of non-cognitive skills into structure and composition effects.
+For example, the difference in the distribution of externalising skills can be written as:
+$$F_{(\theta^{EXT}_{MCS} \mid X_{MCS})} - F_{(\theta^{EXT}_{BCS} \mid X_{BCS})} =
+  \left[ F_{(\theta^{EXT}_{MCS} \mid X_{MCS})} - F_{(\theta^{EXT}_{BCS} \mid X_{MCS})} \right] +
+  \left[ F_{(\theta^{EXT}_{BCS} \mid X_{MCS})} - F_{(\theta^{EXT}_{BCS} \mid X_{BCS})} \right]$$
+
+<<COUNTERF, dependson="FA_SCORES_MAIN">>=
+# perform counterfactuals
+@
+
+<<COUNTERF_PREP_PLOTS, dependson='COUNTERF'>>=
+# prepare counterfactual plots
+@
+
+\begin{landscape}
+\begin{figure}
+<<COUNTERF_EXT, dependson='COUNTERF_PREP_PLOTS', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
+cfp.ext
+@
+\caption{Counterfactual decomposition -- externalising skills}\label{fig:counterf_ext}
+{\scriptsize \emph{Notes}: \par}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+<<COUNTERF_INT, dependson='COUNTERF_PREP_PLOTS', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
+cfp.int
+@
+\caption{Counterfactual decomposition -- internalising skills}\label{fig:counterf_int}
+{\scriptsize \emph{Notes}: \par}
+\end{figure}
+\end{landscape}
+
+<<COUNTERF_PREP_TAB, dependson='COUNTERF'>>=
+# prepare counterfactual tables
+@
+
+<<COUNTERF_INEQ, dependson='COUNTERF_PREP_TABS', results='markup'>>=
+cat("Males EXT")
+print(iqdecomp(logitcf.ext.m))
+
+cat("Males INT")
+print(iqdecomp(logitcf.int.m))
+
+cat("Females EXT")
+print(iqdecomp(logitcf.ext.f))
+
+cat("Females INT")
+print(iqdecomp(logitcf.int.f))
+@
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% COUNTERFACTUAL DECOMPOSITIONS
+
+<<COUNTERF_SEQDEC_ESTIM, dependson="SELECTVARS">>=
+# estimate sequential decompositions
+@
+
+<<COUNTERF_SEQDEC_PREP>>=
+# extract data
+@
+
+<<COUNTERF_SEQDEC_PLOT>>=
+# decomposition plots
+@
+
+\begin{landscape}
+\begin{figure}
+<<COUNTERF_SEQDEC_EXT, dependson='COUNTERF_SEQDEC_PLOT', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
+cfp.seqdec.ext
+@
+\caption{Counterfactual sequential decomposition -- externalising skills}\label{fig:counterf_seq_ext}
+{\scriptsize \emph{Notes}: \par}
+\end{figure}
+\end{landscape}
+
+\begin{landscape}
+\begin{figure}
+<<COUNTERF_SEQDEC_INT, dependson='COUNTERF_SEQDEC_PLOT', dev='tikz', fig.height=10, fig.width=16, out.width='1\\linewidth'>>=
+cfp.seqdec.int
+@
+\caption{Counterfactual sequential decomposition -- internalising skills}\label{fig:counterf_seq_int}
+{\scriptsize \emph{Notes}: \par}
+\end{figure}
+\end{landscape}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% <<FA_INV_AUX, dependson='LOAD_DATA'>>=
+% # robustness models
+% @
+
+% <<FA_SCORES_AUX, dependson='FA_INV_AUX'>>=
+% # produce scores for robustness models
+% @
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Graph for income inequality
+
+\clearpage
+\begin{figure}
+\begin{center}
+<<INCINEQ, dependson="FA_SCORES_MAIN", dev='tikz', fig.height=4, fig.width=4, out.width='.5\\linewidth'>>=
+@
+\end{center}
+\caption{Evolution of income inequality }
+{\scriptsize \emph{Notes}:  \par}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Tables for inequality regressions

### 2018-07-10 codes/R/1_load.R
add cognitive tests

@@ -135,12 +135,12 @@ for (i in 1:ncol(mcs5sdqc)) mcs5sdqcf[,i] <- as.ordered(mcs5sdqc[,i])
 mcs5aux <- cbind(mcs5data[,c("mcsid", 
                              auxvars, 
                              "rwtd",
-                             "nvoc_bastz", "psim_bastz", "patc_bastz")]
+                             "nvoc_basz", "psim_basz", "patc_basz")]
 )
 names(mcs5aux)[names(mcs5aux)=="mcsid"] <- "id"
-names(mcs5aux)[names(mcs5aux)=="nvoc_bastz"] <- "C1" # vocabulary test
-names(mcs5aux)[names(mcs5aux)=="psim_bastz"] <- "C2"
-names(mcs5aux)[names(mcs5aux)=="patc_bastz"]  <- "C3"
+names(mcs5aux)[names(mcs5aux)=="nvoc_basz"] <- "C1" # vocabulary test
+names(mcs5aux)[names(mcs5aux)=="psim_basz"] <- "C2"
+names(mcs5aux)[names(mcs5aux)=="patc_basz"]  <- "C3"
 
 # MERGE -------------------------------------------
 

### 2018-07-31 codes/R/notes/paper_jpube_partinv.Rnw
with partial invariance

--- 2018-07-02 codes/R/paper/paper_jpube.Rnw
@@
 read_chunk(here("codes/R/3_inv_main.R"))
+read_chunk(here("codes/R/3_inv_part.R"))
 read_chunk(here("codes/R/3_inv_aux.R"))
@@
 <<LOAD_DATA>>=
+@
+
+<<FUN_GETPARS>>=
+# load function that gets parameters
 @
@@
 
+<<FA_INV_PART>>=
+# partial invariance (many models)
+@
+
 <<FA_FIT>>=
@@
 
+<<FA_SCORES_MAIN>>=
+# produce scores for main model
+@
+
+<<FA_SCORES_BIND>>=
+# bind scores with data
+@
+
+<<FA_SCORES_ALT>>=
+# produce scores for alternative anchoring model
+@
+
 % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\documentclass[pdftex,10pt,twoside,a4paper]{article}
+<<FACINEQ_MAKE>>=
+# make factor inequality graphs
+@
+
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\documentclass[pdftex,10pt,a4paper]{article}
 \usepackage{amsmath}
@@
 \setlength{\topmargin}{0cm}
-\setlength{\textheight}{24cm}
+\setlength{\textheight}{23cm}
 \setlength{\oddsidemargin}{.0cm}
@@
 \setlength{\footnotesep}{0.4cm}
-\linespread{1.3}
 
@@
 
+% ADD PARTIAL INVARIANCE
+
 \clearpage
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{landscape}
@@
 
-table <- rbind(afitab[[1]],afitab[[2]])
+table <- rbind(fit_main,fit_agesub)
 
@@
              booktabs = T,
-             add.to.row = list(pos = c(list(0,3)),
+             add.to.row = list(pos = c(list(0,4)),
                               command = c(
@@
 Panel A shows results for the whole sample of children in the BCS and MCS cohorts.
-Panel B is restricted to a subsample of children in the age range of maximum overlap between the two cohorts -- 59-61 months.
+Panel B is restricted to a subsample of children in the age range of maximum overlap between the two cohorts (59-61 months).
 \par}
@@
 
+\clearpage
+\begin{table}
+\caption{Estimated parameters -- Threshold and loading invariance model (anchored)}\label{tab:measpars}
+\centering\setstretch{1.2}\scriptsize
+<<RES_FA_PAR, dependson="FA_FIT", results='asis'>>=
+
+table <- getmeaspars(fa.tlmt[[1]], groups = 4, mode = "tl")
+
+# sequences to add rows
+rws <- seq(2, (nrow(table)-1), by = 1) # start from row 2 (exclude model titles)
+rwsl <- as.list(rws)
+
+print.xtable(xtable(table,
+              align="C{0cm}C{.06\\textwidth}C{.04\\textwidth}C{.07\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}",
+              digits = 3
+                    ),
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0)),
+                              command = c(
+                              # header
+                                          " &
+                                            & \\textbf{Loadings}
+                                            & \\multicolumn{2}{c}{\\textbf{Thresholds}}
+                                            & \\multicolumn{3}{c}{\\textbf{Intercepts}}
+                                            & \\multicolumn{3}{c}{\\textbf{Variances}}
+                                            \\\\ \\cmidrule(lr){3-3} \\cmidrule(lr){4-5} \\cmidrule(lr){6-8} \\cmidrule(lr){9-11}
+
+                                            &
+                                            & $\\lambda$
+                                            & $\\tau_1$ & $\\tau_2$
+                                            & \\multicolumn{3}{c}{$\\nu$}
+                                            & \\multicolumn{3}{c}{$\\text{diag}(\\Psi)$} \\\\
+
+                                            Measure
+                                            & Factor
+                                            & All
+                                            & All & All
+                                            & BCS F & MCS M & MCS F
+                                            & BCS F & MCS M & MCS F
+                                            \\\\ "
+                      )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = "\\textwidth",
+             floating = FALSE
+)
+@
+\end{table}
+
+\begin{table}
+\caption{Estimated latent variable parameters -- Threshold and loading invariance model (anchored)}\label{tab:lvpars}
+\centering\setstretch{1.2}\scriptsize
+<<RES_FA_LV, dependson="FA_FIT", results='asis'>>=
+require(xtable)
+
+lvtab <- getlvpars(fa.tlmt[[1]], groups = 4)
+table <- rbind(lvtab$Males,lvtab$Females)
+
+print.xtable(xtable(table,
+                     align="C{0cm}L{.13\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
+                    digits = 3
+                    ),
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0,2,4)),
+                              command = c(
+                              # header
+                      "\\toprule
+                      & \\multicolumn{4}{c}{\\textbf{BCS}}
+                      & \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
+
+                      & Mean
+                      & \\multicolumn{2}{c}{Covariance}
+                      & \\multicolumn{1}{c}{Correlation}
+                      & Mean
+                      & \\multicolumn{2}{c}{Covariance}
+                      & \\multicolumn{1}{c}{Correlation} \\\\
+
+                      \\textbf{Males} \\\\",
+                      "\\textbf{Females} \\\\",
+                      "\\bottomrule"
+                                    )),
+             hline.after = NULL,  # remove default booktabs lines
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = "\\textwidth",
+             floating = FALSE
+)
+
+@
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\clearpage
+
+<<FA_FIT_PART, dependson="FA_INV_PART", results='markup'>>=
+head(partialfit[order(partialfit$rmsea),], 15)
+@
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}
+<<FACDENS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
+@
+\caption{Scores}
+\justify {\scriptsize \emph{Notes}: Density of scores from partial invariance model with constrained intercepts on items 1, 6, 9, 10.
+\par}
+\end{figure}
+
+\begin{figure}
+<<FACDENS_ALT, dependson="FA_SCORES_BIND", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
+# graph x axis boundaries
+maxx <- max(scores2plot_alt[names(scores2plot_alt) %in% c("EXT","INT")]) +.05
+minx <- min(scores2plot_alt[names(scores2plot_alt) %in% c("EXT","INT")]) -.05
+
+# common options
+addopts.dens <- function(x) {
+  x <- x +
+    theme(
+      axis.title.y=element_blank(),
+      legend.position="none"
+    ) +
+    scale_x_continuous(name = "Score", breaks = seq(-3.5,1.5,.5)) +
+    coord_cartesian(xlim = c(minx, maxx), ylim = c(0,.62)) +
+    scale_fill_discrete("") + # remove fill guide title
+    scale_colour_discrete(guide=FALSE) +  # remove colour legend
+    geom_density(alpha = 0.1, bw = "nrd", adjust=1.5, kernel = "epanechnikov")
+  return(x)
+}
+
+# densities of factor scores
+pdext.ebm.m <- ggplot(subset(scores2plot_alt, sex=="M"), aes(x=EXT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("EXT Scores (Males)")
+pdint.ebm.m <- ggplot(subset(scores2plot_alt, sex=="M"), aes(x=INT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("INT Scores (Males)")
+pdext.ebm.f <- ggplot(subset(scores2plot_alt, sex=="F"), aes(x=EXT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("EXT Scores (Females)")
+pdint.ebm.f <- ggplot(subset(scores2plot_alt, sex=="F"), aes(x=INT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("INT Scores (Females)")
+denslist <- list(pdext.ebm.m, pdint.ebm.m, pdext.ebm.f, pdint.ebm.f) 
+denslist <- lapply(denslist, addopts.dens) # apply options to all graphs
+pcol <- plot_grid( denslist[[1]], denslist[[3]], denslist[[2]], denslist[[4]],
+                   align = 'vh',
+                   hjust = -1,
+                   nrow = 2
+)
+# add legend
+legend_b <- get_legend(denslist[[1]] + theme(legend.position="bottom"))
+p <- plot_grid( pcol, legend_b, ncol = 1, rel_heights = c(1, .1))
+p
+
+@
+\caption{Scores -- alternative restrictions}
+\justify {\scriptsize \emph{Notes}: Density of scores from partial invariance model with constrained intercepts on items 3, 6, 8, 11.
+\par}
+\end{figure}
+
+\begin{figure}
+<<FACDENS_ALT2, dependson="FA_SCORES_BIND", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
+# graph x axis boundaries
+maxx <- max(scores2plot_alt2[names(scores2plot_alt2) %in% c("EXT","INT")]) +.05
+minx <- min(scores2plot_alt2[names(scores2plot_alt2) %in% c("EXT","INT")]) -.05
+
+# common options
+addopts.dens <- function(x) {
+  x <- x +
+    theme(
+      axis.title.y=element_blank(),
+      legend.position="none"
+    ) +
+    scale_x_continuous(name = "Score", breaks = seq(-3.5,1.5,.5)) +
+    coord_cartesian(xlim = c(minx, maxx), ylim = c(0,.62)) +
+    scale_fill_discrete("") + # remove fill guide title
+    scale_colour_discrete(guide=FALSE) +  # remove colour legend
+    geom_density(alpha = 0.1, bw = "nrd", adjust=1.5, kernel = "epanechnikov")
+  return(x)
+}
+
+# densities of factor scores
+pdext.ebm.m <- ggplot(subset(scores2plot_alt2, sex=="M"), aes(x=EXT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("EXT Scores (Males)")
+pdint.ebm.m <- ggplot(subset(scores2plot_alt2, sex=="M"), aes(x=INT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("INT Scores (Males)")
+pdext.ebm.f <- ggplot(subset(scores2plot_alt2, sex=="F"), aes(x=EXT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("EXT Scores (Females)")
+pdint.ebm.f <- ggplot(subset(scores2plot_alt2, sex=="F"), aes(x=INT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("INT Scores (Females)")
+denslist <- list(pdext.ebm.m, pdint.ebm.m, pdext.ebm.f, pdint.ebm.f) 
+denslist <- lapply(denslist, addopts.dens) # apply options to all graphs
+pcol <- plot_grid( denslist[[1]], denslist[[3]], denslist[[2]], denslist[[4]],
+                   align = 'vh',
+                   hjust = -1,
+                   nrow = 2
+)
+# add legend
+legend_b <- get_legend(denslist[[1]] + theme(legend.position="bottom"))
+p <- plot_grid( pcol, legend_b, ncol = 1, rel_heights = c(1, .1))
+p
+
+@
+\caption{Scores -- alternative restrictions}
+\justify {\scriptsize \emph{Notes}: Density of scores from partial invariance model with constrained intercepts on items 3, 6, 9, 10.
+\par}
+\end{figure}
+
+\begin{figure}
+<<FACINEQ_FSC, dependson="FA_SCORES_BIND", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
+# plot
+fi <- fi_fsc # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\caption{Inequality by father occupation -- normalised}
+\end{figure}
+
+\begin{figure}
+<<FACINEQ_FSC2, dependson="FA_SCORES_BIND", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
+# plot
+fi <- fi_fsc2 # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\caption{Inequality by father occupation -- not normalised}
+\end{figure}
+
+\begin{figure}
+<<FACINEQ_FSC3, dependson="FA_SCORES_BIND", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
+fi_fsc3 <- plotineq(scores2plot_alt, "fscl5wb", ylab = "Score ", xlab = "Father Occupation (5)",
+                   ylim = c(-.8,1.0))
+fi <- fi_fsc3 # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\caption{Inequality by father occupation -- not normalised, alternative restrictions}
+\end{figure}
+
+\begin{figure}
+<<FACINEQ_PS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
+# plot
+fi <- fi_ps # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\caption{Inequality by mother education -- normalised}
+\end{figure}
+
+\begin{figure}
+<<FACINEQ_PS2, dependson="FA_SCORES_BIND", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
+# plot
+fi <- fi_ps2 # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\caption{Inequality by mother education -- not normalised}
+\end{figure}
+
 \end{document}

### 2018-08-02 codes/R/paper/paper_jpube.Rnw
sent to coauthors 2/8/18

@@
-<<MAIN,cache=FALSE,echo=FALSE,results='hide',message=FALSE,warning=FALSE>>=
-rm(list=ls())         # Clear all objects from memory 
-  opts_chunk$set(concordance=FALSE,
-                 echo=FALSE,
-                 tidy=FALSE,
-                 autodep=TRUE,
-                 cache=T, # REPLACE LATER
-                 message=FALSE, 
-                 warning=FALSE,
-                 size='scriptsize',
-                 results='hide',
-                 comment=NA, 
-                 fig.width=4,
-                 fig.height=4, 
-                 out.width='\\textwidth',
-                 fig.show='hold',
-                 tikzDefaultEngine='pdftex',
-                 tikzDocumentDeclaration = "\\documentclass[10pt]{article}",
-                 tikzMetricPackages = c("\\usepackage[utf8]{inputenc}","\\usepackage[T1]{fontenc}", "\\usetikzlibrary{calc}", "\\usepackage{amssymb}")
-                 )
-
-@
-
-<<READ, cache=FALSE >>=
-library(here)
-read_chunk(here("codes/R/1_load.R"))
-read_chunk(here("codes/R/2_efa.R"))
-read_chunk(here("codes/R/3_inv_main.R"))
-read_chunk(here("codes/R/3_inv_aux.R"))
-read_chunk(here("codes/R/3_inv_fit.R"))
-read_chunk(here("codes/R/4_scores.R"))
-read_chunk(here("codes/R/4_scores_aux.R"))
-read_chunk(here("codes/R/5_graphs.R"))
-read_chunk(here("codes/R/6_selectvars.R"))
-read_chunk(here("codes/R/8_regs_dep.R"))
-read_chunk(here("codes/R/8_regs_outc.R"))
-read_chunk(here("codes/R/9_gelbfig.R"))
-read_chunk(here("codes/R/getpars.R"))
-
-@
-
-<<PREAMBLE >>=
-@
-
-<<LOAD_DATA>>=
-@
-
-<<MEANTABLE>>=
-# make table with means
-@
-
-<<MEANTABLE_PRINT>>=
-# function to print mean table
-printmeantab <- function(tab) {
-  print(xtable(tab,
-               align="L{0cm}C{.04\\textwidth}C{.05\\textwidth}C{.05\\textwidth}L{.4\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}L{.4\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
-               digits = 1
-  ), 
-  size="\\small",
-  include.rownames=F,
-  include.colnames=F,
-  booktabs = T,
-  add.to.row = list(pos = c(list(0)),
-                    command = c(
-                      # header
-                      paste(" & & &
-                            \\multicolumn{4}{c}{\\textbf{BCS}} &
-                            \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){4-7} \\cmidrule(lr){8-11} 
-                            Item & Factor & Num. Categ. &
-                            Wording & Cert. Appl. (\\%) & Smtm. Appl. (\\%) & Appl. (\\%) &
-                            Wording & Cert. True (\\%) & Smwt. True (\\%) & True (\\%) \\\\ \\midrule
-                            "
-                            , sep="")
-                    )),
-  sanitize.text.function = function(x){x},
-  sanitize.rownames.function = function(x){""}, # to not print row numbers
-  tabular.environment = "tabular", width = ".8\\textwidth",
-  floating = FALSE
-  )
-}
-
-@
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-<<FA_INV_MAIN>>=
-# main models only
-@
-
-<<FA_FIT>>=
-# assemble fit indices into matrices
-@
-
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\documentclass[pdftex,10pt,twoside,a4paper]{article}
+\documentclass[pdftex,10pt,a4paper]{article}
 \usepackage{amsmath}
@@
 \maketitle
+
+<<MAIN,cache=FALSE,echo=FALSE,results='hide',message=FALSE,warning=FALSE>>=
+rm(list=ls())         # Clear all objects from memory 
+  opts_chunk$set(concordance=FALSE,
+                 echo=FALSE,
+                 tidy=FALSE,
+                 autodep=TRUE,
+                 cache=T, # REPLACE LATER
+                 message=FALSE, 
+                 warning=FALSE,
+                 size='scriptsize',
+                 results='hide',
+                 comment=NA, 
+                 fig.width=4,
+                 fig.height=4, 
+                 out.width='\\textwidth',
+                 fig.show='hold',
+                 tikzDefaultEngine='pdftex',
+                 tikzDocumentDeclaration = "\\documentclass[10pt]{article}",
+                 tikzMetricPackages = c("\\usepackage[utf8]{inputenc}","\\usepackage[T1]{fontenc}", "\\usetikzlibrary{calc}", "\\usepackage{amssymb}")
+                 )
+
+@
+
+<<READ, cache=FALSE >>=
+library(here)
+read_chunk(here("codes/R/1_load.R"))
+read_chunk(here("codes/R/2_efa.R"))
+read_chunk(here("codes/R/3_inv_main.R"))
+read_chunk(here("codes/R/3_inv_aux.R"))
+read_chunk(here("codes/R/3_inv_fit.R"))
+read_chunk(here("codes/R/4_scores.R"))
+read_chunk(here("codes/R/4_scores_aux.R"))
+read_chunk(here("codes/R/5_graphs.R"))
+read_chunk(here("codes/R/5_graphs_ineq.R"))
+read_chunk(here("codes/R/5_graphs_item.R"))
+read_chunk(here("codes/R/6_selectvars.R"))
+read_chunk(here("codes/R/8_regs_dep.R"))
+read_chunk(here("codes/R/8_regs_outc.R"))
+read_chunk(here("codes/R/99_auxfun.R"))
+
+@
+
+<<PREAMBLE >>=
+@
+
+<<LOAD_DATA>>=
+@
+
+<<AUXFUN>>=
+# load function that gets parameters
+@
+
+<<MEANTABLE>>=
+# make table with means
+@
+
+<<MEANTABLE_PRINT>>=
+# function to print mean table
+printmeantab <- function(tab) {
+  print(xtable(tab,
+               align="L{0cm}C{.04\\textwidth}C{.05\\textwidth}C{.05\\textwidth}L{.35\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}L{.35\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
+               digits = 1
+  ), 
+  size="\\small",
+  include.rownames=F,
+  include.colnames=F,
+  booktabs = T,
+  add.to.row = list(pos = c(list(0)),
+                    command = c(
+                      # header
+                      paste(" & & &
+                            \\multicolumn{4}{c}{\\textbf{BCS}} &
+                            \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){4-7} \\cmidrule(lr){8-11} 
+                            Item & Factor & Num. Categ. &
+                            Wording & Cert. Appl. (\\%) & Smtm. Appl. (\\%) & Appl. (\\%) &
+                            Wording & Cert. True (\\%) & Smwt. True (\\%) & True (\\%) \\\\ \\midrule
+                            "
+                            , sep="")
+                    )),
+  sanitize.text.function = function(x){x},
+  sanitize.rownames.function = function(x){""}, # to not print row numbers
+  tabular.environment = "tabular", width = ".8\\textwidth",
+  floating = FALSE
+  )
+}
+
+@
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+<<FA_INV_MAIN>>=
+# main models only
+@
+
+<<FA_INV_AUX_SYN>>=
+# syntax for robustness models
+@
+
+<<FA_INV_AUX_GEND>>=
+# robustness checks for gender alone
+@
+
+<<FA_FIT>>=
+# assemble fit indices into matrices
+@
+
+<<FA_SCORES_MAIN>>=
+# produce scores for main model
+@
+
+<<FA_SCORES_BIND>>=
+# bind scores with data
+@
+
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+<<FACINEQ_MAKE>>=
+# make factor inequality graphs
+@
+
+<<ITEMINEQ_MAKE>>=
+# make item-level inequality graphs
+@
+
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+<<SELECTVARS>>=
+# define variables and formulas for regression and decomposition
+@
+
+<<REGS_DEP, dependson="SELECTVARS">>=
+# run regressions explaining EXT INT
+@
+
+<<LOAD_DATA_OUTC, dependson="SELECTVARS">>=
+# load outcomes data
+@
+
+<<REGS_OUTC, dependson="LOAD_DATA_OUTC">>=
+# run regressions explaining adolescent/adult outcomes
+@
+
+<<REGS_OUTC_TAB, dependson="REGS_OUTC">>=
+# make tables from regressions
+@
 
@@
 \caption{Measurement invariance fit comparison}\label{tab:fit}
-\centering\setstretch{1.2} \footnotesize
+\centering\setstretch{1.2} \scriptsize
 <<FA_FIT_RES, dependson="FA_FIT", results='asis'>>=
@@
 
-table <- rbind(afitab[[1]],afitab[[2]])
+# MAKE FIT TABLES
+fit_main <- afitab(list(fa.c[[1]], fa.tl[[1]], fa.tli[[1]]), 
+       c("Configural", 
+         "Threshold + Loading Inv", 
+         "Threshold, Loading, + Intercept Inv")
+)
+
+fit_agesub <- afitab(list(fa.c[[2]], fa.tl[[2]], fa.tli[[2]]), 
+                   c("Configural", 
+                     "Threshold + Loading Inv", 
+                     "Threshold, Loading, + Intercept Inv")
+)
+fit_males <- afitab(list(fa.c[[4]], fa.tl[[4]], fa.tli[[4]]), 
+                   c("Configural", 
+                     "Threshold + Loading Inv", 
+                     "Threshold, Loading, + Intercept Inv")
+)
+fit_females <- afitab(list(fa.c[[5]], fa.tl[[5]], fa.tli[[5]]), 
+                   c("Configural", 
+                     "Threshold + Loading Inv", 
+                     "Threshold, Loading, + Intercept Inv")
+)
+
+table <- rbind(fit_main,fit_agesub,fit_males,fit_females)
 
 print(xtable(table,
-                    align="L{0cm}L{.20\\textwidth}C{.05\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}",
+                    align="L{0cm}L{.20\\textwidth}C{.05\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
                     digits = 4
                     ),
-             size="\\small",
+             size="\\footnotesize",
              include.rownames=F,
@@
              booktabs = T,
-             add.to.row = list(pos = c(list(0,3)),
+             add.to.row = list(pos = c(list(0,3,6,9)),
                               command = c(
@@
                   \\\\ \\cmidrule(lr){3-8} \\cmidrule(lr){9-14}
-            Model & Num. params & $\\chi^2$ & RMSEA & SRMR & MFI & CFI & G-hat & $\\chi^2$ $p$ & \\textDelta RMSEA & \\textDelta SRMR &  \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ 
+            Model & Num. params & $\\chi^2$ & RMSE & SRMR & MFI & CFI & G-hat & $\\chi^2$ $p$ & \\textDelta RMSE & \\textDelta SRMR &  \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ 
 & (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8) & (9) & (10) & (11) & (12) & (13) \\\\ \\midrule
-                       \\textbf{A: All ages} \\\\",
-                      "\\newline \\textbf{B: 59-61 months sample} \\\\ \\midrule "
+                       \\textbf{A: Entire sample} \\\\",
+                      "\\newline \\textbf{B: 59-61 months sample} \\\\ \\midrule ",
+                      "\\newline \\textbf{C: Males only} \\\\ \\midrule ",
+                      "\\newline \\textbf{D: Females only} \\\\ \\midrule "
                                     )),
@@
 (3)-(7) and (9)-(13) present alternative fit indices (AFIs), in absolute values and differences from the configural model respectively.
-RMSEA = Root mean squared error of approximation; SRMR = standardised root mean residual; MFI = McDonald non-centrality index; CFI = comparative fit index; G-hat = gamma-hat.
+RMSE = Root mean squared error of approximation; SRMR = standardised root mean residual; MFI = McDonald non-centrality index; CFI = comparative fit index; G-hat = gamma-hat.
 Panel A shows results for the whole sample of children in the BCS and MCS cohorts.
-Panel B is restricted to a subsample of children in the age range of maximum overlap between the two cohorts -- 59-61 months.
+Panel B is restricted to a subsample of children in the age range of maximum overlap between the two cohorts (59-61 months).
 \par}
@@
 
+\clearpage
+\begin{table}
+\caption{Estimated parameters -- Threshold and loading invariance model (anchored)}\label{tab:measpars}
+\centering\setstretch{1.2}\footnotesize
+<<RES_FA_PAR, dependson="FA_FIT", results='asis'>>=
+
+table <- getmeaspars(fa.tlmt[[1]], groups = 4, mode = "tl")
+
+# sequences to add rows
+rws <- seq(2, (nrow(table)-1), by = 1) # start from row 2 (exclude model titles)
+rwsl <- as.list(rws)
+
+print.xtable(xtable(table,
+              align="C{0cm}C{.06\\textwidth}C{.04\\textwidth}C{.07\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}",
+              digits = 3
+                    ),
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0)),
+                              command = c(
+                              # header
+                                          " &
+                                            & \\textbf{Loadings}
+                                            & \\multicolumn{2}{c}{\\textbf{Thresholds}}
+                                            & \\multicolumn{3}{c}{\\textbf{Intercepts}}
+                                            & \\multicolumn{3}{c}{\\textbf{Variances}}
+                                            \\\\ \\cmidrule(lr){3-3} \\cmidrule(lr){4-5} \\cmidrule(lr){6-8} \\cmidrule(lr){9-11}
+
+                                            &
+                                            & $\\lambda$
+                                            & $\\tau_1$ & $\\tau_2$
+                                            & \\multicolumn{3}{c}{$\\nu$}
+                                            & \\multicolumn{3}{c}{$\\text{diag}(\\Psi)$} \\\\
+
+                                            Measure
+                                            & Factor
+                                            & All
+                                            & All & All
+                                            & BCS F & MCS M & MCS F
+                                            & BCS F & MCS M & MCS F
+                                            \\\\ "
+                      )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = "\\textwidth",
+             floating = FALSE
+)
+@
+\end{table}
+
+\begin{table}
+\caption{Estimated latent variable parameters -- Threshold and loading invariance model (anchored)}\label{tab:lvpars}
+\centering\setstretch{1.2}\footnotesize
+<<RES_FA_LV, dependson="FA_FIT", results='asis'>>=
+require(xtable)
+
+lvtab <- getlvpars(fa.tlmt[[1]], groups = 4)
+table <- rbind(lvtab$Males,lvtab$Females)
+
+print.xtable(xtable(table,
+                     align="C{0cm}L{.13\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
+                    digits = 3
+                    ),
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0,2,4)),
+                              command = c(
+                              # header
+                      "\\toprule
+                      & \\multicolumn{4}{c}{\\textbf{BCS}}
+                      & \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
+
+                      & Mean
+                      & \\multicolumn{2}{c}{Covariance}
+                      & \\multicolumn{1}{c}{Correlation}
+                      & Mean
+                      & \\multicolumn{2}{c}{Covariance}
+                      & \\multicolumn{1}{c}{Correlation} \\\\
+
+                      \\textbf{Males} \\\\",
+                      "\\textbf{Females} \\\\",
+                      "\\bottomrule"
+                                    )),
+             hline.after = NULL,  # remove default booktabs lines
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = "\\textwidth",
+             floating = FALSE
+)
+
+@
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}
+<<FACDENS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
+@
+\caption{Scores}
+\justify {\scriptsize \emph{Notes}: Density of scores from final model.
+\par}
+\end{figure}
+
+\begin{figure}
+<<RAWHIST, dependson="FA_SCORES_BIND", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
+@
+\caption{Sum scores}
+\justify {\scriptsize \emph{Notes}: Density of sum scores without factor analysis.
+\par}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\caption{Quantile differences in scores}\label{tab:iqdiff}
+\centering\setstretch{1.2}\footnotesize
+<<INEQ_STATS, dependson="FA_SCORES_BIND", results='asis'>>=
+
+# income 
+tb <- data.frame(
+  aggregate(finaldata$faminc10_infl, by=list(finaldata$cohort), FUN = function(x) quantile(x, c(.1, .25, .75, .9), na.rm=T))
+)
+tb$d9010 <- tb$x[,4]/tb$x[,1]
+tb$d7525 <- tb$x[,3]/tb$x[,2]
+
+tb.ext <- data.frame(
+  aggregate(finaldata$EXT, by=list(finaldata$cohortsex), FUN = function(x) quantile(x, c(.1, .25, .5, .75, .9), na.rm=T))
+)
+tb.ext$d9010 <- abs(tb.ext$x[,"90%"]-tb.ext$x[,"10%"])
+tb.ext$d7525 <- abs(tb.ext$x[,"75%"]-tb.ext$x[,"25%"])
+tb.ext$d9050 <- abs(tb.ext$x[,"90%"]-tb.ext$x[,"50%"])
+tb.ext$d5010 <- abs(tb.ext$x[,"50%"]-tb.ext$x[,"10%"])
+rownames(tb.ext) <- tb.ext$Group.1
+tb.ext <- t(tb.ext[,-c(1,2)]) # remove aggregated quantiles and transpose
+tb.ext <- tb.ext[,c("BCS.M","MCS.M","BCS.F","MCS.F")]
+
+tb.int <- data.frame(
+  aggregate(finaldata$INT, by=list(finaldata$cohortsex), FUN = function(x) quantile(x, c(.1, .25, .5, .75, .9), na.rm=T))
+)
+tb.int$d9010 <- abs(tb.int$x[,"90%"]-tb.int$x[,"10%"])
+tb.int$d7525 <- abs(tb.int$x[,"75%"]-tb.int$x[,"25%"])
+tb.int$d9050 <- abs(tb.int$x[,"90%"]-tb.int$x[,"50%"])
+tb.int$d5010 <- abs(tb.int$x[,"50%"]-tb.int$x[,"10%"])
+rownames(tb.int) <- tb.int$Group.1
+tb.int <- t(tb.int[,-c(1,2)]) # remove aggregated quantiles and transpose
+tb.int <- tb.int[,c("BCS.M","MCS.M","BCS.F","MCS.F")]
+
+# assemble and print
+table <- rbind(tb.ext,tb.int)
+table <- cbind(data.frame(as.matrix(rep(c("90-10", "75-25", "90-50", "50-10"),2))),
+               table)
+
+print(xtable(table,
+                    align="L{0cm}L{.15\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}",
+                    digits = 2
+                    ),
+             size="\\footnotesize",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0,4)),
+                              command = c(
+                              # header
+            " & \\multicolumn{2}{c}{\\textbf{Males}} & \\multicolumn{2}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-3}  \\cmidrule(lr){4-5} 
+                       & BCS & MCS & BCS & MCS  \\\\
+                      \\textbf{Externalising} \\\\ ",
+                      "\\textbf{Internalising} \\\\ \\midrule "
+                                    )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = "\\textwidth",
+             floating = FALSE
+)
+
+@
+\end{table}
+
+\begin{figure}
+<<FACINEQ_FSC, dependson="FA_SCORES_BIND", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
+fi <- fi_fsc # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\caption{Inequality by father occupation -- normalised}
+\end{figure}
+
+\begin{figure}
+<<FACINEQ_PS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
+fi <- fi_ps # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\caption{Inequality by mother education -- normalised}
+\end{figure}
+
+\begin{figure}
+<<FACINEQ_SM, dependson="FA_SCORES_BIND", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
+fi <- fi_sm # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\caption{Inequality by mother smoking in pregnancy -- normalised}
+\end{figure}
+
+\begin{figure}
+<<ITEMINEQ_PS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=8, fig.width=12, out.width='\\linewidth'>>=
+itempl_ps
+@
+\caption{Item-level changes in inequality -- mother education}
+\justify {\scriptsize \emph{Notes}: The graph displays the ratio between low and high SES in the prevalence of each behaviour, separately by cohort.
+\par}
+\end{figure}
+
+\begin{figure}
+<<ITEMINEQ_FSC, dependson="FA_SCORES_BIND", dev='tikz', fig.height=8, fig.width=12, out.width='\\linewidth'>>=
+itempl_fsc
+@
+\caption{Item-level changes in inequality -- father occupation}
+\justify {\scriptsize \emph{Notes}: The graph displays the ratio between low and high SES in the prevalence of each behaviour, separately by cohort.
+\par}
+\end{figure}
+
+\begin{figure}
+<<ITEMINEQ_SM, dependson="FA_SCORES_BIND", dev='tikz', fig.height=8, fig.width=12, out.width='\\linewidth'>>=
+itempl_sm
+@
+\caption{Item-level changes in inequality -- mother smoking in pregnancy}
+\justify {\scriptsize \emph{Notes}: The graph displays the ratio between low and high SES in the prevalence of each behaviour, separately by cohort.
+\par}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{table}
+\caption{Determinants of Externalising Skills}
+\centering\footnotesize
+<<REGS_EXT, results='asis', dependson="REGS_DEP">>=
+
+# select models
+regs <- list(r_ext$BCS.M[["OLS"]],r_ext$MCS.M[["OLS"]],r_ext$BCS.F[["OLS"]],r_ext$MCS.F[["OLS"]])
+
+# robust SEs
+robust_se <- list()
+robust_p  <- list()
+for (i in 1:length(regs)) {
+  cov1              <- sandwich::vcovHC(regs[[i]], type = "HC1")
+  ct                <- lmtest::coeftest(regs[[i]], vcov = cov1)
+  robust_se[[i]] <- ct[, 2]
+  robust_p[[i]] <- ct[, 4]
+}
+
+texreg::texreg(regs,
+               stars = c(0.1, 0.05, 0.01), digits = 3,
+               override.se = robust_se, override.pvalues = robust_p,
+               single.row = FALSE, scriptsize = TRUE,
+               caption = "", label = "",
+               omit.coef = omitcoefs,
+               groups = grouplabs,
+               custom.columns = list("p-value" = r_ext$Mp, "p-value" = r_ext$Fp), custom.col.pos = c(4,6),
+               custom.coef.names = varlabs,
+               custom.model.names = c("Males BCS", "Males MCS", "Females BCS", "Females MCS"),
+               custom.note = "",
+               include.rmse = FALSE,
+               table = FALSE
+               )
+
+@
+\justify {\scriptsize \emph{Notes}: Robust standard errors in parentheses.
+All estimates additionally control for region of birth, mother height, number of previous stillbirths at child's birth, preterm birth, a dummy for missing gestational age, and number of other children in the household at child age 5.
+\par}
+\end{table}
+
+\begin{table}
+\caption{Determinants of Internalising Skills}
+\centering\footnotesize
+<<REGS_INT, results='asis', dependson="REGS_DEP">>=
+
+# select models
+regs <- list(r_int$BCS.M[["OLS"]],r_int$MCS.M[["OLS"]],r_int$BCS.F[["OLS"]],r_int$MCS.F[["OLS"]])
+
+# robust SEs
+robust_se <- list()
+robust_p  <- list()
+for (i in 1:length(regs)) {
+  cov1              <- sandwich::vcovHC(regs[[i]], type = "HC1")
+  ct                <- lmtest::coeftest(regs[[i]], vcov = cov1)
+  robust_se[[i]] <- ct[, 2]
+  robust_p[[i]] <- ct[, 4]
+}
+
+texreg::texreg(regs,
+               stars = c(0.1, 0.05, 0.01), digits = 3,
+               override.se = robust_se, override.pvalues = robust_p,
+               single.row = FALSE, scriptsize = TRUE,
+               caption = "", label = "",
+               omit.coef = omitcoefs,
+               groups = grouplabs,
+               custom.columns = list("p-value" = r_int$Mp, "p-value" = r_int$Fp), custom.col.pos = c(4,6),
+               custom.coef.names = varlabs,
+               custom.model.names = c("Males BCS", "Males MCS", "Females BCS", "Females MCS"),
+               custom.note = "",
+               include.rmse = FALSE,
+               table = FALSE
+               )
+
+@
+\justify {\scriptsize \emph{Notes}: Robust standard errors in parentheses.
+All estimates additionally control for region of birth, mother height, number of previous stillbirths at child's birth, preterm birth, a dummy for missing gestational age, and number of other children in the household at child age 5.
+\par}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+<<REGS_OUTC_TABPR, dependson="REGS_OUTC">>=
+# function to display outcome regression tables
+# numoutc = number of outcomes
+printoutctab <- function(tab, numoutc) {
+print(xtable(tab,
+                    align="R{0cm}L{.2\\textwidth}C{.05\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}",
+                    digits = 3
+                    ),
+             size="\\footnotesize",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = as.list(c(0, 8*seq(1,numoutc-1))),
+                              command = c(
+                              # header
+            " & \\multicolumn{4}{c}{\\textbf{Males}} & \\multicolumn{4}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-5}  \\cmidrule(lr){6-9}
+& Mean & \\multicolumn{3}{c}{Coefficients} & Mean & \\multicolumn{3}{c}{Coefficients} \\\\ \\cmidrule(lr){3-5}  \\cmidrule(lr){7-9}
+            & (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8) \\\\",
+            rep(" \\midrule", numoutc-1)
+            )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular",
+             floating = FALSE
+)
+}
+@
+
+\begin{table}
+\centering\footnotesize
+\caption{Predictors of adolescent and adult outcomes -- BCS}
+<<REGS_OUTC_BCS, results='asis', dependson="REGS_OUTC">>=
+
+# collapse to matrix
+bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab[c("smktry16", "bmi16", "hnvq30", "lhgrpay38")],matrix,ncol=9,byrow=FALSE))
+table <- bcs_outctab_mat
+printoutctab(table, 4)
+
+@
+\justify {\scriptsize \emph{Notes}: Robust standard errors in parentheses.
+All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
+\par}
+\end{table}
+
+\begin{table}
+\centering\footnotesize
+\caption{Predictors of adolescent outcomes -- MCS}
+<<REGS_OUTC_MCS, results='asis', dependson="REGS_OUTC">>=
+
+# collapse to matrix
+mcs_outctab_mat <- do.call(rbind,lapply(mcs_outctab[c("smktry14", "bmi14")],matrix,ncol=9,byrow=FALSE))
+table <- mcs_outctab_mat
+printoutctab(table, 2)
+
+@
+\justify {\scriptsize \emph{Notes}: Robust standard errors in parentheses.
+All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
+\par}
+\end{table}
+
 \end{document}

### 2018-08-02 codes/R/8_regs_outc.R

@@ -52,82 +52,39 @@ mcsoutcvars <- names(mcsoutclabs)
 # REGRESSIONS OF ADOLESCENT OUTCOMES ON EXT/INT & CHILDHOOD VARIABLES
 
 # estimate BCS models
-bcsoutcmod_m_ext <- list()
-bcsoutcmod_m_int <- list()
-bcsoutcmod_m_b <- list()
-bcsoutcmod_m_bc <- list()
-bcsoutcmod_f_ext <- list()
-bcsoutcmod_f_int <- list()
-bcsoutcmod_f_b <- list()
-bcsoutcmod_f_bc <- list()
+bcsoutcmod_b <- list()
+bcsoutcmod_bc <- list()
+bcsoutcmod_bcr <- list()
 for (o in 1:length(bcsoutcvars)) {
-  form_outc_ext  <- as.formula(paste0(bcsoutcvars[o]," ~ EXT + age16 + region + ", pplus(decvars)))
-  form_outc_int  <- as.formula(paste0(bcsoutcvars[o]," ~ INT + age16 + region + ", pplus(decvars)))
   form_outc_b    <- as.formula(paste0(bcsoutcvars[o]," ~ EXT + INT + age16 + region +", pplus(decvars)))
   form_outc_bc   <- as.formula(paste0(bcsoutcvars[o]," ~ EXT + INT + age16 + region + cogscore +", pplus(decvars)))
-  bcsoutcmod_m_ext[[bcsoutcvars[o]]] <- lm(form_outc_ext, data = subset(bcsoutc, sex=="M"))
-  bcsoutcmod_m_int[[bcsoutcvars[o]]] <- lm(form_outc_int, data = subset(bcsoutc, sex=="M"))
-  bcsoutcmod_m_b[[bcsoutcvars[o]]]   <- lm(form_outc_b, data = subset(bcsoutc, sex=="M"))
-  bcsoutcmod_m_bc[[bcsoutcvars[o]]]   <- lm(form_outc_bc, data = subset(bcsoutc, sex=="M"))
-  bcsoutcmod_f_ext[[bcsoutcvars[o]]] <- lm(form_outc_ext, data = subset(bcsoutc, sex=="F"))
-  bcsoutcmod_f_int[[bcsoutcvars[o]]] <- lm(form_outc_int, data = subset(bcsoutc, sex=="F"))
[…]

### 2018-08-02 codes/R/99_auxfun.R

@@ -181,8 +181,8 @@ cellpr <- function(mod, vr) {
            " \\newline ($", prcoef(round(coefs2[2],3)), "$)")
   ) # SE (on new line)
 }
-# print R2
+# print R2 (with padding)
 prr2 <- function(mod) {
-  return(sprintf("%.4f", summary(mod)$adj.r.squared))
+  return(paste0("\\newline ", sprintf("%.4f", summary(mod)$adj.r.squared)))
 }
 

### 2018-09-03 codes/R/99_auxfun.R
new cellpr function

@@ -169,18 +169,32 @@ stars <- function(t) {
   else if (abs(t) <= 1.96 & abs(t) > 1.64) return("^{*}")
   else return("")
 }
+
 # function to extract estimate, put stars and robust SE
-cellpr <- function(mod, vr) {
-  # mod: lm model
+cellpr2 <- function(mode, vr) {
+  # mod: estimates from ML model
   # vrs: variable to extract
   coefs <- lmtest::coeftest(mod, vcov = sandwich::vcovHC(mod, type = "HC1"))
   coefs2 <- coefs[vr,]
   return(
     paste0("$", prcoef(round(coefs2[1],3)), # estimate
            stars(coefs2[3]), "$", # stars
+           " \\newline ($", prcoef(round(coefs2[2],3)), "$)") # SE (on new line)
+  ) 
+}
+
+# function to extract estimate, put stars and (bootstrap) SE
+cellpr <- function(estimtab, vr) {
+  # estimtab: table with estimate, SE, tstat
+  # vrs: variable to extract
+  coefs2 <- estimtab[vr,]
+  return(
+    paste0("$", prcoef(round(coefs2[1],3)), # estimate
+           stars(coefs2[3]), "$", # stars
            " \\newline ($", prcoef(round(coefs2[2],3)), "$)")
   ) # SE (on new line)
 }
+
 # print R2 (with padding)
 prr2 <- function(mod) {
   return(paste0("\\newline ", sprintf("%.4f", summary(mod)$adj.r.squared)))

### 2018-09-03 codes/R/8_regs_dep.R
pvalues in regs dep (not finished)

@@ -2,7 +2,8 @@
 ## ---- REGS_DEP
 
 # function to extract relevant pvalues from interacted model
-extractp <- function(mod) {
+# outstring: whether to output as string or as numeric (so can be averaged for bootstrap)
+extractp <- function(mod, outstring = FALSE) {
   coefs <- summary(mod)$coefficients
   coefs <- data.frame(tail(coefs, dim(coefs)[1]/2)) # select interactions (second half of table)
   colnames(coefs) <- c("est", "se", "t", "pval")
@@ -13,13 +14,16 @@ extractp <- function(mod) {
     filter(var2!="cohortMCS") %>% # eliminate string "cohortMCS"
     filter(!grepl(omitcoefs, var2)) # remove coefficients to omit
   
-  pvalvec <- prnpval(as.vector(coefs2$pval))
+  if (outstring==TRUE) pvalvec <- prnpval(as.vector(coefs2$pval))
+  if (outstring==FALSE) pvalvec <- as.vector(coefs2$pval)
   names(pvalvec) <- coefs2$var2
   
   # insert blanks corresponding to coefficient groups
   pvalvec2 <- NULL
-  for (i in 1:length(grouplabs)) pvalvec2 <- c(pvalvec2, "", rep(NA, length(grouplabs[[i]])))
+  if (outstring==TRUE) for (i in 1:length(grouplabs)) pvalvec2 <- c(pvalvec2, "", rep(NA, length(grouplabs[[i]])))
+  if (outstring==FALSE) for (i in 1:length(grouplabs)) pvalvec2 <- c(pvalvec2, 999, rep(NA, length(grouplabs[[i]])))
   pvalvec2[is.na(pvalvec2)] <- pvalvec
+  if (outstring==FALSE) pvalvec2[pvalvec2 == 999] <- NA
   
   return(pvalvec2)
 }
@@ -82,8 +86,8 @@ for (s in c("M", "F")) {
   }
   
   # extract pvalues of interactions
-  r_ext[[paste0(s,"p")]] <- extractp(lm(form_exti,  data=subset(regdata, sex==s)))
-  r_int[[paste0(s,"p")]] <- extractp(lm(form_inti,  data=subset(regdata, sex==s)))
+  r_ext[[paste0(s,"p")]] <- extractp(lm(form_exti,  data=subset(regdata, sex==s)), outstring = TRUE)
[…]

### 2018-09-03 codes/R/paper/paper_jpube.Rnw
current version of tables and figures

@@
 read_chunk(here("codes/R/3_inv_fit.R"))
-read_chunk(here("codes/R/4_scores.R"))
-read_chunk(here("codes/R/4_scores_aux.R"))
-read_chunk(here("codes/R/5_graphs.R"))
-read_chunk(here("codes/R/5_graphs_ineq.R"))
-read_chunk(here("codes/R/5_graphs_item.R"))
-read_chunk(here("codes/R/6_selectvars.R"))
+read_chunk(here("codes/R/4_bootstrap.R"))
+read_chunk(here("codes/R/5_selectvars.R"))
+read_chunk(here("codes/R/6_graphs.R"))
+read_chunk(here("codes/R/6_graphs_ineq.R"))
+read_chunk(here("codes/R/6_graphs_item.R"))
 read_chunk(here("codes/R/8_regs_dep.R"))
@@
 
-<<FA_SCORES_BIND>>=
-# bind scores with data
+<<FA_BOOTSCORES_READ>>=
+# read bootstrap data
+@
+
+<<SELECTDATA, dependson="FA_BOOTSCORES_READ">>=
+# define data, variables and formulas for regression and decomposition
+@
+
+<<SELECTDATA_OUTC, dependson="SELECTDATA">>=
+# load outcomes data
 @
@@
 
-<<SELECTVARS>>=
-# define variables and formulas for regression and decomposition
-@
-
-<<REGS_DEP, dependson="SELECTVARS">>=
+<<REGS_DEP, dependson="SELECTDATA">>=
 # run regressions explaining EXT INT
@@
 
-<<LOAD_DATA_OUTC, dependson="SELECTVARS">>=
-# load outcomes data
-@
-
-<<REGS_OUTC, dependson="LOAD_DATA_OUTC">>=
+<<REGS_OUTC, dependson="SELECTDATA_OUTC">>=
 # run regressions explaining adolescent/adult outcomes
@@
 \begin{figure}
-<<FACDENS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
+<<FACDENS, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
 @
@@
 \begin{figure}
-<<RAWHIST, dependson="FA_SCORES_BIND", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
+<<RAWHIST, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
 @
@@
 \centering\setstretch{1.2}\footnotesize
-<<INEQ_STATS, dependson="FA_SCORES_BIND", results='asis'>>=
+<<INEQ_STATS, dependson="SELECTDATA", results='asis'>>=
 
@@
 \begin{figure}
-<<FACINEQ_FSC, dependson="FA_SCORES_BIND", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
+<<FACINEQ_FSC, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
 fi <- fi_fsc # select result
@@
 \begin{figure}
-<<FACINEQ_PS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
+<<FACINEQ_PS, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
 fi <- fi_ps # select result
@@
 \begin{figure}
-<<FACINEQ_SM, dependson="FA_SCORES_BIND", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
+<<FACINEQ_SM, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
 fi <- fi_sm # select result
@@
 \begin{figure}
-<<ITEMINEQ_PS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=8, fig.width=12, out.width='\\linewidth'>>=
+<<ITEMINEQ_FSC, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=11, out.width='\\linewidth'>>=
+itempl_fsc
+@
+\caption{Item-level changes in inequality -- father occupation}
+\justify {\scriptsize \emph{Notes}: The graph displays the ratio between low and high SES in the prevalence of each behaviour, separately by cohort.
+\par}
+\end{figure}
+
+\begin{figure}
+<<ITEMINEQ_PS, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=11, out.width='\\linewidth'>>=
 itempl_ps
@@
 \begin{figure}
-<<ITEMINEQ_FSC, dependson="FA_SCORES_BIND", dev='tikz', fig.height=8, fig.width=12, out.width='\\linewidth'>>=
-itempl_fsc
-@
-\caption{Item-level changes in inequality -- father occupation}
-\justify {\scriptsize \emph{Notes}: The graph displays the ratio between low and high SES in the prevalence of each behaviour, separately by cohort.
-\par}
-\end{figure}
-
-\begin{figure}
-<<ITEMINEQ_SM, dependson="FA_SCORES_BIND", dev='tikz', fig.height=8, fig.width=12, out.width='\\linewidth'>>=
+<<ITEMINEQ_SM, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=11, out.width='\\linewidth'>>=
 itempl_sm
@@
 robust_p  <- list()
+bootstrap_se  <- list(r_ext$BCS.M[["OLSbootse"]],r_ext$MCS.M[["OLSbootse"]],r_ext$BCS.F[["OLSbootse"]],r_ext$MCS.F[["OLSbootse"]])
 for (i in 1:length(regs)) {
@@
                stars = c(0.1, 0.05, 0.01), digits = 3,
-               override.se = robust_se, override.pvalues = robust_p,
+               override.se = bootstrap_se, override.pvalues = robust_p,
                single.row = FALSE, scriptsize = TRUE,
@@
 robust_p  <- list()
+bootstrap_se  <- list(r_int$BCS.M[["OLSbootse"]],r_int$MCS.M[["OLSbootse"]],r_int$BCS.F[["OLSbootse"]],r_int$MCS.F[["OLSbootse"]])
 for (i in 1:length(regs)) {
@@
 \centering\footnotesize
-\caption{Predictors of adolescent and adult outcomes -- BCS}
-<<REGS_OUTC_BCS, results='asis', dependson="REGS_OUTC">>=
+\caption{Predictors of adolescent outcomes -- BCS}\label{tab:bcs_adol}
+<<REGS_OUTC_BCS1, results='asis', dependson="REGS_OUTC">>=
 
 # collapse to matrix
-bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab[c("smktry16", "bmi16", "hnvq30", "lhgrpay38")],matrix,ncol=9,byrow=FALSE))
+bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab[c("smktry16", "bmi16", "thn16", "owt16")],matrix,ncol=9,byrow=FALSE))
 table <- bcs_outctab_mat
@@
 \centering\footnotesize
-\caption{Predictors of adolescent outcomes -- MCS}
-<<REGS_OUTC_MCS, results='asis', dependson="REGS_OUTC">>=
+\caption{Predictors of adult outcomes -- BCS}\label{tab:bcs_adult}
+<<REGS_OUTC_BCS2, results='asis', dependson="REGS_OUTC">>=
 
 # collapse to matrix
-mcs_outctab_mat <- do.call(rbind,lapply(mcs_outctab[c("smktry14", "bmi14")],matrix,ncol=9,byrow=FALSE))
-table <- mcs_outctab_mat
-printoutctab(table, 2)
+bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab[c("hnvq30", "lhgrpay38", "obs42")],matrix,ncol=9,byrow=FALSE))
+table <- bcs_outctab_mat
+printoutctab(table, 3)
 
@@
 
+\begin{table}
+\centering\footnotesize
+\caption{Predictors of adolescent outcomes -- MCS}\label{tab:mcs_adol}
+<<REGS_OUTC_MCS, results='asis', dependson="REGS_OUTC">>=
+
+# collapse to matrix
+mcs_outctab_mat <- do.call(rbind,lapply(mcs_outctab[c("smktry14", "bmi14", "thn14", "owt14")],matrix,ncol=9,byrow=FALSE))
+table <- mcs_outctab_mat
+printoutctab(table, 4)
+
+@
+\justify {\scriptsize \emph{Notes}: Robust standard errors in parentheses.
+All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
+\par}
+\end{table}
+
+\clearpage
+{\color{red} 
+\textbf{Interpretation of relationship between early-life socioemotional skills and adolescent/adult outcomes} \\
+Remember that a kid with good externalising skills is a kid that is NOT restless/hyperactive and does not have conduct problems (fighting, tantrums).
+(In fact in some factorisations of the Rutter, this factor separates into two separate factors similar to conduct and hyperactivity).
+A kid with good internalising skills is NOT neurotic, worried, fearful, or solitary.\\
+In this light, starting from \autoref{tab:bcs_adol}: better EXT at 5 implies less smoking at 16 (a less `rebellious' child with less conduct problems?), and better INT at 5 implies more smoking at 16 (a more `social' child?).
+A similar thing is observed for BMI and overweight (although in the latter INT is not significant), but notice that this is limited to females -- more neurotic and restless females are larger.
+Almost none of the BCS children are thin/underweight at 16.
+In most of these adolescent outcomes, cognition does not make a difference.\\
+A similar picture emerges for MCS in \autoref{tab:mcs_adol} -- note that here the age for the adolescent outcomes is 14 and not 16.
+But in MCS it seems that internalising skills matter less.
+There is some small indication that more neurotic females are more likely to be thin, but the prevalence is again really low.
+Also consider that all of these are LPM estimates.\\
+For BCS adult outcomes (\autoref{tab:bcs_adult}), cognition dominates.
+But there is some significance for higher education in females: MORE neurotic (perhaps more focused on studies, care about outcomes) and LESS restless females have a higher likelihood of achieving a higher degree.
+}
+
 \end{document}

### 2018-09-20 codes/R/2_efa.R
fix name

@@ -6,18 +6,18 @@
 rotation <- "oblimin"
 
 # EFA with 11 item scale
-items.bcs <- items.c[items.c$cohort=="BCS",c(grep("X[0-9]", names(items.c), value=T))]
+items.bcs <- cohdata[cohdata$cohort=="BCS",c(grep("X[0-9]", names(cohdata), value=T))]
 nbcs <- dim(items.bcs)[1]
 bcscor <- hetcor(items.bcs, ML=TRUE, std.err = F)$correlations
 bcsfa <- fa(r=bcscor, nfactors=2, rotate = rotation, fm="wls")
 #bcsfa3 <- fa(r=bcscor, nfactors=3, rotate = rotation, fm="wls")
 
-items.bcs.m <- items.c[items.c$cohort=="BCS" & items.c$sex=="M",c(grep("X[0-9]", names(items.c), value=T))]
+items.bcs.m <- cohdata[cohdata$cohort=="BCS" & cohdata$sex=="M",c(grep("X[0-9]", names(cohdata), value=T))]
 nbcs.m <- dim(items.bcs.m)[1]
 bcscor.m <- hetcor(items.bcs.m, ML=TRUE, std.err = F)$correlations
 bcsfa.m <- fa(r=bcscor.m, nfactors=2, rotate = rotation, fm="wls")
 
-items.bcs.f <- items.c[items.c$cohort=="BCS" & items.c$sex=="F",c(grep("X[0-9]", names(items.c), value=T))]
+items.bcs.f <- cohdata[cohdata$cohort=="BCS" & cohdata$sex=="F",c(grep("X[0-9]", names(cohdata), value=T))]
 nbcs.f <- dim(items.bcs.f)[1]
 bcscor.f <- hetcor(items.bcs.f, ML=TRUE, std.err = F)$correlations
 bcsfa.f <- fa(r=bcscor.f, nfactors=2, rotate = rotation, fm="wls")
@@ -63,18 +63,18 @@ mr.bcs.int <- fisherz2r(mean(fz.bcs.int, na.rm=T))
 ## ---- MCSEFA
 
 # EFA with 11 item scale
-items.mcs <- items.c[items.c$cohort=="MCS",c(grep("X[0-9]", names(items.c), value=T))]
+items.mcs <- cohdata[cohdata$cohort=="MCS",c(grep("X[0-9]", names(cohdata), value=T))]
 nmcs <- dim(items.mcs)[1]
[…]

### 2018-09-20 codes/R/paper/paper_jpube.Rnw
move text, reorder tables and figures

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\clearpage
-\section{Measurement Invariance}
-
-Meaningful comparison of latent constructs across different groups of individuals and/or measurement occasions (in our case, cohorts) requires that the available measures have the same relationship with the latent factors.
-This property is denominated measurement invariance \citep{Vandenberg2000,Putnick2016}.
-
-Measurement invariance (MI) is a \emph{formally testable} property.
-Two main frameworks have been used for assessing MI: item-response theory (IRT), or structural equation modeling (SEM).
-This paper focuses on the latter, as it is the most commonly employed.\footnote{For an introduction to MI in an IRT framework, see for example \cite{Tay2015}.} In general, MI analyses in a SEM framework are carried out by specifying and estimating factor models of different invariance levels, and comparing their relative fit.
-It starts with the estimation of a minimally restrictive model, where all measurement parameters are allowed to vary across groups -- net of the necessary normalisations -- and proceeds with progressively restricting sets of parameters to be equal across groups.
-If the fit of more restrictive models is comparable to the configural model, different levels of invariance can be established.
-
-In accordance with the SEM approach, we set up a (confirmatory) item factor analysis (IFA) model to examine measurement invariance.
-In our application, we observe 11 categorical items ($X_1$-$X_{11}$).
-Our exploratory analysis above shows that the correlation structure of these measures can be captured by a two-dimensional latent construct (factor) $\bm{\theta}$ -- externalising and internalising socioemotional skills.
-We make the commonly adopted assumption of a dedicated (or congeneric) factor structure, where each measure is assumed to load on only one latent dimension \citep{Attanasio2017c,Heckman2013}.
-
-To account for the ordered nature of the measures, we specify a threshold model \citep{Muthen1984}.
-Individual $j=1\dots N_c$ in group $c=\{A,B\}$ is assumed to have a latent continuous propensity $X^*_{ijc}$ for each item $i=1,\dots,I$.
-Thus, $X^*_{ijc}$ is the child's propensity to exhibit each of the behaviours described in \autoref{tab:comm}, as reported by the mother.
-This propensity is modelled as a function of item- and group-specific intercepts $\nu_{ic}$ and loadings $\bm{\lambda}_{ic}$, and the individual's level on the latent construct vector $\bm{\theta}_{jc}$, plus an independent error component $u_{ijc}$.
-In our application, the latent factor is bidimensional, and corresponds to the externalising ($\theta_{jc}^{EXT}$) and internalising ($\theta_{jc}^{INT}$) skills.
-
-The propensity for each item can be written as follows:\footnote{The dedicated factor structure is ensured by the sparsity of the loading matrix, i.e.: $$\bm{\Lambda}_{c} \coloneqq \begin{bmatrix} \lambda_{1c}, \dots, \lambda_{6c} & \bm{0} \\ \bm{0} & \lambda_{7c}, \dots, \lambda_{11c} \end{bmatrix}.$$}
-\be\label{eq:propens}
-X^*_{ijc} =\nu_{ic} + \bm{\lambda}_{ic}\bm{\theta}_{jc} + u_{ijc} \qquad \text{for} \quad i = 1,\dots,11 \\
-\ee
-
-or more compactly:
-$$\bm{X}^*_{jc} =\bm{\nu}_c + \bm{\Lambda}_{c}\bm{\theta}_{jc} + \bm{u}_{jc}$$
-
-Additional standard distributional assumptions are placed on the latent constructs and error terms:
-$$\bm{\theta}_{jc} \sim N(\bm{\kappa}_c, \bm{\Phi}_c) \qquad \text{and} \qquad \bm{u}_{jc} \sim N(\bm{0}, \bm{\Psi}_c)$$
-
-The model implies the following expression for the mean and covariance structure of the latent propensities:
-$$\bm{\mu}_c=\bm{\nu}_c + \bm{\Lambda}_c\bm{\kappa}_c \qquad \text{and} \qquad \Sigma_c = \bm{\Lambda}_c \bm{\Phi}_c \bm{\Lambda}_c^\prime + \bm{\Psi}_c.$$
-
-Finally, discreteness in the observed measures $X_{cji}$ is incorporated by introducing item- and group-specific threshold parameters $\tau_{ic}$.
-The observed measures as a function of the propensities $X^*$ can be then written as follows:
-$$X_{ijc} = s \qquad \text{if} \; \tau_{s,ic} \leq X^*_{ijc} < \tau_{s+1,ic} \qquad \text{for} \; s=0,1,2$$
-with $\tau_{0,ic}=-\infty$ and $\tau_{3,ic}=+\infty$.
-
-% FOOTNOTE\footnote{As a consequence of the normality assumption on the error terms $\veps$, the probability that individual $j$ in group $c$ exhibits item $i$ with ``intensity" $s$ is:
-% \begin{align*}
-% \Pr(X_{ijc}=0) &= 1- \Phi(-\tau_{1,ic} + \nu_{ic} + \lambda_{ic}\bm{\theta}_{jc}) \\
-% \Pr(X_{ijc}=1) &= \Phi(-\tau_{1,ic} + \nu_{ic} + \lambda_{ic}\bm{\theta}_{jc})
-% \end{align*}
-% for binary items and
-% \begin{align*}
-% \Pr(X_{ijc}=0) &= 1- \Phi(-\tau_{1,ic} + \nu_{ic} + \lambda_{ic}\bm{\theta}_{jc}) \\
-% \Pr(X_{ijc}=1) &= \Phi(-\tau_{1,ic} + \nu_{ic} + \lambda_{ic}\bm{\theta}_{jc}) - \Phi(-\tau_{2,ic} + \nu_{ic} + \lambda_{ic}\bm{\theta}_{jc})\\
-% \Pr(X_{ijc}=2) &= \Phi(-\tau_{2,ic} +  \nu_{ic} + \lambda_{ic}\bm{\theta}_{jc})
-% \end{align*}
-% for ordinal items. This IFA model can also be written as a 2-parameter Logit IRT model with item difficulty $b$ and item discrimination $a$. In the simplest case for a binary item $1$:
-% $$\text{Probit}(X_{1jc}=1) = a_{1c}(\bm{\theta}_{jc} - b_{1c})$$
-% such that $\tau_{1c}=a_{1c}b_{1c}$ and $a_{1c}=\lambda_{1c}$.}
-
-% Binary items take values:
-% \begin{equation}
-% X_{cji}=
-% \begin{cases}
-% \qquad 0 \qquad \text{if} \; X^*_{cji}<\tau_{1,ci} \\
-% \qquad 1 \qquad \text{if} \; X^*_{cji}\geq\tau_{1,ci}
-% \end{cases}
-% \end{equation}
-% 
-% An ordinal item takes values:
-% 
-% \begin{equation}
-% X_{cji}=
-% \begin{cases}
-% \qquad 0 \qquad \text{if} \; X^*_{cji}<\tau_{1,ci} \\
-% \qquad 1 \qquad \text{if} \; \tau_{1,ci} \leq X^*_{cji}<\tau_{2,ci} \\
-% \qquad 2 \qquad \text{if} \; X^*_{cji}\geq\tau_{2,ci}
-% \end{cases}
-% \end{equation}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-The model specification above cannot identify the parameters unless additional normalisations are imposed.
-A configural model is defined as a minimal set of restrictions that ensure identification of parameters across groups.
-There are infinite, equivalent ways in which the configural model can be parameterised -- the well-known issue of factor indeterminacy.
-Widely used parameterisations for the configural model are:
-\bi[label=$\diamond$]
-\item \underline{Delta parameterisation} {[WE{\textDelta}]} \citep{Wu2016a}
-
-For all groups: $$\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}, \quad \bm{\nu}=\bm{0}, \quad \text{and} \; \text{diag}(\bm{\Sigma})=\bm{I}.$$
-
-\item \underline{Theta parameterisation} {[WE{\textTheta}]} \citep{Wu2016a}
-
-For all groups: $$\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}, \quad \bm{\nu}=\bm{0}, \quad \text{and} \; \text{diag}(\bm{\Psi})=\bm{I}.$$
-
-\item \underline{Anchored parameterisation} {[MT]} \citep{Millsap2004}
-  \bi
-  \item For all groups, normalise a reference loading to 1 for each factor
-  \item Set invariant across groups one threshold per item (e.g. $\tau_{0,Ai} = \tau_{0,Bi}$), and an additional threshold in the reference items above
-  \item In the first group: $\bm{\kappa_A}=\bm{0}$, $\text{diag}(\bm{\Sigma}_A)=\bm{I}$
-  \item Set all intercepts $\bm{\nu}$ to zero
-  \ei
-\ei
-The first two parameterisations (WE{\textDelta} and WE{\textTheta}) normalise the mean and variance of factors to the same constants in both groups, and they leave all loadings and intercepts to be freely estimated; they only differ in whether the additional required normalisation is imposed on the variances of the error terms ($\bm{\Psi}$) or on the diagonal of the covariance matrix of the measures ($\bm{\Sigma}$).
-The MT parameterisation instead proceeds by identifying parameters in one group first, and then imposing cross-group equality constraints to identify parameters in other groups \citep{Wu2016a}.
-
-Once the configural model has been estimated, measurement invariance analysis proceeds by estimating nested models that place increasing restrictions on the item parameters, by constraining them to be equal across groups $c$.
-In the case where available measures are continuous, MI analysis is straightforward \citep{vandeSchoot2012}: one set of parameters is restricted at a time, and the fit of the resulting models are compared to the configural model.
-The hierarchy of the nested models usually proceeds by testing loadings first, and then intercepts (to establish \emph{metric} and \emph{scalar} invariance -- see \citealp{Vandenberg2000}).
-
-Invariance of systems with categorical measures is less well understood.
-In particular, the lack of explicit location and scale in the measures introduces an additional set of parameters (thresholds $\tau$) and makes identification reliant on more stringent normalisations.
-The first comprehensive approach for categorical measures is proposed by \cite{Millsap2004}.
-New identification results in \cite{Wu2016a} indicate that, in the categorical case, invariance properties cannot be examined by simply restricting one set of parameters at a time.
-This is because the identification conditions used in the configural baseline model, while being minimally restrictive on their own, become binding once certain additional restrictions are imposed.
-In light of this, they propose models that identify structures of different invariance levels.
-They find that some restrictions cannot be tested alone against the configural model, because the models they generate are statistically equivalent.
-This is true of loading invariance, and also of threshold invariance in the case when the number of categories of each ordinal item is 3 or less.
-Furthermore, they suggest that comparison of both latent means and variances requires invariance in loadings, thresholds, and intercepts.
-
-{\color{red} ADD SOME MORE COMMENTS ON IDENTIFICATION HERE, to make everything more palatable to economics audience.}
-
-We estimate the sequence of models detailed in \autoref{tab:invparam} by Weighted Least Squares.\footnote{Parameters are estimated by mean- and variance-adjusted weighted least squares (WLSMV) (see \citealp{Muthen2017a}); estimation starts from the items' polychoric correlation matrix, uses diagonally weighted least squares (DWLS), and exploits the full weight matrix to compute robust standard errors and test statistics.
-Robust WLS has proved in simulation studies to be moderately robust to small violations of the normality assumption in the latent underlying measures \citep{Flora2004}, and generally outperforms maximum likelihood in large samples \citep{Beauducel2006,Li2016}.
-All estimates are computed using the \texttt{lavaan} package in \textbf{\textsf{R}} \citep{Rosseel2012}.}
-For the purposes of the analysis, we define groups $c$ as cohort-gender cells, with the reference group being males in the BCS (1970) cohort.
-We then compare the fit of each model against the configural model (i.e.
-the least restrictive).
-According to recommended practice, we diplay a range of fit indices.
-Comparison of $\chi^2$ values across models is the most common strategy.
-However, tests based on $\Delta \chi^2$ are known to display increasing Type I error rates with sample size and model complexity \citep{Sass2014a}.
-The use of approximate fit indices (AFIs) has been proposed alongside $\chi^2$ to overcome these limitation.
-These indices do not have a known sampling distribution, thus making it necessary to rely on rules of thumb to assess what level of \textDelta AFI indicates invariance.
-Nevertheless, AFIs provide valuable corrections for model parsimony.\footnote{Commonly adopted AFIs in studies of measurement invariance are the root mean squared error of approximation (RMSEA) and the Tucker-Lewis index (TLI).
-Simulation evidence by \cite{Cheung2002} shows that these indices can show correlation between overall and relative fit, and suggest relying on additional indices, such as the comparative fit index (CFI, \citealp{Bentler1990}), McDonald non-centrality index (MFI, \citealp{McDonald1989}), and Gamma-hat index \citep{Steiger1989} for the case of ordered measures.
-Commonly accepted thresholds for rejection are $\Delta CFI < - 0.01$, $\Delta MFI < - 0.02$, and $\Delta Gamma hat < - 0.001$.
-\cite{Meade2008}, using the results from a simulation study, suggests stricter thresholds that should apply in a variety of conditions.
-For CFI, a single cutoff value of $.002$ is proposed, while cutoffs for MFI depend on the problem's characteristics; in our case (2 factors, 11 items), they suggest $.0066$.
-\cite{Sass2014a} however cast some doubts of the generalisability of these cutoffs to WLSMV estimators.}
-
-{\color{red} Motivation for choosing RMSEA (does not seem supported in literature, see above).}
-
-The fit of each estimated model using our full sample is compared in Panel A of \autoref{tab:fit}.\footnote{We do not present fit results for the threshold invariance model, as it is statistically equivalent to the configural model and thus its fit is mathematically the same.} The overall fit of the configural model is satisfactory according to widely accepted rules of thumb, with RMSEA close to $0.05$ and CFI close to $.95$ \citep{Hu1999}.
-For all invariance levels, a chi-squared difference would point to a lack of measurement invariance.
-However, as detailed above, this test is expected to lead to substantial Type I error with a relatively large sample size such as ours.
-For the model with restricted thresholds and loadings, most of the \textDelta AFIs indicate that this level of invariance is justified in our data.
-However, further restricting intercepts results in a model where invariance is rejected across the board.
 
@@
 
-\begin{figure}
-<<FACDENS, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
-@
-\caption{Scores}
-\justify {\scriptsize \emph{Notes}: Density of scores from final model.
-\par}
-\end{figure}
-
-\begin{figure}
-<<RAWHIST, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
-@
-\caption{Sum scores}
-\justify {\scriptsize \emph{Notes}: Density of sum scores without factor analysis.
-\par}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \begin{table}
@@
 \end{table}
-
-\begin{figure}
-<<FACINEQ_FSC, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
-fi <- fi_fsc # select result
-gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
-                        fi$ineqlist[[3]],fi$ineqlist[[4]],
-                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
-                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
-@
-\caption{Inequality by father occupation -- normalised}
-\end{figure}
-
-\begin{figure}
-<<FACINEQ_PS, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
-fi <- fi_ps # select result
-gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
-                        fi$ineqlist[[3]],fi$ineqlist[[4]],
-                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
-                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
-@
-\caption{Inequality by mother education -- normalised}
-\end{figure}
-
-\begin{figure}
-<<FACINEQ_SM, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
-fi <- fi_sm # select result
-gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
-                        fi$ineqlist[[3]],fi$ineqlist[[4]],
-                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
-                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
-@
-\caption{Inequality by mother smoking in pregnancy -- normalised}
-\end{figure}
-
-\begin{figure}
-<<ITEMINEQ_FSC, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=11, out.width='\\linewidth'>>=
-itempl_fsc
-@
-\caption{Item-level changes in inequality -- father occupation}
-\justify {\scriptsize \emph{Notes}: The graph displays the ratio between low and high SES in the prevalence of each behaviour, separately by cohort.
-\par}
-\end{figure}
-
-\begin{figure}
-<<ITEMINEQ_PS, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=11, out.width='\\linewidth'>>=
-itempl_ps
-@
-\caption{Item-level changes in inequality -- mother education}
-\justify {\scriptsize \emph{Notes}: The graph displays the ratio between low and high SES in the prevalence of each behaviour, separately by cohort.
-\par}
-\end{figure}
-
-\begin{figure}
-<<ITEMINEQ_SM, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=11, out.width='\\linewidth'>>=
-itempl_sm
-@
-\caption{Item-level changes in inequality -- mother smoking in pregnancy}
-\justify {\scriptsize \emph{Notes}: The graph displays the ratio between low and high SES in the prevalence of each behaviour, separately by cohort.
-\par}
-\end{figure}
 
@@
 
-\clearpage
-{\color{red} 
-\textbf{Interpretation of relationship between early-life socioemotional skills and adolescent/adult outcomes} \\
-Remember that a kid with good externalising skills is a kid that is NOT restless/hyperactive and does not have conduct problems (fighting, tantrums).
-(In fact in some factorisations of the Rutter, this factor separates into two separate factors similar to conduct and hyperactivity).
-A kid with good internalising skills is NOT neurotic, worried, fearful, or solitary.\\
-In this light, starting from \autoref{tab:bcs_adol}: better EXT at 5 implies less smoking at 16 (a less `rebellious' child with less conduct problems?), and better INT at 5 implies more smoking at 16 (a more `social' child?).
-A similar thing is observed for BMI and overweight (although in the latter INT is not significant), but notice that this is limited to females -- more neurotic and restless females are larger.
-Almost none of the BCS children are thin/underweight at 16.
-In most of these adolescent outcomes, cognition does not make a difference.\\
-A similar picture emerges for MCS in \autoref{tab:mcs_adol} -- note that here the age for the adolescent outcomes is 14 and not 16.
-But in MCS it seems that internalising skills matter less.
-There is some small indication that more neurotic females are more likely to be thin, but the prevalence is again really low.
-Also consider that all of these are LPM estimates.\\
-For BCS adult outcomes (\autoref{tab:bcs_adult}), cognition dominates.
-But there is some significance for higher education in females: MORE neurotic (perhaps more focused on studies, care about outcomes) and LESS restless females have a higher likelihood of achieving a higher degree.
-}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}
+<<FACDENS, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
+@
+\caption{Scores}
+\justify {\scriptsize \emph{Notes}: Density of scores from final model.
+\par}
+\end{figure}
+
+\begin{figure}
+<<RAWHIST, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
+@
+\caption{Sum scores}
+\justify {\scriptsize \emph{Notes}: Density of sum scores without factor analysis.
+\par}
+\end{figure}
+
+\begin{figure}
+<<FACINEQ_FSC, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
+fi <- fi_fsc # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\caption{Inequality by father occupation -- normalised}
+\end{figure}
+
+\begin{figure}
+<<FACINEQ_PS, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
+fi <- fi_ps # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\caption{Inequality by mother education -- normalised}
+\end{figure}
+
+\begin{figure}
+<<FACINEQ_SM, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
+fi <- fi_sm # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\caption{Inequality by mother smoking in pregnancy -- normalised}
+\end{figure}
+
+\begin{figure}
+<<ITEMINEQ_FSC, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=11, out.width='\\linewidth'>>=
+itempl_fsc
+@
+\caption{Item-level changes in inequality -- father occupation}
+\justify {\scriptsize \emph{Notes}: The graph displays the ratio between low and high SES in the prevalence of each behaviour, separately by cohort.
+\par}
+\end{figure}
+
+\begin{figure}
+<<ITEMINEQ_PS, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=11, out.width='\\linewidth'>>=
+itempl_ps
+@
+\caption{Item-level changes in inequality -- mother education}
+\justify {\scriptsize \emph{Notes}: The graph displays the ratio between low and high SES in the prevalence of each behaviour, separately by cohort.
+\par}
+\end{figure}
+
+\begin{figure}
+<<ITEMINEQ_SM, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=11, out.width='\\linewidth'>>=
+itempl_sm
+@
+\caption{Item-level changes in inequality -- mother smoking in pregnancy}
+\justify {\scriptsize \emph{Notes}: The graph displays the ratio between low and high SES in the prevalence of each behaviour, separately by cohort.
+\par}
+\end{figure}
 

### 2018-09-22 codes/R/paper/paper_jpube.Rnw
revise EFA, add loadings table

@@
 \usepackage{authblk}
-\title{Inequality in non-cognitive skills: a cross-cohort comparison}
+\title{Inequality in socioemotional skills: a cross-cohort comparison}
 \author[1,2]{Orazio Attanasio}
@@
 \maketitle
+
+\begin{abstract}
+abstract here
+\end{abstract}
+
+% We are grateful to The Centre for Longitudinal Studies, UCL Institute of Education for the use of these data and to the UK Data Archive and UK Data Service for making them available.
+However, they bear no responsibility for the analysis or interpretation of these data.
 
@@
 <<AUXFUN>>=
-# load function that gets parameters
+# load auxiliary functions
 @
@@
 
-<<MEANTABLE_PRINT>>=
-# function to print mean table
-printmeantab <- function(tab) {
-  print(xtable(tab,
-               align="L{0cm}C{.04\\textwidth}C{.05\\textwidth}C{.05\\textwidth}L{.35\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}L{.35\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
-               digits = 1
-  ), 
-  size="\\small",
-  include.rownames=F,
-  include.colnames=F,
-  booktabs = T,
-  add.to.row = list(pos = c(list(0)),
-                    command = c(
-                      # header
-                      paste(" & & &
-                            \\multicolumn{4}{c}{\\textbf{BCS}} &
-                            \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){4-7} \\cmidrule(lr){8-11} 
-                            Item & Factor & Num. Categ. &
-                            Wording & Cert. Appl. (\\%) & Smtm. Appl. (\\%) & Appl. (\\%) &
-                            Wording & Cert. True (\\%) & Smwt. True (\\%) & True (\\%) \\\\ \\midrule
-                            "
-                            , sep="")
-                    )),
-  sanitize.text.function = function(x){x},
-  sanitize.rownames.function = function(x){""}, # to not print row numbers
-  tabular.environment = "tabular", width = ".8\\textwidth",
-  floating = FALSE
-  )
-}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+<<BCSEFA, dependson='LOAD_DATA'>>=
+@
+
+<<BCSEFARES, dependson='BCSEFA'>>=
+@
+
+<<MCSEFA, dependson='LOAD_DATA'>>=
+@
+
+<<MCSEFARES, dependson='MCSEFA'>>=
 @
@@
 
+\clearpage
+\bibliographystyle{chicago}
+\bibliography{/Users/giacomomason/Dropbox/Zotero/cohorts.bib}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\clearpage
 \begin{table}[]
@@
 
-% ADD PARTIAL INVARIANCE
-
-\clearpage
-\bibliographystyle{chicago}
-\bibliography{/Users/giacomomason/Dropbox/Zotero/cohorts.bib}
-
-\clearpage
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{table}[]
-\caption{Item wording and prevalence, by cohort -- females}\label{tab:meantab_f}
+\caption{Item prevalence, by cohort and gender}\label{tab:meantab}
 \centering\setstretch{1.2} \footnotesize
-<<MEANTAB_F, dependson="MEANTABLE", results='asis'>>=
+<<MEANTAB, dependson="MEANTABLE", results='asis'>>=
 require(xtable)
-table <- meantab[[2]] # females
-table <- table[,!names(table) %in% c("bcs.num", "mcs.num")] # drop if unneeded
-printmeantab(table)
+table <- meantab # females
+table <- table[,!names(table) %in% c("bcs.text", "bcs.num", "mcs.text", "mcs.num")] # drop if unneeded
+
+  print(xtable(table,
+               align="L{0cm}C{.04\\textwidth}C{.05\\textwidth}C{.05\\textwidth}L{.12\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
+               digits = 1
+  ), 
+  include.rownames=F,
+  include.colnames=F,
+  booktabs = T,
+  add.to.row = list(pos = c(list(0)),
+                    command = c(
+                      # header
+                      paste(" & & & &
+                            \\multicolumn{6}{c}{\\textbf{Males}} & \\multicolumn{6}{c}{\\textbf{Females}}
+                            \\\\ \\cmidrule(lr){5-10} \\cmidrule(lr){11-16} 
+                            & & & & 
+                            \\multicolumn{3}{c}{\\textbf{BCS}} & \\multicolumn{3}{c}{\\textbf{MCS}}
+                            & \\multicolumn{3}{c}{\\textbf{BCS}} & \\multicolumn{3}{c}{\\textbf{MCS}}
+                            \\\\ \\cmidrule(lr){5-7} \\cmidrule(lr){8-10} \\cmidrule(lr){11-13} \\cmidrule(lr){14-16} 
+                            & & & &
+                            Cert. Appl. & Smtm. Appl. & Appl. &  Cert. True & Smwt. True & True &
+                            Cert. Appl. & Smtm. Appl. & Appl. &  Cert. True & Smwt. True & True 
+                            \\\\
+                            Itm. & Factor & Cat. & Title & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) \\\\
+                            "
+                            , sep="")
+                    )),
+  sanitize.text.function = function(x){x},
+  sanitize.rownames.function = function(x){""}, # to not print row numbers
+  tabular.environment = "tabular",
+  floating = FALSE
+  )
+
 @
+\justify {\scriptsize \emph{Notes}: The table shows the prevalence by gender and cohort for each item of our novel subscale.
+\emph{Itm.} is item number.
+\emph{Factor} is the latent construct to which the item loads -- EXT is Externalising skills, INT is Internalising skills.
+\emph{Cat.} is the number of categories in which the item is coded -- 2 denotes a binary item (applies/does not apply) and 3 denotes a 3-category item.
+\emph{Title} is a short label for the item.
+\emph{Cert.
+/ Smtm.
+Appl.} =  Certainly / sometimes applies.
+\emph{Cert.
+/ Smwt.
+True} = Certainly / somewhat true.
+\par}
 \end{table}
@@
 
-\begin{landscape}
-\begin{table}[]
-\caption{Item wording and prevalence, by cohort -- males}\label{tab:meantab_m}
-\centering\setstretch{1.2} \footnotesize
-<<MEANTAB_M, dependson="MEANTABLE", results='asis'>>=
-require(xtable)
-table <- meantab[[1]] # males
-table <- table[,!names(table) %in% c("bcs.num", "mcs.num")] # drop if unneeded
-printmeantab(table)
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\caption{Suggested number of factors to retain}\label{tab:numfac}
+\centering\setstretch{1.2}\footnotesize
+<<NUMFACTAB, dependson='MCSEFARES', results='asis'>>=
+# combine results
+efatab <- data.frame(cbind(efatab.bcs,efatab.mcs))
+
+efatab %>%
+  tibble::rownames_to_column() %>%
+  kable("latex", escape = F, booktabs = T, linesep = "", align=c('l','c','c','c','c','c','c'),
+        col.names = c("Approach", "All", "Males", "Females", "All", "Males", "Females")) %>%
+  add_header_above(c(" " = 1,  "BCS (1970)" = 3, "MCS (2000/1)" = 3)) %>%
+  column_spec(2:7, width = "3em")
+
 @
+\justify {\scriptsize \emph{Notes}: The table compares the optimal number of factors suggested by different aprroaches for our novel scale; scree test based approaches (optimal coordinates, acceleration factor -- \citealp{Raiche2013}), parallel analysis \citep{Horn1965}, Kaiser's eigenvalue rule \citep{Kaiser1960}, Very Simple Structure (VSS, \citealp{Revelle1979}), Velicer Minimum Average Partial test (MAP, \citealp{Velicer1976}).
+\par}
 \end{table}
-\end{landscape}
+
+\begin{table}
+\caption{Loadings from exploratory factor analysis}\label{tab:load}
+\centering\setstretch{1.2}\footnotesize
+<<EFALOADTAB, dependson='MCSEFARES', results='asis'>>=
+# combine results
+library(kableExtra)
+loadtab <- cbind(meantab$num, meantab$title, data.frame(cbind(bcsfa$loadings, mcsfa$loadings)))
+colnames(loadtab) <- c("item", "title", "bcs1", "bcs2", "mcs1", "mcs2")
+
+loadtab %>%
+  mutate_at(vars(matches("cs")), round, 3) %>%
+  mutate(
+    bcs1 = cell_spec(bcs1, "latex", bold = ifelse(abs(bcs1) > .25, TRUE, FALSE)),
+    bcs2 = cell_spec(bcs2, "latex", bold = ifelse(abs(bcs2) > .25, TRUE, FALSE)),
+    mcs1 = cell_spec(mcs1, "latex", bold = ifelse(abs(mcs1) > .25, TRUE, FALSE)),
+    mcs2 = cell_spec(mcs2, "latex", bold = ifelse(abs(mcs2) > .25, TRUE, FALSE))
+  ) %>%
+  kable("latex", escape = F, booktabs = T, linesep = "", align=c('c','l','c','c','c'),
+        col.names = c("Item", "Title", "Factor 1 (EXT)", "Factor 2 (INT)", "Factor 1 (EXT)", "Factor 2 (INT)")) %>%
+  add_header_above(c(" " = 2,  "BCS (1970)" = 2, "MCS (2000/1)" = 2))
+
+@
+\justify {\scriptsize \emph{Notes}: The table displays the factor loadings obtained from exploratory factor analysis (EFA) on our novel scale, separately by cohort.
+The EFA is based on the decomposition of the polychoric correlation matrix, and uses weighted least squares and oblimin rotation.
+\par}
+\end{table}
 

### 2018-09-22 codes/R/1_load.R

@@ -282,16 +282,15 @@ for (i in 1:length(fadata)) fadata[[i]] <- fadata[[i]][order(fadata[[i]][,"cohor
 ncomm <- 11                           # number of common items
 ncats <- c(3,3,3,3,2,2,3,3,3,3,2)     # number of categories per item
 
+# calculate means by gender
 means <- list()
 X.allg <- list()
-meantab <- list()
-means[[1]] <- matrix(NA, ncomm, 7)    # males
-means[[2]] <- matrix(NA, ncomm, 7)    # females
-X.allg[[1]] <- X.all[X.all$sex==1,]
-X.allg[[2]] <- X.all[X.all$sex==2,]
-
-for (g in c(1,2)) {                   # males/females
-  colnames(means[[g]]) <- c("num", "BCS_ca", "BCS_sa", "BCS_a", "MCS_ca", "MCS_sa", "MCS_a")
+means[["M"]] <- matrix(NA, ncomm, 7)    # males
+means[["F"]] <- matrix(NA, ncomm, 7)    # females
+X.allg[["M"]] <- X.all[X.all$sex==1,]
+X.allg[["F"]] <- X.all[X.all$sex==2,]
+for (g in c("M","F")) {                   # males/females
+  colnames(means[[g]]) <- c("num", paste0(c("BCS_ca", "BCS_sa", "BCS_a", "MCS_ca", "MCS_sa", "MCS_a"), g))
   
   for (i in 1:ncomm) {                # items
     means[[g]][i,1] <- i
@@ -309,67 +308,75 @@ for (g in c(1,2)) {                   # males/females
       means[[g]][i,7] <- t2[2,1]
     }
   }
-  
-  # table with means
-  meantab[[g]]      <- data.frame(row.names = paste("Item", seq(1,ncomm)))
-  meantab[[g]]$num  <- seq(1,ncomm)     # item number
-  meantab[[g]]$fac  <- c(rep("EXT", 6), rep("INT", 5))                                            # factor
-  meantab[[g]]$cats <- as.character(ncats)                                                       # number of categories
-  
-  meantab[[g]]$bcs.text <- c(
-    "Very restless. Often running about or jumping up and down. Hardly ever still",
-    "Is squirmy or fidgety",
-    "Frequently fights other children + \\newline Bullies other children",
[…]

### 2018-09-22 codes/R/2_efa.R

@@ -10,7 +10,6 @@ items.bcs <- cohdata[cohdata$cohort=="BCS",c(grep("X[0-9]", names(cohdata), valu
 nbcs <- dim(items.bcs)[1]
 bcscor <- hetcor(items.bcs, ML=TRUE, std.err = F)$correlations
 bcsfa <- fa(r=bcscor, nfactors=2, rotate = rotation, fm="wls")
-#bcsfa3 <- fa(r=bcscor, nfactors=3, rotate = rotation, fm="wls")
 
 items.bcs.m <- cohdata[cohdata$cohort=="BCS" & cohdata$sex=="M",c(grep("X[0-9]", names(cohdata), value=T))]
 nbcs.m <- dim(items.bcs.m)[1]
@@ -43,14 +42,6 @@ efatab.bcs <- cbind(
 rownames(efatab.bcs) <- c("Optimal Coordinates", "Acceleration Factor", "Parallel Analysis", "Kaiser", 
                           "VSS Compl. 1", "VSS Compl. 2", "Velicer MAP")
 
-# FACTOR ANALYSIS loadings
-cat("\n\n ----------------------- \n BCS Loadings - Males and Females")
-bcsfa$loadings
-cat("\n\n ----------------------- \n BCS Loadings - Males")
-bcsfa.m$loadings
-cat("\n\n ----------------------- \n BCS Loadings - Females")
-bcsfa.f$loadings
-
 ## ---- BCSREL
 # reliability 
 fz.bcs <- fisherz(bcscor)
@@ -67,7 +58,6 @@ items.mcs <- cohdata[cohdata$cohort=="MCS",c(grep("X[0-9]", names(cohdata), valu
 nmcs <- dim(items.mcs)[1]
 mcscor <- hetcor(items.mcs, ML=TRUE, std.err = F)$correlations
 mcsfa <- fa(r=mcscor, nfactors=2, rotate = rotation, fm="wls")
-#mcsfa3 <- fa(r=mcscor, nfactors=3, rotate = rotation, fm="wls")
 
 items.mcs.m <- cohdata[cohdata$cohort=="MCS" & cohdata$sex=="M",c(grep("X[0-9]", names(cohdata), value=T))]
 nmcs.m <- dim(items.mcs.m)[1]
@@ -100,14 +90,6 @@ efatab.mcs <- cbind(
[…]

### 2018-09-26 codes/R/paper/paper_jpube.Rnw
add notes, minor changes to figures, revise tables

@@
 
-% appendices/numeration/references
+% CAPTIONS
+\usepackage[textfont={bf,normalsize},labelfont={bf,normalsize}]{caption}     % to allow multiple lines in captions, and customize the look
+\captionsetup{justification=raggedright,singlelinecheck=false}
+
+% APPENDICES AND SECTION NUMERATION
 \usepackage[nottoc,numbib]{tocbibind} 		% bibliography in TOC
-\usepackage[toc,page]{appendix} %appendix
+\usepackage[toc,page,title]{appendix}       % appendix ('title' makes sections into 'APPENDIX A')
 \newcommand{\tocless}[2]{\bgroup\let\addcontentsline=\nocontentsline#1{#2}\egroup} % for empty section
 \numberwithin{equation}{section} %for section eq numeration
+
+% CROSSREFERENCING
 \usepackage{hyperref}                            % references
@@
 \usepackage{cleveref}
-
 %tikz
@@
 \setlist[itemize]{labelsep=*, leftmargin=*}
-\usepackage[textfont={small},labelfont={bf,small},justification=centering]{caption}     % to allow multiple lines in captions, and customize the look
 
@@
 Panel B is restricted to a subsample of children in the age range of maximum overlap between the two cohorts (59-61 months).
+Panel C and D are limited to the samples of male and female children respectively.
 \par}
@@
 \centering\setstretch{1.2}\footnotesize
+\flushleft\textbf{Panel A: Measurement parameters}
 <<RES_FA_PAR, dependson="FA_FIT", results='asis'>>=
 
-table <- getmeaspars(fa.tlmt[[1]], groups = 4, mode = "tl")
+table <- getmeaspars(finalmod, groups = 4, mode = "tl")
 
@@
 print.xtable(xtable(table,
-              align="C{0cm}C{.06\\textwidth}C{.04\\textwidth}C{.07\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}C{.057\\textwidth}",
+              align="C{0cm}C{.08\\textwidth}C{.09\\textwidth}C{.07\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
               digits = 3
@@
                                             & \\multicolumn{2}{c}{\\textbf{Thresholds}}
-                                            & \\multicolumn{3}{c}{\\textbf{Intercepts}}
-                                            & \\multicolumn{3}{c}{\\textbf{Variances}}
+                                            & \\multicolumn{3}{c}{\\textbf{Intercepts} (BCS M = 0)}
+                                            & \\multicolumn{3}{c}{\\textbf{Variances} (BCS M = 1)}
                                             \\\\ \\cmidrule(lr){3-3} \\cmidrule(lr){4-5} \\cmidrule(lr){6-8} \\cmidrule(lr){9-11}
@@
 
-                                            Measure
+                                            Item
                                             & Factor
@@
 @
-\end{table}
-
-\begin{table}
-\caption{Estimated latent variable parameters -- Threshold and loading invariance model (anchored)}\label{tab:lvpars}
-\centering\setstretch{1.2}\footnotesize
+\vspace{3pt}
+\flushleft\textbf{Panel B: Latent variable parameters}
 <<RES_FA_LV, dependson="FA_FIT", results='asis'>>=
@@
 
-lvtab <- getlvpars(fa.tlmt[[1]], groups = 4)
+lvtab <- getlvpars(finalmod, groups = 4)
 table <- rbind(lvtab$Males,lvtab$Females)
+table <- table[,c("measure", "mean_BCS", "mean_MCS", 
+                  "covext_BCS", "covint_BCS", "covext_MCS", "covint_MCS", 
+                  "corr_BCS", "corr_MCS")] # rearrange
 
 print.xtable(xtable(table,
-                     align="C{0cm}L{.13\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
+                     align="C{0cm}L{.13\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}",
                     digits = 3
@@
                       "\\toprule
-                      & \\multicolumn{4}{c}{\\textbf{BCS}}
-                      & \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9}
+                      & \\multicolumn{2}{c}{\\textbf{Mean}}
+                      & \\multicolumn{4}{c}{\\textbf{Covariance}}
+                      & \\multicolumn{2}{c}{Correlation} \\\\
+                      \\cmidrule(lr){2-3} \\cmidrule(lr){4-7} \\cmidrule(lr){8-9}
 
-                      & Mean
-                      & \\multicolumn{2}{c}{Covariance}
-                      & \\multicolumn{1}{c}{Correlation}
-                      & Mean
-                      & \\multicolumn{2}{c}{Covariance}
-                      & \\multicolumn{1}{c}{Correlation} \\\\
+                      & \\multicolumn{2}{c}{$\\bm{\\kappa}$} & \\multicolumn{4}{c}{$\\bm{\\Phi}$} & & \\\\
+
+                      & BCS & MCS & \\multicolumn{2}{c}{BCS} & \\multicolumn{2}{c}{MCS} & BCS & MCS \\\\
 
@@
 @
+\justify {\scriptsize \emph{Notes}: The table presents estimates for the factor model with loadings $\lambda$ and thresholds $\tau$ restricted to be equal across cohorts.
+Panel A shows estimates of the measurement parameters.
+Loadings and thresholds are the same across all cohorts.
+Intercepts are restricted to zero in the reference group, i.e.
+males in BCS (not shown).
+Variances of the error terms are restricted to one in the reference group, i.e.
+males in BCS (not shown), and for the items that only have two categories (5, 6, 11).
+Panel B shows estimates of the latent variable parameters.
+Means are restricted to zero in all cohort-gender groups, while variances are restricted to one only in the reference group, i.e.
+males in BCS.
+\par}
 \end{table}
@@
                custom.coef.names = varlabs,
-               custom.model.names = c("Males BCS", "Males MCS", "Females BCS", "Females MCS"),
+               custom.model.names = c("\\multicolumn{3}{c}{\\textbf{Males}} & \\multicolumn{3}{c}{\\textbf{Females}} \\\\ 
+                                        \\cmidrule(lr){2-4} \\cmidrule(lr){5-7} 
+                                      & (1) & (2) & (3) & (4) & (5) & (6) \\\\
+                                      & BCS", "MCS", "BCS", "MCS"),
                custom.note = "",
@@
 @
-\justify {\scriptsize \emph{Notes}: Robust standard errors in parentheses.
+\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of children's externalising socioemotional skills at five years of age on family background information.
+The dependent variable is a factor score obtained from the factor model in \autoref{sec:methods}.
+Col.
+(1) and (2) show coefficients and standard errors in parentheses, for male children in the BCS and MCS cohorts separately.
+The latter are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression.
+Col.
+(3) shows the p-value of a test that the coefficient is the same in the two cohorts.
+Col.
+(4) to (6) repeat for female children.
 All estimates additionally control for region of birth, mother height, number of previous stillbirths at child's birth, preterm birth, a dummy for missing gestational age, and number of other children in the household at child age 5.
+See \autoref{tab:harmvar} for a description of the variables used.
 \par}
@@
 @
-\justify {\scriptsize \emph{Notes}: Robust standard errors in parentheses.
+\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of children's internalising socioemotional skills at five years of age on family background information.
+The dependent variable is a factor score obtained from the factor model in \autoref{sec:methods}.
+Col.
+(1) and (2) show coefficients and standard errors in parentheses, for male children in the BCS and MCS cohorts separately.
+The latter are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression.
+Col.
+(3) shows the p-value of a test that the coefficient is the same in the two cohorts.
+Col.
+(4) to (6) repeat for female children.
 All estimates additionally control for region of birth, mother height, number of previous stillbirths at child's birth, preterm birth, a dummy for missing gestational age, and number of other children in the household at child age 5.
+See \autoref{tab:harmvar} for a description of the variables used.
 \par}
@@
 @
-\justify {\scriptsize \emph{Notes}: Robust standard errors in parentheses.
+\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of cohort members' adolescent and adult outcomes on their externalising and internalising socioemotional skills at five years of age.
+Col.
+(1) shows the mean of the outcome for males.
+Col.
+(2) regresses the outcome on the scores obtained from the factor model in \autoref{sec:methods}.
+Col.
+(3) additionally controls for cognitive ability at age five.
+This is a simple factor score obtained by aggregating the available cognitive measures.
+Col.
+(4) uses sum scores (see \autoref{fig:sumscores}) instead of factor scores.
+All standard errors in parentheses are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression.
+Col.
+(5) to (8) repeat for female cohort members.
 All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
+See \autoref{tab:harmvar} for a description of the variables used.
 \par}
@@
 # collapse to matrix
-bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab[c("hnvq30", "lhgrpay38", "obs42")],matrix,ncol=9,byrow=FALSE))
+bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab[c("hnvq30", "lginc34", "lhgrpay38")],matrix,ncol=9,byrow=FALSE))
 table <- bcs_outctab_mat
@@
 @
-\justify {\scriptsize \emph{Notes}: Robust standard errors in parentheses.
+\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of cohort members' adolescent and adult outcomes on their externalising and internalising socioemotional skills at five years of age.
+Col.
+(1) shows the mean of the outcome for males.
+Col.
+(2) regresses the outcome on the scores obtained from the factor model in \autoref{sec:methods}.
+Col.
+(3) additionally controls for cognitive ability at age five.
+This is a simple factor score obtained by aggregating the available cognitive measures.
+Col.
+(4) uses sum scores (see \autoref{fig:sumscores}) instead of factor scores.
+All standard errors in parentheses are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression.
+Col.
+(5) to (8) repeat for female cohort members.
 All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
+See \autoref{tab:harmvar} for a description of the variables used.
+\par}
+\end{table}
+
+\begin{table}
+\centering\footnotesize
+\caption{Predictors of adult outcomes -- BCS}\label{tab:bcs_adult2}
+<<REGS_OUTC_BCS3, results='asis', dependson="REGS_OUTC">>=
+
+# collapse to matrix
+bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab[c("smoke38", "smoke42", "obs42")],matrix,ncol=9,byrow=FALSE))
+table <- bcs_outctab_mat
+printoutctab(table, 3)
+
+@
+\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of cohort members' adolescent and adult outcomes on their externalising and internalising socioemotional skills at five years of age.
+Col.
+(1) shows the mean of the outcome for males.
+Col.
+(2) regresses the outcome on the scores obtained from the factor model in \autoref{sec:methods}.
+Col.
+(3) additionally controls for cognitive ability at age five.
+This is a simple factor score obtained by aggregating the available cognitive measures.
+Col.
+(4) uses sum scores (see \autoref{fig:sumscores}) instead of factor scores.
+All standard errors in parentheses are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression.
+Col.
+(5) to (8) repeat for female cohort members.
+All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
+See \autoref{tab:harmvar} for a description of the variables used.
 \par}
@@
 @
-\caption{Scores}
-\justify {\scriptsize \emph{Notes}: Density of scores from final model.
+\caption{Distribution of factor scores}\label{fig:dens}
+\justify {\scriptsize \emph{Notes}: The figure shows the distribution of the externalising and internalising socioemotional skills scores at age five obtained from the factor model, by gender and cohort.
+The scores are estimated from parameter estimates in \autoref{tab:measpars}, using an Empirical Bayes Modal approach.
+Higher scores correspond to \emph{better} skills.
+The distribution is estimated nonparametrically, using an Epanechnikov kernel.
+The figure also reports the $p$ value from Kolmogorov-Smirnov tests of equality between the distribution in BCS and MCS.
 \par}
@@
 @
-\caption{Sum scores}
-\justify {\scriptsize \emph{Notes}: Density of sum scores without factor analysis.
+\caption{Distribution of sum scores}\label{fig:sumscores}
+\justify {\scriptsize \emph{Notes}: The figure shows the distribution of the externalising and internalising sum scores at age five, by gender and cohort.
+The scores are obtained by assigning 0, 1, or 2 points for each item in the scale in \autoref{tab:scalecomp}.
+Zero points are assigned for `Certainly Applies / True' responses, one point for `Sometimes applies / somewhat true', and two points for `Doesn't apply'.
+Only 0 or 1 points are assigned for items that are coded as having two categories (5,6, and 11).
+Higher scores correspond to \emph{better} skills.
 \par}
@@
 \begin{figure}
-<<FACINEQ_FSC, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
+<<FACINEQ_FSC, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
 fi <- fi_fsc # select result
@@
 @
-\caption{Inequality by father occupation -- normalised}
+\caption{Skill inequality by father's occupation}
+\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socioemotional skills scores by gender, cohort, and father's occupation at age five.
+Father's occupation is based on Registrar General's social class, with classes I to III Non Manual being `White collar' and classes III Manual to V (plus `other') being `Blue collar'.
+`No father figure' is defined as absence of a male figure living in the household.
+The four panels on top present mean and 95\% confidence intervals.
+Given that we cannot compare means of skills, all scores are normalised to take value zero for the `Blue collar' category, so that the gradient is emphasised.
+The bottom two panels present the unconditional distribution of father's occupation.
+\par}
 \end{figure}
@@
 \begin{figure}
-<<FACINEQ_PS, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
+<<FACINEQ_PS, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
 fi <- fi_ps # select result
@@
 @
-\caption{Inequality by mother education -- normalised}
+\caption{Skill inequality by mother's education}
+\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socioemotional skills scores by gender, cohort, and mother's education at age five.
+Mother's education is a dummy for whether the mother continued schooling past the minimum leaving age, based on her date of birth.
+The four panels on top present mean and 95\% confidence intervals.
+Given that we cannot compare means of skills, all scores are normalised to take value zero for the `Compulsory' category, so that the gradient is emphasised.
+The bottom two panels present the unconditional distribution of mother's education.
+\par}
 \end{figure}
@@
 \begin{figure}
-<<FACINEQ_SM, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
+<<FACINEQ_SM, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
 fi <- fi_sm # select result
@@
 @
-\caption{Inequality by mother smoking in pregnancy -- normalised}
-\end{figure}
-
-\begin{figure}
-<<ITEMINEQ_FSC, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=11, out.width='\\linewidth'>>=
-itempl_fsc
-@
-\caption{Item-level changes in inequality -- father occupation}
-\justify {\scriptsize \emph{Notes}: The graph displays the ratio between low and high SES in the prevalence of each behaviour, separately by cohort.
+\caption{Skill inequality by mother's pregnancy smoking}
+\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socioemotional skills scores by gender, cohort, and mother's pregnancy smoking.
+Mother's education is a dummy for whether the mother reported smoking during pregnancy.
+The four panels on top present mean and 95\% confidence intervals.
+Given that we cannot compare means of skills, all scores are normalised to take value zero for the `Non-smoker' category, so that the gradient is emphasised.
+The bottom two panels present the unconditional distribution of mother's pregnancy smoking
 \par}
@@
 \begin{figure}
-<<ITEMINEQ_PS, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=11, out.width='\\linewidth'>>=
-itempl_ps
+<<ITEMINEQ_FSC, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=11, out.width='\\linewidth'>>=
+itempl_fsc
 @
-\caption{Item-level changes in inequality -- mother education}
-\justify {\scriptsize \emph{Notes}: The graph displays the ratio between low and high SES in the prevalence of each behaviour, separately by cohort.
+\caption{Item-level inequality by father's occupation}
+\justify {\scriptsize \emph{Notes}: The graph displays the ratio between the prevalence of each item in our scale in children of white collar vs blue collar fathers, by cohort and gender.
+All items that have three categories in the scale have been dichotomised.
+For example, if the prevalence of the `Restless' behaviours among children of blue collar fathers in the BCS cohort is 7.5\%, and 5\% among white collar fathers, the ratio will be 1.5.
+The error bars represent 95\% confidence intervals.
 \par}
@@
 \begin{figure}
-<<ITEMINEQ_SM, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=11, out.width='\\linewidth'>>=
+<<ITEMINEQ_PS, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=11, out.width='\\linewidth'>>=
+itempl_ps
+@
+\caption{Item-level inequality by mother's education}
+\justify {\scriptsize \emph{Notes}: The graph displays the ratio between the prevalence of each item in our scale in children of educated vs uneducated mothers, by cohort and gender.
+All items that have three categories in the scale have been dichotomised.
+For example, if the prevalence of the `Restless' behaviours among children of mothers with compulsory schooling in the BCS cohort is 7.5\%, and 5\% among mothers with post-compulsory schooling, the ratio will be 1.5.
+The error bars represent 95\% confidence intervals.
+\par}
+\end{figure}
+
+\begin{figure}
+<<ITEMINEQ_SM, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=11, out.width='\\linewidth'>>=
 itempl_sm
 @
-\caption{Item-level changes in inequality -- mother smoking in pregnancy}
-\justify {\scriptsize \emph{Notes}: The graph displays the ratio between low and high SES in the prevalence of each behaviour, separately by cohort.
+\caption{Item-level inequality by mother's pregnancy smoking}
+\justify {\scriptsize \emph{Notes}: The graph displays the ratio between the prevalence of each item in our scale in children of mothers who smoked in pregnancy vs non-smokers, by cohort and gender.
+All items that have three categories in the scale have been dichotomised.
+For example, if the prevalence of the `Restless' behaviours among children of smoker mothers in the BCS cohort is 7.5\%, and 5\% among non-smoker mothers, the ratio will be 1.5.
+The error bars represent 95\% confidence intervals.
 \par}

### 2018-09-26 codes/R/5_selectvars.R

@@ -84,8 +84,11 @@ bcsoutclabs <- c(smktry16 = "Tried smoking (16)",
                  owt16 = "Overweight (16)",
                  thn16 = "Thin (16)",
                  hnvq30 = "Higher education (30)",
+                 lginc34 = "(log) Gross weekly income (34)",
                  lhgrpay38 = "(log) Gross hr. pay (38)",
                  #hscl42 = "Social class I/II (42)",
+                 smoke38 = "Daily smoker (38)",
+                 smoke42 = "Daily smoker (42)",
                  bmi42 = "BMI (42)",
                  #owt42 = "Overweight (42)",
                  obs42 = "Obese (42)"

### 2018-09-26 codes/R/6_graphs.R

@@ -29,21 +29,21 @@ addopts.raw <- function(x) {
       axis.title.y=element_blank(),
       legend.position="none"
     ) +
-    scale_x_continuous(name = "Raw score", breaks = seq(0,10,1)) +
+    scale_x_continuous(name = "Sum score", breaks = seq(0,10,1)) +
     scale_fill_discrete("") + # remove fill guide title
     scale_colour_discrete(guide=FALSE) +  # remove colour legend
     geom_bar(position="dodge", aes(y = ..prop..)) # PLOTS
   return(x)
 }
 
-rawext.m <- ggplot(subset(finaldata, sex=="M"),aes(x=EXT_RAW, fill=cohort)) + ggtitle("EXT Scores (Males)") + coord_cartesian(ylim = c(0,.25))
-rawint.m <- ggplot(subset(finaldata, sex=="M"),aes(x=INT_RAW, fill=cohort)) + ggtitle("INT Scores (Males)") + coord_cartesian(ylim = c(0,.4))
-rawext.f <- ggplot(subset(finaldata, sex=="F"),aes(x=EXT_RAW, fill=cohort)) + ggtitle("EXT Scores (Females)") + coord_cartesian(ylim = c(0,.25))
-rawint.f <- ggplot(subset(finaldata, sex=="F"),aes(x=INT_RAW, fill=cohort)) + ggtitle("INT Scores (Females)") + coord_cartesian(ylim = c(0,.4))
+rawext.m <- ggplot(subset(finaldata, sex=="M"),aes(x=EXT_RAW, fill=cohort)) + ggtitle("Males EXT") + coord_cartesian(ylim = c(0,.25))
+rawint.m <- ggplot(subset(finaldata, sex=="M"),aes(x=INT_RAW, fill=cohort)) + ggtitle("Males INT") + coord_cartesian(ylim = c(0,.4))
+rawext.f <- ggplot(subset(finaldata, sex=="F"),aes(x=EXT_RAW, fill=cohort)) + ggtitle("Females EXT") + coord_cartesian(ylim = c(0,.25))
+rawint.f <- ggplot(subset(finaldata, sex=="F"),aes(x=INT_RAW, fill=cohort)) + ggtitle("Females INT") + coord_cartesian(ylim = c(0,.4))
 rawlist <- list(rawext.m, rawint.m, rawext.f, rawint.f) 
 rawlist <- lapply(rawlist, addopts.raw) # apply options to all graphs
 
[…]

### 2018-09-26 codes/R/6_graphs_ineq.R

@@ -51,6 +51,7 @@ plotineq <- function(data, categ,
     x <- x + xlab(xlab) +
       scale_y_continuous(name = ylab, breaks = round(seq(-.8,.8,.2),1)) +
       coord_cartesian(ylim = ylim) +
+      geom_hline(yintercept=0, color = "black", linetype = "dashed", size=0.5) +
       stat_summary(geom="errorbar", fun.data=mean_cl_normal, width=.2, position=position_dodge(.5)) +
       stat_summary(fun.y=mean, geom="point", size=4, aes(colour=cohort), position=position_dodge(.5)) +
       labs(list(colour="")) + theme(legend.position="none")

### 2018-09-28 codes/R/6_graphs.R
change labels

@@ -92,10 +92,10 @@ addopts.dens <- function(x) {
 }
 
 # densities of factor scores
-pdext.ebm.m <- ggplot(subset(finaldata, sex=="M"), aes(x=EXT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("Males EXT")
-pdint.ebm.m <- ggplot(subset(finaldata, sex=="M"), aes(x=INT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("Males INT")
-pdext.ebm.f <- ggplot(subset(finaldata, sex=="F"), aes(x=EXT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("Females EXT")
-pdint.ebm.f <- ggplot(subset(finaldata, sex=="F"), aes(x=INT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("Females INT")
+pdext.ebm.m <- ggplot(subset(finaldata, sex=="M"), aes(x=EXT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("Males Externalising")
+pdint.ebm.m <- ggplot(subset(finaldata, sex=="M"), aes(x=INT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("Males Internalising")
+pdext.ebm.f <- ggplot(subset(finaldata, sex=="F"), aes(x=EXT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("Females Externalising")
+pdint.ebm.f <- ggplot(subset(finaldata, sex=="F"), aes(x=INT, group=cohort, fill=cohort, colour=cohort)) + ggtitle("Females Internalising")
 denslist <- list(pdext.ebm.m, pdint.ebm.m, pdext.ebm.f, pdint.ebm.f) 
 denslist <- lapply(denslist, addopts.dens) # apply options to all graphs
 # add KS pvalue

### 2018-09-28 codes/R/6_graphs_ineq.R

@@ -108,6 +108,7 @@ plotineq <- function(data, categ,
   out$ineqlist <- ineqlist
   out$histlist <- histlist
   out$legend <- legend_b
+  out$data <- datatemp
   
   return(out)
 }
@@ -116,16 +117,16 @@ plotineq <- function(data, categ,
 # reorder social class to have "other" on the LHS (scl10c)
 finaldata$scl10c <- factor(finaldata$scl10b, levels(finaldata$scl10b)[c(5,1:4)])
 # demean scores for lowest class
-finaldata <- demean(finaldata, "scl10b", "IIINM", "dsc")
+ineqdata_temp <- demean(finaldata, "scl10b", "IIINM", "dsc")
 # make plots
-fi_sc <- plotineq(finaldata, "scl10b", "dsc", ylab = "Score (IIINM = 0)", xlab = "Family Social Class (10)",
+fi_sc <- plotineq(ineqdata_temp, "scl10b", "dsc", ylab = "Score (IIINM = 0)", xlab = "Family Social Class (10)",
                   ylim = c(-.6,.4), vline=4.5)
 
 # FATHER SOCIAL CLASS at 5 --------------------
 # demean scores for lowest class
-finaldata <- demean(finaldata, "fscl5wb", "Blue collar", "dfsc")
+ineqdata_temp <- demean(finaldata, "fscl5wb", "Blue collar", "dfsc")
 # make plots
-fi_fsc <- plotineq(finaldata, "fscl5wb", "dfsc", ylab = "Score (Blue c. = 0)", xlab = "Father Occupation (5)",
+fi_fsc <- plotineq(ineqdata_temp, "fscl5wb", "dfsc", ylab = "Score (Blue c. = 0)", xlab = "Father Occupation (5)",
                    ylim = c(-.4,.4))
 
 # # non-normalised version
@@ -134,19 +135,19 @@ fi_fsc <- plotineq(finaldata, "fscl5wb", "dfsc", ylab = "Score (Blue c. = 0)", x
 
 # MATERNAL SCHOOLING -------------------------
 # demean scores for lowest level
-finaldata <- demean(finaldata, "mysch5b", "17-18", "dys")
-fi_ys <- plotineq(finaldata, "mysch5b", "dys", ylab = "Score (17-18 = 0)", xlab = "Age mother left FTE (5)",
+ineqdata_temp <- demean(finaldata, "mysch5b", "17-18", "dys")
[…]

### 2018-09-28 codes/R/5_selectvars.R
change outcomes, clean employment, NVQ

@@ -16,7 +16,7 @@ mchxbth <-  list(apply(
 decvarslist_int <- c(decvarslist, mchxbth)
 
 # names of the groups
-decvarsgroups <- c("Maternal education (5)", "Maternal employment (5)", "Father occupation (5)","Maternal background (0)", "Pregnancy")
+decvarsgroups <- c("Maternal education (5)", "Maternal employment (5)", "Father occ. (5) - White Collar = 0","Maternal background (0)", "Pregnancy")
 
 # make into vector
 decvars <- unname(unlist(decvarslist))
@@ -49,13 +49,12 @@ mcs14outc$id <- mcs14outc$mcsid
 mcsoutc <- merge(regdata[regdata$cohort=="MCS",], mcs14outc, by="id", all.x = TRUE)
 mcsoutc$cogscore <- as.vector(factor.scores(mcsoutc[,c("C1", "C2", "C3")], fa(mcsoutc[,c("C1", "C2", "C3")], factors=1), method = "Bartlett")$scores)
 
-# recode
-bcsoutc$hscl42 <- NA # high social class
-bcsoutc$hscl42[bcsoutc$scl42 %in% c(1,2)] <- 1
-bcsoutc$hscl42[bcsoutc$scl42 %in% 3:7] <- 0
-bcsoutc$hnvq30 <- NA # high NVQ
-bcsoutc$hnvq30[bcsoutc$nvq30 %in% c(4,5)] <- 1
-bcsoutc$hnvq30[bcsoutc$nvq30 %in% 0:3] <- 0
+# recode NVQ and employment
+bcsoutc <- bcsoutc %>%
+  mutate(hnvq30 = ifelse(nvq30 %in% c(4,5), 1, ifelse(nvq30 %in% c(0,1,2,3), 0, NA))) %>%
+  mutate(hnvq34 = ifelse(nvq34 %in% c(4,5), 1, ifelse(nvq34 %in% c(0,1,2,3), 0, NA))) %>%
+  mutate(emp34 = ifelse(empst34 %in% c(1,2), 1, ifelse(empst34==0, 0, NA))) %>%
+  mutate(emp42 = ifelse(empst42 %in% c(1,2), 1, ifelse(empst42==0, 0, NA)))
 
 # BMI splits
[…]

### 2018-09-28 codes/R/8_regs_dep.R

@@ -37,7 +37,7 @@ decvars_int_reg <- c(decvars_int, c("region"))
 # labels for groups (texreg)
 grouplabs <- list("Maternal education (5)" = 1, 
                   "Maternal employment (5)" = 2,
-                  "Father occupation (5)" = 3:4, 
+                  "Father occ. (5) - White collar = 0" = 3:4, 
                   "Maternal background (0)" = 5:7,
                   "Pregnancy" = 8:10)
 # labels for variables (texreg)

### 2018-09-28 codes/R/8_regs_outc.R

@@ -15,7 +15,7 @@ for (o in 1:length(bcsoutcvars)) { # length(bcsoutcvars)
   form_outc <- list()
   form_outc$b    <- as.formula(paste0(outc, " ~ EXT + INT + age16 + region +", pplus(decvars)))
   form_outc$bc   <- as.formula(paste0(outc, " ~ EXT + INT + age16 + region + cogscore +", pplus(decvars)))
-  form_outc$bcr  <- as.formula(paste0(outc, " ~ EXT_RAW + INT_RAW + age16 + region + cogscore +", pplus(decvars)))
+  form_outc$bcr  <- as.formula(paste0(outc, " ~ EXT_RAWs + INT_RAWs + age16 + region + cogscore +", pplus(decvars)))
   
   for (s in c("M", "F")) {
     # bootstrap SEs
@@ -58,7 +58,7 @@ for (o in 1:length(mcsoutcvars)) { # length(bcsoutcvars)
   form_outc <- list()
   form_outc$b    <- as.formula(paste0(outc," ~ EXT + INT + age14 + region +", pplus(decvars)))
   form_outc$bc   <- as.formula(paste0(mcsoutcvars[o]," ~ EXT + INT + age14 + region + cogscore +", pplus(decvars)))
-  form_outc$bcr  <- as.formula(paste0(mcsoutcvars[o]," ~ EXT_RAW + INT_RAW + age14 + region + cogscore +", pplus(decvars)))
+  form_outc$bcr  <- as.formula(paste0(mcsoutcvars[o]," ~ EXT_RAWs + INT_RAWs + age14 + region + cogscore +", pplus(decvars)))
   bootcoefs <- list()
   
   for (s in c("M", "F")) {
@@ -107,19 +107,19 @@ bcs_outctab <- list()
 for (i in 1:length(bcsoutcvars)) {
   lab <- paste0("\\textit{\\textbf{", bcsoutclabs[i], "}}")
   bcs_outctab[[bcsoutcvars[i]]] <- matrix(c(lab, dvmeans_bcs_m[i], "", "", "", dvmeans_bcs_f[i], "", "", "",   # outcome and means
-                                            "Externalising (factor)", # row label
+                                            "Externalising skills", # row label
                                             "", cellpr(bcsoutcsum$M[[i]]$b, "EXT"), cellpr(bcsoutcsum$M[[i]]$bc, "EXT"), "", # males EXT
[…]

### 2018-09-28 codes/R/paper/paper_jpube.Rnw
Version with external tex

@@
 \title{Inequality in socioemotional skills: a cross-cohort comparison}
-\author[1,2]{Orazio Attanasio}
-\author[1,2]{Richard Blundell}
-\author[1,2]{Gabriella Conti}
-\author[1,2]{Giacomo Mason}
-\affil[1]{\emph{Institute for Fiscal Studies}}
-\affil[1]{\emph{UCL Department of Economics}}
-%\institute{UCL and IFS}
+\author{Orazio Attanasio \hspace{10pt} Richard Blundell \\  Gabriella Conti \hspace{10pt} Giacomo Mason}
+\affil{\emph{Institute for Fiscal Studies} \\ \emph{UCL Department of Economics}}
 \date{\today}
@@
 \begin{document}
+
+\begin{titlepage}
 \maketitle
@@
   \begin{abstract}
-abstract here
+    Abstract here
   \end{abstract}
 
-% We are grateful to The Centre for Longitudinal Studies, UCL Institute of Education for the use of these data and to the UK Data Archive and UK Data Service for making them available.
-However, they bear no responsibility for the analysis or interpretation of these data.
+\vspace{30pt}
+{\footnotesize 
+We thank {\color{orange} participants to the Trans Atlantic Public Economics Seminar }
+We are grateful to The Centre for Longitudinal Studies, UCL Institute of Education for the use of the data and to the UK Data Archive and UK Data Service for making them available.
+However, they bear no responsibility for the analysis or interpretation of the data.
+Giacomo Mason obtained funding thanks to an Economic and Social Research Council graduate scholarship.
+}
+\end{titlepage}
+
+\input{/Users/giacomomason/Dropbox/cohorts_shared/paper_jpube_1intro.tex}
+\input{/Users/giacomomason/Dropbox/cohorts_shared/paper_jpube_2datamethods.tex}
+\input{/Users/giacomomason/Dropbox/cohorts_shared/paper_jpube_3resultsconclusion.tex}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\caption{Quantile differences in scores}\label{tab:iqdiff}
+\centering\setstretch{1.2}\footnotesize
+<<INEQ_STATS, dependson="SELECTDATA", results='asis'>>=
+
+# income 
+tb <- data.frame(
+  aggregate(finaldata$faminc10_infl, by=list(finaldata$cohort), FUN = function(x) quantile(x, c(.1, .25, .75, .9), na.rm=T))
+)
+tb$d9010 <- tb$x[,4]/tb$x[,1]
+tb$d7525 <- tb$x[,3]/tb$x[,2]
+
+tb.ext <- data.frame(
+  aggregate(finaldata$EXT, by=list(finaldata$cohortsex), FUN = function(x) quantile(x, c(.1, .25, .5, .75, .9), na.rm=T))
+)
+tb.ext$d9010 <- abs(tb.ext$x[,"90%"]-tb.ext$x[,"10%"])
+tb.ext$d7525 <- abs(tb.ext$x[,"75%"]-tb.ext$x[,"25%"])
+tb.ext$d9050 <- abs(tb.ext$x[,"90%"]-tb.ext$x[,"50%"])
+tb.ext$d5010 <- abs(tb.ext$x[,"50%"]-tb.ext$x[,"10%"])
+rownames(tb.ext) <- tb.ext$Group.1
+tb.ext <- t(tb.ext[,-c(1,2)]) # remove aggregated quantiles and transpose
+tb.ext <- tb.ext[,c("BCS.M","MCS.M","BCS.F","MCS.F")]
+
+tb.int <- data.frame(
+  aggregate(finaldata$INT, by=list(finaldata$cohortsex), FUN = function(x) quantile(x, c(.1, .25, .5, .75, .9), na.rm=T))
+)
+tb.int$d9010 <- abs(tb.int$x[,"90%"]-tb.int$x[,"10%"])
+tb.int$d7525 <- abs(tb.int$x[,"75%"]-tb.int$x[,"25%"])
+tb.int$d9050 <- abs(tb.int$x[,"90%"]-tb.int$x[,"50%"])
+tb.int$d5010 <- abs(tb.int$x[,"50%"]-tb.int$x[,"10%"])
+rownames(tb.int) <- tb.int$Group.1
+tb.int <- t(tb.int[,-c(1,2)]) # remove aggregated quantiles and transpose
+tb.int <- tb.int[,c("BCS.M","MCS.M","BCS.F","MCS.F")]
+
+# assemble and print
+table <- rbind(tb.ext,tb.int)
+table <- cbind(data.frame(as.matrix(rep(c("90-10", "75-25", "90-50", "50-10"),2))),
+               table)
+
+print(xtable(table,
+                    align="L{0cm}L{.15\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}",
+                    digits = 2
+                    ),
+             size="\\footnotesize",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0,4)),
+                              command = c(
+                              # header
+            " & \\multicolumn{2}{c}{\\textbf{Males}} & \\multicolumn{2}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-3}  \\cmidrule(lr){4-5} 
+                       & BCS (1970) & MCS (2000/1) & BCS (1970) & MCS (2000/1)  \\\\
+                      \\textbf{Externalising} \\\\ ",
+                      "\\textbf{Internalising} \\\\ \\midrule "
+                                    )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = "\\textwidth",
+             floating = FALSE
+)
+
+@
+\justify {\scriptsize \emph{Notes}: The table shows differences between quantiles of the distribution of socioemotional skills, by gender and cohort.
+The distribution is a factor score obtained from the factor model in \autoref{sec:methods}.
+\par}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{landscape}
+\begin{table}
+\caption{Determinants of Socioemotional Skills}\label{tab:determ}
+\centering\footnotesize
+<<REGS_ALL, results='asis', dependson="REGS_DEP">>=
+
+# select models
+regs <- list(r_ext$BCS.M[["OLS"]],r_ext$MCS.M[["OLS"]],r_ext$BCS.F[["OLS"]],r_ext$MCS.F[["OLS"]],
+             r_int$BCS.M[["OLS"]],r_int$MCS.M[["OLS"]],r_int$BCS.F[["OLS"]],r_int$MCS.F[["OLS"]])
+
+# robust SEs
+robust_se <- list()
+robust_p  <- list()
+bootstrap_se  <- list(r_ext$BCS.M[["OLSbootse"]],r_ext$MCS.M[["OLSbootse"]],r_ext$BCS.F[["OLSbootse"]],r_ext$MCS.F[["OLSbootse"]],
+                      r_int$BCS.M[["OLSbootse"]],r_int$MCS.M[["OLSbootse"]],r_int$BCS.F[["OLSbootse"]],r_int$MCS.F[["OLSbootse"]])
+for (i in 1:length(regs)) {
+  cov1              <- sandwich::vcovHC(regs[[i]], type = "HC1")
+  ct                <- lmtest::coeftest(regs[[i]], vcov = cov1)
+  robust_se[[i]] <- ct[, 2]
+  robust_p[[i]] <- ct[, 4]
+}
+
+texreg::texreg(regs,
+               stars = c(0.1, 0.05, 0.01), digits = 3,
+               override.se = bootstrap_se, override.pvalues = robust_p,
+               single.row = FALSE, scriptsize = TRUE,
+               caption = "", label = "",
+               omit.coef = omitcoefs,
+               groups = grouplabs,
+               custom.columns = list("p-value" = r_ext$Mp, "p-value" = r_ext$Fp, "p-value" = r_int$Mp, "p-value" = r_int$Fp), custom.col.pos = c(4,6,8,10),
+               custom.coef.names = varlabs,
+               custom.model.names = c("\\multicolumn{6}{c}{\\textbf{Externalising}} & \\multicolumn{6}{c}{\\textbf{Internalising}} \\\\
+                                        \\cmidrule(lr){2-7} \\cmidrule(lr){8-13}
+                                        & \\multicolumn{3}{c}{\\textbf{Males}} & \\multicolumn{3}{c}{\\textbf{Females}} &
+                                        \\multicolumn{3}{c}{\\textbf{Males}} & \\multicolumn{3}{c}{\\textbf{Females}} \\\\ 
+                                        \\cmidrule(lr){2-4} \\cmidrule(lr){5-7} \\cmidrule(lr){8-10} \\cmidrule(lr){11-13} 
+                                      & (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8) & (9) & (10) & (11) & (12) \\\\
+                                      & BCS", "MCS", "BCS", "MCS", "BCS", "MCS", "BCS", "MCS"),
+               custom.note = "",
+               include.rmse = FALSE, include.rsquared = FALSE,
+               table = FALSE
+               )
+
+@
+\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of children's socioemotional skills at five years of age on family background information.
+The dependent variable is a factor score obtained from the factor model in \autoref{sec:methods}.
+Col.
+(1) and (2) show coefficients and standard errors in parentheses, for male children in the BCS and MCS cohorts separately.
+The latter are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression.
+Col.
+(3) shows the p-value of a test that the coefficient is the same in the two cohorts.
+Col.
+(4) to (6) repeat for female children.
+Col.
+(7) to (12) repeat for internalising skills.
+All estimates additionally control for region of birth, mother height, number of previous stillbirths at child's birth, preterm birth, a dummy for missing gestational age, and number of other children in the household at child age 5.
+See \autoref{tab:harmvar} for a description of the variables used.
+\par}
+\end{table}
+\end{landscape}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+<<REGS_OUTC_TABPR, dependson="REGS_OUTC">>=
+# function to display outcome regression tables
+# numoutc = number of outcomes
+
+# EXCLUDING RAW SCORES
+printoutctab <- function(tab, numoutc) {
+print(xtable(tab,
+                    align="R{0cm}L{.3\\textwidth}C{.05\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}",
+                    digits = 3
+                    ),
+             size="\\footnotesize",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = as.list(c(0, 6*seq(1,numoutc-1))),
+                              command = c(
+                              # header
+            " & \\multicolumn{3}{c}{\\textbf{Males}} & \\multicolumn{3}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-4}  \\cmidrule(lr){5-7}
+& Mean & \\multicolumn{2}{c}{Coefficients} & Mean & \\multicolumn{2}{c}{Coefficients} \\\\ \\cmidrule(lr){3-4}  \\cmidrule(lr){6-7}
+            & (1) & (2) & (3) & (4) & (5) & (6) \\\\",
+            rep(" \\midrule", numoutc-1)
+            )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular",
+             floating = FALSE
+)
+}
+
+# INCLUDING RAW SCORES
+printoutctab_raw <- function(tab, numoutc) {
+print(xtable(tab,
+                    align="R{0cm}L{.2\\textwidth}C{.05\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}",
+                    digits = 3
+                    ),
+             size="\\footnotesize",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = as.list(c(0, 8*seq(1,numoutc-1))),
+                              command = c(
+                              # header
+            " & \\multicolumn{4}{c}{\\textbf{Males}} & \\multicolumn{4}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-5}  \\cmidrule(lr){6-9}
+& Mean & \\multicolumn{3}{c}{Coefficients} & Mean & \\multicolumn{3}{c}{Coefficients} \\\\ \\cmidrule(lr){3-5}  \\cmidrule(lr){7-9}
+            & (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8) \\\\",
+            rep(" \\midrule", numoutc-1)
+            )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular",
+             floating = FALSE
+)
+}
+@
+
+\begin{table}
+\centering\footnotesize
+\caption{Predictors of adolescent outcomes}\label{tab:adol}
+<<REGS_OUTC_ADOL, results='asis', dependson="REGS_OUTC">>=
+
+# collect adolescent outcomes
+outctabs <- list(bcs_outctab$smktry16, bcs_outctab$bmi16, mcs_outctab$smktry14, mcs_outctab$bmi14)
+outctabs <- lapply(outctabs, function(x) x[-4:-5, c(-5,-9)]) # eliminate raw rows and columns
+# collapse to matrix
+outctabs_mat <- do.call(rbind,lapply(outctabs,matrix,ncol=7,byrow=FALSE))
+printoutctab(outctabs_mat, length(outctabs))
+
+@
+\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of cohort members' adolescent outcomes on their externalising and internalising socioemotional skills at five years of age.
+Col.
+(1) shows the mean of the outcome for males.
+Col.
+(2) regresses the outcome on the scores obtained from the factor model in \autoref{sec:methods}.
+Col.
+(3) additionally controls for cognitive ability at age five.
+This is a simple factor score obtained by aggregating the available cognitive measures.
+All standard errors in parentheses are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression.
+Col.
+(4) to (6) repeat for female cohort members.
+All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
+See \autoref{tab:harmvar} for a description of the variables used.
+\par}
+\end{table}
+
+\begin{table}
+\centering\footnotesize
+\caption{Predictors of adult outcomes -- BCS}\label{tab:bcs_adult}
+<<REGS_OUTC_ADULT, results='asis', dependson="REGS_OUTC">>=
+
+# eliminate raw rows and columns
+outcs <- c("hnvq34", "emp34", "lgpay34", "emp42", "lgpay42")
+outctabs <- list()
+for (o in 1:length(outcs)) outctabs[[outcs[o]]] <- bcs_outctab[[outcs[o]]][-4:-5, c(-5,-9)]
+# collapse to matrix
+outctabs_mat <- do.call(rbind,lapply(outctabs,matrix,ncol=7,byrow=FALSE))
+printoutctab(outctabs_mat, length(outcs))
+
+@
+\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of BCS cohort members' adult outcomes on their externalising and internalising socioemotional skills at five years of age.
+Col.
+(1) shows the mean of the outcome for males.
+Col.
+(2) regresses the outcome on the scores obtained from the factor model in \autoref{sec:methods}.
+Col.
+(3) additionally controls for cognitive ability at age five.
+This is a simple factor score obtained by aggregating the available cognitive measures.
+All standard errors in parentheses are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression.
+Col.
+(4) to (6) repeat for female cohort members.
+All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
+See \autoref{tab:harmvar} for a description of the variables used.
+\par}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}
+<<FACDENS, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
+@
+\caption{Distribution of factor scores}\label{fig:dens}
+\justify {\scriptsize \emph{Notes}: The figure shows the distribution of the externalising and internalising socioemotional skills scores at age five obtained from the factor model, by gender and cohort.
+The scores are estimated from parameter estimates in \autoref{tab:measpars}, using an Empirical Bayes Modal approach.
+Higher scores correspond to \emph{better} skills.
+The distribution is estimated nonparametrically, using an Epanechnikov kernel.
+The figure also reports the $p$ value from Kolmogorov-Smirnov tests of equality between the distribution in BCS and MCS.
+\par}
+\end{figure}
+
+\begin{figure}
+<<FACINEQ_PS, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
+fi <- fi_ps # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\caption{Skill inequality by mother's education}\label{fig:ineqps}
+\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socioemotional skills scores by gender, cohort, and mother's education at age five.
+Mother's education is a dummy for whether the mother continued schooling past the minimum leaving age, based on her date of birth.
+The four panels on top present mean and 95\% confidence intervals.
+Given that we cannot compare means of skills, all scores are normalised to take value zero for the `Compulsory' category, so that the gradient is emphasised.
+The bottom two panels present the unconditional distribution of mother's education.
+\par}
+\end{figure}
+
+\begin{figure}
+<<FACINEQ_SM, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
+fi <- fi_sm # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\caption{Skill inequality by mother's pregnancy smoking}\label{fig:ineqsm}
+\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socioemotional skills scores by gender, cohort, and mother's pregnancy smoking.
+Mother's education is a dummy for whether the mother reported smoking during pregnancy.
+The four panels on top present mean and 95\% confidence intervals.
+Given that we cannot compare means of skills, all scores are normalised to take value zero for the `Non-smoker' category, so that the gradient is emphasised.
+The bottom two panels present the unconditional distribution of mother's pregnancy smoking
+\par}
+\end{figure}
+
+\begin{figure}
+<<FACINEQ_FSC, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
+fi <- fi_fsc # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\caption{Skill inequality by father's occupation}\label{fig:ineqfsc}
+\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socioemotional skills scores by gender, cohort, and father's occupation at age five.
+Father's occupation is based on Registrar General's social class, with classes I to III Non Manual being `White collar' and classes III Manual to V (plus `other') being `Blue collar'.
+`No father figure' is defined as absence of a male figure living in the household.
+The four panels on top present mean and 95\% confidence intervals.
+Given that we cannot compare means of skills, all scores are normalised to take value zero for the `Blue collar' category, so that the gradient is emphasised.
+The bottom two panels present the unconditional distribution of father's occupation.
+\par}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \clearpage
-\begin{table}[]
-\centering\setstretch{1.2} \footnotesize
-\caption{Parameterisations for measurement invariance}\label{tab:invparam}
-\begin{tabular}{L{.18\textwidth}L{.45\textwidth}L{.3\textwidth}}
-\toprule
-\textbf{Invariance level}     & \textbf{Description} & \textbf{Restrictions} \\ \midrule \\
-Configural (WE{\textTheta})   
-    & \bi[label=$\cdot$] \item Minimally restrictive model for identification \ei
-    & For all groups: $\begin{array}{|l} \text{diag}(\bm{\Phi}) = \bm{I} \\
-                      \bm{\kappa}=\bm{0} \\
-                      \bm{\nu}=\bm{0} \\
-                      \text{diag}(\bm{\Psi})=\bm{I} \end{array}$
-                      \\ \\
-Threshold invariance
-    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ to be equal across groups
-                  \item Statistically equivalent to configural (when measures have 3 categories or less)\ei
-    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
-                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
-                      \newline For all groups: $\begin{array}{|l} \text{diag}(\bm{\Phi}) = \bm{I} \\ 
-                                                \bm{\kappa}=\bm{0} \end{array}$
-                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\nu}_A=\bm{0} \\
-                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \end{array}$
-                   \\ \\
-Threshold and Loading invariance
-    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ and loadings $\lambda$ to be equal across groups
-                  \item Allows comparison of latent factor variances \ei
-    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
-                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
-                      \newline $\lambda_{ci} = \lambda_{c^\prime i}$ for all items, $\forall c, c^\prime$
-                      \newline For all groups: $\bm{\kappa}=\bm{0}$
-                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\nu}_A=\bm{0} \\
-                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \\
-                                                  \text{diag}(\bm{\Phi}_A) = \bm{I} \end{array}$ 
-                   \\ \\
-Threshold, Loading, and Intercept invariance
-    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ and loadings $\lambda$ to be equal across groups
-                  \item Restricts intercepts $\nu$ to zero in both groups
-                  \item Allows comparison of latent factor variances \emph{and} means \ei
-    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
-                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
-                      \newline $\lambda_{ci} = \lambda_{c^\prime i}$ for all items, $\forall c, c^\prime$
-                      \newline For all groups: $\bm{\nu}=\bm{0}$
-                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\kappa}_A=\bm{0} \\
-                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \\
-                                                  \text{diag}(\bm{\Phi}_A) = \bm{I} \end{array}$ 
-\\ \\ \bottomrule
-\end{tabular}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
+
+\begin{appendices}
+
+% to number tables and figures with A
+        \setcounter{table}{0}
+        \renewcommand{\thetable}{A\arabic{table}}%
+        \setcounter{figure}{0}
+        \renewcommand{\thefigure}{A\arabic{figure}}%
+
+% Appendix text
+\input{/Users/giacomomason/Dropbox/cohorts_shared/paper_jpube_4appendix.tex}
+
+% Appendix tables
 \clearpage
-
+\section{Appendix tables}
+
+% from tex file
+\input{/Users/giacomomason/Dropbox/cohorts_shared/paper_jpube_5apptables.tex}
+
+\clearpage
 \begin{landscape}
@@
 \begin{table}
-\caption{Estimated parameters -- Threshold and loading invariance model (anchored)}\label{tab:measpars}
+\caption{Parameter estimates from factor model with threshold and loading invariance}\label{tab:measpars}
 \centering\setstretch{1.2}\footnotesize
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \begin{table}
-\caption{Quantile differences in scores}\label{tab:iqdiff}
-\centering\setstretch{1.2}\footnotesize
-<<INEQ_STATS, dependson="SELECTDATA", results='asis'>>=
-
-# income 
-tb <- data.frame(
-  aggregate(finaldata$faminc10_infl, by=list(finaldata$cohort), FUN = function(x) quantile(x, c(.1, .25, .75, .9), na.rm=T))
-)
-tb$d9010 <- tb$x[,4]/tb$x[,1]
-tb$d7525 <- tb$x[,3]/tb$x[,2]
-
-tb.ext <- data.frame(
-  aggregate(finaldata$EXT, by=list(finaldata$cohortsex), FUN = function(x) quantile(x, c(.1, .25, .5, .75, .9), na.rm=T))
-)
-tb.ext$d9010 <- abs(tb.ext$x[,"90%"]-tb.ext$x[,"10%"])
-tb.ext$d7525 <- abs(tb.ext$x[,"75%"]-tb.ext$x[,"25%"])
-tb.ext$d9050 <- abs(tb.ext$x[,"90%"]-tb.ext$x[,"50%"])
-tb.ext$d5010 <- abs(tb.ext$x[,"50%"]-tb.ext$x[,"10%"])
-rownames(tb.ext) <- tb.ext$Group.1
-tb.ext <- t(tb.ext[,-c(1,2)]) # remove aggregated quantiles and transpose
-tb.ext <- tb.ext[,c("BCS.M","MCS.M","BCS.F","MCS.F")]
-
-tb.int <- data.frame(
-  aggregate(finaldata$INT, by=list(finaldata$cohortsex), FUN = function(x) quantile(x, c(.1, .25, .5, .75, .9), na.rm=T))
-)
-tb.int$d9010 <- abs(tb.int$x[,"90%"]-tb.int$x[,"10%"])
-tb.int$d7525 <- abs(tb.int$x[,"75%"]-tb.int$x[,"25%"])
-tb.int$d9050 <- abs(tb.int$x[,"90%"]-tb.int$x[,"50%"])
-tb.int$d5010 <- abs(tb.int$x[,"50%"]-tb.int$x[,"10%"])
-rownames(tb.int) <- tb.int$Group.1
-tb.int <- t(tb.int[,-c(1,2)]) # remove aggregated quantiles and transpose
-tb.int <- tb.int[,c("BCS.M","MCS.M","BCS.F","MCS.F")]
-
-# assemble and print
-table <- rbind(tb.ext,tb.int)
-table <- cbind(data.frame(as.matrix(rep(c("90-10", "75-25", "90-50", "50-10"),2))),
-               table)
-
-print(xtable(table,
-                    align="L{0cm}L{.15\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}",
-                    digits = 2
-                    ),
-             size="\\footnotesize",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = c(list(0,4)),
-                              command = c(
-                              # header
-            " & \\multicolumn{2}{c}{\\textbf{Males}} & \\multicolumn{2}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-3}  \\cmidrule(lr){4-5} 
-                       & BCS & MCS & BCS & MCS  \\\\
-                      \\textbf{Externalising} \\\\ ",
-                      "\\textbf{Internalising} \\\\ \\midrule "
-                                    )),
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabular", width = "\\textwidth",
-             floating = FALSE
-)
-
-@
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{table}
-\caption{Determinants of Externalising Skills}
-\centering\footnotesize
-<<REGS_EXT, results='asis', dependson="REGS_DEP">>=
-
-# select models
-regs <- list(r_ext$BCS.M[["OLS"]],r_ext$MCS.M[["OLS"]],r_ext$BCS.F[["OLS"]],r_ext$MCS.F[["OLS"]])
-
-# robust SEs
-robust_se <- list()
-robust_p  <- list()
-bootstrap_se  <- list(r_ext$BCS.M[["OLSbootse"]],r_ext$MCS.M[["OLSbootse"]],r_ext$BCS.F[["OLSbootse"]],r_ext$MCS.F[["OLSbootse"]])
-for (i in 1:length(regs)) {
-  cov1              <- sandwich::vcovHC(regs[[i]], type = "HC1")
-  ct                <- lmtest::coeftest(regs[[i]], vcov = cov1)
-  robust_se[[i]] <- ct[, 2]
-  robust_p[[i]] <- ct[, 4]
+\centering\scriptsize
+\caption{Predictors of adolescent outcomes, with sum scores}\label{tab:adol_app}
+\resizebox{15cm}{!}{
+<<REGS_OUTC_ADOL_APP, results='asis', dependson="REGS_OUTC">>=
+
+# collect adolescent outcomes
+outctabs <- list(bcs_outctab$smktry16, bcs_outctab$bmi16, mcs_outctab$smktry14, mcs_outctab$bmi14)
+# collapse to matrix
+outctabs_mat <- do.call(rbind,lapply(outctabs,matrix,ncol=9,byrow=FALSE))
+printoutctab_raw(outctabs_mat, length(outctabs))
+
+@
 }
-
-texreg::texreg(regs,
-               stars = c(0.1, 0.05, 0.01), digits = 3,
-               override.se = bootstrap_se, override.pvalues = robust_p,
-               single.row = FALSE, scriptsize = TRUE,
-               caption = "", label = "",
-               omit.coef = omitcoefs,
-               groups = grouplabs,
-               custom.columns = list("p-value" = r_ext$Mp, "p-value" = r_ext$Fp), custom.col.pos = c(4,6),
-               custom.coef.names = varlabs,
-               custom.model.names = c("\\multicolumn{3}{c}{\\textbf{Males}} & \\multicolumn{3}{c}{\\textbf{Females}} \\\\ 
-                                        \\cmidrule(lr){2-4} \\cmidrule(lr){5-7} 
-                                      & (1) & (2) & (3) & (4) & (5) & (6) \\\\
-                                      & BCS", "MCS", "BCS", "MCS"),
-               custom.note = "",
-               include.rmse = FALSE,
-               table = FALSE
-               )
-
-@
-\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of children's externalising socioemotional skills at five years of age on family background information.
-The dependent variable is a factor score obtained from the factor model in \autoref{sec:methods}.
-Col.
-(1) and (2) show coefficients and standard errors in parentheses, for male children in the BCS and MCS cohorts separately.
-The latter are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression.
-Col.
-(3) shows the p-value of a test that the coefficient is the same in the two cohorts.
-Col.
-(4) to (6) repeat for female children.
-All estimates additionally control for region of birth, mother height, number of previous stillbirths at child's birth, preterm birth, a dummy for missing gestational age, and number of other children in the household at child age 5.
-See \autoref{tab:harmvar} for a description of the variables used.
-\par}
-\end{table}
-
-\begin{table}
-\caption{Determinants of Internalising Skills}
-\centering\footnotesize
-<<REGS_INT, results='asis', dependson="REGS_DEP">>=
-
-# select models
-regs <- list(r_int$BCS.M[["OLS"]],r_int$MCS.M[["OLS"]],r_int$BCS.F[["OLS"]],r_int$MCS.F[["OLS"]])
-
-# robust SEs
-robust_se <- list()
-robust_p  <- list()
-bootstrap_se  <- list(r_int$BCS.M[["OLSbootse"]],r_int$MCS.M[["OLSbootse"]],r_int$BCS.F[["OLSbootse"]],r_int$MCS.F[["OLSbootse"]])
-for (i in 1:length(regs)) {
-  cov1              <- sandwich::vcovHC(regs[[i]], type = "HC1")
-  ct                <- lmtest::coeftest(regs[[i]], vcov = cov1)
-  robust_se[[i]] <- ct[, 2]
-  robust_p[[i]] <- ct[, 4]
-}
-
-texreg::texreg(regs,
-               stars = c(0.1, 0.05, 0.01), digits = 3,
-               override.se = bootstrap_se, override.pvalues = robust_p,
-               single.row = FALSE, scriptsize = TRUE,
-               caption = "", label = "",
-               omit.coef = omitcoefs,
-               groups = grouplabs,
-               custom.columns = list("p-value" = r_int$Mp, "p-value" = r_int$Fp), custom.col.pos = c(4,6),
-               custom.coef.names = varlabs,
-               custom.model.names = c("\\multicolumn{3}{c}{\\textbf{Males}} & \\multicolumn{3}{c}{\\textbf{Females}} \\\\ 
-                                        \\cmidrule(lr){2-4} \\cmidrule(lr){5-7} 
-                                      & (1) & (2) & (3) & (4) & (5) & (6) \\\\
-                                      & BCS", "MCS", "BCS", "MCS"),
-               custom.note = "",
-               include.rmse = FALSE,
-               table = FALSE
-               )
-
-@
-\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of children's internalising socioemotional skills at five years of age on family background information.
-The dependent variable is a factor score obtained from the factor model in \autoref{sec:methods}.
-Col.
-(1) and (2) show coefficients and standard errors in parentheses, for male children in the BCS and MCS cohorts separately.
-The latter are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression.
-Col.
-(3) shows the p-value of a test that the coefficient is the same in the two cohorts.
-Col.
-(4) to (6) repeat for female children.
-All estimates additionally control for region of birth, mother height, number of previous stillbirths at child's birth, preterm birth, a dummy for missing gestational age, and number of other children in the household at child age 5.
-See \autoref{tab:harmvar} for a description of the variables used.
-\par}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-<<REGS_OUTC_TABPR, dependson="REGS_OUTC">>=
-# function to display outcome regression tables
-# numoutc = number of outcomes
-printoutctab <- function(tab, numoutc) {
-print(xtable(tab,
-                    align="R{0cm}L{.2\\textwidth}C{.05\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}",
-                    digits = 3
-                    ),
-             size="\\footnotesize",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = as.list(c(0, 8*seq(1,numoutc-1))),
-                              command = c(
-                              # header
-            " & \\multicolumn{4}{c}{\\textbf{Males}} & \\multicolumn{4}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-5}  \\cmidrule(lr){6-9}
-& Mean & \\multicolumn{3}{c}{Coefficients} & Mean & \\multicolumn{3}{c}{Coefficients} \\\\ \\cmidrule(lr){3-5}  \\cmidrule(lr){7-9}
-            & (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8) \\\\",
-            rep(" \\midrule", numoutc-1)
-            )),
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabular",
-             floating = FALSE
-)
-}
-@
-
-\begin{table}
-\centering\footnotesize
-\caption{Predictors of adolescent outcomes -- BCS}\label{tab:bcs_adol}
-<<REGS_OUTC_BCS1, results='asis', dependson="REGS_OUTC">>=
-
-# collapse to matrix
-bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab[c("smktry16", "bmi16", "thn16", "owt16")],matrix,ncol=9,byrow=FALSE))
-table <- bcs_outctab_mat
-printoutctab(table, 4)
-
-@
 \justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of cohort members' adolescent and adult outcomes on their externalising and internalising socioemotional skills at five years of age.
@@
 \centering\footnotesize
-\caption{Predictors of adult outcomes -- BCS}\label{tab:bcs_adult}
-<<REGS_OUTC_BCS2, results='asis', dependson="REGS_OUTC">>=
+\caption{Predictors of adult behaviours, BCS}\label{tab:bcs_adultapp}
+<<REGS_OUTC_BCS_APP, results='asis', dependson="REGS_OUTC">>=
 
 # collapse to matrix
-bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab[c("hnvq30", "lginc34", "lhgrpay38")],matrix,ncol=9,byrow=FALSE))
-table <- bcs_outctab_mat
-printoutctab(table, 3)
+bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab[c("smoke42", "bmi42")],matrix,ncol=9,byrow=FALSE))
+printoutctab_raw(bcs_outctab_mat, 2)
 
@@
 
-\begin{table}
-\centering\footnotesize
-\caption{Predictors of adult outcomes -- BCS}\label{tab:bcs_adult2}
-<<REGS_OUTC_BCS3, results='asis', dependson="REGS_OUTC">>=
-
-# collapse to matrix
-bcs_outctab_mat <- do.call(rbind,lapply(bcs_outctab[c("smoke38", "smoke42", "obs42")],matrix,ncol=9,byrow=FALSE))
-table <- bcs_outctab_mat
-printoutctab(table, 3)
-
-@
-\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of cohort members' adolescent and adult outcomes on their externalising and internalising socioemotional skills at five years of age.
-Col.
-(1) shows the mean of the outcome for males.
-Col.
-(2) regresses the outcome on the scores obtained from the factor model in \autoref{sec:methods}.
-Col.
-(3) additionally controls for cognitive ability at age five.
-This is a simple factor score obtained by aggregating the available cognitive measures.
-Col.
-(4) uses sum scores (see \autoref{fig:sumscores}) instead of factor scores.
-All standard errors in parentheses are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression.
-Col.
-(5) to (8) repeat for female cohort members.
-All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
-See \autoref{tab:harmvar} for a description of the variables used.
-\par}
-\end{table}
-
-\begin{table}
-\centering\footnotesize
-\caption{Predictors of adolescent outcomes -- MCS}\label{tab:mcs_adol}
-<<REGS_OUTC_MCS, results='asis', dependson="REGS_OUTC">>=
-
-# collapse to matrix
-mcs_outctab_mat <- do.call(rbind,lapply(mcs_outctab[c("smktry14", "bmi14", "thn14", "owt14")],matrix,ncol=9,byrow=FALSE))
-table <- mcs_outctab_mat
-printoutctab(table, 4)
-
-@
-\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of cohort members' adolescent and adult outcomes on their externalising and internalising socioemotional skills at five years of age.
-Col.
-(1) shows the mean of the outcome for males.
-Col.
-(2) regresses the outcome on the scores obtained from the factor model in \autoref{sec:methods}.
-Col.
-(3) additionally controls for cognitive ability at age five.
-This is a simple factor score obtained by aggregating the available cognitive measures.
-Col.
-(4) uses sum scores (see \autoref{fig:sumscores}) instead of factor scores.
-All standard errors in parentheses are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression.
-Col.
-(5) to (8) repeat for female cohort members.
-All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
-See \autoref{tab:harmvar} for a description of the variables used.
-\par}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}
-<<FACDENS, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
-@
-\caption{Distribution of factor scores}\label{fig:dens}
-\justify {\scriptsize \emph{Notes}: The figure shows the distribution of the externalising and internalising socioemotional skills scores at age five obtained from the factor model, by gender and cohort.
-The scores are estimated from parameter estimates in \autoref{tab:measpars}, using an Empirical Bayes Modal approach.
-Higher scores correspond to \emph{better} skills.
-The distribution is estimated nonparametrically, using an Epanechnikov kernel.
-The figure also reports the $p$ value from Kolmogorov-Smirnov tests of equality between the distribution in BCS and MCS.
-\par}
-\end{figure}
-
-\begin{figure}
-<<RAWHIST, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% Appendix figures
+\clearpage
+\section{Appendix figures}
+
+\begin{figure}[ht]
+<<RAWHIST, dependson="SELECTDATA", dev='tikz', fig.height=9, fig.width=10, out.width='\\linewidth'>>=
 @
@@
 \begin{figure}
-<<FACINEQ_FSC, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
-fi <- fi_fsc # select result
-gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
-                        fi$ineqlist[[3]],fi$ineqlist[[4]],
-                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
-                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
-@
-\caption{Skill inequality by father's occupation}
-\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socioemotional skills scores by gender, cohort, and father's occupation at age five.
-Father's occupation is based on Registrar General's social class, with classes I to III Non Manual being `White collar' and classes III Manual to V (plus `other') being `Blue collar'.
-`No father figure' is defined as absence of a male figure living in the household.
-The four panels on top present mean and 95\% confidence intervals.
-Given that we cannot compare means of skills, all scores are normalised to take value zero for the `Blue collar' category, so that the gradient is emphasised.
-The bottom two panels present the unconditional distribution of father's occupation.
+<<ITEMINEQ_PS, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=11, out.width='\\linewidth'>>=
+itempl_ps
+@
+\caption{Item-level inequality by mother's education}\label{fig:ineqips}
+\justify {\scriptsize \emph{Notes}: The graph displays the ratio between the prevalence of each item in our scale in children of educated vs uneducated mothers, by cohort and gender.
+All items that have three categories in the scale have been dichotomised.
+For example, if the prevalence of the `Restless' behaviours among children of mothers with compulsory schooling in the BCS cohort is 7.5\%, and 5\% among mothers with post-compulsory schooling, the ratio will be 1.5.
+The error bars represent 95\% confidence intervals.
 \par}
@@
 \begin{figure}
-<<FACINEQ_PS, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
-fi <- fi_ps # select result
-gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
-                        fi$ineqlist[[3]],fi$ineqlist[[4]],
-                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
-                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
-@
-\caption{Skill inequality by mother's education}
-\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socioemotional skills scores by gender, cohort, and mother's education at age five.
-Mother's education is a dummy for whether the mother continued schooling past the minimum leaving age, based on her date of birth.
-The four panels on top present mean and 95\% confidence intervals.
-Given that we cannot compare means of skills, all scores are normalised to take value zero for the `Compulsory' category, so that the gradient is emphasised.
-The bottom two panels present the unconditional distribution of mother's education.
-\par}
-\end{figure}
-
-\begin{figure}
-<<FACINEQ_SM, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
-fi <- fi_sm # select result
-gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
-                        fi$ineqlist[[3]],fi$ineqlist[[4]],
-                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
-                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
-@
-\caption{Skill inequality by mother's pregnancy smoking}
-\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socioemotional skills scores by gender, cohort, and mother's pregnancy smoking.
-Mother's education is a dummy for whether the mother reported smoking during pregnancy.
-The four panels on top present mean and 95\% confidence intervals.
-Given that we cannot compare means of skills, all scores are normalised to take value zero for the `Non-smoker' category, so that the gradient is emphasised.
-The bottom two panels present the unconditional distribution of mother's pregnancy smoking
+<<ITEMINEQ_SM, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=11, out.width='\\linewidth'>>=
+itempl_sm
+@
+\caption{Item-level inequality by mother's pregnancy smoking}\label{fig:ineqism}
+\justify {\scriptsize \emph{Notes}: The graph displays the ratio between the prevalence of each item in our scale in children of mothers who smoked in pregnancy vs non-smokers, by cohort and gender.
+All items that have three categories in the scale have been dichotomised.
+For example, if the prevalence of the `Restless' behaviours among children of smoker mothers in the BCS cohort is 7.5\%, and 5\% among non-smoker mothers, the ratio will be 1.5.
+The error bars represent 95\% confidence intervals.
 \par}
@@
 @
-\caption{Item-level inequality by father's occupation}
+\caption{Item-level inequality by father's occupation}\label{fig:ineqifsc}
 \justify {\scriptsize \emph{Notes}: The graph displays the ratio between the prevalence of each item in our scale in children of white collar vs blue collar fathers, by cohort and gender.
@@
 
-\begin{figure}
-<<ITEMINEQ_PS, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=11, out.width='\\linewidth'>>=
-itempl_ps
-@
-\caption{Item-level inequality by mother's education}
-\justify {\scriptsize \emph{Notes}: The graph displays the ratio between the prevalence of each item in our scale in children of educated vs uneducated mothers, by cohort and gender.
-All items that have three categories in the scale have been dichotomised.
-For example, if the prevalence of the `Restless' behaviours among children of mothers with compulsory schooling in the BCS cohort is 7.5\%, and 5\% among mothers with post-compulsory schooling, the ratio will be 1.5.
-The error bars represent 95\% confidence intervals.
-\par}
-\end{figure}
-
-\begin{figure}
-<<ITEMINEQ_SM, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=11, out.width='\\linewidth'>>=
-itempl_sm
-@
-\caption{Item-level inequality by mother's pregnancy smoking}
-\justify {\scriptsize \emph{Notes}: The graph displays the ratio between the prevalence of each item in our scale in children of mothers who smoked in pregnancy vs non-smokers, by cohort and gender.
-All items that have three categories in the scale have been dichotomised.
-For example, if the prevalence of the `Restless' behaviours among children of smoker mothers in the BCS cohort is 7.5\%, and 5\% among non-smoker mothers, the ratio will be 1.5.
-The error bars represent 95\% confidence intervals.
-\par}
-\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\end{appendices}
 

### 2018-09-28 codes/R/1_load.R

@@ -18,7 +18,6 @@ library(tikzDevice)
 library(gtools)
 library(here)
 library(rio)
-library(dummies)
 library(dplyr)
 library(tidyr)
 library(forcats)
@@ -26,6 +25,7 @@ library(ggplot2)
 library(grid)
 library(gridExtra)
 library(ggpubr)
+library(kableExtra)
 
 
 dir_data  <- here("rdata/")
@@ -178,9 +178,12 @@ cohdata <- cohdata[complete.cases(cohdata[,c(grep("X[0-9]", names(X.all), value=
 # add raw scores
 cohdata$EXT_RAW <- rowSums(apply(cohdata[,paste("X", seq(1,6), sep="")], 2, function(x) as.numeric(x)), na.rm = T)
 cohdata$INT_RAW <- rowSums(apply(cohdata[,paste("X", seq(7,11), sep="")], 2, function(x) as.numeric(x)), na.rm = T)
-cohdata$EXT_RAWr <- residuals(lm(EXT_RAW ~ age, data=cohdata, na.action = na.exclude))
+cohdata$EXT_RAWr <- residuals(lm(EXT_RAW ~ age, data=cohdata, na.action = na.exclude)) # residualised
 cohdata$INT_RAWr <- residuals(lm(INT_RAW ~ age, data=cohdata, na.action = na.exclude))
 
+# standardise raw scores using BCS M
+cohdata$EXT_RAWs <- (cohdata$EXT_RAW - mean(cohdata[cohdata$cohortsex=="BCS.M", 'EXT_RAW']))/sd(cohdata[cohdata$cohortsex=="BCS.M", 'EXT_RAW'])
+cohdata$INT_RAWs <- (cohdata$INT_RAW - mean(cohdata[cohdata$cohortsex=="BCS.M", 'INT_RAW']))/sd(cohdata[cohdata$cohortsex=="BCS.M", 'INT_RAW'])
 
 # FINAL CLEANING/RECODING
 

### 2018-09-30 codes/R/paper/paper_jpube.Rnw
add summary stats table

@@
 read_chunk(here("codes/R/5_selectvars.R"))
+read_chunk(here("codes/R/5_sumstats.R"))
 read_chunk(here("codes/R/6_graphs.R"))
@@
 # load outcomes data
+@
+
+<<SUMSTATS, dependson="SELECTDATA">>=
+# make table of summary stats
 @
@@
 # eliminate raw rows and columns
-outcs <- c("hnvq34", "emp34", "lgpay34", "emp42", "lgpay42")
+outcs <- c("hnvq34", "lgpay34", "emp42", "lgpay42")
 outctabs <- list()
@@
 require(xtable)
-table <- meantab # females
+table <- meantab
 table <- table[,!names(table) %in% c("bcs.text", "bcs.num", "mcs.text", "mcs.num")] # drop if unneeded
@@
 
+\clearpage
+\begin{table}[]
+\caption{Summary statistics}\label{tab:sumstats}
+\centering\setstretch{1.2} \footnotesize
+<<SUMSTATS_TAB, dependson="SUMSTATS", results='asis'>>=
+require(xtable)
+table <- summtab 
+table <- table[,!names(table) %in% c("bcs.text", "bcs.num", "mcs.text", "mcs.num")] # drop if unneeded
+
+  print(xtable(table,
+               align="L{0cm}L{.3\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}",
+               digits = 1
+  ), 
+  include.rownames=F,
+  include.colnames=F,
+  booktabs = T,
+  add.to.row = list(pos = c(list(0)),
+                    command = c(
+                      # header
+                      paste(" & 
+                            \\multicolumn{2}{c}{\\textbf{Males}} & \\multicolumn{2}{c}{\\textbf{Females}}
+                            \\\\ \\cmidrule(lr){2-3} \\cmidrule(lr){4-5} 
+                            & BCS & MCS & BCS & MCS \\\\
+                            "
+                            , sep="")
+                    )),
+  sanitize.text.function = function(x){x},
+  sanitize.rownames.function = function(x){""}, # to not print row numbers
+  tabular.environment = "tabular",
+  floating = FALSE
+  )
+
+@
+\justify {\scriptsize \emph{Notes}: The table shows mean (standard deviation) of harmonised variables used in the analysis.
+\par}
+\end{table}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-09-30 codes/R/8_regs_outc.R

@@ -107,10 +107,10 @@ bcs_outctab <- list()
 for (i in 1:length(bcsoutcvars)) {
   lab <- paste0("\\textit{\\textbf{", bcsoutclabs[i], "}}")
   bcs_outctab[[bcsoutcvars[i]]] <- matrix(c(lab, dvmeans_bcs_m[i], "", "", "", dvmeans_bcs_f[i], "", "", "",   # outcome and means
-                                            "Externalising skills", # row label
+                                            "Externalising skills (5)", # row label
                                             "", cellpr(bcsoutcsum$M[[i]]$b, "EXT"), cellpr(bcsoutcsum$M[[i]]$bc, "EXT"), "", # males EXT
                                             "", cellpr(bcsoutcsum$F[[i]]$b, "EXT"), cellpr(bcsoutcsum$M[[i]]$bc, "EXT"), "", # females EXT
-                                            "Internalising skills", # row label
+                                            "Internalising skills (5)", # row label
                                             "", cellpr(bcsoutcsum$M[[i]]$b, "INT"), cellpr(bcsoutcsum$M[[i]]$bc, "INT"), "", # males EXT
                                             "", cellpr(bcsoutcsum$F[[i]]$b, "INT"), cellpr(bcsoutcsum$F[[i]]$bc, "INT"), "", # females EXT
                                             "Externalising (sum score)", # row label
@@ -119,7 +119,7 @@ for (i in 1:length(bcsoutcvars)) {
                                             "Internalising (sum score)", # row label
                                             "", "", "", cellpr(bcsoutcsum$M[[i]]$bcr, "INT_RAWs"), # males EXT
                                             "", "", "", cellpr(bcsoutcsum$F[[i]]$bcr, "INT_RAWs"), # females EXT
-                                            "Cognitive skills", # row label
+                                            "Cognitive skills (5)", # row label
                                             "", "", cellpr(bcsoutcsum$M[[i]]$bc, "cogscore"), cellpr(bcsoutcsum$M[[i]]$bcr, "cogscore"), # males COG
                                             "", "", cellpr(bcsoutcsum$F[[i]]$bc, "cogscore"), cellpr(bcsoutcsum$F[[i]]$bcr, "cogscore"), # females COG
                                             "\\newline Adj. R$^2$", # row label
@@ -142,10 +142,10 @@ mcs_outctab <- list()
 for (i in 1:length(mcsoutcvars)) {
   lab <- paste0("\\textit{\\textbf{", mcsoutclabs[i], "}}")
[…]

### 2018-09-30 codes/R/99_auxfun.R

@@ -197,7 +197,7 @@ cellpr <- function(estimtab, vr) {
 
 # print R2 (with padding)
 prr2 <- function(mod) {
-  return(paste0("\\newline ", sprintf("%.4f", summary(mod)$adj.r.squared)))
+  return(paste0("\\newline ", sprintf("%.3f", summary(mod)$adj.r.squared)))
 }
 
 # function to print pvals as strings in square brackets

### 2018-10-01 codes/R/paper/paper_jpube.Rnw
add text from coauthors

@@
 \author{Orazio Attanasio \hspace{10pt} Richard Blundell \\  Gabriella Conti \hspace{10pt} Giacomo Mason}
-\affil{\emph{Institute for Fiscal Studies} \\ \emph{UCL Department of Economics}}
+\affil{\emph{Institute for Fiscal Studies} \\ \emph{Department of Economics, University College London}}
 \date{\today}
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
   \begin{abstract}
-    Abstract here
+  We examine changes in inequality in socio-emotional skills very early in life in two British cohorts born 30 years apart.
+We construct socio-emotional scales comparable across cohorts for both boys and girls, using two validated instruments for the measurement of child behaviour.
+We identify two dimensions of socio-emotional skills for each cohort: `internalising' and `externalising', related to the ability of children to focus their concentration and to engage in interpersonal activities, respectively.
+Using recent methodological advances in factor analysis, we establish comparability in the inequality of these early skills across cohorts, but not in their average level.
+We document for the first time that inequality in these early skills has increased across cohorts, especially for boys and at the bottom of the distribution.
+We also document changes in conditional skills gaps across cohorts.
+We find an increase in the socio-emotional skills gap in the younger cohort for children born to mothers with higher socio-economic status (education and employment), and to mothers who smoked during pregnancy.
+The increase in inequality in early socio-emotional skills is particularly pronounced for boys.
+On the other hand, we find a decline in the skills gradient for children without a father figure in the household.
+Lastly, we document that socio-emotional skills measured at a much earlier age than in most of the existing literature are significant predictors of outcomes both in adolescence and adulthood, in particular health and health behaviours.
+Our results show the importance of formally testing comparability of measurements to study skills differences across groups, and in general point to the role of inequalities in the early years for the accumulation of health and human capital across the life course.
+
+  	Keywords: Inequality, Socio-emotional skills, Cohort studies, Measurement invariance
+
+	JEL Classification: J13, J24, I14, I24, C38
   \end{abstract}
-
-\vspace{30pt}
-{\footnotesize 
-We thank {\color{orange} participants to the Trans Atlantic Public Economics Seminar }
-We are grateful to The Centre for Longitudinal Studies, UCL Institute of Education for the use of the data and to the UK Data Archive and UK Data Service for making them available.
+\vspace{20pt}
+
+\thanks{
+\footnotesize{Corresponding Author: Orazio Attanasio, Department of Economics, University College London,
+Gordon Street, London WC1H0AX, UK.
+Email: o.attanasio@ucl.ac.uk.
+We thank participants to the NBER/LSE Trans-Atlantic Public Economics Seminar (TAPES) and our two discussants Hilary Hoynes and Gabriel Ulyssea for excellent comments.
+We are grateful to The Centre for Longitudinal Studies, UCL Institute of Education, for the collection of the data, and to the UK Data Archive and UK Data Service for making them available.
 However, they bear no responsibility for the analysis or interpretation of the data.
-Giacomo Mason obtained funding thanks to an Economic and Social Research Council graduate scholarship.
-}
+Giacomo Mason is funded by an Economics and Social Research Council scholarship.}}
 \end{titlepage}
 
-\input{/Users/giacomomason/Dropbox/cohorts_shared/paper_jpube_1intro.tex}
-\input{/Users/giacomomason/Dropbox/cohorts_shared/paper_jpube_2datamethods.tex}
-\input{/Users/giacomomason/Dropbox/cohorts_shared/paper_jpube_3resultsconclusion.tex}
+\input{text/paper_jpube_1intro.tex}
+\input{text/paper_jpube_1blit.tex}
+\input{text/paper_jpube_2datamethods.tex}
+\input{text/paper_jpube_3resultsconclusion.tex}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}[ht]
+\caption{Subscale of comparable items}\label{tab:scalecomp}
+\centering\setstretch{1.2}
+\begingroup\footnotesize
+\begin{tabular}{C{.03\textwidth}C{.05\textwidth}C{.04\textwidth}L{.12\textwidth}L{.3\textwidth}L{.3\textwidth}}
+  \toprule
+ Itm. & Factor & Cat. & Title & Rutter Wording (BCS 1970) & SDQ Wording (MCS 2000/1) \\ \midrule
+ 1 & EXT & 3 & \textit{Restless} & Very restless. Often running about or jumping up and down. Hardly ever still & Restless, overactive, cannot stay still for long   \\
+   2 & EXT & 3 & \textit{Squirmy/fidgety} & Is squirmy or fidgety & Constantly fidgeting or squirming  \\
+   3 & EXT & 3 & \textit{Fights/bullies} & Frequently fights other children + \newline Bullies other children & Often fights with other children or bullies them  \\
+   4 & EXT & 3 & \textit{Distracted} & Cannot settle to anything for more than a few moments & Easily distracted, concentration wanders  \\
+   5 & EXT & 2 & \textit{Tantrums} & Has temper tantrums & Often has temper tantrums or hot tempers \\
+   6 & EXT & 2 & \textit{Disobedient} & Is often disobedient & (+) Generally obedient, usually does what adults request \\
+   7 & INT & 3 & \textit{Worried} & Often worried, worries about many things  & Many worries, often seems worried \\
+   8 & INT & 3 & \textit{Fearful} & Tends to be fearful or afraid of new things or new situations & Nervous or clingy in new situations, easily loses confidence  \\
+   9 & INT & 3 & \textit{Solitary} & Tends to do things on his/her own, rather solitary & Rather solitary, tends to play alone  \\
+  10 & INT & 3 & \textit{Unhappy} & Often appears miserable, unhappy, tearful or distressed & Often unhappy, down-hearted or tearful \\
+  11 & INT & 2 & \textit{Aches} & Complains of headaches + \newline Complains of stomach-ache or has vomited & Often complains of head- aches, stomach-ache or sickness \\
+   \bottomrule
+\end{tabular}
+\endgroup
+\justify {\scriptsize \emph{Notes}: \emph{Itm.} is item number.
+\emph{Factor} is the latent construct to which the item loads -- EXT is Externalising skills, INT is Internalising skills.
+\emph{Cat.} is the number of categories in which the item is coded -- 2 denotes a binary item (applies/does not apply) and 3 denotes a 3-category item.
+\emph{Title} is a short label for the item.
+\emph{Wording} columns show the actual wording in the scales used in each of the cohort studies.
+Items denoted by (+) are positively coded in the original scale.
+\par}
+\end{table}
 
@@
 # eliminate raw rows and columns
-outcs <- c("hnvq34", "lgpay34", "emp42", "lgpay42")
+outcs <- c("hnvq34", "emp42", "lgpay42")
 outctabs <- list()
@@
 % Appendix text
-\input{/Users/giacomomason/Dropbox/cohorts_shared/paper_jpube_4appendix.tex}
+\input{text/paper_jpube_4appendix.tex}
 
@@
 % from tex file
-\input{/Users/giacomomason/Dropbox/cohorts_shared/paper_jpube_5apptables.tex}
+\input{text/paper_jpube_5apptables.tex}
 
@@
 \begin{table}
-\centering\scriptsize
-\caption{Predictors of adolescent outcomes, with sum scores}\label{tab:adol_app}
-\resizebox{15cm}{!}{
-<<REGS_OUTC_ADOL_APP, results='asis', dependson="REGS_OUTC">>=
-
-# collect adolescent outcomes
-outctabs <- list(bcs_outctab$smktry16, bcs_outctab$bmi16, mcs_outctab$smktry14, mcs_outctab$bmi14)
-# collapse to matrix
-outctabs_mat <- do.call(rbind,lapply(outctabs,matrix,ncol=9,byrow=FALSE))
-printoutctab_raw(outctabs_mat, length(outctabs))
-
-@
-}
-\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of cohort members' adolescent and adult outcomes on their externalising and internalising socioemotional skills at five years of age.
-Col.
-(1) shows the mean of the outcome for males.
-Col.
-(2) regresses the outcome on the scores obtained from the factor model in \autoref{sec:methods}.
-Col.
-(3) additionally controls for cognitive ability at age five.
-This is a simple factor score obtained by aggregating the available cognitive measures.
-Col.
-(4) uses sum scores (see \autoref{fig:sumscores}) instead of factor scores.
-All standard errors in parentheses are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression.
-Col.
-(5) to (8) repeat for female cohort members.
-All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
-See \autoref{tab:harmvar} for a description of the variables used.
-\par}
-\end{table}
-
-\begin{table}
 \centering\footnotesize

### 2018-10-01 codes/R/paper/text/paper_jpube_1blit.tex
add text from coauthors

\section{Literature \label{sec:lit}}

The importance of cognition in predicting life course success is well established in the economics literature. However, in recent years the role played by `non-cognitive' traits is being increasingly investigated. These traits include constructs as different as psychological and preference parameters such as social and emotional skills, locus of control and self-esteem, personality traits (e.g. conscientiousness), and risk aversion and time preferences. Given the vastity of this literature, we briefly review below the main papers on the determinants and consequences of socio-emotional traits which are more directly related to our work, and we refer to other sources for more exhaustive reviews \citep{borghans2008economics,almlund2011personality,goodman2015social,kautz2014fostering}.

\paragraph{Consequences of socio-emotional traits} One of the first papers to show the importance of non-cognitive personality variables for wages was \citet{bowles2001determinants}. \citet{heckman2006effects} suggested that non-cognitive skills are at least as important as cognitive abilities in determining a variety of adults outcomes. \citet{lindqvist2011labor}, using data based on personal interviews conducted by a psychologist during the Swedish military enlistment exam, show that both cognitive and noncognitive abilities are important in the labour market, but for different outcomes: low noncognitive abilities are more correlated with unemployment or low earnings, while cognitive ability is a stronger predictor of wages for skilled workers. \citet{segal2013misbehavior}, using data on young men from the US National Education Longitudinal Survey, shows that eight-grade misbehaviour is important for earnings over and above eight-grade test scores. \citet{layard2014predicts} find that childhood emotional health (operationalised using the same mother-reported Rutter scale we use in the 1970 British cohort study) at ages 5, 10 and 16 is the most important predictor of adulthood life satisfaction and life course success.

There are only few studies in economics specifically studying ``non-cognitive'' traits and health behaviors. \citet{conti2010education} and \citet{conti2011early} were the first to consider three early endowments, including child socio-emotional traits and health in addition to cognition, using rich data from the 1970 British cohort study. They find strong evidence that non-cognitive traits promote health outcomes and healthy behaviors, and than not accounting for them overestimates the effects of cognition; additionally, they document that child cognitive traits are more important predictors of employment and wages than socio-emotional traits or early health. \citet{chiteji2010time} used the US Panel Study of Income Dynamics (PSID) and found that future orientation and self-efficacy (related to emotional stability) are associated with less alcohol consumption and more exercise. \citet{cobb2014healthy} used the Australian HILDA data and found that an internal locus of control (also related to emotional stability, perceived control over one's life) is related to better health behaviours (diet, exercise, alcohol consumption and smoking). \citet{mendolia2014effect} used the Longitudinal Study of Young People in England and found that individuals with external locus of control, low self-esteem, and low levels of work ethics, are more likely to engage in risky health behaviours. \citet{savelyev2017socioemotional} show that the association between personality traits and health behaviours also holds in a high-IQ sample (the Terman Sample). \citet{heckmanJPE2018} use, instead, early risky behaviours to measure socio-emotional traits, and confirm their predictive power for health behaviours and health outcomes.

Very few papers attempt to make cross-cohorts comparisons about the importance of socio-emotional skills. \citet{blanden2007accounting} -- one of the closes study to ours -- examine cognitive skills, non‐cognitive traits, educational attainment and labour market attachment as mediators of the decline in inter-generational mobility in UK between the 1958 and the 1970 cohorts. The authors take great care in selecting non-cognitive items to be as comparable as possible across cohorts, from the Rutter scale at age 10 for the 1970 cohort and from the Bristol Social Adjustment Guide for the 1958 cohort; however, they do not carry out a formal test of measurement invariance and the do not construct factor scores fully comparable across cohorts as we do. Another paper related to ours is the one by \citet{reardon2016recent}, who study recent trends in income, racial, and ethnic school gaps in several dimensions of school readiness, including academic achievement, self-control, and externalizing behavior, at
kindergarten entry, using comparable data from the Early Childhood Longitudinal Studies (ECLS-K and ECLS-B) for cohorts born from the
early 1990s to the 2000–2010 period in the US. They find that readiness gaps narrowed modestly from 1998 to 2010, particularly between high- and low-income students and between White and Hispanic students. Lastly, \citet{deming2017growing} uses a comparable set of skill measures and covariates across survey waves for the NLSY79 and the NLSY97, and finds that the labour market return to social skills was much greater in the 2000s than in the mid-1980s and 1990s.

\paragraph{Determinants of socio-emotional traits} Equally flourishing has been the literature on the determinants of child socio-emotional skills, which ranges from reduced-form, correlational or causal estimates, to more structural approaches. One of the first papers by \citet{segal2008classroom} has shown that a variety of family and school characteristics predict classroom behaviour. \citet{carneiro2013maternal} study the intergenerational impacts of maternal education, using data from the NLSY79 and an instrumental variable strategy; they find strong effects in terms of reduction in children's behavioural problems. \citet{cunha2010estimating} and \citet{attanasio2018estimating} both estimate production functions for child cognitive and socio-emotional development (in US and Colombia, respectively), and find an important role played by parental investments.

Interventions for improving Social and Emotional Learning (SEL) in a school setting have shown significant improvements in socio-emotional skills, attitudes, behaviours, and academic performance \citep{durlak2011impact}, and a substantial positive return on investments \citep{belfield2015economic}; after-school programs have been shown to be equally effective \citep{durlak2010meta}.

Additionally, it has been shown that a key mechanism through which early childhood interventions improve adult socioeconomic and health outcomes is by boosting socio-emotional skills, such as four teacher-reported behavioural outcomes in the project STAR\footnote{Student's effort, initiative, non-participatory  behavior, and how the student is seen to `value' the class.} \citep{chetty2011does}, reductions in externalising behaviour (from the Pupil Behavior Inventory) at ages 7-9 in the Perry Preschool Project \citep{heckman2013understanding,conti2016effects}, or improvements in task orientation at ages 1-2 in the Abecedarian Project \citep{conti2016effects}.

In sum, even if the literature on the determinants and consequences of socio-emotional skills has been booming, most papers use skills measured in late childhood or in adolescence; and no paper in economics formally tests for invariance of measurements across different groups and constructs fully comparable scores. In this paper, we use measures of child socio-emotional development at age 5, hence before formal schooling starts; and we construct comparable scales across the two cohorts we study (the 1970 and the 2000 British cohorts), so that we can investigate changes in inequality in early development, their determinants, and consequences, in a parallel fashion.

### 2018-10-01 codes/R/paper/text/paper_jpube_1intro.tex
add text from coauthors

\section{Introduction}

Human capital is a key determinant of economic growth and performance and of the resources an individual creates and controls over the life cycle. Human capital is also important for various determinants of individual well-being, ranging from health to life satisfaction.  In recent years, the process of human capital accumulation has received considerable attention. There is growing consensus on the fact that human capital is a multidimensional object, with different domains playing different roles in labour market as well as in the determination of other outcomes, including the process of human development. It is also recognised that human capital is the output of a very persistent process, where early years inputs play an important and persistent role.

And yet, there are still large gaps in our knowledge of the process of human capital development. These gaps are partly driven by the scarcity of high quality longitudinal data measuring the evolution over the life cycle of different dimension of human capital. Moreover, there is a lack of consensus on the best measures and on the tools to collect high quality data. As a consequence, even when data are available in different contexts, their comparability is problematic.

In this paper, we focus on an important dimension of human capital, which, so far, has received limited attention: socio-emotional skills. Evidence has shown that gaps in socio-emotional skills emerge at very young ages, and that in the absence of interventions are very persistent across the life cycle \citep{cunha2006interpreting}. However, there is surprisingly little evidence on how inequality in this important dimension of human capital has changed across cohorts. In this paper, we start addressing this gap and focus on the measurement of these skills in two British cohorts: the one of children born in 1970 and the one of children born in 2000. We consider the measurement of socio-emotional skills during early childhood, as these skills have been shown, in a variety of contexts \citep{almlund2011personality} to have important long-run effects. Our goal is to characterise the distribution of socio-emotional skills in these cohorts and compare them. In the last part of the paper, we also consider the predictive power of different socio-emotional skills for health and socioeconomic outcomes.

The main contributions of the paper are four. First, we use two validated scales of childhood behavioural traits and select those items which are comparable across the two cohorts. By performing an exploratory factor analysis, we determine that we need at least two dimensions to characterise socio-emotional skills. We label them as `internalising' and `externalising' skills, the former relating to the ability of children to focus their drive and determination and the latter relating to their ability in engaging in interpersonal activities.

Second, we study the comparability of the measures in the two cohorts. In particular, we test for \emph{measurement invariance} of the items we use to estimate the latent factors. Intuitively, if one assumes that a set measures is related to a latent unobserved factor of interest, one can think of this relationship as being driven by the saliency of each measure and the level. If one uses a given measure as the relevant metric for the relevant factor, its saliency will determine the scale of the factor, while some other parameters, which could be driven by the difficulty of a given test or the social norms and attitudes towards a certain type of behaviour, determine the \emph{average level} of the factor. Comparability of estimated factors across different groups (such as different cohorts) assumes  that both the parameters that determine the saliency of a given set of measures and the level of the factors do not vary across groups. We find that, for the measures we use and for both factors, we cannot reject measurement invariance for the saliency parameters. However, we strongly reject measurement invariance for the level parameters. These results imply that while the level of inequality across the two cohorts in the skills we consider is comparable, we cannot determine whether the \emph{average levels} of the two factors are larger or smaller in one of the two cohorts.

Third, given the results we obtain on measurement invariance, we proceed to compare the level of inequality in the two types of socio-emotional skills across the two cohorts, for both boys and girls. We find that the most recent cohort is more unequal in both dimensions of socio-emotional skills than the 1970 cohort. This result is particularly apparent for boys, and when looking at differences by maternal background. Fourth, we study whether the socio-emotional skills we observe at a young age are an important determinant of a variety of adolescent (and adult, for the older BCS cohort) outcomes. We find that socio-emotional skills at age five are more predictive than cognitive skills for unhealthy behaviours like smoking and measures of health capital such as body mass index. The effect of cognition, instead, dominates for educational and labour market outcomes.

The rest of the paper is organised as follows. In section \ref{sec:data}, we briefly discuss the data we use in the analysis. In section \ref{sec:methods}, we discuss the methods we use to identify the number of dimensions in socio-emotional skills and how we estimate the latent factors that represent them. In section \ref{sec:measinv}, we discuss the comparability of factors estimated with a given set of measures from different groups and the \emph{measurement invariance} tests we use. Section \ref{sec:results} reports our empirical results, while section \ref{sec:conclusions} concludes the paper.

### 2018-10-01 codes/R/paper/text/paper_jpube_2datamethods.tex
add text from coauthors

\section{Data \label{sec:data}}

We use information from two nationally representative longitudinal studies in the UK. The studies follow the lives of children born approximately 30 years apart: the British Cohort Study (BCS) surveys individuals born in 1970, and the Millennium Cohort Study (MCS) includes births between 2000 and 2002. The British Cohort Study includes all individuals born across Great Britain in a single week (April 4-11, 1970). Cohort members' families -- and subsequently members themselves -- were surveyed on multiple occasions. For this paper we augment the information at the five-year survey with data from birth, adolescence (16), and adulthood (30, 38, 42). The Millennium Cohort Study follows individuals born in the UK between September 2000 and January 2002. We use the birth survey, and the sweeps at around 5, 11, and 14 years.\footnote{All data is publicly available at the UK Data Service \citep{Chamberlain2013,Butler2016a,Butler2016b,Butler2017,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2016b,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2016,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2016a,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2017b,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2017,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2017a,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2017c}.}

Our main focus is on socio-emotional skills of children around age five. We take advantage of the longitudinal nature of the cohorts by merging information from surveys before and after age five. From the birth survey, we include information on gestational age and weight at birth, previous stillbirths, parity, maternal smoking in pregnancy, maternal age, height, and marital status. From the five year survey, we extract maternal education, employment status, and the father's occupation. All the above variables are transformed or recoded to maximise comparability between the two studies. Furthermore, we add some adolescent risky behaviours such as smoking and BMI, with the caveat that these are surveyed at 16 and 14 in BCS and MCS respectively. Finally, for the 1970 cohort we also include measures of adult educational attainment, BMI, and income. Variable definitions are available in \autoref{tab:harmvar}.

Ideally, we would compare socio-emotional skills alongside cognitive skills. However, the cognitive tests administered to each cohort have no overlap, even at the item level. We thus use the available cognitive tests in each cohort to estimate simple confirmatory factor model with a single latent dimension, separately by cohort (see \autoref{tab:harmvar} for the tests used). Unlike the other indicators in our analysis, cognitive skills are thus not comparable across cohort.

Another complication arises from the fact that, differently from the British Cohort Study, the Millennium Cohort Study has a stratified design. It oversamples children living in administrative areas characterised by higher socioeconomic deprivation and larger ethnic minority population \citep{Plewis2007}. We rebalance the MCS sample to make it nationally representative by excluding from the analysis a fraction of observations from the oversampled areas, proportionally to their sampling probability.\footnote{See Table 5.5 in \citet{Plewis2007}. This choice is mainly driven by software limitations. The \texttt{lavaan} package in \textbf{\textsf{R}} \citep{Rosseel2012} is the most suitable tool for our invariance analysis, but it does not allow to use weights when outcomes are categorical, as it is the case for the socio-emotional measurements.}
We also restrict our sample to individuals born in England. Finally, we restrict the sample to cases where the respondent in the five-year followup was the natural mother, and where there is complete information on socioemotional skills. The final sample contains 9,545 individuals from the British Cohort Study, and 5,436 from the Millennium Cohort Study.

\section{Dimensions of  socio-emotional skills \label{sec:methods}}

Child socio-emotional skills are an unobservable and difficult to measure construct. Over recent years, the measurement of such skills has evolved and, over time, different measures have been used.  As we discuss below, this makes the comparison  of socio emotional skills across different groups, assessed with different tools, difficult.

A common approach to infer a child's socio-emotional development is based on behavioural screening scales. As part of these tools, mothers (or teachers) indicate whether their children exhibit a series of behaviours -- the \emph{items} of the scale. In the British and Millennium Cohort Studies, two different scales were employed. In the BCS, the Rutter A Scale was used \citep{Rutter1970} while in the MCS cohort, mothers were  administered the Strengths and Difficulties Questionnaire (SDQ, \citealp{Goodman1994,Goodman1997}). The SDQ was created as an update to the Rutter scale. It encompasses more recent advances in child psychopathology, and emphasises positive traits alongside undesirable ones \citep{Stone2010}. The Rutter and SDQ scales are reproduced in \autoref{tab:scales}; they have 23 and 25 items each, respectively. In the child psychiatry and psychology literatures, the Rutter and SDQ behavioural screening scales are regarded as measures of behavioural problems and mental health. However, in our analysis we follow the economics literature, and - after having recoded them accordingly - we interpret them as measures of positive child development \citep{goodman2011population}.

While the Rutter and SDQ scales are similar in their components, there is no a priori reason to expect them to be directly comparable. First, the overlap of behaviours described in the two scales is only partial. Second, the wording of each item is slightly different, both in the description and in the options that can be selected as answers. Third, the different ordering of the items within each scale might lead to order effects. Fourth, and no less importantly, the interpretation of each behaviour by respondents living 30 years apart (1975 vs 2005/6) might differ due to a host of evolving societal norms.

As our goal is to compare socio-emotional skills across the two cohorts, we construct a new scale by retaining the items that are worded in a similar way across the two original Rutter and SDQ scales, and making some slight coding adjustments to maximise comparability. In what follows, we will consider the included items to be the same \emph{measure} in the two cohorts. The wording of the items we will be using in the analysis is presented in \autoref{tab:scalecomp}: we retain 13 items for the BCS (two of them are grouped) and 11 for the MCS with high degree of comparability.\footnote{We exclude from the analysis items that were completely different between the two questionnaires, although we could have included them in the factor analysis and treated them as missing in the cohort were they were not administered. While this could have improved efficiency, we decided to rely on a more coherent set of measures to maximise comparability between the two cohorts.}. Item-level prevalence by cohort and gender is in \autoref{tab:meantab}. We see that, in general, item prevalence is more similar across genders within the same cohort, than across cohorts. For the majority of items, there is a lower prevalence of problematic behaviours in the MCS than in the BCS; however, four items (distracted, tantrums, fearful, aches) show a higher prevalence in 2005 than in 1975. Regardless, a simple cross-cohort comparison of item-level prevalence is misleading because of changing perceptions and norms about what constitutes problematic behaviour in children. The analysis in section \ref{sec:measinv} tackles this issue.

In the remainder of this section, we analyse the properties of the new scale. In particular, we study the \emph{factor structure} of our scale. Namely, we establish how many latent dimensions of socio-emotional skills the scale is capturing, and which items are measuring which dimension. We then estimate the parameters of the factor models that corresponds to our choice of dimension and attribution of specific items to factors. In the following section, we investigate to what extent socio-emotional skills are measured in the same way across cohorts.

\subsection{Exploratory analysis \label{sec:ea}}

The original Rutter scale, used for the BCS cohort, distinguishes behaviours into two subscales: \emph{anti-social} and \emph{neurotic} \citep{Rutter1970}. This two-factor conceptualisation has been validated using data from multiple contexts, and the latent dimensions have been broadly identified as externalising and internalising behaviour problems.\footnote{See for example \cite{Fowler1979,Venables1983,Tremblay1987,Berglund1999,Klein2009}. However, in some cases a three-factor structure was found to better fit the data, with the externalising factor separating into two factors seemingly capturing aggressive and hyperactive behaviours \citep{Behar1974,McGee1985}.} The Strength and Difficulties Questionnaire, used for the MCS cohort, was conceived to have five subscales of five items each. The five subscales are: \emph{hyperactivity}, \emph{emotional symptoms}, \emph{conduct problems}, \emph{peer problems}, and \emph{prosocial}. This five-factor structure has been validated in many contexts \citep{Stone2010}; lower-dimensional structures have been also suggested \citep{Dickey2004}. Recent research has shown that there are some benefits to using broader subscales that correspond to the externalising and internalising factors in Rutter, especially in low-risk or general population samples \citep{Goodman2010}.

We use exploratory factor analysis (EFA) to assess the factors structure of our 11-item scale combining Rutter and SDQ.\footnote{Factor-analytic methods have long been used in psychology, and in recent years they have become increasingly popular in economics, especially to meaningfully aggregate high-dimensional items measuring different aspects of common underlying dimensions of human development. The EFA is performed decomposing the polychoric correlation matrix of the items and using weighted least squares, and the solution is rescaled using oblique factor rotation (\emph{oblimin}). We use the \textbf{\textsf{R}} package \texttt{psych}, version 1.8.4 \citep{Revelle2018}.} We start by investigating the number of latent constructs that are captured by the scale, using different methods developed in the psychometric literature, and now also used in the economics literature. The results are displayed in \autoref{tab:numfac}. As pointed out in \citet{conti2014bayesian}, there is relatively little agreement among procedures; this is the case especially for the Rutter items in the BCS data, where different methods suggest to retain between 1 and 3 factors, while most methods suggest to retain 2 factors in the MCS. In our analysis, we adopt two factors and a dedicated measurement system, where each measure reflects only one factor. This choice is justified both by the child psychology literature cited above, and as compromise to work with the same number of factors in the two cohorts. The two-factor EFA delivers a neat and sensible separation between items, as shown in \autoref{tab:load}: reassuringly, similarly-worded items load on the same factor across the two cohorts, and also the magnitude of the respective loadings (measuring the strength of the association between the item and the factor) is very smilar. Following previous research, we name the first dimension \emph{Externalising skills} (EXT, comprising the items restless, squirmy/fidgety, fights/bullies, distracted, tantrums, and disobedient) and the second dimension \emph{Internalising skills} (INT, comprising the items worried, fearful, solitary, unhappy, and aches).

\subsection{Factor model \label{sec:fm}}

Equipped with the factor structure inferred in the previous section, we specify a multiple-group factor analysis model to formally quantify the strength of the relationship between the observed items in our scale and the two latent socio-emotional skills. We specify two groups of children $c=\{BCS,MCS\}$, corresponding to the two cohorts. Each individual child is denoted by $j=1\dots N_c$, where $N_c$ is the number of children in cohort $c$. For each child $j$ in cohort $c$, we observe categorical items $X_{ijc}$ with $i=1,\dots,11$, corresponding to the eleven maternal reports in \autoref{tab:scalecomp}. We assume that each child is characterised by a latent bi-dimensional vector of externalising and internalising socio-emotional skills $\bm{\theta}_{jc} = (\theta_{jc}^{EXT},\theta_{jc}^{INT})$, as shown by the EFA in the previous section.

Children are assumed to have a latent continuous propensity $X^*_{ijc}$ for each item $i=1,\dots,I$. We model this propensity as a function of item- and cohort-specific intercepts $\nu_{ic}$ and loadings $\bm{\lambda}_{ic}$, and the child's latent skills $\bm{\theta}_{jc}$, plus an independent error component $u_{ijc}$. The propensity for each item can be written as follows:

$$X^*_{ijc} =\nu_{ic} + \bm{\lambda}_{ic}\bm{\theta}_{jc} + u_{ijc} \qquad \text{for} \quad i = 1,\dots,11$$

or more compactly:
\be\label{eq:propens}
\bm{X}^*_{jc} =\bm{\nu}_c + \bm{\Lambda}_{c}\bm{\theta}_{jc} + \bm{u}_{jc}
\ee

We make the common assumption of a dedicated (or congeneric) factor structure, where each measure is assumed to load on only one latent dimension \citep{Heckman2013,conti2010education,Attanasio2017c}. We mirror the structure found in the exploratory factor analysis (see \autoref{tab:load}), and assume that items 1-6 load exclusively on the externalising factor and items 7-11 on the internalising factor.\footnote{The dedicated factor structure corresponds to a sparse loading matrix, i.e.: $$\bm{\Lambda}_{c} \coloneqq \begin{bmatrix} \lambda_{1c}, \dots, \lambda_{6c} & \bm{0} \\ \bm{0} & \lambda_{7c}, \dots, \lambda_{11c} \end{bmatrix}.$$}

The discrete ordered nature of the observed measures $X_{ijc}$ is incorporated by introducing item- and cohort-specific threshold parameters $\tau_{ic}$ \citep{Muthen1984}. The observed measures as a function of the propensities $X^*$ can be then written as follows:
\be\label{eq:thresh}
X_{ijc} = s \qquad \text{if} \; \tau_{s,ic} \leq X^*_{ijc} < \tau_{s+1,ic} \qquad \text{for} \; s=0,1,2
\ee
with $\tau_{0,ic}=-\infty$ and $\tau_{3,ic}=+\infty$. Notice that we recode all ordered items to have higher values for \emph{better} behaviours, so that our latent vectors can be interpreted as favourable skills and not behavioural problems.

The model implies the following expression for the mean and covariance structure of the latent propensities:
$$\bm{\mu}_c=\bm{\nu}_c + \bm{\Lambda}_c\bm{\kappa}_c \qquad \text{and} \qquad \Sigma_c = \bm{\Lambda}_c \bm{\Phi}_c \bm{\Lambda}_c^\prime + \bm{\Psi}_c.$$

The model restrictions in \eqref{eq:propens} and \eqref{eq:thresh} do not identify the parameters without additional assumptions. As per the traditional factor analysis approach, we impose a normal distribution on the latent skills and error terms:\footnote{Recent work has also used mixtures of normals for the latent factors distribution, e.g. \citet{conti2010education}.}
\be\label{eq:norm}
\bm{\theta}_{jc} \sim N(\bm{\kappa}_c, \bm{\Phi}_c) \qquad \text{and} \qquad \bm{u}_{jc} \sim N(\bm{0}, \bm{\Psi}_c).
\ee

Even with these assumptions, there are infinite equivalent parameterisations through which the model can be identified -- the well-known issue of factor indeterminacy. We follow common practice and identify the model by setting the mean $\bm{\kappa}$ and variance $\bm{\Phi}$ of the latent skill factor in both cohorts to zero and one, respectively. Furthermore, we set intercepts to zero and error variances to one. Loadings $\lambda$ and thresholds $\tau$ are instead allowed to vary across cohorts.

\be\label{eq:thetapar}
\text{diag}(\bm{\Phi}_c) = \bm{I}, \quad \bm{\kappa}_c=\bm{0}, \quad \bm{\nu}_c=\bm{0}, \quad \text{and} \; \text{diag}(\bm{\Psi}_c) = \bm{I} \qquad \forall c \in \{BCS,MCS\}.
\ee

The restrictions in \eqref{eq:propens}, \eqref{eq:thresh}, \eqref{eq:norm}, and \eqref{eq:thetapar} define the so-called \emph{configural} model. This is a `minimum' identifiable model, in that it places the least possible restrictions on how parameters are allowed to vary across cohorts. It serves as a basis for our measurement invariance analysis in the next section.\footnote{This set of identifying restrictions is known as \emph{Theta} parameterisation \citep{Wu2016a}. See \aref{asec:midet} for statistically equivalent alternative parameterisations.}

\section{Measurement invariance \label{sec:measinv}}

Any comparison between socioemotional skills across the two cohorts requires that the measures at our disposal have the same relationship with the latent constructs of interest in both cohorts. In other words, the items in our new scale must measure externalising and internalising socioemotional skills in the same way in the BCS and MCS data. This property is denominated measurement invariance \citep{Vandenberg2000a,Putnick2016}.

In the framework of factor analysis, measurement invariance is a formally testable property. In this paper, we follow the recent identification methodology by \cite{Wu2016a}. The configural model defined in the previous section serves as the starting point. Measurement invariance is then assessed by comparing the configural model to a series of hierarchically nested models. These models place increasing restrictions on the item parameters, constraining them to be equal across groups. Their fit is then compared to that of the configural model. Intuitively, if the additional cross-group restrictions have not significantly worsened model fit, one can conclude that a certain level of invariance is achieved. The hierarchy of restrictions is detailed in \autoref{tab:invparam}.

Let's consider examples from our application. A \emph{loading and threshold invariance} model restricts every item's loading $\lambda$ and threshold $\tau$ parameters to have the same value in the two cohorts. It assumes that the items in our scale have the same relationship with latent skills across the two cohorts. In other words, items have the same salience, or informational content relative to skills. If this model fits as well as the configural model, we can be confident that the socioemotional skills of children in the two cohorts can be placed on the same scale, and their \emph{variances} can be compared. To see why, consider equation \eqref{eq:propens}. If the loading matrix $\bm{\Lambda}$ is the same across cohorts, any difference in latent skills $\Delta\bm{\theta}$ will correspond to the same difference in latent propensities $\Delta\bm{X}^*$. Equality of thresholds $\tau$ ensures that propensities $\bm{X}^*$ map into observed items $\bm{X}$ in the same way.

A \emph{loading, threshold, and intercept invariance} model additionally restricts every item's intercept $\nu$ across cohorts. A good relative fit of this model indicates that socioemotional skills can be compared across cohorts in terms of their \emph{means} as well. To see why, consider the following. Since the $\lambda$ and $\nu$ parameters are the same across cohorts, a child in the BCS cohort with a given level of latent skills $\bar{\bm{\theta}}$ will have the same expected latent item propensities $\bm{X}^*$ as a child with the same skills in the MCS cohort. Again, equality of thresholds $\tau$ fixes the mapping between $\bm{X}^*$ and $\bm{X}$.\footnote{We recognise that simultaneous invariance of \emph{all} items is not the minimum requirement for comparability. In theory, the availability of just one invariant item (known as `anchor') would suffice to fix the scale and location of the system. However, partial invariance approaches are hard to implement in practice. Its validity hinges on selecting one (or more) truly invariant anchor, which is challenging on an a priori basis. The full procedure, restricting all parameters of a certain type across groups, does not identify which items are at the source of the invariance. Algorithms have been proposed to deal with this issue \citep{Yoon2007,Cheung2012}, however there are still doubts on their robustness and their applicability to the categorical case \citep{Vandenberg2016}.}

We estimate the sequence of models detailed in \autoref{tab:invparam} by Weighted Least Squares.\footnote{Parameters are estimated by mean- and variance-adjusted weighted least squares (WLSMV) -- see \citet{Muthen1997}; estimation starts from the items' polychoric correlation matrix, uses diagonally weighted least squares (DWLS), and exploits the full weight matrix to compute robust standard errors and test statistics. Robust WLS has proved in simulation studies to be moderately robust to small violations of the normality assumption in the latent underlying measures \citep{Flora2004}, and generally outperforms maximum likelihood in large samples \citep{Beauducel2006,Li2016}. All estimates are computed using the \texttt{lavaan} package (version 0.6-2) in \textbf{\textsf{R}} \citep{Rosseel2012}.}
For the purposes of the analysis, we define groups $c$ as cohort-gender cells, with the reference group being males in the BCS cohort. We then compare the fit of each model against the configural model.

Comparison of $\chi^2$ values across models is a common likelihood-based strategy. However, tests based on $\Delta \chi^2$ are known to display high Type I error rates with large sample size and more complex models such as our own \citep{Sass2014}. In fact, for all invariance levels in our applications a chi-squared difference would point to a lack of measurement invariance. The use of approximate fit indices (AFIs) is therefore recommended alongside $\chi^2$. These indices do not have a known sampling distribution, thus making it necessary to rely on rules of thumb to assess what level of \textDelta AFI indicates invariance. Nevertheless, AFIs are widely used in empirical practice to assess model fit.\footnote{The root mean squared error of approximation (RMSE) and the Tucker-Lewis index (TLI) are traditionally the most used AFIs in empirical practice. Simulation evidence by \cite{Cheung2002} shows that these indices can show correlation between overall and relative fit, and suggest relying on additional indices, such as the comparative fit index (CFI, \citealp{Bentler1990}), McDonald non-centrality index (MFI, \citealp{McDonald1989}), and Gamma-hat index \citep{Steiger1989} for the case of ordered measures. Commonly accepted thresholds for rejection are $\Delta CFI < - 0.01$, $\Delta MFI < - 0.02$, and $\Delta Gamma hat < - 0.001$. \cite{Meade2008}, using the results from a simulation study, suggests stricter thresholds that should apply in a variety of conditions. For CFI, a single cutoff value of $.002$ is proposed, while cutoffs for MFI depend on the problem's characteristics; in our case (2 factors, 11 items), they suggest $.0066$. \cite{Sass2014} however cast some doubts of the generalisability of these cutoffs to WLSMV estimators.}

The fit of each model is compared in Panel A of \autoref{tab:fit}. The model with restricted thresholds and loadings exhibits a comparable fit to the configural model, according to all the AFIs. Invariance of loadings and thresholds across cohorts implies that items in our scale are equally salient in their informational content, and that the latent propensities have equal mapping into the observed items. However, further restricting intercepts results in a model where invariance is rejected across the board.\footnote{We do not present fit results for the threshold-only invariance model, as it is statistically equivalent to the configural model and thus its fit is mathematically the same -- see Table 3 in \citealp{Wu2016a}. The ages at which socio-emotional skills are observed varies slightly between BCS and MCS, due to different sampling and fieldwork schedules. In the MCS cohort, the age distribution has significantly higher variance. In Panel B of \autoref{tab:fit}, we restrict the sample to 59 to 61 months, where the overlap between BCS and MCS is maximised. In Panels C and D, we restrict to male and female children respectively. In all these cases, invariance of thresholds and loadings is achieved, but invariance of intercepts is rejected. We can thus rule out that the lack of intercept invariance comes from differences in ages or invariance across child gender.} In other words, intercept parameters in our model ($\nu$) are estimated to be different between maternal reports in the British and Millennium Cohort Studies. This means that, for a given level of latent skills, mothers in MCS tend to assess behaviours differently from mothers in BCS. Thus, cohort differences in scores on our scale cannot be unequivocally interpreted as differences in the underlying skills, since they might also reflect differences in reporting.

This is an important finding, which has to our knowledge never been acknowledged in this literature. How can this lack of comparability be explained?  A possible interpretation is connected with secular evolution of social and cultural norms about child behaviours. For example, commonly held views of what constitutes a restless, distracted, or unhappy child might have changed between 1975 and 2005/6.\footnote{Calibrating the Rutter and SDQ using a contemporary sample of children cannot rule out this issue. For example, \cite{Collishaw2004} administered both Rutter and SDQ items to parents of a small sample of adolescents in London. They use the mapping between the two questionnaires to impute Rutter scores for mothers who answered the SDQ. This can correct for contemporaneous reporting differences between questionnaires, but cannot tackle reporting differences between samples collected at different times in history.}

To summarise, our measurement invariance analysis shows partial comparability of socioemotional skills across cohorts. In particular, the variance of skills can be compared across cohorts, but mean cohort differences do not necessarily reflect differences in skills. We can use scores from our scale to compare children within the same cohort, but not across cohorts. However, we can also compare within-cohort differences between groups of children, across cohorts. As an example, consider two groups of children A and B in the BCS cohort, and two groups of children C and D in the MCS. We cannot compare the mean level of skills between groups A and C, but we can compare the mean difference between groups A and B with the mean difference between groups C and D. This is the approach we take for the rest of the paper. Refraining from direct cross-cohort comparisons, we interpreting significance and magnitude of within-cohort differences across the cohorts.

### 2018-10-01 codes/R/paper/text/paper_jpube_3resultsconclusion.tex
add text from coauthors

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Results}\label{sec:results}

Parameter estimates from our factor model are presented in \autoref{tab:measpars}. As discussed in the previous section, loadings and thresholds are constrained to have the same value across groups. Intercepts are normalised to zero, and error variances to one, for the reference group -- males in the BCS cohort. We use the estimates from this model to predict a score for each child in our sample along the latent externalising and internalising socio-emotional skill dimensions.\footnote{We use an empirical Bayes modal (EBM) approach to estimate the scores. The parameters are estimated using three sources of information. The first is the distribution of the latent variables $\bm{\theta}$, treated as random parameters with a prior $h(\bm{\theta}, \bm{\Omega})$, conditional on the parameters $\bm{\Omega}$. This prior is assumed to be multivariate normal. The second is the observed data $\bm{X}$, and the third is the estimated parameters $\hat{\bm{\Omega}}$. Data and prior are combined into the posterior distribution $w(\bm{\theta}| \bm{X}, \hat{\bm{\Omega}})$. For further details, see Chapter 7 in \cite{Skrondal2004}.} We plot the distribution of the scores in \autoref{fig:dens}. The unit of measurement is standard deviations of the distribution in the subsample of males in the BCS. Given our measurement invariance results in section \ref{sec:measinv}, we stress that the \emph{location} of these scores should not be directly compared across cohorts. However, the shape of the distribution can be given a cross-cohort interpretation.\footnote{The density of the scored factors can be contrasted with the distribution of sum scores in \autoref{fig:sumscores}. Using raw scores, instead, shows an increase in mass only at the top of the distribution.} It is immediately visible that there is more mass in the tails of the distribution in the 2000 than in the 1970 cohort.

\subsection{Inequality in socioemotional skills}

We find that, both unconditionally and for specific groups, inequality in socio-emotional skills at age five has increased between 1975 and 2005/6. \autoref{tab:iqdiff} shows unconditional inequality statistics, using quantile differences in the distribution of skills by gender and cohort. With the exception of internalising skills in female children, all distributions have widened substantially between the BCS and MCS cohorts. The gap for both externalising and internalising skills between the 90th and the 10th percentiles for males has increased by approximately half a standard deviation. The increase in the gap is more pronounced in the bottom half of the distribution. For females, we see a narrowing at the top (90-50), but a widening at the bottom (50-10) of the distribution, again for both externalising and internalising skills.

Inequality has also increased conditional on socioeconomic status. \autoref{fig:ineqps} shows mean skills by maternal education. We compare mothers who continued education past the compulsory age with mothers who left school at the compulsory leaving age, according to their year of birth. Given lack of comparability in the level of skills across cohort, we normalise the mean in the `Compulsory' group to zero for both cohorts. For both males and females, and for both externalising and internalising skills, the difference in the socio-emotional skills of their children between more and less educated mothers has increased. The size of the increase is around .1 to .15 of a standard deviation. The increase is particularly pronounced for males, for whom it goes from .20 to .30 for externalising and from .12 to .24 for internalising.

\autoref{fig:ineqsm} shows an even starker pattern when comparing children of mothers who smoked in pregnancy with non-smoking mothers. The fact that maternal smoking during pregnancy is a risk factor for offspring behavioural problems is well known in the medical literature \citep{gaysina2013maternal}; less evidence there is however, on whether and to which extent these associations have changed across cohorts. The difference in child skills has increased, from less than .2 to around .4 of a standard deviation, again with the biggest increase experienced by the boys. There is also a significant increase in the gradient by paternal occupation based on social class (\autoref{fig:ineqfsc}), although this is less pronounced if compared to the one based on maternal characteristics. In particular, male children with no father figure living in their household have worse skills compared to children with blue collar fathers in the MCS cohort. Otherwise, skill differences in father's occupation are mostly constant across the two cohorts.\footnote{Figures \ref{fig:ineqips}, \ref{fig:ineqism}, and \ref{fig:ineqifsc} show inequality in the scale items underlying the factor scores used in this section. The increase in inequality across cohorts is still present, but less marked when looking at these single items. This shows the importance of the factor analysis step in aggregating items, explicitly modelling the measurement error, and testing and accounting for (loadings and thresholds) invariance across the two cohorts.} These patterns are in stark contrast with the findings of \citet{reardon2016recent} for the US, who have found a narrowing of the readiness gaps from 1998 to 2010.

We then examine the same patterns as in the previous figures, but conditional on other family background indicators. The aim is to disentangle the relative contribution of each indicator to socio-emotional skills, and how it has changed in the thirty years between the two cohorts. \autoref{tab:determ} shows coefficients from linear regressions of socio-emotional skills at five on contemporaneous and past socioeconomic indicators, by cohort and gender. Coefficients for indicators in BCS and MCS are presented side by side, together with the $p$-value of the hypothesis that coefficients are the same in the two cohorts.\footnote{We also estimated Tobit models to account for the right truncation of the distribution of skills -- see \autoref{fig:dens}. Tobit estimates are extremely similar to the linear estimates in \autoref{tab:determ}, and are available from the authors upon request.}

Overall, the importance of maternal socioeconomic status (education and in particular employment) in determining socio-emotional skills has increased from the BCS to the MCS children. The `premium' in skills for children of better educated and employed mothers is significantly larger, for both boys and girls, internalising and externalising skills. At the same time, the penalty for having a blue-collar father, or not having a father figure at all in the household, has significantly declined across the two cohorts, especially for girls. Being born to an unmarried mother, and to a mother who smoked during pregnancy, is associated with a higher penalty for both dimensions of socio-emotional skills in the latter cohort, but only for males. This is consistent with recent evidence which shows that family disadvantage disproportionately impedes the pre-market development of boys, in terms of higher disciplinary problems, lower achievement scores, and fewer high-school completions \citep{autor2016family}. Girls of non-white ethnicity, instead, have worse internalising and externalising skills in the MCS, a penalty not suffered by 5-year old non-white girls in the BCS. Firstborn boys and girls in the BCS have worse skills, but this difference disappears in the MCS. Lastly, we document an increase in the returns to birth weight, which is more pronounced for boys' internalising skills.

These changes in the relative importance of pregnancy factors and family background characteristics for child socio-emotional skills at age 5 need to be interpreted in the light of the significant changes in the prevalence of such characteristics across cohorts. As shown in \autoref{tab:sumstats}, the age of the mother at birth, proportion of mothers non-smoking in pregnancy, with post-compulsory education and in employment at the age 5 of the child has substantially increased; at the same time, the proportion of households with no father figure has increased, and so the proportion of women unmarried at birth is much higher in the 2000 than in the 1970 cohort. Also, the ethnic structure of the population has changed, with a higher proportion of non-white children in the MCS than in the BCS. In general, this has been a period of significant societal changes, with an almost continual rise in the proportion of women in employment, an older age at first birth and a rise in dual-earning parents families \citep{roantree2018women}. However, here we do not attempt to disentangle whether and to which extent the observed changes in inequality in socio-emotional skills across the two cohorts can be attributed to changes in returns (or penalties) to maternal characteristics (such as education and employment) or to compositional changes, like it has been done for the analysis of wage inequality \citep{blundell2007changes}.

\subsection{Socio-emotional skills and adolescent/adult outcomes}

In this last section, we study the predictive power of socio-emotional skills for adolescent and adult outcomes. We contribute to a vast interdisciplinary literature by examining medium- and long-term impacts of skills measured at an earlier age than usually examined in previous studies, well before the start of formal education. We do so by regressing health and socioeconomic outcomes measured in adolescence and adulthood on the socio-emotional skills scores at age five obtained by our factor model, controlling for the harmonised family background variables at birth and age five (see \autoref{tab:harmvar}). We present results with and without controlling for cognitive skills. As detailed in \autoref{sec:data}, the available cognitive measures are not comparable across cohorts. Still, we control for a factor score that summarises all information on cognitive skills that is available in each cohort, regardless of their comparability.

Socio-emotional skills at five years of age are predictive of adolescent health behaviour and outcomes in both cohorts.\footnote{Unfortunately the strength of the association cannot be directly compared, since the outcomes are measured at different ages: 16 and 14 years for BCS and MCS, respectively.} \autoref{tab:adol} examines adolescent smoking and BMI for both cohorts; \autoref{tab:bcs_adultapp} reports the results for the same outcomes in adulthood (at age 42), for the BCS only. Externalising skills are negatively correlated to subsequent smoking and BMI in both cohorts, for both genders. Recall that a child with high externalising skills exhibits less restless and hyperactive behaviours, and has less anti-social conduct. Our findings are consistent with the body of evidence reviewed in section \ref{sec:lit}, which shows that better socio-emotional skills (measured using different scales and at various points during childhood and adolescence) are negatively associated with smoking. At the same time, internalising skills are positively correlated with smoking (only in the 1970 cohort) and BMI (only for girls), although less strongly than externalising skills. This apparently counterintuitive result makes sense in light of the items in our internalising scale shown in \autoref{tab:scalecomp}. A child with better internalising skills is less solitary, neurotic, and worried. From this perspective, he is likely more sociable and subject to peer influence in health behaviours. This is consistent with the evidence in \citet{goodman2015social}, who find a positive association between child emotional health (measured with items for the internalizing behaviour subscale of the Rutter scale at age 10 in the BCS) and smoking at age 42. Furthermore, in recent work \citet{hsieh2018smoking} have shown personality to be a key mechanism through which peers affect smoking behaviour.

Conditional on socio-emotional skills, cognition has limited predictive power for these behaviours, and only for girls.\footnote{We do not observe significant associations between early socio-emotional skills and other risky behaviours like drug-taking and alcohol consumption. One possible reason might be the relatively young age at which these skills are measured. Results are available upon request.} This is in line with the evidence in \citet{conti2010understanding}, who show that not accounting for non-cognitive traits (a self-regulation factor measured at age 10) overestimates the importance of cognition for predicting health and health behaviours, using data from the British cohort study. \citet{conti2013personality} use rich data on child personality and socio-emotional traits collected at ages 7, 11 and 16 in the 1958 British birth cohort,\footnote{They use the Rutter scale and the Bristol Social Adjustment Guide.} and show that these traits rival the importance of cognition in explaining the education gradient in health behaviours (including smoking and BMI). We show that child socio-emotional skills have greater predictive power for health outcomes and behaviours even when measured at an earlier age.

For the British Cohort Study, we last examine the association between socio-emotional skills at age five and adult education and labour market outcomes. The structure of \autoref{tab:bcs_adult} is similar to \autoref{tab:adol}, but it considers educational achievement, employment, and earnings (conditional on being in paid employment) for the BCS cohort members. For these outcomes, the predictive power of cognitive skills outweighs that of socio-emotional skills, whose predictive power diminishes over time (between the ages 34 and 42), and is driven to insignificance after controlling for cognition. This is consistent with the evidence in \citet{conti2011early}, who show that cognitive endowments at age 10 are more predictive (than socio-emotional and health ones) for employment and wage outcomes in the BCS. Again, we show that the greater predictive power of cognition for socioeconomic outcomes holds even when considering earlier-life measures of child development.

\section{Conclusion}\label{sec:conclusions}

In this paper we have studied inequality in a dimension of human capital which has received limited attention in the literature so far: socio-emotional skills very early in life. In particular, we have focused on the measurements of these skills at age 5 in two British cohorts born 30 years apart: the one of children born in 1970 (British Cohort Study, BCS) and the one of children born in 2000/1 (Millenium Cohort Study, MCS). We have provided several contributions to the recent but flourishing literature on the determinants and consequences of early human development.

We have taken very seriously the issue of comparability of measurements of socio-emotional skills across cohorts. First, we have selected 11 comparable items across two related scales: the Rutter scale in the BCS, and the Strength and Difficulties Questionnaire (SDQ) in the MCS. After examining the latent structure underlying the items, we have identified by means of exploratory factor analysis two dimensions of socio-emotional skills. We have labeled them `internalising' and `externalising' skills, the former related to the ability of children to focus their concentration and the latter to engage in interpersonal activities.

Second, we have formally tested for measurement invariance of the 11 items across the two externalising and internalising scales, following recent methodological advances in factor analysis with categorical outcomes. We have found only partial support for measurement invariance, with the implication that we have only been able to compare how inequality in these socio-emotional skills across the two cohorts has changed, but not whether their average level is larger or smaller in one of the two cohorts. These results sound a warning to research in this area which routinely compares levels of skills across different groups (at different times, or of different gender), without first establishing their comparability.

Third, after having computed comparable scores for both externalising and internalising skills, and for both boys and girls, we have compared how inequality in these skills has changed across the 1970 and the 2000 cohort. We have documented for the first time that inequality in these early skills has increased across cohorts, especially for boys. The cross-cohort increase in the gap is more pronounced at the bottom of the distribution (50-10 percentiles). We have also documented changes in conditional skills gaps across cohorts. In particular, the difference in the socio-emotional skills of their children between mothers of higher and lower socio-economic status (education and employment) has increased. The increase in cross-cohort inequality is even starker when comparing children born to mothers who smoked during pregnancy. In both cases, the increase in inequality is particularly pronounced for boys. On the other hand, the skills penalty arising from the lack of a father figure in the household has substantially declined, especially for girls.

Fourth, we have contributed to the literature on the predictive power of socio-emotional skills by showing that even skills measured at a much earlier age than in previous work are significantly associated with outcomes both in adolescence and adulthood. In particular, socio-emotional skills are more significant predictors of health and health behaviours (smoking and BMI), while cognition has greater predictive power for socioeconomic outcomes (education, employment and wages). Our results show the importance of inequalities in the early years development for the accumulation of health and human capital across the life course.

### 2018-10-01 codes/R/paper/text/paper_jpube_4appendix.tex
add text from coauthors

--- 2018-07-02 codes/R/paper/paper_jpube.Rnw
@@
-<<MAIN,cache=FALSE,echo=FALSE,results='hide',message=FALSE,warning=FALSE>>=
-rm(list=ls())         # Clear all objects from memory 
-  opts_chunk$set(concordance=FALSE,
-                 echo=FALSE,
-                 tidy=FALSE,
-                 autodep=TRUE,
-                 cache=T, # REPLACE LATER
-                 message=FALSE, 
-                 warning=FALSE,
-                 size='scriptsize',
-                 results='hide',
-                 comment=NA, 
-                 fig.width=4,
-                 fig.height=4, 
-                 out.width='\\textwidth',
-                 fig.show='hold',
-                 tikzDefaultEngine='pdftex',
-                 tikzDocumentDeclaration = "\\documentclass[10pt]{article}",
-                 tikzMetricPackages = c("\\usepackage[utf8]{inputenc}","\\usepackage[T1]{fontenc}", "\\usetikzlibrary{calc}", "\\usepackage{amssymb}")
-                 )
+\section{Deriving a common scale of socioemotional skills}\label{asec:scales}
 
-@
+In the BCS data, maternal reports on child socioemotional skills are measured using the Rutter A Scale \citep{Rutter1970} -- see Panel A of \autoref{tab:scales}.
+The Rutter items are rated on three levels: `Does not apply', `Somewhat applies', `Certainly applies'.
+Since they are all behaviours indicating lower skills, we encode all of them in reverse, i.e.
+`Certainly applies' = 0, `Somewhat applies' = 1, `Does not apply' = 2.
+We augment the 19-item Rutter Scale with four additional parent-reported questions from the parental questionnaire, items A to D.
+These are rated on 4 levels: `Never in the last 12 months', `less than once a month', `at least once a month', `at least once a week'.
+we recode these into binary indicators, with `Never' and `Less than once a month' to 1 and zero otherwise.
+To increase comparability between the two scales, we merge together two pairs of items: 4 and 19 (to mirror SDQ item 12 ``Often fights with other children or bullies them"), and A and B (to mirror SDQ item 3 ``Often complains of head-aches, stomach-ache or sickness").
+We assign the lowest category among the two original items to the newly obtained item.
+We also recode items 5 and 14 to binary instead of three categories.
+These items are recorded with a positive phrasing in SDQ, so a 3-category split would be harder to compare.
 
-<<READ, cache=FALSE >>=
-library(here)
-read_chunk(here("codes/R/1_load.R"))
-read_chunk(here("codes/R/2_efa.R"))
-read_chunk(here("codes/R/3_inv_main.R"))
-read_chunk(here("codes/R/3_inv_aux.R"))
-read_chunk(here("codes/R/3_inv_fit.R"))
-read_chunk(here("codes/R/4_scores.R"))
-read_chunk(here("codes/R/4_scores_aux.R"))
-read_chunk(here("codes/R/5_graphs.R"))
-read_chunk(here("codes/R/6_selectvars.R"))
-read_chunk(here("codes/R/8_regs_dep.R"))
-read_chunk(here("codes/R/8_regs_outc.R"))
-read_chunk(here("codes/R/9_gelbfig.R"))
-read_chunk(here("codes/R/getpars.R"))
+In MCS, we use the 25-item strengths and difficulties questionnaire \citep{Goodman1997} -- see Panel B of \autoref{tab:scales}.
+All items are recorded on a 4-point scale: `Not true', `Somewhat true', `Certainly true', `Can't say'.
+We set the latter option to missing and recode the rest in ascending order of skill as for the BCS items, i.e.
+`Certainly true' = 0, `Somewhat true' = 1, `Not true' = 2.
+For comparability with the BCS Rutter scale, we dichotomise items 3 and 5 to make them comparable with , and dichotomise and invert items 7, and 14.
 
-@
+\section{Measurement invariance details}\label{asec:midet}
 
-<<PREAMBLE >>=
-@
+\subsection{Alternative parameterisations for the configural model}
 
-<<LOAD_DATA>>=
-@
-
-<<MEANTABLE>>=
-# make table with means
-@
-
-<<MEANTABLE_PRINT>>=
-# function to print mean table
-printmeantab <- function(tab) {
-  print(xtable(tab,
-               align="L{0cm}C{.04\\textwidth}C{.05\\textwidth}C{.05\\textwidth}L{.4\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}L{.4\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
-               digits = 1
-  ), 
-  size="\\small",
-  include.rownames=F,
-  include.colnames=F,
-  booktabs = T,
-  add.to.row = list(pos = c(list(0)),
-                    command = c(
-                      # header
-                      paste(" & & &
-                            \\multicolumn{4}{c}{\\textbf{BCS}} &
-                            \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){4-7} \\cmidrule(lr){8-11} 
-                            Item & Factor & Num. Categ. &
-                            Wording & Cert. Appl. (\\%) & Smtm. Appl. (\\%) & Appl. (\\%) &
-                            Wording & Cert. True (\\%) & Smwt. True (\\%) & True (\\%) \\\\ \\midrule
-                            "
-                            , sep="")
-                    )),
-  sanitize.text.function = function(x){x},
-  sanitize.rownames.function = function(x){""}, # to not print row numbers
-  tabular.environment = "tabular", width = ".8\\textwidth",
-  floating = FALSE
-  )
-}
-
-@
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-<<FA_INV_MAIN>>=
-# main models only
-@
-
-<<FA_FIT>>=
-# assemble fit indices into matrices
-@
-
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\documentclass[pdftex,10pt,twoside,a4paper]{article}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsfonts}
-\usepackage{amstext}% for resizing the symbol in math
-\usepackage{amsthm}
-\usepackage[utf8]{inputenc}
-\usepackage[english]{babel}
-\usepackage[babel]{csquotes}
-\usepackage{booktabs}
-\usepackage{tocloft}
-\usepackage{dsfont}
-\usepackage{xcolor}
-\usepackage{accents}
-\usepackage{rotating}
-\usepackage{pdflscape}  % for landscape page
-\usepackage{lmodern}  %font
-\usepackage{natbib}
-\usepackage{array}
-\usepackage{graphicx}
-\usepackage{adjustbox}
-\usepackage{tikz}
-\usepackage{bm}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage[multiple]{footmisc}   % for multiple footnotes
-\usepackage{textgreek}            % for greek letters in text
-\usepackage{ragged2e}             % for table notes justification
-\usepackage{setspace} % for spacing
-\onehalfspacing
-
-\setlength{\jot}{11pt} % increase vertical spacing in equations
-\usepackage{txfonts}
-\usepackage[T1]{fontenc}
-
-% define new columns for tables
-\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\renewcommand{\arraystretch}{1.2}%
-
-% appendices/numeration/references
-\usepackage[nottoc,numbib]{tocbibind} 		% bibliography in TOC
-\usepackage[toc,page]{appendix} %appendix
-\newcommand{\tocless}[2]{\bgroup\let\addcontentsline=\nocontentsline#1{#2}\egroup} % for empty section
-\numberwithin{equation}{section} %for section eq numeration
-\usepackage{hyperref}                            % references
-\AtBeginDocument{%
-\renewcommand\sectionautorefname{Section}        % custom autoref names
-\renewcommand\subsectionautorefname{Section}
-}
-\newcommand{\aref}[1]{\hyperref[#1]{Appendix~\ref{#1}}} % for appendix sections
-\usepackage{cleveref}
-
-%tikz
-\usepackage{tikz}
-\usetikzlibrary{shapes,intersections}
-\usetikzlibrary{positioning,arrows, arrows.meta}
-
-%lists
-\usepackage[inline]{enumitem}
-\setlist{noitemsep}
-\setlist[enumerate]{labelsep=*, leftmargin=*}
-\setlist[itemize]{labelsep=*, leftmargin=*}
-\usepackage[textfont={small},labelfont={bf,small},justification=centering]{caption}     % to allow multiple lines in captions, and customize the look
-
-% page setup (from Racine)
-\setlength{\topmargin}{0cm}
-\setlength{\textheight}{24cm}
-\setlength{\oddsidemargin}{.0cm}
-\setlength{\evensidemargin}{.0cm}
-\setlength{\textwidth}{16.5cm}
-\setlength{\footskip}{1.5cm}
-\setlength{\footnotesep}{0.4cm}
-\linespread{1.3}
-
-% additional symbols
-\newcommand{\me}{\mathrm{e}}
-\newcommand{\veps}{\varepsilon}
-\newcommand{\E}{\mathbb{E}}
-\newcommand{\1}{\mathds{1}}
-\newcommand{\Cov}{\mathrm{Cov}}
-\newcommand{\Var}{\mathrm{Var}}
-\newcommand{\Corr}{\mathrm{Corr}}
-\newcommand{\vect}{\mathrm{vec}}
-\newcommand{\vecht}{\mathrm{vech}}
-\newcommand{\ubar}[1]{\underaccent{\bar}{#1}}
-\def\chk{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}
-
-% additional commands
-\newcommand{\bi}{\begin{itemize}}
-\newcommand{\ei}{\end{itemize}}
-\newcommand{\bnm}{\begin{enumerate}}
-\newcommand{\enm}{\end{enumerate}}
-\newcommand{\be}{\begin{equation}}
-\newcommand{\ee}{\end{equation}}
-
-\usepackage{authblk}
-\title{Inequality in non-cognitive skills: a cross-cohort comparison}
-\author[1,2]{Orazio Attanasio}
-\author[1,2]{Richard Blundell}
-\author[1,2]{Gabriella Conti}
-\author[1,2]{Giacomo Mason}
-\affil[1]{\emph{Institute for Fiscal Studies}}
-\affil[1]{\emph{UCL Department of Economics}}
-%\institute{UCL and IFS}
-\date{\today}
-
-\begin{document}
-\maketitle
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\clearpage
-\section{Measurement Invariance}
-
-Meaningful comparison of latent constructs across different groups of individuals and/or measurement occasions (in our case, cohorts) requires that the available measures have the same relationship with the latent factors.
-This property is denominated measurement invariance \citep{Vandenberg2000,Putnick2016}.
-
-Measurement invariance (MI) is a \emph{formally testable} property.
-Two main frameworks have been used for assessing MI: item-response theory (IRT), or structural equation modeling (SEM).
-This paper focuses on the latter, as it is the most commonly employed.\footnote{For an introduction to MI in an IRT framework, see for example \cite{Tay2015}.} In general, MI analyses in a SEM framework are carried out by specifying and estimating factor models of different invariance levels, and comparing their relative fit.
-It starts with the estimation of a minimally restrictive model, where all measurement parameters are allowed to vary across groups -- net of the necessary normalisations -- and proceeds with progressively restricting sets of parameters to be equal across groups.
-If the fit of more restrictive models is comparable to the configural model, different levels of invariance can be established.
-
-In accordance with the SEM approach, we set up a (confirmatory) item factor analysis (IFA) model to examine measurement invariance.
-In our application, we observe 11 categorical items ($X_1$-$X_{11}$).
-Our exploratory analysis above shows that the correlation structure of these measures can be captured by a two-dimensional latent construct (factor) $\bm{\theta}$ -- externalising and internalising socioemotional skills.
-We make the commonly adopted assumption of a dedicated (or congeneric) factor structure, where each measure is assumed to load on only one latent dimension \citep{Attanasio2017c,Heckman2013}.
-
-To account for the ordered nature of the measures, we specify a threshold model \citep{Muthen1984}.
-Individual $j=1\dots N_c$ in group $c=\{A,B\}$ is assumed to have a latent continuous propensity $X^*_{ijc}$ for each item $i=1,\dots,I$.
-Thus, $X^*_{ijc}$ is the child's propensity to exhibit each of the behaviours described in \autoref{tab:comm}, as reported by the mother.
-This propensity is modelled as a function of item- and group-specific intercepts $\nu_{ic}$ and loadings $\bm{\lambda}_{ic}$, and the individual's level on the latent construct vector $\bm{\theta}_{jc}$, plus an independent error component $u_{ijc}$.
-In our application, the latent factor is bidimensional, and corresponds to the externalising ($\theta_{jc}^{EXT}$) and internalising ($\theta_{jc}^{INT}$) skills.
-
-The propensity for each item can be written as follows:\footnote{The dedicated factor structure is ensured by the sparsity of the loading matrix, i.e.: $$\bm{\Lambda}_{c} \coloneqq \begin{bmatrix} \lambda_{1c}, \dots, \lambda_{6c} & \bm{0} \\ \bm{0} & \lambda_{7c}, \dots, \lambda_{11c} \end{bmatrix}.$$}
-\be\label{eq:propens}
-X^*_{ijc} =\nu_{ic} + \bm{\lambda}_{ic}\bm{\theta}_{jc} + u_{ijc} \qquad \text{for} \quad i = 1,\dots,11 \\
-\ee
-
-or more compactly:
-$$\bm{X}^*_{jc} =\bm{\nu}_c + \bm{\Lambda}_{c}\bm{\theta}_{jc} + \bm{u}_{jc}$$
-
-Additional standard distributional assumptions are placed on the latent constructs and error terms:
-$$\bm{\theta}_{jc} \sim N(\bm{\kappa}_c, \bm{\Phi}_c) \qquad \text{and} \qquad \bm{u}_{jc} \sim N(\bm{0}, \bm{\Psi}_c)$$
-
-The model implies the following expression for the mean and covariance structure of the latent propensities:
-$$\bm{\mu}_c=\bm{\nu}_c + \bm{\Lambda}_c\bm{\kappa}_c \qquad \text{and} \qquad \Sigma_c = \bm{\Lambda}_c \bm{\Phi}_c \bm{\Lambda}_c^\prime + \bm{\Psi}_c.$$
-
-Finally, discreteness in the observed measures $X_{cji}$ is incorporated by introducing item- and group-specific threshold parameters $\tau_{ic}$.
-The observed measures as a function of the propensities $X^*$ can be then written as follows:
-$$X_{ijc} = s \qquad \text{if} \; \tau_{s,ic} \leq X^*_{ijc} < \tau_{s+1,ic} \qquad \text{for} \; s=0,1,2$$
-with $\tau_{0,ic}=-\infty$ and $\tau_{3,ic}=+\infty$.
-
-% FOOTNOTE\footnote{As a consequence of the normality assumption on the error terms $\veps$, the probability that individual $j$ in group $c$ exhibits item $i$ with ``intensity" $s$ is:
-% \begin{align*}
-% \Pr(X_{ijc}=0) &= 1- \Phi(-\tau_{1,ic} + \nu_{ic} + \lambda_{ic}\bm{\theta}_{jc}) \\
-% \Pr(X_{ijc}=1) &= \Phi(-\tau_{1,ic} + \nu_{ic} + \lambda_{ic}\bm{\theta}_{jc})
-% \end{align*}
-% for binary items and
-% \begin{align*}
-% \Pr(X_{ijc}=0) &= 1- \Phi(-\tau_{1,ic} + \nu_{ic} + \lambda_{ic}\bm{\theta}_{jc}) \\
-% \Pr(X_{ijc}=1) &= \Phi(-\tau_{1,ic} + \nu_{ic} + \lambda_{ic}\bm{\theta}_{jc}) - \Phi(-\tau_{2,ic} + \nu_{ic} + \lambda_{ic}\bm{\theta}_{jc})\\
-% \Pr(X_{ijc}=2) &= \Phi(-\tau_{2,ic} +  \nu_{ic} + \lambda_{ic}\bm{\theta}_{jc})
-% \end{align*}
-% for ordinal items. This IFA model can also be written as a 2-parameter Logit IRT model with item difficulty $b$ and item discrimination $a$. In the simplest case for a binary item $1$:
-% $$\text{Probit}(X_{1jc}=1) = a_{1c}(\bm{\theta}_{jc} - b_{1c})$$
-% such that $\tau_{1c}=a_{1c}b_{1c}$ and $a_{1c}=\lambda_{1c}$.}
-
-% Binary items take values:
-% \begin{equation}
-% X_{cji}=
-% \begin{cases}
-% \qquad 0 \qquad \text{if} \; X^*_{cji}<\tau_{1,ci} \\
-% \qquad 1 \qquad \text{if} \; X^*_{cji}\geq\tau_{1,ci}
-% \end{cases}
-% \end{equation}
-% 
-% An ordinal item takes values:
-% 
-% \begin{equation}
-% X_{cji}=
-% \begin{cases}
-% \qquad 0 \qquad \text{if} \; X^*_{cji}<\tau_{1,ci} \\
-% \qquad 1 \qquad \text{if} \; \tau_{1,ci} \leq X^*_{cji}<\tau_{2,ci} \\
-% \qquad 2 \qquad \text{if} \; X^*_{cji}\geq\tau_{2,ci}
-% \end{cases}
-% \end{equation}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-The model specification above cannot identify the parameters unless additional normalisations are imposed.
-A configural model is defined as a minimal set of restrictions that ensure identification of parameters across groups.
-There are infinite, equivalent ways in which the configural model can be parameterised -- the well-known issue of factor indeterminacy.
-Widely used parameterisations for the configural model are:
+There are infinite ways to parameterise the configural model defined by \eqref{eq:propens} and \eqref{eq:thresh}. Widely used parameterisations are:
 \bi[label=$\diamond$]
@@
 
-Once the configural model has been estimated, measurement invariance analysis proceeds by estimating nested models that place increasing restrictions on the item parameters, by constraining them to be equal across groups $c$.
-In the case where available measures are continuous, MI analysis is straightforward \citep{vandeSchoot2012}: one set of parameters is restricted at a time, and the fit of the resulting models are compared to the configural model.
-The hierarchy of the nested models usually proceeds by testing loadings first, and then intercepts (to establish \emph{metric} and \emph{scalar} invariance -- see \citealp{Vandenberg2000}).
+\subsection{Identification of models with different levels of invariance}
 
-Invariance of systems with categorical measures is less well understood.
-In particular, the lack of explicit location and scale in the measures introduces an additional set of parameters (thresholds $\tau$) and makes identification reliant on more stringent normalisations.
-The first comprehensive approach for categorical measures is proposed by \cite{Millsap2004}.
+In the case where available measures are continuous, MI analysis is straightforward \citep{vandeSchoot2012}.
+The hierarchy of the nested models usually proceeds by testing loadings first, and then intercepts (to establish \emph{metric} and \emph{scalar} invariance -- see \citealp{Vandenberg2000a}).
+
+Invariance of systems with categorical measures, such as the scale we examine in this paper, is less well understood.
+In particular, the lack of explicit location and scale in the measures introduces an additional set of parameters compared to the continuous case (thresholds $\tau$).
+This makes identification reliant on more stringent normalisations.
+A first comprehensive approach for categorical measures was proposed by \cite{Millsap2004}.
 New identification results in \cite{Wu2016a} indicate that, in the categorical case, invariance properties cannot be examined by simply restricting one set of parameters at a time.
@@
 Furthermore, they suggest that comparison of both latent means and variances requires invariance in loadings, thresholds, and intercepts.
-
-{\color{red} ADD SOME MORE COMMENTS ON IDENTIFICATION HERE, to make everything more palatable to economics audience.}
-
-We estimate the sequence of models detailed in \autoref{tab:invparam} by Weighted Least Squares.\footnote{Parameters are estimated by mean- and variance-adjusted weighted least squares (WLSMV) (see \citealp{Muthen2017a}); estimation starts from the items' polychoric correlation matrix, uses diagonally weighted least squares (DWLS), and exploits the full weight matrix to compute robust standard errors and test statistics.
-Robust WLS has proved in simulation studies to be moderately robust to small violations of the normality assumption in the latent underlying measures \citep{Flora2004}, and generally outperforms maximum likelihood in large samples \citep{Beauducel2006,Li2016}.
-All estimates are computed using the \texttt{lavaan} package in \textbf{\textsf{R}} \citep{Rosseel2012}.}
-For the purposes of the analysis, we define groups $c$ as cohort-gender cells, with the reference group being males in the BCS (1970) cohort.
-We then compare the fit of each model against the configural model (i.e.
-the least restrictive).
-According to recommended practice, we diplay a range of fit indices.
-Comparison of $\chi^2$ values across models is the most common strategy.
-However, tests based on $\Delta \chi^2$ are known to display increasing Type I error rates with sample size and model complexity \citep{Sass2014a}.
-The use of approximate fit indices (AFIs) has been proposed alongside $\chi^2$ to overcome these limitation.
-These indices do not have a known sampling distribution, thus making it necessary to rely on rules of thumb to assess what level of \textDelta AFI indicates invariance.
-Nevertheless, AFIs provide valuable corrections for model parsimony.\footnote{Commonly adopted AFIs in studies of measurement invariance are the root mean squared error of approximation (RMSEA) and the Tucker-Lewis index (TLI).
-Simulation evidence by \cite{Cheung2002} shows that these indices can show correlation between overall and relative fit, and suggest relying on additional indices, such as the comparative fit index (CFI, \citealp{Bentler1990}), McDonald non-centrality index (MFI, \citealp{McDonald1989}), and Gamma-hat index \citep{Steiger1989} for the case of ordered measures.
-Commonly accepted thresholds for rejection are $\Delta CFI < - 0.01$, $\Delta MFI < - 0.02$, and $\Delta Gamma hat < - 0.001$.
-\cite{Meade2008}, using the results from a simulation study, suggests stricter thresholds that should apply in a variety of conditions.
-For CFI, a single cutoff value of $.002$ is proposed, while cutoffs for MFI depend on the problem's characteristics; in our case (2 factors, 11 items), they suggest $.0066$.
-\cite{Sass2014a} however cast some doubts of the generalisability of these cutoffs to WLSMV estimators.}
-
-{\color{red} Motivation for choosing RMSEA (does not seem supported in literature, see above).}
-
-The fit of each estimated model using our full sample is compared in Panel A of \autoref{tab:fit}.\footnote{We do not present fit results for the threshold invariance model, as it is statistically equivalent to the configural model and thus its fit is mathematically the same.} The overall fit of the configural model is satisfactory according to widely accepted rules of thumb, with RMSEA close to $0.05$ and CFI close to $.95$ \citep{Hu1999}.
-For all invariance levels, a chi-squared difference would point to a lack of measurement invariance.
-However, as detailed above, this test is expected to lead to substantial Type I error with a relatively large sample size such as ours.
-For the model with restricted thresholds and loadings, most of the \textDelta AFIs indicate that this level of invariance is justified in our data.
-However, further restricting intercepts results in a model where invariance is rejected across the board.
-
-\begin{table}[]
-\centering\setstretch{1.2} \footnotesize
-\caption{Parameterisations for measurement invariance}\label{tab:invparam}
-\begin{tabular}{L{.18\textwidth}L{.45\textwidth}L{.3\textwidth}}
-\toprule
-\textbf{Invariance level}     & \textbf{Description} & \textbf{Restrictions} \\ \midrule \\
-Configural (WE{\textTheta})   
-    & \bi[label=$\cdot$] \item Minimally restrictive model for identification \ei
-    & For all groups: $\begin{array}{|l} \text{diag}(\bm{\Phi}) = \bm{I} \\
-                      \bm{\kappa}=\bm{0} \\
-                      \bm{\nu}=\bm{0} \\
-                      \text{diag}(\bm{\Psi})=\bm{I} \end{array}$
-                      \\ \\
-Threshold invariance
-    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ to be equal across groups
-                  \item Statistically equivalent to configural (when measures have 3 categories or less)\ei
-    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
-                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
-                      \newline For all groups: $\begin{array}{|l} \text{diag}(\bm{\Phi}) = \bm{I} \\ 
-                                                \bm{\kappa}=\bm{0} \end{array}$
-                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\nu}_A=\bm{0} \\
-                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \end{array}$
-                   \\ \\
-Threshold and Loading invariance
-    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ and loadings $\lambda$ to be equal across groups
-                  \item Allows comparison of latent factor variances \ei
-    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
-                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
-                      \newline $\lambda_{ci} = \lambda_{c^\prime i}$ for all items, $\forall c, c^\prime$
-                      \newline For all groups: $\bm{\kappa}=\bm{0}$
-                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\nu}_A=\bm{0} \\
-                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \\
-                                                  \text{diag}(\bm{\Phi}_A) = \bm{I} \end{array}$ 
-                   \\ \\
-Threshold, Loading, and Intercept invariance
-    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ and loadings $\lambda$ to be equal across groups
-                  \item Restricts intercepts $\nu$ to zero in both groups
-                  \item Allows comparison of latent factor variances \emph{and} means \ei
-    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
-                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
-                      \newline $\lambda_{ci} = \lambda_{c^\prime i}$ for all items, $\forall c, c^\prime$
-                      \newline For all groups: $\bm{\nu}=\bm{0}$
-                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\kappa}_A=\bm{0} \\
-                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \\
-                                                  \text{diag}(\bm{\Phi}_A) = \bm{I} \end{array}$ 
-\\ \\ \bottomrule
-\end{tabular}
-\end{table}
-
-\clearpage
-\bibliographystyle{chicago}
-\bibliography{/Users/giacomomason/Dropbox/Zotero/cohorts.bib}
-
-\clearpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
-
-\begin{landscape}
-\begin{table}[]
-\caption{Item wording and prevalence, by cohort -- females}\label{tab:meantab_f}
-\centering\setstretch{1.2} \footnotesize
-<<MEANTAB_F, dependson="MEANTABLE", results='asis'>>=
-require(xtable)
-table <- meantab[[2]] # females
-table <- table[,!names(table) %in% c("bcs.num", "mcs.num")] # drop if unneeded
-printmeantab(table)
-@
-\end{table}
-\end{landscape}
-
-\begin{landscape}
-\begin{table}[]
-\caption{Item wording and prevalence, by cohort -- males}\label{tab:meantab_m}
-\centering\setstretch{1.2} \footnotesize
-<<MEANTAB_M, dependson="MEANTABLE", results='asis'>>=
-require(xtable)
-table <- meantab[[1]] # males
-table <- table[,!names(table) %in% c("bcs.num", "mcs.num")] # drop if unneeded
-printmeantab(table)
-@
-\end{table}
-\end{landscape}
-
-\begin{landscape}
-\begin{table}[ht]
-\caption{Measurement invariance fit comparison}\label{tab:fit}
-\centering\setstretch{1.2} \footnotesize
-<<FA_FIT_RES, dependson="FA_FIT", results='asis'>>=
-require(xtable)
-
-table <- rbind(afitab[[1]],afitab[[2]])
-
-print(xtable(table,
-                    align="L{0cm}L{.20\\textwidth}C{.05\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.08\\textwidth}",
-                    digits = 4
-                    ),
-             size="\\small",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = c(list(0,3)),
-                              command = c(
-                              # header
-            "& & \\multicolumn{6}{c}{\\textbf{Absolute fit}} & \\multicolumn{6}{c}{\\textbf{Relative fit}} 
-                  \\\\ \\cmidrule(lr){3-8} \\cmidrule(lr){9-14}
-            Model & Num. params & $\\chi^2$ & RMSEA & SRMR & MFI & CFI & G-hat & $\\chi^2$ $p$ & \\textDelta RMSEA & \\textDelta SRMR &  \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ 
-& (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8) & (9) & (10) & (11) & (12) & (13) \\\\ \\midrule
-                       \\textbf{A: All ages} \\\\",
-                      "\\newline \\textbf{B: 59-61 months sample} \\\\ \\midrule "
-                                    )),
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabular", width = "\\textwidth",
-             floating = FALSE
-)
-
-@
-\justify {\scriptsize \emph{Notes}: The table presents fit indices for models of different invariance levels, following \cite{Wu2016a} Col.
-(1) displays the number of estimated parameters for each model.
-Col.
-(2) and (8) present the value of the $\chi^2$ statistic and the pvalue of the test of equality with respect to the configural model.
-Col.
-(3)-(7) and (9)-(13) present alternative fit indices (AFIs), in absolute values and differences from the configural model respectively.
-RMSEA = Root mean squared error of approximation; SRMR = standardised root mean residual; MFI = McDonald non-centrality index; CFI = comparative fit index; G-hat = gamma-hat.
-Panel A shows results for the whole sample of children in the BCS and MCS cohorts.
-Panel B is restricted to a subsample of children in the age range of maximum overlap between the two cohorts -- 59-61 months.
-\par}
-\end{table}
-\end{landscape}
-
-\end{document}
+A summary of the approach by \cite{Wu2016a} is available in \autoref{tab:invparam}.

### 2018-10-01 codes/R/paper/text/paper_jpube_5apptables.tex
add text from coauthors

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{table}[ht!]
\caption{Behavioural screening scales in the BCS and MCS five-year surveys}\label{tab:scales}
\centering\setstretch{1.2} \footnotesize
\begin{tabular}{L{.46\linewidth}L{.001\linewidth}L{.46\linewidth}}
\toprule
\multicolumn{3}{l}{\textbf{Panel A: Rutter A Scale \citep{Rutter1970}} -- British Cohort Study (1975) five-year survey} \\[-1.8em]
\begin{enumerate}
\item Very restless. Often running about or jumping up and down. Hardly ever still.$^\dagger$
\item Is squirmy or fidgety.$^\dagger$
\item Often destroys own or others' belongings.
\item Frequently fights other children.$^\dagger$
\item Not much liked by other children.
\item Often worried, worries about many things.$^\dagger$
\item Tends to do things on his/her own, is rather solitary.$^\dagger$
\item Irritable. Is quick to fly off the handle.
\item Often appears miserable, unhappy, tearful or distressed.$^\dagger$
\item Sometimes takes things belonging to others.
\item Has twitches, mannerisms or tics of the face or body.
\item Frequently sucks thumb or finger.
\end{enumerate}
& &
\begin{enumerate}\setcounter{enumi}{12}
\item Frequently bites nails or fingers.
\item Is often disobedient.$^\dagger$
\item Cannot settle to anything for more than a few moments.$^\dagger$
\item Tends to be fearful or afraid of new things or new situations.$^\dagger$
\item Is over fussy or over particular.
\item Often tells lies.
\item Bullies other children.$^\dagger$
\item[A.] Complains of headaches.$^\dagger$
\item[B.] Complains of stomach-ache or has vomited.$^\dagger$
\item[C.] Complains of biliousness.
\item[D.] Has temper tantrums (that is, complete loss of temper with shouting, angry movements, etc.).$^\dagger$
\end{enumerate}
\\
\midrule
\multicolumn{3}{l}{\textbf{Panel B: Strength and Difficulties Questionnaire \citep{Goodman1997}} -- Millennium Cohort Study (2000/1) five-year survey} \\[-1.8em]
\begin{enumerate}
\item Considerate of other people's feelings.
\item Restless, overactive, cannot stay still for long.$^\dagger$
\item Often complains of head- aches, stomach-ache or sickness.$^\dagger$
\item Shares readily with other children (treats, toys, pencils, etc.).$^+$
\item Often has temper tantrums or hot tempers.$^\dagger$
\item Rather solitary, tends to play alone.$^\dagger$
\item Generally obedient, usually does what adults request.$^{\dagger +}$
\item Many worries, often seems worried.$^\dagger$
\item Helpful if someone is hurt, upset or feeling ill.$^+$
\item Constantly fidgeting or squirming.$^\dagger$
\item Has at least one good friend.$^+$
\item Often fights with other children or bullies them.$^\dagger$
\item Often unhappy, down-hearted or tearful.$^\dagger$
\end{enumerate}
& &
\begin{enumerate}\setcounter{enumi}{13}
\item Generally liked by other children.$^+$
\item Easily distracted, concentration wanders.$^\dagger$
\item Nervous or clingy in new situations, easily loses confidence.$^\dagger$
\item Kind to younger children.$^+$
\item Often lies or cheats.
\item Picked on or bullied by other children.
\item Often volunteers to help others (parents, teachers, other children).$^+$
\item Thinks things out before acting.$^+$
\item Steals from home, school or elsewhere.
\item Gets on better with adults than with other children.
\item Many fears, easily scared.
\item Sees tasks through to the end, good attention span.$^+$
\end{enumerate}
\\
\bottomrule
\end{tabular}
\justify {\scriptsize \emph{Notes}: Items denoted by $^+$ are positively coded in the original scale. Items denoted by $^\dagger$ are retained in the new comparable scale.
\par}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{landscape}
\begin{table}[ht!]
\caption{Description of harmonised variables}\label{tab:harmvar}
\centering\setstretch{1.2} \footnotesize
\begin{tabular}{L{.2\textwidth}C{.1\textwidth}L{.3\textwidth}L{.7\textwidth}}
\toprule
Variable Group & Age & Variable & Note \\
\midrule
Maternal education & 5 & Post-compulsory schooling$^d$ & Whether mother continued schooling past the compulsory age, based on her year of birth. School leaving age in England was changed from 14 to 15 in 1947 and from 15 to 16 in 1972. \\ \\[-1.5em]
Maternal employment & 5 & Employed$^d$ & Includes full time and part time \\ \\[-1.5em]
Father occupation & 5 & White collar (I-IIINM)$^d$ \newline Blue collar (IIIM-V-other)$^d$ \newline No father figure$^d$ & Based on father's Registrar General Social Class classification of occupations. White collar includes I (Professional), II (Managerial/technical), IIINM (Skilled non-manual). Blue collar includes IIIM (Skilled manual), IV (Partly skilled), V (Unskilled), Other, Unemployed, and Armed forces. No father figure is a dummy for children whose father does not live in the same household. \\ \\[-1.5em]
Maternal background & 0/5 & Mother's age at birth \newline Mother's height (cm) \newline Mother unmarried at birth$^d$ \newline Child nonwhite ethnicity$^d$  \newline Number of children in HH & All variables are self-reported by the mother at birth, except for number of children in household (at five years old). Unmarried is only based on marital status, and includes cohabitation. \\ \\[-1.5em]
Pregnancy & 0 & Child is firstborn$^d$ \newline Number of previous stillbirths \newline Mother smoked in pregnancy$^d$ \newline Preterm birth (under 37 weeks gestation)$^d$ \newline (log) birth weight (kg) & Parity, stillbirths, and smoking are self-reported by the mother. Gestational length and birth weight are from hospital records. \\ \\[-1.5em]
Cognitive skills & 5 & & Based on test batteries administered to the cohort member at five. Three tests are used for BCS children: Copy Designs child is asked to copy simple designs adjacently), Human Figure Drawing (child draws an entire human figure), English Picture Vocabulary Test (child identifies the picture referring to a word among four pictures). Three \emph{different} tests are used in the MCS: BAS Naming Vocabulary (child is shown a series of pictures and asked to name it), BAS Picture Similarity (child is shown a row of 4 pictures on a page and places a card with a fifth picture under the one most similar to it), BAS Pattern Construction (child constructs a design by putting together flat squares or solid cubes with patterns on each side).

 \\ \\[-1.5em]
Adolescent outcomes & 16 (BCS) \newline 14 (MCS) & Child tried smoking$^d$ \newline Body Mass Index (BMI) & Smoking is self reported by the child. Height and weight are taken as part of a medical examination. \\ \\[-1.5em]
Adult outcomes (BCS only) & 34 \newline 42 \newline 34, 42 & Higher education$^d$ \newline Employed$^d$ \newline (log) gross weekly pay & Higher education is defined on having a university degree or its vocational equivalent. It corresponds to level 4 or 5 in the National Vocational Qualification (NVQ) equivalence. Employed is a dummy for being in paid employment or self-employment, either full or part time. Gross weekly pay is weekly pre-tax pay from the respondent's main activity, conditional on being a paid employee. \\
\bottomrule
\end{tabular}
\justify {\scriptsize \emph{Notes}: Variables denoted by $^d$ are binary or categorical.
\par}
\end{table}
\end{landscape}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{table}[ht]
\centering\setstretch{1.2} \footnotesize
\caption{Parameterisations for measurement invariance}\label{tab:invparam}
\begin{tabular}{L{.18\textwidth}L{.45\textwidth}L{.3\textwidth}}
\toprule
\textbf{Invariance level}     & \textbf{Description} & \textbf{Restrictions} \\ \midrule \\
Configural (WE{\textTheta})   
    & \bi[label=$\cdot$] \item Minimally restrictive model for identification \ei
    & For all groups: $\begin{array}{|l} \text{diag}(\bm{\Phi}) = \bm{I} \\
                      \bm{\kappa}=\bm{0} \\
                      \bm{\nu}=\bm{0} \\
                      \text{diag}(\bm{\Psi})=\bm{I} \end{array}$
                      \\ \\
Threshold invariance
    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ to be equal across groups
                  \item Statistically equivalent to configural (when measures have 3 categories or less)\ei
    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
                      \newline For all groups: $\begin{array}{|l} \text{diag}(\bm{\Phi}) = \bm{I} \\ 
                                                \bm{\kappa}=\bm{0} \end{array}$
                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\nu}_A=\bm{0} \\
                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \end{array}$
                   \\ \\
Threshold and Loading invariance
    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ and loadings $\lambda$ to be equal across groups
                  \item Allows comparison of latent factor variances \ei
    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
                      \newline $\lambda_{ci} = \lambda_{c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline For all groups: $\bm{\kappa}=\bm{0}$
                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\nu}_A=\bm{0} \\
                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \\
                                                  \text{diag}(\bm{\Phi}_A) = \bm{I} \end{array}$ 
                   \\ \\
Threshold, Loading, and Intercept invariance
    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ and loadings $\lambda$ to be equal across groups
                  \item Restricts intercepts $\nu$ to zero in both groups
                  \item Allows comparison of latent factor variances \emph{and} means \ei
    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
                      \newline $\lambda_{ci} = \lambda_{c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline For all groups: $\bm{\nu}=\bm{0}$
                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\kappa}_A=\bm{0} \\
                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \\
                                                  \text{diag}(\bm{\Phi}_A) = \bm{I} \end{array}$ 
\\ \\ \bottomrule
\end{tabular}
\justify {\scriptsize \emph{Notes}: Adapted from \cite{Wu2016a}.
\par}
\end{table}

### 2018-10-01 codes/R/paper/text/paper_jpube_1introlit.tex
fix bibliography

\section{Introduction}

Human capital is a key determinant of economic growth and performance and of the resources an individual creates and controls over the life cycle. Human capital is also important for various determinants of individual well-being, ranging from health to life satisfaction.  In recent years, the process of human capital accumulation has received considerable attention. There is growing consensus on the fact that human capital is a multidimensional object, with different domains playing different roles in labour market as well as in the determination of other outcomes, including the process of human development. It is also recognised that human capital is the output of a very persistent process, where early years inputs play an important and persistent role.

And yet, there are still large gaps in our knowledge of the process of human capital development. These gaps are partly driven by the scarcity of high quality longitudinal data measuring the evolution over the life cycle of different dimension of human capital. Moreover, there is a lack of consensus on the best measures and on the tools to collect high quality data. As a consequence, even when data are available in different contexts, their comparability is problematic.

In this paper, we focus on an important dimension of human capital, which, so far, has received limited attention: socio-emotional skills. Evidence has shown that gaps in socio-emotional skills emerge at very young ages, and that in the absence of interventions are very persistent across the life cycle \citep{Cunha2006}. However, there is surprisingly little evidence on how inequality in this important dimension of human capital has changed across cohorts. In this paper, we start addressing this gap and focus on the measurement of these skills in two British cohorts: the one of children born in 1970 and the one of children born in 2000. We consider the measurement of socio-emotional skills during early childhood, as these skills have been shown, in a variety of contexts \citep{Almlund2011} to have important long-run effects. Our goal is to characterise the distribution of socio-emotional skills in these cohorts and compare them. In the last part of the paper, we also consider the predictive power of different socio-emotional skills for health and socioeconomic outcomes.

The main contributions of the paper are four. First, we use two validated scales of childhood behavioural traits and select those items which are comparable across the two cohorts. By performing an exploratory factor analysis, we determine that we need at least two dimensions to characterise socio-emotional skills. We label them as `internalising' and `externalising' skills, the former relating to the ability of children to focus their drive and determination and the latter relating to their ability in engaging in interpersonal activities.

Second, we study the comparability of the measures in the two cohorts. In particular, we test for \emph{measurement invariance} of the items we use to estimate the latent factors. Intuitively, if one assumes that a set measures is related to a latent unobserved factor of interest, one can think of this relationship as being driven by the saliency of each measure and the level. If one uses a given measure as the relevant metric for the relevant factor, its saliency will determine the scale of the factor, while some other parameters, which could be driven by the difficulty of a given test or the social norms and attitudes towards a certain type of behaviour, determine the \emph{average level} of the factor. Comparability of estimated factors across different groups (such as different cohorts) assumes  that both the parameters that determine the saliency of a given set of measures and the level of the factors do not vary across groups. We find that, for the measures we use and for both factors, we cannot reject measurement invariance for the saliency parameters. However, we strongly reject measurement invariance for the level parameters. These results imply that while the level of inequality across the two cohorts in the skills we consider is comparable, we cannot determine whether the \emph{average levels} of the two factors are larger or smaller in one of the two cohorts.

Third, given the results we obtain on measurement invariance, we proceed to compare the level of inequality in the two types of socio-emotional skills across the two cohorts, for both boys and girls. We find that the most recent cohort is more unequal in both dimensions of socio-emotional skills than the 1970 cohort. This result is particularly apparent for boys, and when looking at differences by maternal background. Fourth, we study whether the socio-emotional skills we observe at a young age are an important determinant of a variety of adolescent (and adult, for the older BCS cohort) outcomes. We find that socio-emotional skills at age five are more predictive than cognitive skills for unhealthy behaviours like smoking and measures of health capital such as body mass index. The effect of cognition, instead, dominates for educational and labour market outcomes.

The rest of the paper is organised as follows. In section \ref{sec:data}, we briefly discuss the data we use in the analysis. In section \ref{sec:methods}, we discuss the methods we use to identify the number of dimensions in socio-emotional skills and how we estimate the latent factors that represent them. In section \ref{sec:measinv}, we discuss the comparability of factors estimated with a given set of measures from different groups and the \emph{measurement invariance} tests we use. Section \ref{sec:results} reports our empirical results, while section \ref{sec:conclusions} concludes the paper.

\section{Literature \label{sec:lit}}

The importance of cognition in predicting life course success is well established in the economics literature. However, in recent years the role played by `non-cognitive' traits is being increasingly investigated. These traits include constructs as different as psychological and preference parameters such as social and emotional skills, locus of control and self-esteem, personality traits (e.g. conscientiousness), and risk aversion and time preferences. Given the vastity of this literature, we briefly review below the main papers on the determinants and consequences of socio-emotional traits which are more directly related to our work, and we refer to other sources for more exhaustive reviews \citep{Borghans2008,Almlund2011,Goodman2015,Kautz2014}.

\paragraph{Consequences of socio-emotional traits} One of the first papers to show the importance of non-cognitive personality variables for wages was \cite{Bowles2001}. \cite{Heckman2006} suggested that non-cognitive skills are at least as important as cognitive abilities in determining a variety of adults outcomes. \citet{Lindqvist2011}, using data based on personal interviews conducted by a psychologist during the Swedish military enlistment exam, show that both cognitive and noncognitive abilities are important in the labour market, but for different outcomes: low noncognitive abilities are more correlated with unemployment or low earnings, while cognitive ability is a stronger predictor of wages for skilled workers. \citet{Segal2013}, using data on young men from the US National Education Longitudinal Survey, shows that eight-grade misbehaviour is important for earnings over and above eight-grade test scores. \citet{Layard2014} find that childhood emotional health (operationalised using the same mother-reported Rutter scale we use in the 1970 British cohort study) at ages 5, 10 and 16 is the most important predictor of adulthood life satisfaction and life course success.

There are only few studies in economics specifically studying ``non-cognitive'' traits and health behaviors. \citet{Conti2010a} and \citet{Conti2011} were the first to consider three early endowments, including child socio-emotional traits and health in addition to cognition, using rich data from the 1970 British cohort study. They find strong evidence that non-cognitive traits promote health outcomes and healthy behaviors, and than not accounting for them overestimates the effects of cognition; additionally, they document that child cognitive traits are more important predictors of employment and wages than socio-emotional traits or early health. \citet{Chiteji2010} used the US Panel Study of Income Dynamics (PSID) and found that future orientation and self-efficacy (related to emotional stability) are associated with less alcohol consumption and more exercise. \citet{Cobb-Clark2014} used the Australian HILDA data and found that an internal locus of control (also related to emotional stability, perceived control over one's life) is related to better health behaviours (diet, exercise, alcohol consumption and smoking). \citet{Mendolia2014a} used the Longitudinal Study of Young People in England and found that individuals with external locus of control, low self-esteem, and low levels of work ethics, are more likely to engage in risky health behaviours. \citet{Savelyev2017} show that the association between personality traits and health behaviours also holds in a high-IQ sample (the Terman Sample). \citet{Heckman2016a} use, instead, early risky behaviours to measure socio-emotional traits, and confirm their predictive power for health behaviours and health outcomes.

Very few papers attempt to make cross-cohorts comparisons about the importance of socio-emotional skills. \citet{Blanden2007} -- one of the closes study to ours -- examine cognitive skills, non‐cognitive traits, educational attainment and labour market attachment as mediators of the decline in inter-generational mobility in UK between the 1958 and the 1970 cohorts. The authors take great care in selecting non-cognitive items to be as comparable as possible across cohorts, from the Rutter scale at age 10 for the 1970 cohort and from the Bristol Social Adjustment Guide for the 1958 cohort; however, they do not carry out a formal test of measurement invariance and the do not construct factor scores fully comparable across cohorts as we do. Another paper related to ours is the one by \citet{Reardon2016}, who study recent trends in income, racial, and ethnic school gaps in several dimensions of school readiness, including academic achievement, self-control, and externalizing behavior, at
kindergarten entry, using comparable data from the Early Childhood Longitudinal Studies (ECLS-K and ECLS-B) for cohorts born from the
early 1990s to the 2000–2010 period in the US. They find that readiness gaps narrowed modestly from 1998 to 2010, particularly between high- and low-income students and between White and Hispanic students. Lastly, \citet{Deming2017} uses a comparable set of skill measures and covariates across survey waves for the NLSY79 and the NLSY97, and finds that the labour market return to social skills was much greater in the 2000s than in the mid-1980s and 1990s.

\paragraph{Determinants of socio-emotional traits} Equally flourishing has been the literature on the determinants of child socio-emotional skills, which ranges from reduced-form, correlational or causal estimates, to more structural approaches. One of the first papers by \citet{Segal2008} has shown that a variety of family and school characteristics predict classroom behaviour. \citet{Carneiro2013} study the intergenerational impacts of maternal education, using data from the NLSY79 and an instrumental variable strategy; they find strong effects in terms of reduction in children's behavioural problems. \citet{Cunha2010} and \citet{Attanasio2018a} both estimate production functions for child cognitive and socio-emotional development (in US and Colombia, respectively), and find an important role played by parental investments.

Interventions for improving Social and Emotional Learning (SEL) in a school setting have shown significant improvements in socio-emotional skills, attitudes, behaviours, and academic performance \citep{Durlak2011}, and a substantial positive return on investments \citep{Belfield2015}; after-school programs have been shown to be equally effective \citep{Durlak2010}.

Additionally, it has been shown that a key mechanism through which early childhood interventions improve adult socioeconomic and health outcomes is by boosting socio-emotional skills, such as four teacher-reported behavioural outcomes in the project STAR\footnote{Student's effort, initiative, non-participatory  behavior, and how the student is seen to `value' the class.} \citep{Chetty2011a}, reductions in externalising behaviour (from the Pupil Behavior Inventory) at ages 7-9 in the Perry Preschool Project \citep{Heckman2013,Conti2016a}, or improvements in task orientation at ages 1-2 in the Abecedarian Project \citep{Conti2016a}.

In sum, even if the literature on the determinants and consequences of socio-emotional skills has been booming, most papers use skills measured in late childhood or in adolescence; and no paper in economics formally tests for invariance of measurements across different groups and constructs fully comparable scores. In this paper, we use measures of child socio-emotional development at age 5, hence before formal schooling starts; and we construct comparable scales across the two cohorts we study (the 1970 and the 2000 British cohorts), so that we can investigate changes in inequality in early development, their determinants, and consequences, in a parallel fashion.

### 2018-10-01 codes/R/paper/text/paper_jpube_2datamethods.tex
fix bibliography

@@
 In the child psychiatry and psychology literatures, the Rutter and SDQ behavioural screening scales are regarded as measures of behavioural problems and mental health.
-However, in our analysis we follow the economics literature, and - after having recoded them accordingly - we interpret them as measures of positive child development \citep{goodman2011population}.
+However, in our analysis we follow the economics literature, and - after having recoded them accordingly - we interpret them as measures of positive child development \citep{Goodman2011}.
 
@@
 The original Rutter scale, used for the BCS cohort, distinguishes behaviours into two subscales: \emph{anti-social} and \emph{neurotic} \citep{Rutter1970}.
-This two-factor conceptualisation has been validated using data from multiple contexts, and the latent dimensions have been broadly identified as externalising and internalising behaviour problems.\footnote{See for example \cite{Fowler1979,Venables1983,Tremblay1987,Berglund1999,Klein2009}.
+This two-factor conceptualisation has been validated using data from multiple contexts, and the latent dimensions have been broadly identified as externalising and internalising behaviour problems.\footnote{See for example \citet{Fowler1979,Venables1983,Tremblay1987,Berglund1999,Klein2009}.
 However, in some cases a three-factor structure was found to better fit the data, with the externalising factor separating into two factors seemingly capturing aggressive and hyperactive behaviours \citep{Behar1974,McGee1985}.} The Strength and Difficulties Questionnaire, used for the MCS cohort, was conceived to have five subscales of five items each.
@@
 The results are displayed in \autoref{tab:numfac}.
-As pointed out in \citet{conti2014bayesian}, there is relatively little agreement among procedures; this is the case especially for the Rutter items in the BCS data, where different methods suggest to retain between 1 and 3 factors, while most methods suggest to retain 2 factors in the MCS.
+As pointed out in \citet{Conti2014}, there is relatively little agreement among procedures; this is the case especially for the Rutter items in the BCS data, where different methods suggest to retain between 1 and 3 factors, while most methods suggest to retain 2 factors in the MCS.
 In our analysis, we adopt two factors and a dedicated measurement system, where each measure reflects only one factor.
@@
 
-We make the common assumption of a dedicated (or congeneric) factor structure, where each measure is assumed to load on only one latent dimension \citep{Heckman2013,conti2010education,Attanasio2017c}.
+We make the common assumption of a dedicated (or congeneric) factor structure, where each measure is assumed to load on only one latent dimension \citep{Heckman2013,Conti2010a,Attanasio2018a}.
 We mirror the structure found in the exploratory factor analysis (see \autoref{tab:load}), and assume that items 1-6 load exclusively on the externalising factor and items 7-11 on the internalising factor.\footnote{The dedicated factor structure corresponds to a sparse loading matrix, i.e.: $$\bm{\Lambda}_{c} \coloneqq \begin{bmatrix} \lambda_{1c}, \dots, \lambda_{6c} & \bm{0} \\ \bm{0} & \lambda_{7c}, \dots, \lambda_{11c} \end{bmatrix}.$$}
@@
 As per the traditional factor analysis approach, we impose a normal distribution on the latent skills and error terms:\footnote{Recent work has also used mixtures of normals for the latent factors distribution, e.g.
-\citet{conti2010education}.}
+\citet{Conti2010a}.}
 \be\label{eq:norm}
@@
 In the framework of factor analysis, measurement invariance is a formally testable property.
-In this paper, we follow the recent identification methodology by \cite{Wu2016a}.
+In this paper, we follow the recent identification methodology by \citet{Wu2016a}.
 The configural model defined in the previous section serves as the starting point.
@@
 Nevertheless, AFIs are widely used in empirical practice to assess model fit.\footnote{The root mean squared error of approximation (RMSE) and the Tucker-Lewis index (TLI) are traditionally the most used AFIs in empirical practice.
-Simulation evidence by \cite{Cheung2002} shows that these indices can show correlation between overall and relative fit, and suggest relying on additional indices, such as the comparative fit index (CFI, \citealp{Bentler1990}), McDonald non-centrality index (MFI, \citealp{McDonald1989}), and Gamma-hat index \citep{Steiger1989} for the case of ordered measures.
+Simulation evidence by \citet{Cheung2002} shows that these indices can show correlation between overall and relative fit, and suggest relying on additional indices, such as the comparative fit index (CFI, \citealp{Bentler1990}), McDonald non-centrality index (MFI, \citealp{McDonald1989}), and Gamma-hat index \citep{Steiger1989} for the case of ordered measures.
 Commonly accepted thresholds for rejection are $\Delta CFI < - 0.01$, $\Delta MFI < - 0.02$, and $\Delta Gamma hat < - 0.001$.
-\cite{Meade2008}, using the results from a simulation study, suggests stricter thresholds that should apply in a variety of conditions.
+\citet{Meade2008}, using the results from a simulation study, suggests stricter thresholds that should apply in a variety of conditions.
 For CFI, a single cutoff value of $.002$ is proposed, while cutoffs for MFI depend on the problem's characteristics; in our case (2 factors, 11 items), they suggest $.0066$.
-\cite{Sass2014} however cast some doubts of the generalisability of these cutoffs to WLSMV estimators.}
+\citet{Sass2014} however cast some doubts of the generalisability of these cutoffs to WLSMV estimators.}
 
@@
 For example, commonly held views of what constitutes a restless, distracted, or unhappy child might have changed between 1975 and 2005/6.\footnote{Calibrating the Rutter and SDQ using a contemporary sample of children cannot rule out this issue.
-For example, \cite{Collishaw2004} administered both Rutter and SDQ items to parents of a small sample of adolescents in London.
+For example, \citet{Collishaw2004} administered both Rutter and SDQ items to parents of a small sample of adolescents in London.
 They use the mapping between the two questionnaires to impute Rutter scores for mothers who answered the SDQ.

### 2018-10-01 codes/R/paper/text/paper_jpube_3resultsconclusion.tex
fix bibliography

@@
 Data and prior are combined into the posterior distribution $w(\bm{\theta}| \bm{X}, \hat{\bm{\Omega}})$.
-For further details, see Chapter 7 in \cite{Skrondal2004}.} We plot the distribution of the scores in \autoref{fig:dens}.
+For further details, see Chapter 7 in \citet{Skrondal2004}.} We plot the distribution of the scores in \autoref{fig:dens}.
 The unit of measurement is standard deviations of the distribution in the subsample of males in the BCS.
@@
 \autoref{fig:ineqsm} shows an even starker pattern when comparing children of mothers who smoked in pregnancy with non-smoking mothers.
-The fact that maternal smoking during pregnancy is a risk factor for offspring behavioural problems is well known in the medical literature \citep{gaysina2013maternal}; less evidence there is however, on whether and to which extent these associations have changed across cohorts.
+The fact that maternal smoking during pregnancy is a risk factor for offspring behavioural problems is well known in the medical literature \citep{Gaysina2013}; less evidence there is however, on whether and to which extent these associations have changed across cohorts.
 The difference in child skills has increased, from less than .2 to around .4 of a standard deviation, again with the biggest increase experienced by the boys.
@@
 The increase in inequality across cohorts is still present, but less marked when looking at these single items.
-This shows the importance of the factor analysis step in aggregating items, explicitly modelling the measurement error, and testing and accounting for (loadings and thresholds) invariance across the two cohorts.} These patterns are in stark contrast with the findings of \citet{reardon2016recent} for the US, who have found a narrowing of the readiness gaps from 1998 to 2010.
+This shows the importance of the factor analysis step in aggregating items, explicitly modelling the measurement error, and testing and accounting for (loadings and thresholds) invariance across the two cohorts.} These patterns are in stark contrast with the findings of \citet{Reardon2016} for the US, who have found a narrowing of the readiness gaps from 1998 to 2010.
 
@@
 Being born to an unmarried mother, and to a mother who smoked during pregnancy, is associated with a higher penalty for both dimensions of socio-emotional skills in the latter cohort, but only for males.
-This is consistent with recent evidence which shows that family disadvantage disproportionately impedes the pre-market development of boys, in terms of higher disciplinary problems, lower achievement scores, and fewer high-school completions \citep{autor2016family}.
+This is consistent with recent evidence which shows that family disadvantage disproportionately impedes the pre-market development of boys, in terms of higher disciplinary problems, lower achievement scores, and fewer high-school completions \citep{Autor2016a}.
 Girls of non-white ethnicity, instead, have worse internalising and externalising skills in the MCS, a penalty not suffered by 5-year old non-white girls in the BCS.
@@
 Also, the ethnic structure of the population has changed, with a higher proportion of non-white children in the MCS than in the BCS.
-In general, this has been a period of significant societal changes, with an almost continual rise in the proportion of women in employment, an older age at first birth and a rise in dual-earning parents families \citep{roantree2018women}.
-However, here we do not attempt to disentangle whether and to which extent the observed changes in inequality in socio-emotional skills across the two cohorts can be attributed to changes in returns (or penalties) to maternal characteristics (such as education and employment) or to compositional changes, like it has been done for the analysis of wage inequality \citep{blundell2007changes}.
+In general, this has been a period of significant societal changes, with an almost continual rise in the proportion of women in employment, an older age at first birth and a rise in dual-earning parents families \citep{Roantree2018}.
+However, here we do not attempt to disentangle whether and to which extent the observed changes in inequality in socio-emotional skills across the two cohorts can be attributed to changes in returns (or penalties) to maternal characteristics (such as education and employment) or to compositional changes, like it has been done for the analysis of wage inequality \citep{Blundell2007}.
 
@@
 From this perspective, he is likely more sociable and subject to peer influence in health behaviours.
-This is consistent with the evidence in \citet{goodman2015social}, who find a positive association between child emotional health (measured with items for the internalizing behaviour subscale of the Rutter scale at age 10 in the BCS) and smoking at age 42.
-Furthermore, in recent work \citet{hsieh2018smoking} have shown personality to be a key mechanism through which peers affect smoking behaviour.
+This is consistent with the evidence in \citet{Goodman2015}, who find a positive association between child emotional health (measured with items for the internalizing behaviour subscale of the Rutter scale at age 10 in the BCS) and smoking at age 42.
+Furthermore, in recent work \citet{Hsieh2018} have shown personality to be a key mechanism through which peers affect smoking behaviour.
 
@@
 One possible reason might be the relatively young age at which these skills are measured.
-Results are available upon request.} This is in line with the evidence in \citet{conti2010understanding}, who show that not accounting for non-cognitive traits (a self-regulation factor measured at age 10) overestimates the importance of cognition for predicting health and health behaviours, using data from the British cohort study.
-\citet{conti2013personality} use rich data on child personality and socio-emotional traits collected at ages 7, 11 and 16 in the 1958 British birth cohort,\footnote{They use the Rutter scale and the Bristol Social Adjustment Guide.} and show that these traits rival the importance of cognition in explaining the education gradient in health behaviours (including smoking and BMI).
+Results are available upon request.} This is in line with the evidence in \citet{Conti2010}, who show that not accounting for non-cognitive traits (a self-regulation factor measured at age 10) overestimates the importance of cognition for predicting health and health behaviours, using data from the British cohort study.
+\citet{Conti2013a} use rich data on child personality and socio-emotional traits collected at ages 7, 11 and 16 in the 1958 British birth cohort,\footnote{They use the Rutter scale and the Bristol Social Adjustment Guide.} and show that these traits rival the importance of cognition in explaining the education gradient in health behaviours (including smoking and BMI).
 We show that child socio-emotional skills have greater predictive power for health outcomes and behaviours even when measured at an earlier age.
@@
 For these outcomes, the predictive power of cognitive skills outweighs that of socio-emotional skills, whose predictive power diminishes over time (between the ages 34 and 42), and is driven to insignificance after controlling for cognition.
-This is consistent with the evidence in \citet{conti2011early}, who show that cognitive endowments at age 10 are more predictive (than socio-emotional and health ones) for employment and wage outcomes in the BCS.
+This is consistent with the evidence in \citet{Conti2011}, who show that cognitive endowments at age 10 are more predictive (than socio-emotional and health ones) for employment and wage outcomes in the BCS.
 Again, we show that the greater predictive power of cognition for socioeconomic outcomes holds even when considering earlier-life measures of child development.
@@
 In this paper we have studied inequality in a dimension of human capital which has received limited attention in the literature so far: socio-emotional skills very early in life.
-In particular, we have focused on the measurements of these skills at age 5 in two British cohorts born 30 years apart: the one of children born in 1970 (British Cohort Study, BCS) and the one of children born in 2000/1 (Millenium Cohort Study, MCS).
+In particular, we have focused on the measurements of these skills at age 5 in two British cohorts born 30 years apart: the one of children born in 1970 (British Cohort Study, BCS) and the one of children born in 2000/1 (Millennium Cohort Study, MCS).
 We have provided several contributions to the recent but flourishing literature on the determinants and consequences of early human development.

### 2018-10-02 codes/R/paper/text/paper_jpube_1introlit.tex
Sent for IFS WP

@@
 \citet{Chiteji2010} used the US Panel Study of Income Dynamics (PSID) and found that future orientation and self-efficacy (related to emotional stability) are associated with less alcohol consumption and more exercise.
-\citet{Cobb-Clark2014} used the Australian HILDA data and found that an internal locus of control (also related to emotional stability, perceived control over one's life) is related to better health behaviours (diet, exercise, alcohol consumption and smoking).
+\citet{Cobb-Clark2014} used the Australian HILDA data and found that an internal locus of control (also related to emotional stability, perceived control over one's life) is related to better health behaviours (diet, exercise, alcohol consumption anwhad smoking).
 \citet{Mendolia2014a} used the Longitudinal Study of Young People in England and found that individuals with external locus of control, low self-esteem, and low levels of work ethics, are more likely to engage in risky health behaviours.

### 2019-02-27 codes/R/2b_efa_all.R
efa with full scales

@@ -1,0 +1,82 @@
+## ---- BCSEFA_ALL
+# full sample
+bcs5data$bcs5_rutbAB <- bcs5data$bcs5_rutbA
+bcs5data$bcs5_rutbAB[which(bcs5data$bcs5_rutbB == 0)] <- 0
+items_bcs_all <- bcs5data[,
+                          c( 
+                            grep("bcs5_rutc", names(bcs5data), value=T),
+                            "bcs5_rutbAB", "bcs5_rutbD"
+                          )] %>%
+  mutate_all(as.factor)
+
+nbcs_all <- dim(items_bcs_all)[1]
+bcscor_all <- hetcor(items_bcs_all, ML=TRUE, std.err = F)$correlations
+bcsfa_all <- fa(r=bcscor_all, nfactors=5, rotate = "oblimin", fm="wls")
+
+# names
+rutnames1 <- c("Restless*", "Squirmy/Fidgety*", 
+               "Destroys Belongings", "Fights**", "Not Liked",
+               "Worried*", "Solitary*", "Irritable", "Unhappy*",
+               "Steals", "Twitches", "Sucks thumbs", "Bites Nails",
+               "Disobedient*", "Distracted*", "Fearful*", "Fussy",
+               "Lies", "Bullies**")
+names(rutnames1) <- paste0("bcs5_rutc", seq(1:19))
+# rutnames2 <- c("Headaches**", "Stomachaches**", "Bilious", "Tantrums")
+# names(rutnames2) <- paste0("bcs5_rutb", LETTERS[1:4])
+rutnames2 <- c("Headaches, stomachaches**", "Tantrums**")
+names(rutnames2) <- c("bcs5_rutbAB", "bcs5_rutbD")
+
+rutnames <- c(rutnames1, rutnames2)
+ruttitle <- c(as.character(seq(1:19)), "A+B", "D")
+
+bcs_loadtab_all <- data.frame(cbind(ruttitle, rutnames, bcsfa_all$loadings))
+colnames(bcs_loadtab_all) <- c("item", "title", "f1", "f2", "f3", "f4", "f5")
+
+## ---- MCSEFA_ALL
+to_na <- function(column) {
+  ifelse(column %in% c(-9,-8,-1,4), NA, column)
+}
+rotation <- "oblimin"
+items_mcs_all <- mcs5data[,
+                          c( 
[…]

### 2019-02-27 codes/R/7_ineq_quantiles.R
minor table output fix

@@ -42,7 +42,7 @@ qdtable <- bind_rows(bdatlist) %>%
     med = median(value),
     ciu = quantile(value, .975),
     cil = quantile(value, .025),
-    medci = sprintf("%s\\n[%s,%s]", prn(med), prn(cil), prn(ciu))
+    medci = sprintf("$%s$ \\newline $[%s,%s]$", prn(med), prn(cil), prn(ciu))
   ) %>%
   select(-med, -ciu, -cil) %>%
   # unite cohortsex and reorder
@@ -52,15 +52,7 @@ qdtable <- bind_rows(bdatlist) %>%
   spread(cohortsex, medci) %>%
   mutate(
     pcdiff = paste0(str_sub(pcdiff,2,3), " - ",  str_sub(pcdiff,4,5))
-    )
+  )
 
-# qdtable[,-1] %>% # drop first column
-#   kable("latex", escape = F, booktabs = T, linesep = "", align=c('l','c','c','c','c'),
-#         col.names = c("Quantile diff.", "BCS (1970)", "MCS (2000/1)", "BCS (1970)", "MCS (2000/1)")) %>%
-#   add_header_above(c(" " = 1,  "Males" = 2, "Females" = 2)) %>%
-#   column_spec(2:5, width = "3em") %>%
-#   mutate_all(linebreak)
-#   
-  
 
   

### 2019-02-27 codes/R/paper/seskills_jpube_rev.Rnw
new version of main Rnw file

--- 2018-10-01 codes/R/paper/paper_jpube.Rnw
@@
-\documentclass[pdftex,10pt,a4paper]{article}
+\documentclass[pdftex,11pt,a4paper]{article}
 \usepackage{amsmath}
@@
 \affil{\emph{Institute for Fiscal Studies} \\ \emph{Department of Economics, University College London}}
-\date{\today}
+\date{December 2018}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 
-\thanks{
-\footnotesize{Corresponding Author: Orazio Attanasio, Department of Economics, University College London,
+\scriptsize{Corresponding Author: Orazio Attanasio, Department of Economics, University College London,
 Gordon Street, London WC1H0AX, UK.
@@
 We thank participants to the NBER/LSE Trans-Atlantic Public Economics Seminar (TAPES) and our two discussants Hilary Hoynes and Gabriel Ulyssea for excellent comments.
+We're especially thankful to George Ploubidis for his guidance in the early stages of the project.
+This work is based on data from the British Cohort Study and the Millennium Cohort Study.
 We are grateful to The Centre for Longitudinal Studies, UCL Institute of Education, for the collection of the data, and to the UK Data Archive and UK Data Service for making them available.
 However, they bear no responsibility for the analysis or interpretation of the data.
-Giacomo Mason is funded by an Economics and Social Research Council scholarship.}}
+We would like to thank all cohort members and their families, who generously gave their time to participate in the surveys.
+We thank the DynaHEALTH consortium (H2020–633595 DynaHEALTH) for funding and support.
+GM is funded by an Economics and Social Research Council scholarship.}
 \end{titlepage}
 
-\input{text/paper_jpube_1intro.tex}
-\input{text/paper_jpube_1blit.tex}
+\input{text/paper_jpube_1introlit.tex}
 \input{text/paper_jpube_2datamethods.tex}
@@
 read_chunk(here("codes/R/2_efa.R"))
+read_chunk(here("codes/R/2b_efa_all.R"))
 read_chunk(here("codes/R/3_inv_main.R"))
 read_chunk(here("codes/R/3_inv_aux.R"))
+read_chunk(here("codes/R/3_inv_no511.R"))
 read_chunk(here("codes/R/3_inv_fit.R"))
@@
 read_chunk(here("codes/R/6_graphs_item.R"))
+read_chunk(here("codes/R/7_ineq_quantiles.R"))
 read_chunk(here("codes/R/8_regs_dep.R"))
@@
 
+<<BCSEFA_ALL, dependson='LOAD_DATA'>>=
+@
+
+<<MCSEFA_ALL, dependson='LOAD_DATA'>>=
+@
+
+<<NUMFAC_ALL, dependson='MCSEFA'>>=
+@
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+<<FA_INV_AUX_NO511>>=
+# robustness checks without items 5-11
+@
+
 <<FA_FIT>>=
@@
 
+<<FA_SCORES_NO511>>=
+# produce scores for model without items 5 and 11
+@
+
 <<FA_BOOTSCORES_READ>>=
@@
 
+<<QDTABLE, dependson="SELECTDATA">>=
+# make table of quantile differences
+@
+
 % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\clearpage
+\begin{table}[]
+\caption{Summary statistics}\label{tab:sumstats}
+\centering\setstretch{1.2} \footnotesize
+<<SUMSTATS_TAB, dependson="SUMSTATS", results='asis'>>=
+require(xtable)
+table <- summtab_estimall
+
+  print(xtable(table,
+               align="L{0cm}L{.35\\textwidth}C{.12\\textwidth}C{.12\\textwidth}C{.12\\textwidth}C{.12\\textwidth}",
+               digits = 1
+  ), 
+  include.rownames=F,
+  include.colnames=F,
+  booktabs = T,
+  add.to.row = list(pos = c(list(0,10)),
+                    command = c(
+                      # header
+                      paste(" & 
+                            \\multicolumn{2}{c}{\\textbf{Estimation sample}} & \\multicolumn{2}{c}{\\textbf{Full sample}} 
+                            \\\\ \\cmidrule(lr){2-3} \\cmidrule(lr){4-5} 
+                            & \\textbf{BCS} & \\textbf{MCS} & \\textbf{BCS} & \\textbf{MCS} \\\\
+                            & $N=9545$ & $N=5572$ & $N=14063$ & $N=11530$ \\\\
+                            & Mean (SD) & Mean (SD) & Mean (SD) & Wt. Mean (SD) \\\\
+                            "
+                            , sep=""),
+                      "\\textbf{Five-year survey} \\\\"
+                    )),
+  sanitize.text.function = function(x){x},
+  sanitize.rownames.function = function(x){""}, # to not print row numbers
+  tabular.environment = "tabular",
+  floating = FALSE
+  )
+
+@
+\justify {\scriptsize \emph{Notes}: The table shows the mean values of harmonised variables (and the standard deviation, for continuous ones).
+The \emph{estimation sample} is the subsample used in the analysis.
+The \emph{full sample} is the entire sample of non-multiparous children in both cohorts, residing in England at birth.
+Mean estimates for the full sample in the MCS cohort are weighted to account for survey design.
+\par}
+\end{table}
+
+\clearpage
 \begin{table}[ht]
@@
 \emph{Wording} columns show the actual wording in the scales used in each of the cohort studies.
-Items denoted by (+) are positively coded in the original scale.
+Items denoted by (+) are positively worded in the original scale.
 \par}
@@
 \centering\setstretch{1.2}\footnotesize
-<<INEQ_STATS, dependson="SELECTDATA", results='asis'>>=
-
-# income 
-tb <- data.frame(
-  aggregate(finaldata$faminc10_infl, by=list(finaldata$cohort), FUN = function(x) quantile(x, c(.1, .25, .75, .9), na.rm=T))
-)
-tb$d9010 <- tb$x[,4]/tb$x[,1]
-tb$d7525 <- tb$x[,3]/tb$x[,2]
-
-tb.ext <- data.frame(
-  aggregate(finaldata$EXT, by=list(finaldata$cohortsex), FUN = function(x) quantile(x, c(.1, .25, .5, .75, .9), na.rm=T))
-)
-tb.ext$d9010 <- abs(tb.ext$x[,"90%"]-tb.ext$x[,"10%"])
-tb.ext$d7525 <- abs(tb.ext$x[,"75%"]-tb.ext$x[,"25%"])
-tb.ext$d9050 <- abs(tb.ext$x[,"90%"]-tb.ext$x[,"50%"])
-tb.ext$d5010 <- abs(tb.ext$x[,"50%"]-tb.ext$x[,"10%"])
-rownames(tb.ext) <- tb.ext$Group.1
-tb.ext <- t(tb.ext[,-c(1,2)]) # remove aggregated quantiles and transpose
-tb.ext <- tb.ext[,c("BCS.M","MCS.M","BCS.F","MCS.F")]
-
-tb.int <- data.frame(
-  aggregate(finaldata$INT, by=list(finaldata$cohortsex), FUN = function(x) quantile(x, c(.1, .25, .5, .75, .9), na.rm=T))
-)
-tb.int$d9010 <- abs(tb.int$x[,"90%"]-tb.int$x[,"10%"])
-tb.int$d7525 <- abs(tb.int$x[,"75%"]-tb.int$x[,"25%"])
-tb.int$d9050 <- abs(tb.int$x[,"90%"]-tb.int$x[,"50%"])
-tb.int$d5010 <- abs(tb.int$x[,"50%"]-tb.int$x[,"10%"])
-rownames(tb.int) <- tb.int$Group.1
-tb.int <- t(tb.int[,-c(1,2)]) # remove aggregated quantiles and transpose
-tb.int <- tb.int[,c("BCS.M","MCS.M","BCS.F","MCS.F")]
-
-# assemble and print
-table <- rbind(tb.ext,tb.int)
-table <- cbind(data.frame(as.matrix(rep(c("90-10", "75-25", "90-50", "50-10"),2))),
-               table)
-
-print(xtable(table,
-                    align="L{0cm}L{.15\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}",
+<<INEQ_STATS, dependson="QDTABLE", results='asis'>>=
+
+print(xtable(qdtable[-1],
+                    align="L{0cm}L{.15\\textwidth}C{.15\\textwidth}C{.15\\textwidth}C{.15\\textwidth}C{.1\\textwidth}",
                     digits = 2
@@
             " & \\multicolumn{2}{c}{\\textbf{Males}} & \\multicolumn{2}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-3}  \\cmidrule(lr){4-5}
-                       & BCS (1970) & MCS (2000/1) & BCS (1970) & MCS (2000/1)  \\\\
+                     Quantile diff. & BCS (1970) & MCS (2000/1) & BCS (1970) & MCS (2000/1)  \\\\
                       \\textbf{Externalising} \\\\ ",
@@
 \justify {\scriptsize \emph{Notes}: The table shows differences between quantiles of the distribution of socioemotional skills, by gender and cohort.
+Bootstrap confidence intervals with 1,000 repetitions are in brackets.
 The distribution is a factor score obtained from the factor model in \autoref{sec:methods}.
+These distributions are shown in \autoref{fig:dens}.
 \par}
@@
 \caption{Determinants of Socioemotional Skills}\label{tab:determ}
-\centering\footnotesize
+\centering\scriptsize
 <<REGS_ALL, results='asis', dependson="REGS_DEP">>=
+library(texreg)
 
@@
 \bibliographystyle{chicago}
-\bibliography{/Users/giacomomason/Dropbox/Zotero/cohorts.bib}
+\bibliography{cohorts.bib}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \clearpage
-\begin{table}[]
-\caption{Summary statistics}\label{tab:sumstats}
-\centering\setstretch{1.2} \footnotesize
-<<SUMSTATS_TAB, dependson="SUMSTATS", results='asis'>>=
-require(xtable)
-table <- summtab 
-table <- table[,!names(table) %in% c("bcs.text", "bcs.num", "mcs.text", "mcs.num")] # drop if unneeded
-
-  print(xtable(table,
-               align="L{0cm}L{.3\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}",
-               digits = 1
-  ), 
-  include.rownames=F,
-  include.colnames=F,
-  booktabs = T,
-  add.to.row = list(pos = c(list(0)),
-                    command = c(
-                      # header
-                      paste(" & 
-                            \\multicolumn{2}{c}{\\textbf{Males}} & \\multicolumn{2}{c}{\\textbf{Females}}
-                            \\\\ \\cmidrule(lr){2-3} \\cmidrule(lr){4-5} 
-                            & BCS & MCS & BCS & MCS \\\\
-                            "
-                            , sep="")
-                    )),
-  sanitize.text.function = function(x){x},
-  sanitize.rownames.function = function(x){""}, # to not print row numbers
-  tabular.environment = "tabular",
-  floating = FALSE
-  )
-
-@
-\justify {\scriptsize \emph{Notes}: The table shows mean (standard deviation) of harmonised variables used in the analysis.
-\par}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 \begin{table}
-\caption{Suggested number of factors to retain}\label{tab:numfac}
+\caption{Suggested number of factors to retain -- 11-item scale}\label{tab:numfac}
 \centering\setstretch{1.2}\footnotesize
@@
 @
-\justify {\scriptsize \emph{Notes}: The table compares the optimal number of factors suggested by different aprroaches for our novel scale; scree test based approaches (optimal coordinates, acceleration factor -- \citealp{Raiche2013}), parallel analysis \citep{Horn1965}, Kaiser's eigenvalue rule \citep{Kaiser1960}, Very Simple Structure (VSS, \citealp{Revelle1979}), Velicer Minimum Average Partial test (MAP, \citealp{Velicer1976}).
+\justify {\scriptsize \emph{Notes}: The table compares the optimal number of factors suggested by different approaches, for our novel scale: scree test based approaches (optimal coordinates, acceleration factor -- \citealp{Raiche2013}), parallel analysis \citep{Horn1965}, Kaiser's eigenvalue rule \citep{Kaiser1960}, Very Simple Structure (VSS, \citealp{Revelle1979}), Velicer Minimum Average Partial test (MAP, \citealp{Velicer1976}).
 \par}
@@
 \begin{table}
-\caption{Loadings from exploratory factor analysis}\label{tab:load}
+\caption{Loadings from exploratory factor analysis -- 11-item scale}\label{tab:load}
 \centering\setstretch{1.2}\footnotesize
@@
   mutate(
-    bcs1 = cell_spec(bcs1, "latex", bold = ifelse(abs(bcs1) > .25, TRUE, FALSE)),
-    bcs2 = cell_spec(bcs2, "latex", bold = ifelse(abs(bcs2) > .25, TRUE, FALSE)),
-    mcs1 = cell_spec(mcs1, "latex", bold = ifelse(abs(mcs1) > .25, TRUE, FALSE)),
-    mcs2 = cell_spec(mcs2, "latex", bold = ifelse(abs(mcs2) > .25, TRUE, FALSE))
+    bcs1 = cell_spec(bcs1, "latex", bold = ifelse(abs(bcs1) > .3, TRUE, FALSE)),
+    bcs2 = cell_spec(bcs2, "latex", bold = ifelse(abs(bcs2) > .3, TRUE, FALSE)),
+    mcs1 = cell_spec(mcs1, "latex", bold = ifelse(abs(mcs1) > .3, TRUE, FALSE)),
+    mcs2 = cell_spec(mcs2, "latex", bold = ifelse(abs(mcs2) > .3, TRUE, FALSE))
   ) %>%
@@
 
+\begin{table}
+\caption{Suggested number of factors to retain -- Full set of items for BCS and MCS}\label{tab:numfac_all}
+\centering\setstretch{1.2}\footnotesize
+<<NUMFACTAB_ALL, dependson='NUMFAC_ALL', results='asis'>>=
+# combine results
+
+efatab_all %>%
+  tibble::rownames_to_column() %>%
+  kable("latex", escape = F, booktabs = T, linesep = "", align=c('l','c','c'),
+        col.names = c("Approach", "Rutter (BCS)", "SDQ (MCS)")) %>%
+  column_spec(2:3, width = "5em")
+
+@
+\justify {\scriptsize \emph{Notes}: The table compares the optimal number of factors suggested by different approaches: scree test based approaches (optimal coordinates, acceleration factor -- \citealp{Raiche2013}), parallel analysis \citep{Horn1965}, Kaiser's eigenvalue rule \citep{Kaiser1960}, Very Simple Structure (VSS, \citealp{Revelle1979}), Velicer Minimum Average Partial test (MAP, \citealp{Velicer1976}).
+\par}
+\end{table}
+
+\begin{table}
+\caption{Loadings from exploratory factor analysis -- full set of BCS items}\label{tab:load_bcsall}
+\centering\setstretch{1.2}\footnotesize
+<<EFALOADTAB_BCSALL, dependson='BCSEFA_ALL', results='asis'>>=
+
+bcs_loadtab_all %>%
+  mutate_at(vars(starts_with("f")), function(x) round(as.numeric(as.character(x)), 3)) %>%
+  mutate(
+    f1 = cell_spec(f1, "latex", bold = ifelse(abs(f1) > .3, TRUE, FALSE)),
+    f2 = cell_spec(f2, "latex", bold = ifelse(abs(f2) > .3, TRUE, FALSE)),
+    f3 = cell_spec(f3, "latex", bold = ifelse(abs(f3) > .3, TRUE, FALSE)),
+    f4 = cell_spec(f4, "latex", bold = ifelse(abs(f4) > .3, TRUE, FALSE)),
+    f5 = cell_spec(f5, "latex", bold = ifelse(abs(f5) > .3, TRUE, FALSE))
+  ) %>%
+  kable("latex", escape = F, booktabs = T, linesep = "", align=c('c','l','c','c','c','c','c'),
+        col.names = c("Item", "Title", "Factor 1", "Factor 2", "Factor 3", "Factor 4 ", "Factor 5"))
+
+@
+\justify {\scriptsize \emph{Notes}: The table displays the factor loadings obtained from exploratory factor analysis (EFA) on the full set of Rutter items from the BCS.
+The EFA is based on the decomposition of the polychoric correlation matrix, and uses weighted least squares and oblimin rotation.
+Items denoted by * are used in the 11-item scale, and items denoted by ** are combined in the 11-item scale.
+\par}
+\end{table}
+
+\clearpage
+\begin{table}
+\caption{Loadings from exploratory factor analysis -- full set of MCS items}\label{tab:load_mcsall}
+\centering\setstretch{1.2}\footnotesize
+<<EFALOADTAB_MCSALL, dependson='MCSEFA_ALL', results='asis'>>=
+
+mcs_loadtab_all %>%
+  mutate_at(vars(starts_with("f")), function(x) round(as.numeric(as.character(x)), 3)) %>%
+  mutate(
+    f1 = cell_spec(f1, "latex", bold = ifelse(abs(f1) > .3, TRUE, FALSE)),
+    f2 = cell_spec(f2, "latex", bold = ifelse(abs(f2) > .3, TRUE, FALSE)),
+    f3 = cell_spec(f3, "latex", bold = ifelse(abs(f3) > .3, TRUE, FALSE)),
+    f4 = cell_spec(f4, "latex", bold = ifelse(abs(f4) > .3, TRUE, FALSE)),
+    f5 = cell_spec(f5, "latex", bold = ifelse(abs(f5) > .3, TRUE, FALSE))
+  ) %>%
+  kable("latex", escape = F, booktabs = T, linesep = "", align=c('c','l','c','c','c','c','c'),
+        col.names = c("Item", "Title", "Factor 1", "Factor 2", "Factor 3", "Factor 4 ", "Factor 5"))
+
+@
+\justify {\scriptsize \emph{Notes}: The table displays the factor loadings obtained from exploratory factor analysis (EFA) on the full set of SDQ items from the BCS.
+The EFA is based on the decomposition of the polychoric correlation matrix, and uses weighted least squares and oblimin rotation.
+Items denoted by * are used in the 11-item scale.
+\par}
+\end{table}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}[ht]
+\centering\setstretch{1.2} \footnotesize
+\caption{Parameterisations for measurement invariance}\label{tab:invparam}
+\begin{tabular}{L{.18\textwidth}L{.45\textwidth}L{.3\textwidth}}
+\toprule
+\textbf{Invariance level}     & \textbf{Description} & \textbf{Restrictions} \\ \midrule \\
+Configural (WE{\textTheta})   
+    & \bi[label=$\cdot$] \item Minimally restrictive model for identification \ei
+    & For all groups: $\begin{array}{|l} \text{diag}(\bm{\Phi}) = \bm{I} \\
+                      \bm{\kappa}=\bm{0} \\
+                      \bm{\nu}=\bm{0} \\
+                      \text{diag}(\bm{\Psi})=\bm{I} \end{array}$
+                      \\ \\
+Threshold invariance
+    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ to be equal across groups
+                  \item Statistically equivalent to configural (when measures have 3 categories or less)\ei
+    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
+                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
+                      \newline For all groups: $\begin{array}{|l} \text{diag}(\bm{\Phi}) = \bm{I} \\ 
+                                                \bm{\kappa}=\bm{0} \end{array}$
+                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\nu}_A=\bm{0} \\
+                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \end{array}$
+                   \\ \\
+Threshold and Loading invariance
+    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ and loadings $\lambda$ to be equal across groups
+                  \item Allows comparison of latent factor variances \ei
+    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
+                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
+                      \newline $\lambda_{ci} = \lambda_{c^\prime i}$ for all items, $\forall c, c^\prime$
+                      \newline For all groups: $\bm{\kappa}=\bm{0}$
+                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\nu}_A=\bm{0} \\
+                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \\
+                                                  \text{diag}(\bm{\Phi}_A) = \bm{I} \end{array}$ 
+                   \\ \\
+Threshold, Loading, and Intercept invariance
+    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ and loadings $\lambda$ to be equal across groups
+                  \item Restricts intercepts $\nu$ to zero in both groups
+                  \item Allows comparison of latent factor variances \emph{and} means \ei
+    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
+                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
+                      \newline $\lambda_{ci} = \lambda_{c^\prime i}$ for all items, $\forall c, c^\prime$
+                      \newline For all groups: $\bm{\nu}=\bm{0}$
+                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\kappa}_A=\bm{0} \\
+                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \\
+                                                  \text{diag}(\bm{\Phi}_A) = \bm{I} \end{array}$ 
+\\ \\ \bottomrule
+\end{tabular}
+\justify {\scriptsize \emph{Notes}: Adapted from \cite{Wu2016a}.
+\par}
+\end{table}
 
@@
          "Threshold + Loading Inv", 
-         "Threshold, Loading, + Intercept Inv")
+                     "Thr. + Load. + Intercept Inv")
 )
@@
                      "Threshold + Loading Inv", 
-                     "Threshold, Loading, + Intercept Inv")
-)
-fit_males <- afitab(list(fa.c[[4]], fa.tl[[4]], fa.tli[[4]]), 
+                     "Thr. + Load. + Intercept Inv")
+)
+fit_no511 <- afitab(list(fa.c[[6]], fa.tl[[6]], fa.tli[[6]]), 
                    c("Configural", 
                      "Threshold + Loading Inv", 
-                     "Threshold, Loading, + Intercept Inv")
-)
-fit_females <- afitab(list(fa.c[[5]], fa.tl[[5]], fa.tli[[5]]), 
-                   c("Configural", 
-                     "Threshold + Loading Inv", 
-                     "Threshold, Loading, + Intercept Inv")
-)
-
-table <- rbind(fit_main,fit_agesub,fit_males,fit_females)
+                     "Thr. + Load. + Intercept Inv")
+)
+
+table <- rbind(fit_main,fit_agesub,fit_no511)
 
 print(xtable(table,
-                    align="L{0cm}L{.20\\textwidth}C{.05\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
+                    align="L{0cm}L{.25\\textwidth}C{.08\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
                     digits = 4
@@
              booktabs = T,
-             add.to.row = list(pos = c(list(0,3,6,9)),
+             add.to.row = list(pos = c(list(0,3,6)),
                               command = c(
@@
                   \\\\ \\cmidrule(lr){3-8} \\cmidrule(lr){9-14}
-            Model & Num. params & $\\chi^2$ & RMSE & SRMR & MFI & CFI & G-hat & $\\chi^2$ $p$ & \\textDelta RMSE & \\textDelta SRMR &  \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ 
+            Model & Num. par. & $\\chi^2$ & RMSE & SRMR & MFI & CFI & G-hat & $\\chi^2$ $p$ & \\textDelta RMSE & \\textDelta SRMR &  \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ 
 & (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8) & (9) & (10) & (11) & (12) & (13) \\\\ \\midrule
@@
                       "\\newline \\textbf{B: 59-61 months sample} \\\\ \\midrule ",
-                      "\\newline \\textbf{C: Males only} \\\\ \\midrule ",
-                      "\\newline \\textbf{D: Females only} \\\\ \\midrule "
+                      "\\newline \\textbf{C: Excluding items 5 and 11} \\\\ \\midrule "                                    )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = "\\textwidth",
+             floating = FALSE
+)
+
+@
+\justify {\scriptsize \emph{Notes}: The table presents fit indices for models of different invariance levels, following \cite{Wu2016a} Col.
+(1) displays the number of estimated parameters for each model.
+Col.
+(2) and (8) present the value of the $\chi^2$ statistic and the pvalue of the test of equality with respect to the configural model.
+Col.
+(3)-(7) and (9)-(13) present alternative fit indices (AFIs), in absolute values and differences from the configural model respectively.
+RMSE = Root mean squared error of approximation; SRMR = standardised root mean residual; MFI = McDonald non-centrality index; CFI = comparative fit index; G-hat = gamma-hat.
+Panel A shows results for the whole sample of children in the BCS and MCS cohorts.
+Panel B is restricted to a subsample of children in the age range of maximum overlap between the two cohorts (59-61 months).
+Panel C shows results from a model excluding items 5 and 11 of the 11-item subscale.
+\par}
+\end{table}
+\end{landscape}
+
+\begin{landscape}
+\begin{table}[ht]
+\caption{Measurement invariance fit comparison -- separate genders}\label{tab:fit_gend}
+\centering\setstretch{1.2} \scriptsize
+<<FA_FIT_RES_GEND, dependson="FA_FIT", results='asis'>>=
+require(xtable)
+
+# MAKE FIT TABLES
+fit_males <- afitab(list(fa.c[[4]], fa.tl[[4]], fa.tli[[4]]), 
+                   c("Configural", 
+                     "Threshold + Loading Inv", 
+                     "Thr. + Load. + Intercept Inv")
+)
+fit_females <- afitab(list(fa.c[[5]], fa.tl[[5]], fa.tli[[5]]), 
+                   c("Configural", 
+                     "Threshold + Loading Inv", 
+                     "Thr. + Load. + Intercept Inv")
+)
+
+table <- rbind(fit_males,fit_females)
+
+print(xtable(table,
+                    align="L{0cm}L{.25\\textwidth}C{.08\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
+                    digits = 4
+                    ),
+             size="\\footnotesize",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0,3)),
+                              command = c(
+                              # header
+            "& & \\multicolumn{6}{c}{\\textbf{Absolute fit}} & \\multicolumn{6}{c}{\\textbf{Relative fit}} 
+                  \\\\ \\cmidrule(lr){3-8} \\cmidrule(lr){9-14}
+            Model & Num. par. & $\\chi^2$ & RMSE & SRMR & MFI & CFI & G-hat & $\\chi^2$ $p$ & \\textDelta RMSE & \\textDelta SRMR &  \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ 
+& (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8) & (9) & (10) & (11) & (12) & (13) \\\\ \\midrule
+                       \\textbf{A: Males only} \\\\ ",
+                      "\\newline \\textbf{B: Females only} \\\\ \\midrule "
                                     )),
@@
 RMSE = Root mean squared error of approximation; SRMR = standardised root mean residual; MFI = McDonald non-centrality index; CFI = comparative fit index; G-hat = gamma-hat.
-Panel A shows results for the whole sample of children in the BCS and MCS cohorts.
-Panel B is restricted to a subsample of children in the age range of maximum overlap between the two cohorts (59-61 months).
-Panel C and D are limited to the samples of male and female children respectively.
+Panel A shows results for the whole sample of male children in the BCS and MCS cohorts.
+Panel B shows the same for female children.
 \par}
@@
 print.xtable(xtable(table,
-              align="C{0cm}C{.08\\textwidth}C{.09\\textwidth}C{.07\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
+              align="C{0cm}C{.06\\textwidth}C{.06\\textwidth}C{.09\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.065\\textwidth}C{.065\\textwidth}C{.065\\textwidth}C{.065\\textwidth}C{.065\\textwidth}C{.065\\textwidth}",
               digits = 3
@@
 \centering\footnotesize
+\caption{Predictors of adolescent outcomes, BCS -- factor scores vs. sum scores}\label{tab:adol_sum}
+<<REGS_OUTC_ADOL_RAW, results='asis', dependson="REGS_OUTC">>=
+
+# collect adolescent outcomes
+outctabs <- list(bcs_outctab$smktry16, bcs_outctab$bmi16)
+# collapse to matrix
+outctabs_mat <- do.call(rbind,lapply(outctabs,matrix,ncol=9,byrow=FALSE))
+printoutctab_raw(outctabs_mat, length(outctabs))
+
+@
+\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of cohort members' adolescent outcomes on their externalising and internalising socioemotional skills at five years of age.
+Col.
+(1) shows the mean of the outcome for males.
+Col.
+(2) regresses the outcome on the scores obtained from the factor model in \autoref{sec:methods}.
+Col.
+(3) additionally controls for cognitive ability at age five.
+This is a simple factor score obtained by aggregating the available cognitive measures.
+All standard errors in parentheses are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression.
+Col.
+(4) replaces the factor scores used in col.
+(3) with simpler sum scores -- see \autoref{fig:sumscores}.
+Col.
+(5) to (8) repeat for female cohort members.
+All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
+See \autoref{tab:harmvar} for a description of the variables used.
+\par}
+\end{table}
+
+\begin{table}
+\centering\footnotesize
 \caption{Predictors of adult behaviours, BCS}\label{tab:bcs_adultapp}
@@
 
+\begin{figure}[ht]
+<<SCATTER_NO511, dependson="FA_SCORES_NO511", dev='pdf', fig.height=4, fig.width=9, out.width='\\linewidth'>>=
+@
+\caption{Factor scores excluding items 5 and 11}\label{fig:scatter_no511}
+\justify {\scriptsize \emph{Notes}: The figure shows the relationship between factor scores from the full 11-item model (horizontal axis) against factors scores from a model excluding items 5 and 11 (vertical axis).
+Pearson correlations are reported in the plot area.
+\par}
+\end{figure}
+
 \begin{figure}
@@
 \begin{figure}
-<<ITEMINEQ_FSC, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=11, out.width='\\linewidth'>>=
+<<ITEMINEQ_FSC, dependson="SELECTDATA", dev='tikz', fig.height=8, fig.width=11, out.width='\\linewidth'>>=
 itempl_fsc

### 2019-02-28 codes/R/2b_efa_all.R
asterisks

@@ -52,7 +52,7 @@ mcsfa_all <- fa(r=mcscor_all, nfactors=5, rotate = rotation, fm="wls")
 
 # names
 sdqnames <- c("Considerate", "Restless*", "Headaches/stomachaches*",
-              "Shares", "Tantrums", "Solitary*", "Obedient*",
+              "Shares", "Tantrums*", "Solitary*", "Obedient*",
               "Worried*", "Helpful", "Squirmy/Fidgety*", 
               "Good friend", "Fights/Bullies*", "Unhappy*",
               "Liked", "Distracted*", "Clingy", "Kind", "Lies",

### 2019-02-28 codes/R/3_inv_no511.R
alpha change

@@ -39,7 +39,7 @@ fscorrs <- fss %>% # compute correlations
   summarise(label = paste0("r = ", sprintf("%.3f", cor(full, no511))))
   
 ggplot(fss, aes(x = full, y = no511)) + 
-  geom_point(alpha = .3) + 
+  geom_point(alpha = .15) + 
   coord_fixed() +
   facet_grid(cols = vars(skill)) + 
   xlab("Score (all items)") + ylab("Score (excl. items 5 and 11)") + 

### 2019-02-28 codes/R/paper/seskills_jpube_rev.Rnw
new version of text and tables - incorporates most reviewer comments

@@
 
-\begin{table}
-\caption{Quantile differences in scores}\label{tab:iqdiff}
-\centering\setstretch{1.2}\footnotesize
-<<INEQ_STATS, dependson="QDTABLE", results='asis'>>=
-
-print(xtable(qdtable[-1],
-                    align="L{0cm}L{.15\\textwidth}C{.15\\textwidth}C{.15\\textwidth}C{.15\\textwidth}C{.1\\textwidth}",
-                    digits = 2
-                    ),
-             size="\\footnotesize",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = c(list(0,4)),
-                              command = c(
-                              # header
-            " & \\multicolumn{2}{c}{\\textbf{Males}} & \\multicolumn{2}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-3}  \\cmidrule(lr){4-5}
-                     Quantile diff. & BCS (1970) & MCS (2000/1) & BCS (1970) & MCS (2000/1)  \\\\
-                      \\textbf{Externalising} \\\\ ",
-                      "\\textbf{Internalising} \\\\ \\midrule "
-                                    )),
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabular", width = "\\textwidth",
-             floating = FALSE
-)
-
-@
-\justify {\scriptsize \emph{Notes}: The table shows differences between quantiles of the distribution of socioemotional skills, by gender and cohort.
-Bootstrap confidence intervals with 1,000 repetitions are in brackets.
-The distribution is a factor score obtained from the factor model in \autoref{sec:methods}.
-These distributions are shown in \autoref{fig:dens}.
-\par}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{landscape}
-\begin{table}
-\caption{Determinants of Socioemotional Skills}\label{tab:determ}
-\centering\scriptsize
-<<REGS_ALL, results='asis', dependson="REGS_DEP">>=
-library(texreg)
-
-# select models
-regs <- list(r_ext$BCS.M[["OLS"]],r_ext$MCS.M[["OLS"]],r_ext$BCS.F[["OLS"]],r_ext$MCS.F[["OLS"]],
-             r_int$BCS.M[["OLS"]],r_int$MCS.M[["OLS"]],r_int$BCS.F[["OLS"]],r_int$MCS.F[["OLS"]])
-
-# robust SEs
-robust_se <- list()
-robust_p  <- list()
-bootstrap_se  <- list(r_ext$BCS.M[["OLSbootse"]],r_ext$MCS.M[["OLSbootse"]],r_ext$BCS.F[["OLSbootse"]],r_ext$MCS.F[["OLSbootse"]],
-                      r_int$BCS.M[["OLSbootse"]],r_int$MCS.M[["OLSbootse"]],r_int$BCS.F[["OLSbootse"]],r_int$MCS.F[["OLSbootse"]])
-for (i in 1:length(regs)) {
-  cov1              <- sandwich::vcovHC(regs[[i]], type = "HC1")
-  ct                <- lmtest::coeftest(regs[[i]], vcov = cov1)
-  robust_se[[i]] <- ct[, 2]
-  robust_p[[i]] <- ct[, 4]
-}
-
-texreg::texreg(regs,
-               stars = c(0.1, 0.05, 0.01), digits = 3,
-               override.se = bootstrap_se, override.pvalues = robust_p,
-               single.row = FALSE, scriptsize = TRUE,
-               caption = "", label = "",
-               omit.coef = omitcoefs,
-               groups = grouplabs,
-               custom.columns = list("p-value" = r_ext$Mp, "p-value" = r_ext$Fp, "p-value" = r_int$Mp, "p-value" = r_int$Fp), custom.col.pos = c(4,6,8,10),
-               custom.coef.names = varlabs,
-               custom.model.names = c("\\multicolumn{6}{c}{\\textbf{Externalising}} & \\multicolumn{6}{c}{\\textbf{Internalising}} \\\\
-                                        \\cmidrule(lr){2-7} \\cmidrule(lr){8-13}
-                                        & \\multicolumn{3}{c}{\\textbf{Males}} & \\multicolumn{3}{c}{\\textbf{Females}} &
-                                        \\multicolumn{3}{c}{\\textbf{Males}} & \\multicolumn{3}{c}{\\textbf{Females}} \\\\ 
-                                        \\cmidrule(lr){2-4} \\cmidrule(lr){5-7} \\cmidrule(lr){8-10} \\cmidrule(lr){11-13} 
-                                      & (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8) & (9) & (10) & (11) & (12) \\\\
-                                      & BCS", "MCS", "BCS", "MCS", "BCS", "MCS", "BCS", "MCS"),
-               custom.note = "",
-               include.rmse = FALSE, include.rsquared = FALSE,
-               table = FALSE
-               )
-
-@
-\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of children's socioemotional skills at five years of age on family background information.
-The dependent variable is a factor score obtained from the factor model in \autoref{sec:methods}.
-Col.
-(1) and (2) show coefficients and standard errors in parentheses, for male children in the BCS and MCS cohorts separately.
-The latter are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression.
-Col.
-(3) shows the p-value of a test that the coefficient is the same in the two cohorts.
-Col.
-(4) to (6) repeat for female children.
-Col.
-(7) to (12) repeat for internalising skills.
-All estimates additionally control for region of birth, mother height, number of previous stillbirths at child's birth, preterm birth, a dummy for missing gestational age, and number of other children in the household at child age 5.
-See \autoref{tab:harmvar} for a description of the variables used.
-\par}
-\end{table}
-\end{landscape}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-<<REGS_OUTC_TABPR, dependson="REGS_OUTC">>=
-# function to display outcome regression tables
-# numoutc = number of outcomes
-
-# EXCLUDING RAW SCORES
-printoutctab <- function(tab, numoutc) {
-print(xtable(tab,
-                    align="R{0cm}L{.3\\textwidth}C{.05\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}",
-                    digits = 3
-                    ),
-             size="\\footnotesize",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = as.list(c(0, 6*seq(1,numoutc-1))),
-                              command = c(
-                              # header
-            " & \\multicolumn{3}{c}{\\textbf{Males}} & \\multicolumn{3}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-4}  \\cmidrule(lr){5-7}
-& Mean & \\multicolumn{2}{c}{Coefficients} & Mean & \\multicolumn{2}{c}{Coefficients} \\\\ \\cmidrule(lr){3-4}  \\cmidrule(lr){6-7}
-            & (1) & (2) & (3) & (4) & (5) & (6) \\\\",
-            rep(" \\midrule", numoutc-1)
-            )),
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabular",
-             floating = FALSE
-)
-}
-
-# INCLUDING RAW SCORES
-printoutctab_raw <- function(tab, numoutc) {
-print(xtable(tab,
-                    align="R{0cm}L{.2\\textwidth}C{.05\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}",
-                    digits = 3
-                    ),
-             size="\\footnotesize",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = as.list(c(0, 8*seq(1,numoutc-1))),
-                              command = c(
-                              # header
-            " & \\multicolumn{4}{c}{\\textbf{Males}} & \\multicolumn{4}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-5}  \\cmidrule(lr){6-9}
-& Mean & \\multicolumn{3}{c}{Coefficients} & Mean & \\multicolumn{3}{c}{Coefficients} \\\\ \\cmidrule(lr){3-5}  \\cmidrule(lr){7-9}
-            & (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8) \\\\",
-            rep(" \\midrule", numoutc-1)
-            )),
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabular",
-             floating = FALSE
-)
-}
-@
-
-\begin{table}
-\centering\footnotesize
-\caption{Predictors of adolescent outcomes}\label{tab:adol}
-<<REGS_OUTC_ADOL, results='asis', dependson="REGS_OUTC">>=
-
-# collect adolescent outcomes
-outctabs <- list(bcs_outctab$smktry16, bcs_outctab$bmi16, mcs_outctab$smktry14, mcs_outctab$bmi14)
-outctabs <- lapply(outctabs, function(x) x[-4:-5, c(-5,-9)]) # eliminate raw rows and columns
-# collapse to matrix
-outctabs_mat <- do.call(rbind,lapply(outctabs,matrix,ncol=7,byrow=FALSE))
-printoutctab(outctabs_mat, length(outctabs))
-
-@
-\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of cohort members' adolescent outcomes on their externalising and internalising socioemotional skills at five years of age.
-Col.
-(1) shows the mean of the outcome for males.
-Col.
-(2) regresses the outcome on the scores obtained from the factor model in \autoref{sec:methods}.
-Col.
-(3) additionally controls for cognitive ability at age five.
-This is a simple factor score obtained by aggregating the available cognitive measures.
-All standard errors in parentheses are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression.
-Col.
-(4) to (6) repeat for female cohort members.
-All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
-See \autoref{tab:harmvar} for a description of the variables used.
-\par}
-\end{table}
-
-\begin{table}
-\centering\footnotesize
-\caption{Predictors of adult outcomes -- BCS}\label{tab:bcs_adult}
-<<REGS_OUTC_ADULT, results='asis', dependson="REGS_OUTC">>=
-
-# eliminate raw rows and columns
-outcs <- c("hnvq34", "emp42", "lgpay42")
-outctabs <- list()
-for (o in 1:length(outcs)) outctabs[[outcs[o]]] <- bcs_outctab[[outcs[o]]][-4:-5, c(-5,-9)]
-# collapse to matrix
-outctabs_mat <- do.call(rbind,lapply(outctabs,matrix,ncol=7,byrow=FALSE))
-printoutctab(outctabs_mat, length(outcs))
-
-@
-\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of BCS cohort members' adult outcomes on their externalising and internalising socioemotional skills at five years of age.
-Col.
-(1) shows the mean of the outcome for males.
-Col.
-(2) regresses the outcome on the scores obtained from the factor model in \autoref{sec:methods}.
-Col.
-(3) additionally controls for cognitive ability at age five.
-This is a simple factor score obtained by aggregating the available cognitive measures.
-All standard errors in parentheses are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression.
-Col.
-(4) to (6) repeat for female cohort members.
-All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
-See \autoref{tab:harmvar} for a description of the variables used.
-\par}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}
-<<FACDENS, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
-@
-\caption{Distribution of factor scores}\label{fig:dens}
-\justify {\scriptsize \emph{Notes}: The figure shows the distribution of the externalising and internalising socioemotional skills scores at age five obtained from the factor model, by gender and cohort.
-The scores are estimated from parameter estimates in \autoref{tab:measpars}, using an Empirical Bayes Modal approach.
-Higher scores correspond to \emph{better} skills.
-The distribution is estimated nonparametrically, using an Epanechnikov kernel.
-The figure also reports the $p$ value from Kolmogorov-Smirnov tests of equality between the distribution in BCS and MCS.
-\par}
-\end{figure}
-
-\begin{figure}
-<<FACINEQ_PS, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
-fi <- fi_ps # select result
-gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
-                        fi$ineqlist[[3]],fi$ineqlist[[4]],
-                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
-                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
-@
-\caption{Skill inequality by mother's education}\label{fig:ineqps}
-\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socioemotional skills scores by gender, cohort, and mother's education at age five.
-Mother's education is a dummy for whether the mother continued schooling past the minimum leaving age, based on her date of birth.
-The four panels on top present mean and 95\% confidence intervals.
-Given that we cannot compare means of skills, all scores are normalised to take value zero for the `Compulsory' category, so that the gradient is emphasised.
-The bottom two panels present the unconditional distribution of mother's education.
-\par}
-\end{figure}
-
-\begin{figure}
-<<FACINEQ_SM, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
-fi <- fi_sm # select result
-gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
-                        fi$ineqlist[[3]],fi$ineqlist[[4]],
-                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
-                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
-@
-\caption{Skill inequality by mother's pregnancy smoking}\label{fig:ineqsm}
-\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socioemotional skills scores by gender, cohort, and mother's pregnancy smoking.
-Mother's education is a dummy for whether the mother reported smoking during pregnancy.
-The four panels on top present mean and 95\% confidence intervals.
-Given that we cannot compare means of skills, all scores are normalised to take value zero for the `Non-smoker' category, so that the gradient is emphasised.
-The bottom two panels present the unconditional distribution of mother's pregnancy smoking
-\par}
-\end{figure}
-
-\begin{figure}
-<<FACINEQ_FSC, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
-fi <- fi_fsc # select result
-gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
-                        fi$ineqlist[[3]],fi$ineqlist[[4]],
-                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
-                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
-@
-\caption{Skill inequality by father's occupation}\label{fig:ineqfsc}
-\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socioemotional skills scores by gender, cohort, and father's occupation at age five.
-Father's occupation is based on Registrar General's social class, with classes I to III Non Manual being `White collar' and classes III Manual to V (plus `other') being `Blue collar'.
-`No father figure' is defined as absence of a male figure living in the household.
-The four panels on top present mean and 95\% confidence intervals.
-Given that we cannot compare means of skills, all scores are normalised to take value zero for the `Blue collar' category, so that the gradient is emphasised.
-The bottom two panels present the unconditional distribution of father's occupation.
-\par}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
-\bibliographystyle{chicago}
-\bibliography{cohorts.bib}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
-
-\begin{appendices}
-
-% to number tables and figures with A
-        \setcounter{table}{0}
-        \renewcommand{\thetable}{A\arabic{table}}%
-        \setcounter{figure}{0}
-        \renewcommand{\thefigure}{A\arabic{figure}}%
-
-% Appendix text
-\input{text/paper_jpube_4appendix.tex}
-
-% Appendix tables
-\clearpage
-\section{Appendix tables}
-
-% from tex file
-\input{text/paper_jpube_5apptables.tex}
-
-\clearpage
-\begin{landscape}
-\begin{table}[]
-\caption{Item prevalence, by cohort and gender}\label{tab:meantab}
-\centering\setstretch{1.2} \footnotesize
-<<MEANTAB, dependson="MEANTABLE", results='asis'>>=
-require(xtable)
-table <- meantab
-table <- table[,!names(table) %in% c("bcs.text", "bcs.num", "mcs.text", "mcs.num")] # drop if unneeded
-
-  print(xtable(table,
-               align="L{0cm}C{.04\\textwidth}C{.05\\textwidth}C{.05\\textwidth}L{.12\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
-               digits = 1
-  ), 
-  include.rownames=F,
-  include.colnames=F,
-  booktabs = T,
-  add.to.row = list(pos = c(list(0)),
-                    command = c(
-                      # header
-                      paste(" & & & &
-                            \\multicolumn{6}{c}{\\textbf{Males}} & \\multicolumn{6}{c}{\\textbf{Females}}
-                            \\\\ \\cmidrule(lr){5-10} \\cmidrule(lr){11-16} 
-                            & & & & 
-                            \\multicolumn{3}{c}{\\textbf{BCS}} & \\multicolumn{3}{c}{\\textbf{MCS}}
-                            & \\multicolumn{3}{c}{\\textbf{BCS}} & \\multicolumn{3}{c}{\\textbf{MCS}}
-                            \\\\ \\cmidrule(lr){5-7} \\cmidrule(lr){8-10} \\cmidrule(lr){11-13} \\cmidrule(lr){14-16} 
-                            & & & &
-                            Cert. Appl. & Smtm. Appl. & Appl. &  Cert. True & Smwt. True & True &
-                            Cert. Appl. & Smtm. Appl. & Appl. &  Cert. True & Smwt. True & True 
-                            \\\\
-                            Itm. & Factor & Cat. & Title & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) \\\\
-                            "
-                            , sep="")
-                    )),
-  sanitize.text.function = function(x){x},
-  sanitize.rownames.function = function(x){""}, # to not print row numbers
-  tabular.environment = "tabular",
-  floating = FALSE
-  )
-
-@
-\justify {\scriptsize \emph{Notes}: The table shows the prevalence by gender and cohort for each item of our novel subscale.
-\emph{Itm.} is item number.
-\emph{Factor} is the latent construct to which the item loads -- EXT is Externalising skills, INT is Internalising skills.
-\emph{Cat.} is the number of categories in which the item is coded -- 2 denotes a binary item (applies/does not apply) and 3 denotes a 3-category item.
-\emph{Title} is a short label for the item.
-\emph{Cert.
-/ Smtm.
-Appl.} =  Certainly / sometimes applies.
-\emph{Cert.
-/ Smwt.
-True} = Certainly / somewhat true.
-\par}
-\end{table}
-\end{landscape}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
-
-\begin{table}
-\caption{Suggested number of factors to retain -- 11-item scale}\label{tab:numfac}
-\centering\setstretch{1.2}\footnotesize
-<<NUMFACTAB, dependson='MCSEFARES', results='asis'>>=
-# combine results
-efatab <- data.frame(cbind(efatab.bcs,efatab.mcs))
-
-efatab %>%
-  tibble::rownames_to_column() %>%
-  kable("latex", escape = F, booktabs = T, linesep = "", align=c('l','c','c','c','c','c','c'),
-        col.names = c("Approach", "All", "Males", "Females", "All", "Males", "Females")) %>%
-  add_header_above(c(" " = 1,  "BCS (1970)" = 3, "MCS (2000/1)" = 3)) %>%
-  column_spec(2:7, width = "3em")
-
-@
-\justify {\scriptsize \emph{Notes}: The table compares the optimal number of factors suggested by different approaches, for our novel scale: scree test based approaches (optimal coordinates, acceleration factor -- \citealp{Raiche2013}), parallel analysis \citep{Horn1965}, Kaiser's eigenvalue rule \citep{Kaiser1960}, Very Simple Structure (VSS, \citealp{Revelle1979}), Velicer Minimum Average Partial test (MAP, \citealp{Velicer1976}).
-\par}
-\end{table}
-
-\begin{table}
-\caption{Loadings from exploratory factor analysis -- 11-item scale}\label{tab:load}
-\centering\setstretch{1.2}\footnotesize
-<<EFALOADTAB, dependson='MCSEFARES', results='asis'>>=
-# combine results
-library(kableExtra)
-loadtab <- cbind(meantab$num, meantab$title, data.frame(cbind(bcsfa$loadings, mcsfa$loadings)))
-colnames(loadtab) <- c("item", "title", "bcs1", "bcs2", "mcs1", "mcs2")
-
-loadtab %>%
-  mutate_at(vars(matches("cs")), round, 3) %>%
-  mutate(
-    bcs1 = cell_spec(bcs1, "latex", bold = ifelse(abs(bcs1) > .3, TRUE, FALSE)),
-    bcs2 = cell_spec(bcs2, "latex", bold = ifelse(abs(bcs2) > .3, TRUE, FALSE)),
-    mcs1 = cell_spec(mcs1, "latex", bold = ifelse(abs(mcs1) > .3, TRUE, FALSE)),
-    mcs2 = cell_spec(mcs2, "latex", bold = ifelse(abs(mcs2) > .3, TRUE, FALSE))
-  ) %>%
-  kable("latex", escape = F, booktabs = T, linesep = "", align=c('c','l','c','c','c'),
-        col.names = c("Item", "Title", "Factor 1 (EXT)", "Factor 2 (INT)", "Factor 1 (EXT)", "Factor 2 (INT)")) %>%
-  add_header_above(c(" " = 2,  "BCS (1970)" = 2, "MCS (2000/1)" = 2))
-
-@
-\justify {\scriptsize \emph{Notes}: The table displays the factor loadings obtained from exploratory factor analysis (EFA) on our novel scale, separately by cohort.
-The EFA is based on the decomposition of the polychoric correlation matrix, and uses weighted least squares and oblimin rotation.
-\par}
-\end{table}
-
-\begin{table}
-\caption{Suggested number of factors to retain -- Full set of items for BCS and MCS}\label{tab:numfac_all}
-\centering\setstretch{1.2}\footnotesize
-<<NUMFACTAB_ALL, dependson='NUMFAC_ALL', results='asis'>>=
-# combine results
-
-efatab_all %>%
-  tibble::rownames_to_column() %>%
-  kable("latex", escape = F, booktabs = T, linesep = "", align=c('l','c','c'),
-        col.names = c("Approach", "Rutter (BCS)", "SDQ (MCS)")) %>%
-  column_spec(2:3, width = "5em")
-
-@
-\justify {\scriptsize \emph{Notes}: The table compares the optimal number of factors suggested by different approaches: scree test based approaches (optimal coordinates, acceleration factor -- \citealp{Raiche2013}), parallel analysis \citep{Horn1965}, Kaiser's eigenvalue rule \citep{Kaiser1960}, Very Simple Structure (VSS, \citealp{Revelle1979}), Velicer Minimum Average Partial test (MAP, \citealp{Velicer1976}).
-\par}
-\end{table}
-
-\begin{table}
-\caption{Loadings from exploratory factor analysis -- full set of BCS items}\label{tab:load_bcsall}
-\centering\setstretch{1.2}\footnotesize
-<<EFALOADTAB_BCSALL, dependson='BCSEFA_ALL', results='asis'>>=
-
-bcs_loadtab_all %>%
-  mutate_at(vars(starts_with("f")), function(x) round(as.numeric(as.character(x)), 3)) %>%
-  mutate(
-    f1 = cell_spec(f1, "latex", bold = ifelse(abs(f1) > .3, TRUE, FALSE)),
-    f2 = cell_spec(f2, "latex", bold = ifelse(abs(f2) > .3, TRUE, FALSE)),
-    f3 = cell_spec(f3, "latex", bold = ifelse(abs(f3) > .3, TRUE, FALSE)),
-    f4 = cell_spec(f4, "latex", bold = ifelse(abs(f4) > .3, TRUE, FALSE)),
-    f5 = cell_spec(f5, "latex", bold = ifelse(abs(f5) > .3, TRUE, FALSE))
-  ) %>%
-  kable("latex", escape = F, booktabs = T, linesep = "", align=c('c','l','c','c','c','c','c'),
-        col.names = c("Item", "Title", "Factor 1", "Factor 2", "Factor 3", "Factor 4 ", "Factor 5"))
-
-@
-\justify {\scriptsize \emph{Notes}: The table displays the factor loadings obtained from exploratory factor analysis (EFA) on the full set of Rutter items from the BCS.
-The EFA is based on the decomposition of the polychoric correlation matrix, and uses weighted least squares and oblimin rotation.
-Items denoted by * are used in the 11-item scale, and items denoted by ** are combined in the 11-item scale.
-\par}
-\end{table}
-
-\clearpage
-\begin{table}
-\caption{Loadings from exploratory factor analysis -- full set of MCS items}\label{tab:load_mcsall}
-\centering\setstretch{1.2}\footnotesize
-<<EFALOADTAB_MCSALL, dependson='MCSEFA_ALL', results='asis'>>=
-
-mcs_loadtab_all %>%
-  mutate_at(vars(starts_with("f")), function(x) round(as.numeric(as.character(x)), 3)) %>%
-  mutate(
-    f1 = cell_spec(f1, "latex", bold = ifelse(abs(f1) > .3, TRUE, FALSE)),
-    f2 = cell_spec(f2, "latex", bold = ifelse(abs(f2) > .3, TRUE, FALSE)),
-    f3 = cell_spec(f3, "latex", bold = ifelse(abs(f3) > .3, TRUE, FALSE)),
-    f4 = cell_spec(f4, "latex", bold = ifelse(abs(f4) > .3, TRUE, FALSE)),
-    f5 = cell_spec(f5, "latex", bold = ifelse(abs(f5) > .3, TRUE, FALSE))
-  ) %>%
-  kable("latex", escape = F, booktabs = T, linesep = "", align=c('c','l','c','c','c','c','c'),
-        col.names = c("Item", "Title", "Factor 1", "Factor 2", "Factor 3", "Factor 4 ", "Factor 5"))
-
-@
-\justify {\scriptsize \emph{Notes}: The table displays the factor loadings obtained from exploratory factor analysis (EFA) on the full set of SDQ items from the BCS.
-The EFA is based on the decomposition of the polychoric correlation matrix, and uses weighted least squares and oblimin rotation.
-Items denoted by * are used in the 11-item scale.
-\par}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \begin{table}[ht]
@@
 \caption{Parameterisations for measurement invariance}\label{tab:invparam}
-\begin{tabular}{L{.18\textwidth}L{.45\textwidth}L{.3\textwidth}}
+\begin{tabular}{L{.18\textwidth}L{.4\textwidth}L{.45\textwidth}}
 \toprule
@@
 
+\begin{table}
+\caption{Quantile differences in scores}\label{tab:iqdiff}
+\centering\setstretch{1.2}\footnotesize
+<<INEQ_STATS, dependson="QDTABLE", results='asis'>>=
+
+print(xtable(qdtable[-1],
+                    align="L{0cm}L{.15\\textwidth}C{.15\\textwidth}C{.15\\textwidth}C{.15\\textwidth}C{.1\\textwidth}",
+                    digits = 2
+                    ),
+             size="\\footnotesize",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0,4)),
+                              command = c(
+                              # header
+            " & \\multicolumn{2}{c}{\\textbf{Males}} & \\multicolumn{2}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-3}  \\cmidrule(lr){4-5}
+                     Quantile diff. & BCS (1970) & MCS (2000/1) & BCS (1970) & MCS (2000/1)  \\\\
+                      \\textbf{Externalising} \\\\ ",
+                      "\\textbf{Internalising} \\\\ \\midrule "
+                                    )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = "\\textwidth",
+             floating = FALSE
+)
+
+@
+\justify {\scriptsize \emph{Notes}: The table shows differences between quantiles of the distribution of socioemotional skills, by gender and cohort.
+Bootstrap confidence intervals with 1,000 repetitions are in brackets.
+The distribution is a factor score obtained from the factor model in \autoref{sec:methods}.
+These distributions are shown in \autoref{fig:dens}.
+\par}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{landscape}
+\begin{table}
+\caption{Determinants of Socioemotional Skills}\label{tab:determ}
+\centering\scriptsize
+<<REGS_ALL, results='asis', dependson="REGS_DEP">>=
+library(texreg)
+
+# select models
+regs <- list(r_ext$BCS.M[["OLS"]],r_ext$MCS.M[["OLS"]],r_ext$BCS.F[["OLS"]],r_ext$MCS.F[["OLS"]],
+             r_int$BCS.M[["OLS"]],r_int$MCS.M[["OLS"]],r_int$BCS.F[["OLS"]],r_int$MCS.F[["OLS"]])
+
+# robust SEs
+robust_se <- list()
+robust_p  <- list()
+bootstrap_se  <- list(r_ext$BCS.M[["OLSbootse"]],r_ext$MCS.M[["OLSbootse"]],r_ext$BCS.F[["OLSbootse"]],r_ext$MCS.F[["OLSbootse"]],
+                      r_int$BCS.M[["OLSbootse"]],r_int$MCS.M[["OLSbootse"]],r_int$BCS.F[["OLSbootse"]],r_int$MCS.F[["OLSbootse"]])
+for (i in 1:length(regs)) {
+  cov1              <- sandwich::vcovHC(regs[[i]], type = "HC1")
+  ct                <- lmtest::coeftest(regs[[i]], vcov = cov1)
+  robust_se[[i]] <- ct[, 2]
+  robust_p[[i]] <- ct[, 4]
+}
+
+texreg::texreg(regs,
+               stars = c(0.1, 0.05, 0.01), digits = 3,
+               override.se = bootstrap_se, override.pvalues = robust_p,
+               single.row = FALSE, scriptsize = TRUE,
+               caption = "", label = "",
+               omit.coef = omitcoefs,
+               groups = grouplabs,
+               custom.columns = list("p-value" = r_ext$Mp, "p-value" = r_ext$Fp, "p-value" = r_int$Mp, "p-value" = r_int$Fp), custom.col.pos = c(4,6,8,10),
+               custom.coef.names = varlabs,
+               custom.model.names = c("\\multicolumn{6}{c}{\\textbf{Externalising}} & \\multicolumn{6}{c}{\\textbf{Internalising}} \\\\
+                                        \\cmidrule(lr){2-7} \\cmidrule(lr){8-13}
+                                        & \\multicolumn{3}{c}{\\textbf{Males}} & \\multicolumn{3}{c}{\\textbf{Females}} &
+                                        \\multicolumn{3}{c}{\\textbf{Males}} & \\multicolumn{3}{c}{\\textbf{Females}} \\\\ 
+                                        \\cmidrule(lr){2-4} \\cmidrule(lr){5-7} \\cmidrule(lr){8-10} \\cmidrule(lr){11-13} 
+                                      & (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8) & (9) & (10) & (11) & (12) \\\\
+                                      & BCS", "MCS", "BCS", "MCS", "BCS", "MCS", "BCS", "MCS"),
+               custom.note = "",
+               include.rmse = FALSE, include.rsquared = FALSE,
+               table = FALSE
+               )
+
+@
+\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of children's socioemotional skills at five years of age on family background information.
+The dependent variable is a factor score obtained from the factor model in \autoref{sec:methods}.
+Col.
+(1) and (2) show coefficients and standard errors in parentheses, for male children in the BCS and MCS cohorts separately.
+The latter are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression.
+Col.
+(3) shows the p-value of a test that the coefficient is the same in the two cohorts.
+Col.
+(4) to (6) repeat for female children.
+Col.
+(7) to (12) repeat for internalising skills.
+All estimates additionally control for region of birth, mother height, number of previous stillbirths at child's birth, preterm birth, a dummy for missing gestational age, and number of other children in the household at child age 5.
+See \autoref{tab:harmvar} for a description of the variables used.
+\par}
+\end{table}
+\end{landscape}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+<<REGS_OUTC_TABPR, dependson="REGS_OUTC">>=
+# function to display outcome regression tables
+# numoutc = number of outcomes
+
+# EXCLUDING RAW SCORES
+printoutctab <- function(tab, numoutc) {
+print(xtable(tab,
+                    align="R{0cm}L{.3\\textwidth}C{.05\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}",
+                    digits = 3
+                    ),
+             size="\\footnotesize",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = as.list(c(0, 6*seq(1,numoutc-1))),
+                              command = c(
+                              # header
+            " & \\multicolumn{3}{c}{\\textbf{Males}} & \\multicolumn{3}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-4}  \\cmidrule(lr){5-7}
+& Mean & \\multicolumn{2}{c}{Coefficients} & Mean & \\multicolumn{2}{c}{Coefficients} \\\\ \\cmidrule(lr){3-4}  \\cmidrule(lr){6-7}
+            & (1) & (2) & (3) & (4) & (5) & (6) \\\\",
+            rep(" \\midrule", numoutc-1)
+            )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular",
+             floating = FALSE
+)
+}
+
+# INCLUDING RAW SCORES
+printoutctab_raw <- function(tab, numoutc) {
+print(xtable(tab,
+                    align="R{0cm}L{.2\\textwidth}C{.05\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}C{.075\\textwidth}",
+                    digits = 3
+                    ),
+             size="\\footnotesize",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = as.list(c(0, 8*seq(1,numoutc-1))),
+                              command = c(
+                              # header
+            " & \\multicolumn{4}{c}{\\textbf{Males}} & \\multicolumn{4}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-5}  \\cmidrule(lr){6-9}
+& Mean & \\multicolumn{3}{c}{Coefficients} & Mean & \\multicolumn{3}{c}{Coefficients} \\\\ \\cmidrule(lr){3-5}  \\cmidrule(lr){7-9}
+            & (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8) \\\\",
+            rep(" \\midrule", numoutc-1)
+            )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular",
+             floating = FALSE
+)
+}
+@
+
+\begin{table}
+\centering\footnotesize
+\caption{Predictors of adolescent outcomes}\label{tab:adol}
+<<REGS_OUTC_ADOL, results='asis', dependson="REGS_OUTC">>=
+
+# collect adolescent outcomes
+outctabs <- list(bcs_outctab$smktry16, bcs_outctab$bmi16, mcs_outctab$smktry14, mcs_outctab$bmi14)
+outctabs <- lapply(outctabs, function(x) x[-4:-5, c(-5,-9)]) # eliminate raw rows and columns
+# collapse to matrix
+outctabs_mat <- do.call(rbind,lapply(outctabs,matrix,ncol=7,byrow=FALSE))
+printoutctab(outctabs_mat, length(outctabs))
+
+@
+\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of cohort members' adolescent outcomes on their externalising and internalising socioemotional skills at five years of age.
+Col.
+(1) shows the mean of the outcome for males.
+Col.
+(2) regresses the outcome on the scores obtained from the factor model in \autoref{sec:methods}.
+Col.
+(3) additionally controls for cognitive ability at age five.
+This is a simple factor score obtained by aggregating the available cognitive measures.
+All standard errors in parentheses are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression.
+Col.
+(4) to (6) repeat for female cohort members.
+All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
+See \autoref{tab:harmvar} for a description of the variables used.
+\par}
+\end{table}
+
+\begin{table}
+\centering\footnotesize
+\caption{Predictors of adult outcomes -- BCS}\label{tab:bcs_adult}
+<<REGS_OUTC_ADULT, results='asis', dependson="REGS_OUTC">>=
+
+# eliminate raw rows and columns
+outcs <- c("hnvq34", "emp42", "lgpay42")
+outctabs <- list()
+for (o in 1:length(outcs)) outctabs[[outcs[o]]] <- bcs_outctab[[outcs[o]]][-4:-5, c(-5,-9)]
+# collapse to matrix
+outctabs_mat <- do.call(rbind,lapply(outctabs,matrix,ncol=7,byrow=FALSE))
+printoutctab(outctabs_mat, length(outcs))
+
+@
+\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of BCS cohort members' adult outcomes on their externalising and internalising socioemotional skills at five years of age.
+Col.
+(1) shows the mean of the outcome for males.
+Col.
+(2) regresses the outcome on the scores obtained from the factor model in \autoref{sec:methods}.
+Col.
+(3) additionally controls for cognitive ability at age five.
+This is a simple factor score obtained by aggregating the available cognitive measures.
+All standard errors in parentheses are obtained using 1,000 bootstrap repetitions, taking into account the factor estimation stage that precedes the regression.
+Col.
+(4) to (6) repeat for female cohort members.
+All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
+See \autoref{tab:harmvar} for a description of the variables used.
+\par}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}
+<<FACDENS, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
+@
+\caption{Distribution of factor scores}\label{fig:dens}
+\justify {\scriptsize \emph{Notes}: The figure shows the distribution of the externalising and internalising socioemotional skills scores at age five obtained from the factor model, by gender and cohort.
+The scores are estimated from parameter estimates in \autoref{tab:measpars}, using an Empirical Bayes Modal approach.
+Higher scores correspond to \emph{better} skills.
+The distribution is estimated nonparametrically, using an Epanechnikov kernel.
+The figure also reports the $p$ value from Kolmogorov-Smirnov tests of equality between the distribution in BCS and MCS.
+\par}
+\end{figure}
+
+\begin{figure}
+<<FACINEQ_PS, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
+fi <- fi_ps # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\caption{Skill inequality by mother's education}\label{fig:ineqps}
+\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socioemotional skills scores by gender, cohort, and mother's education at age five.
+Mother's education is a dummy for whether the mother continued schooling past the minimum leaving age, based on her date of birth.
+The four panels on top present mean and 95\% confidence intervals.
+Given that we cannot compare means of skills, all scores are normalised to take value zero for the `Compulsory' category, so that the gradient is emphasised.
+The bottom two panels present the unconditional distribution of mother's education.
+\par}
+\end{figure}
+
+\begin{figure}
+<<FACINEQ_SM, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
+fi <- fi_sm # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\caption{Skill inequality by mother's pregnancy smoking}\label{fig:ineqsm}
+\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socioemotional skills scores by gender, cohort, and mother's pregnancy smoking.
+Mother's education is a dummy for whether the mother reported smoking during pregnancy.
+The four panels on top present mean and 95\% confidence intervals.
+Given that we cannot compare means of skills, all scores are normalised to take value zero for the `Non-smoker' category, so that the gradient is emphasised.
+The bottom two panels present the unconditional distribution of mother's pregnancy smoking
+\par}
+\end{figure}
+
+\begin{figure}
+<<FACINEQ_FSC, dependson="SELECTDATA", dev='tikz', fig.height=10, fig.width=10, out.width='\\linewidth'>>=
+fi <- fi_fsc # select result
+gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]],
+                        fi$ineqlist[[3]],fi$ineqlist[[4]],
+                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
+                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
+@
+\caption{Skill inequality by father's occupation}\label{fig:ineqfsc}
+\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socioemotional skills scores by gender, cohort, and father's occupation at age five.
+Father's occupation is based on Registrar General's social class, with classes I to III Non Manual being `White collar' and classes III Manual to V (plus `other') being `Blue collar'.
+`No father figure' is defined as absence of a male figure living in the household.
+The four panels on top present mean and 95\% confidence intervals.
+Given that we cannot compare means of skills, all scores are normalised to take value zero for the `Blue collar' category, so that the gradient is emphasised.
+The bottom two panels present the unconditional distribution of father's occupation.
+\par}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\clearpage
+\bibliographystyle{chicago}
+\bibliography{cohorts.bib}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\clearpage
+
+\begin{appendices}
+
+% to number tables and figures with A
+        \setcounter{table}{0}
+        \renewcommand{\thetable}{A\arabic{table}}%
+        \setcounter{figure}{0}
+        \renewcommand{\thefigure}{A\arabic{figure}}%
+
+% Appendix text
+\input{text/paper_jpube_4appendix.tex}
+
+% Appendix tables
+\clearpage
+\section{Appendix tables}
+
+% from tex file
+\input{text/paper_jpube_5apptables.tex}
+
+\clearpage
+\begin{landscape}
+\begin{table}[]
+\caption{Item prevalence, by cohort and gender}\label{tab:meantab}
+\centering\setstretch{1.2} \footnotesize
+<<MEANTAB, dependson="MEANTABLE", results='asis'>>=
+require(xtable)
+table <- meantab
+table <- table[,!names(table) %in% c("bcs.text", "bcs.num", "mcs.text", "mcs.num")] # drop if unneeded
+
+  print(xtable(table,
+               align="L{0cm}C{.04\\textwidth}C{.05\\textwidth}C{.05\\textwidth}L{.12\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
+               digits = 1
+  ), 
+  include.rownames=F,
+  include.colnames=F,
+  booktabs = T,
+  add.to.row = list(pos = c(list(0)),
+                    command = c(
+                      # header
+                      paste(" & & & &
+                            \\multicolumn{6}{c}{\\textbf{Males}} & \\multicolumn{6}{c}{\\textbf{Females}}
+                            \\\\ \\cmidrule(lr){5-10} \\cmidrule(lr){11-16} 
+                            & & & & 
+                            \\multicolumn{3}{c}{\\textbf{BCS}} & \\multicolumn{3}{c}{\\textbf{MCS}}
+                            & \\multicolumn{3}{c}{\\textbf{BCS}} & \\multicolumn{3}{c}{\\textbf{MCS}}
+                            \\\\ \\cmidrule(lr){5-7} \\cmidrule(lr){8-10} \\cmidrule(lr){11-13} \\cmidrule(lr){14-16} 
+                            & & & &
+                            Cert. Appl. & Smtm. Appl. & Appl. &  Cert. True & Smwt. True & True &
+                            Cert. Appl. & Smtm. Appl. & Appl. &  Cert. True & Smwt. True & True 
+                            \\\\
+                            Itm. & Factor & Cat. & Title & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) & (\\%) \\\\
+                            "
+                            , sep="")
+                    )),
+  sanitize.text.function = function(x){x},
+  sanitize.rownames.function = function(x){""}, # to not print row numbers
+  tabular.environment = "tabular",
+  floating = FALSE
+  )
+
+@
+\justify {\scriptsize \emph{Notes}: The table shows the prevalence by gender and cohort for each item of our novel subscale.
+\emph{Itm.} is item number.
+\emph{Factor} is the latent construct to which the item loads -- EXT is Externalising skills, INT is Internalising skills.
+\emph{Cat.} is the number of categories in which the item is coded -- 2 denotes a binary item (applies/does not apply) and 3 denotes a 3-category item.
+\emph{Title} is a short label for the item.
+\emph{Cert.
+/ Smtm.
+Appl.} =  Certainly / sometimes applies.
+\emph{Cert.
+/ Smwt.
+True} = Certainly / somewhat true.
+\par}
+\end{table}
+\end{landscape}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\clearpage
+
+\begin{table}
+\caption{Suggested number of factors to retain -- 11-item scale}\label{tab:numfac}
+\centering\setstretch{1.2}\footnotesize
+<<NUMFACTAB, dependson='MCSEFARES', results='asis'>>=
+# combine results
+efatab <- data.frame(cbind(efatab.bcs,efatab.mcs))
+
+efatab %>%
+  tibble::rownames_to_column() %>%
+  kable("latex", escape = F, booktabs = T, linesep = "", align=c('l','c','c','c','c','c','c'),
+        col.names = c("Approach", "All", "Males", "Females", "All", "Males", "Females")) %>%
+  add_header_above(c(" " = 1,  "BCS (1970)" = 3, "MCS (2000/1)" = 3)) %>%
+  column_spec(2:7, width = "3em")
+
+@
+\justify {\scriptsize \emph{Notes}: The table compares the optimal number of factors suggested by different approaches, for our novel scale: scree test based approaches (optimal coordinates, acceleration factor -- \citealp{Raiche2013}), parallel analysis \citep{Horn1965}, Kaiser's eigenvalue rule \citep{Kaiser1960}, Very Simple Structure (VSS, \citealp{Revelle1979}), Velicer Minimum Average Partial test (MAP, \citealp{Velicer1976}).
+\par}
+\end{table}
+
+\begin{table}
+\caption{Loadings from exploratory factor analysis -- 11-item scale}\label{tab:load}
+\centering\setstretch{1.2}\footnotesize
+<<EFALOADTAB, dependson='MCSEFARES', results='asis'>>=
+# combine results
+library(kableExtra)
+loadtab <- cbind(meantab$num, meantab$title, data.frame(cbind(bcsfa$loadings, mcsfa$loadings)))
+colnames(loadtab) <- c("item", "title", "bcs1", "bcs2", "mcs1", "mcs2")
+
+loadtab %>%
+  mutate_at(vars(matches("cs")), round, 3) %>%
+  mutate(
+    bcs1 = cell_spec(bcs1, "latex", bold = ifelse(abs(bcs1) > .3, TRUE, FALSE)),
+    bcs2 = cell_spec(bcs2, "latex", bold = ifelse(abs(bcs2) > .3, TRUE, FALSE)),
+    mcs1 = cell_spec(mcs1, "latex", bold = ifelse(abs(mcs1) > .3, TRUE, FALSE)),
+    mcs2 = cell_spec(mcs2, "latex", bold = ifelse(abs(mcs2) > .3, TRUE, FALSE))
+  ) %>%
+  kable("latex", escape = F, booktabs = T, linesep = "", align=c('c','l','c','c','c'),
+        col.names = c("Item", "Title", "Factor 1 (EXT)", "Factor 2 (INT)", "Factor 1 (EXT)", "Factor 2 (INT)")) %>%
+  add_header_above(c(" " = 2,  "BCS (1970)" = 2, "MCS (2000/1)" = 2))
+
+@
+\justify {\scriptsize \emph{Notes}: The table displays the factor loadings obtained from exploratory factor analysis (EFA) on our novel scale, separately by cohort.
+The EFA is performed decomposing the polychoric correlation matrix of the items and using weighted least squares, and the solution is rescaled using oblique factor rotation (\emph{oblimin}).
+We use the \textbf{\textsf{R}} package \texttt{psych}, version 1.8.4 \citep{Revelle2018}.
+\par}
+\end{table}
+
+\begin{table}
+\caption{Suggested number of factors to retain -- Full set of items for BCS and MCS}\label{tab:numfac_all}
+\centering\setstretch{1.2}\footnotesize
+<<NUMFACTAB_ALL, dependson='NUMFAC_ALL', results='asis'>>=
+# combine results
+
+efatab_all %>%
+  tibble::rownames_to_column() %>%
+  kable("latex", escape = F, booktabs = T, linesep = "", align=c('l','c','c'),
+        col.names = c("Approach", "Rutter (BCS)", "SDQ (MCS)")) %>%
+  column_spec(2:3, width = "5em")
+
+@
+\justify {\scriptsize \emph{Notes}: The table compares the optimal number of factors suggested by different approaches: scree test based approaches (optimal coordinates, acceleration factor -- \citealp{Raiche2013}), parallel analysis \citep{Horn1965}, Kaiser's eigenvalue rule \citep{Kaiser1960}, Very Simple Structure (VSS, \citealp{Revelle1979}), Velicer Minimum Average Partial test (MAP, \citealp{Velicer1976}).
+\par}
+\end{table}
+
+\begin{table}
+\caption{Loadings from exploratory factor analysis -- full set of BCS items}\label{tab:load_bcsall}
+\centering\setstretch{1.2}\footnotesize
+<<EFALOADTAB_BCSALL, dependson='BCSEFA_ALL', results='asis'>>=
+
+bcs_loadtab_all %>%
+  mutate_at(vars(starts_with("f")), function(x) round(as.numeric(as.character(x)), 3)) %>%
+  mutate(
+    f1 = cell_spec(f1, "latex", bold = ifelse(abs(f1) > .3, TRUE, FALSE)),
+    f2 = cell_spec(f2, "latex", bold = ifelse(abs(f2) > .3, TRUE, FALSE)),
+    f3 = cell_spec(f3, "latex", bold = ifelse(abs(f3) > .3, TRUE, FALSE)),
+    f4 = cell_spec(f4, "latex", bold = ifelse(abs(f4) > .3, TRUE, FALSE)),
+    f5 = cell_spec(f5, "latex", bold = ifelse(abs(f5) > .3, TRUE, FALSE))
+  ) %>%
+  kable("latex", escape = F, booktabs = T, linesep = "", align=c('c','l','c','c','c','c','c'),
+        col.names = c("Item", "Title", "Factor 1", "Factor 2", "Factor 3", "Factor 4 ", "Factor 5"))
+
+@
+\justify {\scriptsize \emph{Notes}: The table displays the factor loadings obtained from exploratory factor analysis (EFA) on the full set of Rutter items from the BCS.
+Items denoted by * are used in the 11-item scale.
+The EFA is performed decomposing the polychoric correlation matrix of the items and using weighted least squares, and the solution is rescaled using oblique factor rotation (\emph{oblimin}).
+We use the \textbf{\textsf{R}} package \texttt{psych}, version 1.8.4 \citep{Revelle2018}.
+\par}
+\end{table}
+
+\clearpage
+\begin{table}
+\caption{Loadings from exploratory factor analysis -- full set of MCS items}\label{tab:load_mcsall}
+\centering\setstretch{1.2}\footnotesize
+<<EFALOADTAB_MCSALL, dependson='MCSEFA_ALL', results='asis'>>=
+
+mcs_loadtab_all %>%
+  mutate_at(vars(starts_with("f")), function(x) round(as.numeric(as.character(x)), 3)) %>%
+  mutate(
+    f1 = cell_spec(f1, "latex", bold = ifelse(abs(f1) > .3, TRUE, FALSE)),
+    f2 = cell_spec(f2, "latex", bold = ifelse(abs(f2) > .3, TRUE, FALSE)),
+    f3 = cell_spec(f3, "latex", bold = ifelse(abs(f3) > .3, TRUE, FALSE)),
+    f4 = cell_spec(f4, "latex", bold = ifelse(abs(f4) > .3, TRUE, FALSE)),
+    f5 = cell_spec(f5, "latex", bold = ifelse(abs(f5) > .3, TRUE, FALSE))
+  ) %>%
+  kable("latex", escape = F, booktabs = T, linesep = "", align=c('c','l','c','c','c','c','c'),
+        col.names = c("Item", "Title", "Factor 1", "Factor 2", "Factor 3", "Factor 4 ", "Factor 5"))
+
+@
+\justify {\scriptsize \emph{Notes}: The table displays the factor loadings obtained from exploratory factor analysis (EFA) on the full set of SDQ items from the BCS.
+Items denoted by * are used in the 11-item scale.
+The EFA is performed decomposing the polychoric correlation matrix of the items and using weighted least squares, and the solution is rescaled using oblique factor rotation (\emph{oblimin}).
+We use the \textbf{\textsf{R}} package \texttt{psych}, version 1.8.4 \citep{Revelle2018}.
+\par}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{landscape}

### 2019-02-28 codes/R/paper/text/paper_jpube_1introlit.tex
new version of text and tables - incorporates most reviewer comments

@@
 conscientiousness), and risk aversion and time preferences.
-Given the vastity of this literature, we briefly review below the main papers on the determinants and consequences of socio-emotional traits which are more directly related to our work, and we refer to other sources for more exhaustive reviews \citep{Borghans2008,Almlund2011,Goodman2015,Kautz2014}.
+Given the vastness of this literature, we briefly review below the main papers on the determinants and consequences of socio-emotional traits which are more directly related to our work, and we refer to other sources for more exhaustive reviews \citep{Borghans2008,Almlund2011,Goodman2015,Kautz2014}.
 
@@
 Very few papers attempt to make cross-cohorts comparisons about the importance of socio-emotional skills.
-\citet{Blanden2007} -- one of the closes study to ours -- examine cognitive skills, non‐cognitive traits, educational attainment and labour market attachment as mediators of the decline in inter-generational mobility in UK between the 1958 and the 1970 cohorts.
+\citet{Blanden2007} -- one of the closest study to ours -- examine cognitive skills, non‐cognitive traits, educational attainment and labour market attachment as mediators of the decline in inter-generational income mobility in UK between the 1958 and the 1970 cohorts.
 The authors take great care in selecting non-cognitive items to be as comparable as possible across cohorts, from the Rutter scale at age 10 for the 1970 cohort and from the Bristol Social Adjustment Guide for the 1958 cohort; however, they do not carry out a formal test of measurement invariance and the do not construct factor scores fully comparable across cohorts as we do.
@@
 In sum, even if the literature on the determinants and consequences of socio-emotional skills has been booming, most papers use skills measured in late childhood or in adolescence; and no paper in economics formally tests for invariance of measurements across different groups and constructs fully comparable scores.
-In this paper, we use measures of child socio-emotional development at age 5, hence before formal schooling starts; and we construct comparable scales across the two cohorts we study (the 1970 and the 2000 British cohorts), so that we can investigate changes in inequality in early development, their determinants, and consequences, in a parallel fashion.
+In this paper, we use measures of child socio-emotional development at age 5, hence at the very beginning of formal schooling; and we construct comparable scales across the two cohorts we study (the 1970 and the 2000 British cohorts), so that we can investigate changes in inequality in early development, their determinants, and consequences, in a parallel fashion.

### 2019-02-28 codes/R/paper/text/paper_jpube_2datamethods.tex
new version of text and tables - incorporates most reviewer comments

@@
 The studies follow the lives of children born approximately 30 years apart: the British Cohort Study (BCS) surveys individuals born in 1970, and the Millennium Cohort Study (MCS) includes births between 2000 and 2002.
-The British Cohort Study includes all individuals born across Great Britain in a single week (April 4-11, 1970).
+The British Cohort Study includes all individuals born across Great Britain in a single week in 1970.
 Cohort members' families -- and subsequently members themselves -- were surveyed on multiple occasions.
@@
 The Millennium Cohort Study follows individuals born in the UK between September 2000 and January 2002.
-We use the birth survey, and the sweeps at around 5, 11, and 14 years.\footnote{All data is publicly available at the UK Data Service \citep{Chamberlain2013,Butler2016a,Butler2016b,Butler2017,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2016b,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2016,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2016a,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2017b,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2017,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2017a,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2017c}.}
+We use the first (age 0) survey -- carried out at 9 months of age -- and the sweeps at around 5 and 14 years of age.\footnote{All data is publicly available at the UK Data Service \citep{Chamberlain2013,Butler2016a,Butler2016b,Butler2017,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2016b,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2016,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2016a,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2017b,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2017,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2017c}.}
 
@@
 All the above variables are transformed or recoded to maximise comparability between the two studies.
-Furthermore, we add some adolescent risky behaviours such as smoking and BMI, with the caveat that these are surveyed at 16 and 14 in BCS and MCS respectively.
+Furthermore, we add some adolescent outcomes such as smoking and BMI, with the caveat that these are surveyed at different ages -- 16 in BCS and 14 in MCS.
 Finally, for the 1970 cohort we also include measures of adult educational attainment, BMI, and income.
@@
 The \texttt{lavaan} package in \textbf{\textsf{R}} \citep{Rosseel2012} is the most suitable tool for our invariance analysis, but it does not allow to use weights when outcomes are categorical, as it is the case for the socio-emotional measurements.}
-We also restrict our sample to individuals born in England.
-Finally, we restrict the sample to cases where the respondent in the five-year followup was the natural mother, and where there is complete information on socioemotional skills.
-The final sample contains 9,545 individuals from the British Cohort Study, and 5,436 from the Millennium Cohort Study.
+We also restrict our sample to non-multiparous individuals born in England.
+Finally, we restrict the sample to cases where there is complete information on socioemotional skills at five years of age.
+The final sample contains 9,545 individuals from the British Cohort Study, and 5,572 from the Millennium Cohort Study.
+Summary statistics for the full and estimation samples are displayed in \autoref{tab:sumstats}.
+After the rebalancing step, the MCS estimation sample closely mirrors the full sample in terms of average observable characteristics, thus preserving representativeness.
 
@@
 Third, the different ordering of the items within each scale might lead to order effects.
-Fourth, and no less importantly, the interpretation of each behaviour by respondents living 30 years apart (1975 vs 2005/6) might differ due to a host of evolving societal norms.
+Fourth, and no less importantly, the interpretation of each behaviour by respondents living 30 years apart (1975 vs 2006) might differ due to a host of evolving societal norms.
 
@@
 In what follows, we will consider the included items to be the same \emph{measure} in the two cohorts.
-The wording of the items we will be using in the analysis is presented in \autoref{tab:scalecomp}: we retain 13 items for the BCS (two of them are grouped) and 11 for the MCS with high degree of comparability.\footnote{We exclude from the analysis items that were completely different between the two questionnaires, although we could have included them in the factor analysis and treated them as missing in the cohort were they were not administered.
-While this could have improved efficiency, we decided to rely on a more coherent set of measures to maximise comparability between the two cohorts.}.
+The wording of the items we will be using in the analysis is presented in \autoref{tab:scalecomp}: we retain 13 items for the BCS (two of them are grouped) and 11 for the MCS with high degree of comparability.
+We exclude from the analysis items that were completely different between the two questionnaires, although we could have included them in the factor analysis and treated them as missing in the cohort were they were not administered.
+While this could have improved efficiency, we decided to rely on a more coherent set of measures to maximise comparability between the two cohorts.
+More details on the derivation of the scale are available in Appendix \autoref{asec:scales}.
+
 Item-level prevalence by cohort and gender is in \autoref{tab:meantab}.
 We see that, in general, item prevalence is more similar across genders within the same cohort, than across cohorts.
-For the majority of items, there is a lower prevalence of problematic behaviours in the MCS than in the BCS; however, four items (distracted, tantrums, fearful, aches) show a higher prevalence in 2005 than in 1975.
+For the majority of items, there is a lower prevalence of problematic behaviours in the MCS than in the BCS; however, four items (distracted, tantrums, fearful, aches) show a higher prevalence in 2006 than in 1975.
 Regardless, a simple cross-cohort comparison of item-level prevalence is misleading because of changing perceptions and norms about what constitutes problematic behaviour in children.
@@
 In the remainder of this section, we analyse the properties of the new scale.
-In particular, we study the \emph{factor structure} of our scale.
-Namely, we establish how many latent dimensions of socio-emotional skills the scale is capturing, and which items are measuring which dimension.
-We then estimate the parameters of the factor models that corresponds to our choice of dimension and attribution of specific items to factors.
-In the following section, we investigate to what extent socio-emotional skills are measured in the same way across cohorts.
+Following a common approach, we proceed in two steps.
+First, we carry out an \emph{exploratory} step, where we study the factor structure of our scale.
+The aim of this step is to examine the correlation between observed measures in a data-driven way, imposing the least possible assumptions.
+Here, we establish how many latent dimensions of socio-emotional skills the scale is capturing, and which items of our scale are measuring which dimension.
+As a second step, we set up a \emph{confirmatory} factor model.
+This model fixes the number of latent dimensions, and imposes a dedicated measurement structure, based on the insights obtained in the exploratory step.
+This is the model from which the measurement invariance analysis of \autoref{sec:measinv} departs.
 
@@
 The EFA is performed decomposing the polychoric correlation matrix of the items and using weighted least squares, and the solution is rescaled using oblique factor rotation (\emph{oblimin}).
-We use the \textbf{\textsf{R}} package \texttt{psych}, version 1.8.4 \citep{Revelle2018}.} We start by investigating the number of latent constructs that are captured by the scale, using different methods developed in the psychometric literature, and now also used in the economics literature.
+We use the \textbf{\textsf{R}} package \texttt{psych}, version 1.8.4 \citep{Revelle2018}.} We start by investigating the number of latent constructs that are captured by the scale, using different methods developed in the psychometric literature, and recently adopted by the economics literature.
 The results are displayed in \autoref{tab:numfac}.
@@
 This choice is justified both by the child psychology literature cited above, and as compromise to work with the same number of factors in the two cohorts.
-The two-factor EFA delivers a neat and sensible separation between items, as shown in \autoref{tab:load}: reassuringly, similarly-worded items load on the same factor across the two cohorts, and also the magnitude of the respective loadings (measuring the strength of the association between the item and the factor) is very smilar.
-Following previous research, we name the first dimension \emph{Externalising skills} (EXT, comprising the items restless, squirmy/fidgety, fights/bullies, distracted, tantrums, and disobedient) and the second dimension \emph{Internalising skills} (INT, comprising the items worried, fearful, solitary, unhappy, and aches).
+
+The two-factor EFA delivers a neat and sensible separation between items, as shown in \autoref{tab:load}: reassuringly, similarly-worded items load on the same factor across the two cohorts, and also the magnitude of the respective loadings (measuring the strength of the association between the item and the factor) is very similar.
+Following previous research, we name the first dimension \emph{Externalising skills} (EXT, indicating low scores on the items restless, squirmy/fidgety, fights/bullies, distracted, tantrums, and disobedient) and the second dimension \emph{Internalising skills} (INT, indicating low scores on the items worried, fearful, solitary, unhappy, and aches).\footnote{Internalising and externalising dimensions emerge from the exploratory step on our novel 11-item scale.
+Appendix \autoref{asec:fullefa} performs the same exploratory steps on the full set of Rutter items in BCS and SDQ items in MCS.
+It confirms that the items we consider in our subscale have a broadly consistent covariance structure even when considered in the context of their original scales.
+Appendix \autoref{asec:no511} considers the robustness of our results to the exclusion of the items of the scale that perform most poorly.}
 
@@
 Each individual child is denoted by $j=1\dots N_c$, where $N_c$ is the number of children in cohort $c$.
-For each child $j$ in cohort $c$, we observe categorical items $X_{ijc}$ with $i=1,\dots,11$, corresponding to the eleven maternal reports in \autoref{tab:scalecomp}.
-We assume that each child is characterised by a latent bi-dimensional vector of externalising and internalising socio-emotional skills $\bm{\theta}_{jc} = (\theta_{jc}^{EXT},\theta_{jc}^{INT})$, as shown by the EFA in the previous section.
+For each child $j$ in cohort $c$, we observe categorical items $X_{ijc}$ with $i=1,\dots,11$ corresponding to the eleven maternal reports in \autoref{tab:scalecomp}.
+We assume that each child is characterised by a latent bidimensional vector of externalising and internalising socio-emotional skills $\bm{\theta}_{jc} = (\theta_{jc}^{EXT},\theta_{jc}^{INT})$.
 
@@
 with $\tau_{0,ic}=-\infty$ and $\tau_{3,ic}=+\infty$.
-Notice that we recode all ordered items to have higher values for \emph{better} behaviours, so that our latent vectors can be interpreted as favourable skills and not behavioural problems.
-
-The model implies the following expression for the mean and covariance structure of the latent propensities:
+Notice that we recode all ordered items to have higher values for \emph{better} behaviours, so that our latent vectors can be interpreted as favourable skills and not behavioural problems.\footnote{The model implies the following expression for the mean and covariance structure of the latent propensities:
 $$\bm{\mu}_c=\bm{\nu}_c + \bm{\Lambda}_c\bm{\kappa}_c \qquad \text{and} \qquad \Sigma_c = \bm{\Lambda}_c \bm{\Phi}_c \bm{\Lambda}_c^\prime + \bm{\Psi}_c.$$
-
-The model restrictions in \eqref{eq:propens} and \eqref{eq:thresh} do not identify the parameters without additional assumptions.
-As per the traditional factor analysis approach, we impose a normal distribution on the latent skills and error terms:\footnote{Recent work has also used mixtures of normals for the latent factors distribution, e.g.
-\citet{Conti2010a}.}
+As per the traditional factor analysis approach, we impose a normal distribution on the latent skills and error terms. 
 \be\label{eq:norm}
@@
 \ee
+Recent work has also used mixtures of normals for the latent factors distribution, e.g. \citet{Conti2010a}.}
 
-Even with these assumptions, there are infinite equivalent parameterisations through which the model can be identified -- the well-known issue of factor indeterminacy.
-We follow common practice and identify the model by setting the mean $\bm{\kappa}$ and variance $\bm{\Phi}$ of the latent skill factor in both cohorts to zero and one, respectively.
-Furthermore, we set intercepts to zero and error variances to one.
-Loadings $\lambda$ and thresholds $\tau$ are instead allowed to vary across cohorts.
+\section{Measurement invariance \label{sec:measinv}}
 
+\subsection{The configural model \label{sec:config}}
+
+Measurement invariance analysis necessarily starts starts from a minimally restrictive model, denominated \emph{configural} model.
+This is a `minimum' identifiable model, in that it places the least possible restrictions on how parameters are allowed to vary across cohorts.
+It serves as a basis for our measurement invariance analysis.
+
+The restrictions implied in \eqref{eq:propens} and \eqref{eq:thresh} are not sufficient to identify the parameters of the model.
+Even with these assumptions, there are infinite equivalent parameterisations (or rotations) that deliver a minimally restrictive configural model.
+This is the well-known issue of factor indeterminacy, which arises due to the lack of natural units of measurement for the latent factors being assessed.
+
+Further sets of restrictions are thus required to set the location and scale of the latent factors. Among the most straightforward and widely used parameterisations for the configural model are:
+\bi[label=$\diamond$]
+\item \underline{Delta parameterisation} {[WE{\textDelta}]} \citep{Wu2016a}
+
+For all groups: $$\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}, \quad \bm{\nu}=\bm{0}, \quad \text{and} \; \text{diag}(\bm{\Sigma})=\bm{I}.$$
+
+\item \underline{Theta parameterisation} {[WE{\textTheta}]} \citep{Wu2016a}
+
+For all groups: 
 \be\label{eq:thetapar}
-\text{diag}(\bm{\Phi}_c) = \bm{I}, \quad \bm{\kappa}_c=\bm{0}, \quad \bm{\nu}_c=\bm{0}, \quad \text{and} \; \text{diag}(\bm{\Psi}_c) = \bm{I} \qquad \forall c \in \{BCS,MCS\}.
+\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}, \quad \bm{\nu}=\bm{0}, \quad \text{and} \; \text{diag}(\bm{\Psi})=\bm{I}.
 \ee
 
-The restrictions in \eqref{eq:propens}, \eqref{eq:thresh}, \eqref{eq:norm}, and \eqref{eq:thetapar} define the so-called \emph{configural} model.
-This is a `minimum' identifiable model, in that it places the least possible restrictions on how parameters are allowed to vary across cohorts.
-It serves as a basis for our measurement invariance analysis in the next section.\footnote{This set of identifying restrictions is known as \emph{Theta} parameterisation \citep{Wu2016a}.
-See \aref{asec:midet} for statistically equivalent alternative parameterisations.}
+\item \underline{Anchored parameterisation} {[MT]} \citep{Millsap2004}
+  \bi
+  \item For all groups, normalise a reference loading to 1 for each factor
+  \item Set invariant across groups one threshold per item (e.g. $\tau_{0,Ai} = \tau_{0,Bi}$), and an additional threshold in the reference items above
+  \item In the first group: $\bm{\kappa_A}=\bm{0}$, $\text{diag}(\bm{\Sigma}_A)=\bm{I}$
+  \item Set all intercepts $\bm{\nu}$ to zero
+  \ei
+\ei
+The first two parameterisations (WE{\textDelta} and WE{\textTheta}) normalise the mean and variance of factors to the same constants in both groups, and they leave all loadings and intercepts to be freely estimated; they only differ in whether the additional required normalisation is imposed on the variances of the error terms ($\bm{\Psi}$) or on the diagonal of the covariance matrix of the measures ($\bm{\Sigma}$).
+The MT parameterisation instead proceeds by identifying parameters in one group first, and then imposing cross-group equality constraints to identify parameters in other groups \citep{Wu2016a}.
+Still, all of these parameterisations are statistically equivalent.
+The measurement invariance analysis in this paper is based on the Theta parameterisation (WE{\textTheta}), but results are independent on this choice.
+The restrictions in \eqref{eq:propens}, \eqref{eq:thresh}, and \eqref{eq:thetapar} define the so-called \emph{configural} model.
 
-\section{Measurement invariance \label{sec:measinv}}
+\subsection{Nested models \label{sec:nested}}
 
@@
 In this paper, we follow the recent identification methodology by \citet{Wu2016a}.
-The configural model defined in the previous section serves as the starting point.
+The configural model defined above serves as the starting point.
 Measurement invariance is then assessed by comparing the configural model to a series of hierarchically nested models.
@@
 Intuitively, if the additional cross-group restrictions have not significantly worsened model fit, one can conclude that a certain level of invariance is achieved.
-The hierarchy of restrictions is detailed in \autoref{tab:invparam}.
+
+In the case where available measures are continuous, MI analysis is straightforward \citep{vandeSchoot2012}.
+The hierarchy of the nested models usually proceeds by testing loadings first, and then intercepts (to establish \emph{metric} and \emph{scalar} invariance -- see \citealp{Vandenberg2000a}).
+
+Invariance of systems with categorical measures, such as the scale we examine in this paper, is less well understood.
+In particular, the lack of explicit location and scale in the measures introduces an additional set of parameters compared to the continuous case (thresholds $\tau$).
+This makes identification reliant on more stringent normalisations.
+A first comprehensive approach for categorical measures was proposed by \cite{Millsap2004}.
+New identification results in \cite{Wu2016a} indicate that, in the categorical case, invariance properties cannot be examined by simply restricting one set of parameters at a time.
+This is because the identification conditions used in the configural baseline model, while being minimally restrictive on their own, become binding once certain additional restrictions are imposed.
+In light of this, they propose models that identify structures of different invariance levels.
+They find that some restrictions cannot be tested alone against the configural model, because the models they generate are statistically equivalent.
+This is true of loading invariance, and also of threshold invariance in the case when the number of categories of each ordinal item is 3 or less.
+Furthermore, they suggest that comparison of both latent means and variances requires invariance in loadings, thresholds, and intercepts.
+A summary of the approach by \cite{Wu2016a} is available in \autoref{tab:invparam}.
 
@@
 
-We estimate the sequence of models detailed in \autoref{tab:invparam} by Weighted Least Squares.\footnote{Parameters are estimated by mean- and variance-adjusted weighted least squares (WLSMV) -- see \citet{Muthen1997}; estimation starts from the items' polychoric correlation matrix, uses diagonally weighted least squares (DWLS), and exploits the full weight matrix to compute robust standard errors and test statistics.
-Robust WLS has proved in simulation studies to be moderately robust to small violations of the normality assumption in the latent underlying measures \citep{Flora2004}, and generally outperforms maximum likelihood in large samples \citep{Beauducel2006,Li2016}.
-All estimates are computed using the \texttt{lavaan} package (version 0.6-2) in \textbf{\textsf{R}} \citep{Rosseel2012}.}
+We estimate the sequence of models detailed in \autoref{tab:invparam} by mean- and variance-adjusted weighted least squares (WLSMV) -- see \citet{Muthen1997}; estimation starts from the items' polychoric correlation matrix, uses diagonally weighted least squares (DWLS), and exploits the full weight matrix to compute robust standard errors and test statistics.
+Robust WLS has proved in simulation studies to be moderately robust to small violations of the normality assumption in the latent underlying measures \citep{Flora2004}, and generally outperforms maximum likelihood in large samples \citep{Beauducel2006,Li2016}.\footnote{All estimates are computed using the \texttt{lavaan} package (version 0.6-2) in \textbf{\textsf{R}} \citep{Rosseel2012}.}
 For the purposes of the analysis, we define groups $c$ as cohort-gender cells, with the reference group being males in the BCS cohort.
 We then compare the fit of each model against the configural model.
+
+\subsection{Measurement invariance results \label{sec:mires}}
 
@@
 The use of approximate fit indices (AFIs) is therefore recommended alongside $\chi^2$.
-These indices do not have a known sampling distribution, thus making it necessary to rely on rules of thumb to assess what level of \textDelta AFI indicates invariance.
-Nevertheless, AFIs are widely used in empirical practice to assess model fit.\footnote{The root mean squared error of approximation (RMSE) and the Tucker-Lewis index (TLI) are traditionally the most used AFIs in empirical practice.
-Simulation evidence by \citet{Cheung2002} shows that these indices can show correlation between overall and relative fit, and suggest relying on additional indices, such as the comparative fit index (CFI, \citealp{Bentler1990}), McDonald non-centrality index (MFI, \citealp{McDonald1989}), and Gamma-hat index \citep{Steiger1989} for the case of ordered measures.
-Commonly accepted thresholds for rejection are $\Delta CFI < - 0.01$, $\Delta MFI < - 0.02$, and $\Delta Gamma hat < - 0.001$.
-\citet{Meade2008}, using the results from a simulation study, suggests stricter thresholds that should apply in a variety of conditions.
-For CFI, a single cutoff value of $.002$ is proposed, while cutoffs for MFI depend on the problem's characteristics; in our case (2 factors, 11 items), they suggest $.0066$.
-\citet{Sass2014} however cast some doubts of the generalisability of these cutoffs to WLSMV estimators.}
+While these indices successfully adjust for model complexity \citep{Cheung2002}, they do not have a known sampling distribution.
+This makes it necessary to rely on simulation studies, which derive rules of thumb indicating what level of \textDelta AFI is compatible with invariances invariance.
 
-The fit of each model is compared in Panel A of \autoref{tab:fit}.
+Again, just like in the broader context of measurement invariance, most evidence regarding the performance of AFIs pertains to scenarios with continuous measures.
+The root mean squared error of approximation (RMSE) and the Tucker-Lewis index (TLI) are traditionally the most used AFIs in empirical practice.
+Simulation evidence by \citet{Cheung2002} shows that these indices can show correlation between overall and relative fit, and suggest relying on additional indices, such as the comparative fit index (CFI, \citealp{Bentler1990}), McDonald non-centrality index (MFI, \citealp{McDonald1989}), and Gamma-hat index \citep{Steiger1989}.
+Subsequent simulation studies -- e.g.
+\citet{Chen2007} and \citet{Meade2008} -- have updated these thresholds for the continuous case.
+In particular, \citet{Chen2007} suggests the following thresholds for rejecting measurement invariance: \textDelta RMSE $> .015$, \textDelta CFI $< - .010$, \textDelta SRMR $> .010$.
+
+It is not advisable to directly extrapolate rules of thumb derived from simulations with continuous measures to the categorical case \citet{Lubke2004}.
+Recent studies have advanced simulation evidence around the performance of AFIs in measurement invariance analysis with categorical measures.
+\citet{Sass2014} finds that the cutoffs from \citet{Chen2007} might not generalise well to problems estimated by WLSMV, but this is mostly confined to smaller sample sizes and detection of small degrees of non-invariance.
+More recently, \citet{Rutkowski2017} find that a \textDelta RMSE threshold of $.010$ is appropriate for testing equality of slopes and thresholds when the sample size is large, like in our case.
+
+In any case, we present a range of fit indices to provide a more holistic assessment of measurement invariance.
+The fit of each nested model is compared in Panel A of \autoref{tab:fit}.
+The overall fit of the configural model is satisfactory according to all indices, with CFI around $.95$ and RMSE just above $.05$.
+As expected, given our large sample size, $\chi^2$-based tests reject measurement invariance at all levels.
 The model with restricted thresholds and loadings exhibits a comparable fit to the configural model, according to all the AFIs.
+In particular, the \textDelta AFIs fall within the ranges suggested in \citet{Chen2007} and \citet{Rutkowski2017}.
 Invariance of loadings and thresholds across cohorts implies that items in our scale are equally salient in their informational content, and that the latent propensities have equal mapping into the observed items.
-However, further restricting intercepts results in a model where invariance is rejected across the board.\footnote{We do not present fit results for the threshold-only invariance model, as it is statistically equivalent to the configural model and thus its fit is mathematically the same -- see Table 3 in \citealp{Wu2016a}.
+
+However, further restricting intercepts results in a model where invariance is rejected across the board.
+In other words, intercept parameters in our model ($\nu$) are estimated to be different between maternal reports in the British and Millennium Cohort Studies.
+This means that, for a given level of latent skills, mothers in MCS tend to assess behaviours differently from mothers in BCS.
+Thus, cohort differences in scores on our scale cannot be unequivocally interpreted as differences in the underlying skills, since they might also reflect differences in reporting.\footnote{We do not present fit results for the threshold-only invariance model, as it is statistically equivalent to the configural model and thus its fit is mathematically the same -- see Table 3 in \citealp{Wu2016a}.
 The ages at which socio-emotional skills are observed varies slightly between BCS and MCS, due to different sampling and fieldwork schedules.
@@
 In Panel B of \autoref{tab:fit}, we restrict the sample to 59 to 61 months, where the overlap between BCS and MCS is maximised.
-In Panels C and D, we restrict to male and female children respectively.
-In all these cases, invariance of thresholds and loadings is achieved, but invariance of intercepts is rejected.
-We can thus rule out that the lack of intercept invariance comes from differences in ages or invariance across child gender.} In other words, intercept parameters in our model ($\nu$) are estimated to be different between maternal reports in the British and Millennium Cohort Studies.
-This means that, for a given level of latent skills, mothers in MCS tend to assess behaviours differently from mothers in BCS.
-Thus, cohort differences in scores on our scale cannot be unequivocally interpreted as differences in the underlying skills, since they might also reflect differences in reporting.
+In Panel C, we repeat the analysis with the full sample, but excluding the poorest-performing items (5 and 11) -- see Appendix \autoref{asec:no511} for details.
+In Panels A and B of \autoref{tab:fit_gend}, we restrict to male and female children respectively.
+In all these cases, invariance of thresholds and loadings is confirmed, but invariance of intercepts is rejected.
+We can thus rule out that the lack of intercept invariance comes from differences in ages or invariance across child gender.}
 
@@
 A possible interpretation is connected with secular evolution of social and cultural norms about child behaviours.
-For example, commonly held views of what constitutes a restless, distracted, or unhappy child might have changed between 1975 and 2005/6.\footnote{Calibrating the Rutter and SDQ using a contemporary sample of children cannot rule out this issue.
+For example, commonly held views of what constitutes a restless, distracted, or unhappy child might have changed between 1975 and 2006.\footnote{Calibrating the Rutter and SDQ using a contemporary sample of children cannot rule out this issue.
 For example, \citet{Collishaw2004} administered both Rutter and SDQ items to parents of a small sample of adolescents in London.

### 2019-02-28 codes/R/paper/text/paper_jpube_3resultsconclusion.tex
new version of text and tables - incorporates most reviewer comments

@@
 \autoref{fig:ineqsm} shows an even starker pattern when comparing children of mothers who smoked in pregnancy with non-smoking mothers.
-The fact that maternal smoking during pregnancy is a risk factor for offspring behavioural problems is well known in the medical literature \citep{Gaysina2013}; less evidence there is however, on whether and to which extent these associations have changed across cohorts.
+The fact that maternal smoking during pregnancy is a risk factor for offspring behavioural problems is well known in the medical literature \citep{Gaysina2013}; there is less evidence, however, on whether and to which extent these associations have changed across cohorts.
 The difference in child skills has increased, from less than .2 to around .4 of a standard deviation, again with the biggest increase experienced by the boys.
@@
 Being born to an unmarried mother, and to a mother who smoked during pregnancy, is associated with a higher penalty for both dimensions of socio-emotional skills in the latter cohort, but only for males.
-This is consistent with recent evidence which shows that family disadvantage disproportionately impedes the pre-market development of boys, in terms of higher disciplinary problems, lower achievement scores, and fewer high-school completions \citep{Autor2016a}.
-Girls of non-white ethnicity, instead, have worse internalising and externalising skills in the MCS, a penalty not suffered by 5-year old non-white girls in the BCS.
+This is consistent with recent evidence which shows that family disadvantage disproportionately impedes the pre-market development of boys, in terms of higher disciplinary problems, lower achievement scores, and fewer high-school completions \citep{Autor2016a}.\footnote{It is important to underscore that there has been a significant rise is cohabitation between 1975 and 2006.
+It is likely that unmarried mothers in the two cohorts have very different characteristics.
+The choice of this indicator is due to the absence of information on cohabitation in the birth survey for the BCS cohort.} Girls of non-white ethnicity, instead, have worse internalising and externalising skills in the MCS, a penalty not suffered by 5-year old non-white girls in the BCS.
 Firstborn boys and girls in the BCS have worse skills, but this difference disappears in the MCS.
@@
 In general, this has been a period of significant societal changes, with an almost continual rise in the proportion of women in employment, an older age at first birth and a rise in dual-earning parents families \citep{Roantree2018}.
-However, here we do not attempt to disentangle whether and to which extent the observed changes in inequality in socio-emotional skills across the two cohorts can be attributed to changes in returns (or penalties) to maternal characteristics (such as education and employment) or to compositional changes, like it has been done for the analysis of wage inequality \citep{Blundell2007}.
+However, here we do not attempt to disentangle whether and to which extent the observed changes in inequality in socio-emotional skills across the two cohorts can be attributed to changes in returns (or penalties) to maternal characteristics (such as education and employment) or to compositional changes, as has been done for the analysis of wage inequality \citep{Blundell2007}.
 
@@
 We contribute to a vast interdisciplinary literature by examining medium- and long-term impacts of skills measured at an earlier age than usually examined in previous studies, well before the start of formal education.
-We do so by regressing health and socioeconomic outcomes measured in adolescence and adulthood on the socio-emotional skills scores at age five obtained by our factor model, controlling for the harmonised family background variables at birth and age five (see \autoref{tab:harmvar}).
-We present results with and without controlling for cognitive skills.
+We do so by regressing health and socioeconomic outcomes measured in adolescence and adulthood on the socio-emotional skills scores at age five obtained by our factor model, controlling for the harmonised family background variables at birth and age five (see \autoref{tab:harmvar}).\footnote{In \autoref{tab:bcs_adultapp}, we show that the conclusions in this section are not sensitive to the factor scoring methodology used.
+'Raw' score, obtained by simple summing of the item categories in the 11-item subscale have basically equal predictive power to factor scores.} We present results with and without controlling for cognitive skills.
 As detailed in \autoref{sec:data}, the available cognitive measures are not comparable across cohorts.
@@
 A child with better internalising skills is less solitary, neurotic, and worried.
-From this perspective, he is likely more sociable and subject to peer influence in health behaviours.
+From this perspective, she is likely more sociable and subject to peer influence in health behaviours.
 This is consistent with the evidence in \citet{Goodman2015}, who find a positive association between child emotional health (measured with items for the internalizing behaviour subscale of the Rutter scale at age 10 in the BCS) and smoking at age 42.
@@
 
-For the British Cohort Study, we last examine the association between socio-emotional skills at age five and adult education and labour market outcomes.
+Cohort members from the British Cohort Study are now well into their adulthood.
+For this cohort, we can examine the association between socio-emotional skills at age five and adult education and labour market outcomes.
 The structure of \autoref{tab:bcs_adult} is similar to \autoref{tab:adol}, but it considers educational achievement, employment, and earnings (conditional on being in paid employment) for the BCS cohort members.
@@
 Again, we show that the greater predictive power of cognition for socioeconomic outcomes holds even when considering earlier-life measures of child development.
+Similarly, \citet{Goodman2015} use data from the same cohort, but exploit a richer set of measures from the age 10 survey.
+They show that social and emotional skills measured at age 10 are strongly related to a wide range of adult outcomes.
 

### 2019-02-28 codes/R/paper/text/paper_jpube_4appendix.tex
new version of text and tables - incorporates most reviewer comments

@@
 `Certainly applies' = 0, `Somewhat applies' = 1, `Does not apply' = 2.
-We augment the 19-item Rutter Scale with four additional parent-reported questions from the parental questionnaire, items A to D.
+We augment the 19-item Rutter Scale with three additional parent-reported questions from the parental questionnaire, items A, B, and D.
 These are rated on 4 levels: `Never in the last 12 months', `less than once a month', `at least once a month', `at least once a week'.
@@
 
-\section{Measurement invariance details}\label{asec:midet}
+\section{Robustness of exploratory analysis}\label{asec:fullefa}
 
-\subsection{Alternative parameterisations for the configural model}
+In this section, we repeat the exploratory analysis step in \autoref{sec:ea} for the full set of Rutter and SDQ items.
+This is to show that the factor structure emerging from exploratory analysis of the 11-item subscale is consistent with what would emerge considering the original scales in their entirety.
+Again, we proceed by first assessing the optimal number of factors, and then examining the loading obtained from exploratory factor analysis.
 
-There are infinite ways to parameterise the configural model defined by \eqref{eq:propens} and \eqref{eq:thresh}. Widely used parameterisations are:
-\bi[label=$\diamond$]
-\item \underline{Delta parameterisation} {[WE{\textDelta}]} \citep{Wu2016a}
+Results for the optimal number of factors as indicated by different approaches are in \autoref{tab:numfac_all}.
+Similarly to the 11-item subscale, there is not much agreement between methods.
+Since the purpose of this section is to assess the robustness of the 11-item subscale, we adopt a conservative approach by estimating EFA models with the largest number of factors suggested, i.e.
+five.
+In this way, we allow for richer factor solutions, that have more power to disprove our simpler two-factor solution for the novel subscale.
 
-For all groups: $$\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}, \quad \bm{\nu}=\bm{0}, \quad \text{and} \; \text{diag}(\bm{\Sigma})=\bm{I}.$$
+\autoref{tab:load_bcsall} presents factor loadings for the Rutter scale in BCS, with the addition of the "headaches/stomachaches" and "tantrums" items (see Appendix \autoref{asec:scales}).
+The split between externalising and internalising items that we recover in the 11-item scale is almost entirely preserved in the full scale, as seen by items loading on factors 1 and 2.
+The only exception is the "headaches/stomachaches" item, which seems to load on a separate factor.
+We carry out robustness checks for the measurement invariance analysis excluding this item in Appendix \autoref{asec:no511}.
 
-\item \underline{Theta parameterisation} {[WE{\textTheta}]} \citep{Wu2016a}
+The same analysis is repeated for the SDQ scale in MCS in \autoref{tab:load_mcsall}.
+An internalising, emotional dimension (factor 3) emerges neatly, and coherently with the analysis on our subscale.
+The externalising items from our subscales are split across two dimensions in this full-scale EFA: one more related to hyperactivity (factor 2) and one to conduct problems (factor 4).
+This is consistent with the original structure of the SDQ \citep{Goodman1997}.
 
-For all groups: $$\text{diag}(\bm{\Phi}) = \bm{I}, \quad \bm{\kappa}=\bm{0}, \quad \bm{\nu}=\bm{0}, \quad \text{and} \; \text{diag}(\bm{\Psi})=\bm{I}.$$
+\section{Robustness of item choice}\label{asec:no511}
 
-\item \underline{Anchored parameterisation} {[MT]} \citep{Millsap2004}
-  \bi
-  \item For all groups, normalise a reference loading to 1 for each factor
-  \item Set invariant across groups one threshold per item (e.g. $\tau_{0,Ai} = \tau_{0,Bi}$), and an additional threshold in the reference items above
-  \item In the first group: $\bm{\kappa_A}=\bm{0}$, $\text{diag}(\bm{\Sigma}_A)=\bm{I}$
-  \item Set all intercepts $\bm{\nu}$ to zero
-  \ei
-\ei
-The first two parameterisations (WE{\textDelta} and WE{\textTheta}) normalise the mean and variance of factors to the same constants in both groups, and they leave all loadings and intercepts to be freely estimated; they only differ in whether the additional required normalisation is imposed on the variances of the error terms ($\bm{\Psi}$) or on the diagonal of the covariance matrix of the measures ($\bm{\Sigma}$).
-The MT parameterisation instead proceeds by identifying parameters in one group first, and then imposing cross-group equality constraints to identify parameters in other groups \citep{Wu2016a}.
+In deriving our novel 11-item scale, we construct two items for the BCS cohort based on questions that are not in the original Rutter scale -- namely those concerning "headaches/stomachaches" and "tantrums" (see Appendix \autoref{asec:scales} above for details).
+Concerns might arise that introducing these items might somewhat invalidate our main conclusions, rather than provide additional informational content on children's externalising and internalising behaviours and symptoms.
 
-\subsection{Identification of models with different levels of invariance}
+In fact, exploratory factor analysis (on both on the full Rutter and SDQ scales and on the 11-item subscale) shows that these items, numbered 5 and 11, perform poorly and exhibit relatively low factor loadings.
+As a robustness check, we replicate the main results of the paper by excluding them from the subscale.
 
-In the case where available measures are continuous, MI analysis is straightforward \citep{vandeSchoot2012}.
-The hierarchy of the nested models usually proceeds by testing loadings first, and then intercepts (to establish \emph{metric} and \emph{scalar} invariance -- see \citealp{Vandenberg2000a}).
-
-Invariance of systems with categorical measures, such as the scale we examine in this paper, is less well understood.
-In particular, the lack of explicit location and scale in the measures introduces an additional set of parameters compared to the continuous case (thresholds $\tau$).
-This makes identification reliant on more stringent normalisations.
-A first comprehensive approach for categorical measures was proposed by \cite{Millsap2004}.
-New identification results in \cite{Wu2016a} indicate that, in the categorical case, invariance properties cannot be examined by simply restricting one set of parameters at a time.
-This is because the identification conditions used in the configural baseline model, while being minimally restrictive on their own, become binding once certain additional restrictions are imposed.
-In light of this, they propose models that identify structures of different invariance levels.
-They find that some restrictions cannot be tested alone against the configural model, because the models they generate are statistically equivalent.
-This is true of loading invariance, and also of threshold invariance in the case when the number of categories of each ordinal item is 3 or less.
-Furthermore, they suggest that comparison of both latent means and variances requires invariance in loadings, thresholds, and intercepts.
-A summary of the approach by \cite{Wu2016a} is available in \autoref{tab:invparam}.
+Panel C of \autoref{tab:fit} shows that the measurement invariance analysis yields the same qualitative results once these two items are included.
+\autoref{fig:scatter_no511} shows a scatter plot of the factor scores obtained from the factor model with and without items 5 and 11.
+They exhibit very high correlation, thus indicating that our results in \autoref{sec:results} would not substantially change if we omitted the two items with the least informational content.

### 2019-02-28 codes/R/paper/text/paper_jpube_5apptables.tex
new version of text and tables - incorporates most reviewer comments

@@
 \begin{enumerate}
-\item Very restless. Often running about or jumping up and down. Hardly ever still.$^\dagger$
-\item Is squirmy or fidgety.$^\dagger$
+\item Very restless. Often running about or jumping up and down. Hardly ever still.$^*$
+\item Is squirmy or fidgety.$^*$
 \item Often destroys own or others' belongings.
-\item Frequently fights other children.$^\dagger$
+\item Frequently fights other children.$^*$
 \item Not much liked by other children.
-\item Often worried, worries about many things.$^\dagger$
-\item Tends to do things on his/her own, is rather solitary.$^\dagger$
+\item Often worried, worries about many things.$^*$
+\item Tends to do things on his/her own, is rather solitary.$^*$
 \item Irritable. Is quick to fly off the handle.
-\item Often appears miserable, unhappy, tearful or distressed.$^\dagger$
+\item Often appears miserable, unhappy, tearful or distressed.$^*$
 \item Sometimes takes things belonging to others.
@@
 \item Frequently bites nails or fingers.
-\item Is often disobedient.$^\dagger$
-\item Cannot settle to anything for more than a few moments.$^\dagger$
-\item Tends to be fearful or afraid of new things or new situations.$^\dagger$
+\item Is often disobedient.$^*$
+\item Cannot settle to anything for more than a few moments.$^*$
+\item Tends to be fearful or afraid of new things or new situations.$^*$
 \item Is over fussy or over particular.
 \item Often tells lies.
-\item Bullies other children.$^\dagger$
-\item[A.] Complains of headaches.$^\dagger$
-\item[B.] Complains of stomach-ache or has vomited.$^\dagger$
-\item[C.] Complains of biliousness.
-\item[D.] Has temper tantrums (that is, complete loss of temper with shouting, angry movements, etc.).$^\dagger$
+\item Bullies other children.$^*$
+\item[A.] Complains of headaches.$^*$
+\item[B.] Complains of stomach-ache or has vomited.$^*$
+\item[D.] Has temper tantrums (that is, complete loss of temper with shouting, angry movements, etc.).$^*$
 \end{enumerate}
@@
 \item Considerate of other people's feelings.
-\item Restless, overactive, cannot stay still for long.$^\dagger$
-\item Often complains of head- aches, stomach-ache or sickness.$^\dagger$
+\item Restless, overactive, cannot stay still for long.$^*$
+\item Often complains of head- aches, stomach-ache or sickness.$^*$
 \item Shares readily with other children (treats, toys, pencils, etc.).$^+$
-\item Often has temper tantrums or hot tempers.$^\dagger$
-\item Rather solitary, tends to play alone.$^\dagger$
-\item Generally obedient, usually does what adults request.$^{\dagger +}$
-\item Many worries, often seems worried.$^\dagger$
+\item Often has temper tantrums or hot tempers.$^*$
+\item Rather solitary, tends to play alone.$^*$
+\item Generally obedient, usually does what adults request.$^{* +}$
+\item Many worries, often seems worried.$^*$
 \item Helpful if someone is hurt, upset or feeling ill.$^+$
-\item Constantly fidgeting or squirming.$^\dagger$
+\item Constantly fidgeting or squirming.$^*$
 \item Has at least one good friend.$^+$
-\item Often fights with other children or bullies them.$^\dagger$
-\item Often unhappy, down-hearted or tearful.$^\dagger$
+\item Often fights with other children or bullies them.$^*$
+\item Often unhappy, down-hearted or tearful.$^*$
 \end{enumerate}
@@
 \item Generally liked by other children.$^+$
-\item Easily distracted, concentration wanders.$^\dagger$
-\item Nervous or clingy in new situations, easily loses confidence.$^\dagger$
+\item Easily distracted, concentration wanders.$^*$
+\item Nervous or clingy in new situations, easily loses confidence.$^*$
 \item Kind to younger children.$^+$
@@
 \end{tabular}
-\justify {\scriptsize \emph{Notes}: Items denoted by $^+$ are positively coded in the original scale. Items denoted by $^\dagger$ are retained in the new comparable scale.
+\justify {\scriptsize \emph{Notes}: Items denoted by $^+$ are positively worded in the original scale. Items denoted by $^*$ are retained in the new 11-item comparable scale.
 \par}
@@
 \caption{Description of harmonised variables}\label{tab:harmvar}
-\centering\setstretch{1.2} \footnotesize
-\begin{tabular}{L{.2\textwidth}C{.1\textwidth}L{.3\textwidth}L{.7\textwidth}}
+\centering\setstretch{1.2} \scriptsize
+\begin{tabular}{L{.18\textwidth}C{.1\textwidth}L{.3\textwidth}L{.75\textwidth}}
 \toprule
@@
 No father figure is a dummy for children whose father does not live in the same household.
+Father's social class was recorded using the SOC2000 classification in MCS.
+We use the derivation matrices kindly provided by David Pevalin at ISER (available at https://www.iser.essex.ac.uk/archives/nssec/derivations-of-social-class) to map SOC2000 into Registrar General Social Class.
 \\ \\[-1.5em]
@@
 \end{landscape}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{table}[ht]
-\centering\setstretch{1.2} \footnotesize
-\caption{Parameterisations for measurement invariance}\label{tab:invparam}
-\begin{tabular}{L{.18\textwidth}L{.45\textwidth}L{.3\textwidth}}
-\toprule
-\textbf{Invariance level}     & \textbf{Description} & \textbf{Restrictions} \\ \midrule \\
-Configural (WE{\textTheta})   
-    & \bi[label=$\cdot$] \item Minimally restrictive model for identification \ei
-    & For all groups: $\begin{array}{|l} \text{diag}(\bm{\Phi}) = \bm{I} \\
-                      \bm{\kappa}=\bm{0} \\
-                      \bm{\nu}=\bm{0} \\
-                      \text{diag}(\bm{\Psi})=\bm{I} \end{array}$
-                      \\ \\
-Threshold invariance
-    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ to be equal across groups
-                  \item Statistically equivalent to configural (when measures have 3 categories or less)\ei
-    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
-                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
-                      \newline For all groups: $\begin{array}{|l} \text{diag}(\bm{\Phi}) = \bm{I} \\ 
-                                                \bm{\kappa}=\bm{0} \end{array}$
-                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\nu}_A=\bm{0} \\
-                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \end{array}$
-                   \\ \\
-Threshold and Loading invariance
-    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ and loadings $\lambda$ to be equal across groups
-                  \item Allows comparison of latent factor variances \ei
-    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
-                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
-                      \newline $\lambda_{ci} = \lambda_{c^\prime i}$ for all items, $\forall c, c^\prime$
-                      \newline For all groups: $\bm{\kappa}=\bm{0}$
-                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\nu}_A=\bm{0} \\
-                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \\
-                                                  \text{diag}(\bm{\Phi}_A) = \bm{I} \end{array}$ 
-                   \\ \\
-Threshold, Loading, and Intercept invariance
-    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ and loadings $\lambda$ to be equal across groups
-                  \item Restricts intercepts $\nu$ to zero in both groups
-                  \item Allows comparison of latent factor variances \emph{and} means \ei
-    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
-                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
-                      \newline $\lambda_{ci} = \lambda_{c^\prime i}$ for all items, $\forall c, c^\prime$
-                      \newline For all groups: $\bm{\nu}=\bm{0}$
-                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\kappa}_A=\bm{0} \\
-                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \\
-                                                  \text{diag}(\bm{\Phi}_A) = \bm{I} \end{array}$ 
-\\ \\ \bottomrule
-\end{tabular}
-\justify {\scriptsize \emph{Notes}: Adapted from \cite{Wu2016a}.
-\par}
-\end{table}

### 2019-07-02 codes/R/paper/seskills_jpube_rev.Rnw
add funding acknowledgement

@@
 We would like to thank all cohort members and their families, who generously gave their time to participate in the surveys.
-We thank the DynaHEALTH consortium (H2020–633595 DynaHEALTH) for funding and support.
+This project has received funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme (grant agreement No 695300 - HKADeC - ERC-2015-AdG/ERC-2015-AdG.
+We also thank the DynaHEALTH consortium (H2020–633595 DynaHEALTH) for funding and support.
 GM is funded by an Economics and Social Research Council scholarship.}

### 2019-08-25 codes/R/paper/text/paper_jpube_1introlit.tex
text updates by Gabriella

@@
 
-The main contributions of the paper are four.
-First, we use two validated scales of childhood behavioural traits and select those items which are comparable across the two cohorts.
-By performing an exploratory factor analysis, we determine that we need at least two dimensions to characterise socio-emotional skills.
-We label them as `internalising' and `externalising' skills, the former relating to the ability of children to focus their drive and determination and the latter relating to their ability in engaging in interpersonal activities.
-
-Second, we study the comparability of the measures in the two cohorts.
+The main contributions of the paper are three.
+First, we construct a novel scale of of childhood behavioural traits from two validated instruments and assess its comparability across cohorts.
+By performing an exploratory factor analysis, we determine that two dimensions are a parsimonious representation of socio-emotional skills for both cohorts.
+Coherently with previous literature, we label them as `internalising' and `externalising' skills, the former relating to the ability of children to focus their drive and determination and the latter relating to their ability in engaging in interpersonal activities.
+Importantly, for the first time in economics, we study the comparability of the measures in the two cohorts.
 In particular, we test for \emph{measurement invariance} of the items we use to estimate the latent factors.
-Intuitively, if one assumes that a set measures is related to a latent unobserved factor of interest, one can think of this relationship as being driven by the saliency of each measure and the level.
+Intuitively, if one assumes that a set of measures is related to a latent unobserved factor of interest, one can think of this relationship as being driven by the saliency of each measure and the level.
 If one uses a given measure as the relevant metric for the relevant factor, its saliency will determine the scale of the factor, while some other parameters, which could be driven by the difficulty of a given test or the social norms and attitudes towards a certain type of behaviour, determine the \emph{average level} of the factor.
@@
 These results imply that while the level of inequality across the two cohorts in the skills we consider is comparable, we cannot determine whether the \emph{average levels} of the two factors are larger or smaller in one of the two cohorts.
+While this result hinders a comparison in the level of skills, it is of interest per se to find that mothers of English children born in England thirty years apart assess behaviours differently, so that differences in the raw scales cannot be unequivocally interpreted as differences in the underlying skills.
+We believe this is an important finding which deserves a greater degree of attention in the economic literature.
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Literature \label{sec:lit}}
@@
 \citet{Chiteji2010} used the US Panel Study of Income Dynamics (PSID) and found that future orientation and self-efficacy (related to emotional stability) are associated with less alcohol consumption and more exercise.
-\citet{Cobb-Clark2014} used the Australian HILDA data and found that an internal locus of control (also related to emotional stability, perceived control over one's life) is related to better health behaviours (diet, exercise, alcohol consumption anwhad smoking).
+\citet{Cobb-Clark2014} used the Australian HILDA data and found that an internal locus of control (also related to emotional stability, perceived control over one's life) is related to better health behaviours (diet, exercise, alcohol consumption and smoking).
 \citet{Mendolia2014a} used the Longitudinal Study of Young People in England and found that individuals with external locus of control, low self-esteem, and low levels of work ethics, are more likely to engage in risky health behaviours.
-\citet{Savelyev2017} show that the association between personality traits and health behaviours also holds in a high-IQ sample (the Terman Sample).
-\citet{Heckman2016a} use, instead, early risky behaviours to measure socio-emotional traits, and confirm their predictive power for health behaviours and health outcomes.
+\citet{Prevoo2015} construct measures of personality from maternal ratings at 10 and 16 in the British Cohort Study and find that their measure of conscientiousness is positively associated with education and economic outcomes, and negatively associated with body mass index and smoking.
+\citet{Goodman2015} review the interdisciplinary literature and provide a new analysis of the British Cohort Study, including a particular focus on the role of social and emotional skills (defined using a rich set of measurements of the age 10 sweep) in transmitting `top `job' status between parents and their children.
+\citet{Savelyev2019} show that the association between personality traits and health behaviours also holds in a high-IQ sample (the Terman Sample).
+\citet{Heckman2018} use, instead, early risky and reckless behaviours to measure socio-emotional endowments, and confirm their predictive power for education, log wages, smoking and health limits work.
 
-Very few papers attempt to make cross-cohorts comparisons about the importance of socio-emotional skills.
+Few papers attempt to make cross-cohorts comparisons about the importance of socio-emotional skills.
 \citet{Blanden2007} -- one of the closest study to ours -- examine cognitive skills, non‐cognitive traits, educational attainment and labour market attachment as mediators of the decline in inter-generational income mobility in UK between the 1958 and the 1970 cohorts.
 The authors take great care in selecting non-cognitive items to be as comparable as possible across cohorts, from the Rutter scale at age 10 for the 1970 cohort and from the Bristol Social Adjustment Guide for the 1958 cohort; however, they do not carry out a formal test of measurement invariance and the do not construct factor scores fully comparable across cohorts as we do.
-Another paper related to ours is the one by \citet{Reardon2016}, who study recent trends in income, racial, and ethnic school gaps in several dimensions of school readiness, including academic achievement, self-control, and externalizing behavior, at
-kindergarten entry, using comparable data from the Early Childhood Longitudinal Studies (ECLS-K and ECLS-B) for cohorts born from the
-early 1990s to the 2000–2010 period in the US.
+Another paper related to ours is the one by \citet{Reardon2016}, who study recent trends in income, racial, and ethnic school gaps in several dimensions of school readiness, including academic achievement, self-control, and externalizing behavior, at kindergarten entry, using comparable data from the Early Childhood Longitudinal Studies (ECLS-K and ECLS-B) for cohorts born from the early 1990s to the 2000–2010 period in the US.
 They find that readiness gaps narrowed modestly from 1998 to 2010, particularly between high- and low-income students and between White and Hispanic students.
-Lastly, \citet{Deming2017} uses a comparable set of skill measures and covariates across survey waves for the NLSY79 and the NLSY97, and finds that the labour market return to social skills was much greater in the 2000s than in the mid-1980s and 1990s.
+\citet{Landerso2017a} study the sources of differences in social mobility between US and Denmark; for the US, they use the antisocial, headstrong, hyperactivtity subscales from the Behavior Problem Index (BPI) in the Children of the NLSY79 (CNLSY), while for Denmark they use orderliness/organization/neatness grades from the Danish written exams.\footnote{As the authors note (footnote 41) ``Our measures of non-cognitive skills in the two countries are clearly not equivalent.
+The Danish measure of non-cognitive skills is more related to an orderliness/effort measure while the US measure is related to behavioral problems''.} They find that, in both countries, cognitive and non-cognitive skills acquired by age 15 are more important for predicting educational attainment than parental income.
+Lastly, \citet{Deming2017} uses two sets of skill measures and comparable covariates across survey waves for the NLSY79 and the NLSY97,\footnote{He uses the following four variables as measures of social skills in the NLSY79: self-reported sociability in 1981 and at age 6 (retrospective), the number of clubs in which the respondent participated in high school and participation in high school sports; and the following two variables in the NLSY97: two questions that capture the extroversion factor from the Big 5 Personality Inventory (since measures comparable to the NLSY79 are not available in the NLSY97).} and finds that the labour market return to social skills was much greater in the 2000s than in the mid-1980s and 1990s.
+\citet{Zilanawala2019} examine differences in socioemotional and cognitive development among 11-year old children in the UK Millennium Cohort Study and the US Early Childhood Longitudinal Survey-Kindergarten Cohort, and find that family resources explain some cross-national differences, however there appears to be a broader range of family background variables in the UK that influence child development.
+Importantly, none of these papers making comparisons across cohorts or ethnic groups test for measurement invariance like we do.
 
@@
 \citet{Cunha2010} and \citet{Attanasio2018a} both estimate production functions for child cognitive and socio-emotional development (in US and Colombia, respectively), and find an important role played by parental investments.
+\citet{Moroni2019} estimate production functions for child socio-emotional skills (internalising and externalising behaviour) at age 11 in the UK Millennium Cohort Study, and find that the effects of parental inputs which improve the home environment varies as a function both of the level of the inputs themselves and of the development of the child.
 

### 2019-08-25 codes/R/paper/text/paper_jpube_2datamethods.tex
text updates by Gabriella

@@
 This choice is mainly driven by software limitations.
-The \texttt{lavaan} package in \textbf{\textsf{R}} \citep{Rosseel2012} is the most suitable tool for our invariance analysis, but it does not allow to use weights when outcomes are categorical, as it is the case for the socio-emotional measurements.}
-We also restrict our sample to non-multiparous individuals born in England.
-Finally, we restrict the sample to cases where there is complete information on socioemotional skills at five years of age.
+The \texttt{lavaan} package in \textbf{\textsf{R}} \citep{Rosseel2012} is the most suitable tool for our invariance analysis, but it does not allow to use weights when outcomes are categorical, as it is the case for the socio-emotional measurements.} Finally, we also restrict our sample to individuals born in England and to cases where there is complete information on socioemotional skills at five years of age.
 The final sample contains 9,545 individuals from the British Cohort Study, and 5,572 from the Millennium Cohort Study.
@@
 It encompasses more recent advances in child psychopathology, and emphasises positive traits alongside undesirable ones \citep{Stone2010}.
+\citet{Goodman1997} administered both scales to a sample of children, and showed that the scores are highly correlated, and the two measures do not differ in their discriminatory ability.
 The Rutter and SDQ scales are reproduced in \autoref{tab:scales}; they have 23 and 25 items each, respectively.
@@
 
-While the Rutter and SDQ scales are similar in their components, there is no a priori reason to expect them to be directly comparable.
-First, the overlap of behaviours described in the two scales is only partial.
+While the Rutter and SDQ scales are similar in their components (since the latter was developed from the former, see \citet{Goodman1994}), there is no a priori reason to expect them to be directly comparable.
+First, the overlap of behaviours described in the two scales is only partial, given that - by design - the SDQ includes also strengths, in addition to weaknesses.
 Second, the wording of each item is slightly different, both in the description and in the options that can be selected as answers.
@@
 Fourth, and no less importantly, the interpretation of each behaviour by respondents living 30 years apart (1975 vs 2006) might differ due to a host of evolving societal norms.
+Nonetheless, the level of comparibility of the two scales is higher than that of other scales used in comparating work in the literature reviewed in section \ref{sec:lit}.
 
@@
 The wording of the items we will be using in the analysis is presented in \autoref{tab:scalecomp}: we retain 13 items for the BCS (two of them are grouped) and 11 for the MCS with high degree of comparability.
-We exclude from the analysis items that were completely different between the two questionnaires, although we could have included them in the factor analysis and treated them as missing in the cohort were they were not administered.
-While this could have improved efficiency, we decided to rely on a more coherent set of measures to maximise comparability between the two cohorts.
-More details on the derivation of the scale are available in Appendix \autoref{asec:scales}.
+We exclude from the analysis items that were completely different between the two questionnaires to maximise comparability between the two cohorts, as it is standard good practice in the psychometric literature (see for example \citet{Kern2014}).\footnote{Of course, we could have included them in the factor analysis and treated them as missing in the cohort were they were not administered.} More details on the derivation of the scale are available in Appendix \autoref{asec:scales}.
 
@@
 Recent research has shown that there are some benefits to using broader subscales that correspond to the externalising and internalising factors in Rutter, especially in low-risk or general population samples \citep{Goodman2010}.
+Indeed, the internalising and externalising dimensions were introduced in psychology by \citet{Achenbach1966}, who show that they are the two main factors underlying a wide range of psychological measures; as noted in \citet{Achenbach2016}, more than 75,000 articles have been published on internalizing and externalizing problems.
 
-We use exploratory factor analysis (EFA) to assess the factors structure of our 11-item scale combining Rutter and SDQ.\footnote{Factor-analytic methods have long been used in psychology, and in recent years they have become increasingly popular in economics, especially to meaningfully aggregate high-dimensional items measuring different aspects of common underlying dimensions of human development.
+We use exploratory factor analysis (EFA) to assess the factor structure of our new scale, composed of 11 items of the Rutter scale in the BCS and the corresponding items of the SDQ in the MCS.\footnote{Factor-analytic methods have long been used in psychology, and in recent years they have become increasingly popular in economics, especially to meaningfully aggregate high-dimensional items measuring different aspects of common underlying dimensions of human development.
 The EFA is performed decomposing the polychoric correlation matrix of the items and using weighted least squares, and the solution is rescaled using oblique factor rotation (\emph{oblimin}).
@@
 The results are displayed in \autoref{tab:numfac}.
-As pointed out in \citet{Conti2014}, there is relatively little agreement among procedures; this is the case especially for the Rutter items in the BCS data, where different methods suggest to retain between 1 and 3 factors, while most methods suggest to retain 2 factors in the MCS.
-In our analysis, we adopt two factors and a dedicated measurement system, where each measure reflects only one factor.
-This choice is justified both by the child psychology literature cited above, and as compromise to work with the same number of factors in the two cohorts.
+As pointed out in \citet{Conti2014}, there is relatively little agreement among procedures; this is the case especially for the Rutter items in the BCS data, where different methods suggest to retain between 1 and 3 factors, while most methods suggest to retain 2 factors for the SDQ items in the MCS.
+In our analysis, we adopt two factors, which is the average number of factors across all methods for each cohort and gender.
+This choice is also consistent with the child psychology literature cited above, and allows us to work with the same number of factors in the two cohorts.
 
@@
 Appendix \autoref{asec:fullefa} performs the same exploratory steps on the full set of Rutter items in BCS and SDQ items in MCS.
-It confirms that the items we consider in our subscale have a broadly consistent covariance structure even when considered in the context of their original scales.
+It confirms that the items we select for our subscale have a broadly consistent covariance structure even when factor-analysed with the others in their original scales.
 Appendix \autoref{asec:no511} considers the robustness of our results to the exclusion of the items of the scale that perform most poorly.}
@@
 We make the common assumption of a dedicated (or congeneric) factor structure, where each measure is assumed to load on only one latent dimension \citep{Heckman2013,Conti2010a,Attanasio2018a}.
-We mirror the structure found in the exploratory factor analysis (see \autoref{tab:load}), and assume that items 1-6 load exclusively on the externalising factor and items 7-11 on the internalising factor.\footnote{The dedicated factor structure corresponds to a sparse loading matrix, i.e.: $$\bm{\Lambda}_{c} \coloneqq \begin{bmatrix} \lambda_{1c}, \dots, \lambda_{6c} & \bm{0} \\ \bm{0} & \lambda_{7c}, \dots, \lambda_{11c} \end{bmatrix}.$$}
+We mirror the structure found in the exploratory factor analysis (see \autoref{tab:load}), and assume that items 1-6 load exclusively on the externalising factor and items 7-11 on the internalising factor.\footnote{The dedicated factor structure corresponds to a sparse loading matrix, i.e.: $$\bm{\Lambda}_{c} \coloneqq \begin{bmatrix} \lambda_{1c}, \dots, \lambda_{6c} & \bm{0} \\ \bm{0} & \lambda_{7c}, \dots, \lambda_{11c} \end{bmatrix}.$$} This choice of using a dedicated measurement system is consistent with the standard rule of thumb for selecting items based on the magnitude of their loadings, which is indicated as 0.32 by \citet{Tabachnick2001}.
 
@@
 
-Measurement invariance analysis necessarily starts starts from a minimally restrictive model, denominated \emph{configural} model.
+Measurement invariance analysis necessarily starts from a minimally restrictive model, denominated \emph{configural} model.
 This is a `minimum' identifiable model, in that it places the least possible restrictions on how parameters are allowed to vary across cohorts.
@@
   \item For all groups, normalise a reference loading to 1 for each factor
-  \item Set invariant across groups one threshold per item (e.g. $\tau_{0,Ai} = \tau_{0,Bi}$), and an additional threshold in the reference items above
+  \item Set invariant across groups one threshold per item (e.g. $\tau_{0,Ai} = \tau_{0,Bi}$ for two groups $A$ and $B$), and an additional threshold in the reference items above
   \item In the first group: $\bm{\kappa_A}=\bm{0}$, $\text{diag}(\bm{\Sigma}_A)=\bm{I}$
@@
 \ei
-The first two parameterisations (WE{\textDelta} and WE{\textTheta}) normalise the mean and variance of factors to the same constants in both groups, and they leave all loadings and intercepts to be freely estimated; they only differ in whether the additional required normalisation is imposed on the variances of the error terms ($\bm{\Psi}$) or on the diagonal of the covariance matrix of the measures ($\bm{\Sigma}$).
+The first two parameterisations (WE{\textDelta} and WE{\textTheta}) normalise the mean and variance of factors to the same constants in both groups, and they leave all loadings and thresholds to be freely estimated; they only differ in whether the additional required normalisation is imposed on the variances of the error terms ($\bm{\Psi}$) or on the diagonal of the covariance matrix of the measures ($\bm{\Sigma}$).
 The MT parameterisation instead proceeds by identifying parameters in one group first, and then imposing cross-group equality constraints to identify parameters in other groups \citep{Wu2016a}.
@@
 While these indices successfully adjust for model complexity \citep{Cheung2002}, they do not have a known sampling distribution.
-This makes it necessary to rely on simulation studies, which derive rules of thumb indicating what level of \textDelta AFI is compatible with invariances invariance.
+This makes it necessary to rely on simulation studies, which derive rules of thumb indicating what level of \textDelta AFI is compatible with invariance.
 
@@
 \citet{Chen2007} and \citet{Meade2008} -- have updated these thresholds for the continuous case.
-In particular, \citet{Chen2007} suggests the following thresholds for rejecting measurement invariance: \textDelta RMSE $> .015$, \textDelta CFI $< - .010$, \textDelta SRMR $> .010$.
+In particular, \citet{Chen2007} shows in two Monte Carlo studies that the standardised root mean square residual (RMSR) is more sensitive to lack of invariance in factor loadings than in intercepts or residual variances, while the CFI and RMSEA are equally sensitive to all three types of lack of invariance; he suggests the following thresholds for rejecting measurement invariance: \textDelta RMSE $> .015$, \textDelta CFI $< - .010$, \textDelta SRMR $> .010$.
 
-It is not advisable to directly extrapolate rules of thumb derived from simulations with continuous measures to the categorical case \citet{Lubke2004}.
+However, it is not advisable to directly extrapolate rules of thumb derived from simulations with continuous measures to the categorical case \citep{Lubke2004}.
 Recent studies have advanced simulation evidence around the performance of AFIs in measurement invariance analysis with categorical measures.
@@
 The model with restricted thresholds and loadings exhibits a comparable fit to the configural model, according to all the AFIs.
-In particular, the \textDelta AFIs fall within the ranges suggested in \citet{Chen2007} and \citet{Rutkowski2017}.
+In particular, the \textDelta AFIs fall within the ranges suggested in \citet{Chen2007}, \citet{Rutkowski2017} and \citet{Svetina2017}; see also \citet{Svetina2019} for a review of updated guidelines for measurement invariance.
 Invariance of loadings and thresholds across cohorts implies that items in our scale are equally salient in their informational content, and that the latent propensities have equal mapping into the observed items.
@@
 
-This is an important finding, which has to our knowledge never been acknowledged in this literature.
+This is an important finding, which has to our knowledge never been acknowledged in the economic literature.
 How can this lack of comparability be explained?
@@
 In particular, the variance of skills can be compared across cohorts, but mean cohort differences do not necessarily reflect differences in skills.
-We can use scores from our scale to compare children within the same cohort, but not across cohorts.
+We can use scores from our scale to compare children within the same cohort-gender group, but not across cohorts.
 However, we can also compare within-cohort differences between groups of children, across cohorts.

### 2019-08-25 codes/R/paper/text/paper_jpube_3resultsconclusion.tex
text updates by Gabriella

@@
 Given our measurement invariance results in section \ref{sec:measinv}, we stress that the \emph{location} of these scores should not be directly compared across cohorts.
-However, the shape of the distribution can be given a cross-cohort interpretation.\footnote{The density of the scored factors can be contrasted with the distribution of sum scores in \autoref{fig:sumscores}.
-Using raw scores, instead, shows an increase in mass only at the top of the distribution.} It is immediately visible that there is more mass in the tails of the distribution in the 2000 than in the 1970 cohort.
+However, the shape of the distribution can be given a cross-cohort interpretation.
+This result is in sharp contrast with what shown by the simple distribution of sum scores in \ref{fig:sumscores}: using raw scores shows an increase in mass only at the top of the distribution.
+It is immediately visible that there is more mass in the tails of the distribution in the 2000 than in the 1970 cohort.
 
@@
 There is also a significant increase in the gradient by paternal occupation based on social class (\autoref{fig:ineqfsc}), although this is less pronounced if compared to the one based on maternal characteristics.
-In particular, male children with no father figure living in their household have worse skills compared to children with blue collar fathers in the MCS cohort.
+In particular, male children with no father figure living in their household have worse skills (both internalising and externalising) compared to children with blue collar fathers in the MCS cohort.
 Otherwise, skill differences in father's occupation are mostly constant across the two cohorts.\footnote{Figures \ref{fig:ineqips}, \ref{fig:ineqism}, and \ref{fig:ineqifsc} show inequality in the scale items underlying the factor scores used in this section.
 The increase in inequality across cohorts is still present, but less marked when looking at these single items.
-This shows the importance of the factor analysis step in aggregating items, explicitly modelling the measurement error, and testing and accounting for (loadings and thresholds) invariance across the two cohorts.} These patterns are in stark contrast with the findings of \citet{Reardon2016} for the US, who have found a narrowing of the readiness gaps from 1998 to 2010.
+This shows the importance of the factor analysis step in aggregating items, explicitly modelling the measurement error, and testing and accounting for (loadings and thresholds) invariance across the two cohorts.} These patterns are in stark contrast with the findings of \citet{Reardon2016} for the US, who have found a narrowing of the readiness gaps from 1998 to 2010 (however, they have not tested for measurement invariance).
 
@@
 At the same time, the penalty for having a blue-collar father, or not having a father figure at all in the household, has significantly declined across the two cohorts, especially for girls.
-Being born to an unmarried mother, and to a mother who smoked during pregnancy, is associated with a higher penalty for both dimensions of socio-emotional skills in the latter cohort, but only for males.
-This is consistent with recent evidence which shows that family disadvantage disproportionately impedes the pre-market development of boys, in terms of higher disciplinary problems, lower achievement scores, and fewer high-school completions \citep{Autor2016a}.\footnote{It is important to underscore that there has been a significant rise is cohabitation between 1975 and 2006.
+Being born to an unmarried mother, and to a mother who smoked during pregnancy, is associated with a higher penalty for both dimensions of socio-emotional skills in the latter cohort.\footnote{It is important to underscore that there has been a significant rise is cohabitation between 1975 and 2006.
 It is likely that unmarried mothers in the two cohorts have very different characteristics.
-The choice of this indicator is due to the absence of information on cohabitation in the birth survey for the BCS cohort.} Girls of non-white ethnicity, instead, have worse internalising and externalising skills in the MCS, a penalty not suffered by 5-year old non-white girls in the BCS.
+The choice of this indicator is due to the absence of information on cohabitation in the birth survey for the BCS cohort.} Children of non-white ethnicity have worse internalising and externalising skills in the MCS, a penalty almost absent in the BCS (where the prevalence of non-white children was much lower).
 Firstborn boys and girls in the BCS have worse skills, but this difference disappears in the MCS.
-Lastly, we document an increase in the returns to birth weight, which is more pronounced for boys' internalising skills.
+Lastly, we document an increase in the returns to birth weight, which is more pronounced for boys.
 
 These changes in the relative importance of pregnancy factors and family background characteristics for child socio-emotional skills at age 5 need to be interpreted in the light of the significant changes in the prevalence of such characteristics across cohorts.
-As shown in \autoref{tab:sumstats}, the age of the mother at birth, proportion of mothers non-smoking in pregnancy, with post-compulsory education and in employment at the age 5 of the child has substantially increased; at the same time, the proportion of households with no father figure has increased, and so the proportion of women unmarried at birth is much higher in the 2000 than in the 1970 cohort.
-Also, the ethnic structure of the population has changed, with a higher proportion of non-white children in the MCS than in the BCS.
+As shown in \autoref{tab:sumstats}, the age of the mother at birth, and the proportion of mothers non-smoking in pregnancy, with post-compulsory education and in employment at the age 5 of the child has substantially increased; at the same time, the proportion of households with no father figure has increased, and so the proportion of women unmarried at birth is much higher in the 2000 than in the 1970 cohort.
+Also, as noted, the ethnic structure of the population has changed, with a higher proportion of non-white children in the MCS than in the BCS.
 In general, this has been a period of significant societal changes, with an almost continual rise in the proportion of women in employment, an older age at first birth and a rise in dual-earning parents families \citep{Roantree2018}.
@@
 In this last section, we study the predictive power of socio-emotional skills for adolescent and adult outcomes.
-We contribute to a vast interdisciplinary literature by examining medium- and long-term impacts of skills measured at an earlier age than usually examined in previous studies, well before the start of formal education.
-We do so by regressing health and socioeconomic outcomes measured in adolescence and adulthood on the socio-emotional skills scores at age five obtained by our factor model, controlling for the harmonised family background variables at birth and age five (see \autoref{tab:harmvar}).\footnote{In \autoref{tab:bcs_adultapp}, we show that the conclusions in this section are not sensitive to the factor scoring methodology used.
-'Raw' score, obtained by simple summing of the item categories in the 11-item subscale have basically equal predictive power to factor scores.} We present results with and without controlling for cognitive skills.
+We contribute to a vast interdisciplinary literature by examining medium- and long-term impacts of skills measured at an earlier age than usually examined in previous studies, i.e.
+well before the start of formal education.
+We do so by regressing health and socioeconomic outcomes measured in adolescence and adulthood on the socio-emotional skills scores at age five obtained by our factor model, controlling for the harmonised family background variables at birth and age five (see \autoref{tab:harmvar}).\footnote{In tables \ref{tab:adol_sum} and \ref{tab:bcs_adultapp}, we show that the conclusions in this section are not sensitive to the factor scoring methodology used.
+'Raw' scores, obtained by a simple unweighted average of the item categories in the 11-item subscale have basically equal predictive power to factor scores.} We present results with and without controlling for cognitive skills.
 As detailed in \autoref{sec:data}, the available cognitive measures are not comparable across cohorts.
@@
 A child with better internalising skills is less solitary, neurotic, and worried.
-From this perspective, she is likely more sociable and subject to peer influence in health behaviours.
+From this perspective, he/she is likely more sociable and subject to peer influence in health behaviours.
 This is consistent with the evidence in \citet{Goodman2015}, who find a positive association between child emotional health (measured with items for the internalizing behaviour subscale of the Rutter scale at age 10 in the BCS) and smoking at age 42.
@@
 One possible reason might be the relatively young age at which these skills are measured.
-Results are available upon request.} This is in line with the evidence in \citet{Conti2010}, who show that not accounting for non-cognitive traits (a self-regulation factor measured at age 10) overestimates the importance of cognition for predicting health and health behaviours, using data from the British cohort study.
-\citet{Conti2013a} use rich data on child personality and socio-emotional traits collected at ages 7, 11 and 16 in the 1958 British birth cohort,\footnote{They use the Rutter scale and the Bristol Social Adjustment Guide.} and show that these traits rival the importance of cognition in explaining the education gradient in health behaviours (including smoking and BMI).
+Results are available upon request.} This is in line with the evidence in \citet{Conti2010}, who show that not accounting for non-cognitive traits (in their paper, a self-regulation factor measured at age 10) overestimates the importance of cognition for predicting health and health behaviours, using data from the British cohort study.
+Along the same lines, \citet{Conti2013a} use rich data on child personality and socio-emotional traits collected at ages 7, 11 and 16 in the 1958 British birth cohort,\footnote{They use the Rutter scale and the Bristol Social Adjustment Guide.} and show that these traits rival the importance of cognition in explaining the education gradient in health behaviours (including smoking and BMI).
 We show that child socio-emotional skills have greater predictive power for health outcomes and behaviours even when measured at an earlier age.
@@
 The structure of \autoref{tab:bcs_adult} is similar to \autoref{tab:adol}, but it considers educational achievement, employment, and earnings (conditional on being in paid employment) for the BCS cohort members.
-For these outcomes, the predictive power of cognitive skills outweighs that of socio-emotional skills, whose predictive power diminishes over time (between the ages 34 and 42), and is driven to insignificance after controlling for cognition.
+For these outcomes, the predictive power of cognitive skills outweighs that of socio-emotional skills, which are only predictive of educational attainment, and whose predictive power for males is driven to insignificance after controlling for cognition.
 This is consistent with the evidence in  \citet{Conti2011}, who show that cognitive endowments at age 10 are more predictive (than socio-emotional and health ones) for employment and wage outcomes in the BCS.
 Again, we show that the greater predictive power of cognition for socioeconomic outcomes holds even when considering earlier-life measures of child development.
-Similarly, \citet{Goodman2015} use data from the same cohort, but exploit a richer set of measures from the age 10 survey.
-They show that social and emotional skills measured at age 10 are strongly related to a wide range of adult outcomes.
 
@@
 The increase in cross-cohort inequality is even starker when comparing children born to mothers who smoked during pregnancy.
-In both cases, the increase in inequality is particularly pronounced for boys.
-On the other hand, the skills penalty arising from the lack of a father figure in the household has substantially declined, especially for girls.
+On the other hand, the skills penalty arising from the lack of a father figure in the household has substantially declined.
 

### 2019-08-25 codes/R/paper/text/paper_jpube_4appendix.tex
text updates by Gabriella

@@
 The Rutter items are rated on three levels: `Does not apply', `Somewhat applies', `Certainly applies'.
-Since they are all behaviours indicating lower skills, we encode all of them in reverse, i.e.
+Since they are all behaviours indicating lower skills, we recode all of them in reverse, i.e.
 `Certainly applies' = 0, `Somewhat applies' = 1, `Does not apply' = 2.
@@
 
-In MCS, we use the 25-item strengths and difficulties questionnaire \citep{Goodman1997} -- see Panel B of \autoref{tab:scales}.
+In MCS, we use the 25-item Strengths and Difficulties Questionnaire \citep{Goodman1997} -- see Panel B of \autoref{tab:scales}.
 All items are recorded on a 4-point scale: `Not true', `Somewhat true', `Certainly true', `Can't say'.
@@
 `Certainly true' = 0, `Somewhat true' = 1, `Not true' = 2.
-For comparability with the BCS Rutter scale, we dichotomise items 3 and 5 to make them comparable with , and dichotomise and invert items 7, and 14.
+For comparability with the BCS Rutter scale, we dichotomise items 3 and 5, and dichotomise and invert items 7, and 14.
 
@@
 
-In fact, exploratory factor analysis (on both on the full Rutter and SDQ scales and on the 11-item subscale) shows that these items, numbered 5 and 11, perform poorly and exhibit relatively low factor loadings.
+In fact, exploratory factor analysis (on both the full Rutter and SDQ scales and on the 11-item subscale) shows that these items, numbered 5 and 11, perform poorly and exhibit relatively low factor loadings.
 As a robustness check, we replicate the main results of the paper by excluding them from the subscale.

### 2019-08-25 codes/R/paper/text/paper_jpube_5apptables.tex
text updates by Gabriella

@@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{sidewaystable}[ht!]
+\caption{Description of harmonised variables}\label{tab:harmvar}
+\centering\setstretch{1.2} \scriptsize
+\begin{tabular}{L{.18\textwidth}C{.1\textwidth}L{.3\textwidth}L{.75\textwidth}}
+\toprule
+Variable Group & Age & Variable & Note \\
+\midrule
+Maternal education & 5 & Post-compulsory schooling$^d$ & Whether mother continued schooling past the compulsory age, based on her year of birth.
+School leaving age in England was changed from 14 to 15 in 1947 and from 15 to 16 in 1972.
+\\ \\[-1.5em]
+Maternal employment & 5 & Employed$^d$ & Includes full time and part time \\ \\[-1.5em]
+Father occupation & 5 & White collar (I-IIINM)$^d$ \newline Blue collar (IIIM-V-other)$^d$ \newline No father figure$^d$ & Based on father's Registrar General Social Class classification of occupations.
+White collar includes I (Professional), II (Managerial/technical), IIINM (Skilled non-manual).
+Blue collar includes IIIM (Skilled manual), IV (Partly skilled), V (Unskilled), Other, Unemployed, and Armed forces.
+No father figure is a dummy for children whose father does not live in the same household.
+Father's social class was recorded using the SOC2000 classification in MCS.
+We use the derivation matrices kindly provided by David Pevalin at ISER (available at https://www.iser.essex.ac.uk/archives/nssec/derivations-of-social-class) to map SOC2000 into Registrar General Social Class.
+\\ \\[-1.5em]
+Maternal background & 0/5 & Mother's age at birth \newline Mother's height (cm) \newline Mother unmarried at birth$^d$ \newline Child nonwhite ethnicity$^d$  \newline Number of children in HH & All variables are self-reported by the mother at birth, except for number of children in household (at five years old).
+Unmarried is only based on marital status, and includes cohabitation.
+\\ \\[-1.5em]
+Pregnancy & 0 & Child is firstborn$^d$ \newline Number of previous stillbirths \newline Mother smoked in pregnancy$^d$ \newline Preterm birth (under 37 weeks gestation)$^d$ \newline (log) birth weight (kg) & Parity, stillbirths, and smoking are self-reported by the mother.
+Gestational length and birth weight are from hospital records.
+\\ \\[-1.5em]
+Cognitive skills & 5 & & Based on test batteries administered to the cohort member at five.
+Three tests are used for BCS children: Copy Designs (child is asked to copy simple designs adjacently), Human Figure Drawing (child draws an entire human figure), English Picture Vocabulary Test (child identifies the picture referring to a word among four pictures).
+Three tests are used in the MCS: BAS Naming Vocabulary (child is shown a series of pictures and asked to name it), BAS Picture Similarity (child is shown a row of 4 pictures on a page and places a card with a fifth picture under the one most similar to it), BAS Pattern Construction (child constructs a design by putting together flat squares or solid cubes with patterns on each side).
+ \\ \\[-1.5em]
+Adolescent outcomes & 16 (BCS) \newline 14 (MCS) & Child tried smoking$^d$ \newline Body Mass Index (BMI) & Smoking is self reported by the child.
+Height and weight are taken as part of a medical examination.
+\\ \\[-1.5em]
+Adult outcomes (BCS only) & 34 \newline 42 \newline 34, 42 & Higher education$^d$ \newline Employed$^d$ \newline (log) gross weekly pay & Higher education is defined as having a university degree or its vocational equivalent.
+It corresponds to level 4 or 5 in the National Vocational Qualification (NVQ) equivalence.
+Employed is a dummy for being in paid employment or self-employment, either full or part time.
+Gross weekly pay is weekly pre-tax pay from the respondent's main activity, conditional on being a paid employee.
+\\
+\bottomrule
+\end{tabular}
+\justify {\scriptsize \emph{Notes}: Variables denoted by $^d$ are binary or categorical.
+\par}
+\end{sidewaystable}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{landscape}
-\begin{table}[ht!]
-\caption{Description of harmonised variables}\label{tab:harmvar}
-\centering\setstretch{1.2} \scriptsize
-\begin{tabular}{L{.18\textwidth}C{.1\textwidth}L{.3\textwidth}L{.75\textwidth}}
-\toprule
-Variable Group & Age & Variable & Note \\
-\midrule
-Maternal education & 5 & Post-compulsory schooling$^d$ & Whether mother continued schooling past the compulsory age, based on her year of birth.
-School leaving age in England was changed from 14 to 15 in 1947 and from 15 to 16 in 1972.
-\\ \\[-1.5em]
-Maternal employment & 5 & Employed$^d$ & Includes full time and part time \\ \\[-1.5em]
-Father occupation & 5 & White collar (I-IIINM)$^d$ \newline Blue collar (IIIM-V-other)$^d$ \newline No father figure$^d$ & Based on father's Registrar General Social Class classification of occupations.
-White collar includes I (Professional), II (Managerial/technical), IIINM (Skilled non-manual).
-Blue collar includes IIIM (Skilled manual), IV (Partly skilled), V (Unskilled), Other, Unemployed, and Armed forces.
-No father figure is a dummy for children whose father does not live in the same household.
-Father's social class was recorded using the SOC2000 classification in MCS.
-We use the derivation matrices kindly provided by David Pevalin at ISER (available at https://www.iser.essex.ac.uk/archives/nssec/derivations-of-social-class) to map SOC2000 into Registrar General Social Class.
-\\ \\[-1.5em]
-Maternal background & 0/5 & Mother's age at birth \newline Mother's height (cm) \newline Mother unmarried at birth$^d$ \newline Child nonwhite ethnicity$^d$  \newline Number of children in HH & All variables are self-reported by the mother at birth, except for number of children in household (at five years old).
-Unmarried is only based on marital status, and includes cohabitation.
-\\ \\[-1.5em]
-Pregnancy & 0 & Child is firstborn$^d$ \newline Number of previous stillbirths \newline Mother smoked in pregnancy$^d$ \newline Preterm birth (under 37 weeks gestation)$^d$ \newline (log) birth weight (kg) & Parity, stillbirths, and smoking are self-reported by the mother.
-Gestational length and birth weight are from hospital records.
-\\ \\[-1.5em]
-Cognitive skills & 5 & & Based on test batteries administered to the cohort member at five.
-Three tests are used for BCS children: Copy Designs child is asked to copy simple designs adjacently), Human Figure Drawing (child draws an entire human figure), English Picture Vocabulary Test (child identifies the picture referring to a word among four pictures).
-Three \emph{different} tests are used in the MCS: BAS Naming Vocabulary (child is shown a series of pictures and asked to name it), BAS Picture Similarity (child is shown a row of 4 pictures on a page and places a card with a fifth picture under the one most similar to it), BAS Pattern Construction (child constructs a design by putting together flat squares or solid cubes with patterns on each side).
-
- \\ \\[-1.5em]
-Adolescent outcomes & 16 (BCS) \newline 14 (MCS) & Child tried smoking$^d$ \newline Body Mass Index (BMI) & Smoking is self reported by the child.
-Height and weight are taken as part of a medical examination.
-\\ \\[-1.5em]
-Adult outcomes (BCS only) & 34 \newline 42 \newline 34, 42 & Higher education$^d$ \newline Employed$^d$ \newline (log) gross weekly pay & Higher education is defined on having a university degree or its vocational equivalent.
-It corresponds to level 4 or 5 in the National Vocational Qualification (NVQ) equivalence.
-Employed is a dummy for being in paid employment or self-employment, either full or part time.
-Gross weekly pay is weekly pre-tax pay from the respondent's main activity, conditional on being a paid employee.
-\\
-\bottomrule
-\end{tabular}
-\justify {\scriptsize \emph{Notes}: Variables denoted by $^d$ are binary or categorical.
-\par}
-\end{table}
-\end{landscape}

### 2019-09-17 codes/R/paper/seskills_jpube_rev.Rnw
finalise figures and include in manuscript

@@
 read_chunk(here("codes/R/8_regs_outc.R"))
+read_chunk(here("codes/R/9_rif.R"))
 read_chunk(here("codes/R/99_auxfun.R"))
@@
 
+<<RIF_PLOTS_PREP, dependson="SELECTDATA">>=
+# make RIF decomposition plots
+@
+
 % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 @
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+\begin{figure}
+<<RIF_PLOTS, dependson="RIF_PLOTS_PREP", dev='tikz', fig.height=8, fig.width=10, out.width='\\linewidth'>>=
+gridExtra::grid.arrange(rif_plots[[1]],rif_plots[[2]], ncol=1, nrow=2)
+@
+\caption{RIF decomposition of quantile differences}\label{fig:rif}
+\justify {\scriptsize \emph{Notes}: 
+\par}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2019-09-17 codes/R/9_rif.R

@@ -54,6 +54,12 @@ rif_beta_all$component <- factor(
   )
 levels(rif_beta_all$component) <- c("Total Diff.", "Composition", "Specif. Error", "Coefficient", "Reweight. Error")
 
+# common graph options
+bptheme <- theme(axis.title.x = element_blank(),
+                 axis.text.x = element_blank(), axis.line.x = element_blank(),
+                 axis.ticks=element_blank(),
+                 strip.background=element_rect(fill="white"),
+                 panel.spacing = unit(1, "lines"))
 
 # Normal BAR PLOT, by gender
 palette_5 <- c("#d95f02", "#7570b3", "#b5b1e6", "#1b9e77", "#99c9bc")
@@ -66,21 +72,14 @@ for (g in c("Males", "Females")) {
     geom_bar(stat = "identity", position = "dodge") +
     ggtitle(g) +
     geom_hline(yintercept = 0) +
-    ylab("Difference") +
+    ylab("Difference") + ylim(-.2, .4) +
     facet_grid(~ skill + iq) +
     scale_fill_manual("Component", values = palette_5) + 
     bptheme
 }
 
-
 ## ---- RIF_PLOTS_PC
 # Normal BAR PLOT, by gender, percentages
-bptheme <- theme(axis.title.x = element_blank(),
-                 axis.text.x = element_blank(), axis.line.x = element_blank(),
-                 axis.ticks=element_blank(),
-                 strip.background=element_rect(fill="white"),
-                 panel.spacing = unit(1, "lines"))
-
 palette_4 <- c("#7570b3", "#b5b1e6", "#1b9e77", "#99c9bc")
 rif_plots_pc <- list()
 for (g in c("Males", "Females")) {

### 2019-09-18 codes/R/paper/seskills_jpube_rev.Rnw
today's date
ERC acknowledgement

@@
 \affil{\emph{Institute for Fiscal Studies} \\ \emph{Department of Economics, University College London}}
-\date{December 2018}
+\date{\today}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 We would like to thank all cohort members and their families, who generously gave their time to participate in the surveys.
-This project has received funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme (grant agreement No 695300 - HKADeC - ERC-2015-AdG/ERC-2015-AdG.
-We also thank the DynaHEALTH consortium (H2020–633595 DynaHEALTH) for funding and support.
-GM is funded by an Economics and Social Research Council scholarship.}
+This project has received funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme (grant agreement No 695300 - HKADeC - ERC-2015-AdG/ERC-2015-AdG and grant agreement No 819752 - DEVORHBIOSHIP - ERC-2018COG).
+We also thank the DynaHEALTH consortium (H2020 - 633595 DynaHEALTH) for funding and support.
+GM is funded by an Economic and Social Research Council scholarship.}
 \end{titlepage}

### 2020-02-01 codes/R/paper/seskills_jpube_rev.Rnw
add multiple factor options EFA

@@
 <<NUMFAC_ALL, dependson='MCSEFA'>>=
+@
+
+<<BCSEFA_ADD, dependson='BCSEFA'>>=
+@
+
+<<MCSEFA_ADD, dependson='MCSEFA'>>=
 @
@@
 
+\begin{landscape}
+\begin{table}
+\caption{Loadings from exploratory factor analysis with 1 or 3 factors -- 11-item scale}\label{tab:load_add}
+\centering\setstretch{1.2}\footnotesize
+<<EFALOADTAB_ADD, dependson='MCSEFA_ADD', results='asis'>>=
+# combine results
+library(kableExtra)
+loadtab <- cbind(meantab$num, meantab$title, 
+                 data.frame(cbind(
+                   bcsfa1.m$loadings,
+                   bcsfa1.f$loadings,
+                   bcsfa3.f$loadings,
+                   mcsfa1.m$loadings,
+                   mcsfa1.f$loadings,
+                   mcsfa3.f$loadings
+                 ))
+                 )
+colnames(loadtab) <- c("item", "title", 
+                       "bcs1m", "bcs1f", "bcs3f1", "bcs3f2", "bcs3f3",
+                       "mcs1m", "mcs1f", "mcs3f1", "mcs3f2", "mcs3f3"
+)
+
+loadtab %>%
+  mutate_at(vars(matches("cs")), round, 3) %>%
+  mutate_at(
+    vars(matches("cs")),
+    function(x) cell_spec(x, "latex", bold = ifelse(abs(x) > .3, TRUE, FALSE))
+    ) %>%
+  kable("latex", escape = F, booktabs = T, linesep = "", align=c('c','l','c','c','c','c','c','c','c','c','c','c'),
+        col.names = c("Item", "Title", 
+                      "Males", "Females", "Females Fac. 1", "Females Fac. 2", "Females Fac. 3",
+                      "Males", "Females", "Females Fac. 1", "Females Fac. 2", "Females Fac. 3"
+                      )) %>%
+  column_spec(3:12, width = "5em") %>%
+  add_header_above(c(" " = 2,  
+                     "BCS (1970) - 1 Fac." = 2, "BCS (1970) - 3 Fac." = 3, 
+                     "MCS (2000/1) - 1 Fac." = 2, "MCS (2000/1) - 3 Fac." = 3))
+
+@
+\justify {\scriptsize \emph{Notes}: The table displays the factor loadings obtained from exploratory factor analysis (EFA) on our novel scale, separately by cohort.
+The EFA is performed decomposing the polychoric correlation matrix of the items and using weighted least squares, and the solution is rescaled using oblique factor rotation (\emph{oblimin}).
+We use the \textbf{\textsf{R}} package \texttt{psych}, version 1.8.4 \citep{Revelle2018}.
+\par}
+\end{table}
+\end{landscape}
+
 \begin{table}

### 2020-02-01 codes/R/2_efa.R

@@ -68,7 +68,6 @@ nmcs.f <- dim(items.mcs.f)[1]
 mcscor.f <- hetcor(items.mcs.f, ML=TRUE, std.err = F)$correlations
 mcsfa.f <- fa(r=mcscor.f, nfactors=2, rotate = rotation, fm="wls")
 
-
 ## ---- MCSEFARES
 # VSS for number of factors
 ns.mcs <- nScree(mcscor)$Components
@@ -99,4 +98,50 @@ mr.mcs.ext <- fisherz2r(mean(fz.mcs.ext, na.rm=T))
 fz.mcs.int <- fz.mcs[7:11,7:11]
 mr.mcs.int <- fisherz2r(mean(fz.mcs.int, na.rm=T))
 
+
+## ---- BCSEFA_ADD
+# use only 1 factor
+bcsfa1 <- fa(r=bcscor, nfactors=1, rotate = rotation, fm="wls")
+bcsfa1.m <- fa(r=bcscor.m, nfactors=1, rotate = rotation, fm="wls")
+bcsfa1.f <- fa(r=bcscor.f, nfactors=1, rotate = rotation, fm="wls")
+# 3 factors for females
+bcsfa3.f <- fa(r=bcscor.f, nfactors=3, rotate = rotation, fm="wls")
+
+## ---- MCSEFA_ADD
+# use only 1 factor
+mcsfa1 <- fa(r=mcscor, nfactors=1, rotate = rotation, fm="wls")
+mcsfa1.m <- fa(r=mcscor.m, nfactors=1, rotate = rotation, fm="wls")
+mcsfa1.f <- fa(r=mcscor.f, nfactors=1, rotate = rotation, fm="wls")
+# 3 factors for females
+mcsfa3.f <- fa(r=mcscor.f, nfactors=3, rotate = rotation, fm="wls")
+
+loadtab <- cbind(meantab$num, meantab$title, 
+                 data.frame(cbind(
+                   bcsfa1.m$loadings,
+                   bcsfa1.f$loadings,
+                   bcsfa3.f$loadings,
+                   mcsfa1.m$loadings,
+                   mcsfa1.f$loadings,
+                   mcsfa3.f$loadings
+                 ))
+                 )
+colnames(loadtab) <- c("item", "title", 
+                       "bcs1m", "bcs1f", "bcs3f1", "bcs3f2", "bcs3f3",
[…]

### 2020-02-03 codes/R/paper/seskills_jpube_rev.Rnw
add CFA with single factor

@@
 # robustness checks without items 5-11
+@
+
+<<FA_INV_AUX_1FAC>>=
+# robustness checks with 1 factor
 @
@@
 
+\begin{landscape}
+\begin{table}[ht]
+\caption{Measurement invariance fit comparison -- single factor}\label{tab:fit_1f}
+\centering\setstretch{1.2} \scriptsize
+<<FA_FIT_RES_1F, dependson="FA_FIT", results='asis'>>=
+require(xtable)
+
+# MAKE FIT TABLES
+fit_1f <- afitab(list(fa.c.1f, fa.tl.1f, fa.tli.1f), 
+                 c("Configural", 
+                   "Threshold + Loading Inv", 
+                   "Thr. + Load. + Intercept Inv")
+)
+
+print(xtable(fit_1f,
+                    align="L{0cm}L{.22\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
+                    digits = 4
+                    ),
+             size="\\footnotesize",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0)),
+                              command = c(
+                              # header
+            "& & \\multicolumn{6}{c}{\\textbf{Absolute fit}} & \\multicolumn{6}{c}{\\textbf{Relative fit}} 
+                  \\\\ \\cmidrule(lr){3-8} \\cmidrule(lr){9-14}
+            Model & Num. par. & $\\chi^2$ & RMSE & SRMR & MFI & CFI & G-hat & $\\chi^2$ $p$ & \\textDelta RMSE & \\textDelta SRMR &  \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ 
+& (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8) & (9) & (10) & (11) & (12) & (13) \\\\ \\midrule"
+                              )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = "\\textwidth",
+             floating = FALSE
+)
+
+@
+\justify {\scriptsize \emph{Notes}: The table presents fit indices for models of different invariance levels, following \cite{Wu2016a} Col.
+(1) displays the number of estimated parameters for each model.
+Col.
+(2) and (8) present the value of the $\chi^2$ statistic and the pvalue of the test of equality with respect to the configural model.
+Col.
+(3)-(7) and (9)-(13) present alternative fit indices (AFIs), in absolute values and differences from the configural model respectively.
+RMSE = Root mean squared error of approximation; SRMR = standardised root mean residual; MFI = McDonald non-centrality index; CFI = comparative fit index; G-hat = gamma-hat.
+\par}
+\end{table}
+\end{landscape}
+
 \clearpage

### 2020-02-03 codes/R/3_inv_aux.R

@@ -1,3 +1,12 @@
+## ---- FA_INV_AUX_1FAC
+# robustness checks with a single factor
+c.th.4.1f <- glue::glue_collapse(readLines(paste0(dir_syntax, "c_th_4_1f.lav")), sep=" \n  ")
+tl.th.4.1f <- glue::glue_collapse(readLines(paste0(dir_syntax, "tl_th_4_1f.lav")), sep=" \n  ")
+tli.th.4.1f <- glue::glue_collapse(readLines(paste0(dir_syntax, "tli_th_4_1f.lav")), sep=" \n  ")
+
+fa.c.1f   <- cfa(c.th.4.1f, data = fadata[[1]], group = "cohortsex", estimator="wlsmv", parameterization="theta")
+fa.tl.1f   <- cfa(tl.th.4.1f, data = fadata[[1]], group = "cohortsex", estimator="wlsmv", parameterization="theta",group.equal='loadings')
+fa.tli.1f <- cfa(tli.th.4.1f, data = fadata[[1]], group = "cohortsex", estimator="wlsmv", parameterization="theta", group.equal='loadings')
 
 ## ---- FA_INV_AUX_SYN
 # syntax for robustness checks

### 2020-02-11 codes/R/paper/text/paper_jpube_1introlit.tex
Gabriella's text updates

@@
 \section{Introduction}
+
+[\textbf{ADD REFS}]
 
@@
 There is growing consensus on the fact that human capital is a multidimensional object, with different domains playing different roles in labour market as well as in the determination of other outcomes, including the process of human development.
-It is also recognised that human capital is the output of a very persistent process, where early years inputs play an important and persistent role.
+It is also recognised that human capital is the output of a very persistent process, where early years inputs play an important and longlasting role.
 
@@
 
-In this paper, we focus on an important dimension of human capital, which, so far, has received limited attention: socio-emotional skills.
-Evidence has shown that gaps in socio-emotional skills emerge at very young ages, and that in the absence of interventions are very persistent across the life cycle \citep{Cunha2006}.
+In this paper, we focus on an important dimension of human capital, which has been receiving increasing attention in the last few years: socio-emotional skills.
+It has been shown that gaps in socio-emotional skills emerge at very young ages, and that in the absence of interventions are very persistent across the life cycle \citep{Cunha2006}.
 However, there is surprisingly little evidence on how inequality in this important dimension of human capital has changed across cohorts.
@@
 We consider the measurement of socio-emotional skills during early childhood, as these skills have been shown, in a variety of contexts \citep{Almlund2011} to have important long-run effects.
-Our goal is to characterise the distribution of socio-emotional skills in these cohorts and compare them.
+Our goal is to characterise the distributions of socio-emotional skills in these cohorts and compare them.
 In the last part of the paper, we also consider the predictive power of different socio-emotional skills for health and socioeconomic outcomes.
 
-The main contributions of the paper are three.
-First, we construct a novel scale of of childhood behavioural traits from two validated instruments and assess its comparability across cohorts.
+We proceed in four steps.
+First, we construct a novel scale of childhood behavioural traits from two validated instruments and assess its comparability across cohorts.
 By performing an exploratory factor analysis, we determine that two dimensions are a parsimonious representation of socio-emotional skills for both cohorts.
-Coherently with previous literature, we label them as `internalising' and `externalising' skills, the former relating to the ability of children to focus their drive and determination and the latter relating to their ability in engaging in interpersonal activities.
+Coherently with previous literature, we label them as `internalising' and `externalising' skills, the former relating to the ability of children to focus their drive and determination, and the latter relating to their ability to engage in interpersonal activities.
 Importantly, for the first time in economics, we study the comparability of the measures in the two cohorts.
@@
 However, we strongly reject measurement invariance for the level parameters.
-These results imply that while the level of inequality across the two cohorts in the skills we consider is comparable, we cannot determine whether the \emph{average levels} of the two factors are larger or smaller in one of the two cohorts.
-While this result hinders a comparison in the level of skills, it is of interest per se to find that mothers of English children born in England thirty years apart assess behaviours differently, so that differences in the raw scales cannot be unequivocally interpreted as differences in the underlying skills.
+These results imply that while we can compare the inequality in skills across the two cohorts, we cannot determine whether the \emph{average levels} of the two factors are larger or smaller in one of the two cohorts.
+While this result hinders a comparison in the level of skills, it is of interest per se to find that mothers of children born in England thirty years apart assess behaviours differently, so that differences in the raw scales cannot be unequivocally interpreted as differences in the underlying skills.
 We believe this is an important finding which deserves a greater degree of attention in the economic literature.
 
-Third, given the results we obtain on measurement invariance, we proceed to compare the level of inequality in the two types of socio-emotional skills across the two cohorts, for both boys and girls.
+Second, given the results we obtain on measurement invariance, we proceed to compare the inequality in the two types of socio-emotional skills across the two cohorts, for both boys and girls.
 We find that the most recent cohort is more unequal in both dimensions of socio-emotional skills than the 1970 cohort.
 This result is particularly apparent for boys, and when looking at differences by maternal background.
+Third, we formally decompose the increase in inequality in skills into changes in the composition of maternal characteristics and changes in the returns to those characteristics, using recently developed methods based on Recentered Influence Function (RIF) regressions.
+In doing so, we provide the first application of this method to the child development literature.
+
 Fourth, we study whether the socio-emotional skills we observe at a young age are an important determinant of a variety of adolescent (and adult, for the older BCS cohort) outcomes.
@@
 
+Our key contribution in this paper is to bring together two important strands of the literature: on the one hand, the literature on child development and early interventions; on the other hand, the literature on the measurement and the evolution of different types of inequality.
+While the former literature has provided robust evidence on the long-term impacts of a variety of early life circumstances, it has not systematically focused on describing and disentangling the sources of inequality in early human development; at the same time, the latter literature has carefully studied measures such as income, wages and wealth, overlooking other important - yet harder to measure - dimensions.
+In bridging these two literatures, we also apply recent methodological advances in factor analysis and show the importance of testing and constructing comparable aggregates.
+The methodology that we apply in this paper is likely to be relevant in many other settings, for example when measuring trends in inequality in other dimensions (such as satisfaction, mental health or well-being) whose measurement might have changed over time.
+Lastly, it is worth emphasizing that, while learning about the evolution and the determinants of inequality in socio-emotional skills is an interesting exercise in its own right, the ultimate goal of such research would be to uncover how much inequalities in early human development contribute to income or wealth inequality later in life.
+The present paper constitutes a first step towards such an endeavour.
+
 The rest of the paper is organised as follows.
-In section \ref{sec:data}, we briefly discuss the data we use in the analysis.
-In section \ref{sec:methods}, we discuss the methods we use to identify the number of dimensions in socio-emotional skills and how we estimate the latent factors that represent them.
+We start in section \ref{sec:lit} by reviewing the main literature on determinants and consequences of socio-emotional traits.
+In section \ref{sec:data}, we briefly introduce the data we use in the analysis.
+In section \ref{sec:methods}, we present the methods we use to identify the number of dimensions in socio-emotional skills and how we estimate the latent factors that represent them.
 In section \ref{sec:measinv}, we discuss the comparability of factors estimated with a given set of measures from different groups and the \emph{measurement invariance} tests we use.
-Section \ref{sec:results} reports our empirical results, while section \ref{sec:conclusions} concludes the paper.
+Section \ref{sec:results} reports our empirical results on changes in inequality in socio-emotional skills and their predictive power for later outcomes.
+Section \ref{sec:conclusions} concludes the paper.
 
@@
 The importance of cognition in predicting life course success is well established in the economics literature.
-However, in recent years the role played by `non-cognitive' traits is being increasingly investigated.
+However, in recent years the role played by `non-cognitive' traits has been increasingly investigated.
 These traits include constructs as different as psychological and preference parameters such as social and emotional skills, locus of control and self-esteem, personality traits (e.g.
@@
 
-\paragraph{Consequences of socio-emotional traits} One of the first papers to show the importance of non-cognitive personality variables for wages was \cite{Bowles2001}.
-\cite{Heckman2006} suggested that non-cognitive skills are at least as important as cognitive abilities in determining a variety of adults outcomes.
+\paragraph{Consequences of socio-emotional traits} One of the first papers to pioneer the importance of `non-cognitive' variables for wages is \cite{Bowles2001}.
+\cite{Heckman2006} suggest that non-cognitive skills are at least as important as cognitive abilities in determining a variety of adults outcomes.
 \citet{Lindqvist2011}, using data based on personal interviews conducted by a psychologist during the Swedish military enlistment exam, show that both cognitive and noncognitive abilities are important in the labour market, but for different outcomes: low noncognitive abilities are more correlated with unemployment or low earnings, while cognitive ability is a stronger predictor of wages for skilled workers.
 \citet{Segal2013}, using data on young men from the US National Education Longitudinal Survey, shows that eight-grade misbehaviour is important for earnings over and above eight-grade test scores.
-\citet{Layard2014} find that childhood emotional health (operationalised using the same mother-reported Rutter scale we use in the 1970 British cohort study) at ages 5, 10 and 16 is the most important predictor of adulthood life satisfaction and life course success.
+\citet{Layard2014} find that childhood emotional health (operationalised using the same mother-reported Rutter scale we use in the 1970 British cohort study) at ages 5, 10 and 16 is the most important predictor of adult life satisfaction and life course success.
 
-There are only few studies in economics specifically studying ``non-cognitive'' traits and health behaviors.
-\citet{Conti2010a} and \citet{Conti2011} were the first to consider three early endowments, including child socio-emotional traits and health in addition to cognition, using rich data from the 1970 British cohort study.
-They find strong evidence that non-cognitive traits promote health outcomes and healthy behaviors, and than not accounting for them overestimates the effects of cognition; additionally, they document that child cognitive traits are more important predictors of employment and wages than socio-emotional traits or early health.
-\citet{Chiteji2010} used the US Panel Study of Income Dynamics (PSID) and found that future orientation and self-efficacy (related to emotional stability) are associated with less alcohol consumption and more exercise.
-\citet{Cobb-Clark2014} used the Australian HILDA data and found that an internal locus of control (also related to emotional stability, perceived control over one's life) is related to better health behaviours (diet, exercise, alcohol consumption and smoking).
-\citet{Mendolia2014a} used the Longitudinal Study of Young People in England and found that individuals with external locus of control, low self-esteem, and low levels of work ethics, are more likely to engage in risky health behaviours.
+There are only few studies in economics specifically studying ``non-cognitive'' traits and health behaviours.
+\citet{Conti2010a} and \citet{Conti2011} are the first to consider three early endowments, including child socio-emotional traits and health in addition to cognition, using rich data from the 1970 British cohort study.
+They find strong evidence that non-cognitive traits promote health and healthy behaviours, and than not accounting for them overestimates the effects of cognition; additionally, they document that child cognitive traits are more important predictors of employment and wages than socio-emotional traits or early health.
+\citet{Chiteji2010} uses the US Panel Study of Income Dynamics (PSID) and finds that future orientation and self-efficacy (related to emotional stability) are associated with less alcohol consumption and more exercise.
+\citet{Cobb-Clark2014} use the Australian HILDA data and find that an internal locus of control (also related to emotional stability, perceived control over one's life) is related to better health behaviours (diet, exercise, alcohol consumption and smoking).
+\citet{Mendolia2014a} use the Longitudinal Study of Young People in England and find that individuals with external locus of control, low self-esteem, and low levels of work ethics, are more likely to engage in risky health behaviours.
 \citet{Prevoo2015} construct measures of personality from maternal ratings at 10 and 16 in the British Cohort Study and find that their measure of conscientiousness is positively associated with education and economic outcomes, and negatively associated with body mass index and smoking.
@@
 \citet{Blanden2007} -- one of the closest study to ours -- examine cognitive skills, non‐cognitive traits, educational attainment and labour market attachment as mediators of the decline in inter-generational income mobility in UK between the 1958 and the 1970 cohorts.
-The authors take great care in selecting non-cognitive items to be as comparable as possible across cohorts, from the Rutter scale at age 10 for the 1970 cohort and from the Bristol Social Adjustment Guide for the 1958 cohort; however, they do not carry out a formal test of measurement invariance and the do not construct factor scores fully comparable across cohorts as we do.
+The authors take great care in selecting non-cognitive items to be as comparable as possible across cohorts, from the Rutter scale at age 10 for the 1970 cohort and from the Bristol Social Adjustment Guide for the 1958 cohort; however, they do not carry out formal tests of measurement invariance and they do not construct factor scores fully comparable across cohorts as we do.
 Another paper related to ours is the one by \citet{Reardon2016}, who study recent trends in income, racial, and ethnic school gaps in several dimensions of school readiness, including academic achievement, self-control, and externalizing behavior, at kindergarten entry, using comparable data from the Early Childhood Longitudinal Studies (ECLS-K and ECLS-B) for cohorts born from the early 1990s to the 2000–2010 period in the US.
 They find that readiness gaps narrowed modestly from 1998 to 2010, particularly between high- and low-income students and between White and Hispanic students.
-\citet{Landerso2017a} study the sources of differences in social mobility between US and Denmark; for the US, they use the antisocial, headstrong, hyperactivtity subscales from the Behavior Problem Index (BPI) in the Children of the NLSY79 (CNLSY), while for Denmark they use orderliness/organization/neatness grades from the Danish written exams.\footnote{As the authors note (footnote 41) ``Our measures of non-cognitive skills in the two countries are clearly not equivalent.
+\citet{Landerso2017a} study the sources of differences in social mobility between US and Denmark; for the US, they use the antisocial, headstrong and hyperactivtity subscales from the Behavior Problem Index (BPI) in the Children of the NLSY79 (CNLSY), while for Denmark they use orderliness/organization/neatness grades from the Danish written exams.\footnote{As the authors note (footnote 41) ``Our measures of non-cognitive skills in the two countries are clearly not equivalent.
 The Danish measure of non-cognitive skills is more related to an orderliness/effort measure while the US measure is related to behavioral problems''.} They find that, in both countries, cognitive and non-cognitive skills acquired by age 15 are more important for predicting educational attainment than parental income.
 Lastly, \citet{Deming2017} uses two sets of skill measures and comparable covariates across survey waves for the NLSY79 and the NLSY97,\footnote{He uses the following four variables as measures of social skills in the NLSY79: self-reported sociability in 1981 and at age 6 (retrospective), the number of clubs in which the respondent participated in high school and participation in high school sports; and the following two variables in the NLSY97: two questions that capture the extroversion factor from the Big 5 Personality Inventory (since measures comparable to the NLSY79 are not available in the NLSY97).} and finds that the labour market return to social skills was much greater in the 2000s than in the mid-1980s and 1990s.
-\citet{Zilanawala2019} examine differences in socioemotional and cognitive development among 11-year old children in the UK Millennium Cohort Study and the US Early Childhood Longitudinal Survey-Kindergarten Cohort, and find that family resources explain some cross-national differences, however there appears to be a broader range of family background variables in the UK that influence child development.
-Importantly, none of these papers making comparisons across cohorts or ethnic groups test for measurement invariance like we do.
+\citet{Zilanawala2019} examine differences in socio-emotional and cognitive development among 11-year old children in the UK Millennium Cohort Study and the US Early Childhood Longitudinal Study-Kindergarten Cohort, and find that family resources explain some cross-national differences, however there appears to be a broader range of family background variables in the UK that influence child development.
+Importantly, none of these papers making comparisons across countries, cohorts or ethnic groups test for measurement invariance like we do.
 
 \paragraph{Determinants of socio-emotional traits} Equally flourishing has been the literature on the determinants of child socio-emotional skills, which ranges from reduced-form, correlational or causal estimates, to more structural approaches.
-One of the first papers by \citet{Segal2008} has shown that a variety of family and school characteristics predict classroom behaviour.
+One of the first papers by \citep{Segal2008} shows that a variety of family and school characteristics predict classroom behaviour.
 \citet{Carneiro2013} study the intergenerational impacts of maternal education, using data from the NLSY79 and an instrumental variable strategy; they find strong effects in terms of reduction in children's behavioural problems.
-\citet{Cunha2010} and \citet{Attanasio2018a} both estimate production functions for child cognitive and socio-emotional development (in US and Colombia, respectively), and find an important role played by parental investments.
+\citet{Cunha2010} and \citet{Attanasio2020} both estimate production functions for child cognitive and socio-emotional development (in US and Colombia, respectively), and find an important role played by parental investments.
 \citet{Moroni2019} estimate production functions for child socio-emotional skills (internalising and externalising behaviour) at age 11 in the UK Millennium Cohort Study, and find that the effects of parental inputs which improve the home environment varies as a function both of the level of the inputs themselves and of the development of the child.
 
-Interventions for improving Social and Emotional Learning (SEL) in a school setting have shown significant improvements in socio-emotional skills, attitudes, behaviours, and academic performance \citep{Durlak2011}, and a substantial positive return on investments \citep{Belfield2015}; after-school programs have been shown to be equally effective \citep{Durlak2010}.
+Interventions targeting Social and Emotional Learning (SEL) in a school setting have been shown to lead to significant improvements in socio-emotional skills, attitudes, behaviours, and academic performance \citep{Durlak2011}, and a substantial positive return on investments \citep{Belfield2015}; after-school programs have been proved to be equally effective \citep{Durlak2010}.
 
-Additionally, it has been shown that a key mechanism through which early childhood interventions improve adult socioeconomic and health outcomes is by boosting socio-emotional skills, such as four teacher-reported behavioural outcomes in the project STAR\footnote{Student's effort, initiative, non-participatory  behavior, and how the student is seen to `value' the class.} \citep{Chetty2011a}, reductions in externalising behaviour (from the Pupil Behavior Inventory) at ages 7-9 in the Perry Preschool Project \citep{Heckman2013,Conti2016a}, or improvements in task orientation at ages 1-2 in the Abecedarian Project \citep{Conti2016a}.
+Additionally, it has been shown that a key mechanism through which early childhood interventions improve adult socioeconomic and health outcomes is by boosting socio-emotional skills, measured as four teacher-reported behavioural outcomes in the project STAR\footnote{Student's effort, initiative, non-participatory  behavior, and how the student is seen to `value' the class.} \citep{Chetty2011a}, reductions in externalising behaviour (from the Pupil Behavior Inventory) at ages 7-9 in the Perry Preschool Project \citep{Heckman2013,Conti2016a}, or improvements in task orientation at ages 1-2 in the Abecedarian Project \citep{Conti2016a}.
 
 In sum, even if the literature on the determinants and consequences of socio-emotional skills has been booming, most papers use skills measured in late childhood or in adolescence; and no paper in economics formally tests for invariance of measurements across different groups and constructs fully comparable scores.
-In this paper, we use measures of child socio-emotional development at age 5, hence at the very beginning of formal schooling; and we construct comparable scales across the two cohorts we study (the 1970 and the 2000 British cohorts), so that we can investigate changes in inequality in early development, their determinants, and consequences, in a parallel fashion.
+In this paper, we use measures of child socio-emotional development at age 5, hence before the start of elementary school; and we construct comparable scales across the two cohorts we study (the 1970 and the 2000 British cohorts), so that we can investigate changes in inequality in early development, their determinants, and consequences, in a parallel fashion.

### 2020-02-11 codes/R/paper/text/paper_jpube_2datamethods.tex
Gabriella's text updates

@@
 
-We use information from two nationally representative longitudinal studies in the UK.
-The studies follow the lives of children born approximately 30 years apart: the British Cohort Study (BCS) surveys individuals born in 1970, and the Millennium Cohort Study (MCS) includes births between 2000 and 2002.
-The British Cohort Study includes all individuals born across Great Britain in a single week in 1970.
-Cohort members' families -- and subsequently members themselves -- were surveyed on multiple occasions.
-For this paper we augment the information at the five-year survey with data from birth, adolescence (16), and adulthood (30, 38, 42).
-The Millennium Cohort Study follows individuals born in the UK between September 2000 and January 2002.
-We use the first (age 0) survey -- carried out at 9 months of age -- and the sweeps at around 5 and 14 years of age.\footnote{All data is publicly available at the UK Data Service \citep{Chamberlain2013,Butler2016a,Butler2016b,Butler2017,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2016b,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2016,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2016a,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2017b,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2017,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2017c}.}
+We use information from two nationally representative longitudinal studies in the UK, which follow the lives of children born approximately 30 years apart: the British Cohort Study (BCS) and the Millennium Cohort Study (MCS).
+The BCS includes all individuals born in Great Britain in a single week in 1970.
+The cohort members' families -- and subsequently the members themselves -- were surveyed on multiple occasions.
+For this paper we augment the information collected at the five-year survey with data from birth, adolescence (16), and adulthood (30, 38, 42).
+The MCS follows individuals born in the UK between September 2000 and January 2002.
+We use the first survey -- carried out at 9 months of age -- and the sweeps at around 5 and 14 years of age.\footnote{All data is publicly available at the UK Data Service \citep{Chamberlain2013,Butler2016a,Butler2016b,Butler2017,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2016b,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2016,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2016a,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2017b,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2017,UniversityOfLondon.InstituteOfEducation.CentreForLongitudinalStudies2017c}.}
 
@@
 However, the cognitive tests administered to each cohort have no overlap, even at the item level.
-We thus use the available cognitive tests in each cohort to estimate simple confirmatory factor model with a single latent dimension, separately by cohort (see \autoref{tab:harmvar} for the tests used).
-Unlike the other indicators in our analysis, cognitive skills are thus not comparable across cohort.
+We thus use the available cognitive tests in each cohort to estimate simple confirmatory factor models with a single latent dimension, separately by cohort (see \autoref{tab:harmvar} for the tests used).
+Unlike the other indicators in our analysis, cognitive skills are thus not comparable across cohorts.
 
@@
 This choice is mainly driven by software limitations.
-The \texttt{lavaan} package in \textbf{\textsf{R}} \citep{Rosseel2012} is the most suitable tool for our invariance analysis, but it does not allow to use weights when outcomes are categorical, as it is the case for the socio-emotional measurements.} Finally, we also restrict our sample to individuals born in England and to cases where there is complete information on socioemotional skills at five years of age.
+The \texttt{lavaan} package in \textbf{\textsf{R}} \citep{Rosseel2012} is the most suitable tool for our invariance analysis, but it does not allow to use weights when outcomes are categorical, as it is the case for the socio-emotional measurements.} Finally, we also restrict our sample to individuals born in England and to cases where there is complete information on socio-emotional skills at five years of age.
 The final sample contains 9,545 individuals from the British Cohort Study, and 5,572 from the Millennium Cohort Study.
@@
 Over recent years, the measurement of such skills has evolved and, over time, different measures have been used.
-As we discuss below, this makes the comparison  of socio emotional skills across different groups, assessed with different tools, difficult.
+As we discuss below, this makes the comparison  of socio-emotional skills across different groups, assessed with different tools, difficult.
 
@@
 In the British and Millennium Cohort Studies, two different scales were employed.
-In the BCS, the Rutter A Scale was used \citep{Rutter1970} while in the MCS cohort, mothers were  administered the Strengths and Difficulties Questionnaire (SDQ, \citealp{Goodman1994,Goodman1997}).
+In the BCS, the Rutter A Scale was used \citep{Rutter1970} while in the MCS mothers were  administered the Strengths and Difficulties Questionnaire (SDQ, \citealp{Goodman1994,Goodman1997}).
 The SDQ was created as an update to the Rutter scale.
@@
 The Rutter and SDQ scales are reproduced in \autoref{tab:scales}; they have 23 and 25 items each, respectively.
-In the child psychiatry and psychology literatures, the Rutter and SDQ behavioural screening scales are regarded as measures of behavioural problems and mental health.
+In the child psychiatry and psychology literatures, the Rutter and SDQ scales are regarded as measures of behavioural problems and mental health.
 However, in our analysis we follow the economics literature, and - after having recoded them accordingly - we interpret them as measures of positive child development \citep{Goodman2011}.
@@
 Fourth, and no less importantly, the interpretation of each behaviour by respondents living 30 years apart (1975 vs 2006) might differ due to a host of evolving societal norms.
-Nonetheless, the level of comparibility of the two scales is higher than that of other scales used in comparating work in the literature reviewed in section \ref{sec:lit}.
+Nonetheless, the level of comparibility of the two scales is higher than that of other scales used in comparative work in the literature reviewed in section \ref{sec:lit}.
 
@@
 The wording of the items we will be using in the analysis is presented in \autoref{tab:scalecomp}: we retain 13 items for the BCS (two of them are grouped) and 11 for the MCS with high degree of comparability.
-We exclude from the analysis items that were completely different between the two questionnaires to maximise comparability between the two cohorts, as it is standard good practice in the psychometric literature (see for example \citet{Kern2014}).\footnote{Of course, we could have included them in the factor analysis and treated them as missing in the cohort were they were not administered.} More details on the derivation of the scale are available in Appendix \autoref{asec:scales}.
+We exclude from the analysis items that were completely different between the two questionnaires to maximise comparability between the two cohorts, as it is standard good practice in the psychometric literature (see for example \citet{Kern2014}).\footnote{Of course, we could have included them in the factor analysis and treated them as missing in the cohort where they were not administered.} More details on the derivation of the scale are available in Appendix \autoref{asec:scales}.
 
 Item-level prevalence by cohort and gender is in \autoref{tab:meantab}.
-We see that, in general, item prevalence is more similar across genders within the same cohort, than across cohorts.
+We see that, in general, there are more similarities across genders within the same cohort, than across cohorts.
 For the majority of items, there is a lower prevalence of problematic behaviours in the MCS than in the BCS; however, four items (distracted, tantrums, fearful, aches) show a higher prevalence in 2006 than in 1975.
@@
 This model fixes the number of latent dimensions, and imposes a dedicated measurement structure, based on the insights obtained in the exploratory step.
-This is the model from which the measurement invariance analysis of \autoref{sec:measinv} departs.
+This is the model to which we apply the measurement invariance analysis of \autoref{sec:measinv}.
 
@@
 
-The original Rutter scale, used for the BCS cohort, distinguishes behaviours into two subscales: \emph{anti-social} and \emph{neurotic} \citep{Rutter1970}.
+The original Rutter scale, used in the BCS cohort, distinguishes behaviours into two subscales: \emph{anti-social} and \emph{neurotic} \citep{Rutter1970}.
 This two-factor conceptualisation has been validated using data from multiple contexts, and the latent dimensions have been broadly identified as externalising and internalising behaviour problems.\footnote{See for example \citet{Fowler1979,Venables1983,Tremblay1987,Berglund1999,Klein2009}.
-However, in some cases a three-factor structure was found to better fit the data, with the externalising factor separating into two factors seemingly capturing aggressive and hyperactive behaviours \citep{Behar1974,McGee1985}.} The Strength and Difficulties Questionnaire, used for the MCS cohort, was conceived to have five subscales of five items each.
+However, in some cases a three-factor structure was found to better fit the data, with the externalising factor separating into two factors seemingly capturing aggressive and hyperactive behaviours \citep{Behar1974,McGee1985}.} The Strength and Difficulties Questionnaire, used in the MCS cohort, was instead conceived to have five subscales of five items each.
 The five subscales are: \emph{hyperactivity}, \emph{emotional symptoms}, \emph{conduct problems}, \emph{peer problems}, and \emph{prosocial}.
@@
 Recent research has shown that there are some benefits to using broader subscales that correspond to the externalising and internalising factors in Rutter, especially in low-risk or general population samples \citep{Goodman2010}.
-Indeed, the internalising and externalising dimensions were introduced in psychology by \citet{Achenbach1966}, who show that they are the two main factors underlying a wide range of psychological measures; as noted in \citet{Achenbach2016}, more than 75,000 articles have been published on internalizing and externalizing problems.
+Indeed, the internalising and externalising dimensions were introduced in psychology by \citet{Achenbach1966}, who showed that they are the two main factors underlying a wide range of psychological measures; as noted in \citet{Achenbach2016}, more than 75,000 articles have been published on internalizing and externalizing problems.
 
@@
 This is a `minimum' identifiable model, in that it places the least possible restrictions on how parameters are allowed to vary across cohorts.
-It serves as a basis for our measurement invariance analysis.
-
-The restrictions implied in \eqref{eq:propens} and \eqref{eq:thresh} are not sufficient to identify the parameters of the model.
-Even with these assumptions, there are infinite equivalent parameterisations (or rotations) that deliver a minimally restrictive configural model.
+The restrictions implied in \eqref{eq:propens} and \eqref{eq:thresh} are not sufficient to identify the parameters of the model: even with these assumptions, there are infinite equivalent parameterisations (or rotations) that deliver a minimally restrictive configural model.
 This is the well-known issue of factor indeterminacy, which arises due to the lack of natural units of measurement for the latent factors being assessed.
@@
   \bi
-  \item For all groups, normalise a reference loading to 1 for each factor
-  \item Set invariant across groups one threshold per item (e.g. $\tau_{0,Ai} = \tau_{0,Bi}$ for two groups $A$ and $B$), and an additional threshold in the reference items above
-  \item In the first group: $\bm{\kappa_A}=\bm{0}$, $\text{diag}(\bm{\Sigma}_A)=\bm{I}$
-  \item Set all intercepts $\bm{\nu}$ to zero
+  \item For all groups, normalise a reference loading to 1 for each factor.
+  \item Set invariant across groups one threshold per item (e.g. $\tau_{0,Ai} = \tau_{0,Bi}$ for two groups $A$ and $B$), and an additional threshold in the reference items above.
+  \item In the first group: $\bm{\kappa_A}=\bm{0}$, $\text{diag}(\bm{\Sigma}_A)=\bm{I}$.
+  \item Set all intercepts $\bm{\nu}$ to zero.
   \ei
@@
 
-Any comparison between socioemotional skills across the two cohorts requires that the measures at our disposal have the same relationship with the latent constructs of interest in both cohorts.
-In other words, the items in our new scale must measure externalising and internalising socioemotional skills in the same way in the BCS and MCS data.
-This property is denominated measurement invariance \citep{Vandenberg2000a,Putnick2016}.
+Any comparison between socio-emotional skills across the two cohorts requires that the measures at our disposal have the same relationship with the latent constructs of interest in both cohorts.
+In other words, the items in our new scale must measure externalising and internalising socio-emotional skills in the same way in the BCS and MCS data.
+This property is denominated measurement invariance (MI) \citep{Vandenberg2000a,Putnick2016}.
 
@@
 In this paper, we follow the recent identification methodology by \citet{Wu2016a}.
-The configural model defined above serves as the starting point.
+The configural model defined above in section \ref{sec:config} serves as the starting point.
 Measurement invariance is then assessed by comparing the configural model to a series of hierarchically nested models.
@@
 
-In the case where available measures are continuous, MI analysis is straightforward \citep{vandeSchoot2012}.
+In the case where the available measures are continuous, MI analysis is straightforward \citep{vandeSchoot2012}.
 The hierarchy of the nested models usually proceeds by testing loadings first, and then intercepts (to establish \emph{metric} and \emph{scalar} invariance -- see \citealp{Vandenberg2000a}).
@@
 In other words, items have the same salience, or informational content relative to skills.
-If this model fits as well as the configural model, we can be confident that the socioemotional skills of children in the two cohorts can be placed on the same scale, and their \emph{variances} can be compared.
+If this model fits as well as the configural model, we can be confident that the socio-emotional skills of children in the two cohorts can be placed on the same scale, and their \emph{variances} can be compared.
 To see why, consider equation \eqref{eq:propens}.
@@
 A \emph{loading, threshold, and intercept invariance} model additionally restricts every item's intercept $\nu$ across cohorts.
-A good relative fit of this model indicates that socioemotional skills can be compared across cohorts in terms of their \emph{means} as well.
+A good relative fit of this model indicates that socio-emotional skills can be compared across cohorts in terms of their \emph{means} as well.
 To see why, consider the following.
@@
 \citet{Chen2007} and \citet{Meade2008} -- have updated these thresholds for the continuous case.
-In particular, \citet{Chen2007} shows in two Monte Carlo studies that the standardised root mean square residual (RMSR) is more sensitive to lack of invariance in factor loadings than in intercepts or residual variances, while the CFI and RMSEA are equally sensitive to all three types of lack of invariance; he suggests the following thresholds for rejecting measurement invariance: \textDelta RMSE $> .015$, \textDelta CFI $< - .010$, \textDelta SRMR $> .010$.
+In particular, \citet{Chen2007} shows in two Monte Carlo studies that the standardised root mean square residual (RMSR) is more sensitive to lack of invariance in factor loadings than in intercepts or residual variances, while the CFI and RMSEA are equally sensitive to all three types of lack of invariance; he suggests the following thresholds for rejecting measurement invariance: \textDelta RMSE $> .015$, \textDelta CFI $< - .010$, \textDelta RMSR $> .010$.
 
 However, it is not advisable to directly extrapolate rules of thumb derived from simulations with continuous measures to the categorical case \citep{Lubke2004}.
-Recent studies have advanced simulation evidence around the performance of AFIs in measurement invariance analysis with categorical measures.
-\citet{Sass2014} finds that the cutoffs from \citet{Chen2007} might not generalise well to problems estimated by WLSMV, but this is mostly confined to smaller sample sizes and detection of small degrees of non-invariance.
+Recent studies have advanced the simulation-based evidence on the performance of AFIs in measurement invariance analysis with categorical measures.
+\citet{Sass2014} find that the cutoffs from \citet{Chen2007} might not generalise well to problems estimated by WLSMV, but this is mostly confined to smaller sample sizes and detection of small degrees of non-invariance.
 More recently, \citet{Rutkowski2017} find that a \textDelta RMSE threshold of $.010$ is appropriate for testing equality of slopes and thresholds when the sample size is large, like in our case.
 
-In any case, we present a range of fit indices to provide a more holistic assessment of measurement invariance.
+In any case, we present a range of fit indices to provide a more complete assessment of measurement invariance.
 The fit of each nested model is compared in Panel A of \autoref{tab:fit}.
@@
 In particular, the \textDelta AFIs fall within the ranges suggested in \citet{Chen2007}, \citet{Rutkowski2017} and \citet{Svetina2017}; see also \citet{Svetina2019} for a review of updated guidelines for measurement invariance.
-Invariance of loadings and thresholds across cohorts implies that items in our scale are equally salient in their informational content, and that the latent propensities have equal mapping into the observed items.
+Invariance of loadings and thresholds across cohorts implies that the items in our scale are equally salient in their informational content, and that the latent propensities have equal mapping into the observed items.
+
+\textbf{[ADD NEW RESULTS]}
 
@@
 
-To summarise, our measurement invariance analysis shows partial comparability of socioemotional skills across cohorts.
+To summarise, our measurement invariance analysis shows partial comparability of socio-emotional skills across cohorts.
 In particular, the variance of skills can be compared across cohorts, but mean cohort differences do not necessarily reflect differences in skills.

### 2020-02-11 codes/R/paper/text/paper_jpube_3resultsconclusion.tex
Gabriella's text updates

@@
 However, the shape of the distribution can be given a cross-cohort interpretation.
-This result is in sharp contrast with what shown by the simple distribution of sum scores in \ref{fig:sumscores}: using raw scores shows an increase in mass only at the top of the distribution.
-It is immediately visible that there is more mass in the tails of the distribution in the 2000 than in the 1970 cohort.
+This result is in sharp contrast with what shown by the simple distribution of sum scores in \ref{fig:sumscores}: using raw scores we see an increase in mass only at the top of the distribution, while the factor scores clearly show that there is more mass in both tails of the distribution of the 2000 than of the 1970 cohort.
 
-\subsection{Inequality in socioemotional skills}
+\subsection{Inequality in socio-emotional skills}
 
@@
 \autoref{fig:ineqps} shows mean skills by maternal education.
-We compare mothers who continued education past the compulsory age with mothers who left school at the compulsory leaving age, according to their year of birth.
+We compare mothers who continued education with mothers who left school at the minimum compulsory leaving age, according to their year of birth.
 Given lack of comparability in the level of skills across cohort, we normalise the mean in the `Compulsory' group to zero for both cohorts.
@@
 In general, this has been a period of significant societal changes, with an almost continual rise in the proportion of women in employment, an older age at first birth and a rise in dual-earning parents families \citep{Roantree2018}.
-However, here we do not attempt to disentangle whether and to which extent the observed changes in inequality in socio-emotional skills across the two cohorts can be attributed to changes in returns (or penalties) to maternal characteristics (such as education and employment) or to compositional changes, as has been done for the analysis of wage inequality \citep{Blundell2007}.
+
+Hence, we lastly attempt to disentangle whether and to which extent the observed changes in inequality in socio-emotional skills across the two cohorts can be attributed to changes in returns (or penalties) to characteristics such as maternal education, or to compositional changes.
+To this aim, we use the method recently developed by \citet{firpo2018decomposing}\footnote{See also \citet{fortin2011decomposition} for a recent survey of decomposition methods in economics.} as an extension of the Oaxaca-Blinder (OB) decomposition to any distributional measure, that here we apply for the first time to changes in inequality in early childhood development.
+This two-stage procedure first decomposes distributional changes into a `composition effect' and a `coefficient effect' using a reweighting method; then it further divides these two components into the contribution of each explanatory variable, using Recentered Influence Function (RIF) regression \citep{firpo2009unconditional}.
+
+Following \citet{firpo2018decomposing}, we first perform an OB decomposition using the BCS sample and the counterfactual sample (BCS reweighted to be as MCS)\footnote{We use a logit model to construct the weights and the post-double selection lasso \citep{belloni2012sparse} to select the covariates, among the set of the baseline variables in \autoref{tab:sumstats} and their pairwise interactions.} to get the pure composition effect, using the BCS as reference coefficients.
+The total unexplained effect in this decomposition corresponds to the specification error, and allows to assess the importance of departures from the linearity assumption.
+Second, we perform the decomposition using the MCS sample and the counterfactual sample, to obtain the pure coefficient effect (the `unexplained' part); the explained effect in this decomposition corresponds to the reweighting error, which allows to assess the quality of the reweighting.
+
+In \autoref{fig:rif} we present the results of the RIF decomposition for changes in five measures of inequality in socio-emotional skills for the boys, both externalising (top figure) and internalising (bottom figure).
+The results indicate that different factors explain the rise in inequality in the two skills: on the one hand, compositional changes explain, on average, half of the cross-cohort increase in inequality in externalising skills, regardless of the measure considered;\footnote{The coefficient effects are also sizeable, but imprecisely estimated, with the exception of the variance component.} on the other hand, the increase in inequality in internalising skills seems to be entirely explained (even over-explained) by changes in returns (or penalties) to background characteristics.
+Composition and coefficient effects are further decomposed in the contribution of each covariate, and the results presented in \autoref{tab:RIF-EXT-M} and in \autoref{tab:RIF-INT-M}.
+We see in \autoref{tab:RIF-EXT-M} that mother's age and marital status at birth are the two variables that best account for the compositional changes, driving the increase in inequality in externalising skills among the boys for the quantile differences and the variance, respectively.
+This is hardly surprising, given that we have seen in \autoref{tab:sumstats} that the average age of the mother at birth has increased by approximately three years (from 26 to 29 years old), and that the proportion of unmarried mothers has increased dramatically, from 5\% in the BCS to 36\% in the MCS.
+The baseline covariates, instead, do a less impressive job at explaining the changes in coefficients underlying the increase in inequality in internalising skills (\autoref{tab:RIF-INT-M}, note the changes in returns to maternal employment go in the direction of reducing inequality).
+This can be partly explained by the fact that, due to lack of comparable measures across cohorts, we have been unable to account for important determinants of a child's internalising behaviour, such as for example maternal mental health.
+We also notice that, for the quantile differences 75-25 and 90-50, the composition effect is significant but negative; in other words, compositional changes linked to maternal marriage status would have led to a reduction in inequality, especially at the top of the distribution.
+Reassuringly, both the specification and the reweighting error are not significantly different from zero.
+Lastly, the results are not so clear-cut for the girls, who experienced a more muted increase in inequality, concentrated at the bottom of the distribution.
+The RIF results displayed in \autoref{tab:RIF-F} show that no single contributing factor emerges.
 
@@
 
-In this last section, we study the predictive power of socio-emotional skills for adolescent and adult outcomes.
-We contribute to a vast interdisciplinary literature by examining medium- and long-term impacts of skills measured at an earlier age than usually examined in previous studies, i.e.
+In this last section, we study the predictive power of socio-emotional skills for adolescent and adult outcomes, to gain some insights as to whether inequality in the early years could translate into later life inequalities.
+We contribute to a vast interdisciplinary literature by examining medium- and long-term impacts of skills measured at an earlier age than in previous studies, i.e.
 well before the start of formal education.
-We do so by regressing health and socioeconomic outcomes measured in adolescence and adulthood on the socio-emotional skills scores at age five obtained by our factor model, controlling for the harmonised family background variables at birth and age five (see \autoref{tab:harmvar}).\footnote{In tables \ref{tab:adol_sum} and \ref{tab:bcs_adultapp}, we show that the conclusions in this section are not sensitive to the factor scoring methodology used.
+Showing that these early skills are predictive of different later outcomes across various domains provide a key rationale for the role of early intervention in reducing life course inequalities.
+In practice, we proceed by regressing health and socioeconomic outcomes measured in adolescence and adulthood on the socio-emotional skills scores at age five obtained by our factor model, controlling for the harmonised family background variables at birth and age five (see \autoref{tab:harmvar}).\footnote{In tables \ref{tab:adol_sum} and \ref{tab:bcs_adultapp}, we show that the conclusions in this section are not sensitive to the factor scoring methodology used.
 'Raw' scores, obtained by a simple unweighted average of the item categories in the 11-item subscale have basically equal predictive power to factor scores.} We present results with and without controlling for cognitive skills.
@@
 Externalising skills are negatively correlated to subsequent smoking and BMI in both cohorts, for both genders.
-Recall that a child with high externalising skills exhibits less restless and hyperactive behaviours, and has less anti-social conduct.
+Recall that a child with high externalising skills exhibits less restless and hyperactive behaviour, and has less anti-social conduct.
 Our findings are consistent with the body of evidence reviewed in section \ref{sec:lit}, which shows that better socio-emotional skills (measured using different scales and at various points during childhood and adolescence) are negatively associated with smoking.
@@
 From this perspective, he/she is likely more sociable and subject to peer influence in health behaviours.
-This is consistent with the evidence in \citet{Goodman2015}, who find a positive association between child emotional health (measured with items for the internalizing behaviour subscale of the Rutter scale at age 10 in the BCS) and smoking at age 42.
+This is consistent with the evidence in \citet{Goodman2015}, who find a positive association between child emotional health (measured with items from the internalizing behaviour subscale of the Rutter scale at age 10 in the BCS) and smoking at age 42.
 Furthermore, in recent work \citet{Hsieh2018} have shown personality to be a key mechanism through which peers affect smoking behaviour.
+We have also tested the robustness of these findings by jointly estimating by maximum likelihood the measurement system (with the partial invariance constraints) and the two outcome equations for smoking and BMI.
+The results, presented in \autoref{tab:adolML}, are qualitatively similar to those obtained with the two-step method.\footnote{Note that the magnitudes are not exactly comparable because in one-step ML estimation the residual variances of the non-binary measurements also need to be fixed for identification.
+The remaining parameter estimates are also very similar to those in \autoref{tab:measpars} and available from the authors upon request.}
 
@@
 Along the same lines, \citet{Conti2013a} use rich data on child personality and socio-emotional traits collected at ages 7, 11 and 16 in the 1958 British birth cohort,\footnote{They use the Rutter scale and the Bristol Social Adjustment Guide.} and show that these traits rival the importance of cognition in explaining the education gradient in health behaviours (including smoking and BMI).
-We show that child socio-emotional skills have greater predictive power for health outcomes and behaviours even when measured at an earlier age.
+We show that child socio-emotional skills have greater predictive power than cognition for health outcomes and behaviours even when measured at an earlier age than in previous work.
 
@@
 
-In this paper we have studied inequality in a dimension of human capital which has received limited attention in the literature so far: socio-emotional skills very early in life.
+In this paper we have studied inequality in a dimension of human capital which has received less attention than others in the literature so far: socio-emotional skills very early in life.
 In particular, we have focused on the measurements of these skills at age 5 in two British cohorts born 30 years apart: the one of children born in 1970 (British Cohort Study, BCS) and the one of children born in 2000/1 (Millennium Cohort Study, MCS).
-We have provided several contributions to the recent but flourishing literature on the determinants and consequences of early human development.
+We have provided a timely contribution to the recent but flourishing literature on the determinants and consequences of early human development, by bridging it with the inequality literature.
 
@@
 
-Second, we have formally tested for measurement invariance of the 11 items across the two externalising and internalising scales, following recent methodological advances in factor analysis with categorical outcomes.
-We have found only partial support for measurement invariance, with the implication that we have only been able to compare how inequality in these socio-emotional skills across the two cohorts has changed, but not whether their average level is larger or smaller in one of the two cohorts.
+Second, we have formally tested for measurement invariance across the two cohorts (for each gender) of the 11 items comprising the two externalising and internalising scales, following recent methodological advances in factor analysis with categorical outcomes.
+We have found only partial support for measurement invariance, with the implication that we have only been able to compare how inequality in these socio-emotional skills has changed across the two cohorts, but not whether their average level is higher or lower in one of them.
 These results sound a warning to research in this area which routinely compares levels of skills across different groups (at different times, or of different gender), without first establishing their comparability.
 
-Third, after having computed comparable scores for both externalising and internalising skills, and for both boys and girls, we have compared how inequality in these skills has changed across the 1970 and the 2000 cohort.
-We have documented for the first time that inequality in these early skills has increased across cohorts, especially for boys.
+Third, after having computed comparable scores for both externalising and internalising skills, and for both boys and girls, we have compared how inequality in these skills has changed across the 1970 and the 2000 cohorts.
+We have documented for the first time that inequality in these early skills has increased, especially for boys.
 The cross-cohort increase in the gap is more pronounced at the bottom of the distribution (50-10 percentiles).
@@
 On the other hand, the skills penalty arising from the lack of a father figure in the household has substantially declined.
+Moreover, we have formally decomposed the increase in inequality into compositional changes, and changes in returns to maternal characteristics - providing the first child development application of the method recently developed by \citet{firpo2018decomposing}.
+We have found that half of the increase in inequality in externalising skills across cohorts can be explained by compositional changes, with maternal age and marital status at birth being the most important factors; on the other hand, the increase in inequality in internalising skills seems to be entirely driven by changes in returns to maternal characteristics.
 
@@
 In particular, socio-emotional skills are more significant predictors of health and health behaviours (smoking and BMI), while cognition has greater predictive power for socioeconomic outcomes (education, employment and wages).
-Our results show the importance of inequalities in the early years development for the accumulation of health and human capital across the life course.
+Our results ultimately show the importance of inequalities in the early years development for the accumulation of health and human capital across the life course.

### 2020-02-11 codes/R/paper/text/paper_jpube_4appendix.tex
Gabriella's text updates

@@
-\section{Deriving a common scale of socioemotional skills}\label{asec:scales}
+\section{Deriving a common scale of socio-emotional skills}\label{asec:scales}
 
-In the BCS data, maternal reports on child socioemotional skills are measured using the Rutter A Scale \citep{Rutter1970} -- see Panel A of \autoref{tab:scales}.
+In the BCS, maternal reports on child socio-emotional skills are measured using the Rutter A Scale \citep{Rutter1970} -- see Panel A of \autoref{tab:scales}.
 The Rutter items are rated on three levels: `Does not apply', `Somewhat applies', `Certainly applies'.
-Since they are all behaviours indicating lower skills, we recode all of them in reverse, i.e.
+Since they all indicate negative behaviours, we recode all of them in reverse, i.e.
 `Certainly applies' = 0, `Somewhat applies' = 1, `Does not apply' = 2.
 We augment the 19-item Rutter Scale with three additional parent-reported questions from the parental questionnaire, items A, B, and D.
-These are rated on 4 levels: `Never in the last 12 months', `less than once a month', `at least once a month', `at least once a week'.
-we recode these into binary indicators, with `Never' and `Less than once a month' to 1 and zero otherwise.
-To increase comparability between the two scales, we merge together two pairs of items: 4 and 19 (to mirror SDQ item 12 ``Often fights with other children or bullies them"), and A and B (to mirror SDQ item 3 ``Often complains of head-aches, stomach-ache or sickness").
-We assign the lowest category among the two original items to the newly obtained item.
-We also recode items 5 and 14 to binary instead of three categories.
-These items are recorded with a positive phrasing in SDQ, so a 3-category split would be harder to compare.
+These are rated on 4 levels: `Never in the last 12 months', `less than once a month', `at least once a month', `at least once a week'; we recode them into binary indicators, with `Never' and `Less than once a month' to 1 and zero otherwise.
+To increase comparability between the two scales, we also merge together two pairs of Rutter items: 4 and 19 (to mirror SDQ item 12 ``Often fights with other children or bullies them"), and A and B (to mirror SDQ item 3 ``Often complains of head-aches, stomach-ache or sickness"); we assign the lowest category among the two original items to the newly obtained item.
+We also recode the three-category Rutter items 5 and 14 to binary to mimic the split in the MCS, where they are worded positively.
 
-In MCS, we use the 25-item Strengths and Difficulties Questionnaire \citep{Goodman1997} -- see Panel B of \autoref{tab:scales}.
+In the MCS, we use the 25-item Strengths and Difficulties Questionnaire \citep{Goodman1997} -- see Panel B of \autoref{tab:scales}.
 All items are recorded on a 4-point scale: `Not true', `Somewhat true', `Certainly true', `Can't say'.
-We set the latter option to missing and recode the rest in ascending order of skill as for the BCS items, i.e.
-`Certainly true' = 0, `Somewhat true' = 1, `Not true' = 2.
-For comparability with the BCS Rutter scale, we dichotomise items 3 and 5, and dichotomise and invert items 7, and 14.
+We set the latter option to missing and recode the rest so that a greater value represents a higher level of skills, as for the BCS items, i.e.
+`Certainly true' = 0, `Somewhat true' = 1, `Not true' = 2 for the negatively-worded items (and the opposite for the positively-worded ones).
+For comparability with the BCS Rutter scale, we dichotomise items 3 and 5, and dichotomise and invert items 7 and 14.
 
@@
 In this section, we repeat the exploratory analysis step in \autoref{sec:ea} for the full set of Rutter and SDQ items.
-This is to show that the factor structure emerging from exploratory analysis of the 11-item subscale is consistent with what would emerge considering the original scales in their entirety.
-Again, we proceed by first assessing the optimal number of factors, and then examining the loading obtained from exploratory factor analysis.
+This is to show that the factor structure emerging from the exploratory analysis of the 11-item subscale is consistent with what would emerge considering the original scales in their entirety.
+Again, we proceed by first assessing the optimal number of factors, and then examining the loadings obtained from exploratory factor analysis.
 
@@
 The only exception is the "headaches/stomachaches" item, which seems to load on a separate factor.
-We carry out robustness checks for the measurement invariance analysis excluding this item in Appendix \autoref{asec:no511}.
+We then carry out robustness checks for the measurement invariance analysis excluding this item in Appendix \autoref{asec:no511}.
 

### 2020-02-11 codes/R/paper/text/paper_jpube_5apptables.tex
fix table size
Gabriella's text updates

@@
 \centering\setstretch{1.2} \scriptsize
-\begin{tabular}{L{.18\textwidth}C{.1\textwidth}L{.3\textwidth}L{.75\textwidth}}
+\begin{tabular}{L{.15\textwidth}C{.08\textwidth}L{.22\textwidth}L{.55\textwidth}}
 \toprule
@@
 \\ \\[-1.5em]
-Maternal employment & 5 & Employed$^d$ & Includes full time and part time \\ \\[-1.5em]
+Maternal employment & 5 & Employed$^d$ & Includes full time and part time. \\ \\[-1.5em]
 Father occupation & 5 & White collar (I-IIINM)$^d$ \newline Blue collar (IIIM-V-other)$^d$ \newline No father figure$^d$ & Based on father's Registrar General Social Class classification of occupations.

### 2020-03-14 codes/R/paper/text/paper_jpube_1introlit.tex
update text

@@
 
-[\textbf{ADD REFS}]
-
-Human capital is a key determinant of economic growth and performance and of the resources an individual creates and controls over the life cycle.
-Human capital is also important for various determinants of individual well-being, ranging from health to life satisfaction.
-In recent years, the process of human capital accumulation has received considerable attention.
+Human capital is a key determinant of economic growth and performance and of the resources an individual creates and controls over the life cycle \citep{hanushek2008role}.
+Human capital is also important for various determinants of individual well-being, ranging from life satisfaction to health \citep{conti2019developmental}.
+In recent years, the process of human capital accumulation has received considerable attention \citep{almond2018childhood}.
 There is growing consensus on the fact that human capital is a multidimensional object, with different domains playing different roles in labour market as well as in the determination of other outcomes, including the process of human development.
-It is also recognised that human capital is the output of a very persistent process, where early years inputs play an important and longlasting role.
+It is also recognised that human capital is the output of a very persistent process, where early years inputs play an important and longlasting role \citep{Cunha2010}.
 
@@
 Moreover, there is a lack of consensus on the best measures and on the tools to collect high quality data.
-As a consequence, even when data are available in different contexts, their comparability is problematic.
+As a consequence, even when data are available in different contexts, their comparability is problematic \citep{richter2019early}.
 
@@
 First, we construct a novel scale of childhood behavioural traits from two validated instruments and assess its comparability across cohorts.
-By performing an exploratory factor analysis, we determine that two dimensions are a parsimonious representation of socio-emotional skills for both cohorts.
+By performing exploratory and multiple-group factor analyses, we determine that two dimensions are a parsimonious representation of socio-emotional skills for both cohorts.
 Coherently with previous literature, we label them as `internalising' and `externalising' skills, the former relating to the ability of children to focus their drive and determination, and the latter relating to their ability to engage in interpersonal activities.

### 2020-03-14 codes/R/paper/text/paper_jpube_2datamethods.tex
update text

@@
 As pointed out in \citet{Conti2014}, there is relatively little agreement among procedures; this is the case especially for the Rutter items in the BCS data, where different methods suggest to retain between 1 and 3 factors, while most methods suggest to retain 2 factors for the SDQ items in the MCS.
-In our analysis, we adopt two factors, which is the average number of factors across all methods for each cohort and gender.
-This choice is also consistent with the child psychology literature cited above, and allows us to work with the same number of factors in the two cohorts.
 
-The two-factor EFA delivers a neat and sensible separation between items, as shown in \autoref{tab:load}: reassuringly, similarly-worded items load on the same factor across the two cohorts, and also the magnitude of the respective loadings (measuring the strength of the association between the item and the factor) is very similar.
+Given the test results, we perform a series of exploratory factor analyses, assuming a one-, two- or three-factor structure, respectively.
+The results for the 1-factor solution, reported in \autoref{tab:load_add}, show relatively similar loadings for both males and females across the two cohorts, of slightly bigger magnitude for the last four items in the MCS than in the BCS; thus, we retain the 1-factor solution for the measurement invariance analysis, in the first instance.
+The results for the 3-factor solution, instead, also reported in \autoref{tab:load_add}, show a less homogeneous picture:\footnote{We do not perform the EFA with 3 factors for males because this solution is never chosen by any test for the number of factors for the MCS, see \autoref{tab:numfac}.} while the magnitude of the loadings is relatively similar across the two cohorts for the first factor, items 3 and 5 only load on the second factor for the MCS, not for the BCS; more importantly, the EFA clearly shows that the third factor only loads on one single item (item number 9, ``solitary'') for both cohorts.
+Given that a one-item factor implies that the item perfectly proxies for the factor, we are not able to test for measurement invariance in this case.
+Hence, the 3-factor solution is not supported by our EFA results.
+Last, the two-factor EFA is shown in \autoref{tab:load} and delivers a neat and sensible separation between items: similarly-worded items load on the same factor across the two cohorts, and also the magnitude of the respective loadings (measuring the strength of the association between the item and the factor) is very similar.
 Following previous research, we name the first dimension \emph{Externalising skills} (EXT, indicating low scores on the items restless, squirmy/fidgety, fights/bullies, distracted, tantrums, and disobedient) and the second dimension \emph{Internalising skills} (INT, indicating low scores on the items worried, fearful, solitary, unhappy, and aches).\footnote{Internalising and externalising dimensions emerge from the exploratory step on our novel 11-item scale.
@@
 
-Equipped with the factor structure inferred in the previous section, we specify a multiple-group factor analysis model to formally quantify the strength of the relationship between the observed items in our scale and the two latent socio-emotional skills.
+After having studied the factor structure underlying the 11 common items in the previous section, we now specify a multiple-group factor analysis model to formally quantify the strength of the relationship between the observed items in our scale and the latent socio-emotional skills, and to test for invariance across cohorts.
 We specify two groups of children $c=\{BCS,MCS\}$, corresponding to the two cohorts.
@@
 For each child $j$ in cohort $c$, we observe categorical items $X_{ijc}$ with $i=1,\dots,11$ corresponding to the eleven maternal reports in \autoref{tab:scalecomp}.
-We assume that each child is characterised by a latent bidimensional vector of externalising and internalising socio-emotional skills $\bm{\theta}_{jc} = (\theta_{jc}^{EXT},\theta_{jc}^{INT})$.
+Following the EFA results above, we specify two models: one in which we assume that each child is characterised by only one latent skills vector, and another in which we assume that each child is characterised by a latent bi-dimensional vector of externalising and internalising socio-emotional skills $\bm{\theta}_{jc} = (\theta_{jc}^{EXT},\theta_{jc}^{INT})$.
 
@@
 We make the common assumption of a dedicated (or congeneric) factor structure, where each measure is assumed to load on only one latent dimension \citep{Heckman2013,Conti2010a,Attanasio2018a}.
-We mirror the structure found in the exploratory factor analysis (see \autoref{tab:load}), and assume that items 1-6 load exclusively on the externalising factor and items 7-11 on the internalising factor.\footnote{The dedicated factor structure corresponds to a sparse loading matrix, i.e.: $$\bm{\Lambda}_{c} \coloneqq \begin{bmatrix} \lambda_{1c}, \dots, \lambda_{6c} & \bm{0} \\ \bm{0} & \lambda_{7c}, \dots, \lambda_{11c} \end{bmatrix}.$$} This choice of using a dedicated measurement system is consistent with the standard rule of thumb for selecting items based on the magnitude of their loadings, which is indicated as 0.32 by \citet{Tabachnick2001}.
+We mirror the structure found in the exploratory factor analysis above, and assume that all items load on one factor for the 1-factor solution (\autoref{tab:load_add}), and that items 1-6 load exclusively on the externalising factor and items 7-11 on the internalising factor for the 2-factor solution (\autoref{tab:load}).\footnote{The dedicated factor structure in the two-factor case corresponds to a sparse loading matrix, i.e.: $$\bm{\Lambda}_{c} \coloneqq \begin{bmatrix} \lambda_{1c}, \dots, \lambda_{6c} & \bm{0} \\ \bm{0} & \lambda_{7c}, \dots, \lambda_{11c} \end{bmatrix}.$$}
 
@@
 Any comparison between socio-emotional skills across the two cohorts requires that the measures at our disposal have the same relationship with the latent constructs of interest in both cohorts.
-In other words, the items in our new scale must measure externalising and internalising socio-emotional skills in the same way in the BCS and MCS data.
+In other words, the items in our new scale must measure socio-emotional skills in the same way in the BCS and MCS data.
 This property is denominated measurement invariance (MI) \citep{Vandenberg2000a,Putnick2016}.
@@
 The hierarchy of the nested models usually proceeds by testing loadings first, and then intercepts (to establish \emph{metric} and \emph{scalar} invariance -- see \citealp{Vandenberg2000a}).
-
 Invariance of systems with categorical measures, such as the scale we examine in this paper, is less well understood.
@@
 In any case, we present a range of fit indices to provide a more complete assessment of measurement invariance.
-The fit of each nested model is compared in Panel A of \autoref{tab:fit}.
-The overall fit of the configural model is satisfactory according to all indices, with CFI around $.95$ and RMSE just above $.05$.
+We present the measurement invariance results for the 1-factor model in \autoref{tab:fit_1f}, and those for the 2-factor model in \autoref{tab:fit}.
+First, by comparing the fit of each nested model across the 1-factor and the 2-factor models, it is clear that the 1-factor model fits the data significantly worse than the 2-factor model, according to all the criteria considered.\footnote{It is  worth noting that threshold and loading invariance only can be established also in the 1-factor case, i.e.
+intercept invariance is never achieved.} Hence, in our analysis since now on, we adopt the two-factor solution, which is also consistent with the child psychology literature cited above: as mentioned above, we name the two factors externalizing and internalizing skills.
+We now examine the measurement invariance properties of our chosen two-factor solution in greater details.
+Looking at Panel A of \autoref{tab:fit}, we see that the overall fit of the configural model for the chosen 2-factor solution is satisfactory according to all indices, with CFI around $.95$ and RMSE just above $.05$.
 As expected, given our large sample size, $\chi^2$-based tests reject measurement invariance at all levels.
@@
 Invariance of loadings and thresholds across cohorts implies that the items in our scale are equally salient in their informational content, and that the latent propensities have equal mapping into the observed items.
-
-\textbf{[ADD NEW RESULTS]}
 

### 2020-03-14 codes/R/paper/seskills_jpube_rev.Rnw
update with external RIF estimates

@@
 \usepackage{authblk}
-\title{Inequality in socioemotional skills: a cross-cohort comparison}
+\title{Inequality in socio-emotional skills: a cross-cohort comparison}
 \author{Orazio Attanasio \hspace{10pt} Richard Blundell \\  Gabriella Conti \hspace{10pt} Giacomo Mason}
-\affil{\emph{Institute for Fiscal Studies} \\ \emph{Department of Economics, University College London}}
-\date{\today}
+%\affil{\emph{Institute for Fiscal Studies} \\ \emph{Department of Economics, University College London}}
+\date{22 February 2020}
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 We examine changes in inequality in socio-emotional skills very early in life in two British cohorts born 30 years apart.
-We construct socio-emotional scales comparable across cohorts for both boys and girls, using two validated instruments for the measurement of child behaviour.
-We identify two dimensions of socio-emotional skills for each cohort: `internalising' and `externalising', related to the ability of children to focus their concentration and to engage in interpersonal activities, respectively.
+We construct comparable scales using two validated instruments for the measurement of child behaviour and identify two dimensions of socio-emotional skills: `internalising' and `eternalising'.
 Using recent methodological advances in factor analysis, we establish comparability in the inequality of these early skills across cohorts, but not in their average level.
-We document for the first time that inequality in these early skills has increased across cohorts, especially for boys and at the bottom of the distribution.
-We also document changes in conditional skills gaps across cohorts.
-We find an increase in the socio-emotional skills gap in the younger cohort for children born to mothers with higher socio-economic status (education and employment), and to mothers who smoked during pregnancy.
-The increase in inequality in early socio-emotional skills is particularly pronounced for boys.
-On the other hand, we find a decline in the skills gradient for children without a father figure in the household.
-Lastly, we document that socio-emotional skills measured at a much earlier age than in most of the existing literature are significant predictors of outcomes both in adolescence and adulthood, in particular health and health behaviours.
-Our results show the importance of formally testing comparability of measurements to study skills differences across groups, and in general point to the role of inequalities in the early years for the accumulation of health and human capital across the life course.
-
+We document for the first time that inequality in socio-emotional skills has increased across cohorts, especially for boys and at the bottom of the distribution.
+We also formally decompose the sources of the increase in inequality and find that compositional changes explain half of the rise in inequality in externalising skills.
+On the other hand, the increase in inequality in internalising skills seems entirely driven by changes in returns to background characteristics.
+Lastly, we document that socio-emotional skills measured at an earlier age than in most of the existing literature are significant predictors of health and health behaviours.
+Our results show the importance of formally testing comparability of measurements to study skills differences across groups, and in general point to the role of inequalities in the early years for the accumulation of health and human capital across the life course.\\
  Keywords: Inequality, Socio-emotional skills, Cohort studies, Measurement invariance
-
 JEL Classification: J13, J24, I14, I24, C38
@@
 \vspace{20pt}
-
-\scriptsize{Corresponding Author: Orazio Attanasio, Department of Economics, University College London,
-Gordon Street, London WC1H0AX, UK.
-Email: o.attanasio@ucl.ac.uk.
+\scriptsize{Corresponding Author: Gabriella Conti, Department of Economics, University College London, Gordon Street, London WC1H0AX, UK.
+Email: gabriella.conti@ucl.ac.uk.
 We thank participants to the NBER/LSE Trans-Atlantic Public Economics Seminar (TAPES) and our two discussants Hilary Hoynes and Gabriel Ulyssea for excellent comments.
@@
 We would like to thank all cohort members and their families, who generously gave their time to participate in the surveys.
-This project has received funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme (grant agreement No 695300 - HKADeC - ERC-2015-AdG/ERC-2015-AdG and grant agreement No 819752 - DEVORHBIOSHIP - ERC-2018COG).
+This project has received funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme (grant agreement No 695300 - HKADeC - ERC-2015-AdG and grant agreement No 819752 - DEVORHBIOSHIP - ERC-2018COG).
 We also thank the DynaHEALTH consortium (H2020 - 633595 DynaHEALTH) for funding and support.
-GM is funded by an Economic and Social Research Council scholarship.}
+GM was funded by an Economic and Social Research Council scholarship.}
 \end{titlepage}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \input{text/paper_jpube_1introlit.tex}
@@
 \input{text/paper_jpube_3resultsconclusion.tex}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 read_chunk(here("codes/R/8_regs_outc.R"))
-read_chunk(here("codes/R/9_rif.R"))
 read_chunk(here("codes/R/99_auxfun.R"))
@@
 # make item-level inequality graphs
-@
-
-<<RIF_PLOTS_PREP, dependson="SELECTDATA">>=
-# make RIF decomposition plots
 @
@@
 The \emph{estimation sample} is the subsample used in the analysis.
-The \emph{full sample} is the entire sample of non-multiparous children in both cohorts, residing in England at birth.
+The \emph{full sample} is the entire sample of children in both cohorts, residing in England at birth.
 Mean estimates for the full sample in the MCS cohort are weighted to account for survey design.
@@
 \end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{table}
@@
 @
-\justify {\scriptsize \emph{Notes}: The table shows differences between quantiles of the distribution of socioemotional skills, by gender and cohort.
+\justify {\scriptsize \emph{Notes}: The table shows differences between quantiles of the distributions of socio-emotional skills, by gender and cohort.
 Bootstrap confidence intervals with 1,000 repetitions are in brackets.
-The distribution is a factor score obtained from the factor model in \autoref{sec:methods}.
+The factor scores for socio-emotional skills are estimated using an empirical Bayes modal approach, using the parameter estimates from the factor model in Table \ref{tab:measpars}.
 These distributions are shown in \autoref{fig:dens}.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{landscape}
@@
 @
-\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of children's socioemotional skills at five years of age on family background information.
+\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of children's socio-emotional skills at five years of age on family background characteristics.
 The dependent variable is a factor score obtained from the factor model in \autoref{sec:methods}.
@@
 See \autoref{tab:harmvar} for a description of the variables used.
+$^{***}$p$\leq$0.01, $^{**}$p$\leq$0.05, $^{*}$p$\leq$0.1.
 \par}
@@
 @
-\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of cohort members' adolescent outcomes on their externalising and internalising socioemotional skills at five years of age.
+\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of cohort members' adolescent outcomes on their externalising and internalising socio-emotional skills at five years of age.
 Col.
@@
 (4) to (6) repeat for female cohort members.
-All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
+All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in the household), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
 See \autoref{tab:harmvar} for a description of the variables used.
-\par}
-\end{table}
+$^{***}$p$\leq$0.01, $^{**}$p$\leq$0.05, $^{*}$p$\leq$0.1.
+\par}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 @
-\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of BCS cohort members' adult outcomes on their externalising and internalising socioemotional skills at five years of age.
+\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of BCS cohort members' adult outcomes on their externalising and internalising socio-emotional skills at five years of age.
 Col.
@@
 (4) to (6) repeat for female cohort members.
-All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
+All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in the household), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
 See \autoref{tab:harmvar} for a description of the variables used.
+$^{***}$p$\leq$0.01, $^{**}$p$\leq$0.05, $^{*}$p$\leq$0.1.
 \par}
@@
 \caption{Distribution of factor scores}\label{fig:dens}
-\justify {\scriptsize \emph{Notes}: The figure shows the distribution of the externalising and internalising socioemotional skills scores at age five obtained from the factor model, by gender and cohort.
-The scores are estimated from parameter estimates in \autoref{tab:measpars}, using an Empirical Bayes Modal approach.
+\justify {\scriptsize \emph{Notes}: The figure shows the distribution of the externalising and internalising socio-emotional skills scores at age five obtained from the factor model, by gender and cohort.
+The scores are estimated from the parameter estimates in \autoref{tab:measpars}, using an Empirical Bayes Modal approach.
 Higher scores correspond to \emph{better} skills.
 The distribution is estimated nonparametrically, using an Epanechnikov kernel.
-The figure also reports the $p$ value from Kolmogorov-Smirnov tests of equality between the distribution in BCS and MCS.
+The figure also reports the $p$-value from Kolmogorov-Smirnov tests of equality between the distribution in BCS and MCS.
 \par}
@@
 \caption{Skill inequality by mother's education}\label{fig:ineqps}
-\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socioemotional skills scores by gender, cohort, and mother's education at age five.
+\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socio-emotional skills scores by gender, cohort, and mother's education at age five.
 Mother's education is a dummy for whether the mother continued schooling past the minimum leaving age, based on her date of birth.
@@
 \caption{Skill inequality by mother's pregnancy smoking}\label{fig:ineqsm}
-\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socioemotional skills scores by gender, cohort, and mother's pregnancy smoking.
-Mother's education is a dummy for whether the mother reported smoking during pregnancy.
+\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socio-emotional skills scores by gender, cohort, and mother's pregnancy smoking.
+Maternal smoking is a dummy for whether the mother reported smoking during pregnancy.
 The four panels on top present mean and 95\% confidence intervals.
@@
 \caption{Skill inequality by father's occupation}\label{fig:ineqfsc}
-\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socioemotional skills scores by gender, cohort, and father's occupation at age five.
-Father's occupation is based on Registrar General's social class, with classes I to III Non Manual being `White collar' and classes III Manual to V (plus `other') being `Blue collar'.
+\justify {\scriptsize \emph{Notes}: The figure shows unconditional mean values of socio-emotional skills scores by gender, cohort, and father's occupation at age five.
+Father's occupation is based on the Registrar General's social class, with classes I to III Non Manual being `White collar' and classes III Manual to V (plus `other') being `Blue collar'.
 `No father figure' is defined as absence of a male figure living in the household.
@@
 \begin{figure}
-<<RIF_PLOTS, dependson="RIF_PLOTS_PREP", dev='tikz', fig.height=8, fig.width=10, out.width='\\linewidth'>>=
-gridExtra::grid.arrange(rif_plots[[1]],rif_plots[[2]], ncol=1, nrow=2)
-@
-\caption{RIF decomposition of quantile differences}\label{fig:rif}
-\justify {\scriptsize \emph{Notes}: 
+<<RIF_PLOTS, dev='tikz', fig.height=8, fig.width=10, out.width='\\linewidth'>>=
+rif_estimates <- read.csv(here("rif_estimates", "rif_m.csv")) %>%
+  mutate(effect = factor(effect, levels = c("Total Difference", "Composition Effect", "Coefficient Effect"))) %>%
+  mutate(statistic = factor(statistic, levels = c("DQ 90-10", "DQ 75-25", "DQ 50-10", "DQ 90-50", "D Variance")))
+
+ggplot(rif_estimates, aes(fill=effect, y=value, x=statistic)) + 
+    geom_bar(position="dodge", stat="identity") +
+    geom_text(aes(label=significance), position=position_dodge(width=0.9)) +
+    geom_hline(yintercept=0) +
+  facet_wrap(~ skill, ncol = 1) +
+  scale_fill_brewer(palette="OrRd") +
+
+  theme(
+    strip.background = element_blank(),
+    strip.placement = "outside",
+    axis.title.y = element_blank(),
+    axis.title.x = element_blank(),
+    legend.position = "bottom",
+    legend.title = element_blank(),
+    legend.spacing.x = unit(.3, 'cm'),
+    legend.box.just = "center"
+  )
+
+@
+\caption{RIF decomposition of changes in measures of inequality in socio-emotional skills - Males}\label{fig:rif}
+\justify {\scriptsize \emph{Notes}: The figures show the total changes in five measures of inequality in socio-emotional skills between the BCS and the MCS, and decomposes them in composition and coefficient effects, following the RIF decomposition (with reweighting) of \citet{firpo2018decomposing}.
+The top figure presents the decompositions for the externalizing skills score, and the bottom figure for the internalizing skills score.
+The five inequality measures considered are the quantile differences 90-10, 75-25, 50-10, 90-50, and the variance.
+Full results are in \autoref{tab:RIF-EXT-M} and \autoref{tab:RIF-INT-M}.
+Bootstrapped standard errors over the entire procedure (500 replications) were used to compute the $p$-values.
+$^{***}$p$\leq$0.01, $^{**}$p$\leq$0.05, $^{*}$p$\leq$0.1.
 \par}
@@
 \input{text/paper_jpube_5apptables.tex}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\begin{table}
-\caption{Loadings from exploratory factor analysis -- 11-item scale}\label{tab:load}
-\centering\setstretch{1.2}\footnotesize
-<<EFALOADTAB, dependson='MCSEFARES', results='asis'>>=
-# combine results
-library(kableExtra)
-loadtab <- cbind(meantab$num, meantab$title, data.frame(cbind(bcsfa$loadings, mcsfa$loadings)))
-colnames(loadtab) <- c("item", "title", "bcs1", "bcs2", "mcs1", "mcs2")
-
-loadtab %>%
-  mutate_at(vars(matches("cs")), round, 3) %>%
-  mutate(
-    bcs1 = cell_spec(bcs1, "latex", bold = ifelse(abs(bcs1) > .3, TRUE, FALSE)),
-    bcs2 = cell_spec(bcs2, "latex", bold = ifelse(abs(bcs2) > .3, TRUE, FALSE)),
-    mcs1 = cell_spec(mcs1, "latex", bold = ifelse(abs(mcs1) > .3, TRUE, FALSE)),
-    mcs2 = cell_spec(mcs2, "latex", bold = ifelse(abs(mcs2) > .3, TRUE, FALSE))
-  ) %>%
-  kable("latex", escape = F, booktabs = T, linesep = "", align=c('c','l','c','c','c'),
-        col.names = c("Item", "Title", "Factor 1 (EXT)", "Factor 2 (INT)", "Factor 1 (EXT)", "Factor 2 (INT)")) %>%
-  add_header_above(c(" " = 2,  "BCS (1970)" = 2, "MCS (2000/1)" = 2))
-
-@
-\justify {\scriptsize \emph{Notes}: The table displays the factor loadings obtained from exploratory factor analysis (EFA) on our novel scale, separately by cohort.
-The EFA is performed decomposing the polychoric correlation matrix of the items and using weighted least squares, and the solution is rescaled using oblique factor rotation (\emph{oblimin}).
-We use the \textbf{\textsf{R}} package \texttt{psych}, version 1.8.4 \citep{Revelle2018}.
-\par}
-\end{table}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\caption{Loadings from exploratory factor analysis with 2 factors -- 11-item scale}\label{tab:load}
+\centering\setstretch{1.2}\footnotesize
+<<EFALOADTAB, dependson='MCSEFARES', results='asis'>>=
+# combine results
+library(kableExtra)
+loadtab <- cbind(meantab$num, meantab$title, data.frame(cbind(bcsfa$loadings, mcsfa$loadings)))
+colnames(loadtab) <- c("item", "title", "bcs1", "bcs2", "mcs1", "mcs2")
+
+loadtab %>%
+  mutate_at(vars(matches("cs")), round, 3) %>%
+  mutate(
+    bcs1 = cell_spec(bcs1, "latex", bold = ifelse(abs(bcs1) > .3, TRUE, FALSE)),
+    bcs2 = cell_spec(bcs2, "latex", bold = ifelse(abs(bcs2) > .3, TRUE, FALSE)),
+    mcs1 = cell_spec(mcs1, "latex", bold = ifelse(abs(mcs1) > .3, TRUE, FALSE)),
+    mcs2 = cell_spec(mcs2, "latex", bold = ifelse(abs(mcs2) > .3, TRUE, FALSE))
+  ) %>%
+  kable("latex", escape = F, booktabs = T, linesep = "", align=c('c','l','c','c','c'),
+        col.names = c("Item", "Title", "Factor 1 (EXT)", "Factor 2 (INT)", "Factor 1 (EXT)", "Factor 2 (INT)")) %>%
+  add_header_above(c(" " = 2,  "BCS (1970)" = 2, "MCS (2000/1)" = 2))
+
+@
+\justify {\scriptsize \emph{Notes}: The table displays the factor loadings obtained from exploratory factor analysis (EFA) on our novel scale, separately by cohort.
+The EFA is performed decomposing the polychoric correlation matrix of the items and using weighted least squares, and the solution is rescaled using oblique factor rotation (\emph{oblimin}).
+We use the \textbf{\textsf{R}} package \texttt{psych}, version 1.8.4 \citep{Revelle2018}.
+\par}
+\end{table}
+
 \begin{table}
@@
 \end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \clearpage
@@
 @
-\justify {\scriptsize \emph{Notes}: The table displays the factor loadings obtained from exploratory factor analysis (EFA) on the full set of SDQ items from the BCS.
+\justify {\scriptsize \emph{Notes}: The table displays the factor loadings obtained from exploratory factor analysis (EFA) on the full set of SDQ items from the MCS.
 Items denoted by * are used in the 11-item scale.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \begin{table}[ht]
-\caption{Measurement invariance fit comparison}\label{tab:fit}
+\caption{Measurement invariance fit comparison -- single factor}\label{tab:fit_1f}
+\centering\setstretch{1.2} \scriptsize
+<<FA_FIT_RES_1F, dependson="FA_FIT", results='asis'>>=
+require(xtable)
+
+# MAKE FIT TABLES
+fit_1f <- afitab(list(fa.c.1f, fa.tl.1f, fa.tli.1f), 
+                 c("Configural", 
+                   "Threshold + Loading Inv", 
+                   "Thr. + Load. + Intercept Inv")
+)
+
+print(xtable(fit_1f,
+                    align="L{0cm}L{.22\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
+                    digits = 4
+                    ),
+             size="\\footnotesize",
+             include.rownames=F,
+             include.colnames=F,
+             booktabs = T,
+             add.to.row = list(pos = c(list(0)),
+                              command = c(
+                              # header
+            "& & \\multicolumn{6}{c}{\\textbf{Absolute fit}} & \\multicolumn{6}{c}{\\textbf{Relative fit}} 
+                  \\\\ \\cmidrule(lr){3-8} \\cmidrule(lr){9-14}
+            Model & Num. par. & $\\chi^2$ & RMSE & SRMR & MFI & CFI & G-hat & $\\chi^2$ $p$ & \\textDelta RMSE & \\textDelta SRMR &  \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ 
+& (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8) & (9) & (10) & (11) & (12) & (13) \\\\ \\midrule"
+                              )),
+             sanitize.text.function = function(x){x},
+             sanitize.rownames.function = function(x){""}, # to not print row numbers
+             tabular.environment = "tabular", width = "\\textwidth",
+             floating = FALSE
+)
+
+@
+\justify {\scriptsize \emph{Notes}: The table presents fit indices for models of different invariance levels, following \cite{Wu2016a} Col.
+(1) displays the number of estimated parameters for each model.
+Col.
+(2) and (8) present the value of the $\chi^2$ statistic and the $p$-value of the test of equality with respect to the configural model.
+Col.
+(3)-(7) and (9)-(13) present alternative fit indices (AFIs), in absolute values and differences from the configural model respectively.
+RMSE = Root mean squared error of approximation; SRMR = standardised root mean residual; MFI = McDonald non-centrality index; CFI = comparative fit index; G-hat = gamma-hat.
+\par}
+\end{table}
+\end{landscape}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{landscape}
+\begin{table}[ht]
+\caption{Measurement invariance fit comparison -- two factors}\label{tab:fit}
 \centering\setstretch{1.2} \scriptsize
@@
 Col.
-(2) and (8) present the value of the $\chi^2$ statistic and the pvalue of the test of equality with respect to the configural model.
+(2) and (8) present the value of the $\chi^2$ statistic and the $p$-value of the test of equality with respect to the configural model.
 Col.
@@
 \end{landscape}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\begin{landscape}
-\begin{table}[ht]
-\caption{Measurement invariance fit comparison -- single factor}\label{tab:fit_1f}
-\centering\setstretch{1.2} \scriptsize
-<<FA_FIT_RES_1F, dependson="FA_FIT", results='asis'>>=
-require(xtable)
-
-# MAKE FIT TABLES
-fit_1f <- afitab(list(fa.c.1f, fa.tl.1f, fa.tli.1f), 
-                 c("Configural", 
-                   "Threshold + Loading Inv", 
-                   "Thr. + Load. + Intercept Inv")
-)
-
-print(xtable(fit_1f,
-                    align="L{0cm}L{.22\\textwidth}C{.08\\textwidth}C{.08\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.07\\textwidth}C{.07\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
-                    digits = 4
-                    ),
-             size="\\footnotesize",
-             include.rownames=F,
-             include.colnames=F,
-             booktabs = T,
-             add.to.row = list(pos = c(list(0)),
-                              command = c(
-                              # header
-            "& & \\multicolumn{6}{c}{\\textbf{Absolute fit}} & \\multicolumn{6}{c}{\\textbf{Relative fit}} 
-                  \\\\ \\cmidrule(lr){3-8} \\cmidrule(lr){9-14}
-            Model & Num. par. & $\\chi^2$ & RMSE & SRMR & MFI & CFI & G-hat & $\\chi^2$ $p$ & \\textDelta RMSE & \\textDelta SRMR &  \\textDelta MFI & \\textDelta CFI & \\textDelta G-hat \\\\ 
-& (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8) & (9) & (10) & (11) & (12) & (13) \\\\ \\midrule"
-                              )),
-             sanitize.text.function = function(x){x},
-             sanitize.rownames.function = function(x){""}, # to not print row numbers
-             tabular.environment = "tabular", width = "\\textwidth",
-             floating = FALSE
-)
-
-@
-\justify {\scriptsize \emph{Notes}: The table presents fit indices for models of different invariance levels, following \cite{Wu2016a} Col.
-(1) displays the number of estimated parameters for each model.
-Col.
-(2) and (8) present the value of the $\chi^2$ statistic and the pvalue of the test of equality with respect to the configural model.
-Col.
-(3)-(7) and (9)-(13) present alternative fit indices (AFIs), in absolute values and differences from the configural model respectively.
-RMSE = Root mean squared error of approximation; SRMR = standardised root mean residual; MFI = McDonald non-centrality index; CFI = comparative fit index; G-hat = gamma-hat.
-\par}
-\end{table}
-\end{landscape}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
                       & \\multicolumn{4}{c}{\\textbf{Covariance}}
-                      & \\multicolumn{2}{c}{Correlation} \\\\
+                      & \\multicolumn{2}{c}{\\textbf{Correlation}} \\\\
                       \\cmidrule(lr){2-3} \\cmidrule(lr){4-7} \\cmidrule(lr){8-9}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% RIF decomposition tables
+\input{text/paper_jpube_riftables.tex}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{table}
@@
 (5) to (8) repeat for female cohort members.
-All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
+All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in the household), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
 See \autoref{tab:harmvar} for a description of the variables used.
-\par}
-\end{table}
+$^{***}$p$\leq$0.01, $^{**}$p$\leq$0.05, $^{*}$p$\leq$0.1.
+\par}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\centering\footnotesize
+\caption{Predictors of adolescent outcomes, Maximum Likelihood Estimates}\label{tab:adolML}
+\begingroup\footnotesize
+\begin{tabular}{L{.3\textwidth}C{.075\textwidth}C{.075\textwidth}}
+  \toprule
+   & \multicolumn{1}{c}{\textbf{Males}} & \multicolumn{1}{c}{\textbf{Females}} \\ 
+%\cmidrule(lr){2-4}  \cmidrule(lr){5-7}
+%& Mean & \multicolumn{2}{c}{Coefficients} & Mean & \multicolumn{2}{c}{Coefficients} \\ \cmidrule(lr){3-4}  \cmidrule(lr){6-7}
+ %           & (1) & (2) & (3) & (4) & (5) & (6) \\ \midrule
+\textit{\textbf{Tried smoking (BCS - 16)}} &   \\ 
+  Externalising skills (5) &  $-.128^{***}$ \newline ($.042$) &  $-.157^{***}$ \newline ($.039$) \\ 
+  Internalising skills (5) &  $.084^{*}$ \newline ($.046$) & $.080^{**}$ \newline ($.035$)  \\ 
+  Observations &  1575 &  2133 \\ 
+    \midrule\textit{\textbf{BMI (BCS - 16)}} &  \\ 
+  Externalising skills (5) &  $-.211^{**}$ \newline ($.093$) & $-.158$ \newline ($.103$)  \\ 
+  Internalising skills (5) &  $.071$ \newline ($.104$)  & $.127^{*}$ \newline ($.167$)  \\ 
+  Observations &  2011 & 2295  \\ 
+    \midrule\textit{\textbf{Tried smoking (MCS - 14)}} &  \\ 
+  Externalising skills (5) &   $-.063^{**}$ \newline ($.028$)   & $-.049$ \newline ($.034$)  \\ 
+  Internalising skills (5) &   $.010$ \newline ($.043$)   & $.058$ \newline ($.056$)  \\ 
+  Observations &   1997 &  2024  \\ 
+    \midrule\textit{\textbf{BMI (MCS - 14)}} &   \\ 
+  Externalising skills (5) &   $-.269^{**}$ \newline ($.133$)  & $-.439^{***}$ \newline ($.153$) \\ 
+  Internalising skills (5) &   $.091$ \newline ($.201$)   & $.510^{**}$ \newline ($.246$)  \\ 
+  Observations &   2005 &  1936  \\ 
+   \bottomrule
+\end{tabular}
+\endgroup
+\justify {\scriptsize \emph{Notes}: The table shows coefficients from the joint estimation of the (partially invariant) measurement system for externalising and internalising socio-emotional skills at five years of age and of the cohort members' adolescent outcomes.
+Estimation is by maximum likelihood, see \autoref{tab:adol} for the corresponding results obtained via the two-step process.
+All estimates control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in the household), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
+See \autoref{tab:harmvar} for a description of the variables used.
+$^{***}$p$\leq$0.01, $^{**}$p$\leq$0.05, $^{*}$p$\leq$0.1.
+\par}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 @
-\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of cohort members' adolescent and adult outcomes on their externalising and internalising socioemotional skills at five years of age.
+\justify {\scriptsize \emph{Notes}: The table shows coefficients from linear regressions of cohort members' adolescent and adult outcomes on their externalising and internalising socio-emotional skills at five years of age.
 Col.
@@
 (5) to (8) repeat for female cohort members.
-All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in HH), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
+All estimates additionally control for region of birth, maternal education (5), maternal employment (5), father occupation (5), maternal background (age, height, nonwhite ethnicity, number of children in the household), pregnancy (firstborn child, number of previous stillbirths, mother smoked in pregnancy, preterm birth, (log) birth weight).
 See \autoref{tab:harmvar} for a description of the variables used.
+$^{***}$p$\leq$0.01, $^{**}$p$\leq$0.05, $^{*}$p$\leq$0.1.
 \par}
@@
 Zero points are assigned for `Certainly Applies / True' responses, one point for `Sometimes applies / somewhat true', and two points for `Doesn't apply'.
-Only 0 or 1 points are assigned for items that are coded as having two categories (5,6, and 11).
+Only 0 or 1 points are assigned for items that are coded as having two categories (5 and 11).
 Higher scores correspond to \emph{better} skills.

### 2020-03-14 codes/R/5_sumstats.R

@@ -29,7 +29,7 @@ sumlabs <- c(
              "Mother age", "Mother height (m)", "Unmarried", "Nonwhite child",
              "Firstborn child", "Number previous stillbirths", "Mother smoked in pregnancy", "Preterm birth",
              "Missing gest. age", "Birthweight (kg)",
-             "Number of children in HH",
+             "Number of children in the household",
              "Mother has post-compulsory education", "Mother is employed",
              "Father occupation: blue collar", "No father figure" 
 )
